### 2016-05-26 sftrends_manu.Rnw
abstract and schematic figure

\documentclass[letterpaper,12pt,oneside]{article}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{indentfirst}
\usepackage{titlesec}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{outlines}
\usepackage{lineno}
\usepackage{array}
\usepackage{times}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage[position=t]{subfig}
\usepackage{paralist}
\usepackage[noae]{Sweave}
\usepackage{natbib}
\usepackage{array}
\usepackage{pdflscape}
\usepackage{bm}
% \usepackage{showlabels}
\bibpunct{(}{)}{,}{a}{}{,}

% page margins and section title formatting
\linespread{1.5}
\setlength{\footskip}{0.5in}
\titleformat*{\section}{\Large\bf\em}
\titleformat*{\subsection}{\singlespace\large\bf}
\titleformat*{\subsubsection}{\singlespace\normalsize\bf\em}
\titlespacing{\section}{0in}{0in}{0in}
\titlespacing{\subsection}{0in}{0in}{0in}
\titlespacing{\subsubsection}{0in}{0in}{0in}

% cleveref options
\crefname{table}{Table}{Tables}
\crefname{figure}{Fig.}{Figs.}
\renewcommand{\figurename}{Fig.}

% aliased citations
% \defcitealias{FLDEP12}{FLDEP 2012}

%acronyms
\acrodef{sfe}[SFE]{San Francisco Estuary}
\acrodef{wrtds}[WRTDS]{Weighted Regressions on Time, Discharge, and Season}

%for supplemental figures/tables
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

%knitr options
<<setup,include=F,cache=F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign = TRUE, width = 90, digits = 1)

source('R/funcs.r')
library(WRTDStidal)
@

% get the version based on commit date
<<echo = FALSE, cache = FALSE>>=
raw <- system('git log -1', intern = TRUE)
raw <- raw[grep('^Date', raw)]
raw <- paste('Version', raw)
@

% get online bib file
<<echo = FALSE, cache = FALSE>>=
refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
refs <- rawToChar(refs$content)
writeLines(refs, con = file('refs.bib'))
@

\begin{document}

\raggedbottom
\linenumbers
\raggedright
\urlstyle{same}
\setlength{\parindent}{0.5in}
\renewcommand\refname{References \vspace{12pt}}

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality changes in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ \Sexpr{raw}
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\begin{abstract}
\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of long-term time series.  Combined with these methods, multi-decadal datasets of water quality in coastal systems can provide valuable opportunities to gain insight into ecosystem properties and drivers of change.  This study describes the application of an estuarine adaptation of the \ac{wrtds} approach to describe water quality trends in the upper region of the \ac{sfe}, a primary source of nutrients into the system.  This statistical technique is data-driven where the parameterization of the functional model changes smoothly over time following dynamic patterns of season and flow.  By doing so, changes over time that have not been previously quantified can be described, including variation in flow-normalized concentrations, frequency occurrence of extreme events, and response to historical changes in the watershed, all of which are important needs for understanding trends in the \ac{sfe}.  The goal of the analysis is to apply the WRTDS model at multiple stations in the Delta and Suisun Bay regions of \ac{sfe} to provide novel descriptions of water quality changes over time and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  This variation is considered in the context of varying contributions of input flows from the Sacramento and San Joaquin rivers, as well as tidal exchange with the central \ac{sfe}.  The results are described in the context of conceptual relationships between water quality and drivers of change to generate hypotheses of mechanistic relationships using selected examples from the trend descriptions. Overall, this analysis provides an ecological and management-based understanding of historical trends in the \ac{sfe} as a means to interpret potential impacts of recent changes and expected trends in this dynamic system.  An argument is also made for more comprehensive evaluations of long-term monitoring datasets to understand relationships between response endpoints and causal mechanisms in coastal waters.
\end{abstract}
\acresetall

\section{Introduction}

\begin{itemize}
\item Paragraph 1 - how and why are trends interpreted - raw data, surrogates, various methods, what have been implications of using different approaches
\item Paragraph 2 - WRTDS for estuaries, application to Tampa Bay as test set, further validation in Patuxent
\item Paragraph 3 - SF estuary, unique and prominent location, full story is not known (historical context and recent changes), no one has empirically described the data, how is this related to the delta (a vigorous biogeochemical reactor)
\item Paragraph 4 - Study goal and objectives
\begin{itemize}
\item Provide a description of trends - annual, seasonal, spatial, response to flow, change by analytes
\item Detailed description of selected sites in the context of conceptual relationships - 1) nonlinear or extreme quantile changes site TBD, 2) P8 and WWTP improvements, 3) Suisun DIN, SiO2, Chla, and clams
\item What this means for understanding other systems
\end{itemize}
\end{itemize}

\section{Methods}

\begin{itemize}
\item Study location and data - describe conceptual relationships
\item WRTDS method - estuarine adaptation, how was it applied to \ac{sfe}
\end{itemize}

\section{Results}

\section{Discussuion}

What additional information is needed?

method leads to hypothesis generation or mechanistic explanations of change, verification of conceptual links

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/schematic.pdf}
\caption{schematic}
\label{fig:schematic}
\end{figure}

\end{document}

### 2016-05-26 sftrends_manu.tex
abstract and schematic figure

--- 2016-05-26 sftrends_manu.Rnw
@@
-\documentclass[letterpaper,12pt,oneside]{article}
+\documentclass[letterpaper,12pt,oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
@@
 %knitr options
-<<setup,include=F,cache=F>>=
-library(knitr)
-# set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
-options(replace.assign = TRUE, width = 90, digits = 1)
-
-source('R/funcs.r')
-library(WRTDStidal)
-@
 
 % get the version based on commit date
-<<echo = FALSE, cache = FALSE>>=
-raw <- system('git log -1', intern = TRUE)
-raw <- raw[grep('^Date', raw)]
-raw <- paste('Version', raw)
-@
 
 % get online bib file
-<<echo = FALSE, cache = FALSE>>=
-refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
-refs <- rawToChar(refs$content)
-writeLines(refs, con = file('refs.bib'))
-@
 
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
   \vspace{1in} 
-  \\ \Sexpr{raw}
+  \\ Version Date:   Wed May 25 08:44:44 2016 -0500
 	}

### 2016-05-26 R/funcs.R

@@ -1,0 +1,346 @@
+######
+# get legend from an existing ggplot object
+g_legend <- function(a.gplot){
+  tmp <- ggplot_gtable(ggplot_build(a.gplot))
+  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+  legend <- tmp$grobs[[leg]]
+  return(legend)}
+
+######
+# trend maps
+# 
+# library(dplyr)
+# library(ggplot2)
+# library(WRTDStidal)
+# library(maptools)
+# library(tidyr)
+# library(purrr)
+# library(RColorBrewer)
+# library(scales)
+# library(gridExtra)
+# library(ggrepel)
+# source('R/funcs.R')
+#
+# uses delt_dat, mods_nolag, delt_map
+#
+trnd_map <- function(res, 
+  mobrks = c(-Inf, 4, 8, Inf),
+  yrbrks = c(-Inf, 1988, 2000, Inf),
+  molabs = c('JFMA', 'MJJA', 'SOND'),
+  yrlabs = c('1976-1988', '1989-2000', '2001-2012'),
+  buffx = 0.00025,
+  buffy = 0.001,
+  cols =  c('#E41A1C', '#4DAF4A'),
+  sz_rng = c(2, 11),
+  shps = c(24, 25),
+  ndivs = 6,
+  strp_fl = 'lightgrey',
+  bar = FALSE, 
+  stat_labs = TRUE,
+  leg = TRUE){
+
+  # load required data
+  data(delt_dat)
+  data(delt_map)
+  data(mods_nolag)
+
+  ##
+  # get a bounding box for the stations
+  statmeta <- select(delt_dat, Site_Code, Latitude, Longitude) %>% 
+    unique
+  
+  # bounding box
+  lims<- select(statmeta, -Site_Code) %>% 
+    apply(., 2, function(x) range(x, na.rm = TRUE))
+  lims <- c(
+    (1 + buffx) * lims[1, 2], 
+    (1 - buffy) * lims[1, 1], 
+    (1 - buffx) * lims[2, 2], 
+    (1 + buffy) * lims[2, 1]
+    )    
+  names(lims) <- c('left', 'bottom', 'right', 'top')
+  
+  ##
+  # get trends using wrtdstrnd function
+  
+  # get trends, merge with statmeta for lat/lon
+  trnds <- mutate(mods_nolag, 
+    trnd = map(mod, function(x){
[…]

### 2016-05-26 R/schem_figs.R

@@ -1,0 +1,85 @@
+
+library(dplyr)
+library(tidyr)
+library(purrr)
+library(ggplot2)
+library(WRTDStidal)
+library(gridExtra)
+data("mods_nolag")
+
+# model subset and plot subsets
+row <- 7
+mod <- mods_nolag$mod[[row]]
+ylab <-  expression(paste("DIN (mg ", L^-1, ")"))
+ylims <- as.Date(c('1976-01-01', '2013-12-31'))
+xlims <- c(0, 6)
+
+annuals <- prdnrmplot(tmp, annuals = TRUE, plot = F) %>%
+  .$nrms %>% 
+  filter(taus == 0.5)
+seasnls <- prdnrmplot(tmp, annuals = FALSE, plot = F) %>%
+  .$nrms %>% 
+  filter(taus == 0.5)
+residus <- select(mod, date, res, fit0.5) %>% 
+  mutate(resfit = exp(res) - exp(fit0.5)) %>% 
+  na.omit()
+
+# default themes
+pthms <- theme_bw() + 
+  theme(
+    axis.title.y = element_blank(),
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank()
+  )
+
+# plots
+p1 <- ggplot(na.omit(mod), aes(x = date, y = exp(res))) + 
+  geom_line() + 
+  theme_bw() + 
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank()
+  ) +
+  scale_y_continuous('DIN', limits = c(0, 4)) + 
+  scale_x_date('Time', limits = ylims)
+
+p2 <- ggplot(annuals, aes(x = date, y = exp(nrms_value))) + 
+  geom_line() + 
+  pthms +
+  scale_y_continuous(limits = c(0, 4)) + 
+  scale_x_date('Time', limits = ylims)
+
+p3 <- ggplot(seasnls, aes(x = date, y = exp(nrms_value))) + 
+  geom_line() + 
+  pthms +
+  scale_x_date('Time', limits = ylims)
+
+p4 <- ggplot(residus, aes(x = date, y = resfit)) + 
+  geom_point() + 
+  geom_hline(aes(yintercept = 0)) + 
+  pthms +
+  scale_x_date('Time', limits = ylims)
+
+p5 <- dynaplot(mod, mo = 1, logspace = F, years = c(1976, 2000, 2013), col_vec = 'black') + 
+  pthms +
+  theme(legend.position = 'none', strip.text.x = element_blank()) + 
[…]

### 2016-05-26 sftrends_manu.Rnw
manuscript outline and hypotheses

@@
 \begin{singlespace}
-\title{{\bf {\Large Four decades of water quality changes in the upper San Francisco Estuary}}}
+\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
 \author{
@@
 Combined with these methods, multi-decadal datasets of water quality in coastal systems can provide valuable opportunities to gain insight into ecosystem properties and drivers of change.
-This study describes the application of an estuarine adaptation of the \ac{wrtds} approach to describe water quality trends in the upper region of the \ac{sfe}, a primary source of nutrients into the system.
-This statistical technique is data-driven where the parameterization of the functional model changes smoothly over time following dynamic patterns of season and flow.
-By doing so, changes over time that have not been previously quantified can be described, including variation in flow-normalized concentrations, frequency occurrence of extreme events, and response to historical changes in the watershed, all of which are important needs for understanding trends in the \ac{sfe}.
-The goal of the analysis is to apply the WRTDS model at multiple stations in the Delta and Suisun Bay regions of \ac{sfe} to provide novel descriptions of water quality changes over time and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
-This variation is considered in the context of varying contributions of input flows from the Sacramento and San Joaquin rivers, as well as tidal exchange with the central \ac{sfe}.
-The results are described in the context of conceptual relationships between water quality and drivers of change to generate hypotheses of mechanistic relationships using selected examples from the trend descriptions.
+This study describes use of an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}.
+This region is a complex mosaic of inflows that are primary sources of nutrients into the larger Bay.
+To date, a comprehensive evaluation of the long-term monitoring dataset at multiple stations in the Delta has not been conducted despite the importance of nutrient transport from the region for water quality in the entire bay.
+The \ac{wrtds} technique is data-driven where the parameterization of the functional model changes smoothly over time following dynamic patterns of season and flow.
+Water quality trends that have not been previously quantified can be described, including variation in flow-normalized concentrations, frequency occurrence of extreme events, and response to historical changes in the watershed, all of which are important needs for understanding changes in the \ac{sfe}.
+Model results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
+This variation was described in the context of varying contributions of input flows from the Sacramento and San Joaquin rivers, as well as tidal exchange with the central \ac{sfe}.
+Conceptual relationships between water quality and drivers of change were used to generate and test hypotheses of mechanistic relationships using selected examples from the trend descriptions.
 Overall, this analysis provides an ecological and management-based understanding of historical trends in the \ac{sfe} as a means to interpret potential impacts of recent changes and expected trends in this dynamic system.
@@
 
-\begin{itemize}
-\item Paragraph 1 - how and why are trends interpreted - raw data, surrogates, various methods, what have been implications of using different approaches
-\item Paragraph 2 - WRTDS for estuaries, application to Tampa Bay as test set, further validation in Patuxent
-\item Paragraph 3 - SF estuary, unique and prominent location, full story is not known (historical context and recent changes), no one has empirically described the data, how is this related to the delta (a vigorous biogeochemical reactor)
-\item Paragraph 4 - Study goal and objectives
+\begin{enumerate}
+\item How and why are trends interpreted - assessment of raw data, surrogates, various methods (kendall, GAM, WRTDS), what have been implications of using different approaches
+\item WRTDS, original method
+\item WRTDS application to Tampa Bay as test set, further validation in Patuxent
+\item SF estuary, unique and prominent location, full story is eomplex (historical context and recent changes), no one has empirically described the data, how is this related to the delta (a vigorous biogeochemical reactor)
+\item Study goal and objectives
 \begin{itemize}
 \item Provide a description of trends - annual, seasonal, spatial, response to flow, change by analytes
-\item Detailed description of selected sites in the context of conceptual relationships - 1) nonlinear or extreme quantile changes site TBD, 2) P8 and WWTP improvements, 3) Suisun DIN, SiO2, Chla, and clams
+\item Detailed description of selected sites in the context of conceptual relationships - 1) nonlinear or extreme quantile changes, site TBD, 2) P8 and WWTP improvements, 3) Suisun DIN, SiO2, Chla, and clams
 \item What this means for understanding other systems
 \end{itemize}
-\end{itemize}
+\end{enumerate}
 
@@
 \begin{itemize}
-\item Study location and data - describe conceptual relationships
-\item WRTDS method - estuarine adaptation, how was it applied to \ac{sfe}
+\item Study location and data
+\item WRTDS method - conceptual diagram
+\item WRTDS application to delta
+\item case studies and why
+\begin{itemize}
+\item Hypothesis 1: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
+\item Hypothesis 2: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions.
+\item Hypothesis 3: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in stoichiometric ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
+\end{itemize}
 \end{itemize}
@@
 
-\section{Discussuion}
+\begin{itemize}
+\item Trend descriptions
+\item Case studies
+\end{itemize}
 
-What additional information is needed?
-
-method leads to hypothesis generation or mechanistic explanations of change, verification of conceptual links
+\section{Discussion}
 

### 2016-05-26 sftrends_manu.tex
manuscript outline and hypotheses

@@
   \vspace{1in} 
-  \\ Version Date:   Wed May 25 08:44:44 2016 -0500
+  \\ Version Date:   Thu May 26 10:37:57 2016 -0500
 	}

### 2016-06-02 sftrends_manu.Rnw
USGS refs to outline

@@
 \begin{enumerate}
-\item How and why are trends interpreted - assessment of raw data, surrogates, various methods (kendall, GAM, WRTDS), what have been implications of using different approaches
+\item How and why are trends interpreted - assessment of raw data, surrogates, various methods (kendall, GAM, WRTDS), what have been implications of using different approaches, see Kratzer USGS report http://pubs.usgs.gov/sir/2010/5228/pdf/sir20105228.pdf and data http://pubs.usgs.gov/sir/2010/5228/
 \item WRTDS, original method

### 2016-06-02 sftrends_manu.tex
USGS refs to outline

@@
   \vspace{1in} 
-  \\ Version Date:   Thu May 26 10:37:57 2016 -0500
+  \\ Version Date:   Thu May 26 14:03:03 2016 -0500
 	}

### 2016-06-13 sftrends_manu.Rnw
clam, chl plots

@@
 library(WRTDStidal)
+library(ggplot2)
+library(dplyr)
+library(gridExtra)
+library(grid)
+library(purrr)
+library(tidyr)
 @
@@
 \noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of long-term time series.
-Combined with these methods, multi-decadal datasets of water quality in coastal systems can provide valuable opportunities to gain insight into ecosystem properties and drivers of change.
+Combined with these methods, multi-decadal datasets of water quality in coastal systems can provide valuable opportunities to gain insights into ecosystem properties and drivers of change.
 This study describes use of an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}.
@@
 \begin{itemize}
-\item Study location and data
-\item WRTDS method - conceptual diagram
+\item Study location (Bay and Delta) and data
+\item WRTDS method - conceptual diagram \cref{fig:schematic}
 \item WRTDS application to delta
@@
 
+\clearpage
+\begin{singlespace}
+\bibliographystyle{apalike_mine}
+\bibliography{refs}
+\end{singlespace}
+\clearpage
+
 \begin{figure}
@@
 \includegraphics[width=1\textwidth,page=1]{figs/schematic.pdf}
-\caption{schematic}
+\caption{Conceptual diagram illustrating use of \ac{wrtds} to decompose trends in observed nitrogen time series and potential forcing factors that can explain model output.
+Results from the model are described as annual and seasonal trends, changes in flow-nutrient dynamics for different time periods, and residual variation independent of time, flow, and season.
+Relationships between environmental factors (climate, local, regional/historical) and nitrogen trends are more easily related to the separate components of the observed time series using results from the model.
+}
 \label{fig:schematic}   
@@
 
+<<clmchl, fig.height = 6, fig.width = 10, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
+Invasion by \\textit{Potamocorbula} clams in 1987 and displacement of \\textit{Corbicula sp.} was shown by changes in clam density (top left, annual means).
+A coincident decrease in chloropyll concentration was also observed (bottom left).
+A weak but significant ($p < 0.001$) relationship between clam biomass and chlorophyll concentration was observed (right).'>>=
+# clam and model data at D7
+data(clams)
+data(diat_dat)
+
+# format chl data and clam data for first two plots
+chldat <- filter(diat_dat, resvar %in% 'chl') %>% 
+  .$mod %>% 
+  .[[1]] %>% 
+  annual_agg %>% 
+  select(date, year, res, norm0.5) %>% 
+  mutate(res = exp(res), norm0.5 = exp(norm0.5)) %>% 
+  rename(yr = date)
+
+clmdat <- group_by(clams, yr, species) %>% 
+  summarize(
+    clams_smp = mean(clams_smp, na.rm = TRUE),
+    biomass = mean(biomass, na.rm = TRUE)
+    ) %>% 
+  ungroup %>% 
+  mutate(yr = as.Date(paste0(yr, '-10-01')))
+
+# make sure date range is same on both
+xlims <- as.Date(c('1976-10-01', '2014-10-01'))
+
+# annual clam time series
+p1 <- ggplot(clmdat, aes(x = yr, y = clams_smp, fill = species)) + 
+  geom_bar(stat = 'identity') + 
+  theme_minimal() + 
+  theme(
+    axis.line.y = element_line(), 
+    axis.line.x = element_line(),
+    # axis.title.x = element_blank(),
+    legend.position = 'top', 
+    legend.title = element_blank()
+    ) + 
+  scale_fill_brewer(
+    palette = 'Set2',
+    labels = parse(text = c('italic(Corbicula)', 'italic(Potamocorbula)'))
+    ) + 
+  scale_x_date('', limits = xlims) +
+  scale_y_continuous('Clams per sample')
+
+# annual chlorophyll time series with flow-norm
+p2 <- ggplot(chldat, aes(x = yr, y = norm0.5)) + 
+  geom_point(aes(y = res, shape = 'Observed'), size = 3) + 
+  geom_line(aes(color = 'Flow-normalized')) + 
+  theme_minimal() + 
+  theme(
+    axis.line.y = element_line(), 
+    axis.line.x = element_line(),
+    # axis.title.x = element_blank(), 
+    legend.position = 'top', 
+    legend.title = element_blank(), 
+    legend.box = 'horizontal'
+    ) + 
+  scale_colour_manual(values = c('black', 'black')) +
+  scale_x_date('', limits = xlims) +
+  scale_y_continuous(expression(paste("Chl-", italic(a), " (", italic(mu), "g ", L^-1, ")")))
+
+# chlorophyll by month
+chldat <- filter(diat_dat, resvar %in% 'chl') %>% 
+  .$mod %>% 
+  .[[1]] %>% 
+  select(year, month, res, norm0.5) %>% 
+  mutate(res = exp(res), norm0.5 = exp(norm0.5)) %>% 
+  rename(yr = year, mo = month) %>% 
+  group_by(yr, mo) %>% 
+  summarize(res = mean(res, na.rm = TRUE), norm0.5 = mean(norm0.5, na.rm = TRUE)) %>% 
+  ungroup
+
+# clam by month
+clmdat <- mutate(clams, mo = lubridate::month(date)) %>% 
+  group_by(yr, mo, species) %>% 
+  summarize(
+    clams_smp = mean(clams_smp, na.rm = TRUE),
+    biomass = mean(biomass, na.rm = TRUE)
+    ) %>% 
+  ungroup %>% 
+  filter(species %in% 'Potamocorbula')
+
+# combine chlorophyll and clam by year, month
+toplo <- left_join(clmdat, chldat, by = c('yr', 'mo')) %>% 
+  mutate(date = as.Date(paste0(yr, '-', mo, '-01')))
+
+# linear model of concentration x biomass
+mod <- lm(log(1 + norm0.5) ~ log(1 + biomass), data = toplo) %>% 
+  coefficients
+xvals <- range(toplo$biomass, na.rm = TRUE)
+xvals <- seq(xvals[1], xvals[2], length = 100)
+yvals <- (exp(mod[1] + mod[2] * log(1 + xvals))) - 1
+
+xlab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
+ylab <- expression(paste("Flow-normalized chlorophyll (   ", italic(mu), "g ", L^-1, ")"))
+
+# scatter plot of concetration x biomass with model 
+p3 <- ggplot(toplo, aes(x = biomass, y = norm0.5)) + 
+  geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
+  geom_line(data = data.frame(x = xvals, y = yvals), aes(x, y, colour = 'black'), lwd = 1) +
+  theme_minimal() + 
+  theme(
+    axis.line.y = element_line(), 
+    axis.line.x = element_line(),
+    legend.position = 'top', 
+    legend.title = element_blank(), 
+    legend.box = 'horizontal'
+    ) +
+  scale_colour_manual(label = 'log(1 + chl) = 1.1 - 0.09log(1 + biomass)', values = 'black') +
+  scale_x_continuous(xlab) + 
+  scale_y_continuous(ylab, limits = c(0.8, 4))
+
+# align widths of plots in first column
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3])
+
+# Set the widths
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+
+# combine all
+grid.arrange(
+  arrangeGrob(pA, pB, ncol = 1, heights = c(1, 1)), 
+  p3, ncol = 2, widths = c(1, 0.5)
+  )
+@
+
 \end{document}

### 2016-06-13 sftrends_manu.tex
clam, chl plots

@@
   \vspace{1in} 
-  \\ Version Date:   Thu May 26 14:03:03 2016 -0500
+  \\ Version Date:   Thu Jun 2 15:15:19 2016 -0500
 	}
@@
 
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\textwidth]{figs/clmchl-1}
+
+}
+
+\caption{Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
+Invasion by \textit{Potamocorbula} clams in 1987 and displacement of \textit{Corbicula sp.} was shown by changes in clam density (top left, annual means).
+A coincident decrease in chloropyll concentration was also observed (bottom left).
+A weak but significant ($p < 0.001$) relationship between clam biomass and chlorophyll concentration was observed (right).}\label{fig:clmchl}
+\end{figure}
+
 \end{document}

### 2016-06-14 sftrends_manu.Rnw
dynaplot june to clam plot

@@
 library(tidyr)
+library(Hmisc)
 @
@@
 \begin{enumerate}
-\item How and why are trends interpreted - assessment of raw data, surrogates, various methods (kendall, GAM, WRTDS), what have been implications of using different approaches, see Kratzer USGS report http://pubs.usgs.gov/sir/2010/5228/pdf/sir20105228.pdf and data http://pubs.usgs.gov/sir/2010/5228/
-\item WRTDS, original method
-\item WRTDS application to Tampa Bay as test set, further validation in Patuxent
-\item SF estuary, unique and prominent location, full story is eomplex (historical context and recent changes), no one has empirically described the data, how is this related to the delta (a vigorous biogeochemical reactor)
+\item How and why are trends interpreted - assessment of raw data, surrogates, various methods (kendall, GAM, WRTDS), what have been implications of using different approaches, see Kratzer USGS report http://pubs.usgs.gov/sir/2010/5228/pdf/sir20105228.pdf and data http://pubs.usgs.gov/sir/2010/5228/, need to interpret eutrophication trends in estuaries - it's confusing \citep{Cloern10}
+\item WRTDS, original method \citep{Hirsch10,Hirsch15}
+\item WRTDS application to Tampa Bay as test set \citep{Beck15}, further validation in Patuxent and other tidal waters \cite{Beck15b}
+\item SF estuary, unique and prominent location, full story is complex (historical context and recent changes) \citep{Cloern12b}, why is the delta important (a vigorous biogeochemical reactor) \citep{Jassby00,Jassby02,Jassby08}, no one has empirically described the data in the delta using data-intensive methods
 \item Study goal and objectives
@@
 
-\begin{itemize}
-\item Trend descriptions
-\item Case studies
-\end{itemize}
+\subsection{Trends}
+
+\subsection{Selected examples}
+
+\subsubsection{Clam invasion in Suisun Bay}
+Data from \citep{Crauder16}, \citet{Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
+Figure 10 in \citet{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
+\cref{fig:clmchl}, \cref{tab:chlsio}
 
@@
 
-<<clmchl, fig.height = 6, fig.width = 10, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
-Invasion by \\textit{Potamocorbula} clams in 1987 and displacement of \\textit{Corbicula sp.} was shown by changes in clam density (top left, annual means).
-A coincident decrease in chloropyll concentration was also observed (bottom left).
+<<clmchl, fig.height = 7, fig.width = 10, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
+Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (top left, annual means).
+A coincident decrease in chloropyhll concentration was also observed (bottom left).
 A weak but significant ($p < 0.001$) relationship between clam biomass and chlorophyll concentration was observed (right).'>>=
@@
 data(diat_dat)
+data(chld7)
 
@@
   annual_agg %>% 
+  mutate(year = lubridate::year(date)) %>% 
   select(date, year, res, norm0.5) %>% 
@@
 clmdat <- group_by(clams, yr, species) %>% 
-  summarize(
+  dplyr::summarize(
     clams_smp = mean(clams_smp, na.rm = TRUE),
@@
   group_by(yr, mo) %>% 
-  summarize(res = mean(res, na.rm = TRUE), norm0.5 = mean(norm0.5, na.rm = TRUE)) %>% 
+  dplyr::summarize(res = mean(res, na.rm = TRUE), norm0.5 = mean(norm0.5, na.rm = TRUE)) %>% 
   ungroup
@@
   group_by(yr, mo, species) %>% 
-  summarize(
+  dplyr::summarize(
     clams_smp = mean(clams_smp, na.rm = TRUE),
@@
 
-# align widths of plots in first column
+# dynaplot for chl, june
+cols <- RColorBrewer::brewer.pal(10, 'Set2')[1:2]
+ylab <- attr(chld7, 'reslab')
+p4 <- dynaplot(chld7, mo = 6, col_vec = cols) +
+  theme_minimal() +
+  theme(
+    axis.line.y = element_line(), 
+    axis.line.x = element_line(),
+    legend.position = 'top', 
+    legend.title = element_blank(), 
+    legend.box = 'horizontal',
+    strip.background = element_blank(),
+    strip.text.x = element_blank()
+    ) + 
+  scale_x_continuous('Salinity (standardized 0-1)') + 
+  scale_y_continuous(ylab, limits = c(0.4, 3.8))
+
+# align widths of plots in first column, first two
 pA <- ggplot_gtable(ggplot_build(p1))
@@
 
+# align widths of plots in first column, first two
+pC <- ggplot_gtable(ggplot_build(p3))
+pD <- ggplot_gtable(ggplot_build(p4))
+maxWidth = unit.pmax(pc$widths[2:3], pD$widths[2:3])
+
+# Set the widths
+pC$widths[2:3] <- maxWidth
+pD$widths[2:3] <- maxWidth
+
 # combine all
@@
   arrangeGrob(pA, pB, ncol = 1, heights = c(1, 1)), 
-  p3, ncol = 2, widths = c(1, 0.5)
+  arrangeGrob(pC, pD, ncol = 1, heights = c(1, 1)), 
+  ncol = 2, widths = c(1, 0.6)
   )
@@
 
+\clearpage
+%%%%%%
+% tables
+
+% trends in sio and chl at d7
+<<chlsio>>=
+
+data(diat_dat)
+
+sioraw <- filter(diat_dat, resvar %in% 'sio2') %>% 
+  .$mod %>% 
+  .[[1]]
+chlraw <- filter(diat_dat, resvar %in% 'chl') %>% 
+  .$mod %>% 
+  .[[1]]
+
+mobrks <- c(-Inf, 3, 6, 9, Inf)
+yrbrks <- c(-Inf, 1985, 1994, 2003, Inf)
+molabs <- c('JFM', 'AMJ', 'JAS', 'OND')
+yrlabs <- c('1976-1985', '1986-1994', '1995-2003', '2004-2013')
+
+# summarize sio2 and chl trends
+sio <- wrtdstrnd(sioraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
+sio <- wrtdstrnd(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+  .[1, ] %>% 
+  rbind(., sio) %>% 
+  data.frame(res = 'sio', .)
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+  .[1, ] %>% 
+  rbind(., chl) %>% 
+  data.frame(res = 'chl', .)
+
+# combine for trends for table
+totab <- rbind(sio, chl) %>% 
+  mutate(
+    chg = round(chg, 1), 
+    chg = ifelse(chg < 0, chg, paste0('\\textit{\\textbf{', chg , '}}')), 
+    ave = as.character(round(ave, 1))
+    ) %>% 
+  gather('var', 'val', chg:ave) %>% 
+  unite('col', res, var) %>% 
+  spread(col, val) %>% 
+  mutate(cat = factor(cat, levels = c('1976-2013', yrlabs, molabs), labels = c('1976-2013', yrlabs, molabs))) %>% 
+  arrange(cat) %>% 
+  data.frame(sep = c(rep('Annual', 5), rep('Seasonal', 4)), .)
+
+# final table formatting
+segs <- c('All', 'Annual', 'Seasonal')
+cats <- totab$cat
+totab <- totab[,-c(1:2)]
+vars <- c('Chl-\\textit{a}', 'SiO$_2$')
+names(totab) <- rep(c('Mean', '\\% change'), 2)
+
+cap.val<-'Summaries of flow-normalized trends in chlorophyll and silicon dioxide concentrations for different time periods.
+Summaries are means and percent changes based on annual means within the time periods.
+Increasing values are in bold-italics.'
+
+latex( 
+  totab,
+  file = '',
+  rowlabel = 'Period',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = segs,
+  n.rgroup = c(1, rep(4, 2)),
+  cgroup = vars,
+  n.cgroup = c(2, 2),
+  rowname = cats,
+  label = 'tab:chlsio'
+  )
+
+@
+
 \end{document}

### 2016-06-14 sftrends_manu.tex
dynaplot june to clam plot

@@
   \vspace{1in} 
-  \\ Version Date:   Thu Jun 2 15:15:19 2016 -0500
+  \\ Version Date:   Mon Jun 13 16:52:18 2016 -0500
 	}
@@
 
-\begin{figure}[!ht]
+\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unit.pmax(pc\$widths[2:3], pD\$widths[2:3]): object 'pc' not found}}\end{kframe}\begin{figure}[!ht]
 
@@
 
-\caption{Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
-Invasion by \textit{Potamocorbula} clams in 1987 and displacement of \textit{Corbicula sp.} was shown by changes in clam density (top left, annual means).
-A coincident decrease in chloropyll concentration was also observed (bottom left).
+\caption[Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay]{Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
+Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (top left, annual means).
+A coincident decrease in chloropyhll concentration was also observed (bottom left).
 A weak but significant ($p < 0.001$) relationship between clam biomass and chlorophyll concentration was observed (right).}\label{fig:clmchl}
@@
 
+\clearpage
+%%%%%%
+% tables
+
+% trends in sio and chl at d7
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:chlsio")%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in chlorophyll and silicon dioxide concentrations for different time periods.
+Summaries are means and percent changes based on annual means within the time periods.
+Increasing values are in bold-italics.\label{tab:chlsio}}
+\begin{center}
+\begin{tabular}{lllcll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
+\cline{2-3} \cline{5-6}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\hline
+{\bfseries All}&&&&&\tabularnewline
+~~1976-2013&3.5&-80.1&&12.5&\textit{\textbf{33.6}}\tabularnewline
+\hline
+{\bfseries Annual}&&&&&\tabularnewline
+~~1976-1985&7.8&-62.3&&10.5&\textit{\textbf{8.9}}\tabularnewline
+~~1986-1994&2.2&-62.1&&12.4&\textit{\textbf{15.6}}\tabularnewline
+~~1995-2003&1.6&\textit{\textbf{20.2}}&&13.7&\textit{\textbf{4.5}}\tabularnewline
+~~2004-2013&2&\textit{\textbf{31.3}}&&13.6&-2\tabularnewline
+\hline
+{\bfseries Seasonal}&&&&&\tabularnewline
+~~JFM&1.9&-43.2&&14.8&\textit{\textbf{7.3}}\tabularnewline
+~~AMJ&4.3&-72.8&&12.9&\textit{\textbf{25.7}}\tabularnewline
+~~JAS&4.8&-88.4&&11.1&\textit{\textbf{70.8}}\tabularnewline
+~~OND&2.1&-79.4&&11.7&\textit{\textbf{41.7}}\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
+
 \end{document}

### 2016-06-14 R/dat_proc.R

@@ -1,0 +1,65 @@
+######
+# some additional data processing for the figure
+
+######
+# a smoother chlorophyll model for d7 
+
+data(diat_dat)
+data(delt_dat)
+
+# resvar label lookup
+lablk <- list(
+  shrt = c('chl', 'sio2'),
+  lngs = c(
+    expression(paste('ln-chlorophyll a (ug ', L^-1, ')')),
+    expression(paste('ln-silicon dioxide (mg ', L^-1, ')'))
+    )
+  )
+
+# flovar label lookup
+stalk <- list(
+  shrt = c('sal'),
+  lngs = c('Salinity')
+  )
+
+i <- 4
+
+# data, response variable label
+dat <- diat_dat[i, ]$data[[1]]
+resvar <- diat_dat[i, ]$resvar
+flovar <- diat_dat[i, ]$flovar
+sta <- diat_dat[i, ]$Site_Code
+reslab <- with(lablk, lngs[shrt == resvar])
+flolab <- with(stalk, shrt[lngs == flovar])
+
+# prep data as tidal object
+tomod <- select(dat, Date, resval, floval, lim) %>% 
+  rename(
+    res = resval, 
+    flo = floval
+  ) %>% 
+  data.frame %>% 
+  tidal(., 
+    reslab = reslab, 
+    flolab = expression(paste('ln-flow (standardized)'))
+  )
+
+# salinity data at D7 to pred
+topred <- filter(delt_dat, Site_Code == 'D7') %>% 
+    mutate(flo = log(1 + sal)) %>% # salinity is only variable with zeroes
+    rename(date = Date) %>% 
+    select(date, flo) %>% 
+    filter(date >= min(tomod$date) & date <= max(tomod$date)) %>% 
+    na.omit %>% 
+    data.frame
+
+# create model and exit
+mod <- wrtds(tomod, tau = c(0.1, 0.5, 0.9), wins = list(0.5, 20, 0.5), flo_div = 30, min_obs = 150)
+
+# get predictions, norms from obs flow data
+out <- mod %>% 
+  respred(dat_pred = topred) %>% 
+  resnorm
+
+chld7 <- out
+save(chld7, file = 'data/chld7.RData', compress = 'xz')

### 2016-06-16 sftrends_manu.Rnw
more subsections

@@
 %acronyms
+\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
 \acrodef{sfe}[SFE]{San Francisco Estuary}
@@
 \item SF estuary, unique and prominent location, full story is complex (historical context and recent changes) \citep{Cloern12b}, why is the delta important (a vigorous biogeochemical reactor) \citep{Jassby00,Jassby02,Jassby08}, no one has empirically described the data in the delta using data-intensive methods
+
+San Francisco Bay on the Pacific Coast of the United States is one of the most prominent estuaries in the western hemisphere.
+Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.
+Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed historical trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system.
+The unique ecological and social context of the Bay, including a rich source of monitoring data from the last four decades, provides a valuable opportunity to gain insight into ecosystem properties of estuaries.
+
 \item Study goal and objectives
@@
 
-\begin{itemize}
-\item Study location (Bay and Delta) and data
-\item WRTDS method - conceptual diagram \cref{fig:schematic}
-\item WRTDS application to delta
-\item case studies and why
-\begin{itemize}
-\item Hypothesis 1: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
-\item Hypothesis 2: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions.
-\item Hypothesis 3: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in stoichiometric ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
-\end{itemize}
-\end{itemize}
+\subsection{Study location and data}
+
+The \ac{sfe}... exchange with ocean, freshwater inputs, runoff variation (drought years), climate, support of natural resources
+\subsection{Analysis method and application}
+
+\begin{equation}
+\ln\left(N\right) = \beta_0 + \beta_1 t + \beta_2 \ln\left(Q\right) + \beta_3 \sin\left(2\pi t\right) + \beta_4 \cos\left(2\pi t\right)
+\end{equation}
+
+\subsection{Case studies}
+\subsubsection{Disaggregating observed nitrogen time series}
+Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
+\subsubsection{Effects of wastewater treatment}
+Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions.
+\subsubsection{Effects of biological invasions}
+Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in stoichiometric ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
 
@@
 
-<<clmchl, fig.height = 7, fig.width = 10, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
+<<clmchl, fig.height = 7, fig.width = 10, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
 Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (top left, annual means).
-A coincident decrease in chloropyhll concentration was also observed (bottom left).
-A weak but significant ($p < 0.001$) relationship between clam biomass and chlorophyll concentration was observed (right).'>>=
+A coincident decrease in \\ac{chla} concentration was also observed (bottom left).
+A weak but significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in the top right.
+Flow relationships with \\ac{chla} concentration have also changed over time (bottom right, observations from June).
+Chlorophyll shows a slight positive then dominantly negative association with increasing flow (decreasing salinity) early in the time series, whereas the trend is reversed in recent years.'>>=
+
 # clam and model data at D7
@@
 ylab <- attr(chld7, 'reslab')
-p4 <- dynaplot(chld7, mo = 6, col_vec = cols) +
+
+p4 <- dynaplot(chld7, mo = 6, col_vec = cols, floscl = F) +
   theme_minimal() +
@@
     ) + 
-  scale_x_continuous('Salinity (standardized 0-1)') + 
+  scale_x_continuous('ln-Salinity (psu)') + 
   scale_y_continuous(ylab, limits = c(0.4, 3.8))
@@
 pD <- ggplot_gtable(ggplot_build(p4))
-maxWidth = unit.pmax(pc$widths[2:3], pD$widths[2:3])
+maxWidth = unit.pmax(pC$widths[2:3], pD$widths[2:3])
 

### 2016-06-16 sftrends_manu.tex
more subsections

@@
   \vspace{1in} 
-  \\ Version Date:   Mon Jun 13 16:52:18 2016 -0500
+  \\ Version Date:   Tue Jun 14 18:01:49 2016 -0500
 	}
@@
 
-\begin{itemize}
-\item Study location (Bay and Delta) and data
-\item WRTDS method - conceptual diagram \cref{fig:schematic}
-\item WRTDS application to delta
-\item case studies and why
-\begin{itemize}
-\item Hypothesis 1: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
-\item Hypothesis 2: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions.
-\item Hypothesis 3: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in stoichiometric ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
-\end{itemize}
-\end{itemize}
+\subsection{Study location and data}
+
+The \ac{sfe}... The delta...
+
+\subsection{Analysis method and application}
+
+\begin{equation}
+\ln\left(N\right) = \beta_0 + \beta_1 t + \beta_2 \ln\left(Q\right) + \beta_3 \sin\left(2\pi t\right) + \beta_4 \cos\left(2\pi t\right)
+\end{equation}
+
+\subsection{Case studies}
+\subsubsection{Disaggregating observed nitrogen time series}
+Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
+\subsubsection{Effects of wastewater treatment}
+Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions.
+\subsubsection{Effects of biological invasions}
+Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in stoichiometric ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
 
@@
 
-\begin{kframe}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unit.pmax(pc\$widths[2:3], pD\$widths[2:3]): object 'pc' not found}}\end{kframe}\begin{figure}[!ht]
+\begin{figure}[!ht]
 
@@
 
-\caption[Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay]{Trends in clam abundance and chlorophyll concentration from 1976 to 2014 at station D7 in Suisun Bay.
+\caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
 Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (top left, annual means).
-A coincident decrease in chloropyhll concentration was also observed (bottom left).
-A weak but significant ($p < 0.001$) relationship between clam biomass and chlorophyll concentration was observed (right).}\label{fig:clmchl}
+A coincident decrease in \ac{chla} concentration was also observed (bottom left).
+A weak but significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in the top right.
+Flow relationships with \ac{chla} concentration have also changed over time (bottom right, observations from June).
+Chlorophyll shows a slight positive then dominantly negative association with increasing flow (decreasing salinity) early in the time series, whereas the trend is reversed in recent years.}\label{fig:clmchl}
 \end{figure}

### 2016-06-16 R/dat_proc.R

@@ -1,6 +1,10 @@
 ######
 # some additional data processing for the figure
 
+library(WRTDStidal)
+library(tidyr)
+library(dplyr)
+
 ######
 # a smoother chlorophyll model for d7 
 
@@ -63,3 +67,65 @@ out <- mod %>%
 
 chld7 <- out
 save(chld7, file = 'data/chld7.RData', compress = 'xz')
+
+
+######
+# a smoother din model for c10
+data(mods_nolag)
+data(flow_dat)
+data(delt_dat)
+
+# resvar label lookup
+lablk <- list(
+  shrt = c('din', 'nh', 'no23'),
+  lngs = c(
+    expression(paste('ln-dissolved inorganic nitrogen (mg ', L^-1, ')')),
+    expression(paste('ln-ammonium (mg ', L^-1, ')')),
+    expression(paste('ln-nitrite/nitrate (mg ', L^-1, ')'))
+    )
+  )
+
+# flovar label lookup
+stalk <- list(
+  shrt = c('sjr', 'sac', 'sal'),
+  lngs = c('San Joaquin', 'Sacramento', 'Salinity')
+  )
+
+# C10, din
+i <- 1
+
+# data, respons variable label
+dat <- mods_nolag[i, ]$data[[1]]
+resvar <- mods_nolag[i, ]$resvar
+flovar <- mods_nolag[i, ]$flovar
+sta <- mods_nolag[i, ]$Site_Code
+reslab <- with(lablk, lngs[shrt == resvar])
+flolab <- with(stalk, shrt[lngs == flovar])
+  
+# prep data as tidal object
+tomod <- select(dat, Date, resval, flolag, lim) %>% 
+  rename(
+    res = resval, 
+    flo = flolag
+  ) %>% 
+  data.frame %>% 
+  tidal(., 
+    reslab = reslab, 
+    flolab = expression(paste('ln-flow (standardized)'))
+  )
+    
+topred <- filter(flow_dat, station == flolab) %>% 
+  mutate(flo = log(q)) %>% 
+  rename(date = Date) %>% 
+  select(date, flo) %>% 
+  filter(date >= min(tomod$date) & date <= max(tomod$date)) %>% 
+  na.omit %>% 
+  data.frame
+    
+# create model and exit
+mod <- wrtds(tomod, tau = c(0.1, 0.5, 0.9), wins = list(0.5, 15, 0.5), flo_div = 30, min_obs = 150)
+
[…]

### 2016-06-16 README.Rmd
readme with file descriptions, dinc10 file added

---
title: "README"
output: 
  html_document:
    keep_md: yes
    highlight: tango
author: "Marcus W. Beck, beck.marcus@epa.gov"
---

### Files

Files marked with * were created in `sf_trends.RProj`, all others created in `R\dat_proc.R`

`chld7.RData` WRTDS model for chl at D7

`clams.RData`* clam data for D7 from Crauder 2016, includes sample date, water year (`yr` starting in October), biomass in g/m2, clams per sample (`clams_smp`) in no/m2, depth (m), growth rate (`gr`) in m3/m2/d, mean size (mm), number of grabs (`no_grabs`), recruitment per unit area (`recruit_area`) in recruit/0.05 m2, and species as corbicula or potamocorbula.

`delt_dat.RData`* Processed wq time series data `dwr_wq.RData`, includes all nitrogen analytes and current/active stations in the delta, also includes matched and smoothed flow records from `flocor.RData` results

`diat_dat.RData`* additional WRTDS models for D7, including chlorphyll and sio2

`flow_dat.RData`* time series of daily flow estimates for the delta, input stations from Novick et al (Fig 2) were used

`mods_nolag.RData`* complete dataset for wrtds, including model results.  This is the same file as `mods_lag.RData` except the matched flow variables are not lagged. Nested, do not load without loading tidyr.

`dinc10.RData` WRTDS model for DIN at C10

### 2016-06-16 R/dat_proc.R

@@ -129,3 +129,7 @@ mod <- wrtds(tomod, tau = c(0.1, 0.5, 0.9), wins = list(0.5, 15, 0.5), flo_div =
 out <- mod %>% 
   respred(dat_pred = topred) %>% 
   resnorm
+
+# save
+dinc10 <- out
+save(dinc10, file = 'data/dinc10.RData')

### 2016-06-17 sftrends_manu.Rnw
figures for C10 din

@@
 \acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
+\acrodef{din}[DIN]{dissolved inorganic nitrogen}
 \acrodef{sfe}[SFE]{San Francisco Estuary}
@@
 \subsection{Case studies}
+These are science questions that are relevant outside of the region.
 \subsubsection{Disaggregating observed nitrogen time series}
@@
 Figure 10 in \citet{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
+
+Note the decrease in Potamocorbula abundance in 2011, 2012.
+These are wet years where abundance/biomass of the clams is driven down by lower salinity.
+Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.
+The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.
+This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
+
+Further, chlorophyll trends early in the time series generally show a decrease with high flow with a distinct maximum at moderate flow.
+This may suggest stratification events at moderate flow contributed to phytoplankton blooms early in the time series.
+Water withdrawals later in the time series could have also altered environmental conditions to reduce the frequency occurrence of stratification events.
+Look into this more...
+
+What about biomass/density relationships for Potamocorbula?
+Although clam density increases throughout the period,  What about initial decrease in chlorophyll prior to clam invasion?
+Is this related to water withdrawals (i.e., decrease in stratification events at moderate flow)?
+
 \cref{fig:clmchl}, \cref{tab:chlsio}
@@
 
+<<dinc10, fig.height = 8, fig.width = 7, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
+Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).
+The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.
+Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
+Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.'>>=
+# data
+data(dinc10)
+
+# din at c10
+toplo <- dinc10
+
+# globals
+ylab <- expression(paste('DIN (mg ', L^-1, ')'))
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(),
+    axis.line.y = element_line(), 
+    axis.title.x = element_blank()
+  )
+col_vec <- RColorBrewer::brewer.pal(9, 'Set1')[c(3, 2, 1)]
+col_lim <- c(0, 3.8)
+xlims <- range(toplo$date)
+txt <- data.frame(
+  x = min(toplo$date), 
+  y =  c(5, 2.4, 0.5, 7), 
+  txt = c('a', 'b', 'c', 'd')
+)
+size <- 7
+
+# observed din
+toplo1 <- select(toplo, date, res) %>% 
+  na.omit()
+p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
+  geom_line() + 
+  ptheme + 
+  scale_y_continuous(ylab) +
+  theme(
+    axis.text.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, -24, 6), 'pt')
+  ) +
+  geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
+  scale_x_date(limits = xlims)
+
+# din prdnrmplot
+p2 <- prdnrmplot(toplo, logspace = F, col_vec = col_vec) + 
+  ptheme + 
+  scale_y_continuous(ylab) +
+  theme(
+    legend.position = 'top', 
+    axis.text.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, -12, 6), 'pt')
+  ) +
+  geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
+  scale_x_date(limits = xlims)
+
+# flo/observed diffs
+toplo2 <- prdnrmplot(toplo, plot = F)
+toplo2 <- left_join(toplo2$fits, toplo2$nrms, by = c('date', 'taus')) %>% 
+  mutate(flodiff = fits_value - nrms_value) %>% 
+  filter(taus == 0.5)
+toplo3 <- toplo
+toplo3$flo <- scales::rescale(toplo3$flo, to = attr(toplo, 'floobs_rng'))
+ylab <- expression(paste('Pred - norm DIN (mg ', L^-1, ')'))
+p3 <- ggplot(toplo2, aes(x = date, y = flodiff)) + 
+  geom_bar(stat = 'identity') + 
+  theme_minimal() +
+  ptheme + 
+  theme(
+    axis.text.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, -12, 6), 'pt')
+    ) + 
+  scale_y_continuous(ylab) +
+  scale_x_date(limits = xlims) +
+  geom_text(data = txt[3, ], aes(x = x, y = y, label = txt), size = size)
+
+# flo with loess
+ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+p4 <- ggplot(toplo3, aes(x = date, y = flo)) + 
+  geom_line() +
+  stat_smooth(se = F) + 
+  ptheme + 
+  scale_y_continuous(ylab) +
+  scale_x_date(limits = range(toplo3$date)) +
+  geom_text(data = txt[4, ], aes(x = x, y = y, label = txt), size = size)
+
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+pC <- ggplot_gtable(ggplot_build(p3))
+pD <- ggplot_gtable(ggplot_build(p4))
+maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[2:3], pD$widths[2:3])
+
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+pC$widths[2:3] <- maxWidth
+pD$widths[2:3] <- maxWidth
+
+grid.arrange(
+  pA, pB, pC, pD,
+  heights = c(0.9, 1.3, 1, 1.2),
+  ncol = 1
+  )
+@
+
+<<dinc10dyna, fig.height = 6, fig.width = 9, out.width = '\\textwidth', fig.cap = 'Modelled relationships between \\ac{din}, flow, and time at station C10.
+The top figure shows the annual and seasonal variation over the entire time series and the bottom figure shows annual variation for selected months to remove seasonal variation.
+Warmer colors indicate higher \\ac{din} concentrations.
+The y-axis on the bottom figure is truncated by the fifth and ninety-fifth percentiles of flow within each month.
+Model results are for the fiftieth conditional quantile of \\ac{din}.'>>=
+
+# data
+data(dinc10)
+
+# din at c10
+toplo <- dinc10
+
+# globals
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(),
+    axis.line.y = element_line(), 
+    axis.title.x = element_blank()
+  )
+col_lim <- c(0, 2.9)
+txt <- data.frame(
+  x = min(toplo$date), 
+  y =  c(5, 2.4, 0.5, 6), 
+  txt = c('a', 'b', 'c', 'd')
+)
+size <- 7
+
+# dynaplot, all
+col_vec = NULL
+attr(toplo, 'reslab') <- expression(paste("DIN (mg ", L^-1, ")"))
+ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+p1 <- gridplot(toplo, month = 'all', logspace = F, tau = 0.5, col_lim = col_lim, floscl = F, col_vec = rev(col_vec)) +
+  scale_y_continuous(ylab, expand = c(0, 0)) + 
+  ptheme + 
+  theme(legend.position = 'top')
+
+# dynaplot selected months
+p2 <- gridplot(toplo, logspace = F, tau = 0.5, col_lim = col_lim, month = c(1, 4, 7, 10), ncol = 4, floscl = F) + 
+  ptheme + 
+  theme(
+    strip.background = element_blank(),
+    legend.position = 'none', 
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank()
+    ) + 
+  scale_y_continuous(ylab, expand = c(0, 0))
+
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
+
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+
+grid.arrange(
+  pA, pB,
+  heights = c(1.2, 1),
+  ncol = 1
+  )
+@
+
 <<clmchl, fig.height = 7, fig.width = 10, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
-Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (top left, annual means).
-A coincident decrease in \\ac{chla} concentration was also observed (bottom left).
-A weak but significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in the top right.
-Flow relationships with \\ac{chla} concentration have also changed over time (bottom right, observations from June).
+Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means).
+A coincident decrease in \\ac{chla} concentration was also observed (c).
+A weak but significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (b).
+Flow relationships with \\ac{chla} concentration have also changed over time (d, observations from June).
 Chlorophyll shows a slight positive then dominantly negative association with increasing flow (decreasing salinity) early in the time series, whereas the trend is reversed in recent years.'>>=
@@
 
+# subfig labs
+txt <- data.frame(
+  x = c(as.numeric(xlims[1]), as.numeric(xlims[2]), 25, 0.25),
+  y =  c(10000, 11, 4, 3.8), 
+  txt = c('a', 'c', 'b', 'd')
+)
+size <- 7
+
 # annual clam time series
@@
   scale_x_date('', limits = xlims) +
-  scale_y_continuous('Clams per sample')
+  scale_y_continuous('Clams per sample') +
+  with(txt[1, ], annotate('text', x = xlims[1], y = y, label = txt, size = size))
 
@@
   scale_x_date('', limits = xlims) +
-  scale_y_continuous(expression(paste("Chl-", italic(a), " (", italic(mu), "g ", L^-1, ")")))
+  scale_y_continuous(expression(paste("Chl-", italic(a), " (", italic(mu), "g ", L^-1, ")"))) +
+  with(txt[2, ], annotate('text', x = xlims[2], y = y, label = txt, size = size))
 
@@
   scale_x_continuous(xlab) + 
-  scale_y_continuous(ylab, limits = c(0.8, 4))
+  scale_y_continuous(ylab, limits = c(0.8, 4)) +
+  geom_text(data = txt[3, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
 
@@
   scale_x_continuous('ln-Salinity (psu)') + 
-  scale_y_continuous(ylab, limits = c(0.4, 3.8))
+  scale_y_continuous(ylab, limits = c(0.4, 3.8)) +
+  annotate('text', x = txt[4, 'x'], y = txt[4, 'y'], label = txt[4, 'txt'], size = size)
 

### 2016-06-17 sftrends_manu.tex
figures for C10 din

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Jun 14 18:01:49 2016 -0500
+  \\ Version Date:   Thu Jun 16 16:53:30 2016 -0500
 	}
@@
 
-The \ac{sfe}... The delta...
-
+The \ac{sfe}... exchange with ocean, freshwater inputs, runoff variation (drought years), climate, support of natural resources
 \subsection{Analysis method and application}
@@
 
+{\centering \includegraphics[width=0.75\textwidth]{figs/dinc10-1}
+
+}
+
+\caption{Time series of \ac{din} and flow at station C10.
+Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual predictions from \ac{wrtds} at different conditional quantiles ($\tau$ = 0.1, 0.5, 0.9).
+The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.
+Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
+Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.}\label{fig:dinc10}
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\textwidth]{figs/dinc10dyna-1}
+
+}
+
+\caption{Modelled relationships between \ac{din}, flow, and time at station C10.
+The top figure shows the annual and seasonal variation over the entire time series and the bottom figure shows annual variation for selected months to remove seasonal variation.
+Warmer colors indicate higher \ac{din} concentrations.
+The y-axis on the bottom figure is truncated by the fifth and ninety-fifth percentiles of flow within each month.
+Model results are for the fiftieth conditional quantile of \ac{din}.}\label{fig:dinc10dyna}
+\end{figure}
+
+\begin{figure}[!ht]
+
 {\centering \includegraphics[width=\textwidth]{figs/clmchl-1}
@@
 \caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
-Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (top left, annual means).
-A coincident decrease in \ac{chla} concentration was also observed (bottom left).
-A weak but significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in the top right.
-Flow relationships with \ac{chla} concentration have also changed over time (bottom right, observations from June).
+Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means).
+A coincident decrease in \ac{chla} concentration was also observed (c).
+A weak but significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (b).
+Flow relationships with \ac{chla} concentration have also changed over time (d, observations from June).
 Chlorophyll shows a slight positive then dominantly negative association with increasing flow (decreasing salinity) early in the time series, whereas the trend is reversed in recent years.}\label{fig:clmchl}

### 2016-06-17 R/dat_proc.R

@@ -68,7 +68,6 @@ out <- mod %>%
 chld7 <- out
 save(chld7, file = 'data/chld7.RData', compress = 'xz')
 
-
 ######
 # a smoother din model for c10
 data(mods_nolag)
@@ -92,7 +91,7 @@ stalk <- list(
   )
 
 # C10, din
-i <- 1
+i <- which(mods_nolag$Site_Code == 'C10' & mods_nolag$resvar == 'din')
 
 # data, respons variable label
 dat <- mods_nolag[i, ]$data[[1]]

### 2016-06-17 sftrends_manu.Rnw
some points about first case study

@@
 
+nine discrete sampling stations with data from 1976 to 2012, \ac{din}, ammonium, total nitrate, effects of different flow variables
+
+Annual, seasonal trends of flow-normalized predictions
+
 \subsection{Case studies}
@@
 \subsubsection{Disaggregating observed nitrogen time series}
-Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
+Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear and complex, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
 \subsubsection{Effects of wastewater treatment}
@@
 \subsection{Selected examples}
+
+\subsubsection{Disaggregating observed nitrogen time series}
+
+\cref{fig:dinc10}, \cref{fig:dinc10dyna}
+
+Emphasize the information the model provides relative to the observed time series.
+A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
+The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
+Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
+The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
+Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
+The quantile distributions showed similar trends over time in both predicted values and flow-normalized predictions, although some exceptions were observed.
+In particular, high flow (1984, 2008) reduced concentrations of all quantiles but the magniutude of the effect increased at higher quantiles (i.e., the effect was disproportionate).
+The opposite was observed for low flow, i.e., the ninetieth percent showed the greatest increase for low flow.
+
+Emphasize the summer/fall change in the 2000s, why is this?  Relate to conceptual diagram.
 
@@
 <<dinc10, fig.height = 8, fig.width = 7, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
-Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).
+Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).
 The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.

### 2016-06-17 sftrends_manu.tex
some points about first case study

@@
   \vspace{1in} 
-  \\ Version Date:   Thu Jun 16 16:53:30 2016 -0500
+  \\ Version Date:   Fri Jun 17 16:39:22 2016 -0500
 	}
@@
 \subsection{Selected examples}
+
+\subsubsection{Disaggregating observed nitrogen time series}
+
+\cref{fig:dinc10}, \cref{fig:dinc10dyna}
+
+Emphasize the information the model provides relative to the observed time series.
+A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
+The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
+Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
+The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
+Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
+The quantile distributions showed similar trends over time in both predicted values and flow-normalized predictions, although some exceptions were observed.
+In particular, high flow (1984, 2008) reduced concentrations of all quantiles but the magniutude of the effect increased at higher quantiles (i.e., the effect was disproportionate).
+The opposite was observed for low flow, i.e., the ninetieth percent showed the greatest increase for low flow.
+
+Emphasize the summer/fall change in the 2000s, why is this?
 
@@
 \caption{Time series of \ac{din} and flow at station C10.
-Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual predictions from \ac{wrtds} at different conditional quantiles ($\tau$ = 0.1, 0.5, 0.9).
+Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} at different conditional quantiles ($\tau$ = 0.1, 0.5, 0.9).
 The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.

### 2016-06-21 sftrends_manu.tex
nutrient load data for stockton and tracy

@@
 % get online bib file
+\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in curl::curl\_fetch\_memory(url, handle = handle): Timeout was reached}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in rawToChar(refs\$content): object 'refs' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in writeLines(refs, con = file("{}refs.bib"{})): object 'refs' not found}}\end{kframe}
 
@@
   \vspace{1in} 
-  \\ Version Date:   Fri Jun 17 16:39:22 2016 -0500
+  \\ Version Date:   Fri Jun 17 17:17:59 2016 -0500
 	}
@@
 
-Emphasize the summer/fall change in the 2000s, why is this?
+Emphasize the summer/fall change in the 2000s, why is this?  Relate to conceptual diagram.
 

### 2016-06-22 sftrends_manu.tex
tracy discharge info to draft

--- 2016-06-17 sftrends_manu.tex
@@
   \vspace{1in} 
-  \\ Version Date:   Fri Jun 17 16:39:22 2016 -0500
+  \\ Version Date:   Tue Jun 21 13:55:26 2016 -0500
 	}
@@
 Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions.
+See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
 \subsubsection{Effects of biological invasions}

### 2016-06-23 sftrends_manu.Rnw
tracy discharge info to draft
mods_nolag_mean added, not used yet

@@
 
-The \ac{sfe}... exchange with ocean, freshwater inputs, runoff variation (drought years), climate, support of natural resources
+The \ac{sfe}...
+exchange with ocean, freshwater inputs (rivers and wshed drainage), runoff variation (drought years), climate, support of natural resources, subembayments, POTW (publicly owned treatment works), nutrient loading relative to other locations
 \subsection{Analysis method and application}
@@
 
-Emphasize the summer/fall change in the 2000s, why is this?  Relate to conceptual diagram.
+Emphasize the summer/fall change in the 2000s, why is this?
+Check \citep{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
+Relate to conceptual diagram.
 

### 2016-06-23 sftrends_manu.tex
mods_nolag_mean added, not used yet

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Jun 21 13:55:26 2016 -0500
+  \\ Version Date:   Wed Jun 22 07:22:19 2016 -0500
 	}

### 2016-07-13 sftrends_manu.Rnw
c10 d7 trends plot

@@
 \subsubsection{Effects of biological invasions}
-Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in stoichiometric ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
+Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
 
@@
 \subsubsection{Clam invasion in Suisun Bay}
+
 Data from \citep{Crauder16}, \citet{Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
@@
 \section{Discussion}
+
+What do nitrogen trends mean?
+Have to interpret relative to trends in other variables.
+A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
+A change in chlorohpyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
+However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
+Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
+See Senn slide 14 (from burial?).
+The WRTDS model lets us at least address trends in the context of season, time, and flow.
+This allows for more improved interpretion relative to observing raw data.
+Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
+What about other variables (light level as suspended particulate matter, temperature)?
 
@@
 
+<<d7c10trnds, fig.cap = 'Flow-normalized trends of annual (left) and seasonal (right) variation in DIN, chl-{\\it\\footnotesize a}, and SiO$_2$ at station D7 (top) and C10 (bottom).
+Covariation between nutrients, chl-{\\it\\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.
+Seasonal changes at D7 are most pronounced during the summer.', fig.width = 9, fig.height = 10>>=
+data(h3dat)
+
+# separate each site
+toplo1 <- filter(h3dat, Site_Code %in% 'D7')
+toplo2 <- filter(h3dat, Site_Code %in% 'C10')
+
+# y axis labels
+ylab1 <- expression(paste("DIN (mg ", L^-1, ")"))
+ylab2 <- expression(paste("Chl-", italic(a), " (", italic(mu), "g ", L^-1, ")"))
+ylab3 <- expression(paste("Si", O[2], " (mg ", L^-1, ")"))
+
+# generic theme for all plots
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(),
+    axis.line.y = element_line(), 
+    axis.title.x = element_blank(), 
+    axis.text.x = element_blank()
+  )
+
+# colors
+cols <- RColorBrewer::brewer.pal(9, 'Set1')[1:3]
+
+## first set of plots
+pdin1 <- prdnrmplot(toplo1$mod[[2]], col_vec = cols[2], logspace = F, min_mo = 10) +
+  ptheme + 
+  scale_y_continuous(ylab1) +
+  theme(legend.position = 'top', legend.title = element_blank())
+pleg1 <- g_legend(pdin1)
+pdin1 <- pdin1 + theme(legend.position = 'none')
+
+pdinsy1 <- seasyrplot(toplo1$mod[[2]], col_vec = cols, logspace = F, predicted = F) +
+  ptheme + 
+  theme(legend.position = 'top', axis.title.y = element_blank())
+pleg2 <- g_legend(pdinsy1)
+pdinsy1 <- pdinsy1 + theme(legend.position = 'none')
+
+pchl1 <- prdnrmplot(toplo1$mod[[1]], col_vec = cols[2], logspace = F, min_mo = 10) +
+  ptheme + 
+  theme(legend.position = 'none') + 
+  scale_y_continuous(ylab2)
+
+pchlsy1 <- seasyrplot(toplo1$mod[[1]], col_vec = cols, logspace = F, predicted = F) +
+  ptheme + 
+  theme(legend.position = 'none', axis.title.y = element_blank())
+
+psio1 <- prdnrmplot(toplo1$mod[[3]], col_vec = cols[2], logspace = F, min_mo = 10) +
+  ptheme + 
+  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text()) +
+  scale_x_date('Year') +
+  scale_y_continuous(ylab3)
+
+psiosy1 <- seasyrplot(toplo1$mod[[3]], col_vec = cols, logspace = F, predicted = F) + 
+  ptheme + 
+  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text(), axis.title.y = element_blank()) +
+  scale_x_date('Day of year', date_labels = "%m/%d")
+
+# Set the widths
+pA1 <- ggplot_gtable(ggplot_build(pdin1))
+pB1 <- ggplot_gtable(ggplot_build(pchl1))
+pC1 <- ggplot_gtable(ggplot_build(psio1))
+maxWidth = unit.pmax(pA1$widths[2:3], pB1$widths[2:3], pC1$widths[2:3])
+pA1$widths[2:3] <- maxWidth
+pB1$widths[2:3] <- maxWidth
+pC1$widths[2:3] <- maxWidth
+
+# Set the widths for seasyrplots
+pAsy1 <- ggplot_gtable(ggplot_build(pdinsy1))
+pBsy1 <- ggplot_gtable(ggplot_build(pchlsy1))
+pCsy1 <- ggplot_gtable(ggplot_build(psiosy1))
+maxWidth = unit.pmax(pAsy1$widths[2:3], pBsy1$widths[2:3], pCsy1$widths[2:3])
+pAsy1$widths[2:3] <- maxWidth
+pBsy1$widths[2:3] <- maxWidth
+pCsy1$widths[2:3] <- maxWidth
+
+## second set of plots
+pdin2 <- prdnrmplot(toplo2$mod[[2]], col_vec = cols[2], logspace = F, min_mo = 10) +
+  ptheme + 
+  scale_y_continuous(ylab1) +
+  theme(legend.position = 'none')
+
+pdinsy2 <- seasyrplot(toplo2$mod[[2]], col_vec = cols, logspace = F, predicted = F) +
+  ptheme + 
+  theme(legend.position = 'none', axis.title.y = element_blank())
+
+pchl2 <- prdnrmplot(toplo2$mod[[1]], col_vec = cols[2], logspace = F, min_mo = 10) +
+  ptheme + 
+  theme(legend.position = 'none') + 
+  scale_y_continuous(ylab2)
+
+pchlsy2 <- seasyrplot(toplo2$mod[[1]], col_vec = cols, logspace = F, predicted = F) +
+  ptheme + 
+  theme(legend.position = 'none', axis.title.y = element_blank())
+
+psio2 <- prdnrmplot(toplo2$mod[[3]], col_vec = cols[2], logspace = F, min_mo = 10) +
+  ptheme + 
+  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text()) +
+  scale_x_date('Year') +
+  scale_y_continuous(ylab3)
+
+psiosy2 <- seasyrplot(toplo2$mod[[3]], col_vec = cols, logspace = F, predicted = F) + 
+  ptheme + 
+  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text(), axis.title.y = element_blank()) +
+  scale_x_date('Day of year', date_labels = "%m/%d")
+
+# Set the widths
+pA2 <- ggplot_gtable(ggplot_build(pdin2))
+pB2 <- ggplot_gtable(ggplot_build(pchl2))
+pC2 <- ggplot_gtable(ggplot_build(psio2))
+maxWidth = unit.pmax(pA2$widths[2:3], pB2$widths[2:3], pC2$widths[2:3])
+pA2$widths[2:3] <- maxWidth
+pB2$widths[2:3] <- maxWidth
+pC2$widths[2:3] <- maxWidth
+
+# Set the widths for seasyrplots
+pAsy2 <- ggplot_gtable(ggplot_build(pdinsy2))
+pBsy2 <- ggplot_gtable(ggplot_build(pchlsy2))
+pCsy2 <- ggplot_gtable(ggplot_build(psiosy2))
+maxWidth = unit.pmax(pAsy2$widths[2:3], pBsy2$widths[2:3], pCsy2$widths[2:3])
+pAsy2$widths[2:3] <- maxWidth
+pBsy2$widths[2:3] <- maxWidth
+pCsy2$widths[2:3] <- maxWidth
+
+grid.arrange(
+  arrangeGrob(
+    ncol = 1, heights = c(0.05, 1),
+    arrangeGrob(pleg1, pleg2, ncol = 2),
+    arrangeGrob(
+      arrangeGrob(
+        textGrob('(a) Station D7, Suisun Bay'), heights = c(0.1, 1), 
+        arrangeGrob(
+          arrangeGrob(pA1, pB1, pC1, ncol = 1, heights = c(0.85, 0.85, 1)),
+          arrangeGrob(pAsy1, pBsy1, pCsy1, ncol = 1, heights = c(0.85, 0.85, 1)),
+          ncol = 2
+        )
+      ),
+      arrangeGrob(
+        textGrob('(b) Station C10, Delta'), heights = c(0.1, 1),
+        arrangeGrob(
+          arrangeGrob(pA2, pB2, pC2, ncol = 1, heights = c(0.85, 0.85, 1)),
+          arrangeGrob(pAsy2, pBsy2, pCsy2, ncol = 1, heights = c(0.85, 0.85, 1)),
+          ncol = 2
+        )
+      ),
+      ncol = 1 
+    )
+  )
+)
+
+@
+
 \clearpage
@@
 
-cap.val<-'Summaries of flow-normalized trends in chlorophyll and silicon dioxide concentrations for different time periods.
+cap.val<-'Summaries of flow-normalized trends in chlorophyll ($\\mu$g $L^{-1}$) and silicon dioxide (mg $L^{-1}$) concentrations for different time periods.
 Summaries are means and percent changes based on annual means within the time periods.

### 2016-07-13 sftrends_manu.tex
c10 d7 trends plot

@@
   \vspace{1in} 
-  \\ Version Date:   Wed Jun 22 07:22:19 2016 -0500
+  \\ Version Date:   Thu Jun 23 14:00:35 2016 -0500
 	}
@@
 
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figs/d7c10trnds-1}
+
+}
+
+\caption{Flow-normalized trends of annual (left) and seasonal (right) variation in DIN, chl-{\it\footnotesize a}, and SiO$_2$ at station D7 (top) and C10 (bottom).
+Covariation between nutrients, chl-{\it\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.
+Seasonal changes at D7 are most pronounced during the summer.}\label{fig:d7c10trnds}
+\end{figure}
+
 \clearpage
@@
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in chlorophyll and silicon dioxide concentrations for different time periods.
+\caption{Summaries of flow-normalized trends in chlorophyll ($\mu$g $L^{-1}$) and silicon dioxide (mg $L^{-1}$) concentrations for different time periods.
 Summaries are means and percent changes based on annual means within the time periods.

### 2016-07-13 sftrends_manu.Rnw
d7 and c10 trend tables added

@@
 
-% trends in sio and chl at d7
-<<chlsio>>=
+% trends in din, chl, sio2 at d7
+<<d7chg>>=
 
-data(diat_dat)
+data(h3dat)
 
-sioraw <- filter(diat_dat, resvar %in% 'sio2') %>% 
+d7dat <- filter(h3dat, Site_Code %in% 'D7')
+
+dinraw <- filter(d7dat, resvar %in% 'din') %>% 
   .$mod %>% 
   .[[1]]
-chlraw <- filter(diat_dat, resvar %in% 'chl') %>% 
+chlraw <- filter(d7dat, resvar %in% 'chl') %>% 
+  .$mod %>% 
+  .[[1]]
+sioraw <- filter(d7dat, resvar %in% 'sio2') %>% 
   .$mod %>% 
@@
 
-# summarize sio2 and chl trends
+# summarize din, sio2, chl trends
+din <- wrtdstrnd(dinraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
+din <- wrtdstrnd(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+  .[1, ] %>% 
+  rbind(., din) %>% 
+  data.frame(res = 'din', .)
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+  .[1, ] %>% 
+  rbind(., chl) %>% 
+  data.frame(res = 'chl', .)
 sio <- wrtdstrnd(sioraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
@@
   data.frame(res = 'sio', .)
-chl <- wrtdstrnd(chlraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
-chl <- wrtdstrnd(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
-  .[1, ] %>% 
-  rbind(., chl) %>% 
-  data.frame(res = 'chl', .)
 
 # combine for trends for table
-totab <- rbind(sio, chl) %>% 
+totab <- rbind(din, chl, sio) %>% 
   mutate(
@@
   arrange(cat) %>% 
-  data.frame(sep = c(rep('Annual', 5), rep('Seasonal', 4)), .)
+  data.frame(sep = c(rep('Annual', 5), rep('Seasonal', 4)), .) %>% 
+  select(sep, cat, din_ave, din_chg, chl_ave, chl_chg, sio_ave, sio_chg)
 
@@
 totab <- totab[,-c(1:2)]
-vars <- c('Chl-\\textit{a}', 'SiO$_2$')
-names(totab) <- rep(c('Mean', '\\% change'), 2)
+vars <- c('DIN', 'Chl-\\textit{a}', 'SiO$_2$')
+names(totab) <- rep(c('Mean', '\\% change'), 3)
 
-cap.val<-'Summaries of flow-normalized trends in chlorophyll ($\\mu$g $L^{-1}$) and silicon dioxide (mg $L^{-1}$) concentrations for different time periods.
+cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station D7.
 Summaries are means and percent changes based on annual means within the time periods.
@@
   cgroup = vars,
-  n.cgroup = c(2, 2),
+  n.cgroup = c(2, 2, 2),
   rowname = cats,
-  label = 'tab:chlsio'
+  label = 'tab:d7chg'
+  )
+
+@
+
+% trends in din, chl, sio2 at c10
+<<c10chg>>=
+
+data(h3dat)
+
+c10dat <- filter(h3dat, Site_Code %in% 'C10')
+
+dinraw <- filter(c10dat, resvar %in% 'din') %>% 
+  .$mod %>% 
+  .[[1]]
+chlraw <- filter(c10dat, resvar %in% 'chl') %>% 
+  .$mod %>% 
+  .[[1]]
+sioraw <- filter(c10dat, resvar %in% 'sio2') %>% 
+  .$mod %>% 
+  .[[1]]
+
+mobrks <- c(-Inf, 3, 6, 9, Inf)
+yrbrks <- c(-Inf, 1985, 1994, 2003, Inf)
+molabs <- c('JFM', 'AMJ', 'JAS', 'OND')
+yrlabs <- c('1976-1985', '1986-1994', '1995-2003', '2004-2013')
+
+# summarize din, sio2, chl trends
+din <- wrtdstrnd(dinraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
+din <- wrtdstrnd(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+  .[1, ] %>% 
+  rbind(., din) %>% 
+  data.frame(res = 'din', .)
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+  .[1, ] %>% 
+  rbind(., chl) %>% 
+  data.frame(res = 'chl', .)
+sio <- wrtdstrnd(sioraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
+sio <- wrtdstrnd(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+  .[1, ] %>% 
+  rbind(., sio) %>% 
+  data.frame(res = 'sio', .)
+
+# combine for trends for table
+totab <- rbind(din, chl, sio) %>% 
+  mutate(
+    chg = round(chg, 1), 
+    chg = ifelse(chg < 0, chg, paste0('\\textit{\\textbf{', chg , '}}')), 
+    ave = as.character(round(ave, 1))
+    ) %>% 
+  gather('var', 'val', chg:ave) %>% 
+  unite('col', res, var) %>% 
+  spread(col, val) %>% 
+  mutate(cat = factor(cat, levels = c('1976-2013', yrlabs, molabs), labels = c('1976-2013', yrlabs, molabs))) %>% 
+  arrange(cat) %>% 
+  data.frame(sep = c(rep('Annual', 5), rep('Seasonal', 4)), .) %>% 
+  select(sep, cat, din_ave, din_chg, chl_ave, chl_chg, sio_ave, sio_chg)
+
+# final table formatting
+segs <- c('All', 'Annual', 'Seasonal')
+cats <- totab$cat
+totab <- totab[,-c(1:2)]
+vars <- c('DIN', 'Chl-\\textit{a}', 'SiO$_2$')
+names(totab) <- rep(c('Mean', '\\% change'), 3)
+
+cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station C10.
+Summaries are means and percent changes based on annual means within the time periods.
+Increasing values are in bold-italics.'
+
+latex( 
+  totab,
+  file = '',
+  rowlabel = 'Period',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = segs,
+  n.rgroup = c(1, rep(4, 2)),
+  cgroup = vars,
+  n.cgroup = c(2, 2, 2),
+  rowname = cats,
+  label = 'tab:c10chg'
   )

### 2016-07-13 sftrends_manu.tex
d7 and c10 trend tables added

@@
   \vspace{1in} 
-  \\ Version Date:   Thu Jun 23 14:00:35 2016 -0500
+  \\ Version Date:   Wed Jul 13 15:18:11 2016 -0500
 	}
@@
 
-% trends in sio and chl at d7
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:chlsio")%
+% trends in din, chl, sio2 at d7
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:d7chg")%
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in chlorophyll ($\mu$g $L^{-1}$) and silicon dioxide (mg $L^{-1}$) concentrations for different time periods.
+\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station D7.
 Summaries are means and percent changes based on annual means within the time periods.
-Increasing values are in bold-italics.\label{tab:chlsio}}
+Increasing values are in bold-italics.\label{tab:d7chg}}
 \begin{center}
-\begin{tabular}{lllcll}
+\begin{tabular}{lllcllcll}
 \hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
-\cline{2-3} \cline{5-6}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
+\cline{2-3} \cline{5-6} \cline{8-9}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
 \hline
-{\bfseries All}&&&&&\tabularnewline
-~~1976-2013&3.5&-80.1&&12.5&\textit{\textbf{33.6}}\tabularnewline
+{\bfseries All}&&&&&&&&\tabularnewline
+~~1976-2013&0.4&-4.2&&4.6&-82.4&&12.5&\textit{\textbf{30}}\tabularnewline
 \hline
-{\bfseries Annual}&&&&&\tabularnewline
-~~1976-1985&7.8&-62.3&&10.5&\textit{\textbf{8.9}}\tabularnewline
-~~1986-1994&2.2&-62.1&&12.4&\textit{\textbf{15.6}}\tabularnewline
-~~1995-2003&1.6&\textit{\textbf{20.2}}&&13.7&\textit{\textbf{4.5}}\tabularnewline
-~~2004-2013&2&\textit{\textbf{31.3}}&&13.6&-2\tabularnewline
+{\bfseries Annual}&&&&&&&&\tabularnewline
+~~1976-1985&0.4&-15.2&&10.2&-67.9&&10.7&\textit{\textbf{8.4}}\tabularnewline
+~~1986-1994&0.4&\textit{\textbf{15.4}}&&3.3&-48.8&&12.3&\textit{\textbf{13.5}}\tabularnewline
+~~1995-2003&0.5&\textit{\textbf{4.3}}&&2.1&-3.9&&13.5&\textit{\textbf{3.7}}\tabularnewline
+~~2004-2013&0.5&-10.6&&2.5&\textit{\textbf{31.7}}&&13.7&-1.8\tabularnewline
 \hline
-{\bfseries Seasonal}&&&&&\tabularnewline
-~~JFM&1.9&-43.2&&14.8&\textit{\textbf{7.3}}\tabularnewline
-~~AMJ&4.3&-72.8&&12.9&\textit{\textbf{25.7}}\tabularnewline
-~~JAS&4.8&-88.4&&11.1&\textit{\textbf{70.8}}\tabularnewline
-~~OND&2.1&-79.4&&11.7&\textit{\textbf{41.7}}\tabularnewline
+{\bfseries Seasonal}&&&&&&&&\tabularnewline
+~~JFM&0.5&-4.9&&2.7&-56&&15&\textit{\textbf{4.4}}\tabularnewline
+~~AMJ&0.4&\textit{\textbf{17.4}}&&6.1&-74.3&&12.9&\textit{\textbf{20.9}}\tabularnewline
+~~JAS&0.3&\textit{\textbf{40.6}}&&6.2&-89&&10.9&\textit{\textbf{79.1}}\tabularnewline
+~~OND&0.5&\textit{\textbf{13.1}}&&2.7&-83.8&&11.7&\textit{\textbf{33.7}}\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
+
+% trends in din, chl, sio2 at c10
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:c10chg")%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station C10.
+Summaries are means and percent changes based on annual means within the time periods.
+Increasing values are in bold-italics.\label{tab:c10chg}}
+\begin{center}
+\begin{tabular}{lllcllcll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
+\cline{2-3} \cline{5-6} \cline{8-9}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\hline
+{\bfseries All}&&&&&&&&\tabularnewline
+~~1976-2013&1.3&-16.9&&27.6&-4.9&&14.5&-23.1\tabularnewline
+\hline
+{\bfseries Annual}&&&&&&&&\tabularnewline
+~~1976-1985&1.3&\textit{\textbf{16.6}}&&26.6&-20.5&&15.7&-5\tabularnewline
+~~1986-1994&1.4&\textit{\textbf{6.6}}&&27.5&\textit{\textbf{23.7}}&&14.8&-3.9\tabularnewline
+~~1995-2003&1.4&-1.1&&29.3&-15.2&&14.1&-5.1\tabularnewline
+~~2004-2013&1.2&-32.6&&27.3&\textit{\textbf{11}}&&13&-9.6\tabularnewline
+\hline
+{\bfseries Seasonal}&&&&&&&&\tabularnewline
+~~JFM&1.4&-20.5&&12&\textit{\textbf{14.8}}&&14.3&-16.9\tabularnewline
+~~AMJ&1&-26.4&&43&\textit{\textbf{3.1}}&&13.1&-27.5\tabularnewline
+~~JAS&1.3&-3.7&&46.6&-13.6&&14.8&-26.1\tabularnewline
+~~OND&1.6&-6.8&&8.7&-18.5&&15.6&-16\tabularnewline
 \hline

### 2016-07-13 README.Rmd
nutrient load data for stockton and tracy
tracy discharge info to draft
c10 d7 trends plot
update clams plot with biomass flow reg

@@
 
-`chld7.RData` WRTDS model for chl at D7
-
 `clams.RData`* clam data for D7 from Crauder 2016, includes sample date, water year (`yr` starting in October), biomass in g/m2, clams per sample (`clams_smp`) in no/m2, depth (m), growth rate (`gr`) in m3/m2/d, mean size (mm), number of grabs (`no_grabs`), recruitment per unit area (`recruit_area`) in recruit/0.05 m2, and species as corbicula or potamocorbula.
@@
 
-`diat_dat.RData`* additional WRTDS models for D7, including chlorphyll and sio2
+`dinc10.RData` WRTDS model for DIN at C10
 
 `flow_dat.RData`* time series of daily flow estimates for the delta, input stations from Novick et al (Fig 2) were used
+
+`h3dat.RData`* mean models to support third hypothesis/case study in manuscript, wrtds mean models for sio2, din, chla at c10, d7
 
@@
 
-`dinc10.RData` WRTDS model for DIN at C10
+`potw_load.RData`* nutrient load data from Stockton and Tracy wwtp, kg/day

### 2016-07-13 sftrends_manu.Rnw
update clams plot with biomass flow reg

@@
 
-\cref{fig:clmchl}, \cref{tab:chlsio}
+\cref{fig:clmchl}, \cref{fig:d7c10trnds}, \cref{tab:d7chg}, \cref{tab:c10chg}
 
@@
 
-<<clmchl, fig.height = 7, fig.width = 10, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
+<<clmchl, fig.height = 7, fig.width = 8, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
 Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means).
@@
 data(clams)
-data(diat_dat)
-data(chld7)
-
-# format chl data and clam data for first two plots
-chldat <- filter(diat_dat, resvar %in% 'chl') %>% 
-  .$mod %>% 
-  .[[1]] %>% 
-  annual_agg %>% 
-  mutate(year = lubridate::year(date)) %>% 
-  select(date, year, res, norm0.5) %>% 
-  mutate(res = exp(res), norm0.5 = exp(norm0.5)) %>% 
-  rename(yr = date)
+data(h3dat)
+data(flow_dat)
+chld7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'chl')$mod[[1]]
 
@@
 txt <- data.frame(
-  x = c(as.numeric(xlims[1]), as.numeric(xlims[2]), 25, 0.25),
-  y =  c(10000, 11, 4, 3.8), 
-  txt = c('a', 'c', 'b', 'd')
+  x = c(as.numeric(as.Date('1979-10-01')), 2500, 25, 0.25),
+  y =  c(10000, 15, 4, 3.3), 
+  txt = c('(a)', '(c)', '(b)', '(d)')
 )
-size <- 7
+size <- 6
 
+##
 # annual clam time series
@@
     ) + 
-  scale_x_date('', limits = xlims) +
+  scale_x_date('Year', limits = xlims) +
   scale_y_continuous('Clams per sample') +
-  with(txt[1, ], annotate('text', x = xlims[1], y = y, label = txt, size = size))
+  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5)) +
+  with(txt[1, ], annotate('text', x = xlims[1] + 365 * 2, y = y, label = txt, size = size))
 
-# annual chlorophyll time series with flow-norm
-p2 <- ggplot(chldat, aes(x = yr, y = norm0.5)) + 
-  geom_point(aes(y = res, shape = 'Observed'), size = 3) + 
-  geom_line(aes(color = 'Flow-normalized')) + 
+##
+# clam biomass related to daily sacramento flow
+clmdat <- group_by(clams, yr, species) %>% 
+  dplyr::summarize(
+    clams_smp = mean(clams_smp, na.rm = TRUE),
+    biomass = mean(biomass, na.rm = TRUE)
+    ) %>% 
+  ungroup %>% 
+  mutate(yr = as.Date(paste0(yr, '-10-01')))
+
+flodat <- filter(flow_dat, station %in% 'sac')
+clmflo <- select(clams, biomass, clams_smp, species, date) %>% 
+  rename(Date = date) %>% 
+  left_join(., flodat, by = 'Date') %>% 
+  filter(species %in% 'Potamocorbula')
+
+# linear model of concentration x biomass
+mod <- lm(log(1 + biomass) ~ log(1 + q), data = clmflo) %>% 
+  coefficients
+xvals <- range(clmflo$q, na.rm = TRUE)
+xvals <- seq(xvals[1], xvals[2], length = 100)
+yvals <- (exp(mod[1] + mod[2] * log(1 + xvals))) - 1
+
+ylab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
+xlab <- expression(paste("Sacramento daily flow (", m^3, ' ', s^-1, ")"))
+
+# scatter plot of concentration x biomass with model 
+p2 <- ggplot(clmflo, aes(x = q, y = biomass)) + 
+  geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
+  geom_line(data = data.frame(x = xvals, y = yvals), aes(x, y, colour = 'black'), lwd = 1) +
   theme_minimal() + 
@@
     axis.line.x = element_line(),
-    # axis.title.x = element_blank(), 
     legend.position = 'top', 
@@
     ) +
-  scale_colour_manual(values = c('black', 'black')) +
-  scale_x_date('', limits = xlims) +
-  scale_y_continuous(expression(paste("Chl-", italic(a), " (", italic(mu), "g ", L^-1, ")"))) +
-  with(txt[2, ], annotate('text', x = xlims[2], y = y, label = txt, size = size))
+  scale_colour_manual(label = 'log(1 + biomass) = 5.2 - 0.6log(1 + flow)', values = 'black') +
+  scale_x_continuous(xlab) + 
+  scale_y_continuous(ylab, limits = c(0,15)) +
+  geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
 
+##
 # chlorophyll by month
-chldat <- filter(diat_dat, resvar %in% 'chl') %>% 
+chldat <- filter(h3dat, resvar %in% 'chl' & Site_Code %in% 'D7') %>% 
   .$mod %>% 
   .[[1]] %>% 
-  select(year, month, res, norm0.5) %>% 
-  mutate(res = exp(res), norm0.5 = exp(norm0.5)) %>% 
-  rename(yr = year, mo = month) %>% 
+  select(year, month, res, bt_norm) %>% 
+  mutate(res = exp(res)) %>% 
+  rename(yr = year, mo = month, bt_res = res) %>% 
   group_by(yr, mo) %>% 
-  dplyr::summarize(res = mean(res, na.rm = TRUE), norm0.5 = mean(norm0.5, na.rm = TRUE)) %>% 
+  dplyr::summarize(res = mean(bt_res, na.rm = TRUE), bt_norm = mean(bt_norm, na.rm = TRUE)) %>% 
   ungroup
@@
 # linear model of concentration x biomass
-mod <- lm(log(1 + norm0.5) ~ log(1 + biomass), data = toplo) %>% 
+mod <- lm(log(1 + bt_norm) ~ log(1 + biomass), data = toplo) %>% 
   coefficients
@@
 xlab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
-ylab <- expression(paste("Flow-normalized chlorophyll (   ", italic(mu), "g ", L^-1, ")"))
+ylab <- expression(paste("Flow-normalized chlorophyll (", italic(mu), "g ", L^-1, ")"))
 
 # scatter plot of concetration x biomass with model 
-p3 <- ggplot(toplo, aes(x = biomass, y = norm0.5)) + 
+p3 <- ggplot(toplo, aes(x = biomass, y = bt_norm)) + 
   geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
@@
     ) +
-  scale_colour_manual(label = 'log(1 + chl) = 1.1 - 0.09log(1 + biomass)', values = 'black') +
+  scale_colour_manual(label = 'log(1 + chl) = 1.35 - 0.1log(1 + biomass)', values = 'black') +
   scale_x_continuous(xlab) + 
@@
 
+##
 # dynaplot for chl, june
@@
   scale_x_continuous('ln-Salinity (psu)') + 
-  scale_y_continuous(ylab, limits = c(0.4, 3.8)) +
+  scale_y_continuous(ylab, limits = c(0.4, 3.3)) +
+  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) + 
   annotate('text', x = txt[4, 'x'], y = txt[4, 'y'], label = txt[4, 'txt'], size = size)
@@
   arrangeGrob(pA, pB, ncol = 1, heights = c(1, 1)), 
-  arrangeGrob(pC, pD, ncol = 1, heights = c(1, 1)), 
-  ncol = 2, widths = c(1, 0.6)
+  arrangeGrob(pC, pD, ncol = 1, heights = c(1.05, 1)), 
+  ncol = 2, widths = c(1, 1)
   )
@@
   ptheme + 
-  theme(legend.position = 'top', axis.title.y = element_blank())
+  theme(legend.position = 'top', axis.title.y = element_blank()) + 
+  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10))
 pleg2 <- g_legend(pdinsy1)

### 2016-07-13 R/dat_proc.R

@@ -6,69 +6,6 @@ library(tidyr)
 library(dplyr)
 
 ######
-# a smoother chlorophyll model for d7 
-
-data(diat_dat)
-data(delt_dat)
-
-# resvar label lookup
-lablk <- list(
-  shrt = c('chl', 'sio2'),
-  lngs = c(
-    expression(paste('ln-chlorophyll a (ug ', L^-1, ')')),
-    expression(paste('ln-silicon dioxide (mg ', L^-1, ')'))
-    )
-  )
-
-# flovar label lookup
-stalk <- list(
-  shrt = c('sal'),
-  lngs = c('Salinity')
-  )
-
-i <- 4
-
-# data, response variable label
-dat <- diat_dat[i, ]$data[[1]]
-resvar <- diat_dat[i, ]$resvar
-flovar <- diat_dat[i, ]$flovar
-sta <- diat_dat[i, ]$Site_Code
-reslab <- with(lablk, lngs[shrt == resvar])
-flolab <- with(stalk, shrt[lngs == flovar])
-
-# prep data as tidal object
-tomod <- select(dat, Date, resval, floval, lim) %>% 
-  rename(
-    res = resval, 
-    flo = floval
-  ) %>% 
-  data.frame %>% 
-  tidal(., 
-    reslab = reslab, 
-    flolab = expression(paste('ln-flow (standardized)'))
-  )
-
-# salinity data at D7 to pred
-topred <- filter(delt_dat, Site_Code == 'D7') %>% 
-    mutate(flo = log(1 + sal)) %>% # salinity is only variable with zeroes
-    rename(date = Date) %>% 
-    select(date, flo) %>% 
-    filter(date >= min(tomod$date) & date <= max(tomod$date)) %>% 
-    na.omit %>% 
-    data.frame
-
-# create model and exit
-mod <- wrtds(tomod, tau = c(0.1, 0.5, 0.9), wins = list(0.5, 20, 0.5), flo_div = 30, min_obs = 150)
-
-# get predictions, norms from obs flow data
-out <- mod %>% 
-  respred(dat_pred = topred) %>% 
-  resnorm
-
-chld7 <- out
-save(chld7, file = 'data/chld7.RData', compress = 'xz')
-
-######
 # a smoother din model for c10
 data(mods_nolag)
 data(flow_dat)

### 2016-07-18 sftrends_manu.Rnw
load figure for tracy, needs caption

@@
 
-\subsubsection{Clam invasion in Suisun Bay}
+\subsubsection{Effects of wastewater treatment}
+
+\subsubsection{Effects of biological invasions}
 
@@
 
+<<tracy, fig.height = 5, fig.width = 8, out.width = '\\textwidth', fig.cap = 'test'>>=
+data(potw_load)
+
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(),
+    axis.line.y = element_line(), 
+    axis.title.x = element_blank()
+  )
+lncol <- RColorBrewer::brewer.pal(9, 'Set1')[2]
+
+toplo1 <- filter(potw_load, loc %in% 'tracy' & var %in% c('tn'))
+toplo2 <- filter(potw_load, loc %in% 'tracy' & var %in% c('nh4', 'no3', 'orgn'))
+
+ylab <- expression(paste("Total nitrogen load (kg ", d^-1, ")"))
+leglabs <- list(
+  expression(NH [4] ^ '+'),
+  expression(NO [3] ^ '2-'),
+  expression('organic N')
+  )
+
+p1 <- ggplot(toplo1, aes(x = date, y = val)) + 
+  geom_line(colour = lncol) + 
+  scale_x_date(expand = c(0, 0)) + 
+  scale_y_continuous(ylab, expand = c(0, 0), limits = c(50, 850)) + 
+  ptheme
+
+p2 <- ggplot(toplo2, aes(x = date, y = val, group = var, fill = var)) + 
+  geom_area(stat = 'identity', position = 'fill') + 
+  scale_x_date(expand = c(0, 0)) + 
+  scale_fill_brewer(palette = 'Blues', labels = leglabs) + 
+  scale_y_continuous('% of total nitrogen', expand = c(0, 0)) + 
+  ptheme + 
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank()
+    )
+
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3])
+
+# Set the widths
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+
+grid.arrange(pA, pB, ncol = 1, heights = c(0.8, 1))
+@
+
 <<clmchl, fig.height = 7, fig.width = 8, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
@@
   x = c(as.numeric(as.Date('1979-10-01')), 2500, 25, 0.25),
-  y =  c(10000, 15, 4, 3.3), 
+  y =  c(10000, 14, 4, 3.3), 
   txt = c('(a)', '(c)', '(b)', '(d)')

### 2016-07-19 sftrends_manu.Rnw
tracy plot finished, h2dat added

@@
 library(Hmisc)
+library(ggrepel)
 @
@@
 
+Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:tracy}).  Nitrate is the primary constituent of total nitrogen after 2007.
+
 \subsubsection{Effects of biological invasions}
@@
 txt <- data.frame(
-  x = min(toplo$date), 
+  x = max(toplo$date), 
   y =  c(5, 2.4, 0.5, 7), 
-  txt = c('a', 'b', 'c', 'd')
+  txt = c('(a)', '(b)', '(c)', '(d)')
 )
@@
 
-<<tracy, fig.height = 5, fig.width = 8, out.width = '\\textwidth', fig.cap = 'test'>>=
+<<tracy, fig.height = 5, fig.width = 8, out.width = '\\textwidth', fig.cap = 'Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in May, 2007 to include nitrification/denitrification and tertiary filtration causing a reduction in total nitrogen effluent discharged to the Delta.
+Reductions were primarily observed for ammonium.'>>=
 data(potw_load)
@@
 
-p1 <- ggplot(toplo1, aes(x = date, y = val)) + 
-  geom_line(colour = lncol) + 
+lndate <- as.Date('2007/05/01')
+
+p1 <- ggplot(toplo1) + 
+  geom_line(colour = lncol, aes(x = date, y = val)) + 
   scale_x_date(expand = c(0, 0)) + 
   scale_y_continuous(ylab, expand = c(0, 0), limits = c(50, 850)) + 
+  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
+  geom_label_repel(
+    data = data.frame(x = lndate, y = 600), 
+    aes(x, y, label = 'WWTP upgrades'), 
+    box.padding = grid::unit(2, "lines"), 
+    nudge_x = 1
+    ) + 
   ptheme
@@
   scale_y_continuous('% of total nitrogen', expand = c(0, 0)) + 
+  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
   ptheme + 

### 2016-07-19 sftrends_manu.tex
update clams plot with biomass flow reg
load figure for tracy, needs caption
tracy plot finished, h2dat added

@@
   \vspace{1in} 
-  \\ Version Date:   Wed Jul 13 15:18:11 2016 -0500
+  \\ Version Date:   Mon Jul 18 18:28:15 2016 -0500
 	}
@@
 
-\subsubsection{Clam invasion in Suisun Bay}
+\subsubsection{Effects of wastewater treatment}
+
+Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:tracy}).  Nitrate is the primary constituent of total nitrogen after 2007.
+
+\subsubsection{Effects of biological invasions}
 
@@
 Model results are for the fiftieth conditional quantile of \ac{din}.}\label{fig:dinc10dyna}
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\textwidth]{figs/tracy-1}
+
+}
+
+\caption[Nitrogen load measurements (kg d]{Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in May, 2007 to include nitrification/denitrification and tertiary filtration causing a reduction in total nitrogen effluent discharged to the Delta.
+Reductions were primarily observed for ammonium.}\label{fig:tracy}
 \end{figure}

### 2016-07-19 sftrends_manu.Rnw
p8 trends figure done

@@
 
+<<p8trnds, fig.height = 8, fig.width = 9, out.width = '\\textwidth', fig.cap = 'Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \\ac{wrtds} (a, bottom), and relationships with flow over time (b).
+Nitrite/nitrate trends are on the left and ammonium trends are on the right.
+Wastewater treatment plant upgrades at the City of Tracy (San Joaquin County), were completed in May 2007 (\\cref{fig:tracy}), coincident with a dramatic decrease in ammonium at P8.'>>=
+data(h2dat)
+
+nomod <-  filter(h2dat, resvar == 'no23') %>% 
+  .$mod %>% 
+  .[[1]]
+
+nhmod <- filter(h2dat, resvar == 'nh') %>% 
+  .$mod %>% 
+  .[[1]]
+
+# globals
+ylab1 <- expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-', ' (mg ', L^-1, ')'))
+ylab2 <- expression(paste(NH [4] ^ '2+', ' (mg ', L^-1, ')'))
+flolab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(),
+    axis.line.y = element_line()
+  )
+col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
+dynacol_vec <- RColorBrewer::brewer.pal(9, 'Spectral')
+xlims <- range(nomod$date)
+
+# observed time series
+toplo1 <- select(nomod, date, res) %>% 
+  na.omit()
+toplo2 <- select(nhmod, date, res) %>% 
+  na.omit()
+p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
+  geom_line() + 
+  ptheme + 
+  scale_y_continuous(ylab1) +
+  theme(
+    axis.text.x = element_blank(),
+    axis.title.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, -12, 6), 'pt')
+  ) +
+  # geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
+  scale_x_date(limits = xlims) 
+p2 <- ggplot(toplo2, aes(x = date, y = exp(res))) + 
+  geom_line() + 
+  ptheme + 
+  scale_y_continuous(ylab2) +
+  theme(
+    axis.text.x = element_blank(),
+    axis.title.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, -12, 6), 'pt')
+  ) +
+  # geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
+  scale_x_date(limits = xlims)
+
+# prdnrmplots
+p3 <- prdnrmplot(nomod, logspace = F, col_vec = col_vec, min_mo = 10) + 
+  ptheme + 
+  scale_y_continuous(ylab1) +
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank(),
+    axis.title.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, 6, 6), 'pt')
+  ) +
+  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
+  scale_x_date(limits = xlims) 
+pleg1 <- g_legend(p3)
+p3 <- p3 + theme(legend.position = 'none')
+p4 <- prdnrmplot(nhmod, logspace = F, col_vec = col_vec, min_mo = 10) + 
+  ptheme + 
+  scale_y_continuous(ylab2) +
+  theme(
+    legend.position = 'none',
+    legend.title = element_blank(),
+    axis.title.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, 6, 6), 'pt')
+  ) +
+  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
+  scale_x_date(limits = xlims)
+
+# dynaplots
+p5 <- dynaplot(nomod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F) + 
+  ptheme + 
+  scale_y_continuous(ylab1) +
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank()
+  ) +
+  scale_x_continuous(flolab) + 
+  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10))
+  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
+pleg2 <- g_legend(p5)
+p5 <- p5 + theme(legend.position = 'none')
+p6 <- dynaplot(nhmod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F) + 
+  ptheme + 
+  scale_y_continuous(ylab2) +
+  scale_x_continuous(flolab) +
+  theme(
+    legend.position = 'none', 
+    legend.title = element_blank()
+  )# +
+  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
+
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+pC <- ggplot_gtable(ggplot_build(p3))
+pD <- ggplot_gtable(ggplot_build(p4))
+pE <- ggplot_gtable(ggplot_build(p5))
+pF <- ggplot_gtable(ggplot_build(p6))
+
+maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[2:3], pD$widths[2:3], pE$widths[2:3], pF$widths[2:3])
+
+# Set the widths
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+pC$widths[2:3] <- maxWidth
+pD$widths[2:3] <- maxWidth
+pE$widths[2:3] <- maxWidth
+pF$widths[2:3] <- maxWidth
+
+grid.arrange(
+  textGrob('(a) Observed, predicted, and flow-normalized trends'),
+  arrangeGrob(pA, pB, ncol = 2), 
+  pleg1,
+  arrangeGrob(pC, pD, ncol = 2),
+  textGrob('(b) Nutrient/flow relationships over time'),
+  pleg2, 
+  arrangeGrob(pE, pF, ncol = 2), 
+  ncol = 1,
+  heights = c(0.2, 0.8, 0.25, 1, 0.2, 0.25, 2)
+  )
+@
+
 <<clmchl, fig.height = 7, fig.width = 8, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.

### 2016-07-19 sftrends_manu.tex
p8 trends figure done

@@
   \vspace{1in} 
-  \\ Version Date:   Mon Jul 18 18:28:15 2016 -0500
+  \\ Version Date:   Tue Jul 19 09:56:38 2016 -0500
 	}
@@
 
+{\centering \includegraphics[width=\textwidth]{figs/p8trnds-1}
+
+}
+
+\caption{Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \ac{wrtds} (a, bottom), and relationships with flow over time (b).
+Nitrite/nitrate trends are on the left and ammonium trends are on the right.
+Wastewater treatment plant upgrades at the City of Tracy (San Joaquin County), were completed in May 2007 (\cref{fig:tracy}), coincident with a dramatic decrease in ammonium at P8.}\label{fig:p8trnds}
+\end{figure}
+
+\begin{figure}[!ht]
+
 {\centering \includegraphics[width=\textwidth]{figs/clmchl-1}

### 2016-07-19 sftrends_manu.Rnw
second case study figs and tabs done

@@
 \acrodef{wrtds}[WRTDS]{Weighted Regressions on Time, Discharge, and Season}
+\acrodef{wwtp}[WWTP]{wastewater treatment plant}
 
@@
 
+What was reduction in ammonium starting in 2002?
+
 \subsubsection{Effects of biological invasions}
@@
 \section{Discussion}
+
+Trends as percent change depend on the mean value, lower values will have larger percent changes.
 
@@
 ylab1 <- expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-', ' (mg ', L^-1, ')'))
-ylab2 <- expression(paste(NH [4] ^ '2+', ' (mg ', L^-1, ')'))
+ylab2 <- expression(paste(NH [4] ^ '+', ' (mg ', L^-1, ')'))
 flolab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
@@
 cols <- RColorBrewer::brewer.pal(9, 'Set1')[1:3]
+col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
 
 ## first set of plots
-pdin1 <- prdnrmplot(toplo1$mod[[2]], col_vec = cols[2], logspace = F, min_mo = 10) +
+pdin1 <- prdnrmplot(toplo1$mod[[2]], col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-pchl1 <- prdnrmplot(toplo1$mod[[1]], col_vec = cols[2], logspace = F, min_mo = 10) +
+pchl1 <- prdnrmplot(toplo1$mod[[1]], col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-psio1 <- prdnrmplot(toplo1$mod[[3]], col_vec = cols[2], logspace = F, min_mo = 10) +
+psio1 <- prdnrmplot(toplo1$mod[[3]], col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 ## second set of plots
-pdin2 <- prdnrmplot(toplo2$mod[[2]], col_vec = cols[2], logspace = F, min_mo = 10) +
+pdin2 <- prdnrmplot(toplo2$mod[[2]], col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-pchl2 <- prdnrmplot(toplo2$mod[[1]], col_vec = cols[2], logspace = F, min_mo = 10) +
+pchl2 <- prdnrmplot(toplo2$mod[[1]], col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-psio2 <- prdnrmplot(toplo2$mod[[3]], col_vec = cols[2], logspace = F, min_mo = 10) +
+psio2 <- prdnrmplot(toplo2$mod[[3]], col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
+% trends in no23, nh at p8
+<<p8chg>>=
+
+data(h2dat)
+dt <- as.Date('2007-05-01')
+
+noraw_bf <- filter(h2dat, resvar %in% 'no23') %>% 
+  .$mod %>% 
+  .[[1]] %>% 
+  .[which(.$date < dt), ]
+noraw_af <- filter(h2dat, resvar %in% 'no23') %>% 
+  .$mod %>% 
+  .[[1]] %>% 
+  .[which(.$date >= dt), ]
+nhraw_bf <- filter(h2dat, resvar %in% 'nh') %>% 
+  .$mod %>% 
+  .[[1]] %>% 
+  .[which(.$date < dt), ]
+nhraw_af <- filter(h2dat, resvar %in% 'nh') %>% 
+  .$mod %>% 
+  .[[1]] %>% 
+  .[which(.$date >= dt), ]
+
+mobrks <- c(0, 3, 6, 9, 12)
+yrbrks_bf <- c(1975, 2007)
+yrbrks_af <- c(2006, 2013)
+molabs <- c('JFM', 'AMJ', 'JAS', 'OND')
+yrlabs_bf <- '1976-2007'
+yrlabs_af <- '2008-2013'
+faclevs <- c('1976-2007_before', '2008-2013_after', paste0(molabs, '_before'), paste0(molabs, '_after'))
+
+# summarize din, sio2, chl trends
+no_bf <- wrtdstrnd(noraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf, aves = TRUE, min_mo = 10) %>% 
+  mutate(period = 'before', var = 'no')
+no_af <- wrtdstrnd(noraw_af, mobrks, yrbrks_af, molabs, yrlabs_af, aves = TRUE, min_mo = 10) %>% 
+  mutate(period = 'after', var = 'no')
+nh_bf <- wrtdstrnd(nhraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf, aves = TRUE, min_mo = 10) %>% 
+  mutate(period = 'before', var = 'nh')
+nh_af <- wrtdstrnd(nhraw_af, mobrks, yrbrks_af, molabs, yrlabs_af, aves = TRUE, min_mo = 10) %>% 
+  mutate(period = 'after', var = 'nh')
+
+# combine for trends for table
+totab <- rbind(no_bf, no_af, nh_bf, nh_af) %>% 
+  mutate(
+    chg = round(chg, 1), 
+    chg = ifelse(chg < 0, chg, paste0('\\textit{\\textbf{', chg , '}}')), 
+    ave = as.character(round(ave, 2))
+    ) %>% 
+  gather('est', 'val', chg:ave) %>% 
+  unite('col', var, est) %>% 
+  spread(col, val) %>% 
+  unite('period', cat, period) %>% 
+  mutate(period = factor(period, levels = faclevs, labels = faclevs)) %>% 
+  arrange(period) %>% 
+  mutate(cat = c(yrlabs_bf, yrlabs_af, molabs, molabs)) %>% 
+  data.frame(sep = c(rep('Annual', 2), rep('Seasonal, pre', 4), rep('Seasonal, post', 4)), .) %>% 
+  select(sep, cat, no_ave, no_chg, nh_ave, nh_chg)
+
+# final table formatting
+segs <- unique(totab$sep)
+cats <- totab$cat
+totab <- totab[,-c(1:2)]
+vars <- c('NO$_{2}^{-}$/NO$_{3}^{2-}$', 'NH$_{4}^{+}$')
+names(totab) <- rep(c('Mean', '\\% change'), 2)
+
+cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8.
+Upgrades were completed in May 2007 at the City of Tracy \\ac{wwtp} (San Joaquin County, \\cref{fig:tracy}).
+Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).
+Increasing values are in bold-italics.'
+
+latex( 
+  totab,
+  file = '',
+  rowlabel = 'Period',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = segs,
+  n.rgroup = c(2, rep(4, 2)),
+  cgroup = vars,
+  n.cgroup = c(2, 2),
+  rowname = cats,
+  label = 'tab:p8chg'
+  )
+
+@
+
 % trends in din, chl, sio2 at d7
@@
 
-mobrks <- c(-Inf, 3, 6, 9, Inf)
-yrbrks <- c(-Inf, 1985, 1994, 2003, Inf)
+mobrks <- c(0, 3, 6, 9, 12)
+yrbrks <- c(1975, 1985, 1994, 2003, 2013)
 molabs <- c('JFM', 'AMJ', 'JAS', 'OND')
@@
 # summarize din, sio2, chl trends
-din <- wrtdstrnd(dinraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
-din <- wrtdstrnd(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+din <- wrtdstrnd(dinraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE, min_mo = 10)
+din <- wrtdstrnd(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE, min_mo = 10) %>% 
   .[1, ] %>% 
@@
   data.frame(res = 'din', .)
-chl <- wrtdstrnd(chlraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
-chl <- wrtdstrnd(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE, min_mo = 10)
+chl <- wrtdstrnd(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE, min_mo = 10) %>% 
   .[1, ] %>% 
@@
   data.frame(res = 'chl', .)
-sio <- wrtdstrnd(sioraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE)
-sio <- wrtdstrnd(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE) %>% 
+sio <- wrtdstrnd(sioraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE, min_mo = 10)
+sio <- wrtdstrnd(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE, min_mo = 10) %>% 
   .[1, ] %>% 
@@
 
-cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station D7.
+cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
 Summaries are means and percent changes based on annual means within the time periods.

### 2016-07-19 sftrends_manu.tex
second case study figs and tabs done

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Jul 19 09:56:38 2016 -0500
+  \\ Version Date:   Tue Jul 19 14:18:21 2016 -0500
 	}
@@
 
+% trends in no23, nh at p8
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg")%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
+Upgrades were completed in May 2007 at the City of Tracy \ac{wwtp} (San Joaquin County, \cref{fig:tracy}).
+Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).
+Increasing values are in bold-italics.\label{tab:p8chg}}
+\begin{center}
+\begin{tabular}{lllcll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
+\cline{2-3} \cline{5-6}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\hline
+{\bfseries Annual}&&&&&\tabularnewline
+~~1976-2007&1.27&\textit{\textbf{75.7}}&&0.33&\textit{\textbf{71.3}}\tabularnewline
+~~2008-2013&1.47&-6&&0.11&-62.6\tabularnewline
+\hline
+{\bfseries Seasonal, pre}&&&&&\tabularnewline
+~~JFM&1.46&\textit{\textbf{77.7}}&&0.49&\textit{\textbf{57.1}}\tabularnewline
+~~AMJ&1&\textit{\textbf{75.3}}&&0.13&\textit{\textbf{63.4}}\tabularnewline
+~~JAS&1.08&\textit{\textbf{97.9}}&&0.13&\textit{\textbf{151.9}}\tabularnewline
+~~OND&1.53&\textit{\textbf{81.9}}&&0.54&\textit{\textbf{102.4}}\tabularnewline
+\hline
+{\bfseries Seasonal, post}&&&&&\tabularnewline
+~~JFM&1.76&-3.4&&0.14&-47\tabularnewline
+~~AMJ&0.94&-5.5&&0.07&-45.3\tabularnewline
+~~JAS&1.17&-10&&0.07&-50.3\tabularnewline
+~~OND&1.99&-4.6&&0.15&-58.5\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
+
 % trends in din, chl, sio2 at d7
@@
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station D7.
+\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
 Summaries are means and percent changes based on annual means within the time periods.
@@
 {\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&0.4&-4.2&&4.6&-82.4&&12.5&\textit{\textbf{30}}\tabularnewline
+~~1976-2013&0.4&\textit{\textbf{27.8}}&&4.4&-81.6&&12.6&\textit{\textbf{30.2}}\tabularnewline
 \hline
 {\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&0.4&-15.2&&10.2&-67.9&&10.7&\textit{\textbf{8.4}}\tabularnewline
-~~1986-1994&0.4&\textit{\textbf{15.4}}&&3.3&-48.8&&12.3&\textit{\textbf{13.5}}\tabularnewline
-~~1995-2003&0.5&\textit{\textbf{4.3}}&&2.1&-3.9&&13.5&\textit{\textbf{3.7}}\tabularnewline
-~~2004-2013&0.5&-10.6&&2.5&\textit{\textbf{31.7}}&&13.7&-1.8\tabularnewline
+~~1976-1985&0.4&\textit{\textbf{10.2}}&&9.8&-63.9&&10.8&\textit{\textbf{9}}\tabularnewline
+~~1986-1994&0.4&\textit{\textbf{14.9}}&&3.4&-48.8&&12.2&\textit{\textbf{14.2}}\tabularnewline
+~~1995-2003&0.5&\textit{\textbf{4}}&&2.1&-4.8&&13.5&\textit{\textbf{3.7}}\tabularnewline
+~~2004-2013&0.5&-8.6&&2.5&\textit{\textbf{28.5}}&&13.7&-1.8\tabularnewline
 \hline
@@
 {\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&1.3&-16.9&&27.6&-4.9&&14.5&-23.1\tabularnewline
+~~1976-2013&1.3&-16.5&&27.5&\textit{\textbf{9}}&&14.4&-21.7\tabularnewline
 \hline
 {\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&1.3&\textit{\textbf{16.6}}&&26.6&-20.5&&15.7&-5\tabularnewline
-~~1986-1994&1.4&\textit{\textbf{6.6}}&&27.5&\textit{\textbf{23.7}}&&14.8&-3.9\tabularnewline
-~~1995-2003&1.4&-1.1&&29.3&-15.2&&14.1&-5.1\tabularnewline
-~~2004-2013&1.2&-32.6&&27.3&\textit{\textbf{11}}&&13&-9.6\tabularnewline
+~~1976-1985&1.3&\textit{\textbf{12.1}}&&26.1&-3.3&&15.7&-4.3\tabularnewline
+~~1986-1994&1.4&\textit{\textbf{6.3}}&&27.5&\textit{\textbf{23.5}}&&14.8&-4\tabularnewline
+~~1995-2003&1.4&-0.8&&29.3&-15.3&&14.1&-5\tabularnewline
+~~2004-2013&1.2&-29.4&&27.1&\textit{\textbf{5.1}}&&13.1&-8.7\tabularnewline
 \hline

### 2016-07-20 sftrends_manu.Rnw
some text about second case study, new h2dat file

@@
 \acrodef{din}[DIN]{dissolved inorganic nitrogen}
+\acrodef{jas}[JAS]{July-August-September}
 \acrodef{sfe}[SFE]{San Francisco Estuary}
@@
 \subsubsection{Effects of wastewater treatment}
-Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) shift in the flow/nutrient relatinship before and after upgrade related to change in load contributions.
+Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) different nitrogen species will have different changes depending on change in load outputs.
 See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
@@
 
-Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:tracy}).  Nitrate is the primary constituent of total nitrogen after 2007.
-
+Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:tracy}).
+Nitrate is the primary constituent of total nitrogen after 2007.
+Organic nitrogen is a larger percentage of the total after nitrification.
 What was reduction in ammonium starting in 2002?
+
+Nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}), with ammonium showing the largest reduction.
+Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
+Percent changes are shown in \cref{tab:p8chg}, where both nitrogen species shows large percent increases prior to \ac{wwtp} upgrades followed by decreases after upgrades with ammonium showing the largest pecentage.
+Seasonally, increases prior to upgrades were most apparent in the \ac{jas} months for both analytes.
+Seasonal reductions post-upgrades were also largest in \ac{jas} for nitrite/nitrate, whereas percent reductions were similar across all monthly groupings for ammonium.
+
+Relationships of nitrogen with flow showed the typical inverse flow/concentration dynamic with flushing at high flow, although patterns differed by nitrogen species.
+Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations in the winter.
+Additionally, strength of the flow/nutrient relationship changed throughout the time series the year where the strongest relationship differed by analyte.
+Nitrite/nitrate typically had the strongest relationship flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
 
@@
 Trends as percent change depend on the mean value, lower values will have larger percent changes.
+
+Second case study showed typical inverse relationships between nutrients and flow, more flow means greater flushing and dilution of nutrient concentrations.
+Conversely, low flow means less flushing and higher nutrient concentrations, although this may not always be observed if the available nutrients are biologically available.
+Low-flow events during warmer months show the lowest ammonium concentrations, which corresponds to seasonal maxima in chlorophyll concentration.
+A similar but weaker relationship was observed with nitrite/nitrate where increased flow was related to decreased concentration and lower concentrations overall were observed in the summer.
+However, low-flow events still had higher concentrations than high-flow events in July, as compared to ammonium which was low regardless of flow.
+This suggests that ammonium concentrations are driving phytoplankton production at P8.
+Annual trends in chlorophyll concentration (not shown) showed an overall decrease from the 1970s to present, although a slight peak is observed in the 2000s.
+This peak is likely related to the maximum ammonium concentration shown in \cref{fig:p8trnds}.
+Moreover, flow/chlorophyll relationships have generally been constant throughout the period of record such that a change in flow has not been related to a change in phytoplankton production.
+This suggests that nutrient loads that contribute to production at P8 are primarily from point sources at \ac{wwtp} outflows as a change in flow does not affect the load output.
+But what are watershed loads?
 

### 2016-07-20 sftrends_manu.tex
some text about second case study, new h2dat file

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Jul 19 14:18:21 2016 -0500
+  \\ Version Date:   Tue Jul 19 17:50:17 2016 -0500
 	}
@@
 
-Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:tracy}).  Nitrate is the primary constituent of total nitrogen after 2007.
+Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:tracy}).
+Nitrate is the primary constituent of total nitrogen after 2007.
+Organic nitrogen is a larger percentage of the total after nitrification.
+What was reduction in ammonium starting in 2002?
 
-What was reduction in ammonium starting in 2002?
+Nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}), with ammonium showing the largest reduction.
+Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
+Percent changes are shown in \cref{tab:p8chg}, where both nitrogen species shows large percent increases prior to \ac{wwtp} upgrades followed by decreases after upgrades with ammonium showing the largest pecentage.
+Seasonally, increases prior to upgrades were most apparent in the \ac{jas} months for both analytes.
+Seasonal reductions post-upgrades were also largest in \ac{jas} for nitrite/nitrate, whereas percent reductions were similar across all monthly groupings for ammonium.
+
+Relationships of nitrogen with flow showed the typical inverse flow/concentration dynamic with flushing at high flow, although patterns differed by nitrogen species.
+Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations in the winter.
+Additionally, strength of the flow/nutrient relationship changed throughout the time series the year where the strongest relationship differed by analyte.
+Nitrite/nitrate typically had the strongest relationship flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
 

### 2016-09-30 sftrends_manu.Rnw
trends map added as figure

@@
 opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
-options(replace.assign = TRUE, width = 90, digits = 1)
+options(replace.assign = TRUE, width = 90)
 
@@
 
+<<trends, fig.height = 6, fig.width = 7, out.width = '0.65\\textwidth', fig.cap = 'Percent changes in nitrogen analytes for (a) annual and (b) seasonal periods in the record.
+Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
+Points are colored for direction (red increasing, green decreasing) and sized for relative magnitude.
+Station names are shown in the top left panel.'>>=
+trnd_map(molabs = NULL, sz_rng = c(2, 8), leg_lab = '(a) Annual % change')
+trnd_map(yrlabs = NULL, sz_rng = c(2, 8), stat_labs = F, leg_lab = '(b) Seasonal % change')
+@
+
 <<dinc10, fig.height = 8, fig.width = 7, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
@@
 Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.'>>=
+
 # data
@@
     axis.line.y = element_line(), 
-    axis.title.x = element_blank()
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm") 
   )
@@
     axis.text.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, -24, 6), 'pt')
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
   ) +
@@
     axis.text.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, -12, 6), 'pt')
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
   ) +
@@
     axis.text.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, -12, 6), 'pt')
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
     ) + 
@@
     axis.line.y = element_line(), 
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"),
     axis.title.x = element_blank()
@@
     axis.line.x = element_line(), 
-    axis.line.y = element_line()
+    axis.line.y = element_line(), 
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm")
   )
@@
     axis.title.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, -12, 6), 'pt')
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
   ) +
@@
     axis.title.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 6, 6), 'pt')
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
   ) +
@@
   ncol = 1,
-  heights = c(0.2, 0.8, 0.25, 1, 0.2, 0.25, 2)
+  heights = c(0.2, 0.8, 0.25, 0.9, 0.2, 0.25, 2)
   )
@@
   theme(
+    axis.line.x = element_line(), 
     axis.line.y = element_line(), 
-    axis.line.x = element_line(),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
     # axis.title.x = element_blank(),
@@
   theme(
+    axis.line.x = element_line(), 
     axis.line.y = element_line(), 
-    axis.line.x = element_line(),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
     legend.position = 'top', 
@@
   theme(
+    axis.line.x = element_line(), 
     axis.line.y = element_line(), 
-    axis.line.x = element_line(),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"),
     legend.position = 'top', 
@@
     axis.line.y = element_line(), 
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"),  
     axis.title.x = element_blank(), 

### 2016-09-30 sftrends_manu.tex
trends map added as figure

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Jul 19 17:50:17 2016 -0500
+  \\ Version Date:   Wed Jul 20 11:11:39 2016 -0500
 	}
@@
 
+{\centering \includegraphics[width=0.65\textwidth]{figs/trends-1} 
+\includegraphics[width=0.65\textwidth]{figs/trends-2}
+
+}
+
+\caption[Percent changes in nitrogen analytes for (a) annual and (b) seasonal periods in the record]{Percent changes in nitrogen analytes for (a) annual and (b) seasonal periods in the record.
+Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
+Points are colored for direction (red increasing, green decreasing) and sized for relative magnitude.
+Station names are shown in the top left panel.}\label{fig:trends}
+\end{figure}
+
+\begin{figure}[!ht]
+
 {\centering \includegraphics[width=0.75\textwidth]{figs/dinc10-1}

### 2016-09-30 R/funcs.R

@@ -23,27 +23,41 @@ g_legend <- function(a.gplot){
 #
 # uses delt_dat, mods_nolag, delt_map
 #
-trnd_map <- function(res, 
+trnd_map <- function(res = c('din', 'nh', 'no23'), 
   mobrks = c(-Inf, 4, 8, Inf),
   yrbrks = c(-Inf, 1988, 2000, Inf),
   molabs = c('JFMA', 'MJJA', 'SOND'),
   yrlabs = c('1976-1988', '1989-2000', '2001-2012'),
   buffx = 0.00025,
   buffy = 0.001,
-  cols =  c('#E41A1C', '#4DAF4A'),
+  cols =  c('#D53E4F', '#4DAF4A'),
   sz_rng = c(2, 11),
   shps = c(24, 25),
   ndivs = 6,
   strp_fl = 'lightgrey',
   bar = FALSE, 
   stat_labs = TRUE,
-  leg = TRUE){
+  leg = TRUE, 
+  leg_lab = 'Percent change'){
 
   # load required data
   data(delt_dat)
   data(delt_map)
   data(mods_nolag)
 
+  # response labels 
+  reslabs <- list(
+    shrt = c('din', 'nh', 'no23'),
+    expr = c(
+      'DIN', 
+      expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-')),
+      expression(paste(NH [4] ^ '+'))
+    )
+  )
+  
+  # make response labels
+  mods_nolag <- mutate(mods_nolag, reslabs = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr))
+  
   ##
   # get a bounding box for the stations
   statmeta <- select(delt_dat, Site_Code, Latitude, Longitude) %>% 
@@ -83,29 +97,47 @@ trnd_map <- function(res,
   # base delta map
   pbase <- ggplot(delt_map, aes(x = long, y = lat)) + 
     geom_polygon(aes(group = group, fill = hole), colour = "cornflowerblue") +
-    theme_bw() +
+    theme_bw() + 
     theme(
-      axis.title.x=element_blank(),
-      axis.title.y=element_blank(),
-      panel.background=element_blank(),panel.grid.major=element_blank(),
-      panel.grid.minor=element_blank(),plot.background=element_blank(), 
-      legend.position = 'none'
-      ) +  
+      plot.background = element_rect(fill='transparent', 
+        colour = NA),
+      panel.background = element_rect(fill='transparent', 
+        colour = NA),
[…]

### 2016-09-30 sftrends_manu.Rnw
trends table added

@@
 
-<<trends, fig.height = 6, fig.width = 7, out.width = '0.65\\textwidth', fig.cap = 'Percent changes in nitrogen analytes for (a) annual and (b) seasonal periods in the record.
+<<trndmap, fig.height = 6, fig.width = 7, out.width = '0.65\\textwidth', fig.cap = 'Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record.
 Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
@@
 % tables
+
+% overall trends
+<<trnds>>=
+data(trnds_chg)
+data(trnds_ave)
+
+# percent changes
+tab_chg <- trnds_chg
+
+# increase in bold italic
+tab_chg[, -c(1:2)] <- round(tab_chg[, -c(1:2)], 1)
+tab_chg[, -c(1:2)][sign(tab_chg[, -c(1:2)]) != -1] <- paste0('\\textit{\\textbf{', tab_chg[, -c(1:2)][sign(tab_chg[, -c(1:2)]) != -1], '}}')
+tab_chg <- gather(tab_chg, 'var', 'chg', -Site_Code, -resvar) %>% 
+  mutate(chg = paste0('\\footnotesize{(', chg, ')}'))
+
+# averages
+tab_ave <- gather(trnds_ave, 'var', 'ave', -Site_Code, -resvar) %>% 
+  mutate(ave = round(ave, 1))
+
+tab <- left_join(tab_ave, tab_chg, by = c('Site_Code', 'resvar', 'var')) %>% 
+  unite(val, ave, chg, sep = ' ') %>% 
+  spread(var, val) %>% 
+  arrange(resvar, Site_Code)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
+Summaries are averages (mg L$^{-1}$) and percent changes in parentheses (increasing in bold-italic).
+Changes are based on the difference between the ending and starting estimates for the estimates within each period.
+See \\cref{fig:trndmap} for a summary of spatial trends.'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = c(9, 9, 9),
+  cgroup = c('Annual', 'Seasonal'),
+  n.cgroup = c(3, 3),
+  rowname = stats,
+  label = 'tab:trnds'
+  )
+
+@
 

### 2016-09-30 sftrends_manu.tex
trends table added

--- 2016-07-20 sftrends_manu.tex
@@
   \vspace{1in} 
-  \\ Version Date:   Tue Jul 19 17:50:17 2016 -0500
+  \\ Version Date:   Fri Sep 30 16:40:38 2016 -0500
 	}
@@
 
+{\centering \includegraphics[width=0.65\textwidth]{figs/trndmap-1} 
+\includegraphics[width=0.65\textwidth]{figs/trndmap-2}
+
+}
+
+\caption[Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record]{Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record.
+Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
+Points are colored for direction (red increasing, green decreasing) and sized for relative magnitude.
+Station names are shown in the top left panel.}\label{fig:trndmap}
+\end{figure}
+
+\begin{figure}[!ht]
+
 {\centering \includegraphics[width=0.75\textwidth]{figs/dinc10-1}
@@
 % tables
+
+% overall trends
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual", "Seasonal"), n.cgroup = c(3,         3), rowname = stats, label = "tab:trnds")%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
+Summaries are averages (mg L$^{-1}$) and percent changes in parentheses (increasing in bold-italic).
+Changes are based on the difference between the ending and starting estimates for the estimates within each period.
+See \cref{fig:trndmap} for a summary of spatial trends.\label{tab:trnds}}
+\begin{center}
+\begin{tabular}{llllclll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{3}{c}{\bfseries Annual}&\multicolumn{1}{c}{\bfseries }&\multicolumn{3}{c}{\bfseries Seasonal}\tabularnewline
+\cline{2-4} \cline{6-8}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1988}&\multicolumn{1}{c}{1989-2000}&\multicolumn{1}{c}{2001-2012}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{JFMA}&\multicolumn{1}{c}{MJJA}&\multicolumn{1}{c}{SOND}\tabularnewline
+\hline
+{\bfseries DIN}&&&&&&&\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{38.4}})}&1.3 \footnotesize{(\textit{\textbf{16.2}})}&1.2 \footnotesize{(-39.3)}&&1.1 \footnotesize{(-25.7)}&1.1 \footnotesize{(-10.1)}&1.5 \footnotesize{(\textit{\textbf{1.7}})}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{33.6}})}&0.4 \footnotesize{(\textit{\textbf{27.4}})}&0.4 \footnotesize{(-12.7)}&&0.4 \footnotesize{(\textit{\textbf{30.7}})}&0.3 \footnotesize{(\textit{\textbf{54.8}})}&0.5 \footnotesize{(\textit{\textbf{42.8}})}\tabularnewline
+~~D19&0.4 \footnotesize{(-20.2)}&0.4 \footnotesize{(\textit{\textbf{6.5}})}&0.4 \footnotesize{(-22.6)}&&0.6 \footnotesize{(-28.6)}&0.3 \footnotesize{(-35.1)}&0.4 \footnotesize{(-15.6)}\tabularnewline
+~~D26&0.4 \footnotesize{(-2.4)}&0.4 \footnotesize{(\textit{\textbf{10.2}})}&0.4 \footnotesize{(-15.4)}&&0.5 \footnotesize{(-10.9)}&0.3 \footnotesize{(-3.4)}&0.5 \footnotesize{(-2.6)}\tabularnewline
+~~D28&0.4 \footnotesize{(-20.5)}&0.4 \footnotesize{(\textit{\textbf{2.7}})}&0.4 \footnotesize{(-32.6)}&&0.6 \footnotesize{(-32.9)}&0.2 \footnotesize{(-53.3)}&0.4 \footnotesize{(-48.8)}\tabularnewline
+~~D4&0.3 \footnotesize{(-9.1)}&0.4 \footnotesize{(\textit{\textbf{20.7}})}&0.4 \footnotesize{(-4.3)}&&0.5 \footnotesize{(\textit{\textbf{11.4}})}&0.3 \footnotesize{(\textit{\textbf{4.8}})}&0.4 \footnotesize{(\textit{\textbf{10.7}})}\tabularnewline
+~~D6&0.4 \footnotesize{(-9.6)}&0.4 \footnotesize{(\textit{\textbf{21.9}})}&0.5 \footnotesize{(-7.1)}&&0.5 \footnotesize{(-4.7)}&0.4 \footnotesize{(\textit{\textbf{45.6}})}&0.4 \footnotesize{(\textit{\textbf{34.7}})}\tabularnewline
+~~D7&0.4 \footnotesize{(-26.4)}&0.4 \footnotesize{(\textit{\textbf{20.4}})}&0.5 \footnotesize{(-2.4)}&&0.5 \footnotesize{(-6.9)}&0.4 \footnotesize{(\textit{\textbf{19}})}&0.4 \footnotesize{(\textit{\textbf{21.4}})}\tabularnewline
+~~P8&1.1 \footnotesize{(\textit{\textbf{74.1}})}&1.5 \footnotesize{(\textit{\textbf{37.9}})}&1.6 \footnotesize{(-31)}&&1.4 \footnotesize{(\textit{\textbf{31.6}})}&1 \footnotesize{(\textit{\textbf{52.8}})}&1.8 \footnotesize{(\textit{\textbf{45.4}})}\tabularnewline
+\hline
+{\bfseries NH$_{4}^{+}$}&&&&&&&\tabularnewline
+~~C10&0.1 \footnotesize{(-18.7)}&0.1 \footnotesize{(-58.6)}&0 \footnotesize{(-43.1)}&&0.1 \footnotesize{(-78)}&0 \footnotesize{(-96.2)}&0.1 \footnotesize{(-91.7)}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{74.6}})}&0.2 \footnotesize{(\textit{\textbf{43}})}&0.3 \footnotesize{(-18.6)}&&0.2 \footnotesize{(\textit{\textbf{81.8}})}&0.2 \footnotesize{(\textit{\textbf{95.8}})}&0.3 \footnotesize{(\textit{\textbf{74.6}})}\tabularnewline
+~~D19&0.1 \footnotesize{(\textit{\textbf{14.5}})}&0.1 \footnotesize{(\textit{\textbf{7.3}})}&0 \footnotesize{(-11.6)}&&0.1 \footnotesize{(-16)}&0 \footnotesize{(-21.8)}&0.1 \footnotesize{(-4.8)}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{32}})}&0.1 \footnotesize{(\textit{\textbf{9.3}})}&0.1 \footnotesize{(-17.6)}&&0.1 \footnotesize{(-0.9)}&0.1 \footnotesize{(\textit{\textbf{7.9}})}&0.1 \footnotesize{(\textit{\textbf{18.7}})}\tabularnewline
+~~D28&0.1 \footnotesize{(-9.3)}&0 \footnotesize{(-18.2)}&0 \footnotesize{(-12.5)}&&0.1 \footnotesize{(-42.3)}&0 \footnotesize{(-24.7)}&0.1 \footnotesize{(-34.7)}\tabularnewline
+~~D4&0.1 \footnotesize{(\textit{\textbf{10.5}})}&0.1 \footnotesize{(\textit{\textbf{47.8}})}&0.1 \footnotesize{(\textit{\textbf{0.2}})}&&0.1 \footnotesize{(\textit{\textbf{66.9}})}&0 \footnotesize{(\textit{\textbf{33.6}})}&0.1 \footnotesize{(\textit{\textbf{68.3}})}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{19.9}})}&0.1 \footnotesize{(\textit{\textbf{33.3}})}&0.1 \footnotesize{(-7.3)}&&0.1 \footnotesize{(\textit{\textbf{36}})}&0.1 \footnotesize{(\textit{\textbf{56.4}})}&0.1 \footnotesize{(\textit{\textbf{45.3}})}\tabularnewline
+~~D7&0.1 \footnotesize{(-18.1)}&0.1 \footnotesize{(\textit{\textbf{34.4}})}&0.1 \footnotesize{(-0.3)}&&0.1 \footnotesize{(\textit{\textbf{21.2}})}&0 \footnotesize{(-16)}&0.1 \footnotesize{(\textit{\textbf{34}})}\tabularnewline
+~~P8&0.2 \footnotesize{(\textit{\textbf{178.6}})}&0.3 \footnotesize{(\textit{\textbf{49.2}})}&0.2 \footnotesize{(-86.1)}&&0.3 \footnotesize{(-52.7)}&0.1 \footnotesize{(-23.9)}&0.3 \footnotesize{(-61.1)}\tabularnewline
+\hline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&&&&\tabularnewline
+~~C10&1 \footnotesize{(\textit{\textbf{48.1}})}&1.2 \footnotesize{(\textit{\textbf{23}})}&1.1 \footnotesize{(-38.1)}&&1 \footnotesize{(-12.5)}&1 \footnotesize{(-0.2)}&1.4 \footnotesize{(\textit{\textbf{20.4}})}\tabularnewline
+~~C3&0.1 \footnotesize{(-4.4)}&0.1 \footnotesize{(\textit{\textbf{11.7}})}&0.2 \footnotesize{(-0.8)}&&0.2 \footnotesize{(-2.3)}&0.1 \footnotesize{(\textit{\textbf{29.4}})}&0.2 \footnotesize{(\textit{\textbf{3.3}})}\tabularnewline
+~~D19&0.4 \footnotesize{(-22)}&0.4 \footnotesize{(\textit{\textbf{11.1}})}&0.3 \footnotesize{(-23.7)}&&0.5 \footnotesize{(-28.6)}&0.2 \footnotesize{(-34)}&0.4 \footnotesize{(-16.3)}\tabularnewline
+~~D26&0.3 \footnotesize{(-9)}&0.3 \footnotesize{(\textit{\textbf{12.4}})}&0.3 \footnotesize{(-14.8)}&&0.4 \footnotesize{(-13.4)}&0.2 \footnotesize{(-11.8)}&0.3 \footnotesize{(-4.7)}\tabularnewline
+~~D28&0.4 \footnotesize{(-29.1)}&0.4 \footnotesize{(\textit{\textbf{7}})}&0.3 \footnotesize{(-34.7)}&&0.5 \footnotesize{(-34.5)}&0.2 \footnotesize{(-59.1)}&0.4 \footnotesize{(-53.8)}\tabularnewline
+~~D4&0.3 \footnotesize{(-8.2)}&0.3 \footnotesize{(\textit{\textbf{16.6}})}&0.3 \footnotesize{(-3.7)}&&0.4 \footnotesize{(\textit{\textbf{5.4}})}&0.3 \footnotesize{(\textit{\textbf{6.6}})}&0.3 \footnotesize{(\textit{\textbf{7.3}})}\tabularnewline
+~~D6&0.3 \footnotesize{(-9.5)}&0.4 \footnotesize{(\textit{\textbf{16.5}})}&0.4 \footnotesize{(-7.5)}&&0.4 \footnotesize{(-17.9)}&0.3 \footnotesize{(\textit{\textbf{42.6}})}&0.3 \footnotesize{(\textit{\textbf{27.6}})}\tabularnewline
+~~D7&0.3 \footnotesize{(-7.5)}&0.4 \footnotesize{(\textit{\textbf{20.5}})}&0.4 \footnotesize{(-7.9)}&&0.4 \footnotesize{(-12.7)}&0.3 \footnotesize{(\textit{\textbf{41.9}})}&0.3 \footnotesize{(\textit{\textbf{24.4}})}\tabularnewline
+~~P8&0.9 \footnotesize{(\textit{\textbf{61.3}})}&1.2 \footnotesize{(\textit{\textbf{32.7}})}&1.3 \footnotesize{(-12.5)}&&1.1 \footnotesize{(\textit{\textbf{52.2}})}&0.9 \footnotesize{(\textit{\textbf{60.6}})}&1.4 \footnotesize{(\textit{\textbf{67.3}})}\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
 

### 2016-09-30 R/funcs.R

@@ -161,7 +161,7 @@ trnd_map <- function(res = c('din', 'nh', 'no23'),
     )  + 
     scale_shape_manual(values = shps) + 
     scale_size(range = sz_rng) + 
-    scale_fill_manual(values=c(cols, "cornflowerblue", "aliceblue"), guide="none") +
+    scale_fill_manual(values=c(cols, "cornflowerblue", "white"), guide="none") +
     facet_grid(reslabs ~ cat, labeller = label_parsed) +
     theme(strip.background=element_rect(fill = 'white', colour = 'white'),
       panel.border = element_rect(colour = "grey")

### 2016-09-30 R/funcs.R
fixed label error in trnd_map function

@@ -50,8 +50,8 @@ trnd_map <- function(res = c('din', 'nh', 'no23'),
     shrt = c('din', 'nh', 'no23'),
     expr = c(
       'DIN', 
-      expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-')),
-      expression(paste(NH [4] ^ '+'))
+      expression(paste(NH [4] ^ '+')),
+      expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
     )
   )
   

### 2016-10-12 sftrends_manu.tex
fixed label error in trnd_map function
updating data from new wrtds preds

--- 2016-07-20 sftrends_manu.tex
@@
 \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
 
@@
   \vspace{1in} 
-  \\ Version Date:   Tue Jul 19 17:50:17 2016 -0500
+  \\ Version Date:   Fri Sep 30 17:57:52 2016 -0500
 	}
@@
 
-\caption[Nitrogen load measurements (kg d]{Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
+\caption[Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County]{Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
 Wastewater discharge requirements were implemented in May, 2007 to include nitrification/denitrification and tertiary filtration causing a reduction in total nitrogen effluent discharged to the Delta.
@@
 
-\begin{figure}[!ht]
+\begin{kframe}
 
-{\centering \includegraphics[width=\textwidth]{figs/p8trnds-1}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
 
-}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
 
-\caption{Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \ac{wrtds} (a, bottom), and relationships with flow over time (b).
-Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Tracy (San Joaquin County), were completed in May 2007 (\cref{fig:tracy}), coincident with a dramatic decrease in ammonium at P8.}\label{fig:p8trnds}
-\end{figure}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: Invalid input: date\_trans works with objects of class Date only}}
 
-\begin{figure}[!ht]
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: Invalid input: date\_trans works with objects of class Date only}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in tmp\$grobs[[leg]]: attempt to select less than one element in get1index}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}dynaplot"{}): no applicable method for 'dynaplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'p5' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'p5' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}dynaplot"{}): no applicable method for 'dynaplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'p5' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'p6' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unit.pmax(pA\$widths[2:3], pB\$widths[2:3], pC\$widths[2:3], pD\$widths[2:3], : object 'pE' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pE\$widths[2:3] <- maxWidth: object 'pE' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pF\$widths[2:3] <- maxWidth: object 'pF' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in arrangeGrob(...): object 'pleg1' not found}}\end{kframe}
+
+\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): 'yr' column not found in lhs, cannot join}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'bt\_norm' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in seq.default(xvals[1], xvals[2], length = 100): 'from' cannot be NA, NaN or infinite}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}dynaplot"{}): no applicable method for 'dynaplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'biomass' not found}}\end{kframe}\begin{figure}[!ht]
 
@@
 
-\begin{figure}[!ht]
+\begin{kframe}
 
-{\centering \includegraphics[width=\maxwidth]{figs/d7c10trnds-1}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
 
-}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pdin1' not found}}
 
-\caption{Flow-normalized trends of annual (left) and seasonal (right) variation in DIN, chl-{\it\footnotesize a}, and SiO$_2$ at station D7 (top) and C10 (bottom).
-Covariation between nutrients, chl-{\it\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.
-Seasonal changes at D7 are most pronounced during the summer.}\label{fig:d7c10trnds}
-\end{figure}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'pdin1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}seasyrplot"{}): no applicable method for 'seasyrplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pdinsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'pdinsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}seasyrplot"{}): no applicable method for 'seasyrplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}seasyrplot"{}): no applicable method for 'seasyrplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pdin1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pchl1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'psio1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unit.pmax(pA1\$widths[2:3], pB1\$widths[2:3], pC1\$widths[2:3]): object 'pA1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pA1\$widths[2:3] <- maxWidth: object 'pA1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pB1\$widths[2:3] <- maxWidth: object 'pB1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pC1\$widths[2:3] <- maxWidth: object 'pC1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pdinsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pchlsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'psiosy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unit.pmax(pAsy1\$widths[2:3], pBsy1\$widths[2:3], pCsy1\$widths[2:3]): object 'pAsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pAsy1\$widths[2:3] <- maxWidth: object 'pAsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pBsy1\$widths[2:3] <- maxWidth: object 'pBsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pCsy1\$widths[2:3] <- maxWidth: object 'pCsy1' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}seasyrplot"{}): no applicable method for 'seasyrplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}seasyrplot"{}): no applicable method for 'seasyrplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}prdnrmplot"{}): no applicable method for 'prdnrmplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}seasyrplot"{}): no applicable method for 'seasyrplot' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pdin2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pchl2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'psio2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unit.pmax(pA2\$widths[2:3], pB2\$widths[2:3], pC2\$widths[2:3]): object 'pA2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pA2\$widths[2:3] <- maxWidth: object 'pA2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pB2\$widths[2:3] <- maxWidth: object 'pB2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pC2\$widths[2:3] <- maxWidth: object 'pC2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pdinsy2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'pchlsy2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in plot\_clone(plot): object 'psiosy2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unit.pmax(pAsy2\$widths[2:3], pBsy2\$widths[2:3], pCsy2\$widths[2:3]): object 'pAsy2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pAsy2\$widths[2:3] <- maxWidth: object 'pAsy2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pBsy2\$widths[2:3] <- maxWidth: object 'pBsy2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in pCsy2\$widths[2:3] <- maxWidth: object 'pCsy2' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in arrangeGrob(pleg1, pleg2, ncol = 2): object 'pleg1' not found}}\end{kframe}
 
@@
 
-% trends in no23, nh at p8
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg")%
+% overall trends
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual", "Seasonal"), n.cgroup = c(3,         3), rowname = stats, label = "tab:trnds")%
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in May 2007 at the City of Tracy \ac{wwtp} (San Joaquin County, \cref{fig:tracy}).
-Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).
-Increasing values are in bold-italics.\label{tab:p8chg}}
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
+Summaries are averages (mg L$^{-1}$) and percent changes in parentheses (increasing in bold-italic).
+Changes are based on the difference between the ending and starting estimates for the estimates within each period.
+See \cref{fig:trndmap} for a summary of spatial trends.\label{tab:trnds}}
 \begin{center}
-\begin{tabular}{lllcll}
+\begin{tabular}{llllclll}
 \hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
-\cline{2-3} \cline{5-6}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{3}{c}{\bfseries Annual}&\multicolumn{1}{c}{\bfseries }&\multicolumn{3}{c}{\bfseries Seasonal}\tabularnewline
+\cline{2-4} \cline{6-8}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1988}&\multicolumn{1}{c}{1989-2000}&\multicolumn{1}{c}{2001-2012}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{JFMA}&\multicolumn{1}{c}{MJJA}&\multicolumn{1}{c}{SOND}\tabularnewline
 \hline
-{\bfseries Annual}&&&&&\tabularnewline
-~~1976-2007&1.27&\textit{\textbf{75.7}}&&0.33&\textit{\textbf{71.3}}\tabularnewline
-~~2008-2013&1.47&-6&&0.11&-62.6\tabularnewline
+{\bfseries DIN}&&&&&&&\tabularnewline
+~~C10&1.3 \footnotesize{(\textit{\textbf{29.9}})}&1.4 \footnotesize{(\textit{\textbf{17.6}})}&1.3 \footnotesize{(-39)}&&1.4 \footnotesize{(-25.6)}&1.1 \footnotesize{(-10)}&1.5 \footnotesize{(-0.2)}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{31}})}&0.4 \footnotesize{(\textit{\textbf{23.5}})}&0.5 \footnotesize{(-7.9)}&&0.4 \footnotesize{(\textit{\textbf{29.5}})}&0.4 \footnotesize{(\textit{\textbf{78.8}})}&0.5 \footnotesize{(\textit{\textbf{45.1}})}\tabularnewline
+~~D19&0.5 \footnotesize{(-16.5)}&0.5 \footnotesize{(\textit{\textbf{7}})}&0.4 \footnotesize{(-27.5)}&&0.7 \footnotesize{(-35.9)}&0.3 \footnotesize{(-25.5)}&0.4 \footnotesize{(-24.6)}\tabularnewline
+~~D26&0.4 \footnotesize{(-12)}&0.5 \footnotesize{(\textit{\textbf{11.5}})}&0.4 \footnotesize{(-19.8)}&&0.6 \footnotesize{(-24.6)}&0.3 \footnotesize{(-9.7)}&0.4 \footnotesize{(-6.7)}\tabularnewline
+~~D28&0.5 \footnotesize{(-16.7)}&0.5 \footnotesize{(\textit{\textbf{5.9}})}&0.4 \footnotesize{(-41.1)}&&0.7 \footnotesize{(-35.8)}&0.3 \footnotesize{(-49)}&0.4 \footnotesize{(-52.3)}\tabularnewline
+~~D4&0.4 \footnotesize{(-23.7)}&0.4 \footnotesize{(\textit{\textbf{23.7}})}&0.4 \footnotesize{(-8.4)}&&0.5 \footnotesize{(-14.3)}&0.3 \footnotesize{(\textit{\textbf{1.7}})}&0.4 \footnotesize{(\textit{\textbf{1.6}})}\tabularnewline
+~~D6&0.4 \footnotesize{(-9)}&0.5 \footnotesize{(\textit{\textbf{22.6}})}&0.5 \footnotesize{(-7.8)}&&0.5 \footnotesize{(-15.8)}&0.4 \footnotesize{(\textit{\textbf{51}})}&0.4 \footnotesize{(\textit{\textbf{18.9}})}\tabularnewline
+~~D7&0.4 \footnotesize{(-21.5)}&0.5 \footnotesize{(\textit{\textbf{22.9}})}&0.5 \footnotesize{(-10.1)}&&0.5 \footnotesize{(-29.9)}&0.4 \footnotesize{(\textit{\textbf{28.5}})}&0.4 \footnotesize{(-0.2)}\tabularnewline
+~~P8&1.2 \footnotesize{(\textit{\textbf{60.1}})}&1.7 \footnotesize{(\textit{\textbf{36.7}})}&1.8 \footnotesize{(-32.7)}&&1.9 \footnotesize{(\textit{\textbf{37}})}&1.1 \footnotesize{(\textit{\textbf{39.1}})}&1.8 \footnotesize{(\textit{\textbf{50.1}})}\tabularnewline
 \hline
-{\bfseries Seasonal, pre}&&&&&\tabularnewline
-~~JFM&1.46&\textit{\textbf{77.7}}&&0.49&\textit{\textbf{57.1}}\tabularnewline
-~~AMJ&1&\textit{\textbf{75.3}}&&0.13&\textit{\textbf{63.4}}\tabularnewline
-~~JAS&1.08&\textit{\textbf{97.9}}&&0.13&\textit{\textbf{151.9}}\tabularnewline
-~~OND&1.53&\textit{\textbf{81.9}}&&0.54&\textit{\textbf{102.4}}\tabularnewline
+{\bfseries NH$_{4}^{+}$}&&&&&&&\tabularnewline
+~~C10&0.2 \footnotesize{(-20.2)}&0.1 \footnotesize{(-54.7)}&0 \footnotesize{(-23.2)}&&0.1 \footnotesize{(-73.6)}&0 \footnotesize{(-81.8)}&0.1 \footnotesize{(-87.9)}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{88.5}})}&0.3 \footnotesize{(\textit{\textbf{35.9}})}&0.3 \footnotesize{(-3.3)}&&0.2 \footnotesize{(\textit{\textbf{113}})}&0.3 \footnotesize{(\textit{\textbf{150.4}})}&0.3 \footnotesize{(\textit{\textbf{101.1}})}\tabularnewline
+~~D19&0.1 \footnotesize{(\textit{\textbf{29.6}})}&0.1 \footnotesize{(-10.7)}&0.1 \footnotesize{(-19.8)}&&0.1 \footnotesize{(-24.4)}&0 \footnotesize{(-22.4)}&0.1 \footnotesize{(-2.8)}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{43.6}})}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(-12.7)}&&0.1 \footnotesize{(-3.8)}&0.1 \footnotesize{(\textit{\textbf{34.6}})}&0.1 \footnotesize{(\textit{\textbf{17.1}})}\tabularnewline
+~~D28&0.1 \footnotesize{(-15.5)}&0.1 \footnotesize{(-14.9)}&0 \footnotesize{(-16.2)}&&0.1 \footnotesize{(-42.3)}&0 \footnotesize{(-34.8)}&0.1 \footnotesize{(-35.2)}\tabularnewline
+~~D4&0.1 \footnotesize{(\textit{\textbf{17.3}})}&0.1 \footnotesize{(\textit{\textbf{64}})}&0.1 \footnotesize{(-2.9)}&&0.1 \footnotesize{(\textit{\textbf{71.7}})}&0 \footnotesize{(\textit{\textbf{36.8}})}&0.1 \footnotesize{(\textit{\textbf{39}})}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{55.4}})}&0.1 \footnotesize{(\textit{\textbf{31.1}})}&0.1 \footnotesize{(-5.7)}&&0.1 \footnotesize{(\textit{\textbf{44.1}})}&0.1 \footnotesize{(\textit{\textbf{69.8}})}&0.1 \footnotesize{(\textit{\textbf{56}})}\tabularnewline
+~~D7&0.1 \footnotesize{(\textit{\textbf{36}})}&0.1 \footnotesize{(\textit{\textbf{35.6}})}&0.1 \footnotesize{(-6.1)}&&0.1 \footnotesize{(\textit{\textbf{36.5}})}&0 \footnotesize{(\textit{\textbf{27.2}})}&0.1 \footnotesize{(\textit{\textbf{11.5}})}\tabularnewline
+~~P8&0.2 \footnotesize{(\textit{\textbf{186.1}})}&0.4 \footnotesize{(\textit{\textbf{47.1}})}&0.2 \footnotesize{(-86.8)}&&0.4 \footnotesize{(-58.9)}&0.1 \footnotesize{(-21.8)}&0.3 \footnotesize{(-51.9)}\tabularnewline
 \hline
-{\bfseries Seasonal, post}&&&&&\tabularnewline
-~~JFM&1.76&-3.4&&0.14&-47\tabularnewline
-~~AMJ&0.94&-5.5&&0.07&-45.3\tabularnewline
-~~JAS&1.17&-10&&0.07&-50.3\tabularnewline
-~~OND&1.99&-4.6&&0.15&-58.5\tabularnewline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&&&&\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{34.1}})}&1.4 \footnotesize{(\textit{\textbf{23.3}})}&1.3 \footnotesize{(-39.1)}&&1.3 \footnotesize{(-16.4)}&1.1 \footnotesize{(-6)}&1.4 \footnotesize{(\textit{\textbf{18.7}})}\tabularnewline
+~~C3&0.1 \footnotesize{(-7.3)}&0.2 \footnotesize{(\textit{\textbf{6.5}})}&0.2 \footnotesize{(-9.6)}&&0.2 \footnotesize{(-14)}&0.1 \footnotesize{(\textit{\textbf{20.7}})}&0.2 \footnotesize{(-2.9)}\tabularnewline
+~~D19&0.4 \footnotesize{(-23.7)}&0.4 \footnotesize{(\textit{\textbf{9}})}&0.4 \footnotesize{(-28.1)}&&0.6 \footnotesize{(-37.3)}&0.3 \footnotesize{(-33.4)}&0.4 \footnotesize{(-33.7)}\tabularnewline
+~~D26&0.3 \footnotesize{(-21.4)}&0.4 \footnotesize{(\textit{\textbf{14.2}})}&0.3 \footnotesize{(-21.6)}&&0.5 \footnotesize{(-28.6)}&0.3 \footnotesize{(-17.4)}&0.3 \footnotesize{(-13.2)}\tabularnewline
+~~D28&0.4 \footnotesize{(-22)}&0.5 \footnotesize{(\textit{\textbf{5.6}})}&0.4 \footnotesize{(-43.7)}&&0.6 \footnotesize{(-38.5)}&0.3 \footnotesize{(-57.9)}&0.4 \footnotesize{(-59.7)}\tabularnewline
+~~D4&0.3 \footnotesize{(-26.5)}&0.3 \footnotesize{(\textit{\textbf{15.8}})}&0.3 \footnotesize{(-9.3)}&&0.4 \footnotesize{(-23.4)}&0.3 \footnotesize{(-1.7)}&0.3 \footnotesize{(-7.3)}\tabularnewline
+~~D6&0.3 \footnotesize{(-22.6)}&0.4 \footnotesize{(\textit{\textbf{19.6}})}&0.4 \footnotesize{(-8.5)}&&0.4 \footnotesize{(-25.1)}&0.3 \footnotesize{(\textit{\textbf{30.3}})}&0.3 \footnotesize{(\textit{\textbf{2.8}})}\tabularnewline
+~~D7&0.4 \footnotesize{(-30.1)}&0.4 \footnotesize{(\textit{\textbf{19.3}})}&0.4 \footnotesize{(-11.3)}&&0.4 \footnotesize{(-35.6)}&0.3 \footnotesize{(\textit{\textbf{12.7}})}&0.4 \footnotesize{(-30.6)}\tabularnewline
+~~P8&1.1 \footnotesize{(\textit{\textbf{46}})}&1.3 \footnotesize{(\textit{\textbf{30.4}})}&1.6 \footnotesize{(-16.3)}&&1.5 \footnotesize{(\textit{\textbf{60}})}&1 \footnotesize{(\textit{\textbf{40}})}&1.5 \footnotesize{(\textit{\textbf{60}})}\tabularnewline
 \hline
@@
 
+% trends in no23, nh at p8
+\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in rbind(no\_bf, no\_af, nh\_bf, nh\_af): object 'no\_bf' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in unique(totab\$sep): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in names(totab) <- rep(c("{}Mean"{}, "{}\textbackslash{}\textbackslash{}\% change"{}), 2): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in latex(totab, file = "{}"{}, rowlabel = "{}Period"{}, caption = cap.val, : object 'totab' not found}}\end{kframe}
+
 % trends in din, chl, sio2 at d7
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:d7chg")%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
-Summaries are means and percent changes based on annual means within the time periods.
-Increasing values are in bold-italics.\label{tab:d7chg}}
-\begin{center}
-\begin{tabular}{lllcllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
-\cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
-\hline
-{\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&0.4&\textit{\textbf{27.8}}&&4.4&-81.6&&12.6&\textit{\textbf{30.2}}\tabularnewline
-\hline
-{\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&0.4&\textit{\textbf{10.2}}&&9.8&-63.9&&10.8&\textit{\textbf{9}}\tabularnewline
-~~1986-1994&0.4&\textit{\textbf{14.9}}&&3.4&-48.8&&12.2&\textit{\textbf{14.2}}\tabularnewline
-~~1995-2003&0.5&\textit{\textbf{4}}&&2.1&-4.8&&13.5&\textit{\textbf{3.7}}\tabularnewline
-~~2004-2013&0.5&-8.6&&2.5&\textit{\textbf{28.5}}&&13.7&-1.8\tabularnewline
-\hline
-{\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~JFM&0.5&-4.9&&2.7&-56&&15&\textit{\textbf{4.4}}\tabularnewline
-~~AMJ&0.4&\textit{\textbf{17.4}}&&6.1&-74.3&&12.9&\textit{\textbf{20.9}}\tabularnewline
-~~JAS&0.3&\textit{\textbf{40.6}}&&6.2&-89&&10.9&\textit{\textbf{79.1}}\tabularnewline
-~~OND&0.5&\textit{\textbf{13.1}}&&2.7&-83.8&&11.7&\textit{\textbf{33.7}}\tabularnewline
-\hline
-\end{tabular}\end{center}
+\begin{kframe}
 
-\end{table}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in rbind(din, chl, sio): object 'din' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in names(totab) <- rep(c("{}Mean"{}, "{}\textbackslash{}\textbackslash{}\% change"{}), 3): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in latex(totab, file = "{}"{}, rowlabel = "{}Period"{}, caption = cap.val, : object 'totab' not found}}\end{kframe}
 
 % trends in din, chl, sio2 at c10
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:c10chg")%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station C10.
-Summaries are means and percent changes based on annual means within the time periods.
-Increasing values are in bold-italics.\label{tab:c10chg}}
-\begin{center}
-\begin{tabular}{lllcllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
-\cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
-\hline
-{\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&1.3&-16.5&&27.5&\textit{\textbf{9}}&&14.4&-21.7\tabularnewline
-\hline
-{\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&1.3&\textit{\textbf{12.1}}&&26.1&-3.3&&15.7&-4.3\tabularnewline
-~~1986-1994&1.4&\textit{\textbf{6.3}}&&27.5&\textit{\textbf{23.5}}&&14.8&-4\tabularnewline
-~~1995-2003&1.4&-0.8&&29.3&-15.3&&14.1&-5\tabularnewline
-~~2004-2013&1.2&-29.4&&27.1&\textit{\textbf{5.1}}&&13.1&-8.7\tabularnewline
-\hline
-{\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~JFM&1.4&-20.5&&12&\textit{\textbf{14.8}}&&14.3&-16.9\tabularnewline
-~~AMJ&1&-26.4&&43&\textit{\textbf{3.1}}&&13.1&-27.5\tabularnewline
-~~JAS&1.3&-3.7&&46.6&-13.6&&14.8&-26.1\tabularnewline
-~~OND&1.6&-6.8&&8.7&-18.5&&15.6&-16\tabularnewline
-\hline
-\end{tabular}\end{center}
+\begin{kframe}
 
-\end{table}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in UseMethod("{}select\_"{}): no applicable method for 'select\_' applied to an object of class "{}NULL"{}}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in rbind(din, chl, sio): object 'din' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in names(totab) <- rep(c("{}Mean"{}, "{}\textbackslash{}\textbackslash{}\% change"{}), 3): object 'totab' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in latex(totab, file = "{}"{}, rowlabel = "{}Period"{}, caption = cap.val, : object 'totab' not found}}\end{kframe}
 

### 2016-10-12 R/dat_proc.R

@@ -1,71 +1,0 @@
-######
-# some additional data processing for the figure
-
-library(WRTDStidal)
-library(tidyr)
-library(dplyr)
-
-######
-# a smoother din model for c10
-data(mods_nolag)
-data(flow_dat)
-data(delt_dat)
-
-# resvar label lookup
-lablk <- list(
-  shrt = c('din', 'nh', 'no23'),
-  lngs = c(
-    expression(paste('ln-dissolved inorganic nitrogen (mg ', L^-1, ')')),
-    expression(paste('ln-ammonium (mg ', L^-1, ')')),
-    expression(paste('ln-nitrite/nitrate (mg ', L^-1, ')'))
-    )
-  )
-
-# flovar label lookup
-stalk <- list(
-  shrt = c('sjr', 'sac', 'sal'),
-  lngs = c('San Joaquin', 'Sacramento', 'Salinity')
-  )
-
-# C10, din
-i <- which(mods_nolag$Site_Code == 'C10' & mods_nolag$resvar == 'din')
-
-# data, respons variable label
-dat <- mods_nolag[i, ]$data[[1]]
-resvar <- mods_nolag[i, ]$resvar
-flovar <- mods_nolag[i, ]$flovar
-sta <- mods_nolag[i, ]$Site_Code
-reslab <- with(lablk, lngs[shrt == resvar])
-flolab <- with(stalk, shrt[lngs == flovar])
-  
-# prep data as tidal object
-tomod <- select(dat, Date, resval, flolag, lim) %>% 
-  rename(
-    res = resval, 
-    flo = flolag
-  ) %>% 
-  data.frame %>% 
-  tidal(., 
-    reslab = reslab, 
-    flolab = expression(paste('ln-flow (standardized)'))
-  )
-    
-topred <- filter(flow_dat, station == flolab) %>% 
-  mutate(flo = log(q)) %>% 
-  rename(date = Date) %>% 
-  select(date, flo) %>% 
-  filter(date >= min(tomod$date) & date <= max(tomod$date)) %>% 
-  na.omit %>% 
-  data.frame
-    
-# create model and exit
-mod <- wrtds(tomod, tau = c(0.1, 0.5, 0.9), wins = list(0.5, 15, 0.5), flo_div = 30, min_obs = 150)
-
-# get predictions, norms from obs flow data
-out <- mod %>% 
-  respred(dat_pred = topred) %>% 
-  resnorm
-
-# save
-dinc10 <- out
-save(dinc10, file = 'data/dinc10.RData')

### 2016-10-13 sftrends_manu.Rnw
figs tabs updated with new results, different seasonal comps

@@
 # set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
+opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T)
 options(replace.assign = TRUE, width = 90)
@@
 
-<<trndmap, fig.height = 6, fig.width = 7, out.width = '0.65\\textwidth', fig.cap = 'Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record.
+<<trndmap, fig.height = 6, fig.width = 9, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record.
 Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
 Points are colored for direction (red increasing, green decreasing) and sized for relative magnitude.
-Station names are shown in the top left panel.'>>=
-trnd_map(molabs = NULL, sz_rng = c(2, 8), leg_lab = '(a) Annual % change')
-trnd_map(yrlabs = NULL, sz_rng = c(2, 8), stat_labs = F, leg_lab = '(b) Seasonal % change')
+Station names are shown in the top left panel.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'>>=
+data(mods)
+mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
+yrbrks <- c(1975, 1995, 2000, 2008, 2014)
+molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
+yrlabs <- c('1976-1995', '1996-2000', '2001-2008', '2009-2014')
+
+trnd_map(mods = mods, molabs = NULL, yrlabs = yrlabs, yrbrks = yrbrks, sz_rng = c(2, 8), leg_lab = '(a) Annual % change')
+trnd_map(mods = mods, yrlabs = NULL, molabs = molabs, mobrks = mobrks, sz_rng = c(2, 8), stat_labs = F, leg_lab = '(b) Seasonal % change')
 @
 
-<<dinc10, fig.height = 8, fig.width = 7, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
+<<dinc10, fig.height = 8, fig.width = 7, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
 Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).
@@
 # data
-data(dinc10)
+data(h1dat)
 
 # din at c10
-toplo <- dinc10
+toplo <- filter(h1dat, resvar == 'din') %>% 
+  .$data %>% 
+  .[[1]]
 
@@
   )
-col_vec <- RColorBrewer::brewer.pal(9, 'Set1')[c(3, 2, 1)]
+col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
 col_lim <- c(0, 3.8)
@@
     axis.text.x = element_blank(),
+    legend.title = element_blank(),
     plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
@@
 toplo2 <- prdnrmplot(toplo, plot = F)
-toplo2 <- left_join(toplo2$fits, toplo2$nrms, by = c('date', 'taus')) %>% 
-  mutate(flodiff = fits_value - nrms_value) %>% 
-  filter(taus == 0.5)
+toplo2 <- left_join(toplo2$fits, toplo2$nrms, by = c('date')) %>% 
+  mutate(flodiff = fits_variable - nrms_variable)
 toplo3 <- toplo
@@
 
-<<dinc10dyna, fig.height = 6, fig.width = 9, out.width = '\\textwidth', fig.cap = 'Modelled relationships between \\ac{din}, flow, and time at station C10.
+<<dinc10dyna, fig.height = 6, fig.width = 9, cache = T, out.width = '\\textwidth', fig.cap = 'Modelled relationships between \\ac{din}, flow, and time at station C10.
 The top figure shows the annual and seasonal variation over the entire time series and the bottom figure shows annual variation for selected months to remove seasonal variation.
@@
   )
-col_lim <- c(0, 2.9)
+col_lim <- c(0, 3.8)
 txt <- data.frame(
@@
 ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
-p1 <- gridplot(toplo, month = 'all', logspace = F, tau = 0.5, col_lim = col_lim, floscl = F, col_vec = rev(col_vec)) +
+p1 <- gridplot(toplo, month = 'all', logspace = F, col_lim = col_lim, floscl = F, col_vec = rev(col_vec)) +
   scale_y_continuous(ylab, expand = c(0, 0)) + 
@@
 # dynaplot selected months
-p2 <- gridplot(toplo, logspace = F, tau = 0.5, col_lim = col_lim, month = c(1, 4, 7, 10), ncol = 4, floscl = F) + 
+p2 <- gridplot(toplo, logspace = F, col_lim = col_lim, month = c(1, 4, 7, 10), ncol = 4, floscl = F) + 
   ptheme + 
@@
 
-<<tracy, fig.height = 5, fig.width = 8, out.width = '\\textwidth', fig.cap = 'Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
+<<tracy, fig.height = 5, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
 Wastewater discharge requirements were implemented in May, 2007 to include nitrification/denitrification and tertiary filtration causing a reduction in total nitrogen effluent discharged to the Delta.
@@
 
-<<p8trnds, fig.height = 8, fig.width = 9, out.width = '\\textwidth', fig.cap = 'Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \\ac{wrtds} (a, bottom), and relationships with flow over time (b).
+<<p8trnds, fig.height = 8, fig.width = 9, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \\ac{wrtds} (a, bottom), and relationships with flow over time (b).
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
@@
 nomod <-  filter(h2dat, resvar == 'no23') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]]
@@
 nhmod <- filter(h2dat, resvar == 'nh') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]]
@@
 
-<<clmchl, fig.height = 7, fig.width = 8, out.width = '\\textwidth', fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
+<<clmchl, fig.height = 7, fig.width = 8, out.width = '\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
 Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means).
@@
 data(flow_dat)
-chld7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'chl')$mod[[1]]
+chld7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'chl')$data[[1]]
 
@@
 chldat <- filter(h3dat, resvar %in% 'chl' & Site_Code %in% 'D7') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]] %>% 
@@
 Covariation between nutrients, chl-{\\it\\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.
-Seasonal changes at D7 are most pronounced during the summer.', fig.width = 9, fig.height = 10>>=
+Seasonal changes at D7 are most pronounced during the summer.', fig.width = 9, fig.height = 10, cache = T>>=
 data(h3dat)
@@
 ## first set of plots
-pdin1 <- prdnrmplot(toplo1$mod[[2]], col_vec = col_vec, logspace = F, min_mo = 10) +
+d7din<- filter(toplo1, resvar == 'din') %>% 
+  .$data %>% 
+  .[[1]]
+pdin1 <- prdnrmplot(d7din, col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-pdinsy1 <- seasyrplot(toplo1$mod[[2]], col_vec = cols, logspace = F, predicted = F) +
+pdinsy1 <- seasyrplot(d7din, col_vec = cols, logspace = F, predicted = F) +
   ptheme + 
@@
 
-pchl1 <- prdnrmplot(toplo1$mod[[1]], col_vec = col_vec, logspace = F, min_mo = 10) +
+d7chl<- filter(toplo1, resvar == 'chl') %>% 
+  .$data %>% 
+  .[[1]]
+pchl1 <- prdnrmplot(d7chl, col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-pchlsy1 <- seasyrplot(toplo1$mod[[1]], col_vec = cols, logspace = F, predicted = F) +
+pchlsy1 <- seasyrplot(d7chl, col_vec = cols, logspace = F, predicted = F) +
   ptheme + 
@@
 
-psio1 <- prdnrmplot(toplo1$mod[[3]], col_vec = col_vec, logspace = F, min_mo = 10) +
+d7sio2 <- filter(toplo1, resvar == 'sio2') %>% 
+  .$data %>% 
+  .[[1]]
+psio1 <- prdnrmplot(d7sio2, col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-psiosy1 <- seasyrplot(toplo1$mod[[3]], col_vec = cols, logspace = F, predicted = F) + 
+psiosy1 <- seasyrplot(d7sio2, col_vec = cols, logspace = F, predicted = F) + 
   ptheme + 
@@
 ## second set of plots
-pdin2 <- prdnrmplot(toplo2$mod[[2]], col_vec = col_vec, logspace = F, min_mo = 10) +
+c10din <- filter(toplo2, resvar == 'din') %>% 
+  .$data %>% 
+  .[[1]]
+pdin2 <- prdnrmplot(c10din, col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-pdinsy2 <- seasyrplot(toplo2$mod[[2]], col_vec = cols, logspace = F, predicted = F) +
+pdinsy2 <- seasyrplot(c10din, col_vec = cols, logspace = F, predicted = F) +
   ptheme + 
@@
 
-pchl2 <- prdnrmplot(toplo2$mod[[1]], col_vec = col_vec, logspace = F, min_mo = 10) +
+c10chl <- filter(toplo2, resvar == 'chl') %>% 
+  .$data %>% 
+  .[[1]]
+pchl2 <- prdnrmplot(c10chl, col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-pchlsy2 <- seasyrplot(toplo2$mod[[1]], col_vec = cols, logspace = F, predicted = F) +
+pchlsy2 <- seasyrplot(c10chl, col_vec = cols, logspace = F, predicted = F) +
   ptheme + 
@@
 
-psio2 <- prdnrmplot(toplo2$mod[[3]], col_vec = col_vec, logspace = F, min_mo = 10) +
+c10sio2 <- filter(toplo2, resvar == 'sio2') %>% 
+  .$data %>% 
+  .[[1]]
+psio2 <- prdnrmplot(c10sio2, col_vec = col_vec, logspace = F, min_mo = 10) +
   ptheme + 
@@
 
-psiosy2 <- seasyrplot(toplo2$mod[[3]], col_vec = cols, logspace = F, predicted = F) + 
+psiosy2 <- seasyrplot(c10sio2, col_vec = cols, logspace = F, predicted = F) + 
   ptheme + 
@@
   spread(var, val) %>% 
-  arrange(resvar, Site_Code)
+  arrange(resvar, Site_Code) %>% 
+  select(Site_Code, resvar, `1976-1995`, `1996-2014`, Spring, Summer, Fall, Winter)
 
@@
 Summaries are averages (mg L$^{-1}$) and percent changes in parentheses (increasing in bold-italic).
-Changes are based on the difference between the ending and starting estimates for the estimates within each period.
-See \\cref{fig:trndmap} for a summary of spatial trends.'
+Changes are based on the difference between the ending and starting estimates within each period.
+See \\cref{fig:trndmap} for a summary of spatial trends.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
   cgroup = c('Annual', 'Seasonal'),
-  n.cgroup = c(3, 3),
+  n.cgroup = c(2, 4),
   rowname = stats,
@@
 noraw_bf <- filter(h2dat, resvar %in% 'no23') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]] %>% 
@@
 noraw_af <- filter(h2dat, resvar %in% 'no23') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]] %>% 
@@
 nhraw_bf <- filter(h2dat, resvar %in% 'nh') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]] %>% 
@@
 nhraw_af <- filter(h2dat, resvar %in% 'nh') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]] %>% 
@@
 
-mobrks <- c(0, 3, 6, 9, 12)
+mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
 yrbrks_bf <- c(1975, 2007)
 yrbrks_af <- c(2006, 2013)
-molabs <- c('JFM', 'AMJ', 'JAS', 'OND')
+molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
 yrlabs_bf <- '1976-2007'
@@
 Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).
-Increasing values are in bold-italics.'
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
 
@@
 dinraw <- filter(d7dat, resvar %in% 'din') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]]
 chlraw <- filter(d7dat, resvar %in% 'chl') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]]
 sioraw <- filter(d7dat, resvar %in% 'sio2') %>% 
-  .$mod %>% 
-  .[[1]]
-
-mobrks <- c(0, 3, 6, 9, 12)
+  .$data %>% 
+  .[[1]]
+
+mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
 yrbrks <- c(1975, 1985, 1994, 2003, 2013)
-molabs <- c('JFM', 'AMJ', 'JAS', 'OND')
+molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
 yrlabs <- c('1976-1985', '1986-1994', '1995-2003', '2004-2013')
@@
 Summaries are means and percent changes based on annual means within the time periods.
-Increasing values are in bold-italics.'
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
 
@@
 dinraw <- filter(c10dat, resvar %in% 'din') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]]
 chlraw <- filter(c10dat, resvar %in% 'chl') %>% 
-  .$mod %>% 
+  .$data %>% 
   .[[1]]
 sioraw <- filter(c10dat, resvar %in% 'sio2') %>% 
-  .$mod %>% 
-  .[[1]]
-
-mobrks <- c(0, 3, 6, 9, 12)
+  .$data %>% 
+  .[[1]]
+
+mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
 yrbrks <- c(1975, 1985, 1994, 2003, 2013)
-molabs <- c('JFM', 'AMJ', 'JAS', 'OND')
+molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
 yrlabs <- c('1976-1985', '1986-1994', '1995-2003', '2004-2013')

### 2016-10-13 sftrends_manu.tex
figs tabs updated with new results, different seasonal comps

--- 2016-09-30 sftrends_manu.tex
@@
   \vspace{1in} 
-  \\ Version Date:   Wed Jul 20 11:11:39 2016 -0500
+  \\ Version Date:   Wed Oct 12 11:34:29 2016 -0500
 	}
@@
 
-{\centering \includegraphics[width=0.65\textwidth]{figs/trends-1} 
-\includegraphics[width=0.65\textwidth]{figs/trends-2}
+{\centering \includegraphics[width=0.75\textwidth]{figs/trndmap-1} 
+\includegraphics[width=0.75\textwidth]{figs/trndmap-2}
 
@@
 
-\caption[Percent changes in nitrogen analytes for (a) annual and (b) seasonal periods in the record]{Percent changes in nitrogen analytes for (a) annual and (b) seasonal periods in the record.
+\caption[Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record]{Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record.
 Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
 Points are colored for direction (red increasing, green decreasing) and sized for relative magnitude.
-Station names are shown in the top left panel.}\label{fig:trends}
+Station names are shown in the top left panel.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}
 \end{figure}
@@
 
+% overall trends
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual", "Seasonal"), n.cgroup = c(2,         4), rowname = stats, label = "tab:trnds")%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
+Summaries are averages (mg L$^{-1}$) and percent changes in parentheses (increasing in bold-italic).
+Changes are based on the difference between the ending and starting estimates within each period.
+See \cref{fig:trndmap} for a summary of spatial trends.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trnds}}
+\begin{center}
+\begin{tabular}{lllcllll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{2}{c}{\bfseries Annual}&\multicolumn{1}{c}{\bfseries }&\multicolumn{4}{c}{\bfseries Seasonal}\tabularnewline
+\cline{2-3} \cline{5-8}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1995}&\multicolumn{1}{c}{1996-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
+\hline
+{\bfseries DIN}&&&&&&&\tabularnewline
+~~C10&1.3 \footnotesize{(\textit{\textbf{35}})}&1.4 \footnotesize{(-28.6)}&&1.1 \footnotesize{(-25.8)}&1.2 \footnotesize{(-6.3)}&1.5 \footnotesize{(\textit{\textbf{5.5}})}&1.7 \footnotesize{(-20.7)}\tabularnewline
+~~C3&0.4 \footnotesize{(\textit{\textbf{51.1}})}&0.5 \footnotesize{(\textit{\textbf{3}})}&&0.4 \footnotesize{(\textit{\textbf{62.2}})}&0.3 \footnotesize{(\textit{\textbf{71.9}})}&0.5 \footnotesize{(\textit{\textbf{54.4}})}&0.5 \footnotesize{(\textit{\textbf{18.3}})}\tabularnewline
+~~D19&0.5 \footnotesize{(-5.5)}&0.4 \footnotesize{(-30.7)}&&0.5 \footnotesize{(-33.9)}&0.3 \footnotesize{(-25.4)}&0.3 \footnotesize{(-21.7)}&0.7 \footnotesize{(-33.5)}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.3}})}&0.5 \footnotesize{(-19.6)}&&0.5 \footnotesize{(-19.2)}&0.3 \footnotesize{(-10.2)}&0.4 \footnotesize{(-3.5)}&0.6 \footnotesize{(-21.7)}\tabularnewline
+~~D28&0.5 \footnotesize{(-3.1)}&0.4 \footnotesize{(-45.6)}&&0.5 \footnotesize{(-32.9)}&0.2 \footnotesize{(-56.5)}&0.3 \footnotesize{(-57.4)}&0.8 \footnotesize{(-40.2)}\tabularnewline
+~~D4&0.4 \footnotesize{(-8.1)}&0.4 \footnotesize{(-7.5)}&&0.4 \footnotesize{(-13.6)}&0.3 \footnotesize{(\textit{\textbf{4.3}})}&0.4 \footnotesize{(-1.2)}&0.5 \footnotesize{(-15)}\tabularnewline
+~~D6&0.4 \footnotesize{(\textit{\textbf{11.6}})}&0.5 \footnotesize{(-6.1)}&&0.5 \footnotesize{(-5)}&0.4 \footnotesize{(\textit{\textbf{64.1}})}&0.4 \footnotesize{(\textit{\textbf{21}})}&0.5 \footnotesize{(-14.6)}\tabularnewline
+~~D7&0.4 \footnotesize{(-1.7)}&0.5 \footnotesize{(-11.8)}&&0.5 \footnotesize{(-23.8)}&0.4 \footnotesize{(\textit{\textbf{49.9}})}&0.4 \footnotesize{(\textit{\textbf{10.4}})}&0.6 \footnotesize{(-34.6)}\tabularnewline
+~~P8&1.4 \footnotesize{(\textit{\textbf{81.1}})}&1.8 \footnotesize{(-15.8)}&&1.5 \footnotesize{(\textit{\textbf{35.5}})}&1 \footnotesize{(\textit{\textbf{42}})}&1.6 \footnotesize{(\textit{\textbf{55}})}&2.3 \footnotesize{(\textit{\textbf{38.7}})}\tabularnewline
+\hline
+{\bfseries NH$_{4}^{+}$}&&&&&&&\tabularnewline
+~~C10&0.1 \footnotesize{(-37.7)}&0 \footnotesize{(-55.5)}&&0.1 \footnotesize{(-68.6)}&0 \footnotesize{(-85)}&0.1 \footnotesize{(-90.8)}&0.2 \footnotesize{(-78.8)}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{127}})}&0.3 \footnotesize{(\textit{\textbf{14.1}})}&&0.2 \footnotesize{(\textit{\textbf{165.1}})}&0.2 \footnotesize{(\textit{\textbf{133.3}})}&0.3 \footnotesize{(\textit{\textbf{107.7}})}&0.2 \footnotesize{(\textit{\textbf{81.2}})}\tabularnewline
+~~D19&0.1 \footnotesize{(\textit{\textbf{35.6}})}&0.1 \footnotesize{(-32.6)}&&0.1 \footnotesize{(-32.9)}&0 \footnotesize{(-21)}&0 \footnotesize{(-6.4)}&0.1 \footnotesize{(-12)}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{58.4}})}&0.1 \footnotesize{(-18.7)}&&0.1 \footnotesize{(\textit{\textbf{3.3}})}&0.1 \footnotesize{(\textit{\textbf{38.1}})}&0.1 \footnotesize{(\textit{\textbf{21.1}})}&0.1 \footnotesize{(\textit{\textbf{0.8}})}\tabularnewline
+~~D28&0.1 \footnotesize{(-10)}&0 \footnotesize{(-31.8)}&&0.1 \footnotesize{(-42.8)}&0 \footnotesize{(-36.9)}&0 \footnotesize{(-42.2)}&0.1 \footnotesize{(-35.7)}\tabularnewline
+~~D4&0.1 \footnotesize{(\textit{\textbf{74.6}})}&0.1 \footnotesize{(\textit{\textbf{0.3}})}&&0.1 \footnotesize{(\textit{\textbf{54.2}})}&0 \footnotesize{(\textit{\textbf{30.8}})}&0.1 \footnotesize{(\textit{\textbf{23.8}})}&0.1 \footnotesize{(\textit{\textbf{56.9}})}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{100.2}})}&0.1 \footnotesize{(-3.5)}&&0.1 \footnotesize{(\textit{\textbf{42.2}})}&0.1 \footnotesize{(\textit{\textbf{74.1}})}&0.1 \footnotesize{(\textit{\textbf{49.2}})}&0.1 \footnotesize{(\textit{\textbf{39.8}})}\tabularnewline
+~~D7&0.1 \footnotesize{(\textit{\textbf{88.2}})}&0.1 \footnotesize{(-13.3)}&&0.1 \footnotesize{(\textit{\textbf{26.8}})}&0 \footnotesize{(\textit{\textbf{25.1}})}&0.1 \footnotesize{(\textit{\textbf{4.3}})}&0.1 \footnotesize{(\textit{\textbf{16.8}})}\tabularnewline
+~~P8&0.3 \footnotesize{(\textit{\textbf{267.4}})}&0.3 \footnotesize{(-85.1)}&&0.2 \footnotesize{(-51.8)}&0.1 \footnotesize{(-12)}&0.3 \footnotesize{(-44.5)}&0.6 \footnotesize{(-61.2)}\tabularnewline
+\hline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&&&&\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{42.8}})}&1.3 \footnotesize{(-26.9)}&&1.1 \footnotesize{(-21)}&1.2 \footnotesize{(-2.3)}&1.4 \footnotesize{(\textit{\textbf{23.3}})}&1.5 \footnotesize{(-5.6)}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{0.7}})}&0.2 \footnotesize{(-6.6)}&&0.2 \footnotesize{(\textit{\textbf{0.9}})}&0.1 \footnotesize{(\textit{\textbf{18}})}&0.1 \footnotesize{(\textit{\textbf{4.3}})}&0.2 \footnotesize{(-17.4)}\tabularnewline
+~~D19&0.4 \footnotesize{(-11.7)}&0.4 \footnotesize{(-31.1)}&&0.4 \footnotesize{(-32.3)}&0.2 \footnotesize{(-36.7)}&0.3 \footnotesize{(-31.6)}&0.6 \footnotesize{(-38.4)}\tabularnewline
+~~D26&0.3 \footnotesize{(-9.8)}&0.4 \footnotesize{(-20.1)}&&0.4 \footnotesize{(-23.5)}&0.2 \footnotesize{(-18.7)}&0.3 \footnotesize{(-10.5)}&0.5 \footnotesize{(-26.4)}\tabularnewline
+~~D28&0.4 \footnotesize{(-10.5)}&0.4 \footnotesize{(-47.2)}&&0.5 \footnotesize{(-34.6)}&0.2 \footnotesize{(-66.3)}&0.3 \footnotesize{(-64.4)}&0.7 \footnotesize{(-45.7)}\tabularnewline
+~~D4&0.3 \footnotesize{(-14.6)}&0.3 \footnotesize{(-9.8)}&&0.3 \footnotesize{(-19.2)}&0.3 \footnotesize{(\textit{\textbf{1.1}})}&0.3 \footnotesize{(-8)}&0.4 \footnotesize{(-25.9)}\tabularnewline
+~~D6&0.3 \footnotesize{(-5.6)}&0.4 \footnotesize{(-8)}&&0.4 \footnotesize{(-14.5)}&0.3 \footnotesize{(\textit{\textbf{38.8}})}&0.3 \footnotesize{(\textit{\textbf{4}})}&0.4 \footnotesize{(-23.2)}\tabularnewline
+~~D7&0.4 \footnotesize{(-13.4)}&0.4 \footnotesize{(-13.6)}&&0.4 \footnotesize{(-27.3)}&0.3 \footnotesize{(\textit{\textbf{28.7}})}&0.4 \footnotesize{(-23.7)}&0.5 \footnotesize{(-46.8)}\tabularnewline
+~~P8&1.1 \footnotesize{(\textit{\textbf{59.8}})}&1.5 \footnotesize{(\textit{\textbf{3.3}})}&&1.2 \footnotesize{(\textit{\textbf{53.9}})}&1 \footnotesize{(\textit{\textbf{38.8}})}&1.4 \footnotesize{(\textit{\textbf{58.3}})}&1.8 \footnotesize{(\textit{\textbf{62.3}})}\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
+
 % trends in no23, nh at p8
@@
 Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).
-Increasing values are in bold-italics.\label{tab:p8chg}}
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:p8chg}}
 \begin{center}
@@
 {\bfseries Annual}&&&&&\tabularnewline
-~~1976-2007&1.27&\textit{\textbf{75.7}}&&0.33&\textit{\textbf{71.3}}\tabularnewline
-~~2008-2013&1.47&-6&&0.11&-62.6\tabularnewline
+~~1976-2007&1.3&\textit{\textbf{74.3}}&&0.34&\textit{\textbf{93.1}}\tabularnewline
+~~2008-2013&1.5&-7.3&&0.11&-62\tabularnewline
 \hline
 {\bfseries Seasonal, pre}&&&&&\tabularnewline
-~~JFM&1.46&\textit{\textbf{77.7}}&&0.49&\textit{\textbf{57.1}}\tabularnewline
-~~AMJ&1&\textit{\textbf{75.3}}&&0.13&\textit{\textbf{63.4}}\tabularnewline
-~~JAS&1.08&\textit{\textbf{97.9}}&&0.13&\textit{\textbf{151.9}}\tabularnewline
-~~OND&1.53&\textit{\textbf{81.9}}&&0.54&\textit{\textbf{102.4}}\tabularnewline
+~~Spring&1.24&\textit{\textbf{79.5}}&&0.25&\textit{\textbf{57.1}}\tabularnewline
+~~Summer&0.99&\textit{\textbf{95}}&&0.09&\textit{\textbf{143.5}}\tabularnewline
+~~Fall&1.32&\textit{\textbf{77}}&&0.29&\textit{\textbf{137.7}}\tabularnewline
+~~Winter&1.67&\textit{\textbf{69.2}}&&0.69&\textit{\textbf{72.5}}\tabularnewline
 \hline
 {\bfseries Seasonal, post}&&&&&\tabularnewline
-~~JFM&1.76&-3.4&&0.14&-47\tabularnewline
-~~AMJ&0.94&-5.5&&0.07&-45.3\tabularnewline
-~~JAS&1.17&-10&&0.07&-50.3\tabularnewline
-~~OND&1.99&-4.6&&0.15&-58.5\tabularnewline
+~~Spring&1.27&\textit{\textbf{0.9}}&&0.09&-49.2\tabularnewline
+~~Summer&0.92&-18.7&&0.06&-49.2\tabularnewline
+~~Fall&1.58&-8.7&&0.1&-58.2\tabularnewline
+~~Winter&2.17&-5.1&&0.18&-62.1\tabularnewline
 \hline
@@
 Summaries are means and percent changes based on annual means within the time periods.
-Increasing values are in bold-italics.\label{tab:d7chg}}
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:d7chg}}
 \begin{center}
@@
 {\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&0.4&\textit{\textbf{27.8}}&&4.4&-81.6&&12.6&\textit{\textbf{30.2}}\tabularnewline
+~~1976-2013&0.4&-7.8&&4.1&-76.7&&12.7&\textit{\textbf{21.8}}\tabularnewline
 \hline
 {\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&0.4&\textit{\textbf{10.2}}&&9.8&-63.9&&10.8&\textit{\textbf{9}}\tabularnewline
-~~1986-1994&0.4&\textit{\textbf{14.9}}&&3.4&-48.8&&12.2&\textit{\textbf{14.2}}\tabularnewline
-~~1995-2003&0.5&\textit{\textbf{4}}&&2.1&-4.8&&13.5&\textit{\textbf{3.7}}\tabularnewline
-~~2004-2013&0.5&-8.6&&2.5&\textit{\textbf{28.5}}&&13.7&-1.8\tabularnewline
+~~1976-1985&0.4&-22.5&&9.3&-57.1&&10.9&-0.1\tabularnewline
+~~1986-1994&0.4&\textit{\textbf{29.4}}&&3&-61.7&&12.2&\textit{\textbf{18.6}}\tabularnewline
+~~1995-2003&0.5&-0.1&&2&\textit{\textbf{21.1}}&&13.7&\textit{\textbf{5.2}}\tabularnewline
+~~2004-2013&0.5&-11.9&&2.4&\textit{\textbf{34.2}}&&13.7&-1.7\tabularnewline
 \hline
 {\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~JFM&0.5&-4.9&&2.7&-56&&15&\textit{\textbf{4.4}}\tabularnewline
-~~AMJ&0.4&\textit{\textbf{17.4}}&&6.1&-74.3&&12.9&\textit{\textbf{20.9}}\tabularnewline
-~~JAS&0.3&\textit{\textbf{40.6}}&&6.2&-89&&10.9&\textit{\textbf{79.1}}\tabularnewline
-~~OND&0.5&\textit{\textbf{13.1}}&&2.7&-83.8&&11.7&\textit{\textbf{33.7}}\tabularnewline
+~~Spring&0.5&-23.8&&4.2&-60.3&&14.5&\textit{\textbf{0.6}}\tabularnewline
+~~Summer&0.4&\textit{\textbf{49.9}}&&6.4&-82.3&&11.2&\textit{\textbf{51.3}}\tabularnewline
+~~Fall&0.4&\textit{\textbf{10.4}}&&3.7&-84.9&&11.1&\textit{\textbf{43.3}}\tabularnewline
+~~Winter&0.6&-34.6&&2&-63.8&&14.2&\textit{\textbf{4.2}}\tabularnewline
 \hline
@@
 Summaries are means and percent changes based on annual means within the time periods.
-Increasing values are in bold-italics.\label{tab:c10chg}}
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:c10chg}}
 \begin{center}
@@
 {\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&1.3&-16.5&&27.5&\textit{\textbf{9}}&&14.4&-21.7\tabularnewline
+~~1976-2013&1.4&-10.1&&33.2&-20.6&&14.4&-19.1\tabularnewline
 \hline
 {\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&1.3&\textit{\textbf{12.1}}&&26.1&-3.3&&15.7&-4.3\tabularnewline
-~~1986-1994&1.4&\textit{\textbf{6.3}}&&27.5&\textit{\textbf{23.5}}&&14.8&-4\tabularnewline
-~~1995-2003&1.4&-0.8&&29.3&-15.3&&14.1&-5\tabularnewline
-~~2004-2013&1.2&-29.4&&27.1&\textit{\textbf{5.1}}&&13.1&-8.7\tabularnewline
+~~1976-1985&1.3&\textit{\textbf{22.8}}&&27.5&-21.6&&15.8&-3.2\tabularnewline
+~~1986-1994&1.4&\textit{\textbf{1.2}}&&35.3&\textit{\textbf{31.5}}&&14.8&-4.1\tabularnewline
+~~1995-2003&1.6&\textit{\textbf{16.4}}&&41&-27.5&&14.2&-3.3\tabularnewline
+~~2004-2013&1.3&-36.3&&30.2&-12.8&&13.2&-7.6\tabularnewline
 \hline
 {\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~JFM&1.4&-20.5&&12&\textit{\textbf{14.8}}&&14.3&-16.9\tabularnewline
-~~AMJ&1&-26.4&&43&\textit{\textbf{3.1}}&&13.1&-27.5\tabularnewline
-~~JAS&1.3&-3.7&&46.6&-13.6&&14.8&-26.1\tabularnewline
-~~OND&1.6&-6.8&&8.7&-18.5&&15.6&-16\tabularnewline
+~~Spring&1.1&-25.8&&26.2&\textit{\textbf{3.1}}&&13.2&-20\tabularnewline
+~~Summer&1.2&-6.3&&79.6&-23.2&&13.9&-25.4\tabularnewline
+~~Fall&1.5&\textit{\textbf{5.5}}&&19&-39.8&&15.6&-18\tabularnewline
+~~Winter&1.7&-20.7&&7.7&\textit{\textbf{7.3}}&&15.1&-14.1\tabularnewline
 \hline

### 2016-10-13 R/funcs.R

@@ -23,8 +23,8 @@ g_legend <- function(a.gplot){
 #
 # uses delt_dat, mods_nolag, delt_map
 #
-trnd_map <- function(res = c('din', 'nh', 'no23'), 
-  mobrks = c(-Inf, 4, 8, Inf),
+trnd_map <- function(res = c('din', 'nh', 'no23'), mods,
+  mobrks = list(c(1, 2, 3, 4), c(5, 6, 7, 8), c(9, 10, 11, 12)),
   yrbrks = c(-Inf, 1988, 2000, Inf),
   molabs = c('JFMA', 'MJJA', 'SOND'),
   yrlabs = c('1976-1988', '1989-2000', '2001-2012'),
@@ -43,7 +43,6 @@ trnd_map <- function(res = c('din', 'nh', 'no23'),
   # load required data
   data(delt_dat)
   data(delt_map)
-  data(mods_nolag)
 
   # response labels 
   reslabs <- list(
@@ -56,7 +55,7 @@ trnd_map <- function(res = c('din', 'nh', 'no23'),
   )
   
   # make response labels
-  mods_nolag <- mutate(mods_nolag, reslabs = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr))
+  mods <- mutate(mods, reslabs = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr))
   
   ##
   # get a bounding box for the stations
@@ -76,17 +75,40 @@ trnd_map <- function(res = c('din', 'nh', 'no23'),
   
   ##
   # get trends using wrtdstrnd function
+
+  # need placeholders for null molabs
+  if(is.null(molabs)){
+    
+    molabstmp <- letters[1:length(mobrks)]
+    
+    # get trends
+    trnds <- mutate(mods, 
+      trnd = map(data, function(x){
+        wrtdstrnd(x, mobrks, yrbrks, molabstmp, yrlabs)
+        })
+      )
   
-  # get trends, merge with statmeta for lat/lon
-  trnds <- mutate(mods_nolag, 
-    trnd = map(mod, function(x){
-      wrtdstrnd(x, mobrks, yrbrks, molabs, yrlabs, tau = 0.5)
-      })
-    ) %>% 
-    select(-data, -mod) %>% 
-    unnest %>% 
-    left_join(., statmeta, by = 'Site_Code') %>% 
-    data.frame
+  }
+  
+  # need placeholders for null yrlabs
+  if(is.null(yrlabs)){
[…]

### 2016-10-18 sftrends_manu.Rnw
started study location text

@@
 This region is a complex mosaic of inflows that are primary sources of nutrients into the larger Bay.
-To date, a comprehensive evaluation of the long-term monitoring dataset at multiple stations in the Delta has not been conducted despite the importance of nutrient transport from the region for water quality in the entire bay.
+To date, a comprehensive evaluation of flow-normalized trends using the long-term monitoring dataset at multiple stations in the Delta has not been conducted despite the importance of nutrient transport from the region for water quality in the entire bay.
 The \ac{wrtds} technique is data-driven where the parameterization of the functional model changes smoothly over time following dynamic patterns of season and flow.
@@
 
-The \ac{sfe}...
-exchange with ocean, freshwater inputs (rivers and wshed drainage), runoff variation (drought years), climate, support of natural resources, subembayments, POTW (publicly owned treatment works), nutrient loading relative to other locations
+The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest bay on the Pacific coast of North America.
+The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land in the Central Valley.
+Water enters the Bay through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
+The \ac{sfe} system is divided into several sub-bays, including Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.
+Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \citep{Jassby00}.
+Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \citep{Cloern12b}.
+The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns.
+Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \citep{Cloern15}.
+Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \citep{Cloern07}.
+
+Nutrient loading in \ac{sfe} is comparable to other locations that exhibit symptomatic effects of cultural eutrophication, although productivity has historically not been affected by excess nutrients \citep{Jassby02,Kimmerer12}.
+Orthoposphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
+Annual nutrient export from the Delta region as been estimated as approximateily 30 thousand kg d$-1$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
+Although nitrogen inputs are considerable, several
+
+POTW (publicly owned treatment works), nutrient loading relative to other locations, recent trends
+
+The delta region...
+
 \subsection{Analysis method and application}
+
+nine discrete sampling stations with data from 1976 to 2012, \ac{din}, ammonium, total nitrate, effects of different flow variables
+
+\subsection{Time series modelling}
 
@@
 \end{equation}
-
-nine discrete sampling stations with data from 1976 to 2012, \ac{din}, ammonium, total nitrate, effects of different flow variables
 

### 2016-10-18 sftrends_manu.tex
started study location text

@@
   \vspace{1in} 
-  \\ Version Date:   Wed Oct 12 11:34:29 2016 -0500
+  \\ Version Date:   Thu Oct 13 08:56:33 2016 -0500
 	}

### 2016-10-19 sftrends_manu.Rnw
more site desription

@@
 
-Nutrient loading in \ac{sfe} is comparable to other locations that exhibit symptomatic effects of cultural eutrophication, although productivity has historically not been affected by excess nutrients \citep{Jassby02,Kimmerer12}.
-Orthoposphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
-Annual nutrient export from the Delta region as been estimated as approximateily 30 thousand kg d$-1$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
-Although nitrogen inputs are considerable, several
-
-POTW (publicly owned treatment works), nutrient loading relative to other locations, recent trends
-
-The delta region...
+Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.
+Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
+Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
+Although nitrogen inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
+Recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass, increased occurrence of hypoxic conditions, and increased abundance of phytoplankton species associated with harmful algal blooms.
+These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.
+
+The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality in response to nutrient inputs into the Bay.
+The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers.
+Quantitative descriptions of nutrient dynamics in the Delta are complicated by the numerous sources that contribute input loads.
+Sources are not limited to the watershed, such that nutrient concentrations can also originate from local wastewater inputs and internal loading from sediment processes.
+Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have further complicated the interpretation of trends in nutrient concentrations.
+A comprehensive evaluation of nutrient dynamics using mass-balance models showed that
+
+Accordingly, quantitative descriptions of nutrient dynamics in the Delta must explicitly account for the effects of flow changes to better understand variation both within the Delta and potential mechanisms of downstream tranport.
+
+Two data sources that provide multi-decadal time series of nutrients and daily flow records were used to develop a quantitative description fo
 

### 2016-10-19 sftrends_manu.tex
more site desription

@@
   \vspace{1in} 
-  \\ Version Date:   Thu Oct 13 08:56:33 2016 -0500
+  \\ Version Date:   Tue Oct 18 17:52:14 2016 -0500
 	}
@@
 
-Nutrient loading in \ac{sfe} is comparable to other locations that exhibit symptomatic effects of cultural eutrophication, although productivity has historically not been affected by excess nutrients \citep{Jassby02,Kimmerer12}.
-Orthoposphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
-Annual nutrient export from the Delta region as been estimated as approximateily 30 thousand kg d$-1$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
-Although nitrogen inputs are considerable, several
+Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.
+Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
+Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
+Although nitrogen inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
+Recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass, increased occurrence of hypoxic conditions, and increased abundance of phytoplankton species associated with harmful algal blooms.
+These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.
 
-POTW (publicly owned treatment works), nutrient loading relative to other locations, recent trends
+The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.
+The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers to the lower Bay.
+Quantitative descriptions of nutrient dynamics in the Delta are complicated by the numerous sources that contribute input loads.
+Sources are not limited to the watershed, such that local inputs from wastewater and internal loading from sediment processes also represent important contributions.
+Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have complicated the interpretation of trends in nutrient concentrations.
+Accordingly, quantitative descriptions of nutrient dynamics in the Delta must explicitly account for the effects of flow changes to better understand variation both within the Delta and potential mechanisms of downstream tranport.
 
-The delta region...
+Two data sources that provide multi-decadal time series of nutrients and daily flow records were used to develop a quantitative description fo
 

### 2016-10-20 sftrends_manu.tex
author updates

@@
   {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
-  \\{\bf {\normalsize Thomas Jabusch$^2$}}
+  \\{\bf {\normalsize Thomas Jabusch$^2$, Phil Trowbridge$^2$}}
   \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
@@
   \vspace{1in} 
-  \\ Version Date:   Tue Oct 18 17:52:14 2016 -0500
+  \\ Version Date:   Wed Oct 19 16:59:28 2016 -0500
 	}
@@
 
-The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.
-The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers to the lower Bay.
+The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality in response to nutrient inputs into the Bay.
+The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers.
 Quantitative descriptions of nutrient dynamics in the Delta are complicated by the numerous sources that contribute input loads.
-Sources are not limited to the watershed, such that local inputs from wastewater and internal loading from sediment processes also represent important contributions.
-Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have complicated the interpretation of trends in nutrient concentrations.
+Sources are not limited to the watershed, such that nutrient concentrations can also originate from local wastewater inputs and internal loading from sediment processes.
+Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have further complicated the interpretation of trends in nutrient concentrations.
+A comprehensive evaluation of nutrient dynamics using mass-balance models showed that
+
 Accordingly, quantitative descriptions of nutrient dynamics in the Delta must explicitly account for the effects of flow changes to better understand variation both within the Delta and potential mechanisms of downstream tranport.

### 2016-10-25 sftrends_manu.Rnw
author updates
through study site description

--- 2016-10-18 sftrends_manu.Rnw
@@
 % aliased citations
-% \defcitealias{FLDEP12}{FLDEP 2012}
+\defcitealias{SutulaIR}{Sutula et al. in review}
 
@@
 \acrodef{din}[DIN]{dissolved inorganic nitrogen}
+\acrodef{emp}[EMP]{Environmental Monitoring Program}
+\acrodef{iep}[IEP]{Interagency Ecological Program}
 \acrodef{jas}[JAS]{July-August-September}
@@
 
-\subsection{Study location and data}
+\subsection{Study system}
 
@@
 
-Nutrient loading in \ac{sfe} is comparable to other locations that exhibit symptomatic effects of cultural eutrophication, although productivity has historically not been affected by excess nutrients \citep{Jassby02,Kimmerer12}.
-Orthoposphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
-Annual nutrient export from the Delta region as been estimated as approximateily 30 thousand kg d$-1$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
-Although nitrogen inputs are considerable, several
-
-POTW (publicly owned treatment works), nutrient loading relative to other locations, recent trends
-
-The delta region...
+Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.
+Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
+Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
+Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
+However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.
+These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.
+
+The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.
+The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \citep{Jassby00,Jassby08,Novick15}.
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.
+A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}.
+For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.
+Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
+Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \citep{Jassby00,Jassby08}.
+Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
+
+\subsection{Data sources}
+
+Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.
+The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \citep{IEP13}.
+The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.
+Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.
+All samples were processed with adequate QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.
+Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 to 2012.
+Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+These stations were chosen based on the continuity of the water quality time series and geographic location for understanding trends.
+Data were minimally processed with the exception of averaging replicates that occurred on the same day.
+Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).
+The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
+
+Daily flow estimates for the Delta region were obtained from the
 

### 2016-10-25 sftrends_manu.tex
through study site description

--- 2016-10-19 sftrends_manu.tex
@@
   \vspace{1in} 
-  \\ Version Date:   Tue Oct 18 17:52:14 2016 -0500
+  \\ Version Date:   Thu Oct 20 16:04:44 2016 -0500
 	}
@@
 Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
-Although nitrogen inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
-Recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass, increased occurrence of hypoxic conditions, and increased abundance of phytoplankton species associated with harmful algal blooms.
+Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
+However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.
 These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.
@@
 The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.
-The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers to the lower Bay.
-Quantitative descriptions of nutrient dynamics in the Delta are complicated by the numerous sources that contribute input loads.
-Sources are not limited to the watershed, such that local inputs from wastewater and internal loading from sediment processes also represent important contributions.
-Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have complicated the interpretation of trends in nutrient concentrations.
-Accordingly, quantitative descriptions of nutrient dynamics in the Delta must explicitly account for the effects of flow changes to better understand variation both within the Delta and potential mechanisms of downstream tranport.
+The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \citep{Jassby00,Jassby08,Novick15}.
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.
+A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}.
+For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.
+Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
+Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \citep{Jassby00,Jassby08}.
+Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
 
-Two data sources that provide multi-decadal time series of nutrients and daily flow records were used to develop a quantitative description fo
+\subsection{Data sources}
+
+Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.
+The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \citep{IEP13}.
+The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.
+Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.
+All samples were processed with adequate QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.
+Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 to 2012.
+Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+These stations were chosen based on the continuity of the water quality time series and geographic location for understanding trends.
+Data were minimally processed with the exception of averaging replicates that occurred on the same day.
+Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).
+The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
+
+Daily flow estimates for the Delta region were obtained from the
 

### 2016-10-31 sftrends_manu.Rnw
almost done with analysis methods

@@
 Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.
-All samples were processed with adequate QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.
+All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.
 Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 to 2012.
 Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
-These stations were chosen based on the continuity of the water quality time series and geographic location for understanding trends.
+These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.
 Data were minimally processed with the exception of averaging replicates that occurred on the same day.
@@
 
-Daily flow estimates for the Delta region were obtained from the
+Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \citep{IEP16}.
+Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.
+The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends.
+Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.
+Initial analyses indicated that model fit was not signifcantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
+As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.
+The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.
+Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.
 
@@
 
-nine discrete sampling stations with data from 1976 to 2012, \ac{din}, ammonium, total nitrate, effects of different flow variables
-
-\subsection{Time series modelling}
+A total of twenty-seven \ac{wrtds} models were created, one for each nitrogen at each station.
+The functional form of \ac{wrtds} is a simple regression that models the log-transformed response variable as a function of time, flow, and season:
 
@@
 
-Annual, seasonal trends of flow-normalized predictions
+\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
+The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.
+The \ac{wrtds} model is a moving window regression that fits a unique set of parameters at each observation point in the time series.
+A unique set of weights is used for each regression to control the relevance of observations used to fit the model to the observation at the center of the window.
+The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow.
+The final vector used to fit the model at each point weights observations more similar to the center of the window with more importance.
+The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.
+As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series.
+The original \ac{wrtds} method is described in more detail in \citep{Hirsch10}.
+The \ac{wrtds} model applied to the Delta time series followed methods in \citet{Beck15}, which were based on a tidal adaptation of the original method.
+The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
+Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme evens.
+The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.
+
+A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
+Flow-normalized trends have value for the interpretation of changes that could have been caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}).
+Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect trend evaluation.
+A flow-normalized value is the average of predictions at a given time using all flow values that are expected to occur for the relevant month across years in the record.
+Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
+Specifically, flow-normalized trends in each analyte were summarized as percent changes from start to end of annual groupings from 1976-1995, 1996-2000, 2001-2008, and 2009-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
 

### 2016-10-31 sftrends_manu.tex
almost done with analysis methods

@@
   \vspace{1in} 
-  \\ Version Date:   Thu Oct 20 16:04:44 2016 -0500
+  \\ Version Date:   Tue Oct 25 16:16:08 2016 -0500
 	}

### 2016-11-01 sftrends_manu.Rnw
working on case study descriptions

@@
 % aliased citations
+\defcitealias{BeckIP}{Beck and Murphy in press}
 \defcitealias{SutulaIR}{Sutula et al. in review}
@@
 The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.
+All analyses used the WRTDStidal package for R \citep{Beck16b,RDCT16}
 
 A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
-Flow-normalized trends have value for the interpretation of changes that could have been caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}).
-Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect trend evaluation.
-A flow-normalized value is the average of predictions at a given time using all flow values that are expected to occur for the relevant month across years in the record.
+Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citealt{Beck15}).
+Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
+A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
-Specifically, flow-normalized trends in each analyte were summarized as percent changes from start to end of annual groupings from 1976-1995, 1996-2000, 2001-2008, and 2009-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
+Specifically, flow-normalized trends in each analyte were summarized as both averages and percent changes from the beginning to end of annual groupings from 1976-1995, 1996-2000, 2001-2008, and 2009-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
+These annual and seasonal groupings were chosen for continuity with additional research efforts \citep[i.e.,][]{Jabusch16}.
+Trends within each seasonal grouping were based on annual mean estimates for the first and last three years of flow-normalized estimates to reduce the effects of abnormal years or missing data.
+Annual trends were based only on the first and last years in the time period due to limited number of years in some groupings \citepalias{BeckIP}.
 
 \subsection{Case studies}
-These are science questions that are relevant outside of the region.
+
+Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more detailed understanding of trends that can lead to additional hypotheses for further evaluation.
+The stations were also chosen to address ecological and management-based questions that have relevance outside of the region, having importance for our understanding of estuarine processes that influence eutrophication trends on decadal time scales.
+The selected case studies focused on \begin{inparaenum}[1\upshape)]
+\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
+\item effects of wastewater treatment upgrades upstream of P8, and
+\item effects of biological invasion on nutrient dynamics in Suisun Bay.
+\end{inparaenum}
+Each case study is built around hypotheses that results from \ac{wrtds} models are expected to support, both as a general description and as a means for additional testing with alternative methods.
+
 \subsubsection{Disaggregating observed nitrogen time series}
-Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear and complex, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
+
+Multiple biological and physical factors influence nutrient concentrations at different temporal scales \citep[][, references therein]{Cloern10}.
+As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex.
+A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, is expected to reveal \begin{inparaenum}[1\upshape)]
+\item an annual trend independent of a seasonal trend, and 
+\item varying flow contribution to nutrient dynamics, either as a difference between predicted and flow-normalized results or changes in the nutrient versus flow relationship at different annual periods.
+\end{inparaenum}
+Results from this case study were used to demonstrate the breadth of information that can be obtained from the observed time series with \ac{wrtds}.
+
 \subsubsection{Effects of wastewater treatment}
@@
 data(mods)
+
 mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))

### 2016-11-01 sftrends_manu.tex
working on case study descriptions

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Oct 25 16:16:08 2016 -0500
+  \\ Version Date:   Mon Oct 31 12:28:10 2016 -0500
 	}
@@
 The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.
+All analyses used the WRTDStidal package for R \citep{Beck16b,RDCT16}
 
 A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
-Flow-normalized trends have value for the interpretation of changes that could have been caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}).
-Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect trend evaluation.
-A flow-normalized value is the average of predictions at a given time using all flow values that are expected to occur for the relevant month across years in the record.
+Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citealt{Beck15}).
+Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
+A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
-Specifically, flow-normalized trends in each analyte were summarized as percent changes from start to end of annual groupings from 1976-1995, 1996-2000, 2001-2008, and 2009-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
+Specifically, flow-normalized trends in each analyte were summarized as both averages and percent changes from the beginning to end of annual groupings from 1976-1995, 1996-2000, 2001-2008, and 2009-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
+These annual and seasonal groupings were chosen for continuity with additional research efforts \citep[i.e.,][]{Jabusch16}.
+Trends within each seasonal grouping were based on annual mean estimates for the first and last three years of flow-normalized estimates to reduce the effects of abnormal years or missing data.
+Annual trends were based only on the first and last years in the time period due to limited number of years in some groupings \citepalias{BeckIP}.
 
 \subsection{Case studies}
-These are science questions that are relevant outside of the region.
+
+Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more detailed understanding of trends that can lead to additional hypotheses for further evaluation.
+The stations were also chosen to address ecological and management-based questions that have relevance outside of the region, having importance for our understanding of estuarine processes that influence eutrophication trends on decadal time scales.
+The selected case studies focused on \begin{inparaenum}[1\upshape)]
+\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
+\item effects of wastewater treatment upgrades upstream of P8, and
+\item effects of biological invasion on nutrient dynamics in Suisun Bay.
+\end{inparaenum}
+Each case study is built around hypotheses that results from \ac{wrtds} models are expected to support, both as a general description and as a means for additional testing with alternative methods.
+
 \subsubsection{Disaggregating observed nitrogen time series}
-Hypothesis: Because multiple factors influence nutrient concentrations at different times, relationships between nutrients, time, and flow/salinity are non-linear and complex, so we expect 1) annual trend independent of seasonal trend, 2) changes in seasonal amplitudes and quantile trends over time, 3) varying flow contribution, either as difference between predicted/flow-normalized results or changes in nutrient v flow scatterplots at different annual periods.
+
+Multiple biological and physical factors influence nutrient concentrations at different temporal scales \citep[][, references therein]{Cloern10}.
+As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex.
+A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, is exptected reveal \begin{inparaenum}[1\upshape)]
+\item an annual trend independent of a seasonal trend, and 
+\item varying flow contribution, either as a difference between predicted and flow-normalized results or changes in the nutrient versus flow relationshps at different annual periods.
+\end{inparaenum}
+Results from this case study were used to demonstrate the breadth of information that can be obtained from the observed time series with \ac{wrtds}.
+
 \subsubsection{Effects of wastewater treatment}

### 2016-11-10 sftrends_manu.tex
alpine92 ref added

@@
   \vspace{1in} 
-  \\ Version Date:   Mon Oct 31 12:28:10 2016 -0500
+  \\ Version Date:   Tue Nov 1 11:08:12 2016 -0500
 	}
@@
 As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex.
-A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, is exptected reveal \begin{inparaenum}[1\upshape)]
+A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, is expected to reveal \begin{inparaenum}[1\upshape)]
 \item an annual trend independent of a seasonal trend, and 
-\item varying flow contribution, either as a difference between predicted and flow-normalized results or changes in the nutrient versus flow relationshps at different annual periods.
+\item varying flow contribution to nutrient dynamics, either as a difference between predicted and flow-normalized results or changes in the nutrient versus flow relationship at different annual periods.
 \end{inparaenum}
@@
 This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
+See suggestions in \citet{Alpine92} regarding flow/grazer relationships in the Bay.
 

### 2016-11-17 README.Rmd
tracy plot finished, h2dat added
replaced tracy data with stockton data

@@
 
+`h2dat.RData`* mean models to support second hypothesis/case study in manuscript, wrtds mean models for no23, nh at p8
+
 `h3dat.RData`* mean models to support third hypothesis/case study in manuscript, wrtds mean models for sio2, din, chla at c10, d7
@@
 
-`potw_load.RData`* nutrient load data from Stockton and Tracy wwtp, kg/day
+`stock_conc.RData` summarized effluent data from Stockton wwtp, mg/L
+
+`stock_load.RData` nutrient load data from Stockton, kg/day

### 2016-11-17 sftrends_manu.Rnw
alpine92 ref added
replaced tracy data with stockton data

@@
 
-Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more detailed understanding of trends that can lead to additional hypotheses for further evaluation.
-The stations were also chosen to address ecological and management-based questions that have relevance outside of the region, having importance for our understanding of estuarine processes that influence eutrophication trends on decadal time scales.
+Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more detailed understanding of trends for further evaluation.
+The stations werechosen to address ecological and management-based questions that have relevance outside of the region, having importance for our understanding of estuarine processes that influence eutrophication trends on decadal time scales.
 The selected case studies focused on \begin{inparaenum}[1\upshape)]
@@
 \end{inparaenum}
-Each case study is built around hypotheses that results from \ac{wrtds} models are expected to support, both as a general description and as a means for additional testing with alternative methods.
+Each case study is built around hypotheses that results from \ac{wrtds} models are expected to support, both as a general description and for additional testing with alternative methods.
 
@@
 \subsubsection{Effects of wastewater treatment}
+
 Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) different nitrogen species will have different changes depending on change in load outputs.
-See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
+See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}, but this is about Tracy
 \subsubsection{Effects of biological invasions}
+
 Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
@@
 
-Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:tracy}).
+Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:stock}).
 Nitrate is the primary constituent of total nitrogen after 2007.
@@
 
-<<tracy, fig.height = 5, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in May, 2007 to include nitrification/denitrification and tertiary filtration causing a reduction in total nitrogen effluent discharged to the Delta.
-Reductions were primarily observed for ammonium.'>>=
-data(potw_load)
+<<stock, fig.height = 5, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
+data(stock_conc)
+
+stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))
 
@@
 
-toplo1 <- filter(potw_load, loc %in% 'tracy' & var %in% c('tn'))
-toplo2 <- filter(potw_load, loc %in% 'tracy' & var %in% c('nh4', 'no3', 'orgn'))
-
-ylab <- expression(paste("Total nitrogen load (kg ", d^-1, ")"))
+toplo1 <- stock_conc %>% 
+  select(-val2) %>% 
+  spread(var, val) %>% 
+  mutate(TN = NH3 + NO2 + NO3)
+toplo2 <- stock_conc
+
+ylab1 <- expression(paste("Total nitrogen (mg ", L^-1, ")"))
+ylab2 <- '% of total nitrogen'
+xrng <- range(c(toplo1$date, toplo2$date))
+yrng <- c(0, 50)
 leglabs <- list(
   expression(NH [4] ^ '+'),
-  expression(NO [3] ^ '2-'),
-  expression('organic N')
-  )
-
-lndate <- as.Date('2007/05/01')
+  expression(NO [2] ^ '-'),
+  expression(NO [3] ^ '2-')
+  )
+
+lndate <- as.Date('2006/01/01')
 
 p1 <- ggplot(toplo1) + 
-  geom_line(colour = lncol, aes(x = date, y = val)) + 
-  scale_x_date(expand = c(0, 0)) + 
-  scale_y_continuous(ylab, expand = c(0, 0), limits = c(50, 850)) + 
+  geom_line(colour = lncol, aes(x = date, y = TN)) + 
+  scale_x_date(expand = c(0, 0), limits = xrng) + 
+  scale_y_continuous(ylab1, expand = c(0, 0), limits = yrng) + 
   geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
   geom_label_repel(
-    data = data.frame(x = lndate, y = 600), 
+    data = data.frame(x = lndate, y = 35), 
     aes(x, y, label = 'WWTP upgrades'), 
@@
 
-p2 <- ggplot(toplo2, aes(x = date, y = val, group = var, fill = var)) + 
-  geom_area(stat = 'identity', position = 'fill') + 
-  scale_x_date(expand = c(0, 0)) + 
+p2 <- ggplot(toplo2) + 
+  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'fill') + 
+  scale_x_date(expand = c(0, 0), limits = xrng) + 
   scale_fill_brewer(palette = 'Blues', labels = leglabs) + 
-  scale_y_continuous('% of total nitrogen', expand = c(0, 0)) + 
   geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
+  scale_y_continuous(ylab2, expand = c(0, 0)) + 
   ptheme + 
@@
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Tracy (San Joaquin County), were completed in May 2007 (\\cref{fig:tracy}), coincident with a dramatic decrease in ammonium at P8.'>>=
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County), were completed in May 2007 (\\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.'>>=
 data(h2dat)
@@
 cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in May 2007 at the City of Tracy \\ac{wwtp} (San Joaquin County, \\cref{fig:tracy}).
+Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, \\cref{fig:stock}).
 Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).

### 2016-11-17 sftrends_manu.tex
replaced tracy data with stockton data

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Nov 1 11:08:12 2016 -0500
+  \\ Version Date:   Thu Nov 10 16:43:55 2016 -0600
 	}
@@
 
-{\centering \includegraphics[width=\textwidth]{figs/tracy-1}
+{\centering \includegraphics[width=\textwidth]{figs/stock-1}
 
@@
 
-\caption[Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County]{Nitrogen load measurements (kg d$^{-1}$) at the City of Tracy Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in May, 2007 to include nitrification/denitrification and tertiary filtration causing a reduction in total nitrogen effluent discharged to the Delta.
-Reductions were primarily observed for ammonium.}\label{fig:tracy}
+\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
 \end{figure}
@@
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Tracy (San Joaquin County), were completed in May 2007 (\cref{fig:tracy}), coincident with a dramatic decrease in ammonium at P8.}\label{fig:p8trnds}
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County), were completed in May 2007 (\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.}\label{fig:p8trnds}
 \end{figure}
@@
 \caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in May 2007 at the City of Tracy \ac{wwtp} (San Joaquin County, \cref{fig:tracy}).
+Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
 Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).

### 2016-11-18 R/funcs.R
added trnds_sk data object

@@ -7,6 +7,39 @@ g_legend <- function(a.gplot){
   return(legend)}
 
 ######
+# some stupid function for formatting p-values, no asterisks
+# one value at a time
+p_form <- function(x, tex_out = F){
+  if(x < 0.005) out <-'< 0.005'
+  else{
+    if(x < 0.05 & x >= 0.005) out <- as.character(format(round(x, 3), nsmall = 3))
+    else out <- 'ns'
+    }
+  if(tex_out == F) return(out)
+  else{
+    if(out == '< 0.005') '$p<$ 0.005'
+    else{
+      if(out == 'ns') '$p = $ ns'
+      else paste('$p = $', out, sep = ' ')
+      }
+    }
+  }
+
+######
+# another stupid function for formatting p-values as asterisks
+p_ast <- function(x){
+  
+  sig_cats <- c('**', '*', '')
+  sig_vals <- c(-Inf, 0.005, 0.05, Inf)
+  
+  out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
+  out <- as.character(out)
+  
+  return(out)
+  
+}
+
+######
 # trend maps
 # 
 # library(dplyr)

### 2016-11-18 sftrends_manu.Rnw
tables updated with kendall results

@@
 
-A total of twenty-seven \ac{wrtds} models were created, one for each nitrogen at each station.
+A total of twenty-seven \ac{wrtds} models were created, one for each nitrogen analyte at each station.
 The functional form of \ac{wrtds} is a simple regression that models the log-transformed response variable as a function of time, flow, and season:
@@
 <<trnds>>=
-data(trnds_chg)
-data(trnds_ave)
+data(trnds_sk)
 
 # percent changes
-tab_chg <- trnds_chg
-
-# increase in bold italic
-tab_chg[, -c(1:2)] <- round(tab_chg[, -c(1:2)], 1)
-tab_chg[, -c(1:2)][sign(tab_chg[, -c(1:2)]) != -1] <- paste0('\\textit{\\textbf{', tab_chg[, -c(1:2)][sign(tab_chg[, -c(1:2)]) != -1], '}}')
-tab_chg <- gather(tab_chg, 'var', 'chg', -Site_Code, -resvar) %>% 
-  mutate(chg = paste0('\\footnotesize{(', chg, ')}'))
-
-# averages
-tab_ave <- gather(trnds_ave, 'var', 'ave', -Site_Code, -resvar) %>% 
-  mutate(ave = round(ave, 1))
-
-tab <- left_join(tab_ave, tab_chg, by = c('Site_Code', 'resvar', 'var')) %>% 
-  unite(val, ave, chg, sep = ' ') %>% 
-  spread(var, val) %>% 
+tab <- trnds_sk %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
   arrange(resvar, Site_Code) %>% 
@@
 cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
-Summaries are averages (mg L$^{-1}$) and percent changes in parentheses (increasing in bold-italic).
-Changes are based on the difference between the ending and starting estimates within each period.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 See \\cref{fig:trndmap} for a summary of spatial trends.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
   rowname = stats,
-  label = 'tab:trnds'
+  label = 'tab:trnds',
+  insert.bottom = foot.val
   )
@@
 data(h2dat)
-dt <- as.Date('2007-05-01')
+dt <- as.Date('2006-01-01')
 
@@
 mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
-yrbrks_bf <- c(1975, 2007)
+yrbrks_bf <- c(1975, 2005)
 yrbrks_af <- c(2006, 2013)
 molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
-yrlabs_bf <- '1976-2007'
-yrlabs_af <- '2008-2013'
-faclevs <- c('1976-2007_before', '2008-2013_after', paste0(molabs, '_before'), paste0(molabs, '_after'))
+yrlabs_bf <- '1976-2006'
+yrlabs_af <- '2007-2013'
+faclevs <- c('1976-2006_before', '2007-2013_after', paste0(molabs, '_before'), paste0(molabs, '_after'))
 
 # summarize din, sio2, chl trends
-no_bf <- wrtdstrnd(noraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf, aves = TRUE, min_mo = 10) %>% 
+no_bf <- wrtdstrnd_sk(noraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf) %>% 
   mutate(period = 'before', var = 'no')
-no_af <- wrtdstrnd(noraw_af, mobrks, yrbrks_af, molabs, yrlabs_af, aves = TRUE, min_mo = 10) %>% 
+no_af <- wrtdstrnd_sk(noraw_af, mobrks, yrbrks_af, molabs, yrlabs_af) %>% 
   mutate(period = 'after', var = 'no')
-nh_bf <- wrtdstrnd(nhraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf, aves = TRUE, min_mo = 10) %>% 
+nh_bf <- wrtdstrnd_sk(nhraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf) %>% 
   mutate(period = 'before', var = 'nh')
-nh_af <- wrtdstrnd(nhraw_af, mobrks, yrbrks_af, molabs, yrlabs_af, aves = TRUE, min_mo = 10) %>% 
+nh_af <- wrtdstrnd_sk(nhraw_af, mobrks, yrbrks_af, molabs, yrlabs_af) %>% 
   mutate(period = 'after', var = 'nh')
@@
 totab <- rbind(no_bf, no_af, nh_bf, nh_af) %>% 
+  select(-tau, -slope, -chitest) %>% 
   mutate(
-    chg = round(chg, 1), 
-    chg = ifelse(chg < 0, chg, paste0('\\textit{\\textbf{', chg , '}}')), 
-    ave = as.character(round(ave, 2))
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      perchg, 
+      paste0('\\textit{\\textbf{', perchg, '}}')
+      ), 
+    ztest = p_ast(ztest)
   ) %>% 
-  gather('est', 'val', chg:ave) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  gather('est', 'val', med:perchg) %>% 
   unite('col', var, est) %>% 
@@
   data.frame(sep = c(rep('Annual', 2), rep('Seasonal, pre', 4), rep('Seasonal, post', 4)), .) %>% 
-  select(sep, cat, no_ave, no_chg, nh_ave, nh_chg)
+  select(sep, cat, no_med, no_perchg, nh_med, nh_perchg)
 
@@
 vars <- c('NO$_{2}^{-}$/NO$_{3}^{2-}$', 'NH$_{4}^{+}$')
-names(totab) <- rep(c('Mean', '\\% change'), 2)
+names(totab) <- rep(c('Median', '\\% change'), 2)
 
@@
 Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, \\cref{fig:stock}).
-Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Increasing values are in bold-italics.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
 
@@
   rowname = cats,
-  label = 'tab:p8chg'
+  label = 'tab:p8chg',
+  insert.bottom = foot.val
   )
@@
 # summarize din, sio2, chl trends
-din <- wrtdstrnd(dinraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE, min_mo = 10)
-din <- wrtdstrnd(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE, min_mo = 10) %>% 
+din <- wrtdstrnd_sk(dinraw, mobrks, yrbrks, molabs, yrlabs)
+din <- wrtdstrnd_sk(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
   .[1, ] %>% 
@@
   data.frame(res = 'din', .)
-chl <- wrtdstrnd(chlraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE, min_mo = 10)
-chl <- wrtdstrnd(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE, min_mo = 10) %>% 
+chl <- wrtdstrnd_sk(chlraw, mobrks, yrbrks, molabs, yrlabs)
+chl <- wrtdstrnd_sk(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
   .[1, ] %>% 
@@
   data.frame(res = 'chl', .)
-sio <- wrtdstrnd(sioraw, mobrks, yrbrks, molabs, yrlabs, aves = TRUE, min_mo = 10)
-sio <- wrtdstrnd(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013'), aves = TRUE, min_mo = 10) %>% 
+sio <- wrtdstrnd_sk(sioraw, mobrks, yrbrks, molabs, yrlabs)
+sio <- wrtdstrnd_sk(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
   .[1, ] %>% 
@@
 totab <- rbind(din, chl, sio) %>% 
+  select(res, cat, med, perchg, ztest) %>% 
   mutate(
-    chg = round(chg, 1), 
-    chg = ifelse(chg < 0, chg, paste0('\\textit{\\textbf{', chg , '}}')), 
-    ave = as.character(round(ave, 1))
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      perchg, 
+      paste0('\\textit{\\textbf{', perchg, '}}')
+      ), 
+    ztest = p_ast(ztest)
   ) %>% 
-  gather('var', 'val', chg:ave) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  gather('var', 'val', med:perchg) %>% 
   unite('col', res, var) %>% 
@@
   data.frame(sep = c(rep('Annual', 5), rep('Seasonal', 4)), .) %>% 
-  select(sep, cat, din_ave, din_chg, chl_ave, chl_chg, sio_ave, sio_chg)
+  select(sep, cat, din_med, din_perchg, chl_med, chl_perchg, sio_med, sio_perchg)
 
@@
 vars <- c('DIN', 'Chl-\\textit{a}', 'SiO$_2$')
-names(totab) <- rep(c('Mean', '\\% change'), 3)
+names(totab) <- rep(c('Median', '\\% change'), 3)
 
 cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
-Summaries are means and percent changes based on annual means within the time periods.
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Increasing values are in bold-italics.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
 
@@
   rowname = cats,
-  label = 'tab:d7chg'
+  label = 'tab:d7chg', 
+  insert.bottom = foot.val
   )
@@
 vars <- c('DIN', 'Chl-\\textit{a}', 'SiO$_2$')
-names(totab) <- rep(c('Mean', '\\% change'), 3)
-
-cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station C10.
-Summaries are means and percent changes based on annual means within the time periods.
+names(totab) <- rep(c('Median', '\\% change'), 3)
+
+cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station C10.
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Increasing values are in bold-italics.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
 
@@
   rowname = cats,
-  label = 'tab:c10chg'
+  label = 'tab:c10chg', 
+  insert.bottom = foot.val
   )

### 2016-11-18 sftrends_manu.tex
added trnds_sk data object
tables updated with kendall results

@@
   \vspace{1in} 
-  \\ Version Date:   Thu Nov 10 16:43:55 2016 -0600
+  \\ Version Date:   Fri Nov 18 13:04:31 2016 -0600
 	}
@@
 % overall trends
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual", "Seasonal"), n.cgroup = c(2,         4), rowname = stats, label = "tab:trnds")%
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual", "Seasonal"), n.cgroup = c(2,         4), rowname = stats, label = "tab:trnds", insert.bottom = foot.val)%
 \begin{table}[!tbp]
 \caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
-Summaries are averages (mg L$^{-1}$) and percent changes in parentheses (increasing in bold-italic).
-Changes are based on the difference between the ending and starting estimates within each period.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 See \cref{fig:trndmap} for a summary of spatial trends.
@@
 {\bfseries DIN}&&&&&&&\tabularnewline
-~~C10&1.3 \footnotesize{(\textit{\textbf{35}})}&1.4 \footnotesize{(-28.6)}&&1.1 \footnotesize{(-25.8)}&1.2 \footnotesize{(-6.3)}&1.5 \footnotesize{(\textit{\textbf{5.5}})}&1.7 \footnotesize{(-20.7)}\tabularnewline
-~~C3&0.4 \footnotesize{(\textit{\textbf{51.1}})}&0.5 \footnotesize{(\textit{\textbf{3}})}&&0.4 \footnotesize{(\textit{\textbf{62.2}})}&0.3 \footnotesize{(\textit{\textbf{71.9}})}&0.5 \footnotesize{(\textit{\textbf{54.4}})}&0.5 \footnotesize{(\textit{\textbf{18.3}})}\tabularnewline
-~~D19&0.5 \footnotesize{(-5.5)}&0.4 \footnotesize{(-30.7)}&&0.5 \footnotesize{(-33.9)}&0.3 \footnotesize{(-25.4)}&0.3 \footnotesize{(-21.7)}&0.7 \footnotesize{(-33.5)}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.3}})}&0.5 \footnotesize{(-19.6)}&&0.5 \footnotesize{(-19.2)}&0.3 \footnotesize{(-10.2)}&0.4 \footnotesize{(-3.5)}&0.6 \footnotesize{(-21.7)}\tabularnewline
-~~D28&0.5 \footnotesize{(-3.1)}&0.4 \footnotesize{(-45.6)}&&0.5 \footnotesize{(-32.9)}&0.2 \footnotesize{(-56.5)}&0.3 \footnotesize{(-57.4)}&0.8 \footnotesize{(-40.2)}\tabularnewline
-~~D4&0.4 \footnotesize{(-8.1)}&0.4 \footnotesize{(-7.5)}&&0.4 \footnotesize{(-13.6)}&0.3 \footnotesize{(\textit{\textbf{4.3}})}&0.4 \footnotesize{(-1.2)}&0.5 \footnotesize{(-15)}\tabularnewline
-~~D6&0.4 \footnotesize{(\textit{\textbf{11.6}})}&0.5 \footnotesize{(-6.1)}&&0.5 \footnotesize{(-5)}&0.4 \footnotesize{(\textit{\textbf{64.1}})}&0.4 \footnotesize{(\textit{\textbf{21}})}&0.5 \footnotesize{(-14.6)}\tabularnewline
-~~D7&0.4 \footnotesize{(-1.7)}&0.5 \footnotesize{(-11.8)}&&0.5 \footnotesize{(-23.8)}&0.4 \footnotesize{(\textit{\textbf{49.9}})}&0.4 \footnotesize{(\textit{\textbf{10.4}})}&0.6 \footnotesize{(-34.6)}\tabularnewline
-~~P8&1.4 \footnotesize{(\textit{\textbf{81.1}})}&1.8 \footnotesize{(-15.8)}&&1.5 \footnotesize{(\textit{\textbf{35.5}})}&1 \footnotesize{(\textit{\textbf{42}})}&1.6 \footnotesize{(\textit{\textbf{55}})}&2.3 \footnotesize{(\textit{\textbf{38.7}})}\tabularnewline
+~~C10&1.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.4 \footnotesize{(-3.1)**}&&1.1 \footnotesize{(-0.1)}&1.2 \footnotesize{(\textit{\textbf{0.1}})}&1.4 \footnotesize{(\textit{\textbf{0.6}})**}&1.7 \footnotesize{(\textit{\textbf{0.1}})}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.5 \footnotesize{(-0.1)**}&&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.3 \footnotesize{(\textit{\textbf{1.9}})**}&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+~~D19&0.4 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(-1.9)**}&&0.5 \footnotesize{(-0.4)**}&0.2 \footnotesize{(\textit{\textbf{0}})}&0.3 \footnotesize{(\textit{\textbf{0.2}})}&0.7 \footnotesize{(-0.5)**}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-1.2)**}&&0.5 \footnotesize{(\textit{\textbf{0}})}&0.3 \footnotesize{(\textit{\textbf{0.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.5}})**}&0.6 \footnotesize{(-0.2)**}\tabularnewline
+~~D28&0.4 \footnotesize{(\textit{\textbf{0.1}})**}&0.4 \footnotesize{(-3.1)**}&&0.5 \footnotesize{(-0.5)**}&0.2 \footnotesize{(-0.9)**}&0.3 \footnotesize{(-1.1)**}&0.8 \footnotesize{(-0.8)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(-0.3)**}&&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.3 \footnotesize{(\textit{\textbf{0.8}})**}&0.4 \footnotesize{(\textit{\textbf{0.5}})**}&0.5 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
+~~D6&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.5 \footnotesize{(-0.3)**}&&0.5 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.6}})**}&0.4 \footnotesize{(\textit{\textbf{0.9}})**}&0.5 \footnotesize{(\textit{\textbf{0.2}})*}\tabularnewline
+~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}&&0.5 \footnotesize{(-0.1)}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.4 \footnotesize{(\textit{\textbf{0.6}})**}&0.6 \footnotesize{(-0.2)}\tabularnewline
+~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}&&1.4 \footnotesize{(\textit{\textbf{0.7}})**}&1 \footnotesize{(\textit{\textbf{1.4}})**}&1.6 \footnotesize{(\textit{\textbf{2}})**}&2.2 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
 \hline
 {\bfseries NH$_{4}^{+}$}&&&&&&&\tabularnewline
-~~C10&0.1 \footnotesize{(-37.7)}&0 \footnotesize{(-55.5)}&&0.1 \footnotesize{(-68.6)}&0 \footnotesize{(-85)}&0.1 \footnotesize{(-90.8)}&0.2 \footnotesize{(-78.8)}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{127}})}&0.3 \footnotesize{(\textit{\textbf{14.1}})}&&0.2 \footnotesize{(\textit{\textbf{165.1}})}&0.2 \footnotesize{(\textit{\textbf{133.3}})}&0.3 \footnotesize{(\textit{\textbf{107.7}})}&0.2 \footnotesize{(\textit{\textbf{81.2}})}\tabularnewline
-~~D19&0.1 \footnotesize{(\textit{\textbf{35.6}})}&0.1 \footnotesize{(-32.6)}&&0.1 \footnotesize{(-32.9)}&0 \footnotesize{(-21)}&0 \footnotesize{(-6.4)}&0.1 \footnotesize{(-12)}\tabularnewline
-~~D26&0.1 \footnotesize{(\textit{\textbf{58.4}})}&0.1 \footnotesize{(-18.7)}&&0.1 \footnotesize{(\textit{\textbf{3.3}})}&0.1 \footnotesize{(\textit{\textbf{38.1}})}&0.1 \footnotesize{(\textit{\textbf{21.1}})}&0.1 \footnotesize{(\textit{\textbf{0.8}})}\tabularnewline
-~~D28&0.1 \footnotesize{(-10)}&0 \footnotesize{(-31.8)}&&0.1 \footnotesize{(-42.8)}&0 \footnotesize{(-36.9)}&0 \footnotesize{(-42.2)}&0.1 \footnotesize{(-35.7)}\tabularnewline
-~~D4&0.1 \footnotesize{(\textit{\textbf{74.6}})}&0.1 \footnotesize{(\textit{\textbf{0.3}})}&&0.1 \footnotesize{(\textit{\textbf{54.2}})}&0 \footnotesize{(\textit{\textbf{30.8}})}&0.1 \footnotesize{(\textit{\textbf{23.8}})}&0.1 \footnotesize{(\textit{\textbf{56.9}})}\tabularnewline
-~~D6&0.1 \footnotesize{(\textit{\textbf{100.2}})}&0.1 \footnotesize{(-3.5)}&&0.1 \footnotesize{(\textit{\textbf{42.2}})}&0.1 \footnotesize{(\textit{\textbf{74.1}})}&0.1 \footnotesize{(\textit{\textbf{49.2}})}&0.1 \footnotesize{(\textit{\textbf{39.8}})}\tabularnewline
-~~D7&0.1 \footnotesize{(\textit{\textbf{88.2}})}&0.1 \footnotesize{(-13.3)}&&0.1 \footnotesize{(\textit{\textbf{26.8}})}&0 \footnotesize{(\textit{\textbf{25.1}})}&0.1 \footnotesize{(\textit{\textbf{4.3}})}&0.1 \footnotesize{(\textit{\textbf{16.8}})}\tabularnewline
-~~P8&0.3 \footnotesize{(\textit{\textbf{267.4}})}&0.3 \footnotesize{(-85.1)}&&0.2 \footnotesize{(-51.8)}&0.1 \footnotesize{(-12)}&0.3 \footnotesize{(-44.5)}&0.6 \footnotesize{(-61.2)}\tabularnewline
+~~C10&0.1 \footnotesize{(-3.4)**}&0 \footnotesize{(-5.2)**}&&0 \footnotesize{(-4.4)**}&0 \footnotesize{(-6.9)**}&0.1 \footnotesize{(-8.7)**}&0.2 \footnotesize{(-4.3)**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{3.7}})**}&0.3 \footnotesize{(\textit{\textbf{0}})}&&0.2 \footnotesize{(\textit{\textbf{2.9}})**}&0.2 \footnotesize{(\textit{\textbf{2.8}})**}&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.2 \footnotesize{(\textit{\textbf{2}})**}\tabularnewline
+~~D19&0 \footnotesize{(\textit{\textbf{0.4}})**}&0 \footnotesize{(-1.7)**}&&0.1 \footnotesize{(-0.7)**}&0 \footnotesize{(-0.5)**}&0 \footnotesize{(\textit{\textbf{0}})}&0.1 \footnotesize{(-0.6)**}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{2.2}})**}&0.1 \footnotesize{(-1.5)**}&&0.1 \footnotesize{(\textit{\textbf{0}})}&0.1 \footnotesize{(\textit{\textbf{1}})**}&0.1 \footnotesize{(\textit{\textbf{0.5}})**}&0.1 \footnotesize{(-0.2)*}\tabularnewline
+~~D28&0 \footnotesize{(-1.1)**}&0 \footnotesize{(-1.4)**}&&0 \footnotesize{(-1.2)**}&0 \footnotesize{(-0.4)*}&0 \footnotesize{(-0.9)**}&0.1 \footnotesize{(-1.4)**}\tabularnewline
+~~D4&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0}})}&&0.1 \footnotesize{(\textit{\textbf{1.7}})**}&0 \footnotesize{(\textit{\textbf{1.3}})**}&0.1 \footnotesize{(\textit{\textbf{1.2}})**}&0.1 \footnotesize{(\textit{\textbf{1.4}})**}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{2.4}})**}&0.1 \footnotesize{(-0.5)**}&&0.1 \footnotesize{(\textit{\textbf{0.6}})**}&0.1 \footnotesize{(\textit{\textbf{1.4}})**}&0.1 \footnotesize{(\textit{\textbf{1.1}})**}&0.1 \footnotesize{(\textit{\textbf{0.7}})**}\tabularnewline
+~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}&&0.1 \footnotesize{(\textit{\textbf{0.5}})*}&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0.4}})**}&0.1 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
+~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}&&0.2 \footnotesize{(-0.5)}&0.1 \footnotesize{(\textit{\textbf{1.3}})**}&0.2 \footnotesize{(\textit{\textbf{1.9}})*}&0.6 \footnotesize{(\textit{\textbf{1}})}\tabularnewline
 \hline
 {\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&&&&\tabularnewline
-~~C10&1.2 \footnotesize{(\textit{\textbf{42.8}})}&1.3 \footnotesize{(-26.9)}&&1.1 \footnotesize{(-21)}&1.2 \footnotesize{(-2.3)}&1.4 \footnotesize{(\textit{\textbf{23.3}})}&1.5 \footnotesize{(-5.6)}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{0.7}})}&0.2 \footnotesize{(-6.6)}&&0.2 \footnotesize{(\textit{\textbf{0.9}})}&0.1 \footnotesize{(\textit{\textbf{18}})}&0.1 \footnotesize{(\textit{\textbf{4.3}})}&0.2 \footnotesize{(-17.4)}\tabularnewline
-~~D19&0.4 \footnotesize{(-11.7)}&0.4 \footnotesize{(-31.1)}&&0.4 \footnotesize{(-32.3)}&0.2 \footnotesize{(-36.7)}&0.3 \footnotesize{(-31.6)}&0.6 \footnotesize{(-38.4)}\tabularnewline
-~~D26&0.3 \footnotesize{(-9.8)}&0.4 \footnotesize{(-20.1)}&&0.4 \footnotesize{(-23.5)}&0.2 \footnotesize{(-18.7)}&0.3 \footnotesize{(-10.5)}&0.5 \footnotesize{(-26.4)}\tabularnewline
-~~D28&0.4 \footnotesize{(-10.5)}&0.4 \footnotesize{(-47.2)}&&0.5 \footnotesize{(-34.6)}&0.2 \footnotesize{(-66.3)}&0.3 \footnotesize{(-64.4)}&0.7 \footnotesize{(-45.7)}\tabularnewline
-~~D4&0.3 \footnotesize{(-14.6)}&0.3 \footnotesize{(-9.8)}&&0.3 \footnotesize{(-19.2)}&0.3 \footnotesize{(\textit{\textbf{1.1}})}&0.3 \footnotesize{(-8)}&0.4 \footnotesize{(-25.9)}\tabularnewline
-~~D6&0.3 \footnotesize{(-5.6)}&0.4 \footnotesize{(-8)}&&0.4 \footnotesize{(-14.5)}&0.3 \footnotesize{(\textit{\textbf{38.8}})}&0.3 \footnotesize{(\textit{\textbf{4}})}&0.4 \footnotesize{(-23.2)}\tabularnewline
-~~D7&0.4 \footnotesize{(-13.4)}&0.4 \footnotesize{(-13.6)}&&0.4 \footnotesize{(-27.3)}&0.3 \footnotesize{(\textit{\textbf{28.7}})}&0.4 \footnotesize{(-23.7)}&0.5 \footnotesize{(-46.8)}\tabularnewline
-~~P8&1.1 \footnotesize{(\textit{\textbf{59.8}})}&1.5 \footnotesize{(\textit{\textbf{3.3}})}&&1.2 \footnotesize{(\textit{\textbf{53.9}})}&1 \footnotesize{(\textit{\textbf{38.8}})}&1.4 \footnotesize{(\textit{\textbf{58.3}})}&1.8 \footnotesize{(\textit{\textbf{62.3}})}\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{1.4}})**}&1.4 \footnotesize{(-3)**}&&1.1 \footnotesize{(\textit{\textbf{0.1}})}&1.2 \footnotesize{(\textit{\textbf{0.2}})}&1.3 \footnotesize{(\textit{\textbf{1.2}})**}&1.5 \footnotesize{(\textit{\textbf{0.8}})**}\tabularnewline
+~~C3&0.1 \footnotesize{(-0.1)**}&0.2 \footnotesize{(\textit{\textbf{0.7}})**}&&0.2 \footnotesize{(\textit{\textbf{0.5}})**}&0.1 \footnotesize{(\textit{\textbf{0.7}})**}&0.1 \footnotesize{(\textit{\textbf{0.9}})**}&0.2 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
+~~D19&0.4 \footnotesize{(-0.1)**}&0.4 \footnotesize{(-1.9)**}&&0.4 \footnotesize{(-0.3)*}&0.2 \footnotesize{(-0.3)*}&0.3 \footnotesize{(-0.1)}&0.6 \footnotesize{(-0.6)**}\tabularnewline
+~~D26&0.3 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1.1)**}&&0.4 \footnotesize{(\textit{\textbf{0}})}&0.2 \footnotesize{(\textit{\textbf{0.2}})*}&0.3 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-0.1)}\tabularnewline
+~~D28&0.4 \footnotesize{(-0.2)**}&0.4 \footnotesize{(-3.1)**}&&0.5 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-1.5)**}&0.3 \footnotesize{(-1.4)**}&0.7 \footnotesize{(-0.8)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(-0.4)**}&&0.3 \footnotesize{(\textit{\textbf{0.2}})*}&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(\textit{\textbf{0.2}})*}&0.4 \footnotesize{(\textit{\textbf{0.1}})}\tabularnewline
+~~D6&0.3 \footnotesize{(\textit{\textbf{1.3}})**}&0.4 \footnotesize{(-0.3)**}&&0.4 \footnotesize{(\textit{\textbf{0.2}})*}&0.4 \footnotesize{(\textit{\textbf{1.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(\textit{\textbf{0.1}})}\tabularnewline
+~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}&&0.4 \footnotesize{(-0.3)**}&0.4 \footnotesize{(\textit{\textbf{1.2}})**}&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-0.5)**}\tabularnewline
+~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}&&1.2 \footnotesize{(\textit{\textbf{1}})**}&1 \footnotesize{(\textit{\textbf{1.2}})**}&1.4 \footnotesize{(\textit{\textbf{1.7}})**}&1.6 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
 \hline
@@
 
-\end{table}
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
 
 % trends in no23, nh at p8
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg")%
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg", insert.bottom = foot.val)%
 \begin{table}[!tbp]
@@
 Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
-Summaries are means and percent changes based on annual means within the pre- and post-upgrade time periods (1976-2007, 2008-2013).
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Increasing values are in bold-italics.
@@
 \cline{2-3} \cline{5-6}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
 \hline
 {\bfseries Annual}&&&&&\tabularnewline
-~~1976-2007&1.3&\textit{\textbf{74.3}}&&0.34&\textit{\textbf{93.1}}\tabularnewline
-~~2008-2013&1.5&-7.3&&0.11&-62\tabularnewline
+~~1976-2006&1.3&\textit{\textbf{2}}**&&0.2&\textit{\textbf{2.8}}**\tabularnewline
+~~2007-2013&1.4&-1.9**&&0.1&-16.6**\tabularnewline
 \hline
 {\bfseries Seasonal, pre}&&&&&\tabularnewline
-~~Spring&1.24&\textit{\textbf{79.5}}&&0.25&\textit{\textbf{57.1}}\tabularnewline
-~~Summer&0.99&\textit{\textbf{95}}&&0.09&\textit{\textbf{143.5}}\tabularnewline
-~~Fall&1.32&\textit{\textbf{77}}&&0.29&\textit{\textbf{137.7}}\tabularnewline
-~~Winter&1.67&\textit{\textbf{69.2}}&&0.69&\textit{\textbf{72.5}}\tabularnewline
+~~Spring&1.2&\textit{\textbf{1.6}}**&&0.2&\textit{\textbf{1.4}}**\tabularnewline
+~~Summer&1&\textit{\textbf{2.4}}**&&0.1&\textit{\textbf{3.3}}**\tabularnewline
+~~Fall&1.3&\textit{\textbf{2.2}}**&&0.2&\textit{\textbf{4.9}}**\tabularnewline
+~~Winter&1.5&\textit{\textbf{2.1}}**&&0.7&\textit{\textbf{4.8}}**\tabularnewline
 \hline
 {\bfseries Seasonal, post}&&&&&\tabularnewline
-~~Spring&1.27&\textit{\textbf{0.9}}&&0.09&-49.2\tabularnewline
-~~Summer&0.92&-18.7&&0.06&-49.2\tabularnewline
-~~Fall&1.58&-8.7&&0.1&-58.2\tabularnewline
-~~Winter&2.17&-5.1&&0.18&-62.1\tabularnewline
+~~Spring&1.3&-1.6**&&0.1&-16.2**\tabularnewline
+~~Summer&0.9&-4.3**&&0.1&-15.7**\tabularnewline
+~~Fall&1.5&-1.7**&&0.1&-19.3**\tabularnewline
+~~Winter&2.2&-0.8**&&0.2&-26.7**\tabularnewline
 \hline
@@
 
-\end{table}
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
 
 % trends in din, chl, sio2 at d7
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:d7chg")%
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:d7chg", insert.bottom = foot.val)%
 \begin{table}[!tbp]
 \caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
-Summaries are means and percent changes based on annual means within the time periods.
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Increasing values are in bold-italics.
@@
 \cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
 \hline
 {\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&0.4&-7.8&&4.1&-76.7&&12.7&\textit{\textbf{21.8}}\tabularnewline
+~~1976-2013&0.4&\textit{\textbf{0.6}}**&&3&-6.7**&&12.7&\textit{\textbf{0.7}}**\tabularnewline
 \hline
 {\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&0.4&-22.5&&9.3&-57.1&&10.9&-0.1\tabularnewline
-~~1986-1994&0.4&\textit{\textbf{29.4}}&&3&-61.7&&12.2&\textit{\textbf{18.6}}\tabularnewline
-~~1995-2003&0.5&-0.1&&2&\textit{\textbf{21.1}}&&13.7&\textit{\textbf{5.2}}\tabularnewline
-~~2004-2013&0.5&-11.9&&2.4&\textit{\textbf{34.2}}&&13.7&-1.7\tabularnewline
+~~1976-1985&0.4&-2.1**&&8.8&-10.7**&&10.2&-0.2\tabularnewline
+~~1986-1994&0.4&\textit{\textbf{3.6}}**&&2.6&-13.5**&&11.9&\textit{\textbf{2.3}}**\tabularnewline
+~~1995-2003&0.5&-0.1&&1.8&\textit{\textbf{1.9}}**&&13.3&\textit{\textbf{0.7}}**\tabularnewline
+~~2004-2013&0.5&-1.3**&&2.1&\textit{\textbf{2.9}}**&&13.1&-0.3**\tabularnewline
 \hline
 {\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~Spring&0.5&-23.8&&4.2&-60.3&&14.5&\textit{\textbf{0.6}}\tabularnewline
-~~Summer&0.4&\textit{\textbf{49.9}}&&6.4&-82.3&&11.2&\textit{\textbf{51.3}}\tabularnewline
-~~Fall&0.4&\textit{\textbf{10.4}}&&3.7&-84.9&&11.1&\textit{\textbf{43.3}}\tabularnewline
-~~Winter&0.6&-34.6&&2&-63.8&&14.2&\textit{\textbf{4.2}}\tabularnewline
+~~Spring&0.5&-0.1&&3.4&-1&&14.7&\textit{\textbf{0.1}}**\tabularnewline
+~~Summer&0.4&\textit{\textbf{1.5}}**&&3.4&-8.8**&&12.2&\textit{\textbf{1.2}}**\tabularnewline
+~~Fall&0.4&\textit{\textbf{0.6}}**&&1.7&-8.8**&&12.1&\textit{\textbf{1}}**\tabularnewline
+~~Winter&0.6&-0.2&&1.4&-3.1**&&14.5&\textit{\textbf{0.3}}**\tabularnewline
 \hline
@@
 
-\end{table}
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
 
 % trends in din, chl, sio2 at c10
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:c10chg")%
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:c10chg", insert.bottom = foot.val)%
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg $L^{-1}$), chlorophyll ($\mu$g $L^{-1}$), and silicon dioxide (mg $L^{-1}$) concentrations for different time periods at station C10.
-Summaries are means and percent changes based on annual means within the time periods.
+\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station C10.
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Increasing values are in bold-italics.
@@
 \cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Mean}&\multicolumn{1}{c}{\% change}\tabularnewline
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
 \hline
 {\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&1.4&-10.1&&33.2&-20.6&&14.4&-19.1\tabularnewline
+~~1976-2013&1.3&\textit{\textbf{0.1}}**&&20.4&\textit{\textbf{0.1}}**&&14.7&-0.6**\tabularnewline
 \hline
 {\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&1.3&\textit{\textbf{22.8}}&&27.5&-21.6&&15.8&-3.2\tabularnewline
-~~1986-1994&1.4&\textit{\textbf{1.2}}&&35.3&\textit{\textbf{31.5}}&&14.8&-4.1\tabularnewline
-~~1995-2003&1.6&\textit{\textbf{16.4}}&&41&-27.5&&14.2&-3.3\tabularnewline
-~~2004-2013&1.3&-36.3&&30.2&-12.8&&13.2&-7.6\tabularnewline
+~~1976-1985&1.3&\textit{\textbf{3}}**&&19.9&-2.9**&&15.9&-0.2**\tabularnewline
+~~1986-1994&1.3&\textit{\textbf{0.4}}**&&19.6&\textit{\textbf{0.2}}*&&15&-0.3**\tabularnewline
+~~1995-2003&1.5&\textit{\textbf{2.4}}**&&21.2&-0.3**&&14.4&-0.4**\tabularnewline
+~~2004-2013&1.3&-5.2**&&21.4&-0.4**&&13.2&-1**\tabularnewline
 \hline
 {\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~Spring&1.1&-25.8&&26.2&\textit{\textbf{3.1}}&&13.2&-20\tabularnewline
-~~Summer&1.2&-6.3&&79.6&-23.2&&13.9&-25.4\tabularnewline
-~~Fall&1.5&\textit{\textbf{5.5}}&&19&-39.8&&15.6&-18\tabularnewline
-~~Winter&1.7&-20.7&&7.7&\textit{\textbf{7.3}}&&15.1&-14.1\tabularnewline
+~~Spring&1.1&-0.1&&27.2&\textit{\textbf{1.1}}**&&13.2&-0.7**\tabularnewline
+~~Summer&1.2&\textit{\textbf{0.1}}&&73.4&-0.3&&13.5&-0.9**\tabularnewline
+~~Fall&1.4&\textit{\textbf{0.6}}**&&14.4&-1.2**&&15.7&-0.5**\tabularnewline
+~~Winter&1.7&\textit{\textbf{0.1}}&&7.6&\textit{\textbf{0.7}}**&&15.2&-0.3**\tabularnewline
 \hline
@@
 
-\end{table}
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
 

### 2016-11-18 sftrends_manu.Rnw
working on subplot map

@@
 library(ggrepel)
+library(ggtree)
 @
@@
 trnd_map(mods = mods, yrlabs = NULL, molabs = molabs, mobrks = mobrks, sz_rng = c(2, 8), stat_labs = F, leg_lab = '(b) Seasonal % change')
+@
+
+<<trndmap2, fig.height = 9, fig.width = 6, cache = T, out.width = '0.75\\textwidth', fig.cap = 'test'>>=
+
+res = c('din', 'nh', 'no23')
+mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
+yrbrks <- c(-Inf, 1995, Inf)
+molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
+yrlabs <- c('1976-1995', '1996-2014')
+buffx = 0.00025
+buffy = 0.001
+cols =  c('#D53E4F', '#4DAF4A')
+sz_rng = c(2, 11)
+shps = c(24, 25)
+ndivs = 6
+strp_fl = 'lightgrey'
+bar = FALSE
+stat_labs = TRUE
+leg = TRUE
+leg_lab = 'Percent change'
+
+# load required data
+data(delt_dat)
+data(delt_map)
+data(mods)
+
+# response labels 
+reslabs <- list(
+  shrt = c('din', 'nh', 'no23'),
+  expr = c(
+    'DIN', 
+    expression(paste(NH [4] ^ '+')),
+    expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
+  )
+)
+
+# make response labels
+mods <- mutate(mods, reslabs = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr))
+
+##
+# get a bounding box for the stations
+statmeta <- select(delt_dat, Site_Code, Latitude, Longitude) %>% 
+  unique
+
+# bounding box
+lims<- select(statmeta, -Site_Code) %>% 
+  apply(., 2, function(x) range(x, na.rm = TRUE))
+lims <- c(
+  (1 + buffx) * lims[1, 2], 
+  (1 - buffy) * lims[1, 1], 
+  (1 - buffx) * lims[2, 2], 
+  (1 + buffy) * lims[2, 1]
+  )    
+names(lims) <- c('left', 'bottom', 'right', 'top')
+
+# get trends
+trnds <- mutate(mods, 
+  trnd = map(data, function(x){
+    wrtdstrnd_sk(x, mobrks, yrbrks, molabs, yrlabs)
+    })
+  )
+
+cols <- RColorBrewer::brewer.pal(11, 'RdYlGn')
+pal <- colorRampPalette(rev(cols))(1000)
+
+# convert delta map to ggplot compatible object
+delt_map <- fortify(delt_map)
+
+# base delta map
+pbase <- ggplot(delt_map, aes(x = long, y = lat)) + 
+  geom_polygon(aes(group = group, fill = hole), colour = "cornflowerblue") +
+  theme_bw() + 
+  theme(
+    plot.background = element_rect(fill='transparent', 
+      colour = NA),
+    panel.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.key = element_rect(fill = 'transparent', 
+      colour = NA),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.minor = element_blank(), 
+    axis.title.x = element_blank(), 
+    axis.title.y = element_blank(), 
+    legend.position = 'none', 
+    plot.margin = grid::unit(c(3, 3, 12, 3), 'pt'), 
+    axis.text.x = element_text(size = 8), 
+    axis.text.y = element_text(size = 8), 
+    strip.background=element_rect(fill = 'white', colour = 'white'),
+    panel.border = element_rect(colour = "grey")
+  ) +
+  coord_fixed(ratio = 1, xlim = lims[c(1, 3)], ylim = lims[c(2, 4)]) +
+  scale_fill_manual(values=c("cornflowerblue", "white"), guide="none")
+
+yrng <- c(-5, 5)
+
+# unnest trend summary
+toplo <-select(trnds, -data) %>% 
+  unnest %>% 
+  select(Site_Code, resvar, reslabs, cat, ztest, perchg) %>% 
+  left_join(., statmeta, by = 'Site_Code') %>% 
+  data.frame %>% 
+  mutate(
+    rnks = scales::rescale(perchg, c(1, 1000)), 
+    rnks = round(rnks), 
+    cols = pal[rnks],
+    ind = 1:nrow(.), 
+    cattype = ifelse(cat %in% molabs, 'mo', 'yr'), 
+    perchg = pmax(yrng[1], perchg),
+    perchg = pmin(yrng[2], perchg)
+  )
+
+unis <- with(toplo, paste(resvar, cattype)) %>% 
+  unique
+
+sites <- unique(toplo$Site_Code)
+
+plo_ls <- vector('list', length = length(unis))
+names(plo_ls) <- unis
+
+for(uni in unis){
+
+  splits <- strsplit(uni, ' ')[[1]]
+  resvart <- splits[1]
+  cattypet <- splits[2]
+
+  toplo_tmp <- filter(toplo, resvar %in% resvart & cattype %in% cattypet)
+
+  bars <- setNames(lapply(sites, function(i){
+
+    subplo <- filter(toplo_tmp, Site_Code %in% i)
+    ggplot(subplo, aes(x=cat, y = perchg, fill = factor(cat))) +
+      geom_bar(stat = 'identity', colour = 'black') + 
+      scale_fill_manual(values = as.character(subplo$cols)) + 
+      theme_tree() + 
+      xlab(NULL) + 
+      ylab(NULL) + 
+      theme_transparent() +
+      theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
+      scale_y_continuous(limits = yrng)
+
+  }), sites)
+
+  ##
+  # create some plots
+  pall <- pbase
+  for(i in 1:length(bars)){
+
+      nms <- names(bars)[i]
+      dat <- statmeta[statmeta$Site_Code %in% nms,]
+      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, 0.08, 0.08)
+
+  }
+
+  pall <- pall + ggtitle(uni)
+  plo_ls[[uni]] <- pall
+
+}
+
+grid.arrange(plo_ls[[1]], plo_ls[[2]], plo_ls[[3]], plo_ls[[4]], plo_ls[[5]], plo_ls[[6]], ncol = 2)
+
 @

### 2016-11-20 sftrends_manu.Rnw
working on barplot map

--- 2016-11-18 sftrends_manu.Rnw
@@
 trnd_map(mods = mods, yrlabs = NULL, molabs = molabs, mobrks = mobrks, sz_rng = c(2, 8), stat_labs = F, leg_lab = '(b) Seasonal % change')
+@
+
+<<trndmap2, fig.height = 9, fig.width = 6, cache = T, out.width = '0.75\\textwidth', fig.cap = 'test'>>=
+
+res = c('din', 'nh', 'no23')
+mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
+yrbrks <- c(-Inf, 1995, Inf)
+molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
+yrlabs <- c('1976-1995', '1996-2014')
+buffx = 0.00025
+buffy = 0.002
+cols =  c('#D53E4F', '#4DAF4A')
+sz_rng = c(2, 11)
+shps = c(24, 25)
+ndivs = 6
+strp_fl = 'grey90'
+bar = FALSE
+stat_labs = TRUE
+leg = TRUE
+leg_lab = 'Percent change'
+
+# load required data
+data(delt_dat)
+data(delt_map)
+data(mods)
+
+##
+# get a bounding box for the stations
+statmeta <- select(delt_dat, Site_Code, Latitude, Longitude) %>% 
+  unique
+
+# bounding box
+lims<- select(statmeta, -Site_Code) %>% 
+  apply(., 2, function(x) range(x, na.rm = TRUE))
+lims <- c(
+  (1 + buffx) * lims[1, 2], 
+  (1 - buffy) * lims[1, 1], 
+  (1 - buffx) * lims[2, 2], 
+  (1 + buffy) * lims[2, 1]
+  )    
+names(lims) <- c('left', 'bottom', 'right', 'top')
+
+# get trends
+trnds <- mutate(mods,
+  trnd = map(data, function(x){
+    wrtdstrnd_sk(x, mobrks, yrbrks, molabs, yrlabs)
+    })
+  )
+
+# response labels 
+reslabs <- c('din an', 'nh an', 'no23 an', 'din mo', 'nh mo', 'no23 mo')
+reslevs <- c(
+  'DIN, annual', 
+  expression(paste(NH [4] ^ '+', ', annual')),
+  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-', ', annual')),
+  'DIN, seasonal', 
+  expression(paste(NH [4] ^ '+', ', seasonal')),
+  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-', ', seasonal'))
+)
+
+# convert delta map to ggplot compatible object
+delt_map <- fortify(delt_map)
+
+# base delta map
+pbase <- ggplot(delt_map, aes(x = long, y = lat)) + 
+  geom_polygon(aes(group = group, fill = hole), colour = "cornflowerblue") +
+  theme_bw() + 
+  theme(
+    plot.background = element_rect(fill='transparent', 
+      colour = NA),
+    panel.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.key = element_rect(fill = 'transparent', 
+      colour = NA),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.minor = element_blank(), 
+    axis.title.x = element_blank(), 
+    axis.title.y = element_blank(), 
+    legend.position = 'none', 
+    plot.margin = grid::unit(c(3, 3, 3, 3), 'pt'), 
+    axis.text.x = element_text(size = 8), 
+    axis.text.y = element_text(size = 8), 
+    strip.background=element_rect(fill = 'white', colour = 'white'),
+    panel.border = element_rect(colour = "grey")
+  ) +
+  coord_fixed(ratio = 1, xlim = lims[c(1, 3)], ylim = lims[c(2, 4)]) +
+  scale_fill_manual(values=c("cornflowerblue", "white"), guide="none")
+
+yrng <- c(-5, 5)
+
+# color ramp, red for pos, green for neg
+colpos <- RColorBrewer::brewer.pal(9, 'Reds')[3:9]
+colneg <- RColorBrewer::brewer.pal(9, 'Greens')[3:9]
+palneg <-  colorRampPalette(colneg)(1000)
+palpos <-   colorRampPalette(colpos)(1000)
+
+# plot(1:2000, col = c(palneg, palpos), pch = 16, cex = 4)
+# points(1000, 1000)
+
+# unnest data frame, assign colors by rank and sign
+toplo <- select(trnds, -data) %>% 
+  unnest %>% 
+  select(Site_Code, resvar, resvar, cat, ztest, perchg) %>% 
+  left_join(., statmeta, by = 'Site_Code') %>% 
+  data.frame %>% 
+  mutate(
+    perchg = pmax(yrng[1], perchg),
+    perchg = pmin(yrng[2], perchg),
+    negs = sign(perchg),  
+    rnks = scales::rescale(abs(perchg), c(1, 1000)), 
+    rnks = round(rnks), 
+    cols = ifelse(negs == 1, palpos[rnks], palneg[rnks]),
+    ind = 1:nrow(.), 
+    cattype = ifelse(cat %in% molabs, 'mo', 'an')
+  )
+# plot(perchg ~ ind, dat = toplo, col = cols)
+
+unis <- with(toplo, paste(resvar, cattype)) %>% 
+  unique %>% 
+  sort
+
+sites <- unique(toplo$Site_Code)
+
+plo_ls <- vector('list', length = length(unis))
+names(plo_ls) <- unis
+
+for(uni in unis){
+
+  # get variable, agg period
+  splits <- strsplit(uni, ' ')[[1]]
+  resvart <- splits[1]
+  cattypet <- splits[2]
+  toplo_tmp <- filter(toplo, resvar %in% resvart & cattype %in% cattypet)
+
+  # create individual bar plots as ggplot list
+  bars <- setNames(lapply(sites, function(i){
+
+    subplo <- filter(toplo_tmp, Site_Code %in% i)
+    ggplot(subplo, aes(x=cat, y = perchg, fill = factor(cat))) +
+      geom_bar(stat = 'identity', colour = 'black') + 
+      scale_fill_manual(values = as.character(subplo$cols)) + 
+      theme_tree() + 
+      xlab(NULL) + 
+      ylab(NULL) + 
+      theme_transparent() +
+      theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
+      scale_y_continuous(limits = yrng)
+
+  }), sites)
+
+  ##
+  # add the barplots to the base map
+
+  # add site labels if plot is first
+  pall <- pbase
+  if(which(uni == unis) == 1)
+    pall <- pbase + 
+      geom_label_repel(
+        data = statmeta, 
+        aes(x = Longitude, y = Latitude, label = Site_Code), 
+        label.r = unit(0, "lines"),
+        box.padding = unit(1.9, "lines"),
+        point.padding = unit(0, "lines"),
+        force = 2, size = 2,
+        fill = strp_fl
+      )
+
+  for(i in 1:length(bars)){
+
+      nms <- names(bars)[i]
+      dat <- statmeta[statmeta$Site_Code %in% nms,]
+      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, 0.08, 0.1)
+
+  }
+
+  tval <- reslevs[which(uni == reslabs)]
+  pall <- pall + ggtitle(tval)
+  plo_ls[[uni]] <- pall
+
+}
+
+grid.arrange(plo_ls[[1]], plo_ls[[2]], plo_ls[[3]], plo_ls[[4]], plo_ls[[5]], plo_ls[[6]], ncol = 2)
+
 @

### 2016-11-21 sftrends_manu.Rnw
new trends map with bar plots is done

--- 2016-11-18 sftrends_manu.Rnw
@@
 
-<<trndmap, fig.height = 6, fig.width = 9, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record.
-Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
-Points are colored for direction (red increasing, green decreasing) and sized for relative magnitude.
+<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for (a) annual and (b) seasonal aggregations.
+Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
 Station names are shown in the top left panel.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'>>=
-data(mods)
-
+Three values less than 5\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trnds} for actual values.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 7, cache = T, out.width = '\\textwidth'>>=
+
+res = c('din', 'nh', 'no23')
 mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
-yrbrks <- c(1975, 1995, 2000, 2008, 2014)
+yrbrks <- c(-Inf, 1995, Inf)
 molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
-yrlabs <- c('1976-1995', '1996-2000', '2001-2008', '2009-2014')
-
-trnd_map(mods = mods, molabs = NULL, yrlabs = yrlabs, yrbrks = yrbrks, sz_rng = c(2, 8), leg_lab = '(a) Annual % change')
-trnd_map(mods = mods, yrlabs = NULL, molabs = molabs, mobrks = mobrks, sz_rng = c(2, 8), stat_labs = F, leg_lab = '(b) Seasonal % change')
+yrlabs <- c('1976-1995', '1996-2014')
+buffx = 0.00035
+buffy = 0.0022
+cols =  c('#D53E4F', '#4DAF4A')
+sz_rng = c(2, 11)
+shps = c(24, 25)
+ndivs = 6
+strp_fl = 'grey90'
+bar = FALSE
+stat_labs = TRUE
+leg = TRUE
+leg_lab = 'Percent change'
+
+# load required data
+data(delt_dat)
+data(delt_map)
+data(trnds_sk)
+
+##
+# get a bounding box for the stations
+statmeta <- select(delt_dat, Site_Code, Latitude, Longitude) %>% 
+  unique
+
+# bounding box
+lims<- select(statmeta, -Site_Code) %>% 
+  apply(., 2, function(x) range(x, na.rm = TRUE))
+lims <- c(
+  (1 + buffx) * lims[1, 2], 
+  (1 - buffy) * lims[1, 1], 
+  (1 - buffx) * lims[2, 2], 
+  (1 + buffy) * lims[2, 1]
+  )    
+names(lims) <- c('left', 'bottom', 'right', 'top')
+
+# response labels 
+reslabs <- c('din an', 'nh an', 'no23 an', 'din mo', 'nh mo', 'no23 mo')
+reslevs <- c(
+  'DIN', 
+  expression(paste(NH [4] ^ '+')),
+  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-')),
+  'DIN', 
+  expression(paste(NH [4] ^ '+')),
+  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
+)
+
+# convert delta map to ggplot compatible object
+delt_map <- fortify(delt_map)
+
+# base delta map
+alph <- 1
+pbase <- ggplot(delt_map, aes(x = long, y = lat)) + 
+  geom_polygon(aes(group = group, fill = hole), colour = alpha("lightblue", alph)) +
+  theme_bw() + 
+  theme(
+    plot.background = element_rect(fill='transparent', 
+      colour = NA),
+    panel.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.key = element_rect(fill = 'transparent', 
+      colour = NA),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.minor = element_blank(), 
+    axis.title.x = element_blank(), 
+    axis.title.y = element_blank(), 
+    legend.position = 'none', 
+    plot.margin = grid::unit(c(3, 3, 3, 3), 'pt'), 
+    axis.text.x = element_text(size = 8), 
+    axis.text.y = element_text(size = 8), 
+    strip.background=element_rect(fill = 'white', colour = 'white'),
+    panel.border = element_rect(colour = "grey")
+  ) +
+  coord_fixed(ratio = 1, xlim = lims[c(1, 3)], ylim = lims[c(2, 4)]) +
+  scale_fill_manual(values=c(alpha("lightblue", alph), "white"), guide="none")
+
+yrng <- c(-5, 5)
+
+# color ramp, red for pos, green for neg
+colpos <- RColorBrewer::brewer.pal(9, 'Reds')[3:9]
+colneg <- RColorBrewer::brewer.pal(9, 'Greens')[3:9]
+palneg <-  colorRampPalette(colneg)(1000)
+palpos <-   colorRampPalette(colpos)(1000)
+
+# plot(1:2000, col = c(palneg, palpos), pch = 16, cex = 4)
+# points(1000, 1000)
+
+# unnest data frame, assign colors by rank and sign
+toplo <- select(trnds_sk, Site_Code, resvar, resvar, cat, ztest, perchg) %>% 
+  left_join(., statmeta, by = 'Site_Code') %>% 
+  data.frame %>% 
+  mutate(
+    perchg = pmax(yrng[1], perchg),
+    perchg = pmin(yrng[2], perchg),
+    negs = sign(perchg),  
+    rnks = scales::rescale(abs(perchg), c(1, 1000)), 
+    rnks = round(rnks), 
+    cols = ifelse(negs == 1, palpos[rnks], palneg[rnks]),
+    ind = 1:nrow(.), 
+    cattype = ifelse(cat %in% molabs, 'mo', 'an')
+  )
+# plot(perchg ~ ind, dat = toplo, col = cols)
+
+unis <- with(toplo, paste(resvar, cattype)) %>% 
+  unique %>% 
+  sort
+
+sites <- unique(toplo$Site_Code)
+
+plo_ls <- vector('list', length = length(unis))
+names(plo_ls) <- unis
+
+for(uni in unis){
+
+  # get variable, agg period
+  splits <- strsplit(uni, ' ')[[1]]
+  resvart <- splits[1]
+  cattypet <- splits[2]
+  toplo_tmp <- filter(toplo, resvar %in% resvart & cattype %in% cattypet)
+
+  # create individual bar plots as ggplot list
+  bars <- setNames(lapply(sites, function(i){
+
+    subplo <- filter(toplo_tmp, Site_Code %in% i)
+    ggplot(subplo, aes(x=cat, y = perchg, fill = factor(cat))) +
+      geom_bar(stat = 'identity', colour = 'black', size = 0.2) + 
+      scale_fill_manual(values = as.character(subplo$cols)) + 
+      theme_tree() + 
+      xlab(NULL) + 
+      ylab(NULL) + 
+      theme_transparent() +
+      theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
+      scale_y_continuous(limits = yrng)
+
+  }), sites)
+
+  ##
+  # add the barplots to the base map
+
+  # add site labels if plot is first
+  pall <- pbase
+  set.seed(5)
+  if(which(uni == unis) == 1)
+    pall <- pbase + 
+      geom_label_repel(
+        data = statmeta, 
+        aes(x = Longitude, y = Latitude, label = Site_Code), 
+        label.r = unit(0, "lines"),
+        box.padding = unit(2, "lines"),
+        label.padding = unit(0.15, "lines"), 
+        point.padding = unit(0.1, "lines"),
+        force = 2, size = 2,
+        fill = strp_fl
+      )
+
+  # keep bar widths the same for cattypes
+  subwid <- 0.1
+  if(grepl('an$', uni))
+    subwid <- subwid/2
+
+  for(i in 1:length(bars)){
+
+      nms <- names(bars)[i]
+      dat <- statmeta[statmeta$Site_Code %in% nms,]
+      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, subwid, 0.14)
+
+  }
+
+  tval <- reslevs[which(uni == reslabs)]
+  pall <- pall + 
+    geom_text(data = data.frame(x = lims[1], y = lims[4], tval = as.character(tval)), 
+      aes(x = x, y = y, label = tval), parse = TRUE, hjust = 0, vjust = 1)
+  plo_ls[[uni]] <- pall
+
+}
+
+# setup legend
+leg <- ggplot(data.frame(x = 1:10, y = 1:10, z = seq(-5, 5, length = 10)), aes(x, y, colour = z)) + 
+  geom_point() + 
+  scale_colour_gradientn(colours = c(rev(colneg), colpos)) + 
+  guides(colour = guide_colourbar(title = expression('% chg yr' ^-1), barheight = 5, barwidth = 0.5))
+leg <- g_legend(leg)
+
+grid.arrange(
+  arrangeGrob(plo_ls[[1]], plo_ls[[3]], plo_ls[[5]], ncol = 1, top = c('Annual (1976-1995, 1996-2014)')), 
+  arrangeGrob(plo_ls[[2]], plo_ls[[4]], plo_ls[[6]], ncol = 1, top = c('Seasonal (Sp, Su, Fa, Wi)')),
+  leg,
+  ncol = 3, 
+  widths = c(1, 1, 0.4)
+  )
+
 @

### 2016-11-21 sftrends_manu.tex
working on subplot map
working on barplot map
new trends map with bar plots is done

@@
   \vspace{1in} 
-  \\ Version Date:   Fri Nov 18 13:04:31 2016 -0600
+  \\ Version Date:   Sun Nov 20 17:44:15 2016 -0600
 	}
@@
 
-{\centering \includegraphics[width=0.75\textwidth]{figs/trndmap-1} 
-\includegraphics[width=0.75\textwidth]{figs/trndmap-2}
+{\centering \includegraphics[width=\textwidth]{figs/trndmap-1}
 
@@
 
-\caption[Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record]{Percent changes in nitrogen analytes for (a) annual and (b) seasonal (monthly) periods in the record.
-Changes are based on the difference between the ending and starting estimates for the flow-normalized estimates within each period.
-Points are colored for direction (red increasing, green decreasing) and sized for relative magnitude.
+\caption{Percent change per year in nitrogen analytes for (a) annual and (b) seasonal aggregations.
+Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
 Station names are shown in the top left panel.
+Three values less than 5\% change were truncated for visual clarity of the bar plots, see \cref{tab:trnds} for actual values.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}

### 2016-11-21 sftrends_manu.Rnw
methods draft done

@@
 \acrodef{wwtp}[WWTP]{wastewater treatment plant}
+\acrodefplural{wwtp}[WWTPs]{wastewater treatment plants}
 
@@
 The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.
+However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.
 These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.
+The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations \citep[e.g.,][]{Dugdale07,Parker12,Glibert14}.
 
@@
 Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.
-
 \subsection{Analysis method and application}
@@
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
-Specifically, flow-normalized trends in each analyte were summarized as both averages and percent changes from the beginning to end of annual groupings from 1976-1995, 1996-2000, 2001-2008, and 2009-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
-These annual and seasonal groupings were chosen for continuity with additional research efforts \citep[i.e.,][]{Jabusch16}.
-Trends within each seasonal grouping were based on annual mean estimates for the first and last three years of flow-normalized estimates to reduce the effects of abnormal years or missing data.
-Annual trends were based only on the first and last years in the time period due to limited number of years in some groupings \citepalias{BeckIP}.
+Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
+These annual and seasonal groupings were chosen for continuity with similar comparisons reported in \citet{Jabusch16}.
+
+Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
+This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82}.
+Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
+The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.
+Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.
+The results were interpreted relative to those in \citet{Jabusch16} that were based on observed data to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
+The analysis was also similar to that in \citet{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression.
+Functions in the EnvStats package in R were used for the seasonal Kendall tests \citep{Millard13}.
 
@@
 
-Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more detailed understanding of trends for further evaluation.
-The stations werechosen to address ecological and management-based questions that have relevance outside of the region, having importance for our understanding of estuarine processes that influence eutrophication trends on decadal time scales.
+Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
+The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends on decadal time scales.
 The selected case studies focused on \begin{inparaenum}[1\upshape)]
@@
 \end{inparaenum}
-Each case study is built around hypotheses that results from \ac{wrtds} models are expected to support, both as a general description and for additional testing with alternative methods.
+Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
 
@@
 
-Multiple biological and physical factors influence nutrient concentrations at different temporal scales \citep[][, references therein]{Cloern10}.
-As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex.
-A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, is expected to reveal \begin{inparaenum}[1\upshape)]
+Multiple biological and physical factors influence nutrient concentrations at different temporal scales \citep[][references therein]{Cloern10}.
+As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex as compared to more conventional methods for time series analysis.
+A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, was expected to reveal \begin{inparaenum}[1\upshape)]
 \item an annual trend independent of a seasonal trend, and 
@@
 
-Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) different nitrogen species will have different changes depending on change in load outputs.
-See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}, but this is about Tracy
+Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.
+As noted in \citep{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., \citealt{Cornwell14,Novick15}).
+Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}.
+Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
+\item a shift in load contributions before/after upgrade,
+\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
+\item different nitrogen species will have different changes depending on change in load outputs. 
+\end{inparaenum}
+The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
+%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
+
 \subsubsection{Effects of biological invasions}
 
-Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.
+The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \citep{Cohen98}.
+The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \citep{Carlton90,Nichols90,Werner93,Crauder16}.
+In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
+Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \citep{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}.
+Abundance of clams is linked to inflow rates such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.
+
+We hypothesized that \ac{wrtds} models applied to water quality obsevations in the upper estuary would emphasize \begin{inparaenum}[1\upshape)]
+\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
+\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
+\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
+\end{inparaenum}
+The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
 
@@
 \subsection{Trends}
+
+Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in \citet{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
+\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from \citet{Jabusch16}.
+Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).
 
@@
 This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
-See suggestions in \citet{Alpine92} regarding flow/grazer relationships in the Bay.
+See suggestions in \citet{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
 

### 2016-11-21 sftrends_manu.tex
methods draft done

@@
   \vspace{1in} 
-  \\ Version Date:   Sun Nov 20 17:44:15 2016 -0600
+  \\ Version Date:   Mon Nov 21 10:45:35 2016 -0600
 	}

### 2016-12-08 README.Rmd
added trnds_sk data object
redid readme, deleted unused data files

--- 2016-06-16 README.Rmd
@@
 
-Files marked with * were created in `sf_trends.RProj`, all others created in `R\dat_proc.R`
+Files were created in `sf_trends.RProj`
 
-`chld7.RData` WRTDS model for chl at D7
+`clams.RData` clam data for D7 from Crauder 2016, includes sample date, water year (`yr` starting in October), biomass in g/m2, clams per sample (`clams_smp`) in no/m2, depth (m), growth rate (`gr`) in m3/m2/d, mean size (mm), number of grabs (`no_grabs`), recruitment per unit area (`recruit_area`) in recruit/0.05 m2, and species as corbicula or potamocorbula.
 
-`clams.RData`* clam data for D7 from Crauder 2016, includes sample date, water year (`yr` starting in October), biomass in g/m2, clams per sample (`clams_smp`) in no/m2, depth (m), growth rate (`gr`) in m3/m2/d, mean size (mm), number of grabs (`no_grabs`), recruitment per unit area (`recruit_area`) in recruit/0.05 m2, and species as corbicula or potamocorbula.
+`delt_dat.RData` Processed wq time series data `dwr_wq.RData`, includes all nitrogen analytes and current/active stations in the delta, also includes matched and smoothed flow records from `flocor.RData` results
 
-`delt_dat.RData`* Processed wq time series data `dwr_wq.RData`, includes all nitrogen analytes and current/active stations in the delta, also includes matched and smoothed flow records from `flocor.RData` results
+`delt_map.RData` SpatialPolygon object of generalized delta map
 
-`diat_dat.RData`* additional WRTDS models for D7, including chlorphyll and sio2
+`flow_dat.RData` time series of daily flow estimates for the delta, input stations from Novick et al (Fig 2) were used
 
-`flow_dat.RData`* time series of daily flow estimates for the delta, input stations from Novick et al (Fig 2) were used
+`h1dat.RData` mean models to support first hypothesis/case study in manuscript, wrtds mean models at c10
 
-`mods_nolag.RData`* complete dataset for wrtds, including model results.
-This is the same file as `mods_lag.RData` except the matched flow variables are not lagged.
-Nested, do not load without loading tidyr.
+`h2dat.RData` mean models to support second hypothesis/case study in manuscript, wrtds mean models for no23, nh at p8
 
-`dinc10.RData` WRTDS model for DIN at C10
+`h3dat.RData` mean models to support third hypothesis/case study in manuscript, wrtds mean models for sio2, din, chla at c10, d7
+
+`mods.RData` complete dataset for wrtds, including model results.
+
+`stock_conc.RData` summarized effluent data from Stockton wwtp, mg/L
+
+`stock_load.RData` nutrient load data from Stockton, kg/day
+
+`trnds_ave.RData` trend summary of averages by annual, monthly categories
+
+`trnds_chg.RData` trend summary of percent changes by annual, monthly categories
+
+`trnds_sk.RData` trend summary using seasonal kendall on flow-normalized results

### 2016-12-09 sftrends_manu.Rnw
fig 5 now only tot nitrogen

@@
 
-<<stock, fig.height = 5, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
 Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
@@
 ylab1 <- expression(paste("Total nitrogen (mg ", L^-1, ")"))
-ylab2 <- '% of total nitrogen'
 xrng <- range(c(toplo1$date, toplo2$date))
@@
 
-p1 <- ggplot(toplo1) + 
-  geom_line(colour = lncol, aes(x = date, y = TN)) + 
+p1 <- ggplot(toplo2) + 
+  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack') + 
   scale_x_date(expand = c(0, 0), limits = xrng) + 
-  scale_y_continuous(ylab1, expand = c(0, 0), limits = yrng) + 
+  scale_fill_brewer(palette = 'Blues', labels = leglabs) + 
   geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
@@
     ) +
-  ptheme
-
-p2 <- ggplot(toplo2) + 
-  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'fill') + 
-  scale_x_date(expand = c(0, 0), limits = xrng) + 
-  scale_fill_brewer(palette = 'Blues', labels = leglabs) + 
-  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
-  scale_y_continuous(ylab2, expand = c(0, 0)) + 
+  scale_y_continuous(ylab1, expand = c(0, 0)) + 
   ptheme + 
@@
 
-# align widths of plots in first column, first two
-pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p2))
-maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3])
-
-# Set the widths
-pA$widths[2:3] <- maxWidth
-pB$widths[2:3] <- maxWidth
-
-grid.arrange(pA, pB, ncol = 1, heights = c(0.8, 1))
+p1
 @

### 2016-12-11 sftrends_manu.Rnw
added seas by year to trend map

@@
 library(ggtree)
+library(forcats)
 @
@@
 
-<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for (a) annual and (b) seasonal aggregations.
+<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2014.
 Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
 Station names are shown in the top left panel.
-Three values less than 5\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trnds} for actual values.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 7, cache = T, out.width = '\\textwidth'>>=
-
-res = c('din', 'nh', 'no23')
-mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
-yrbrks <- c(-Inf, 1995, Inf)
-molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
-yrlabs <- c('1976-1995', '1996-2014')
-buffx = 0.00035
-buffy = 0.0022
-cols =  c('#D53E4F', '#4DAF4A')
-sz_rng = c(2, 11)
-shps = c(24, 25)
-ndivs = 6
-strp_fl = 'grey90'
-bar = FALSE
-stat_labs = TRUE
-leg = TRUE
-leg_lab = 'Percent change'
+Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trnds} for actual values.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 8, cache = T, out.width = '\\textwidth'>>=
 
@@
 data(delt_map)
-data(trnds_sk)
+data(trnds_seasyr)
 
@@
 # bounding box
+buffx <- 0.00035
+buffy <- 0.0022
 lims<- select(statmeta, -Site_Code) %>% 
@@
 
+# need to shift locations of some stations
+statmeta[statmeta$Site_Code %in% 'D26', 'Latitude'] <- 38.12
+statmeta[statmeta$Site_Code %in% 'D26', 'Longitude'] <--121.55
+statmeta[statmeta$Site_Code %in% 'D7', 'Latitude'] <- 38.13
+statmeta[statmeta$Site_Code %in% 'D7', 'Longitude'] <--122.015 
+statmeta[statmeta$Site_Code %in% 'C10', 'Latitude'] <-37.68 
+statmeta[statmeta$Site_Code %in% 'C10', 'Longitude'] <--121.28
+
 # response labels 
-reslabs <- c('din an', 'nh an', 'no23 an', 'din mo', 'nh mo', 'no23 mo')
-reslevs <- c(
-  'DIN', 
-  expression(paste(NH [4] ^ '+')),
-  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-')),
+reslabs <- c('din an', 'nh an', 'no23 an', 'din mobef', 'nh mobef', 'no23 mobef', 'din moaft', 'nh moaft', 'no23 moaft')
+reslevs <- rep(c(
   'DIN', 
@@
   expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
-)
+  ), 3)
 
@@
 
-yrng <- c(-5, 5)
+# bounds on limits
+yrng <- c(-9, 9)
 
@@
 # unnest data frame, assign colors by rank and sign
-toplo <- select(trnds_sk, Site_Code, resvar, resvar, cat, ztest, perchg) %>% 
+toplo <- select(trnds_seasyr, Site_Code, resvar, ann, cat, norm) %>% 
+  rename(perchg = norm) %>% 
   left_join(., statmeta, by = 'Site_Code') %>% 
@@
     cols = ifelse(negs == 1, palpos[rnks], palneg[rnks]),
-    ind = 1:nrow(.), 
-    cattype = ifelse(cat %in% molabs, 'mo', 'an')
-  )
-# plot(perchg ~ ind, dat = toplo, col = cols)
-
+    ind = 1:nrow(.)
+    ) %>%
+  unite('cattype', cat, ann, sep = '', remove = F) %>%
+  mutate(
+    cattype = factor(cattype),
+    cattype = fct_collapse(cattype, 
+      an = c('1976-1995bef', '1996-2014aft'),
+      mobef = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'bef'), 
+      moaft = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'aft') 
+    )
+  )
+
+# unite response varibles and summary period combos
 unis <- with(toplo, paste(resvar, cattype)) %>% 
@@
         force = 2, size = 2,
-        fill = strp_fl
+        fill = 'grey90'
       )
@@
       dat <- statmeta[statmeta$Site_Code %in% nms,]
-      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, subwid, 0.14)
+      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, subwid, 0.148) # change bar heights here
 
@@
 # setup legend
-leg <- ggplot(data.frame(x = 1:10, y = 1:10, z = seq(-5, 5, length = 10)), aes(x, y, colour = z)) + 
+rngs <- seq(-9, 7, by = 3)
+leg <- ggplot(data.frame(x = seq_along(rngs), y = seq_along(rngs), z = rngs), aes(x, y, colour = z)) + 
   geom_point() + 
   scale_colour_gradientn(colours = c(rev(colneg), colpos)) + 
-  guides(colour = guide_colourbar(title = expression('% chg yr' ^-1), barheight = 5, barwidth = 0.5))
+  guides(colour = guide_colourbar(title = expression('% chg yr' ^-1), barheight = 0.5, barwidth = 8)) + 
+  theme(legend.position = 'top')
 leg <- g_legend(leg)
@@
 grid.arrange(
-  arrangeGrob(plo_ls[[1]], plo_ls[[3]], plo_ls[[5]], ncol = 1, top = c('Annual (1976-1995, 1996-2014)')), 
-  arrangeGrob(plo_ls[[2]], plo_ls[[4]], plo_ls[[6]], ncol = 1, top = c('Seasonal (Sp, Su, Fa, Wi)')),
   leg,
-  ncol = 3, 
-  widths = c(1, 1, 0.4)
+  arrangeGrob(
+    arrangeGrob (plo_ls[[1]], plo_ls[[4]], plo_ls[[7]], ncol = 1, top = c('(a) 1976-1995, 1996-2014')), 
+    arrangeGrob(plo_ls[[3]], plo_ls[[6]], plo_ls[[9]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1976-1995)')),
+    arrangeGrob(plo_ls[[2]], plo_ls[[5]], plo_ls[[8]], ncol = 1, top = c('(c) Sp, Su, Fa, Wi (1996-2014)')),
+    ncol = 3
+    ),
+  ncol = 1, 
+  heights = c(0.1, 1)
   )
@@
 
-% overall trends
+% overall trends, years
 <<trnds>>=

### 2016-12-11 sftrends_manu.tex
redid readme, deleted unused data files
fig 5 now only tot nitrogen
added seas by year to trend map

@@
   \vspace{1in} 
-  \\ Version Date:   Mon Nov 21 10:45:35 2016 -0600
+  \\ Version Date:   Fri Dec 9 16:31:55 2016 -0600
 	}
@@
 
-\caption{Percent change per year in nitrogen analytes for (a) annual and (b) seasonal aggregations.
+\caption{Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2014.
 Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
 Station names are shown in the top left panel.
-Three values less than 5\% change were truncated for visual clarity of the bar plots, see \cref{tab:trnds} for actual values.
+Values less than -9\% change were truncated for visual clarity of the bar plots, see \cref{tab:trnds} for actual values.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}
@@
 
-% overall trends
+% overall trends, years
 %latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual", "Seasonal"), n.cgroup = c(2,         4), rowname = stats, label = "tab:trnds", insert.bottom = foot.val)%

### 2016-12-12 README.Rmd
broke out trend tables into different annual seasonal groups

@@
 
+`trnds_seasyr.RData` collection of three data frames for seasonal kendall on observed, fit, norm data, different annual and seasonal periods
+
 `stock_conc.RData` summarized effluent data from Stockton wwtp, mg/L
@@
 `stock_load.RData` nutrient load data from Stockton, kg/day
-
-`trnds_ave.RData` trend summary of averages by annual, monthly categories
-
-`trnds_chg.RData` trend summary of percent changes by annual, monthly categories
-
-`trnds_sk.RData` trend summary using seasonal kendall on flow-normalized results

### 2016-12-12 sftrends_manu.Rnw
broke out trend tables into different annual seasonal groups

@@
 
-<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2014.
+<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.
 Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
@@
 data(trnds_seasyr)
+
+# combine all
+trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
+  select(Site_Code, resvar, cat, perchg, ann, trndvar) %>% 
+  spread(trndvar, perchg) %>% 
+  mutate(Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')))
 
@@
     cattype = fct_collapse(cattype, 
-      an = c('1976-1995bef', '1996-2014aft'),
+      an = c('1976-1995bef', '1996-2013aft'),
       mobef = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'bef'), 
@@
   if(grepl('an$', uni))
-    subwid <- subwid/2
+    subwid <- 0.06
 
@@
   arrangeGrob(
-    arrangeGrob (plo_ls[[1]], plo_ls[[4]], plo_ls[[7]], ncol = 1, top = c('(a) 1976-1995, 1996-2014')), 
+    arrangeGrob (plo_ls[[1]], plo_ls[[4]], plo_ls[[7]], ncol = 1, top = c('(a) 1976-1995, 1996-2013')), 
     arrangeGrob(plo_ls[[3]], plo_ls[[6]], plo_ls[[9]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1976-1995)')),
-    arrangeGrob(plo_ls[[2]], plo_ls[[5]], plo_ls[[8]], ncol = 1, top = c('(c) Sp, Su, Fa, Wi (1996-2014)')),
+    arrangeGrob(plo_ls[[2]], plo_ls[[5]], plo_ls[[8]], ncol = 1, top = c('(c) Sp, Su, Fa, Wi (1996-2013)')),
     ncol = 3
@@
 % overall trends, years
-<<trnds>>=
-data(trnds_sk)
+<<trndsann>>=
+data(trnds_seasyr)
 
 # percent changes
-tab <- trnds_sk %>% 
+tab <- trnds_nrm %>% 
+  filter(cat %in% c('1976-1995', '1996-2013')) %>% 
   select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
@@
   arrange(resvar, Site_Code) %>% 
-  select(Site_Code, resvar, `1976-1995`, `1996-2014`, Spring, Summer, Fall, Winter)
+  select(Site_Code, resvar, `1976-1995`, `1996-2013`)
 
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \\cref{fig:trndmap} for a summary of spatial trends.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+See \\cref{fig:trndmap} for a summary of spatial trends.'
 foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
@@
   n.rgroup = c(9, 9, 9),
-  cgroup = c('Annual', 'Seasonal'),
-  n.cgroup = c(2, 4),
+  cgroup = c('Annual'),
+  n.cgroup = c(2),
   rowname = stats,
-  label = 'tab:trnds',
+  label = 'tab:trndsann',
+  insert.bottom = foot.val
+  )
+
+@
+
+% overall trends, seasons, first annual periodyears
+<<trndsbef>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(
+    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
+    ann %in% 'bef'
+    ) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+See \\cref{fig:trndmap} for a summary of spatial trends.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = c(9, 9, 9),
+  cgroup = c('Seasonal, 1976-1995'),
+  n.cgroup = c(4),
+  rowname = stats,
+  label = 'tab:trndsbef',
+  insert.bottom = foot.val
+  )
+
+@
+
+% overall trends, seasons, first annual periodyears
+<<trndsaft>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(
+    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
+    ann %in% 'aft'
+    ) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+See \\cref{fig:trndmap} for a summary of spatial trends.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = c(9, 9, 9),
+  cgroup = c('Seasonal, 1996-2013'),
+  n.cgroup = c(4),
+  rowname = stats,
+  label = 'tab:trndsaft',
   insert.bottom = foot.val

### 2016-12-12 sftrends_manu.tex
broke out trend tables into different annual seasonal groups

@@
   \vspace{1in} 
-  \\ Version Date:   Fri Dec 9 16:31:55 2016 -0600
+  \\ Version Date:   Sun Dec 11 11:23:55 2016 -0600
 	}
@@
 
-\caption{Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2014.
+\caption{Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.
 Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
@@
 % overall trends, years
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual", "Seasonal"), n.cgroup = c(2,         4), rowname = stats, label = "tab:trnds", insert.bottom = foot.val)%
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann", insert.bottom = foot.val)%
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and different time periods.
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+See \cref{fig:trndmap} for a summary of spatial trends.\label{tab:trndsann}}
+\begin{center}
+\begin{tabular}{lll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{2}{c}{\bfseries Annual}\tabularnewline
+\cline{2-3}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1995}&\multicolumn{1}{c}{1996-2013}\tabularnewline
+\hline
+{\bfseries DIN}&&\tabularnewline
+~~C10&1.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.4 \footnotesize{(-3.1)**}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.5 \footnotesize{(-0.1)**}\tabularnewline
+~~D19&0.4 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-1.2)**}\tabularnewline
+~~D28&0.4 \footnotesize{(\textit{\textbf{0.1}})**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
+~~D6&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.5 \footnotesize{(-0.3)**}\tabularnewline
+~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}\tabularnewline
+~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}\tabularnewline
+\hline
+{\bfseries NH$_{4}^{+}$}&&\tabularnewline
+~~C10&0.1 \footnotesize{(-3.4)**}&0 \footnotesize{(-5.2)**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{3.7}})**}&0.3 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
+~~D19&0 \footnotesize{(\textit{\textbf{0.4}})**}&0 \footnotesize{(-1.7)**}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{2.2}})**}&0.1 \footnotesize{(-1.5)**}\tabularnewline
+~~D28&0 \footnotesize{(-1.1)**}&0 \footnotesize{(-1.4)**}\tabularnewline
+~~D4&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{2.4}})**}&0.1 \footnotesize{(-0.5)**}\tabularnewline
+~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}\tabularnewline
+~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}\tabularnewline
+\hline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{1.4}})**}&1.4 \footnotesize{(-3)**}\tabularnewline
+~~C3&0.1 \footnotesize{(-0.1)**}&0.2 \footnotesize{(\textit{\textbf{0.7}})**}\tabularnewline
+~~D19&0.4 \footnotesize{(-0.1)**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
+~~D26&0.3 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1.1)**}\tabularnewline
+~~D28&0.4 \footnotesize{(-0.2)**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(-0.4)**}\tabularnewline
+~~D6&0.3 \footnotesize{(\textit{\textbf{1.3}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
+~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}\tabularnewline
+~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
+
+% overall trends, seasons, first annual periodyears
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef", insert.bottom = foot.val)%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
@@
 See \cref{fig:trndmap} for a summary of spatial trends.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trnds}}
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsbef}}
 \begin{center}
-\begin{tabular}{lllcllll}
+\begin{tabular}{lllll}
 \hline\hline
-\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{2}{c}{\bfseries Annual}&\multicolumn{1}{c}{\bfseries }&\multicolumn{4}{c}{\bfseries Seasonal}\tabularnewline
-\cline{2-3} \cline{5-8}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1995}&\multicolumn{1}{c}{1996-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1976-1995}\tabularnewline
+\cline{2-5}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
 \hline
-{\bfseries DIN}&&&&&&&\tabularnewline
-~~C10&1.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.4 \footnotesize{(-3.1)**}&&1.1 \footnotesize{(-0.1)}&1.2 \footnotesize{(\textit{\textbf{0.1}})}&1.4 \footnotesize{(\textit{\textbf{0.6}})**}&1.7 \footnotesize{(\textit{\textbf{0.1}})}\tabularnewline
-~~C3&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.5 \footnotesize{(-0.1)**}&&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.3 \footnotesize{(\textit{\textbf{1.9}})**}&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
-~~D19&0.4 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(-1.9)**}&&0.5 \footnotesize{(-0.4)**}&0.2 \footnotesize{(\textit{\textbf{0}})}&0.3 \footnotesize{(\textit{\textbf{0.2}})}&0.7 \footnotesize{(-0.5)**}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-1.2)**}&&0.5 \footnotesize{(\textit{\textbf{0}})}&0.3 \footnotesize{(\textit{\textbf{0.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.5}})**}&0.6 \footnotesize{(-0.2)**}\tabularnewline
-~~D28&0.4 \footnotesize{(\textit{\textbf{0.1}})**}&0.4 \footnotesize{(-3.1)**}&&0.5 \footnotesize{(-0.5)**}&0.2 \footnotesize{(-0.9)**}&0.3 \footnotesize{(-1.1)**}&0.8 \footnotesize{(-0.8)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(-0.3)**}&&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.3 \footnotesize{(\textit{\textbf{0.8}})**}&0.4 \footnotesize{(\textit{\textbf{0.5}})**}&0.5 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
-~~D6&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.5 \footnotesize{(-0.3)**}&&0.5 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.6}})**}&0.4 \footnotesize{(\textit{\textbf{0.9}})**}&0.5 \footnotesize{(\textit{\textbf{0.2}})*}\tabularnewline
-~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}&&0.5 \footnotesize{(-0.1)}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.4 \footnotesize{(\textit{\textbf{0.6}})**}&0.6 \footnotesize{(-0.2)}\tabularnewline
-~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}&&1.4 \footnotesize{(\textit{\textbf{0.7}})**}&1 \footnotesize{(\textit{\textbf{1.4}})**}&1.6 \footnotesize{(\textit{\textbf{2}})**}&2.2 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
+{\bfseries DIN}&&&&\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{1.1}})**}&1.2 \footnotesize{(\textit{\textbf{0.3}})}&1.3 \footnotesize{(\textit{\textbf{0.5}})**}&1.7 \footnotesize{(\textit{\textbf{1.2}})**}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.3 \footnotesize{(\textit{\textbf{2.3}})**}&0.4 \footnotesize{(\textit{\textbf{2.4}})**}&0.4 \footnotesize{(\textit{\textbf{1.9}})**}\tabularnewline
+~~D19&0.5 \footnotesize{(\textit{\textbf{0.3}})}&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.7 \footnotesize{(-0.2)}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(\textit{\textbf{0.4}})*}&0.4 \footnotesize{(\textit{\textbf{1}})**}&0.6 \footnotesize{(\textit{\textbf{0.3}})}\tabularnewline
+~~D28&0.5 \footnotesize{(\textit{\textbf{0.8}})*}&0.2 \footnotesize{(\textit{\textbf{0.3}})}&0.3 \footnotesize{(\textit{\textbf{0.5}})*}&0.8 \footnotesize{(-0.3)}\tabularnewline
+~~D4&0.4 \footnotesize{(\textit{\textbf{0.2}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.5 \footnotesize{(-0.5)}\tabularnewline
+~~D6&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{4.6}})**}&0.4 \footnotesize{(\textit{\textbf{1.4}})**}&0.5 \footnotesize{(-0.7)*}\tabularnewline
+~~D7&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.6 \footnotesize{(-2.4)**}\tabularnewline
+~~P8&1.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.9 \footnotesize{(\textit{\textbf{2.4}})**}&1.3 \footnotesize{(\textit{\textbf{3.1}})**}&1.9 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
 \hline
-{\bfseries NH$_{4}^{+}$}&&&&&&&\tabularnewline
-~~C10&0.1 \footnotesize{(-3.4)**}&0 \footnotesize{(-5.2)**}&&0 \footnotesize{(-4.4)**}&0 \footnotesize{(-6.9)**}&0.1 \footnotesize{(-8.7)**}&0.2 \footnotesize{(-4.3)**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{3.7}})**}&0.3 \footnotesize{(\textit{\textbf{0}})}&&0.2 \footnotesize{(\textit{\textbf{2.9}})**}&0.2 \footnotesize{(\textit{\textbf{2.8}})**}&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.2 \footnotesize{(\textit{\textbf{2}})**}\tabularnewline
-~~D19&0 \footnotesize{(\textit{\textbf{0.4}})**}&0 \footnotesize{(-1.7)**}&&0.1 \footnotesize{(-0.7)**}&0 \footnotesize{(-0.5)**}&0 \footnotesize{(\textit{\textbf{0}})}&0.1 \footnotesize{(-0.6)**}\tabularnewline
-~~D26&0.1 \footnotesize{(\textit{\textbf{2.2}})**}&0.1 \footnotesize{(-1.5)**}&&0.1 \footnotesize{(\textit{\textbf{0}})}&0.1 \footnotesize{(\textit{\textbf{1}})**}&0.1 \footnotesize{(\textit{\textbf{0.5}})**}&0.1 \footnotesize{(-0.2)*}\tabularnewline
-~~D28&0 \footnotesize{(-1.1)**}&0 \footnotesize{(-1.4)**}&&0 \footnotesize{(-1.2)**}&0 \footnotesize{(-0.4)*}&0 \footnotesize{(-0.9)**}&0.1 \footnotesize{(-1.4)**}\tabularnewline
-~~D4&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0}})}&&0.1 \footnotesize{(\textit{\textbf{1.7}})**}&0 \footnotesize{(\textit{\textbf{1.3}})**}&0.1 \footnotesize{(\textit{\textbf{1.2}})**}&0.1 \footnotesize{(\textit{\textbf{1.4}})**}\tabularnewline
-~~D6&0.1 \footnotesize{(\textit{\textbf{2.4}})**}&0.1 \footnotesize{(-0.5)**}&&0.1 \footnotesize{(\textit{\textbf{0.6}})**}&0.1 \footnotesize{(\textit{\textbf{1.4}})**}&0.1 \footnotesize{(\textit{\textbf{1.1}})**}&0.1 \footnotesize{(\textit{\textbf{0.7}})**}\tabularnewline
-~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}&&0.1 \footnotesize{(\textit{\textbf{0.5}})*}&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0.4}})**}&0.1 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
-~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}&&0.2 \footnotesize{(-0.5)}&0.1 \footnotesize{(\textit{\textbf{1.3}})**}&0.2 \footnotesize{(\textit{\textbf{1.9}})*}&0.6 \footnotesize{(\textit{\textbf{1}})}\tabularnewline
+{\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
+~~C10&0.1 \footnotesize{(-2.3)**}&0 \footnotesize{(-6.8)**}&0.1 \footnotesize{(-7.1)**}&0.3 \footnotesize{(-1.5)**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.2 \footnotesize{(\textit{\textbf{4}})**}&0.3 \footnotesize{(\textit{\textbf{3.8}})**}&0.2 \footnotesize{(\textit{\textbf{2.9}})**}\tabularnewline
+~~D19&0.1 \footnotesize{(\textit{\textbf{0.4}})*}&0 \footnotesize{(-1.7)**}&0 \footnotesize{(\textit{\textbf{1.2}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{1.4}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.1 \footnotesize{(\textit{\textbf{2.3}})**}\tabularnewline
+~~D28&0.1 \footnotesize{(-0.5)}&0 \footnotesize{(-3.7)**}&0 \footnotesize{(-1.6)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
+~~D4&0.1 \footnotesize{(\textit{\textbf{1.7}})**}&0 \footnotesize{(\textit{\textbf{1}})**}&0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textit{\textbf{2}})**}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{2.9}})**}&0.1 \footnotesize{(\textit{\textbf{2.8}})**}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
+~~D7&0.1 \footnotesize{(\textit{\textbf{3.3}})**}&0 \footnotesize{(\textit{\textbf{2}})**}&0.1 \footnotesize{(-2.8)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
+~~P8&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.1 \footnotesize{(\textit{\textbf{1.8}})**}&0.2 \footnotesize{(\textit{\textbf{7}})**}&0.6 \footnotesize{(\textit{\textbf{7}})**}\tabularnewline
 \hline
-{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&&&&\tabularnewline
-~~C10&1.2 \footnotesize{(\textit{\textbf{1.4}})**}&1.4 \footnotesize{(-3)**}&&1.1 \footnotesize{(\textit{\textbf{0.1}})}&1.2 \footnotesize{(\textit{\textbf{0.2}})}&1.3 \footnotesize{(\textit{\textbf{1.2}})**}&1.5 \footnotesize{(\textit{\textbf{0.8}})**}\tabularnewline
-~~C3&0.1 \footnotesize{(-0.1)**}&0.2 \footnotesize{(\textit{\textbf{0.7}})**}&&0.2 \footnotesize{(\textit{\textbf{0.5}})**}&0.1 \footnotesize{(\textit{\textbf{0.7}})**}&0.1 \footnotesize{(\textit{\textbf{0.9}})**}&0.2 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
-~~D19&0.4 \footnotesize{(-0.1)**}&0.4 \footnotesize{(-1.9)**}&&0.4 \footnotesize{(-0.3)*}&0.2 \footnotesize{(-0.3)*}&0.3 \footnotesize{(-0.1)}&0.6 \footnotesize{(-0.6)**}\tabularnewline
-~~D26&0.3 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1.1)**}&&0.4 \footnotesize{(\textit{\textbf{0}})}&0.2 \footnotesize{(\textit{\textbf{0.2}})*}&0.3 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-0.1)}\tabularnewline
-~~D28&0.4 \footnotesize{(-0.2)**}&0.4 \footnotesize{(-3.1)**}&&0.5 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-1.5)**}&0.3 \footnotesize{(-1.4)**}&0.7 \footnotesize{(-0.8)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(-0.4)**}&&0.3 \footnotesize{(\textit{\textbf{0.2}})*}&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(\textit{\textbf{0.2}})*}&0.4 \footnotesize{(\textit{\textbf{0.1}})}\tabularnewline
-~~D6&0.3 \footnotesize{(\textit{\textbf{1.3}})**}&0.4 \footnotesize{(-0.3)**}&&0.4 \footnotesize{(\textit{\textbf{0.2}})*}&0.4 \footnotesize{(\textit{\textbf{1.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(\textit{\textbf{0.1}})}\tabularnewline
-~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}&&0.4 \footnotesize{(-0.3)**}&0.4 \footnotesize{(\textit{\textbf{1.2}})**}&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-0.5)**}\tabularnewline
-~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}&&1.2 \footnotesize{(\textit{\textbf{1}})**}&1 \footnotesize{(\textit{\textbf{1.2}})**}&1.4 \footnotesize{(\textit{\textbf{1.7}})**}&1.6 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
+~~C10&1.1 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(\textit{\textbf{0.6}})**}&1.2 \footnotesize{(\textit{\textbf{1.3}})**}&1.5 \footnotesize{(\textit{\textbf{1.8}})**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{0.7}})**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+~~D19&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.2 \footnotesize{(-0.3)}&0.3 \footnotesize{(\textit{\textbf{0.3}})}&0.6 \footnotesize{(-0.9)*}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.6}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.3}})*}&0.5 \footnotesize{(-0.3)}\tabularnewline
+~~D28&0.5 \footnotesize{(\textit{\textbf{0.7}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.2}})}&0.7 \footnotesize{(-1)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.1}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.4 \footnotesize{(-0.8)*}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.1}})**}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.4 \footnotesize{(-1)**}\tabularnewline
+~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textit{\textbf{3.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.4 \footnotesize{(-3.6)**}\tabularnewline
+~~P8&1.2 \footnotesize{(\textit{\textbf{2}})**}&0.9 \footnotesize{(\textit{\textbf{2.3}})**}&1.1 \footnotesize{(\textit{\textbf{2}})**}&1.4 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
+
+% overall trends, seasons, first annual periodyears
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft", insert.bottom = foot.val)%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+See \cref{fig:trndmap} for a summary of spatial trends.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsaft}}
+\begin{center}
+\begin{tabular}{lllll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1996-2013}\tabularnewline
+\cline{2-5}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
+\hline
+{\bfseries DIN}&&&&\tabularnewline
+~~C10&1.1 \footnotesize{(-4.1)**}&1.3 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-2)**}&1.7 \footnotesize{(-3.4)**}\tabularnewline
+~~C3&0.5 \footnotesize{(\textit{\textbf{0.5}})}&0.4 \footnotesize{(\textit{\textbf{0.1}})}&0.6 \footnotesize{(-0.2)}&0.5 \footnotesize{(-0.6)**}\tabularnewline
+~~D19&0.5 \footnotesize{(-2.8)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.6)**}&0.7 \footnotesize{(-2.3)**}\tabularnewline
+~~D26&0.5 \footnotesize{(-1.9)**}&0.3 \footnotesize{(-1.7)**}&0.4 \footnotesize{(-1)**}&0.6 \footnotesize{(-0.8)**}\tabularnewline
+~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-4.9)**}&0.2 \footnotesize{(-4.9)**}&0.7 \footnotesize{(-2.1)**}\tabularnewline
+~~D4&0.4 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.9)**}&0.5 \footnotesize{(\textit{\textbf{0.6}})**}\tabularnewline
+~~D6&0.5 \footnotesize{(-0.2)*}&0.5 \footnotesize{(-1)**}&0.5 \footnotesize{(-0.3)*}&0.5 \footnotesize{(-0.1)}\tabularnewline
+~~D7&0.5 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-1.3)**}&0.4 \footnotesize{(-0.4)**}&0.6 \footnotesize{(-0.2)}\tabularnewline
+~~P8&1.5 \footnotesize{(-1.9)**}&1.2 \footnotesize{(-3.5)**}&1.8 \footnotesize{(-2.4)**}&2.7 \footnotesize{(-2.2)**}\tabularnewline
+\hline
+{\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
+~~C10&0 \footnotesize{(-4.2)**}&0 \footnotesize{(-6.1)**}&0 \footnotesize{(-8.5)**}&0.1 \footnotesize{(-7.3)**}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{1}})**}&0.3 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-0.1)}\tabularnewline
+~~D19&0 \footnotesize{(-1.9)**}&0 \footnotesize{(-0.4)}&0 \footnotesize{(-2.2)**}&0.1 \footnotesize{(-1.8)**}\tabularnewline
+~~D26&0.1 \footnotesize{(-1.2)**}&0.1 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-1.9)**}&0.1 \footnotesize{(-1.4)**}\tabularnewline
+~~D28&0 \footnotesize{(-1.7)**}&0 \footnotesize{(-0.2)}&0 \footnotesize{(-2.4)**}&0.1 \footnotesize{(-3.1)**}\tabularnewline
+~~D4&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-0.3)}&0.1 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+~~D6&0.1 \footnotesize{(-0.7)**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0.1 \footnotesize{(-0.3)**}\tabularnewline
+~~D7&0.1 \footnotesize{(-2.2)**}&0 \footnotesize{(-2.1)**}&0.1 \footnotesize{(\textit{\textbf{1.3}})**}&0.1 \footnotesize{(-0.4)*}\tabularnewline
+~~P8&0.2 \footnotesize{(-8.7)**}&0.1 \footnotesize{(-6.3)**}&0.2 \footnotesize{(-10.4)**}&0.5 \footnotesize{(-13.1)**}\tabularnewline
+\hline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
+~~C10&1.1 \footnotesize{(-4.2)**}&1.2 \footnotesize{(-3.2)**}&1.6 \footnotesize{(-1.9)**}&1.6 \footnotesize{(-3.3)**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.2 \footnotesize{(\textit{\textbf{1.7}})**}&0.2 \footnotesize{(-0.4)}\tabularnewline
+~~D19&0.4 \footnotesize{(-2.9)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.5)**}&0.6 \footnotesize{(-2.2)**}\tabularnewline
+~~D26&0.4 \footnotesize{(-1.9)**}&0.2 \footnotesize{(-1.6)**}&0.3 \footnotesize{(-0.6)*}&0.5 \footnotesize{(-0.6)**}\tabularnewline
+~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-5.4)**}&0.2 \footnotesize{(-5.2)**}&0.7 \footnotesize{(-1.7)**}\tabularnewline
+~~D4&0.3 \footnotesize{(-0.1)}&0.3 \footnotesize{(-1)**}&0.3 \footnotesize{(-1)**}&0.4 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.4)*}&0.4 \footnotesize{(-0.1)}\tabularnewline
+~~D7&0.4 \footnotesize{(-0.6)**}&0.4 \footnotesize{(-1.2)**}&0.4 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-0.3)*}\tabularnewline
+~~P8&1.3 \footnotesize{(-1.1)**}&1.1 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
 \hline

### 2016-12-12 sftrends_manu.Rnw
minor edits to methods

@@
 
+Something about flow
+
 San Francisco Bay on the Pacific Coast of the United States is one of the most prominent estuaries in the western hemisphere.
@@
 All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.
-Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 to 2012.
+Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 through 2013.
 Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
@@
 Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.
-Initial analyses indicated that model fit was not signifcantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
+Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
 As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.
@@
 Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.
+
 \subsection{Analysis method and application}
@@
 The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.
-As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series.
+As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \citep{Moyer12,Beck15}.
 The original \ac{wrtds} method is described in more detail in \citep{Hirsch10}.
@@
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
-Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter).
-These annual and seasonal groupings were chosen for continuity with similar comparisons reported in \citet{Jabusch16}.
+Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
+These annual and seasonal groupings were chosen for continuity with similar comparisons reported in \citet{Jabusch16} and as approximate twenty year midway points in the time series.
 
@@
 Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.
-The results were interpreted relative to those in \citet{Jabusch16} that were based on observed data to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
-The analysis was also similar to that in \citet{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression.
+Seasonal Kendall tests weere also used to describe trends in the model predictions for the observed data.
+These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
+The analysis was similar to that in \citet{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression.
 Functions in the EnvStats package in R were used for the seasonal Kendall tests \citep{Millard13}.
@@
 Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
-The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends on decadal time scales.
+The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends over several years.
 The selected case studies focused on \begin{inparaenum}[1\upshape)]
@@
 Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}.
-Abundance of clams is linked to inflow rates such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.
-
-We hypothesized that \ac{wrtds} models applied to water quality obsevations in the upper estuary would emphasize \begin{inparaenum}[1\upshape)]
+The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.
+
+We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
 \item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
@@
 Station names are shown in the top left panel.
-Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trnds} for actual values.
+Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 8, cache = T, out.width = '\\textwidth'>>=
@@
   )
+
 @

### 2016-12-12 sftrends_manu.tex
minor edits to methods

@@
   \vspace{1in} 
-  \\ Version Date:   Sun Dec 11 11:23:55 2016 -0600
+  \\ Version Date:   Mon Dec 12 13:46:45 2016 -0600
 	}
@@
 \item SF estuary, unique and prominent location, full story is complex (historical context and recent changes) \citep{Cloern12b}, why is the delta important (a vigorous biogeochemical reactor) \citep{Jassby00,Jassby02,Jassby08}, no one has empirically described the data in the delta using data-intensive methods
+
+Something about flow
 
@@
 Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}.
-Abundance of clams is linked to inflow rates such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.
+The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.
 
-We hypothesized that \ac{wrtds} models applied to water quality obsevations in the upper estuary would emphasize \begin{inparaenum}[1\upshape)]
+We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
 \item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
@@
 Station names are shown in the top left panel.
-Values less than -9\% change were truncated for visual clarity of the bar plots, see \cref{tab:trnds} for actual values.
+Values less than -9\% change were truncated for visual clarity of the bar plots, see \cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}

### 2016-12-13 sftrends_manu.Rnw
intro draft done, need to edit add more refs

@@
 % aliased citations
+\defcitealias{Beck15}{Beck and Hagy 2015}
 \defcitealias{BeckIP}{Beck and Murphy in press}
@@
 
-\begin{enumerate}
-\item How and why are trends interpreted - assessment of raw data, surrogates, various methods (kendall, GAM, WRTDS), what have been implications of using different approaches, see Kratzer USGS report http://pubs.usgs.gov/sir/2010/5228/pdf/sir20105228.pdf and data http://pubs.usgs.gov/sir/2010/5228/, need to interpret eutrophication trends in estuaries - it's confusing \citep{Cloern10}
-\item WRTDS, original method \citep{Hirsch10,Hirsch15}
-\item WRTDS application to Tampa Bay as test set \citep{Beck15}, further validation in Patuxent and other tidal waters \cite{Beck15b}
-\item SF estuary, unique and prominent location, full story is complex (historical context and recent changes) \citep{Cloern12b}, why is the delta important (a vigorous biogeochemical reactor) \citep{Jassby00,Jassby02,Jassby08}, no one has empirically described the data in the delta using data-intensive methods
-
-Something about flow
-
-San Francisco Bay on the Pacific Coast of the United States is one of the most prominent estuaries in the western hemisphere.
+Environmental condition is affected my multiple processes acting in a hierarchy across space and time \citep{Oneill89,Levin92}.
+Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.
+Visual examination of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables.
+Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \citep[\cref{fig:schematic},][]{Bhangu97,Champely97,Chang08,Halliday12}.
+As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years.
+Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes.
+For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system.
+Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes.
+Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
+
+Appropriate methods for the analysis of change depend largely on the question of interest and on charactericstics of the environmental dataset.
+Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}.
+Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
+Increased availability of multi-decadal datasets, particularly for high profile environments, has prompted the development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs.
+These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
+The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Hirsch14,Hirsch15}.
+This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
+More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citepalias{Beck15} and the Patuxent River Estuary \citepalias{BeckIP} and tidally-infleucned time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
+These studies have demonstrated potential for the use \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into drivers of change in aquatic systems.
+
+The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere.
 Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.
-Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed historical trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system.
+Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed historical trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
 The unique ecological and social context of the Bay, including a rich source of monitoring data from the last four decades, provides a valuable opportunity to gain insight into ecosystem properties of estuaries.
-
-\item Study goal and objectives
-\begin{itemize}
-\item Provide a description of trends - annual, seasonal, spatial, response to flow, change by analytes
-\item Detailed description of selected sites in the context of conceptual relationships - 1) nonlinear or extreme quantile changes, site TBD, 2) P8 and WWTP improvements, 3) Suisun DIN, SiO2, Chla, and clams
-\item What this means for understanding other systems
-\end{itemize}
-\end{enumerate}
+The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay.
+A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades.
+Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport suggesting an approach that explicitly considers flow effects is critical for trend analysis.
+To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
+%\citep{Jassby00,Jassby02,Jassby08}
+
+The goal of this study was to provide a comprehensive description of nutrient trends in the Delta that will inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay.
+We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts.
+The specific objectives were to \begin{inparaenum}[1\upshape)]
+\item quantify and interpret trends over four decades at nine stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
+\item provide detailed descriptions of three case studies in the context of conceptual relationships modelled with \ac{wrtds}.
+\end{inparaenum}
+The second objective evaluated three specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of \ac{wwtp} upgrades on water quality, and effects of biological invasion by benthic filter feeders on primary production.
+Although quantitative descriptions of change can be ends in themselves, the results were expected to have greater impact as a means to more detailed understanding of ecosystem properties.
+Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.
+Overall, this analysis is expected to further an ecological and management-based understanding of dynamics in San Francisco Bay, with implications for water quality restoration and protection of this prominent system.
 
@@
 However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.
-These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.
+These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12b}.
 The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations \citep[e.g.,][]{Dugdale07,Parker12,Glibert14}.
@@
 The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.
-Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.
+Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citepalias{Beck15}.
 
@@
 The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.
-As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \citep{Moyer12,Beck15}.
+As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series (\citealt{Moyer12}, \citetalias{Beck15}).
 The original \ac{wrtds} method is described in more detail in \citep{Hirsch10}.
-The \ac{wrtds} model applied to the Delta time series followed methods in \citet{Beck15}, which were based on a tidal adaptation of the original method.
+The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method (\citetalias{Beck15}).
 The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
@@
 A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
-Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citealt{Beck15}).
+Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citetalias{Beck15}).
 Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
@@
 Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}.
-The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.
+The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}.
 

### 2016-12-13 sftrends_manu.tex
intro draft done, need to edit add more refs

@@
   \vspace{1in} 
-  \\ Version Date:   Mon Dec 12 13:46:45 2016 -0600
+  \\ Version Date:   Mon Dec 12 17:29:39 2016 -0600
 	}
@@
 
-\begin{enumerate}
-\item How and why are trends interpreted - assessment of raw data, surrogates, various methods (kendall, GAM, WRTDS), what have been implications of using different approaches, see Kratzer USGS report http://pubs.usgs.gov/sir/2010/5228/pdf/sir20105228.pdf and data http://pubs.usgs.gov/sir/2010/5228/, need to interpret eutrophication trends in estuaries - it's confusing \citep{Cloern10}
-\item WRTDS, original method \citep{Hirsch10,Hirsch15}
-\item WRTDS application to Tampa Bay as test set \citep{Beck15}, further validation in Patuxent and other tidal waters \cite{Beck15b}
-\item SF estuary, unique and prominent location, full story is complex (historical context and recent changes) \citep{Cloern12b}, why is the delta important (a vigorous biogeochemical reactor) \citep{Jassby00,Jassby02,Jassby08}, no one has empirically described the data in the delta using data-intensive methods
+Environmental condition is affected my multiple processes acting in a hierarchy across space and time \citep{Oneill89,Levin92}.
+Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.
+Visual examination of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables.
+Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \citep[\cref{fig:schematic},][]{Bhangu97,Champely97,Chang08,Halliday12}.
+As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years.
+Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes.
+For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system.
+Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes.
+Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
 
-Something about flow
+Appropriate methods for the analysis of change depend largely on the question of interest and on charactericstics of the environmental dataset.
+Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}.
+Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
+Increased availability of multi-decadal datasets, particularly for high profile environments, has prompted the development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs.
+These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
+The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Hirsch14,Hirsch15}.
+This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
+More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citepalias{Beck15} and the Patuxent River Estuary \citepalias{BeckIP} and tidally-infleucned time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
+These studies have demonstrated potential for the use \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into drivers of change in aquatic systems.
 
-San Francisco Bay on the Pacific Coast of the United States is one of the most prominent estuaries in the western hemisphere.
+The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere.
 Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.
-Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed historical trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system.
+Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed historical trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
 The unique ecological and social context of the Bay, including a rich source of monitoring data from the last four decades, provides a valuable opportunity to gain insight into ecosystem properties of estuaries.
+The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay.
+A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades.
+Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport suggesting an approach that explicitly considers flow effects is critical for trend analysis.
+To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
+%\citep{Jassby00,Jassby02,Jassby08}
 
-\item Study goal and objectives
-\begin{itemize}
-\item Provide a description of trends - annual, seasonal, spatial, response to flow, change by analytes
-\item Detailed description of selected sites in the context of conceptual relationships - 1) nonlinear or extreme quantile changes, site TBD, 2) P8 and WWTP improvements, 3) Suisun DIN, SiO2, Chla, and clams
-\item What this means for understanding other systems
-\end{itemize}
-\end{enumerate}
+The goal of this study was to provide a comprehensive description of nutrient trends in the Delta that will inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay.
+We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts.
+The specific objectives were to \begin{inparaenum}[1\upshape)]
+\item quantify and interpret trends over four decades at nine stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
+\item provide detailed descriptions of three case studies in the context of conceptual relationships modelled with \ac{wrtds}.
+\end{inparaenum}
+The second objective evaluated three specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of \ac{wwtp} upgrades on water quality, and effects of biological invasion by benthic filter feeders on primary production.
+Although quantitative descriptions of change can be ends in themselves, the results were expected to have greater impact as a means to more detailed understanding of ecosystem properties.
+Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.
+Overall, this analysis is expected to further an ecological and management-based understanding of dynamics in San Francisco Bay, with implications for water quality restoration and protection of this prominent system.
 

### 2016-12-13 sftrends_manu.Rnw
more cites to intro

@@
 Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.
-Visual examination of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables.
+Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables.
 Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \citep[\cref{fig:schematic},][]{Bhangu97,Champely97,Chang08,Halliday12}.
@@
 Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes.
-For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system.
-Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes.
+For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}.
+Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
 Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
@@
 Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
-Increased availability of multi-decadal datasets, particularly for high profile environments, has prompted the development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs.
+Increased availability of multi-decadal datasets, particularly for high profile environments, has prompted the development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}.
 These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
-The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Hirsch14,Hirsch15}.
+The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Zhang16}.
 This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
-More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citepalias{Beck15} and the Patuxent River Estuary \citepalias{BeckIP} and tidally-infleucned time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
+More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citepalias{Beck15} and the Patuxent River Estuary \citepalias{BeckIP}, and tidally-infleunced time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
 These studies have demonstrated potential for the use \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into drivers of change in aquatic systems.
 
-The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere.
+The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \citep{Cloern15}.
 Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.
-Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed historical trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
-The unique ecological and social context of the Bay, including a rich source of monitoring data from the last four decades, provides a valuable opportunity to gain insight into ecosystem properties of estuaries.
-The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay.
-A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades.
-Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport suggesting an approach that explicitly considers flow effects is critical for trend analysis.
+Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
+The unique ecological and social context of the Bay provides a valuable opportunity to gain insight into ecosystem properties of estuaries that define water quality dynamics at different scales.
+The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay \citep{Jassby00,Jassby02,Jassby08}.
+A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades \citep{Jabusch16}.
+Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \citep{Novick15}, suggesting an approach that explicitly considers flow effects is critical for trend analysis.
 To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
-%\citep{Jassby00,Jassby02,Jassby08}
 

### 2016-12-13 sftrends_manu.tex
more cites to intro

@@
   \vspace{1in} 
-  \\ Version Date:   Mon Dec 12 17:29:39 2016 -0600
+  \\ Version Date:   Tue Dec 13 15:03:51 2016 -0600
 	}

### 2016-12-14 sftrends_manu.Rnw
added map figure

@@
 
-The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.
+The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).
 The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \citep{Jassby00,Jassby08,Novick15}.
@@
 All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.
-Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 through 2013.
+Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
 Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
@@
 \label{fig:schematic}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/delt_map.pdf}
+\caption{The \acl{sfe} Delta and monitoring stations used for analysis.
+The Delta drains the combined watershed from the Sacramento and San Joaquin rivers (bottom left).
+All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \citealt{IEP13}).}
+\label{fig:delt_map}   
 \end{figure}

### 2016-12-14 sftrends_manu.tex
added map figure

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Dec 13 15:03:51 2016 -0600
+  \\ Version Date:   Tue Dec 13 16:33:55 2016 -0600
 	}

### 2016-12-14 sftrends_manu.Rnw
added trncomp figures

@@
 
+Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.
+
 \subsection{Selected examples}
@@
 @
+
+<<trndcomp, echo = F, warning = F, message = F, cache = T>>=
+data(trnds_seasyr)
+
+# response labels 
+reslabs <- list(
+  shrt = c('din', 'nh', 'no23'),
+  expr = c(
+    'DIN', 
+    expression(paste(NH [4] ^ '+')),
+    expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
+  )
+)
+
+# combine all
+trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
+  select(Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
+  filter(!trndvar %in% 'res') %>% 
+  mutate(
+    Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')),
+    ztest = ifelse(ztest < 0.05, '(< 0.05)', '(ns)'),
+    resvar = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr), 
+    trndvar = factor(trndvar, levels = c('fits', 'norm'), labels = c('Pred.', 'Flow-norm.'))
+  ) %>% 
+  unite('trnd', trndvar, ztest, sep = ' ')
+
+# ggplot theme
+mythm <- theme_minimal() +
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank(),
+    axis.title.y = element_blank(),
+    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.major = element_blank(), 
+    panel.grid.minor = element_blank()
+  )
+
+# year only
+toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
+p1 <- ggplot(toplo1, aes(x = tau, y = Site_Code)) + 
+  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed
+    ) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  mythm + 
+  xlab(expression(tau)) +
+  scale_shape_manual(values = c(16, 1, 17, 2)) +
+  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
+  scale_x_continuous(limits = c(-1, 1)) + 
+  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6)
+
+# seasonal, first years
+toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
+p2 <- ggplot(toplo2, aes(x = tau, y = Site_Code)) + 
+  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed
+    ) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  mythm + 
+  xlab(expression(tau)) +
+  scale_shape_manual(values = c(16, 1, 17, 2)) +
+  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
+  scale_x_continuous(limits = c(-1, 1)) + 
+  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6)
+
+# seasonal, first years
+toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
+p3 <- ggplot(toplo3, aes(x = tau, y = Site_Code)) + 
+  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed
+    ) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  mythm + 
+  xlab(expression(tau)) +
+  scale_shape_manual(values = c(16, 1, 17, 2)) +
+  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
+  scale_x_continuous(limits = c(-1, 1)) + 
+  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6)
+
+pdf('figs/trndcomp1.pdf', height = 5, width = 9, family = 'serif')
+p1
+dev.off()
+
+pdf('figs/trndcomp2.pdf', height = 8, width = 9, family = 'serif')
+p2
+dev.off()
+
+pdf('figs/trndcomp3.pdf', height = 8, width = 9, family = 'serif')
+p3
+dev.off()
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
+\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
+Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Trends are shown separately for different annual groupings.
+See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
+\label{fig:trndcomp1}   
+\end{figure}
 
@@
 
+\clearpage
+
+% supplementary material
+\beginsupplement
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
+\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
+Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Trends are shown separately for different seasonal groupings from 1976-1995.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp2}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
+\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
+Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Trends are shown separately for different seasonal groupings from 1996-2013.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp3}   
+\end{figure}
+
 \end{document}

### 2016-12-14 sftrends_manu.tex
added trncomp figures

@@
   \vspace{1in} 
-  \\ Version Date:   Tue Dec 13 16:33:55 2016 -0600
+  \\ Version Date:   Wed Dec 14 15:30:51 2016 -0600
 	}
@@
 Seasonal changes at D7 are most pronounced during the summer.}\label{fig:d7c10trnds}
+\end{figure}
+
+pdf 
+  2 
+pdf 
+  2 
+pdf 
+  2
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
+\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
+Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Trends are shown separately for different annual groupings.
+See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
+\label{fig:trndcomp1}   
 \end{figure}

### 2016-12-15 sftrends_manu.Rnw
yaxis lab to flow comp plots

@@
     legend.title = element_blank(),
-    axis.title.y = element_blank(),
     panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
@@
   scale_x_continuous(limits = c(-1, 1)) + 
-  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6)
+  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
+  scale_y_discrete('Stations')
 

### 2017-01-08 sftrends_manu.Rnw
removed tex aux files

@@
 \raggedbottom
-\linenumbers
+% \linenumbers
 \raggedright
@@
 
-Environmental condition is affected my multiple processes acting in a hierarchy across space and time \citep{Oneill89,Levin92}.
+Environmental conditions are affected by multiple processes acting in a hierarchy across space and time \citep{Oneill89,Levin92}.
 Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.
@@
 
-Appropriate methods for the analysis of change depend largely on the question of interest and on charactericstics of the environmental dataset.
+Appropriate methods for the analysis of change depend largely on the question of interest and on characteristics of the environmental dataset.
 Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}.
 Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
-Increased availability of multi-decadal datasets, particularly for high profile environments, has prompted the development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}.
+Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated recent development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}.
 These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
@@
 
-<<trndcomp, echo = F, warning = F, message = F, cache = T>>=
+<<trndcomp, echo = F, warning = F, message = F, cache = T, results = 'hide'>>=
 data(trnds_seasyr)

### 2017-01-08 sftrends_manu.tex
yaxis lab to flow comp plots
removed tex aux files

@@
   \vspace{1in} 
-  \\ Version Date:   Wed Dec 14 15:30:51 2016 -0600
+  \\ Version Date:   Thu Dec 15 16:04:34 2016 -0600
 	}
@@
 
-pdf 
-  2 
-pdf 
-  2 
-pdf 
-  2
-
 \begin{figure}

### 2017-01-08 sftrends_manu.Rnw
achemso style

@@
-\documentclass[letterpaper,12pt,oneside]{article}
-\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
-\usepackage{setspace}
-\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
+\documentclass[journal = esthag, manuscript = article]{achemso}
+
+% \usepackage[version=3]{mhchem}
+\usepackage[T1]{fontenc}
+\usepackage[colorlinks  =true, allcolors=Blue]{hyperref}
 \usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{indentfirst}
-\usepackage{titlesec}
-\usepackage{multirow}
-\usepackage{booktabs}
-\usepackage{graphicx}
-\usepackage{verbatim}
-\usepackage{rotating}
-\usepackage{tabularx}
-\usepackage{outlines}
 \usepackage{lineno}
-\usepackage{array}
-\usepackage{times}
 \usepackage{cleveref}
 \usepackage{acronym}
-\usepackage[position=t]{subfig}
 \usepackage{paralist}
-\usepackage[noae]{Sweave}
-\usepackage{natbib}
-\usepackage{array}
-\usepackage{pdflscape}
 \usepackage{bm}
+\usepackage{fixltx2e}
 % \usepackage{showlabels}
-\bibpunct{(}{)}{,}{a}{}{,}
-
-% page margins and section title formatting
-\linespread{1.5}
-\setlength{\footskip}{0.5in}
-\titleformat*{\section}{\Large\bf\em}
-\titleformat*{\subsection}{\singlespace\large\bf}
-\titleformat*{\subsubsection}{\singlespace\normalsize\bf\em}
-\titlespacing{\section}{0in}{0in}{0in}
-\titlespacing{\subsection}{0in}{0in}{0in}
-\titlespacing{\subsubsection}{0in}{0in}{0in}
 
@@
 \renewcommand{\figurename}{Fig.}
-
-% aliased citations
-\defcitealias{Beck15}{Beck and Hagy 2015}
-\defcitealias{BeckIP}{Beck and Murphy in press}
-\defcitealias{SutulaIR}{Sutula et al. in review}
 
@@
 
+\author{Marcus W. Beck}
+\affiliation{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, Gulf Breeze, FL}
+\email{beck.marcus@epa.gov}
+\phone{+1 (850)9342480}
+\fax{+1 (850)9342401}
+\author{David Senn}
+\author{Thomas Jabusch}
+\author{Phil Trowbridge}
+\affiliation{San Francisco Estuary Institute, Richmond, CA}
+
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
+
 \begin{document}
-
-\raggedbottom
 % \linenumbers
-\raggedright
-\urlstyle{same}
-\setlength{\parindent}{0.5in}
-\renewcommand\refname{References \vspace{12pt}}
-
-\begin{singlespace}
-\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
-\author{
-  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
-  \\{\bf {\normalsize Thomas Jabusch$^2$, Phil Trowbridge$^2$}}
-  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
-  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
-	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
-	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
-  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
-	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
-	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
-  \vspace{1in} 
-  \\ \Sexpr{raw}
-	}
-\date{}
-\maketitle
-\end{singlespace}
-\clearpage
 
@@
 
-Environmental conditions are affected by multiple processes acting in a hierarchy across space and time \citep{Oneill89,Levin92}.
+Environmental conditions are affected by multiple processes acting in a hierarchy across space and time \cite{Oneill89,Levin92}.
 Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.
 Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables.
-Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \citep[\cref{fig:schematic},][]{Bhangu97,Champely97,Chang08,Halliday12}.
+Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis (\cref{fig:schematic})\cite{Bhangu97,Champely97,Chang08,Halliday12}.
 As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years.
 Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes.
-For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}.
-Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
+For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}.
+Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}.
 Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
@@
 Appropriate methods for the analysis of change depend largely on the question of interest and on characteristics of the environmental dataset.
-Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}.
+Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}.
 Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
-Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated recent development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}.
+Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated recent development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}.
 These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
-The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Zhang16}.
+The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Zhang16}.
 This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
-More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citepalias{Beck15} and the Patuxent River Estuary \citepalias{BeckIP}, and tidally-infleunced time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
+More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-infleunced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
 These studies have demonstrated potential for the use \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into drivers of change in aquatic systems.
 
-The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \citep{Cloern15}.
+The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \cite{Cloern15}.
 Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.
-Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
+Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
 The unique ecological and social context of the Bay provides a valuable opportunity to gain insight into ecosystem properties of estuaries that define water quality dynamics at different scales.
-The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay \citep{Jassby00,Jassby02,Jassby08}.
-A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades \citep{Jabusch16}.
-Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \citep{Novick15}, suggesting an approach that explicitly considers flow effects is critical for trend analysis.
+The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay \cite{Jassby00,Jassby02,Jassby08}.
+A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades \cite{Jabusch16}.
+Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting an approach that explicitly considers flow effects is critical for trend analysis.
 To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
@@
 The \ac{sfe} system is divided into several sub-bays, including Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.
-Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \citep{Jassby00}.
-Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \citep{Cloern12b}.
+Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
+Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \cite{Cloern12b}.
 The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns.
-Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \citep{Cloern15}.
-Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \citep{Cloern07}.
-
-Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.
-Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
-Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
-Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.
-These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12b}.
-The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations \citep[e.g.,][]{Dugdale07,Parker12,Glibert14}.
+Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.
+Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \cite{Cloern07}.
+
+Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication (e.g., Chesapeake Bay).\cite{Kemp05}  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
+Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.
+Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \cite{Thompson08,Crauder16}.
+However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.
+These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \cite{Cloern07}, biological invasions \cite{Cohen98}, and departures from the historical flow record \cite{Enright09,Cloern12b}.
+The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
 
 The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).
-The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \citep{Jassby00,Jassby08,Novick15}.
+The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.
 Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.
-A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}.
+A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}.
 For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.
 Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
-Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \citep{Jassby00,Jassby08}.
+Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.
 Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
@@
 Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.
-The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \citep{IEP13}.
+The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.
 The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.
 Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.
-All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.
+All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.
 Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
@@
 
-Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \citep{IEP16}.
+Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.
 Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.
@@
 The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.
-Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citepalias{Beck15}.
+Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.
 
@@
 The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.
-As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series (\citealt{Moyer12}, \citetalias{Beck15}).
-The original \ac{wrtds} method is described in more detail in \citep{Hirsch10}.
-The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method (\citetalias{Beck15}).
-The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
+As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.
+The original \ac{wrtds} method is described in more detail in Ref.~\citenum{Hirsch10}.
+The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method\cite{Beck15}.
+The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
 Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme evens.
 The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.
-All analyses used the WRTDStidal package for R \citep{Beck16b,RDCT16}
+All analyses used the WRTDStidal package for R \cite{Beck16b,RDCT16}
 
 A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
-Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citetalias{Beck15}).
+Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic})\cite{Beck15}.
 Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
@@
 Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
-These annual and seasonal groupings were chosen for continuity with similar comparisons reported in \citet{Jabusch16} and as approximate twenty year midway points in the time series.
+These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midway points in the time series.
 
 Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
-This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82}.
+This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82}.
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
@@
 These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
-The analysis was similar to that in \citet{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression.
-Functions in the EnvStats package in R were used for the seasonal Kendall tests \citep{Millard13}.
+The analysis was similar to that in Ref.~\citenum{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression.
+Functions in the EnvStats package in R were used for the seasonal Kendall tests \cite{Millard13}.
 
@@
 
-Multiple biological and physical factors influence nutrient concentrations at different temporal scales \citep[][references therein]{Cloern10}.
+Multiple biological and physical factors influence nutrient concentrations at different temporal scales \cite{Cloern10}.
 As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex as compared to more conventional methods for time series analysis.
@@
 Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.
-As noted in \citep{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.
+As noted in \cite{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.
 Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., \citealt{Cornwell14,Novick15}).
 Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}.
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}.
 Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
@@
 
-The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \citep{Cohen98}.
-The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \citep{Carlton90,Nichols90,Werner93,Crauder16}.
+The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \cite{Cohen98}.
+The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \cite{Carlton90,Nichols90,Werner93,Crauder16}.
 In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
-Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \citep{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}.
-The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}.
+Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}.
+The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
 
@@
 
-\section{Results}
+\section{Results and Discussion}
+
+\subsection{Observed Data}
 
@@
 
-Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in \citet{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
-\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from \citet{Jabusch16}.
+Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in Ref.~\citenum{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
+\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from Ref.~\citenum{Jabusch16}.
 Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).
@@
 Emphasize the summer/fall change in the 2000s, why is this?
-Check \citep{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
+Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
 Relate to conceptual diagram.
@@
 
-Data from \citep{Crauder16}, \citet{Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
-Figure 10 in \citet{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
+Data from \cite{Crauder16,Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
+Figure 10 in Ref.~\citenum{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
 
@@
 This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
-See suggestions in \citet{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
+See suggestions in Ref.~\citenum{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
 
@@
 
-\clearpage
-\begin{singlespace}
-\bibliographystyle{apalike_mine}
 \bibliography{refs}
-\end{singlespace}
-\clearpage
 
@@
 % supplementary material
+\begin{suppinfo}
 \beginsupplement
@@
 
+\end{suppinfo}
+
 \end{document}

### 2017-01-08 sftrends_manu.tex
achemso style

@@
-\documentclass[letterpaper,12pt,oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[journal = esthag, manuscript = article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \usepackage{alltt}
-\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
-\usepackage{setspace}
-\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
+
+% \usepackage[version=3]{mhchem}
+\usepackage[T1]{fontenc}
+\usepackage[colorlinks  =true, allcolors=Blue]{hyperref}
 \usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{indentfirst}
-\usepackage{titlesec}
-\usepackage{multirow}
-\usepackage{booktabs}
-\usepackage{graphicx}
-\usepackage{verbatim}
-\usepackage{rotating}
-\usepackage{tabularx}
-\usepackage{outlines}
 \usepackage{lineno}
-\usepackage{array}
-\usepackage{times}
 \usepackage{cleveref}
 \usepackage{acronym}
-\usepackage[position=t]{subfig}
 \usepackage{paralist}
-\usepackage[noae]{Sweave}
-\usepackage{natbib}
-\usepackage{array}
-\usepackage{pdflscape}
 \usepackage{bm}
+\usepackage{fixltx2e}
 % \usepackage{showlabels}
-\bibpunct{(}{)}{,}{a}{}{,}
-
-% page margins and section title formatting
-\linespread{1.5}
-\setlength{\footskip}{0.5in}
-\titleformat*{\section}{\Large\bf\em}
-\titleformat*{\subsection}{\singlespace\large\bf}
-\titleformat*{\subsubsection}{\singlespace\normalsize\bf\em}
-\titlespacing{\section}{0in}{0in}{0in}
-\titlespacing{\subsection}{0in}{0in}{0in}
-\titlespacing{\subsubsection}{0in}{0in}{0in}
 
@@
 
+\author{Marcus W. Beck}
+\affiliation{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, Gulf Breeze, FL}
+\email{beck.marcus@epa.gov}
+\phone{+1 (850)9342480}
+\fax{+1 (850)9342401}
+\author{David Senn}
+\author{Thomas Jabusch}
+\author{Phil Trowbridge}
+\affiliation{San Francisco Estuary Institute, Richmond, CA}
+
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
-
-\raggedbottom
 % \linenumbers
-\raggedright
-\urlstyle{same}
-\setlength{\parindent}{0.5in}
-\renewcommand\refname{References \vspace{12pt}}
-
-\begin{singlespace}
-\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
-\author{
-  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
-  \\{\bf {\normalsize Thomas Jabusch$^2$, Phil Trowbridge$^2$}}
-  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
-  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
-	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
-	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
-  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
-	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
-	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
-  \vspace{1in} 
-  \\ Version Date:   Thu Dec 15 16:04:34 2016 -0600
-	}
-\date{}
-\maketitle
-\end{singlespace}
-\clearpage
 
@@
 The \ac{sfe} system is divided into several sub-bays, including Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.
-Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \citep{Jassby00}.
-Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \citep{Cloern12b}.
+Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
+Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \cite{Cloern12b}.
 The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns.
-Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \citep{Cloern15}.
-Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \citep{Cloern07}.
+Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.
+Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \cite{Cloern07}.
 
-Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.
-Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
-Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
-Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.
-These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12b}.
-The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations \citep[e.g.,][]{Dugdale07,Parker12,Glibert14}.
+Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication (e.g., Chesapeake Bay).\cite{Kemp05}  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
+Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.
+Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \cite{Thompson08,Crauder16}.
+However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.
+These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \cite{Cloern07}, biological invasions \cite{Cohen98}, and departures from the historical flow record \cite{Enright09,Cloern12b}.
+The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
 
 The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).
-The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \citep{Jassby00,Jassby08,Novick15}.
+The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.
 Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.
-A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}.
+A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}.
 For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.
 Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
-Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \citep{Jassby00,Jassby08}.
+Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.
 Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

### 2017-01-09 sftrends_manu.Rnw
fig schematic gone

@@
 \crefname{table}{Table}{Tables}
-\crefname{figure}{Fig.}{Figs.}
-\renewcommand{\figurename}{Fig.}
+\crefname{figure}{Figure}{Figures}
+\renewcommand{\figurename}{Figure}
 
@@
 Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables.
-Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis (\cref{fig:schematic})\cite{Bhangu97,Champely97,Chang08,Halliday12}.
+Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
 As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years.
@@
 All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.
-Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
-Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
+Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
 These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.
@@
 The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
-Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme evens.
+Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme events.
 The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.
-All analyses used the WRTDStidal package for R \cite{Beck16b,RDCT16}
+All analyses used the WRTDStidal package for R \cite{Beck16b,RDCT16}.
 
 A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
-Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic})\cite{Beck15}.
+Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders\cite{Beck15}.
 Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
@@
 As noted in \cite{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., \citealt{Cornwell14,Novick15}).
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., )\cite{Cornwell14,Novick15}.
 Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
@@
 
-Trends as percent change depend on the mean value, lower values will have larger percent changes.
+Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta.
 
@@
 \centering
-\includegraphics[width=1\textwidth,page=1]{figs/schematic.pdf}
-\caption{Conceptual diagram illustrating use of \ac{wrtds} to decompose trends in observed nitrogen time series and potential forcing factors that can explain model output.
-Results from the model are described as annual and seasonal trends, changes in flow-nutrient dynamics for different time periods, and residual variation independent of time, flow, and season.
-Relationships between environmental factors (climate, local, regional/historical) and nitrogen trends are more easily related to the separate components of the observed time series using results from the model.
-}
-\label{fig:schematic}   
-\end{figure}
-
-\begin{figure}
-\centering
 \includegraphics[width=1\textwidth,page=1]{figs/delt_map.pdf}
@@
 The Delta drains the combined watershed from the Sacramento and San Joaquin rivers (bottom left).
-All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \citealt{IEP13}).}
+All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.}
 \label{fig:delt_map}   

### 2017-01-09 sftrends_manu.Rnw
map update with md10

@@
 The specific objectives were to \begin{inparaenum}[1\upshape)]
-\item quantify and interpret trends over four decades at nine stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
+\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
 \item provide detailed descriptions of three case studies in the context of conceptual relationships modelled with \ac{wrtds}.
@@
 \label{fig:delt_map}   
+\end{figure}
+
+<<trndcomp, echo = F, warning = F, message = F, cache = T, results = 'hide'>>=
+data(trnds_seasyr)
+
+# response labels 
+reslabs <- list(
+  shrt = c('din', 'nh', 'no23'),
+  expr = c(
+    'DIN', 
+    expression(paste(NH [4] ^ '+')),
+    expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
+  )
+)
+
+# combine all
+trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
+  select(Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
+  filter(!trndvar %in% 'res') %>% 
+  mutate(
+    Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')),
+    ztest = ifelse(ztest < 0.05, '(< 0.05)', '(ns)'),
+    resvar = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr), 
+    trndvar = factor(trndvar, levels = c('fits', 'norm'), labels = c('Pred.', 'Flow-norm.'))
+  ) %>% 
+  unite('trnd', trndvar, ztest, sep = ' ')
+
+# ggplot theme
+mythm <- theme_minimal() +
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank(),
+    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.major = element_blank(), 
+    panel.grid.minor = element_blank()
+  )
+
+# year only
+toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
+p1 <- ggplot(toplo1, aes(x = tau, y = Site_Code)) + 
+  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed
+    ) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  mythm + 
+  xlab(expression(tau)) +
+  scale_shape_manual(values = c(16, 1, 17, 2)) +
+  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
+  scale_x_continuous(limits = c(-1, 1)) + 
+  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
+  scale_y_discrete('Stations')
+
+# seasonal, first years
+toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
+p2 <- ggplot(toplo2, aes(x = tau, y = Site_Code)) + 
+  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed
+    ) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  mythm + 
+  xlab(expression(tau)) +
+  scale_shape_manual(values = c(16, 1, 17, 2)) +
+  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
+  scale_x_continuous(limits = c(-1, 1)) + 
+  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
+  scale_y_discrete('Stations')
+
+# seasonal, first years
+toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
+p3 <- ggplot(toplo3, aes(x = tau, y = Site_Code)) + 
+  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed
+    ) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  mythm + 
+  xlab(expression(tau)) +
+  scale_shape_manual(values = c(16, 1, 17, 2)) +
+  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
+  scale_x_continuous(limits = c(-1, 1)) + 
+  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
+  scale_y_discrete('Stations')
+
+pdf('figs/trndcomp1.pdf', height = 5, width = 9, family = 'serif')
+p1
+dev.off()
+
+pdf('figs/trndcomp2.pdf', height = 8, width = 9, family = 'serif')
+p2
+dev.off()
+
+pdf('figs/trndcomp3.pdf', height = 8, width = 9, family = 'serif')
+p3
+dev.off()
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
+\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
+Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Trends are shown separately for different annual groupings.
+See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
+\label{fig:trndcomp1}   
 \end{figure}
@@
 @
-
-<<trndcomp, echo = F, warning = F, message = F, cache = T, results = 'hide'>>=
-data(trnds_seasyr)
-
-# response labels 
-reslabs <- list(
-  shrt = c('din', 'nh', 'no23'),
-  expr = c(
-    'DIN', 
-    expression(paste(NH [4] ^ '+')),
-    expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
-  )
-)
-
-# combine all
-trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
-  select(Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
-  filter(!trndvar %in% 'res') %>% 
-  mutate(
-    Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')),
-    ztest = ifelse(ztest < 0.05, '(< 0.05)', '(ns)'),
-    resvar = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr), 
-    trndvar = factor(trndvar, levels = c('fits', 'norm'), labels = c('Pred.', 'Flow-norm.'))
-  ) %>% 
-  unite('trnd', trndvar, ztest, sep = ' ')
-
-# ggplot theme
-mythm <- theme_minimal() +
-  theme(
-    legend.position = 'top', 
-    legend.title = element_blank(),
-    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"), 
-    panel.grid.major = element_blank(), 
-    panel.grid.minor = element_blank()
-  )
-
-# year only
-toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
-p1 <- ggplot(toplo1, aes(x = tau, y = Site_Code)) + 
-  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
-  facet_grid(cat~resvar, labeller = label_parsed
-    ) + 
-  geom_vline(xintercept = 0, linetype = 'dashed') + 
-  mythm + 
-  xlab(expression(tau)) +
-  scale_shape_manual(values = c(16, 1, 17, 2)) +
-  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
-  scale_x_continuous(limits = c(-1, 1)) + 
-  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
-  scale_y_discrete('Stations')
-
-# seasonal, first years
-toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
-p2 <- ggplot(toplo2, aes(x = tau, y = Site_Code)) + 
-  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
-  facet_grid(cat~resvar, labeller = label_parsed
-    ) + 
-  geom_vline(xintercept = 0, linetype = 'dashed') + 
-  mythm + 
-  xlab(expression(tau)) +
-  scale_shape_manual(values = c(16, 1, 17, 2)) +
-  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
-  scale_x_continuous(limits = c(-1, 1)) + 
-  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
-  scale_y_discrete('Stations')
-
-# seasonal, first years
-toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
-p3 <- ggplot(toplo3, aes(x = tau, y = Site_Code)) + 
-  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
-  facet_grid(cat~resvar, labeller = label_parsed
-    ) + 
-  geom_vline(xintercept = 0, linetype = 'dashed') + 
-  mythm + 
-  xlab(expression(tau)) +
-  scale_shape_manual(values = c(16, 1, 17, 2)) +
-  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
-  scale_x_continuous(limits = c(-1, 1)) + 
-  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
-  scale_y_discrete('Stations')
-
-pdf('figs/trndcomp1.pdf', height = 5, width = 9, family = 'serif')
-p1
-dev.off()
-
-pdf('figs/trndcomp2.pdf', height = 8, width = 9, family = 'serif')
-p2
-dev.off()
-
-pdf('figs/trndcomp3.pdf', height = 8, width = 9, family = 'serif')
-p3
-dev.off()
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
-\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
-Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
-Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
-Trends are shown separately for different annual groupings.
-See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
-\label{fig:trndcomp1}   
-\end{figure}
 

### 2017-01-09 sftrends_manu.Rnw
obsdat fig added

@@
 Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
-Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+Representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
 These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.
@@
 \end{figure}
+
+<<obsdat, fig.cap = 'Observed \\ac{din} from ten stations in the upper \\ac{sfe} Delta.
+Data were collected monthly and evaluated with \\ac{wrtds} models from 1976 to 2013.
+Note different y-axis scales.
+See \\cref{fig:delt_map} for station locations.', fig.height = 8, fig.width = 6, cache = T, out.width = '0.8\\textwidth'>>=
+data(delt_dat)
+
+missvals <- c('1996-01-01', '2004-04-01') %>% 
+  as.Date
+missvals <- seq(missvals[1], missvals[2], by = 'month')
+missvals <- data.frame(Site_Code = 'D19', Date = missvals, nh = 0, no23 = 0)
+
+toplo <- delt_dat %>% 
+  select(Site_Code, Date, nh, no23) %>% 
+  rbind(., missvals) %>% 
+  gather('var', 'val', nh:no23) %>% 
+  mutate(Site_Code = factor(Site_Code, 
+    levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3'))
+    )
+
+mythm <- theme_minimal() +
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank(),
+    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.title.x = element_blank(),
+    axis.text.y = element_text(size = 6),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.major = element_blank(), 
+    panel.grid.minor = element_blank()
+  )
+
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(4, 7)]
+ylab <- expression(paste('DIN (mg ', L^-1, ')'))
+
+leglabs <- list(
+  expression(NH [4] ^ '+'),
+  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
+  )
+
+ggplot(toplo, aes(x = Date)) +
+  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack') + 
+  scale_x_date(expand = c(0, 0)) + 
+  scale_y_continuous(ylab, expand = c(0, 0)) + 
+  facet_grid(Site_Code~., scales = 'free_y') +
+  scale_fill_manual(values = cols, labels = leglabs) +
+  mythm
+
+@
 

### 2017-01-09 sftrends_manu.tex
map update with md10
obsdat fig added

@@
 
+{\centering \includegraphics[width=0.8\textwidth]{figs/obsdat-1}
+
+}
+
+\caption{Observed \ac{din} from ten stations in the upper \ac{sfe} Delta.
+Data were collected monthly and evaluated with \ac{wrtds} models from 1976 to 2013.
+Note different y-axis scales.
+See \cref{fig:delt_map} for station locations.}\label{fig:obsdat}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
+\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
+Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Trends are shown separately for different annual groupings.
+See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
+\label{fig:trndcomp1}   
+\end{figure}
+
+\begin{figure}[!ht]
+
 {\centering \includegraphics[width=\textwidth]{figs/trndmap-1}
@@
 Seasonal changes at D7 are most pronounced during the summer.}\label{fig:d7c10trnds}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
-\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
-Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
-Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
-Trends are shown separately for different annual groupings.
-See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
-\label{fig:trndcomp1}   
 \end{figure}

### 2017-01-09 sftrends_manu.Rnw
annual per chg removed from trndmap

@@
 
-<<obsdat, fig.cap = 'Observed \\ac{din} from ten stations in the upper \\ac{sfe} Delta.
+<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta.
 Data were collected monthly and evaluated with \\ac{wrtds} models from 1976 to 2013.
 Note different y-axis scales.
-See \\cref{fig:delt_map} for station locations.', fig.height = 8, fig.width = 6, cache = T, out.width = '0.8\\textwidth'>>=
+See \\cref{fig:delt_map} for station locations.', fig.height = 7, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
 data(delt_dat)
@@
   select(Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
-  filter(!trndvar %in% 'res') %>% 
+  filter(!trndvar %in% 'fits') %>% 
   mutate(
     Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')),
-    ztest = ifelse(ztest < 0.05, '(< 0.05)', '(ns)'),
+    ztest = ifelse(ztest < 0.05, '(p < 0.05)', '(ns)'),
     resvar = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr), 
-    trndvar = factor(trndvar, levels = c('fits', 'norm'), labels = c('Pred.', 'Flow-norm.'))
+    trndvar = factor(trndvar, levels = c('res', 'norm'), labels = c('Obs.', 'Flow-norm.'))
   ) %>% 
@@
   xlab(expression(tau)) +
-  scale_shape_manual(values = c(16, 1, 17, 2)) +
+  scale_shape_manual(values = c(1, 16, 2, 17)) +
   scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
@@
 Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 8, cache = T, out.width = '\\textwidth'>>=
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 5.3, cache = T, out.width = '0.7\\textwidth'>>=
 
@@
     ) %>%
+  filter(!cat %in% c('1976-1995', '1996-2013')) %>% # remove year categories
   unite('cattype', cat, ann, sep = '', remove = F) %>%
@@
     cattype = fct_collapse(cattype, 
-      an = c('1976-1995bef', '1996-2013aft'),
       mobef = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'bef'), 
@@
   set.seed(5)
-  if(which(uni == unis) == 1)
+  if(which(uni == unis) == 2)
     pall <- pbase + 
@@
         force = 2, size = 2,
-        fill = 'grey90'
+        fill = 'grey95'
       )
@@
   arrangeGrob(
-    arrangeGrob (plo_ls[[1]], plo_ls[[4]], plo_ls[[7]], ncol = 1, top = c('(a) 1976-1995, 1996-2013')), 
-    arrangeGrob(plo_ls[[3]], plo_ls[[6]], plo_ls[[9]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1976-1995)')),
-    arrangeGrob(plo_ls[[2]], plo_ls[[5]], plo_ls[[8]], ncol = 1, top = c('(c) Sp, Su, Fa, Wi (1996-2013)')),
-    ncol = 3
+    arrangeGrob(plo_ls[[2]], plo_ls[[4]], plo_ls[[6]], ncol = 1, top = c('(a) Sp, Su, Fa, Wi (1976-1995)')),
+    arrangeGrob(plo_ls[[1]], plo_ls[[3]], plo_ls[[5]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1996-2013)')),
+    ncol = 2
     ),
@@
   )
-lncol <- RColorBrewer::brewer.pal(9, 'Set1')[2]
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(3, 5, 8)]
 
@@
   scale_x_date(expand = c(0, 0), limits = xrng) + 
-  scale_fill_brewer(palette = 'Blues', labels = leglabs) + 
+  scale_fill_manual(values = cols, labels = leglabs) + 
   geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +

### 2017-01-09 sftrends_manu.Rnw
dynaplot to fig dinc10

@@
 \usepackage{fixltx2e}
-% \usepackage{showlabels}
+\usepackage{showlabels}
 
@@
 \item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
-\item provide detailed descriptions of three case studies in the context of conceptual relationships modelled with \ac{wrtds}.
+\item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
 \end{inparaenum}
-The second objective evaluated three specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of \ac{wwtp} upgrades on water quality, and effects of biological invasion by benthic filter feeders on primary production.
+The second objective evaluated two specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of \ac{wwtp} upgrades on water quality, and effects of biological invasion by benthic filter feeders on primary production.
 Although quantitative descriptions of change can be ends in themselves, the results were expected to have greater impact as a means to more detailed understanding of ecosystem properties.
@@
 
-\subsection{Case studies}
-
-Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
+\section{Results and Discussion}
+
+\subsection{Observed Data}
+
+\subsection{Trends}
+
+kendall tests - observed
+
+What about flow? \cref{fig:dinc10}, \cref{fig:dinc10dyna}
+
+Emphasize the information the model provides relative to the observed time series.
+A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
+The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
+Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
+The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
+Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
+Emphasize the summer/fall change in the 2000s, why is this?
+Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
+Relate to conceptual diagram.
+
+Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in Ref.~\citenum{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
+\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from Ref.~\citenum{Jabusch16}.
+Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).
+
+Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.
+
+\subsection{Selected examples}
+
+Two stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
 The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends over several years.
 The selected case studies focused on \begin{inparaenum}[1\upshape)]
-\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
 \item effects of wastewater treatment upgrades upstream of P8, and
@@
 Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
-
-\subsubsection{Disaggregating observed nitrogen time series}
-
-Multiple biological and physical factors influence nutrient concentrations at different temporal scales \cite{Cloern10}.
-As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex as compared to more conventional methods for time series analysis.
-A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, was expected to reveal \begin{inparaenum}[1\upshape)]
-\item an annual trend independent of a seasonal trend, and 
-\item varying flow contribution to nutrient dynamics, either as a difference between predicted and flow-normalized results or changes in the nutrient versus flow relationship at different annual periods.
-\end{inparaenum}
-Results from this case study were used to demonstrate the breadth of information that can be obtained from the observed time series with \ac{wrtds}.
 
@@
 
+Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:stock}).
+Nitrate is the primary constituent of total nitrogen after 2007.
+Organic nitrogen is a larger percentage of the total after nitrification.
+What was reduction in ammonium starting in 2002?
+
+Nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}), with ammonium showing the largest reduction.
+Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
+Percent changes are shown in \cref{tab:p8chg}, where both nitrogen species shows large percent increases prior to \ac{wwtp} upgrades followed by decreases after upgrades with ammonium showing the largest pecentage.
+Seasonally, increases prior to upgrades were most apparent in the \ac{jas} months for both analytes.
+Seasonal reductions post-upgrades were also largest in \ac{jas} for nitrite/nitrate, whereas percent reductions were similar across all monthly groupings for ammonium.
+
+Relationships of nitrogen with flow showed the typical inverse flow/concentration dynamic with flushing at high flow, although patterns differed by nitrogen species.
+Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations in the winter.
+Additionally, strength of the flow/nutrient relationship changed throughout the time series the year where the strongest relationship differed by analyte.
+Nitrite/nitrate typically had the strongest relationship flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
+
 \subsubsection{Effects of biological invasions}
@@
 The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
-
-\section{Results and Discussion}
-
-\subsection{Observed Data}
-
-\subsection{Trends}
-
-Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in Ref.~\citenum{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
-\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from Ref.~\citenum{Jabusch16}.
-Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).
-
-Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.
-
-\subsection{Selected examples}
-
-\subsubsection{Disaggregating observed nitrogen time series}
-
-\cref{fig:dinc10}, \cref{fig:dinc10dyna}
-
-Emphasize the information the model provides relative to the observed time series.
-A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
-The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
-Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
-The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
-Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
-The quantile distributions showed similar trends over time in both predicted values and flow-normalized predictions, although some exceptions were observed.
-In particular, high flow (1984, 2008) reduced concentrations of all quantiles but the magniutude of the effect increased at higher quantiles (i.e., the effect was disproportionate).
-The opposite was observed for low flow, i.e., the ninetieth percent showed the greatest increase for low flow.
-
-Emphasize the summer/fall change in the 2000s, why is this?
-Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
-Relate to conceptual diagram.
-
-\subsubsection{Effects of wastewater treatment}
-
-Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:stock}).
-Nitrate is the primary constituent of total nitrogen after 2007.
-Organic nitrogen is a larger percentage of the total after nitrification.
-What was reduction in ammonium starting in 2002?
-
-Nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}), with ammonium showing the largest reduction.
-Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
-Percent changes are shown in \cref{tab:p8chg}, where both nitrogen species shows large percent increases prior to \ac{wwtp} upgrades followed by decreases after upgrades with ammonium showing the largest pecentage.
-Seasonally, increases prior to upgrades were most apparent in the \ac{jas} months for both analytes.
-Seasonal reductions post-upgrades were also largest in \ac{jas} for nitrite/nitrate, whereas percent reductions were similar across all monthly groupings for ammonium.
-
-Relationships of nitrogen with flow showed the typical inverse flow/concentration dynamic with flushing at high flow, although patterns differed by nitrogen species.
-Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations in the winter.
-Additionally, strength of the flow/nutrient relationship changed throughout the time series the year where the strongest relationship differed by analyte.
-Nitrite/nitrate typically had the strongest relationship flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
-
-\subsubsection{Effects of biological invasions}
 
@@
 \centering
-\includegraphics[width=1\textwidth,page=1]{figs/delt_map.pdf}
+\includegraphics[width=0.8\textwidth,page=1]{figs/delt_map.pdf}
 \caption{The \acl{sfe} Delta and monitoring stations used for analysis.
@@
 \end{figure}
+
+<<dinc10, fig.height = 9, fig.width = 7, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
+Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).
+The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.
+Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
+Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
+Subfigure (e) shows the modelled relationships between \\ac{din}, flow, and time.'>>=
+
+# data
+data(h1dat)
+
+# din at c10
+toplo <- filter(h1dat, resvar == 'din') %>% 
+  .$data %>% 
+  .[[1]]
+
+# globals
+ylab <- expression(paste('DIN (mg ', L^-1, ')'))
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(),
+    axis.line.y = element_line(), 
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    legend.position = 'top',    
+    legend.title = element_blank(),
+    axis.text.x = element_blank(),
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
+  )
+col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
+col_lim <- c(0, 3.8)
+xlims <- range(toplo$date)
+txt <- data.frame(
+  x = max(toplo$date) - 300, 
+  y =  c(5, 2.4, 0.6, 7, 7), 
+  txt = c('(a)', '(b)', '(c)', '(d)', '(e)')
+)
+size <- 5
+
+# observed din
+toplo1 <- select(toplo, date, res) %>% 
+  na.omit()
+p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
+  geom_line() + 
+  scale_y_continuous(ylab) +
+  geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
+  scale_x_date(limits = xlims, expand = c(0, 0)) +
+  ptheme
+
+# din prdnrmplot
+p2 <- prdnrmplot(toplo, logspace = F, col_vec = col_vec) + 
+  scale_y_continuous(ylab) +
+  geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
+  scale_x_date(limits = xlims, expand = c(0, 0)) +
+  ptheme
+
+# flo/observed diffs
+toplo2 <- prdnrmplot(toplo, plot = F)
+toplo2 <- left_join(toplo2$fits, toplo2$nrms, by = c('date')) %>% 
+  mutate(flodiff = fits_variable - nrms_variable)
+toplo3 <- toplo
+toplo3$flo <- scales::rescale(toplo3$flo, to = attr(toplo, 'floobs_rng'))
+ylab <- expression(paste('Pred - norm DIN (mg ', L^-1, ')'))
+p3 <- ggplot(toplo2, aes(x = date, y = flodiff)) + 
+  geom_bar(stat = 'identity') + 
+  scale_y_continuous(ylab) +
+  scale_x_date(limits = xlims, expand = c(0, 0)) +
+  geom_text(data = txt[3, ], aes(x = x, y = y, label = txt), size = size) +
+  ptheme
+
+# flo with loess
+ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+p4 <- ggplot(toplo3, aes(x = date, y = flo)) + 
+  geom_line() +
+  stat_smooth(se = F) + 
+  scale_y_continuous(ylab) +
+  scale_x_date(limits = range(toplo3$date), expand = c(0, 0)) +
+  geom_text(data = txt[4, ], aes(x = x, y = y, label = txt), size = size) +
+  ptheme
+
+# dynaplot, all
+txte <- txt[5, , drop = FALSE]
+txte[, 'x'] <- lubridate::decimal_date(txte[, 'x'])
+txte$res <- 0
+attr(toplo, 'reslab') <- expression(paste("DIN (mg ", L^-1, ")"))
+ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+p5 <- gridplot(toplo, month = 'all', logspace = F, col_lim = col_lim, floscl = F) +
+  scale_y_continuous(ylab, expand = c(0, 0)) +
+  geom_text(data = txte, aes(x = x, y = y, label = txt), size = size) +
+  theme(
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
+    )
+
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+pC <- ggplot_gtable(ggplot_build(p3))
+pD <- ggplot_gtable(ggplot_build(p4))
+pE <- ggplot_gtable(ggplot_build(p5))
+maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[2:3], pD$widths[2:3], pE$widths[2:3])
+
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+pC$widths[2:3] <- maxWidth
+pD$widths[2:3] <- maxWidth
+pE$widths[2:3] <- maxWidth
+
+grid.arrange(
+  pA, pB, pC, pD, pE,
+  heights = c(1, 1.3, 1, 1, 1.5),
+  ncol = 1
+  )
+
+@
 
@@
 
-<<dinc10, fig.height = 8, fig.width = 7, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
-Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).
-The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.
-Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
-Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.'>>=
-
-# data
-data(h1dat)
-
-# din at c10
-toplo <- filter(h1dat, resvar == 'din') %>% 
-  .$data %>% 
-  .[[1]]
-
-# globals
-ylab <- expression(paste('DIN (mg ', L^-1, ')'))
-ptheme <- theme_minimal() + 
-  theme(
-    axis.line.x = element_line(),
-    axis.line.y = element_line(), 
-    axis.title.x = element_blank(),
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm") 
-  )
-col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
-col_lim <- c(0, 3.8)
-xlims <- range(toplo$date)
-txt <- data.frame(
-  x = max(toplo$date), 
-  y =  c(5, 2.4, 0.5, 7), 
-  txt = c('(a)', '(b)', '(c)', '(d)')
-)
-size <- 7
-
-# observed din
-toplo1 <- select(toplo, date, res) %>% 
-  na.omit()
-p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
-  geom_line() + 
-  ptheme + 
-  scale_y_continuous(ylab) +
-  theme(
-    axis.text.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-  ) +
-  geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
-  scale_x_date(limits = xlims)
-
-# din prdnrmplot
-p2 <- prdnrmplot(toplo, logspace = F, col_vec = col_vec) + 
-  ptheme + 
-  scale_y_continuous(ylab) +
-  theme(
-    legend.position = 'top', 
-    axis.text.x = element_blank(),
-    legend.title = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-  ) +
-  geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
-  scale_x_date(limits = xlims)
-
-# flo/observed diffs
-toplo2 <- prdnrmplot(toplo, plot = F)
-toplo2 <- left_join(toplo2$fits, toplo2$nrms, by = c('date')) %>% 
-  mutate(flodiff = fits_variable - nrms_variable)
-toplo3 <- toplo
-toplo3$flo <- scales::rescale(toplo3$flo, to = attr(toplo, 'floobs_rng'))
-ylab <- expression(paste('Pred - norm DIN (mg ', L^-1, ')'))
-p3 <- ggplot(toplo2, aes(x = date, y = flodiff)) + 
-  geom_bar(stat = 'identity') + 
-  theme_minimal() +
-  ptheme + 
-  theme(
-    axis.text.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-    ) + 
-  scale_y_continuous(ylab) +
-  scale_x_date(limits = xlims) +
-  geom_text(data = txt[3, ], aes(x = x, y = y, label = txt), size = size)
-
-# flo with loess
-ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
-p4 <- ggplot(toplo3, aes(x = date, y = flo)) + 
-  geom_line() +
-  stat_smooth(se = F) + 
-  ptheme + 
-  scale_y_continuous(ylab) +
-  scale_x_date(limits = range(toplo3$date)) +
-  geom_text(data = txt[4, ], aes(x = x, y = y, label = txt), size = size)
-
-# align widths of plots in first column, first two
-pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p2))
-pC <- ggplot_gtable(ggplot_build(p3))
-pD <- ggplot_gtable(ggplot_build(p4))
-maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[2:3], pD$widths[2:3])
-
-pA$widths[2:3] <- maxWidth
-pB$widths[2:3] <- maxWidth
-pC$widths[2:3] <- maxWidth
-pD$widths[2:3] <- maxWidth
-
-grid.arrange(
-  pA, pB, pC, pD,
-  heights = c(0.9, 1.3, 1, 1.2),
-  ncol = 1
-  )
-
-@
-
-<<dinc10dyna, fig.height = 6, fig.width = 9, cache = T, out.width = '\\textwidth', fig.cap = 'Modelled relationships between \\ac{din}, flow, and time at station C10.
-The top figure shows the annual and seasonal variation over the entire time series and the bottom figure shows annual variation for selected months to remove seasonal variation.
-Warmer colors indicate higher \\ac{din} concentrations.
-The y-axis on the bottom figure is truncated by the fifth and ninety-fifth percentiles of flow within each month.
-Model results are for the fiftieth conditional quantile of \\ac{din}.'>>=
-
-# data
-data(h1dat)
-
-# din at c10
-toplo <- filter(h1dat, resvar == 'din') %>% 
-  .$data %>% 
-  .[[1]]
-
-# globals
+<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
+data(stock_conc)
+
+stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))
+
 ptheme <- theme_minimal() + 
@@
     axis.ticks.length = unit(.1, "cm"),
-    axis.title.x = element_blank()
-  )
-col_lim <- c(0, 3.8)
-txt <- data.frame(
-  x = min(toplo$date), 
-  y =  c(5, 2.4, 0.5, 6), 
-  txt = c('a', 'b', 'c', 'd')
-)
-size <- 7
-
-# dynaplot, all
-col_vec = NULL
-attr(toplo, 'reslab') <- expression(paste("DIN (mg ", L^-1, ")"))
-ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
-p1 <- gridplot(toplo, month = 'all', logspace = F, col_lim = col_lim, floscl = F, col_vec = rev(col_vec)) +
-  scale_y_continuous(ylab, expand = c(0, 0)) + 
-  ptheme + 
-  theme(legend.position = 'top')
-
-# dynaplot selected months
-p2 <- gridplot(toplo, logspace = F, col_lim = col_lim, month = c(1, 4, 7, 10), ncol = 4, floscl = F) + 
-  ptheme + 
-  theme(
-    strip.background = element_blank(),
-    legend.position = 'none', 
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank()
-    ) + 
-  scale_y_continuous(ylab, expand = c(0, 0))
-
-# align widths of plots in first column, first two
-pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p2))
-maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
-
-pA$widths[2:3] <- maxWidth
-pB$widths[2:3] <- maxWidth
-
-grid.arrange(
-  pA, pB,
-  heights = c(1.2, 1),
-  ncol = 1
-  )
-@
-
-<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
-data(stock_conc)
-
-stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))
-
-ptheme <- theme_minimal() + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"),
-    axis.title.x = element_blank()
+    axis.title.x = element_blank(),
+    legend.position = 'top', 
+    legend.title = element_blank()
   )
@@
   scale_y_continuous(ylab1, expand = c(0, 0)) + 
-  ptheme + 
-  theme(
-    legend.position = 'top', 
-    legend.title = element_blank()
-    )
+  ptheme
 

### 2017-01-09 sftrends_manu.tex
annual per chg removed from trndmap
dynaplot to fig dinc10

@@
 
-\subsection{Case studies}
+\section{Results and Discussion}
 
-Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
+\subsection{Observed Data}
+
+\subsection{Trends}
+
+kendall tests - observed
+
+What about flow? \cref{fig:dinc10}, \cref{fig:dinc10dyna}
+
+Emphasize the information the model provides relative to the observed time series.
+A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
+The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
+Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
+The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
+Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
+Emphasize the summer/fall change in the 2000s, why is this?
+Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
+Relate to conceptual diagram.
+
+Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in Ref.~\citenum{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
+\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from Ref.~\citenum{Jabusch16}.
+Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).
+
+Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.
+
+\subsection{Selected examples}
+
+Two stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
 The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends over several years.
 The selected case studies focused on \begin{inparaenum}[1\upshape)]
-\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
 \item effects of wastewater treatment upgrades upstream of P8, and
@@
 %See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
-
-\subsubsection{Effects of biological invasions}
-
-The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \cite{Cohen98}.
-The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \cite{Carlton90,Nichols90,Werner93,Crauder16}.
-In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
-Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}.
-The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
-
-We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
-\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
-\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
-\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
-\end{inparaenum}
-The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
-
-\section{Results and Discussion}
-
-\subsection{Observed Data}
-
-\subsection{Trends}
-
-Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in Ref.~\citenum{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
-\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from Ref.~\citenum{Jabusch16}.
-Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).
-
-Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.
-
-\subsection{Selected examples}
-
-\subsubsection{Disaggregating observed nitrogen time series}
-
-\cref{fig:dinc10}, \cref{fig:dinc10dyna}
-
-Emphasize the information the model provides relative to the observed time series.
-A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
-The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
-Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
-The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
-Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
-The quantile distributions showed similar trends over time in both predicted values and flow-normalized predictions, although some exceptions were observed.
-In particular, high flow (1984, 2008) reduced concentrations of all quantiles but the magniutude of the effect increased at higher quantiles (i.e., the effect was disproportionate).
-The opposite was observed for low flow, i.e., the ninetieth percent showed the greatest increase for low flow.
-
-Emphasize the summer/fall change in the 2000s, why is this?
-Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
-Relate to conceptual diagram.
-
-\subsubsection{Effects of wastewater treatment}
 
@@
 \subsubsection{Effects of biological invasions}
+
+The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \cite{Cohen98}.
+The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \cite{Carlton90,Nichols90,Werner93,Crauder16}.
+In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
+Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}.
+The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
+
+We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
+\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
+\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
+\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
+\end{inparaenum}
+The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figs/obsdat-1}
+{\centering \includegraphics[width=0.9\textwidth]{figs/obsdat-1}
 
@@
 
-\caption{Observed \ac{din} from ten stations in the upper \ac{sfe} Delta.
+\caption{Observed \ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \ac{sfe} Delta.
 Data were collected monthly and evaluated with \ac{wrtds} models from 1976 to 2013.
@@
 
-{\centering \includegraphics[width=\textwidth]{figs/trndmap-1}
+{\centering \includegraphics[width=0.75\textwidth]{figs/dinc10-1}
+
+}
+
+\caption{Time series of \ac{din} and flow at station C10.
+Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} at different conditional quantiles ($\tau$ = 0.1, 0.5, 0.9).
+The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.
+Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
+Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
+Subfigure (e) shows the modelled relationships between \ac{din}, flow, and time.}\label{fig:dinc10}
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=0.7\textwidth]{figs/trndmap-1}
 
@@
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=0.75\textwidth]{figs/dinc10-1}
-
-}
-
-\caption{Time series of \ac{din} and flow at station C10.
-Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} at different conditional quantiles ($\tau$ = 0.1, 0.5, 0.9).
-The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.
-Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
-Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.}\label{fig:dinc10}
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\textwidth]{figs/dinc10dyna-1}
-
-}
-
-\caption{Modelled relationships between \ac{din}, flow, and time at station C10.
-The top figure shows the annual and seasonal variation over the entire time series and the bottom figure shows annual variation for selected months to remove seasonal variation.
-Warmer colors indicate higher \ac{din} concentrations.
-The y-axis on the bottom figure is truncated by the fifth and ninety-fifth percentiles of flow within each month.
-Model results are for the fiftieth conditional quantile of \ac{din}.}\label{fig:dinc10dyna}
 \end{figure}

### 2017-01-09 sftrends_manu.Rnw
stock load fig to supp

@@
 \usepackage{paralist}
-\usepackage{bm}
-\usepackage{fixltx2e}
+\usepackage{setspace}
 \usepackage{showlabels}
@@
      }
+
+% turn this off for submission
+\SectionNumbersOn
 
@@
 
-Environmental conditions are affected by multiple processes acting in a hierarchy across space and time \cite{Oneill89,Levin92}.
-Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.
-Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables.
+Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.
+Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.
 Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
@@
 
-\section{Methods}
+\section{Materials and Methods}
 
@@
 Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
-Representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
 These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.
+Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.
 Data were minimally processed with the exception of averaging replicates that occurred on the same day.
@@
 A total of twenty-seven \ac{wrtds} models were created, one for each nitrogen analyte at each station.
-The functional form of \ac{wrtds} is a simple regression that models the log-transformed response variable as a function of time, flow, and season:
+The functional form of \ac{wrtds} is a simple regression\cite{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:
 
@@
 As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.
-The original \ac{wrtds} method is described in more detail in Ref.~\citenum{Hirsch10}.
 The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method\cite{Beck15}.
@@
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
-The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.
-Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.
+The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.
 Seasonal Kendall tests weere also used to describe trends in the model predictions for the observed data.
 These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
-The analysis was similar to that in Ref.~\citenum{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression.
 Functions in the EnvStats package in R were used for the seasonal Kendall tests \cite{Millard13}.
@@
 
+<<echo = F>>=
+data(delt_dat)
+
+@
+
+The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
+In general, long-term (inter-annual) trends were not easily observed from the raw data.
+
+averages by location
+averages by years, seasons
+seasonality vs no seasonality
+ammonium v nitrite/nitrate ratio, overall and by season
+spikes
+
 \subsection{Trends}
 
-kendall tests - observed
+kendall tests - observed, \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}
 
@@
 
-\section{Discussion}
+\subsection{Implications}
 
@@
 
+\section{References} % omit on submit
+\begin{singlespace}
 \bibliography{refs}
+\end{singlespace}
 
@@
     axis.ticks.length = unit(.1, "cm"), 
-    panel.grid.major = element_blank(), 
+    panel.grid.major = element_line(size = 0.25),
     panel.grid.minor = element_blank()
@@
 
-cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(4, 7)]
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
+  rev
 ylab <- expression(paste('DIN (mg ', L^-1, ')'))
@@
 ggplot(toplo, aes(x = Date)) +
-  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack') + 
+  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
   scale_x_date(expand = c(0, 0)) + 
@@
 
-<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
-data(stock_conc)
-
-stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))
-
-ptheme <- theme_minimal() + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"),
-    axis.title.x = element_blank(),
-    legend.position = 'top', 
-    legend.title = element_blank()
-  )
-cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(3, 5, 8)]
-
-toplo1 <- stock_conc %>% 
-  select(-val2) %>% 
-  spread(var, val) %>% 
-  mutate(TN = NH3 + NO2 + NO3)
-toplo2 <- stock_conc
-
-ylab1 <- expression(paste("Total nitrogen (mg ", L^-1, ")"))
-xrng <- range(c(toplo1$date, toplo2$date))
-yrng <- c(0, 50)
-leglabs <- list(
-  expression(NH [4] ^ '+'),
-  expression(NO [2] ^ '-'),
-  expression(NO [3] ^ '2-')
-  )
-
-lndate <- as.Date('2006/01/01')
-
-p1 <- ggplot(toplo2) + 
-  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack') + 
-  scale_x_date(expand = c(0, 0), limits = xrng) + 
-  scale_fill_manual(values = cols, labels = leglabs) + 
-  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
-  geom_label_repel(
-    data = data.frame(x = lndate, y = 35), 
-    aes(x, y, label = 'WWTP upgrades'), 
-    box.padding = grid::unit(2, "lines"), 
-    nudge_x = 1
-    ) +
-  scale_y_continuous(ylab1, expand = c(0, 0)) + 
-  ptheme
-
-p1
-@
-
 <<p8trnds, fig.height = 8, fig.width = 9, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \\ac{wrtds} (a, bottom), and relationships with flow over time (b).
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County), were completed in May 2007 (\\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.'>>=
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.'>>=
 data(h2dat)
@@
   na.omit()
+lndate <- as.Date('2006/01/01')
 p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
@@
   ) +
-  # geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
+  geom_label_repel(
+    data = data.frame(x = lndate, y = 35), 
+    aes(x, y, label = 'WWTP upgrades'), 
+    box.padding = grid::unit(2, "lines"), 
+    nudge_x = 1
+    ) + 
   scale_x_date(limits = xlims) 
@@
   ) +
-  # geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
+  geom_label_repel(
+    data = data.frame(x = lndate, y = 35), 
+    aes(x, y, label = 'WWTP upgrades'), 
+    box.padding = grid::unit(2, "lines"), 
+    nudge_x = 1
+    ) + 
   scale_x_date(limits = xlims)
@@
   ) +
-  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
   scale_x_date(limits = xlims) 
@@
   ) +
-  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
   scale_x_date(limits = xlims)
@@
   guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10))
-  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
 pleg2 <- g_legend(p5)
@@
     legend.title = element_blank()
-  )# +
-  # geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
+  )
 
@@
 
+<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
+data(stock_conc)
+
+stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))
+
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(), 
+    axis.line.y = element_line(), 
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"),
+    axis.title.x = element_blank(),
+    legend.position = 'top', 
+    legend.title = element_blank()
+  )
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(3, 5, 8)] %>% 
+  rev
+
+toplo1 <- stock_conc %>% 
+  select(-val2) %>% 
+  spread(var, val) %>% 
+  mutate(TN = NH3 + NO2 + NO3)
+toplo2 <- stock_conc
+
+ylab1 <- expression(paste("Total nitrogen (mg ", L^-1, ")"))
+xrng <- range(c(toplo1$date, toplo2$date))
+yrng <- c(0, 50)
+leglabs <- list(
+  expression(NH [4] ^ '+'),
+  expression(NO [2] ^ '-'),
+  expression(NO [3] ^ '2-')
+  )
+
+lndate <- as.Date('2006/01/01')
+
+p1 <- ggplot(toplo2) + 
+  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
+  scale_x_date(expand = c(0, 0), limits = xrng) + 
+  scale_fill_manual(values = cols, labels = leglabs) + 
+  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
+  geom_label_repel(
+    data = data.frame(x = lndate, y = 35), 
+    aes(x, y, label = 'WWTP upgrades'), 
+    box.padding = grid::unit(2, "lines"), 
+    nudge_x = 1
+    ) +
+  scale_y_continuous(ylab1, expand = c(0, 0)) + 
+  ptheme
+
+p1
+@
+
 \end{suppinfo}

### 2017-01-09 sftrends_manu.tex
stock load fig to supp

@@
 
+The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
+In general, long-term (inter-annual) trends were not easily observed from the raw data.
+
+averages by location
+averages by years, seasons
+seasonality vs no seasonality
+ammonium v nitrite/nitrate ratio, overall and by season
+spikes
+
 \subsection{Trends}
 
-kendall tests - observed
+kendall tests - observed, \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}
 
@@
 
-{\centering \includegraphics[width=\textwidth]{figs/stock-1}
-
-}
-
-\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
-\end{figure}
-
-\begin{figure}[!ht]
-
 {\centering \includegraphics[width=\textwidth]{figs/p8trnds-1}
@@
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County), were completed in May 2007 (\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.}\label{fig:p8trnds}
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.}\label{fig:p8trnds}
 \end{figure}
@@
 
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\textwidth]{figs/stock-1}
+
+}
+
+\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
+\end{figure}
+
 \end{suppinfo}

### 2017-01-10 sftrends_manu.Rnw
fixed theme issues in p8trnd

@@
 
+obs_sm <- group_by(delt_dat, Site_Code)
 @
@@
 
-What about flow? \cref{fig:dinc10}, \cref{fig:dinc10dyna}
+What about flow? \cref{fig:dinc10}
 
@@
 Relate to conceptual diagram.
-
-Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in Ref.~\citenum{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change.
-\cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from Ref.~\citenum{Jabusch16}.
-Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).
 
@@
 \includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
-\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
-Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
 Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
@@
     axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm")
-  )
+    axis.ticks.length = unit(.1, "cm"),
+    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
+  )
+
 col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
@@
   na.omit()
-lndate <- as.Date('2006/01/01')
 p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
   geom_line() + 
-  ptheme + 
+  scale_x_date(limits = xlims) +
   scale_y_continuous(ylab1) +
+  ptheme %+replace%
   theme(
     axis.text.x = element_blank(),
-    axis.title.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-  ) +
-  geom_label_repel(
-    data = data.frame(x = lndate, y = 35), 
-    aes(x, y, label = 'WWTP upgrades'), 
-    box.padding = grid::unit(2, "lines"), 
-    nudge_x = 1
-    ) + 
-  scale_x_date(limits = xlims) 
+    axis.title.x = element_blank()
+    )
 p2 <- ggplot(toplo2, aes(x = date, y = exp(res))) + 
   geom_line() +
-  ptheme + 
+  scale_x_date(limits = xlims) +
   scale_y_continuous(ylab2) +
+  ptheme %+replace% 
   theme(
     axis.text.x = element_blank(),
-    axis.title.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-  ) +
-  geom_label_repel(
-    data = data.frame(x = lndate, y = 35), 
-    aes(x, y, label = 'WWTP upgrades'), 
-    box.padding = grid::unit(2, "lines"), 
-    nudge_x = 1
-    ) + 
-  scale_x_date(limits = xlims)
+    axis.title.x = element_blank()
+    )
 
@@
 p3 <- prdnrmplot(nomod, logspace = F, col_vec = col_vec, min_mo = 10) + 
-  ptheme + 
+  scale_x_date(limits = xlims) +
   scale_y_continuous(ylab1) +
+  ptheme %+replace% 
   theme(
@@
     legend.title = element_blank(),
-    axis.title.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-  ) +
-  scale_x_date(limits = xlims) 
+    axis.title.x = element_blank()
+  )
 pleg1 <- g_legend(p3)
-p3 <- p3 + theme(legend.position = 'none')
+p3 <- p3 + 
+  ptheme %+replace% 
+  theme(
+    legend.position = 'none', 
+    axis.title.x = element_blank()
+    )
 p4 <- prdnrmplot(nhmod, logspace = F, col_vec = col_vec, min_mo = 10) + 
-  ptheme + 
+  scale_x_date(limits = xlims) +
   scale_y_continuous(ylab2) +
+  ptheme %+replace% 
   theme(
@@
     legend.title = element_blank(),
-    axis.title.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-  ) +
-  scale_x_date(limits = xlims)
+    axis.title.x = element_blank()
+  )
 
@@
 p5 <- dynaplot(nomod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F) + 
-  ptheme + 
+  scale_x_continuous(flolab) + 
   scale_y_continuous(ylab1) +
+  ptheme %+replace% 
   theme(
@@
   ) +
-  scale_x_continuous(flolab) + 
   guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10))
 pleg2 <- g_legend(p5)
-p5 <- p5 + theme(legend.position = 'none')
+p5 <- p5 + ptheme %+replace% theme(legend.position = 'none')
 p6 <- dynaplot(nhmod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F) + 
-  ptheme + 
+  scale_x_continuous(flolab) +
   scale_y_continuous(ylab2) +
-  scale_x_continuous(flolab) +
+  ptheme %+replace% 
   theme(
@@
 \includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
-\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
-Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
 Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
@@
 \includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
-\caption{Results from seasonal Kendall tests on model predictions from \ac{wrtds}.
-Trends were evaluated using predicted (triangles) and flow-normalized (circles) results to identify potential effects of flow.
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
 Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).

### 2017-01-10 sftrends_manu.Rnw
better reporting of censored data

@@
 
+% censored percentages
+<<echo = F, cache = T>>=
+data(mods)
+cvals <- mutate(mods, 
+  cper = map(data, function(x){
+
+    x <- !na.omit(x$not_cens)
+    round(100 * sum(x)/length(x), 1)
+
+  })) %>% 
+  select(Site_Code, resvar, cper) %>% 
+  unnest %>% 
+  arrange(-cper) %>% 
+  filter(cper > 3)
+@
+
 Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.
@@
 Data were minimally processed with the exception of averaging replicates that occurred on the same day.
-Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).
 The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
+Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.
+The most censored observations were observed for ammonium time series at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\%), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).
 
@@
 
-obs_sm <- group_by(delt_dat, Site_Code)
+data(mods)
+
+obs_sm <- select(delt_dat, Site_Code, Location, Date, nh, no23, din) %>% 
+  gather('var', 'val', nh:din) %>% 
+  group_by(Location, var) %>% 
+  summarise(val = mean(val, na.rm = T))
+
 @
@@
 In general, long-term (inter-annual) trends were not easily observed from the raw data.
+Total \ac{din} for most stations was dominated by nitrite/nitrate whereas ammonium was present at lower concentrations.
+However, C3 had a majority of \ac{din} composed of ammonium and others had larger concentrations ammonium relative to nitrite/nitrate depending on the time of year (e.g., P8, D26).
+By location, observed concentrations decreased on average from the upper Delta (C3, C10, MD10, P8), to the middle (D19, D26, D28), and to the Suisun stations(D4, D6, D7).
 
@@
   mutate(Site_Code = factor(Site_Code, 
-    levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3'))
+    levels = rev(c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'MD10', 'C10', 'C3')))
     )

### 2017-01-10 sftrends_manu.tex
better reporting of censored data

@@
 
+% censored percentages
+
 Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.
@@
 Data were minimally processed with the exception of averaging replicates that occurred on the same day.
-Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).
 The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
+Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.
+The most censored observations were observed for ammonium time series at sites C10 (25.4\%), D28 (17.8\%), D19 (12\%), D7 (7.9\%), and D4 (6.4\%).
 

### 2017-01-10 sftrends_manu.Rnw
observed data results done

@@
 
+% for unit macros
+\newcommand{\mgl}{mg L$^{-1}$}
+\newcommand{\mugl}{$\mu$g L$^{-1}$}
+
 % turn this off for submission
@@
 library(WRTDStidal)
-library(ggplot2)
-library(dplyr)
+library(tidyverse)
 library(gridExtra)
 library(grid)
-library(purrr)
-library(tidyr)
 library(Hmisc)
@@
 library(ggtree)
-library(forcats)
+library(NADA)
 @
@@
 data(mods)
-cvals <- mutate(mods, 
+cvals_all <- mutate(mods, 
   cper = map(data, function(x){
@@
   select(Site_Code, resvar, cper) %>% 
-  unnest %>% 
-  arrange(-cper) %>% 
+  unnest
+
+cvals <- arrange(cvals_all, -cper) %>% 
   filter(cper > 3)
@@
 
-<<echo = F>>=
-data(delt_dat)
-
+% mle summaries of obs data
+<<echo = F, cache = T>>=
 data(mods)
 
-obs_sm <- select(delt_dat, Site_Code, Location, Date, nh, no23, din) %>% 
-  gather('var', 'val', nh:din) %>% 
-  group_by(Location, var) %>% 
-  summarise(val = mean(val, na.rm = T))
+# averages across Location
+obslo <- select(mods, Location, resvar, data) %>% 
+  filter(resvar == 'din') %>% 
+  unnest %>% 
+  select(Location, resvar, res, not_cens) %>% 
+  na.omit %>% 
+  group_by(Location) %>%
+  mutate(res = exp(res)) %>% 
+  nest %>% 
+  mutate(
+    ests = map(data, function(x){
+      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>%   
+        mean %>% 
+        as.list
+    })
+  ) %>% 
+  select(-data) %>% 
+  mutate(ests = map(ests, bind_rows)) %>% 
+  unnest %>% 
+  split(., .$Location)
+
+# averages by site
+obssi <- select(mods, Site_Code, resvar, data) %>% 
+  unnest %>% 
+  select(Site_Code, resvar, res, not_cens) %>% 
+  na.omit %>% 
+  group_by(Site_Code, resvar) %>%
+  mutate(res = exp(res)) %>% 
+  nest %>% 
+  mutate(
+    ests = map(data, function(x){
+      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>% 
+        mean %>% 
+        as.list
+    })
+  ) %>% 
+  select(-data) %>% 
+  mutate(ests = map(ests, bind_rows)) %>% 
+  unnest %>% 
+  group_by(resvar) %>% 
+  filter(mean %in% range(mean)) %>% 
+  arrange(resvar, -mean) %>% 
+  split(., .$resvar)
+
+# averages across years
+obsyr <- select(mods, Site_Code, resvar, data) %>% 
+  unnest %>% 
+  select(Site_Code, resvar, date, res, not_cens) %>% 
+  na.omit %>% 
+  mutate(
+    res = exp(res), 
+    yr = lubridate::year(date), 
+    yr = ifelse(yr < 1996, '1976-1995', '1996-2013') 
+    ) %>% 
+  group_by(resvar, yr) %>% 
+  nest %>% 
+  mutate(
+    ests = map(data, function(x){
+      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>% 
+        mean %>% 
+        as.list
+    })
+  ) %>% 
+  select(-data) %>% 
+  mutate(ests = map(ests, bind_rows)) %>% 
+  unnest %>% 
+  split(., .$resvar)
+
+# averages by season
+obsmo <- select(mods, Site_Code, resvar, data) %>% 
+  unnest %>% 
+  select(Site_Code, resvar, date, res, not_cens) %>% 
+  na.omit %>% 
+  mutate(
+    res = exp(res), 
+    mo = lubridate::month(date), 
+    mo = factor(mo), 
+    mo = fct_collapse(mo, 
+      Spring = c('3', '4', '5'),
+      Summer = c('6', '7', '8'), 
+      Fall = c('9', '10', '11'), 
+      Winter = c('12', '1', '2')
+      )
+    ) %>% 
+  group_by(resvar, mo) %>% 
+  nest %>% 
+  mutate(
+    ests = map(data, function(x){
+      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>% 
+        mean %>% 
+        as.list
+    })
+  ) %>% 
+  select(-data) %>% 
+  mutate(ests = map(ests, bind_rows)) %>% 
+  unnest
 
@@
 In general, long-term (inter-annual) trends were not easily observed from the raw data.
-Total \ac{din} for most stations was dominated by nitrite/nitrate whereas ammonium was present at lower concentrations.
-However, C3 had a majority of \ac{din} composed of ammonium and others had larger concentrations ammonium relative to nitrite/nitrate depending on the time of year (e.g., P8, D26).
-By location, observed concentrations decreased on average from the upper Delta (C3, C10, MD10, P8), to the middle (D19, D26, D28), and to the Suisun stations(D4, D6, D7).
-
-averages by location
-averages by years, seasons
-seasonality vs no seasonality
-ammonium v nitrite/nitrate ratio, overall and by season
-spikes
+Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisuan Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
+Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
+Observed concentrations were also higher later in the time series for all analytes.
+For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}.
+Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.
+However, observed seasonality patterns were inconsistent between sites.
+For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).
 
 \subsection{Trends}
+
+Formal trend analyses of the observed data suggested that
 

### 2017-01-10 sftrends_manu.tex
observed data results done

--- 2017-01-10 sftrends_manu.tex
@@
      }
+
+% for unit macros
+\newcommand{\mgl}{mg L$^{-1}$}
+\newcommand{\mugl}{$\mu$g L$^{-1}$}
 
@@
 
+% mle summaries of obs data
+
 The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
 In general, long-term (inter-annual) trends were not easily observed from the raw data.
-
-averages by location
-averages by years, seasons
-seasonality vs no seasonality
-ammonium v nitrite/nitrate ratio, overall and by season
-spikes
+Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.16$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
+Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
+Observed concentrations were also higher later in the time series for all analytes.
+For example, average \ac{din} across all stations was $0.62$$\pm$$0.01$ \mgl   for 1976-1995, compared to $0.72$$\pm$$0.01$ for 1996-2013.
+Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.
+However, observed seasonality patterns were inconsistent between sites.
+For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).
 
 \subsection{Trends}
+
+Formal trend analyses of the observed data suggested that
 

### 2017-01-10 R/funcs.R

@@ -40,6 +40,50 @@ p_ast <- function(x){
 }
 
 ######
+# scientific notation from R to LaTeX
+# x is numeric to convert
+# pow is minimum exponent (positive or negative) to use for notation
+# digits numeric for rounding
+# showdollar logical if output is enclosed in dollar for latex
+# from https://dankelley.github.io/r/2015/03/22/scinot.html
+scinot <- function(x, pow = 2, digits = 2, showDollar = TRUE)
+{
+  
+  x <- as.numeric(x)
+  
+  # only do this is x is not equal to zero
+  if(x != 0){
+    
+    sign <- ""
+    if (x < 0) {
+        sign <- "-"
+        x <- -x
+    }
+    exponent <- floor(log10(x))
+    if (exponent & abs(exponent) > pow) {
+        xx <- round(x / 10^exponent, digits=digits)
+        e <- paste("\\times 10^{", as.integer(exponent), "}", sep="")
+    } else {
+        xx <- round(x, digits=digits)
+        e <- ""
+    }
+    
+    out <- paste(sign, xx, e, sep="")
+    
+  } else {
+    
+    out <- x
+    
+  }
+  
+  # add dollars if true
+  if (showDollar) out <- paste("$", out, "$", sep="")
+  
+  return(out)
+  
+}
+
+######
 # trend maps
 # 
 # library(dplyr)

### 2017-01-10 sftrends_manu.Rnw
trnd figures in blue and per chng per year

@@
 The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.
-Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.
+Salinity has been used as a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.
 
@@
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
-The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.
-Seasonal Kendall tests weere also used to describe trends in the model predictions for the observed data.
+The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
+Seasonal Kendall tests were also used to describe trends in the model predictions for the observed data.
 These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
@@
 
-Formal trend analyses of the observed data suggested that
+Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years.
+Trends from 1976-1995 and 1996-2013 varied considerably between sites and analytes.
+Overall,
 
@@
 
+# color gradients
+blues <- RColorBrewer::brewer.pal(9, 'Blues')
+collo <- blues[5] #'grey60'
+colhi <- blues[9] #'black'
+
 # combine all
@@
   theme(
-    legend.position = 'top', 
     legend.title = element_blank(),
@@
     panel.grid.major = element_blank(), 
-    panel.grid.minor = element_blank()
-  )
+    panel.grid.minor = element_blank(), 
+    legend.position = 'none'
+  )
+
+# stupid legend
+toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
+pleg <- ggplot(toplo1, aes(x = perchg, y = Site_Code)) + 
+  geom_point(aes(pch = trnd), colour = blues[7], size = 4, alpha = 0.8) + 
+  scale_shape_manual(values = c(1, 16, 2, 17)) +
+  scale_colour_gradient(low = collo, high = colhi) +
+  mythm + 
+  theme(legend.position = 'top')
+pleg <- g_legend(pleg)
 
 # year only
-toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
-p1 <- ggplot(toplo1, aes(x = tau, y = Site_Code)) + 
-  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
-  facet_grid(cat~resvar, labeller = label_parsed
+p1 <- ggplot(toplo1, aes(x = perchg, y = Site_Code)) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') +
+  geom_point(aes(pch = trnd, colour = abs(perchg)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed, scales = 'free_x'
     ) + 
-  geom_vline(xintercept = 0, linetype = 'dashed') + 
   mythm + 
-  xlab(expression(tau)) +
+  xlab(expression('% chg yr' ^-1)) +
   scale_shape_manual(values = c(1, 16, 2, 17)) +
-  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
-  scale_x_continuous(limits = c(-1, 1)) + 
+  scale_colour_gradient(low = collo, high = colhi, guide = F) + 
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
@@
 toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
-p2 <- ggplot(toplo2, aes(x = tau, y = Site_Code)) + 
-  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
-  facet_grid(cat~resvar, labeller = label_parsed
+p2 <- ggplot(toplo2, aes(x = perchg, y = Site_Code)) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') +
+  geom_point(aes(pch = trnd, colour = abs(perchg)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed, scales = 'free_x'
     ) +
-  geom_vline(xintercept = 0, linetype = 'dashed') + 
   mythm + 
-  xlab(expression(tau)) +
+  xlab(expression('% chg yr' ^-1)) +
   scale_shape_manual(values = c(1, 16, 2, 17)) +
-  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
-  scale_x_continuous(limits = c(-1, 1)) + 
+  scale_colour_gradient(low = collo, high = colhi, guide = F) + 
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
@@
 toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
-p3 <- ggplot(toplo3, aes(x = tau, y = Site_Code)) + 
-  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
-  facet_grid(cat~resvar, labeller = label_parsed
+p3 <- ggplot(toplo3, aes(x = perchg, y = Site_Code)) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  geom_point(aes(pch = trnd, colour = abs(perchg)), size = 4, alpha = 0.8) + 
+  facet_grid(cat~resvar, labeller = label_parsed, scales = 'free_x'
     ) + 
-  geom_vline(xintercept = 0, linetype = 'dashed') + 
   mythm + 
-  xlab(expression(tau)) +
+  xlab(expression('% chg yr' ^-1)) +
   scale_shape_manual(values = c(1, 16, 2, 17)) +
-  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
-  scale_x_continuous(limits = c(-1, 1)) + 
+  scale_colour_gradient(low = collo, high = colhi, guide = F) + 
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
@@
 pdf('figs/trndcomp1.pdf', height = 5, width = 9, family = 'serif')
-p1
+grid.arrange(pleg, p1, ncol = 1, heights = c(0.1, 1))
 dev.off()
@@
 pdf('figs/trndcomp2.pdf', height = 8, width = 9, family = 'serif')
-p2
+grid.arrange(pleg, p2, ncol = 1, heights = c(0.06, 1))
 dev.off()
@@
 pdf('figs/trndcomp3.pdf', height = 8, width = 9, family = 'serif')
-p3
+grid.arrange(pleg, p3, ncol = 1, heights = c(0.06, 1))
 dev.off()
@@
 \caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
 Trends are shown separately for different annual groupings.
@@
 <<dinc10, fig.height = 9, fig.width = 7, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
-Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).
+Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} for the conditional mean response.
 The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.
-Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
+Subfigure (c) shows the difference between the model predictions and flow-normalized predictions.
 Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
@@
 \caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
 Trends are shown separately for different seasonal groupings from 1976-1995.
@@
 \caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$).
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
 Trends are shown separately for different seasonal groupings from 1996-2013.

### 2017-01-10 sftrends_manu.tex
trnd figures in blue and per chng per year

@@
 \caption{Time series of \ac{din} and flow at station C10.
-Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} at different conditional quantiles ($\tau$ = 0.1, 0.5, 0.9).
+Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} for the conditional mean response.
 The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.
-Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.
+Subfigure (c) shows the difference between the model predictions and flow-normalized predictions.
 Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.

### 2017-01-11 sftrends_manu.Rnw
mods and trends data files updated with md10
added md10 data to text, combiend d7 clam figs

@@
 A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
-A change in chlorohpyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
+A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
 However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
@@
   mutate(
-    Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')),
+    Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3')),
     ztest = ifelse(ztest < 0.05, '(p < 0.05)', '(ns)'),
@@
 
-<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.
+<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013.
 Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
 Station names are shown in the top left panel.
-Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values.
+Station locations have been jittered to reduce overlap.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 5.3, cache = T, out.width = '0.7\\textwidth'>>=
@@
   spread(trndvar, perchg) %>% 
-  mutate(Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')))
+  mutate(Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3')))
 
@@
 # need to shift locations of some stations
-statmeta[statmeta$Site_Code %in% 'D26', 'Latitude'] <- 38.12
+statmeta[statmeta$Site_Code %in% 'D26', 'Latitude'] <- 38.14
 statmeta[statmeta$Site_Code %in% 'D26', 'Longitude'] <--121.55
@@
 statmeta[statmeta$Site_Code %in% 'C10', 'Longitude'] <--121.28
+statmeta[statmeta$Site_Code %in% 'D28', 'Latitude'] <-37.94
+statmeta[statmeta$Site_Code %in% 'MD10', 'Latitude'] <-38.1
+statmeta[statmeta$Site_Code %in% 'P8', 'Latitude'] <-37.93
+statmeta[statmeta$Site_Code %in% 'D19', 'Latitude'] <-38.06
 
@@
 # bounds on limits
-yrng <- c(-9, 9)
+yrng <- max(abs(range(trnds_seasyr$norm)))
+yrng <- c(-1 * yrng, yrng)
 
@@
       dat <- statmeta[statmeta$Site_Code %in% nms,]
-      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, subwid, 0.148) # change bar heights here
+      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, subwid, 0.21) # change bar heights here
 
@@
 # setup legend
-rngs <- seq(-9, 7, by = 3)
+rngs <- seq(yrng[1], yrng[2], length = length(c(colneg, colpos)))
 leg <- ggplot(data.frame(x = seq_along(rngs), y = seq_along(rngs), z = rngs), aes(x, y, colour = z)) + 
@@
 # dynaplots
-p5 <- dynaplot(nomod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F) + 
+p5 <- dynaplot(nomod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F, size = 0.5) + 
   scale_x_continuous(flolab) + 
@@
 p5 <- p5 + ptheme %+replace% theme(legend.position = 'none')
-p6 <- dynaplot(nhmod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F) + 
+p6 <- dynaplot(nhmod, logspace = F, col_vec = dynacol_vec, month = c(1, 4, 7, 10), floscl = F, size = 0.5) + 
   scale_x_continuous(flolab) +
@@
 
-<<clmchl, fig.height = 7, fig.width = 8, out.width = '\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
-Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means).
-A coincident decrease in \\ac{chla} concentration was also observed (c).
-A weak but significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (b).
-Flow relationships with \\ac{chla} concentration have also changed over time (d, observations from June).
-Chlorophyll shows a slight positive then dominantly negative association with increasing flow (decreasing salinity) early in the time series, whereas the trend is reversed in recent years.'>>=
-
+<<clmchl, fig.height = 9, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
+Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means),with biomass linked to salinity (b).
+A coincident decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d).
+A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (e).
+Flow relationships with \\ac{chla} concentration have also changed over time (f, observations from June).'>>=
 # clam and model data at D7
@@
 data(flow_dat)
-chld7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'chl')$data[[1]]
+chld7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'chl')$mod[[1]]
 
@@
 
-# subfig labs
-txt <- data.frame(
-  x = c(as.numeric(as.Date('1979-10-01')), 2500, 25, 0.25),
-  y =  c(10000, 14, 4, 3.3), 
-  txt = c('(a)', '(c)', '(b)', '(d)')
-)
-size <- 6
-
-##
-# annual clam time series
-p1 <- ggplot(clmdat, aes(x = yr, y = clams_smp, fill = species)) + 
-  geom_bar(stat = 'identity') + 
-  theme_minimal() + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"), 
-    # axis.title.x = element_blank(),
-    legend.position = 'top', 
-    legend.title = element_blank()
-    ) + 
-  scale_fill_brewer(
-    palette = 'Set2',
-    labels = parse(text = c('italic(Corbicula)', 'italic(Potamocorbula)'))
-    ) + 
-  scale_x_date('Year', limits = xlims) +
-  scale_y_continuous('Clams per sample') +
-  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5)) +
-  with(txt[1, ], annotate('text', x = xlims[1] + 365 * 2, y = y, label = txt, size = size))
-
-##
-# clam biomass related to daily sacramento flow
-clmdat <- group_by(clams, yr, species) %>% 
-  dplyr::summarize(
-    clams_smp = mean(clams_smp, na.rm = TRUE),
-    biomass = mean(biomass, na.rm = TRUE)
-    ) %>% 
-  ungroup %>% 
-  mutate(yr = as.Date(paste0(yr, '-10-01')))
-
-flodat <- filter(flow_dat, station %in% 'sac')
-clmflo <- select(clams, biomass, clams_smp, species, date) %>% 
-  rename(Date = date) %>% 
-  left_join(., flodat, by = 'Date') %>% 
-  filter(species %in% 'Potamocorbula')
-
-# linear model of concentration x biomass
-mod <- lm(log(1 + biomass) ~ log(1 + q), data = clmflo) %>% 
-  coefficients
-xvals <- range(clmflo$q, na.rm = TRUE)
-xvals <- seq(xvals[1], xvals[2], length = 100)
-yvals <- (exp(mod[1] + mod[2] * log(1 + xvals))) - 1
-
-ylab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
-xlab <- expression(paste("Sacramento daily flow (", m^3, ' ', s^-1, ")"))
-
-# scatter plot of concentration x biomass with model 
-p2 <- ggplot(clmflo, aes(x = q, y = biomass)) + 
-  geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
-  geom_line(data = data.frame(x = xvals, y = yvals), aes(x, y, colour = 'black'), lwd = 1) +
-  theme_minimal() + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"), 
-    legend.position = 'top', 
-    legend.title = element_blank(), 
-    legend.box = 'horizontal'
-    ) +
-  scale_colour_manual(label = 'log(1 + biomass) = 5.2 - 0.6log(1 + flow)', values = 'black') +
-  scale_x_continuous(xlab) + 
-  scale_y_continuous(ylab, limits = c(0,15)) +
-  geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
-
-##
-# chlorophyll by month
-chldat <- filter(h3dat, resvar %in% 'chl' & Site_Code %in% 'D7') %>% 
-  .$data %>% 
-  .[[1]] %>% 
-  select(year, month, res, bt_norm) %>% 
-  mutate(res = exp(res)) %>% 
-  rename(yr = year, mo = month, bt_res = res) %>% 
-  group_by(yr, mo) %>% 
-  dplyr::summarize(res = mean(bt_res, na.rm = TRUE), bt_norm = mean(bt_norm, na.rm = TRUE)) %>% 
-  ungroup
-
-# clam by month
-clmdat <- mutate(clams, mo = lubridate::month(date)) %>% 
-  group_by(yr, mo, species) %>% 
-  dplyr::summarize(
-    clams_smp = mean(clams_smp, na.rm = TRUE),
-    biomass = mean(biomass, na.rm = TRUE)
-    ) %>% 
-  ungroup %>% 
-  filter(species %in% 'Potamocorbula')
-
-# combine chlorophyll and clam by year, month
-toplo <- left_join(clmdat, chldat, by = c('yr', 'mo')) %>% 
-  mutate(date = as.Date(paste0(yr, '-', mo, '-01')))
-
-# linear model of concentration x biomass
-mod <- lm(log(1 + bt_norm) ~ log(1 + biomass), data = toplo) %>% 
-  coefficients
-xvals <- range(toplo$biomass, na.rm = TRUE)
-xvals <- seq(xvals[1], xvals[2], length = 100)
-yvals <- (exp(mod[1] + mod[2] * log(1 + xvals))) - 1
-
-xlab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
-ylab <- expression(paste("Flow-normalized chlorophyll (", italic(mu), "g ", L^-1, ")"))
-
-# scatter plot of concetration x biomass with model 
-p3 <- ggplot(toplo, aes(x = biomass, y = bt_norm)) + 
-  geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
-  geom_line(data = data.frame(x = xvals, y = yvals), aes(x, y, colour = 'black'), lwd = 1) +
-  theme_minimal() + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"),
-    legend.position = 'top', 
-    legend.title = element_blank(), 
-    legend.box = 'horizontal'
-    ) +
-  scale_colour_manual(label = 'log(1 + chl) = 1.35 - 0.1log(1 + biomass)', values = 'black') +
-  scale_x_continuous(xlab) + 
-  scale_y_continuous(ylab, limits = c(0.8, 4)) +
-  geom_text(data = txt[3, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size)
-
-##
-# dynaplot for chl, june
+# colors 
 cols <- RColorBrewer::brewer.pal(10, 'Set2')[1:2]
-ylab <- attr(chld7, 'reslab')
-
-p4 <- dynaplot(chld7, mo = 6, col_vec = cols, floscl = F) +
-  theme_minimal() +
+
+# theme
+pmar <- grid::unit(c(0, 6, 0, 6), 'pt')
+thjust <- 0; tvjust <- -10
+mythm <- theme_minimal() + 
   theme(
@@
     legend.box = 'horizontal',    
-    strip.background = element_blank(),
-    strip.text.x = element_blank()
+    plot.title = element_text(hjust = thjust, vjust = tvjust),
+    plot.margin = pmar
+    )
+
+##
+# annual clam time series
+p1 <- ggplot(clmdat, aes(x = yr, y = clams_smp, fill = species)) + 
+  geom_bar(stat = 'identity') + 
+  mythm +
+  scale_fill_brewer(
+    palette = 'Set2',
+    labels = parse(text = c('italic(Corbicula)', 'italic(Potamocorbula)'))
     ) + 
-  scale_x_continuous('ln-Salinity (psu)') + 
-  scale_y_continuous(ylab, limits = c(0.4, 3.3)) +
-  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) + 
-  annotate('text', x = txt[4, 'x'], y = txt[4, 'y'], label = txt[4, 'txt'], size = size)
-
-# align widths of plots in first column, first two
-pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p2))
-maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3])
-
-# Set the widths
-pA$widths[2:3] <- maxWidth
-pB$widths[2:3] <- maxWidth
-
-# align widths of plots in first column, first two
-pC <- ggplot_gtable(ggplot_build(p3))
-pD <- ggplot_gtable(ggplot_build(p4))
-maxWidth = unit.pmax(pC$widths[2:3], pD$widths[2:3])
-
-# Set the widths
-pC$widths[2:3] <- maxWidth
-pD$widths[2:3] <- maxWidth
-
-# combine all
-grid.arrange(
-  arrangeGrob(pA, pB, ncol = 1, heights = c(1, 1)), 
-  arrangeGrob(pC, pD, ncol = 1, heights = c(1.05, 1)), 
-  ncol = 2, widths = c(1, 1)
-  )
-@
-
-<<d7c10trnds, fig.cap = 'Flow-normalized trends of annual (left) and seasonal (right) variation in DIN, chl-{\\it\\footnotesize a}, and SiO$_2$ at station D7 (top) and C10 (bottom).
-Covariation between nutrients, chl-{\\it\\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.
-Seasonal changes at D7 are most pronounced during the summer.', fig.width = 9, fig.height = 10, cache = T>>=
-data(h3dat)
-
-# separate each site
-toplo1 <- filter(h3dat, Site_Code %in% 'D7')
-toplo2 <- filter(h3dat, Site_Code %in% 'C10')
-
-# y axis labels
-ylab1 <- expression(paste("DIN (mg ", L^-1, ")"))
-ylab2 <- expression(paste("Chl-", italic(a), " (", italic(mu), "g ", L^-1, ")"))
-ylab3 <- expression(paste("Si", O[2], " (mg ", L^-1, ")"))
-
-# generic theme for all plots
-ptheme <- theme_minimal() + 
+  scale_x_date('Year', limits = xlims) +
+  scale_y_continuous('Clams per sample') +
+  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5)) +
+  labs(title = '(a)')
+
+##
+# clam biomass related to daily sacramento flow
+clmdat <- group_by(clams, yr, species) %>% 
+  dplyr::summarize(
+    clams_smp = mean(clams_smp, na.rm = TRUE),
+    biomass = mean(biomass, na.rm = TRUE)
+    ) %>% 
+  ungroup %>% 
+  mutate(yr = as.Date(paste0(yr, '-10-01')))
+
+flodat <- filter(flow_dat, station %in% 'sac')
+clmflo <- select(clams, biomass, clams_smp, species, date) %>% 
+  rename(Date = date) %>% 
+  left_join(., flodat, by = 'Date') %>% 
+  filter(species %in% 'Potamocorbula')
+
+# linear model of concentration x biomass
+mod <- lm(log(1 + biomass) ~ log(1 + q), data = clmflo) %>% 
+  coefficients
+xvals <- range(clmflo$q, na.rm = TRUE)
+xvals <- seq(xvals[1], xvals[2], length = 100)
+yvals <- (exp(mod[1] + mod[2] * log(1 + xvals))) - 1
+
+ylab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
+xlab <- expression(paste("Sacramento daily flow (", m^3, ' ', s^-1, ")"))
+
+# scatter plot of concentration x biomass with model 
+p2 <- ggplot(clmflo, aes(x = q, y = biomass)) + 
+  geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
+  geom_line(data = data.frame(x = xvals, y = yvals), aes(x, y, colour = 'black'), lwd = 1) +
+  mythm + 
+  scale_colour_manual(label = 'log(1 + biomass) = 5.2 - 0.6log(1 + flow)', values = 'black') +
+  scale_x_continuous(xlab) + 
+  scale_y_continuous(ylab, limits = c(0,15)) +
+  labs(title = '(b)')
+
+# prdnrmplot
+ylab <- expression(paste("ln-chlorophyll (", italic(mu), "g ", L^-1, ")"))
+
+p3 <- prdnrmplot(chld7, predicted = F, col_vec = cols, min_mo = 10) + 
+  mythm + 
+  scale_x_date('Year') +
+  scale_y_continuous(ylab) +
+  labs(title = '(c)')
+
+# seasyrplot
+p4 <- seasyrplot(chld7, predicted = F, col_vec = cols, min_mo = 10) + 
   theme(
@@
     axis.ticks.length = unit(.1, "cm"), 
-    axis.title.x = element_blank(), 
-    axis.text.x = element_blank()
-  )
-
-# colors
-cols <- RColorBrewer::brewer.pal(9, 'Set1')[1:3]
-col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
-
-## first set of plots
-d7din<- filter(toplo1, resvar == 'din') %>% 
-  .$data %>% 
-  .[[1]]
-pdin1 <- prdnrmplot(d7din, col_vec = col_vec, logspace = F, min_mo = 10) +
-  ptheme + 
-  scale_y_continuous(ylab1) +
-  theme(legend.position = 'top', legend.title = element_blank())
-pleg1 <- g_legend(pdin1)
-pdin1 <- pdin1 + theme(legend.position = 'none')
-
-pdinsy1 <- seasyrplot(d7din, col_vec = cols, logspace = F, predicted = F) +
-  ptheme + 
-  theme(legend.position = 'top', axis.title.y = element_blank()) + 
-  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10))
-pleg2 <- g_legend(pdinsy1)
-pdinsy1 <- pdinsy1 + theme(legend.position = 'none')
-
-d7chl<- filter(toplo1, resvar == 'chl') %>% 
-  .$data %>% 
-  .[[1]]
-pchl1 <- prdnrmplot(d7chl, col_vec = col_vec, logspace = F, min_mo = 10) +
-  ptheme + 
-  theme(legend.position = 'none') + 
-  scale_y_continuous(ylab2)
-
-pchlsy1 <- seasyrplot(d7chl, col_vec = cols, logspace = F, predicted = F) +
-  ptheme + 
-  theme(legend.position = 'none', axis.title.y = element_blank())
-
-d7sio2 <- filter(toplo1, resvar == 'sio2') %>% 
-  .$data %>% 
-  .[[1]]
-psio1 <- prdnrmplot(d7sio2, col_vec = col_vec, logspace = F, min_mo = 10) +
-  ptheme + 
-  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text()) +
-  scale_x_date('Year') +
-  scale_y_continuous(ylab3)
-
-psiosy1 <- seasyrplot(d7sio2, col_vec = cols, logspace = F, predicted = F) + 
-  ptheme + 
-  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text(), axis.title.y = element_blank()) +
-  scale_x_date('Day of year', date_labels = "%m/%d")
+    legend.title = element_blank(), 
+    panel.border = element_blank(), 
+    axis.title.x = element_text(),
+    plot.title = element_text(hjust = thjust, vjust = tvjust),
+    plot.margin = pmar
+    ) + 
+  scale_x_date('Day of year', date_labels = "%m/%d") +
+  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) +
+  scale_y_continuous(ylab) + 
+  labs(title = '(d)')
+
+##
+# chlorophyll by month
+chldat <- filter(h3dat, resvar %in% 'chl' & Site_Code %in% 'D7') %>% 
+  .$mod %>% 
+  .[[1]] %>% 
+  select(year, month, res, bt_norm) %>% 
+  mutate(res = exp(res)) %>% 
+  rename(yr = year, mo = month, bt_res = res) %>% 
+  group_by(yr, mo) %>% 
+  dplyr::summarize(res = mean(bt_res, na.rm = TRUE), bt_norm = mean(bt_norm, na.rm = TRUE)) %>% 
+  ungroup
+
+# clam by month
+clmdat <- mutate(clams, mo = lubridate::month(date)) %>% 
+  group_by(yr, mo, species) %>% 
+  dplyr::summarize(
+    clams_smp = mean(clams_smp, na.rm = TRUE),
+    biomass = mean(biomass, na.rm = TRUE)
+    ) %>% 
+  ungroup %>% 
+  filter(species %in% 'Potamocorbula')
+
+# combine chlorophyll and clam by year, month
+toplo <- left_join(clmdat, chldat, by = c('yr', 'mo')) %>% 
+  mutate(date = as.Date(paste0(yr, '-', mo, '-01')))
+
+# linear model of concentration x biomass
+mod <- lm(log(1 + bt_norm) ~ log(1 + biomass), data = toplo) %>% 
+  coefficients
+xvals <- range(toplo$biomass, na.rm = TRUE)
+xvals <- seq(xvals[1], xvals[2], length = 100)
+yvals <- (exp(mod[1] + mod[2] * log(1 + xvals))) - 1
+
+xlab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
+
+# scatter plot of concentration x biomass with model 
+p5 <- ggplot(toplo, aes(x = biomass, y = bt_norm)) + 
+  geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
+  geom_line(data = data.frame(x = xvals, y = yvals), aes(x, y, colour = 'black'), lwd = 1) +
+  mythm + 
+  scale_colour_manual(label = 'log(1 + chl) = 1.35 - 0.1log(1 + biomass)', values = 'black') +
+  scale_x_continuous(xlab) + 
+  scale_y_continuous(ylab, limits = c(0.8, 4)) +
+  labs(title = '(e)')
+
+##
+# dynaplot for chl, june
+p6 <- dynaplot(chld7, mo = 6, col_vec = cols, floscl = F, size = 0.5) +
+  mythm %+replace%
+  theme(
+    strip.background = element_blank(),
+    strip.text.x = element_blank()
+    ) + 
+  scale_x_continuous('Salinity (psu)') + 
+  scale_y_continuous(ylab, limits = c(0.4, 3.3)) +
+  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) + 
+  labs(title = '(f)')
+
+##
+# dimension formatting
+
+# align widths of plots in first column
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p3))
+pC <- ggplot_gtable(ggplot_build(p5))
+maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[c(2:3)])
 
 # Set the widths
-pA1 <- ggplot_gtable(ggplot_build(pdin1))
-pB1 <- ggplot_gtable(ggplot_build(pchl1))
-pC1 <- ggplot_gtable(ggplot_build(psio1))
-maxWidth = unit.pmax(pA1$widths[2:3], pB1$widths[2:3], pC1$widths[2:3])
-pA1$widths[2:3] <- maxWidth
-pB1$widths[2:3] <- maxWidth
-pC1$widths[2:3] <- maxWidth
-
-# Set the widths for seasyrplots
-pAsy1 <- ggplot_gtable(ggplot_build(pdinsy1))
-pBsy1 <- ggplot_gtable(ggplot_build(pchlsy1))
-pCsy1 <- ggplot_gtable(ggplot_build(psiosy1))
-maxWidth = unit.pmax(pAsy1$widths[2:3], pBsy1$widths[2:3], pCsy1$widths[2:3])
-pAsy1$widths[2:3] <- maxWidth
-pBsy1$widths[2:3] <- maxWidth
-pCsy1$widths[2:3] <- maxWidth
-
-## second set of plots
-c10din <- filter(toplo2, resvar == 'din') %>% 
-  .$data %>% 
-  .[[1]]
-pdin2 <- prdnrmplot(c10din, col_vec = col_vec, logspace = F, min_mo = 10) +
-  ptheme + 
-  scale_y_continuous(ylab1) +
-  theme(legend.position = 'none')
-
-pdinsy2 <- seasyrplot(c10din, col_vec = cols, logspace = F, predicted = F) +
-  ptheme + 
-  theme(legend.position = 'none', axis.title.y = element_blank())
-
-c10chl <- filter(toplo2, resvar == 'chl') %>% 
-  .$data %>% 
-  .[[1]]
-pchl2 <- prdnrmplot(c10chl, col_vec = col_vec, logspace = F, min_mo = 10) +
-  ptheme + 
-  theme(legend.position = 'none') + 
-  scale_y_continuous(ylab2)
-
-pchlsy2 <- seasyrplot(c10chl, col_vec = cols, logspace = F, predicted = F) +
-  ptheme + 
-  theme(legend.position = 'none', axis.title.y = element_blank())
-
-c10sio2 <- filter(toplo2, resvar == 'sio2') %>% 
-  .$data %>% 
-  .[[1]]
-psio2 <- prdnrmplot(c10sio2, col_vec = col_vec, logspace = F, min_mo = 10) +
-  ptheme + 
-  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text()) +
-  scale_x_date('Year') +
-  scale_y_continuous(ylab3)
-
-psiosy2 <- seasyrplot(c10sio2, col_vec = cols, logspace = F, predicted = F) + 
-  ptheme + 
-  theme(legend.position = 'none', axis.title.x = element_text(), axis.text.x = element_text(), axis.title.y = element_blank()) +
-  scale_x_date('Day of year', date_labels = "%m/%d")
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+pC$widths[2:3] <- maxWidth
+
+# align widths of plots in second column
+pD <- ggplot_gtable(ggplot_build(p2))
+pE <- ggplot_gtable(ggplot_build(p4))
+pF <- ggplot_gtable(ggplot_build(p6))
+maxWidth = unit.pmax(pD$widths[2:3], pE$widths[2:3], pF$widths[2:3])
 
 # Set the widths
-pA2 <- ggplot_gtable(ggplot_build(pdin2))
-pB2 <- ggplot_gtable(ggplot_build(pchl2))
-pC2 <- ggplot_gtable(ggplot_build(psio2))
-maxWidth = unit.pmax(pA2$widths[2:3], pB2$widths[2:3], pC2$widths[2:3])
-pA2$widths[2:3] <- maxWidth
-pB2$widths[2:3] <- maxWidth
-pC2$widths[2:3] <- maxWidth
-
-# Set the widths for seasyrplots
-pAsy2 <- ggplot_gtable(ggplot_build(pdinsy2))
-pBsy2 <- ggplot_gtable(ggplot_build(pchlsy2))
-pCsy2 <- ggplot_gtable(ggplot_build(psiosy2))
-maxWidth = unit.pmax(pAsy2$widths[2:3], pBsy2$widths[2:3], pCsy2$widths[2:3])
-pAsy2$widths[2:3] <- maxWidth
-pBsy2$widths[2:3] <- maxWidth
-pCsy2$widths[2:3] <- maxWidth
-
+pD$widths[2:3] <- maxWidth
+pE$widths[2:3] <- maxWidth
+pF$widths[2:3] <- maxWidth
+
+# combine all
 grid.arrange(
-  arrangeGrob(
-    ncol = 1, heights = c(0.05, 1),
-    arrangeGrob(pleg1, pleg2, ncol = 2),
-    arrangeGrob(
-      arrangeGrob(
-        textGrob('(a) Station D7, Suisun Bay'), heights = c(0.1, 1), 
-        arrangeGrob(
-          arrangeGrob(pA1, pB1, pC1, ncol = 1, heights = c(0.85, 0.85, 1)),
-          arrangeGrob(pAsy1, pBsy1, pCsy1, ncol = 1, heights = c(0.85, 0.85, 1)),
-          ncol = 2
-        )
-      ),
-      arrangeGrob(
-        textGrob('(b) Station C10, Delta'), heights = c(0.1, 1),
-        arrangeGrob(
-          arrangeGrob(pA2, pB2, pC2, ncol = 1, heights = c(0.85, 0.85, 1)),
-          arrangeGrob(pAsy2, pBsy2, pCsy2, ncol = 1, heights = c(0.85, 0.85, 1)),
-          ncol = 2
-        )
-      ),
-      ncol = 1 
-    )
-  )
-)
-
+  arrangeGrob(pA, pB, pC, ncol = 1, heights = c(1, 1, 1)), 
+  arrangeGrob(pD, pE, pF, ncol = 1, heights = c(1, 1, 1)), 
+  ncol = 2, widths = c(1, 1)
+  )
 @
@@
   rgroup = vars,
-  n.rgroup = c(9, 9, 9),
+  n.rgroup = rep(nrow(tab)/3, 3),
   cgroup = c('Annual'),
@@
   rgroup = vars,
-  n.rgroup = c(9, 9, 9),
+  n.rgroup = rep(nrow(tab)/3, 3),
   cgroup = c('Seasonal, 1976-1995'),
@@
   rgroup = vars,
-  n.rgroup = c(9, 9, 9),
+  n.rgroup = rep(nrow(tab)/3, 3),
   cgroup = c('Seasonal, 1996-2013'),
@@
 
-% trends in din, chl, sio2 at c10
-<<c10chg>>=
-
-data(h3dat)
-
-c10dat <- filter(h3dat, Site_Code %in% 'C10')
-
-dinraw <- filter(c10dat, resvar %in% 'din') %>% 
-  .$data %>% 
-  .[[1]]
-chlraw <- filter(c10dat, resvar %in% 'chl') %>% 
-  .$data %>% 
-  .[[1]]
-sioraw <- filter(c10dat, resvar %in% 'sio2') %>% 
-  .$data %>% 
-  .[[1]]
-
-mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
-yrbrks <- c(1975, 1985, 1994, 2003, 2013)
-molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
-yrlabs <- c('1976-1985', '1986-1994', '1995-2003', '2004-2013')
-
-# summarize din, sio2, chl trends
-din <- wrtdstrnd_sk(dinraw, mobrks, yrbrks, molabs, yrlabs)
-din <- wrtdstrnd_sk(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
-  .[1, ] %>% 
-  rbind(., din) %>% 
-  data.frame(res = 'din', .)
-chl <- wrtdstrnd_sk(chlraw, mobrks, yrbrks, molabs, yrlabs)
-chl <- wrtdstrnd_sk(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
-  .[1, ] %>% 
-  rbind(., chl) %>% 
-  data.frame(res = 'chl', .)
-sio <- wrtdstrnd_sk(sioraw, mobrks, yrbrks, molabs, yrlabs)
-sio <- wrtdstrnd_sk(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
-  .[1, ] %>% 
-  rbind(., sio) %>% 
-  data.frame(res = 'sio', .)
-
-# combine for trends for table
-totab <- rbind(din, chl, sio) %>% 
-  select(res, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      perchg, 
-      paste0('\\textit{\\textbf{', perchg, '}}')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  gather('var', 'val', med:perchg) %>% 
-  unite('col', res, var) %>% 
-  spread(col, val) %>% 
-  mutate(cat = factor(cat, levels = c('1976-2013', yrlabs, molabs), labels = c('1976-2013', yrlabs, molabs))) %>% 
-  arrange(cat) %>% 
-  data.frame(sep = c(rep('Annual', 5), rep('Seasonal', 4)), .) %>% 
-  select(sep, cat, din_med, din_perchg, chl_med, chl_perchg, sio_med, sio_perchg)
-
-# final table formatting
-segs <- c('All', 'Annual', 'Seasonal')
-cats <- totab$cat
-totab <- totab[,-c(1:2)]
-vars <- c('DIN', 'Chl-\\textit{a}', 'SiO$_2$')
-names(totab) <- rep(c('Median', '\\% change'), 3)
-
-cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station C10.
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-
-latex( 
-  totab,
-  file = '',
-  rowlabel = 'Period',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = segs,
-  n.rgroup = c(1, rep(4, 2)),
-  cgroup = vars,
-  n.cgroup = c(2, 2, 2),
-  rowname = cats,
-  label = 'tab:c10chg', 
-  insert.bottom = foot.val
-  )
-
-@
-
 \clearpage

### 2017-01-11 sftrends_manu.tex
added md10 data to text, combiend d7 clam figs

@@
 
-\caption{Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.
+\caption[Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013]{Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013.
 Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
 Station names are shown in the top left panel.
-Values less than -9\% change were truncated for visual clarity of the bar plots, see \cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values.
+Station locations have been jittered to reduce overlap.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}
@@
 
-{\centering \includegraphics[width=\textwidth]{figs/clmchl-1}
+{\centering \includegraphics[width=0.9\textwidth]{figs/clmchl-1}
 
@@
 
-\caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2014 at station D7 in Suisun Bay.
-Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means).
-A coincident decrease in \ac{chla} concentration was also observed (c).
-A weak but significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (b).
-Flow relationships with \ac{chla} concentration have also changed over time (d, observations from June).
-Chlorophyll shows a slight positive then dominantly negative association with increasing flow (decreasing salinity) early in the time series, whereas the trend is reversed in recent years.}\label{fig:clmchl}
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figs/d7c10trnds-1}
-
-}
-
-\caption{Flow-normalized trends of annual (left) and seasonal (right) variation in DIN, chl-{\it\footnotesize a}, and SiO$_2$ at station D7 (top) and C10 (bottom).
-Covariation between nutrients, chl-{\it\footnotesize a}, and SiO$_2$ is observed at D7 but not C10, although an overall decrease in SiO$_2$ at C10 is shown.
-Seasonal changes at D7 are most pronounced during the summer.}\label{fig:d7c10trnds}
+\caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
+Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means),with biomass linked to salinity (b).
+A coincident decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d).
+A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (e).
+Flow relationships with \ac{chla} concentration have also changed over time (f, observations from June).}\label{fig:clmchl}
 \end{figure}
@@
 % overall trends, years
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann", insert.bottom = foot.val)%
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann", insert.bottom = foot.val)%
 \begin{table}[!tbp]
@@
 ~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}\tabularnewline
+~~MD10&0.4 \footnotesize{(-1.1)**}&0.3 \footnotesize{(-2.4)**}\tabularnewline
 ~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}\tabularnewline
@@
 ~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}\tabularnewline
+~~MD10&0.1 \footnotesize{(-2.8)**}&0 \footnotesize{(-1.1)**}\tabularnewline
 ~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}\tabularnewline
@@
 ~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}\tabularnewline
+~~MD10&0.4 \footnotesize{(-1)**}&0.3 \footnotesize{(-2.5)**}\tabularnewline
 ~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}\tabularnewline
@@
 % overall trends, seasons, first annual periodyears
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef", insert.bottom = foot.val)%
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef", insert.bottom = foot.val)%
 \begin{table}[!tbp]
@@
 ~~D7&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.6 \footnotesize{(-2.4)**}\tabularnewline
+~~MD10&0.6 \footnotesize{(-0.3)}&0.2 \footnotesize{(-3.6)**}&0.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.3 \footnotesize{(-0.3)*}\tabularnewline
 ~~P8&1.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.9 \footnotesize{(\textit{\textbf{2.4}})**}&1.3 \footnotesize{(\textit{\textbf{3.1}})**}&1.9 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
@@
 ~~D7&0.1 \footnotesize{(\textit{\textbf{3.3}})**}&0 \footnotesize{(\textit{\textbf{2}})**}&0.1 \footnotesize{(-2.8)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
+~~MD10&0.1 \footnotesize{(-1.8)**}&0 \footnotesize{(-6.5)**}&0 \footnotesize{(-3.3)**}&0.2 \footnotesize{(\textit{\textbf{0.4}})}\tabularnewline
 ~~P8&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.1 \footnotesize{(\textit{\textbf{1.8}})**}&0.2 \footnotesize{(\textit{\textbf{7}})**}&0.6 \footnotesize{(\textit{\textbf{7}})**}\tabularnewline
@@
 ~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textit{\textbf{3.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.4 \footnotesize{(-3.6)**}\tabularnewline
+~~MD10&0.5 \footnotesize{(-0.2)}&0.2 \footnotesize{(-3.6)**}&0.2 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(-0.5)*}\tabularnewline
 ~~P8&1.2 \footnotesize{(\textit{\textbf{2}})**}&0.9 \footnotesize{(\textit{\textbf{2.3}})**}&1.1 \footnotesize{(\textit{\textbf{2}})**}&1.4 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
@@
 % overall trends, seasons, first annual periodyears
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = c(9,         9, 9), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft", insert.bottom = foot.val)%
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft", insert.bottom = foot.val)%
 \begin{table}[!tbp]
@@
 ~~D7&0.5 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-1.3)**}&0.4 \footnotesize{(-0.4)**}&0.6 \footnotesize{(-0.2)}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.3)**}&0.2 \footnotesize{(-3.7)**}&0.2 \footnotesize{(-4.4)**}&1 \footnotesize{(-1.8)**}\tabularnewline
 ~~P8&1.5 \footnotesize{(-1.9)**}&1.2 \footnotesize{(-3.5)**}&1.8 \footnotesize{(-2.4)**}&2.7 \footnotesize{(-2.2)**}\tabularnewline
@@
 ~~D7&0.1 \footnotesize{(-2.2)**}&0 \footnotesize{(-2.1)**}&0.1 \footnotesize{(\textit{\textbf{1.3}})**}&0.1 \footnotesize{(-0.4)*}\tabularnewline
+~~MD10&0 \footnotesize{(-1.4)*}&0 \footnotesize{(-0.1)}&0 \footnotesize{(-0.8)**}&0.1 \footnotesize{(-4.3)**}\tabularnewline
 ~~P8&0.2 \footnotesize{(-8.7)**}&0.1 \footnotesize{(-6.3)**}&0.2 \footnotesize{(-10.4)**}&0.5 \footnotesize{(-13.1)**}\tabularnewline
@@
 ~~D7&0.4 \footnotesize{(-0.6)**}&0.4 \footnotesize{(-1.2)**}&0.4 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-0.3)*}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.6)**}&0.1 \footnotesize{(-4.5)**}&0.2 \footnotesize{(-5.4)**}&1 \footnotesize{(-1.4)**}\tabularnewline
 ~~P8&1.3 \footnotesize{(-1.1)**}&1.1 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
@@
 
-% trends in din, chl, sio2 at c10
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:c10chg", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station C10.
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:c10chg}}
-\begin{center}
-\begin{tabular}{lllcllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
-\cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
-\hline
-{\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&1.3&\textit{\textbf{0.1}}**&&20.4&\textit{\textbf{0.1}}**&&14.7&-0.6**\tabularnewline
-\hline
-{\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&1.3&\textit{\textbf{3}}**&&19.9&-2.9**&&15.9&-0.2**\tabularnewline
-~~1986-1994&1.3&\textit{\textbf{0.4}}**&&19.6&\textit{\textbf{0.2}}*&&15&-0.3**\tabularnewline
-~~1995-2003&1.5&\textit{\textbf{2.4}}**&&21.2&-0.3**&&14.4&-0.4**\tabularnewline
-~~2004-2013&1.3&-5.2**&&21.4&-0.4**&&13.2&-1**\tabularnewline
-\hline
-{\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~Spring&1.1&-0.1&&27.2&\textit{\textbf{1.1}}**&&13.2&-0.7**\tabularnewline
-~~Summer&1.2&\textit{\textbf{0.1}}&&73.4&-0.3&&13.5&-0.9**\tabularnewline
-~~Fall&1.4&\textit{\textbf{0.6}}**&&14.4&-1.2**&&15.7&-0.5**\tabularnewline
-~~Winter&1.7&\textit{\textbf{0.1}}&&7.6&\textit{\textbf{0.7}}**&&15.2&-0.3**\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
-
 \clearpage

### 2017-01-11 sftrends_manu.Rnw
text for observed trends added

@@
 library(NADA)
+library(english)
 @
@@
 \begin{document}
-% \linenumbers
+\linenumbers
 
@@
 
-The goal of this study was to provide a comprehensive description of nutrient trends in the Delta that will inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay.
+The goal of this study was to provide a comprehensive description of nutrient trends in the Delta to inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay.
 We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts.
@@
 The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given many nutrients sources and the volume of water that is exchanged through the system with natural and anthropogenic processes.
 A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}.
@@
 
-A total of twenty-seven \ac{wrtds} models were created, one for each nitrogen analyte at each station.
+A total of thirty \ac{wrtds} models were created, one for each nitrogen analyte at each station.
 The functional form of \ac{wrtds} is a simple regression\cite{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:
@@
 Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick29}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisuan Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
 Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
-Observed concentrations were also higher later in the time series for all analytes.
+Mean observed concentrations were also higher later in the time series for all analytes.
 For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}.
@@
 
+<<echo = F, cache = T>>=
+data(trnds_seasyr)
+
+trndsyr <- filter(trnds_obs, cat %in% c('1976-1995', '1996-2013')) %>% 
+  select(Location, Site_Code, resvar, cat, ztest, perchg)
+
+obsdir <- filter(trndsyr, ztest < 0.05) %>% 
+  mutate(dirchg = ifelse(perchg < 0, 'down', 'up')) %>% 
+  group_by(cat, resvar) %>% 
+  summarise(
+    up = sum(dirchg == 'up'), 
+    down = sum(dirchg =='down'),
+    tots = sum(up, down),
+    up = as.character(english(up)),
+    down = as.character(english(down)),
+    tots = as.character(english(tots))
+  ) %>% 
+  as.data.frame %>%
+  split(., .$resvar)
+P8dec <- filter(trndsyr, 
+    Site_Code %in% 'P8' &
+    resvar %in% 'nh'
+  ) 
+@
+
 Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years.
-Trends from 1976-1995 and 1996-2013 varied considerably between sites and analytes.
-Overall,
-
-kendall tests - observed, \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}
-
-What about flow? \cref{fig:dinc10}
+Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).
+More sites had decreasing trends for the observed data from 1996-2013.
+\Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing).
+Trends by location (upper Delta, middle, and Suisun stations) were not apparent, suggesting individual sites had trends that differed independent of relative location.
+For example, P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites.
+Trends by season were similar such that increases were generally observed in all seasons for the first time period from 1976-1995 and decreases were observed for the second (\cref{fig:trndcomp2,fig:trndcomp3}), although trends were noisier and significant changes were less common compared to the annual comparisons.
+
+Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
+These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are potentially confounded by flow effects.
+A simple monotonic change over time could reflect moblization or dilation effect of flow rather than an empirical system response to changes in nutrient sources.
+As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data.
+YThe raw data are presented in \cref{fig:dinc10}a and the annual predicted and flow-normalized results form \ac{wrtds} are shown in \cref{fig:dinc10}b.
 
@@
 
-\cref{fig:clmchl}, \cref{fig:d7c10trnds}, \cref{tab:d7chg}, \cref{tab:c10chg}
+\cref{fig:clmchl}, \cref{tab:d7chg}
 

### 2017-01-11 sftrends_manu.tex
text for observed trends added

@@
 Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick29}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.16$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
-Observed concentrations were also higher later in the time series for all analytes.
+Mean observed concentrations were also higher later in the time series for all analytes.
 For example, average \ac{din} across all stations was $0.62$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.72$$\pm$$0.01$ for 1996-2013.
@@
 Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years.
-Trends from 1976-1995 and 1996-2013 varied considerably between sites and analytes.
-Overall,
+Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).
+More sites had decreasing trends for the observed data from 1996-2013.
+Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing).
+Trends by location (upper Delta, middle, and Suisun stations) were not apparent, suggesting individual sites had trends that differed independent of relative location.
+For example, P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites.
+Trends by season were similar such that increases were generally observed in all seasons for the first time period from 1976-1995 and decreases were observed for the second (\cref{fig:trndcomp2,fig:trndcomp3}), although trends were noisier and significant changes were less common compared to the annual comparisons.
 
-kendall tests - observed, \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}
-
-What about flow? \cref{fig:dinc10}
+Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
+These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are potentially confounded by flow effects.
+A simple monotonic change over time could reflect moblization or dilation effect of flow rather than a response to nutrient soruces.
+As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data.
+YThe raw data are presented in \cref{fig:dinc10}a and the annual predicted and flow-normalized results form \ac{wrtds} are shown in \cref{fig:dinc10}b.
 

### 2017-01-11 R/funcs.R

@@ -347,6 +347,14 @@ trnd_map <- function(res = c('din', 'nh', 'no23'), mods,
   
 }
 
+# capitalize all words in a string
+cap <- function(x) {
+  x <- as.character(x)
+  s <- strsplit(x, " ")[[1]]
+  paste(toupper(substring(s, 1,1)), substring(s, 2),
+      sep="", collapse=" ")
+}
+
 ######
 # gam equivalent of dynaplot function
 # looks at changes in in the chlorophyll-flow relationships by month over different years

### 2017-01-13 sftrends_manu.Rnw
fig 4e range truncated

@@
 In general, long-term (inter-annual) trends were not easily observed from the raw data.
-Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick29}.
+\ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisuan Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
@@
 For example, P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites.
-Trends by season were similar such that increases were generally observed in all seasons for the first time period from 1976-1995 and decreases were observed for the second (\cref{fig:trndcomp2,fig:trndcomp3}), although trends were noisier and significant changes were less common compared to the annual comparisons.
+Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
+Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.
 
 Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
-These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are potentially confounded by flow effects.
-A simple monotonic change over time could reflect moblization or dilation effect of flow rather than an empirical system response to changes in nutrient sources.
+These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
+Change over for time in the observed reflect could reflect mobilization or dilution effects of flow on concentration rather than an empirical system response to changes in nutrient sources.
 As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data.
-YThe raw data are presented in \cref{fig:dinc10}a and the annual predicted and flow-normalized results form \ac{wrtds} are shown in \cref{fig:dinc10}b.
+Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b.
+In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
+The difference between the two is shown in \cref{fig:dinc10}c such that years with model predictions greater or less than the flow-normalized values correspond wiht long-term trends in flow shown in \cref{fig:dinc10}d.
+For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.
+Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
+Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range
 
@@
 Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
-Subfigure (e) shows the modelled relationships between \\ac{din}, flow, and time.'>>=
+Subfigure (e) shows the modelled relationships between \\ac{din}, flow, and time.
+Flow values in (e) are truncated by 5\\textsuperscript{th} and 95\\textsuperscript{th} percentiles of observed.'>>=
 
@@
   x = max(toplo$date) - 300, 
-  y =  c(5, 2.4, 0.6, 7, 7), 
+  y =  c(5, 2.4, 0.6, 7, 5.85), 
   txt = c('(a)', '(b)', '(c)', '(d)', '(e)')
@@
 ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
-p5 <- gridplot(toplo, month = 'all', logspace = F, col_lim = col_lim, floscl = F) +
+p5 <- gridplot(toplo, month = 'all', logspace = F, floscl = F) +
   scale_y_continuous(ylab, expand = c(0, 0)) +

### 2017-01-13 sftrends_manu.tex
fig 4e range truncated

@@
 In general, long-term (inter-annual) trends were not easily observed from the raw data.
-Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick29}.
+\ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.16$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
@@
 For example, P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites.
-Trends by season were similar such that increases were generally observed in all seasons for the first time period from 1976-1995 and decreases were observed for the second (\cref{fig:trndcomp2,fig:trndcomp3}), although trends were noisier and significant changes were less common compared to the annual comparisons.
+Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
+Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.
 
 Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
-These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are potentially confounded by flow effects.
-A simple monotonic change over time could reflect moblization or dilation effect of flow rather than a response to nutrient soruces.
+These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
+Change over for time in the observed reflect could reflect mobilization or dilution effects of flow on concentration rather than an empirical system response to changes in nutrient sources.
 As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data.
-YThe raw data are presented in \cref{fig:dinc10}a and the annual predicted and flow-normalized results form \ac{wrtds} are shown in \cref{fig:dinc10}b.
+Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b.
+In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
+The difference between the two is shown in \cref{fig:dinc10}c such that years with model predictions greater or less than the flow-normalized values correspond wiht long-term trends in flow shown in \cref{fig:dinc10}d.
+For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.
+Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
+Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range
 
@@
 Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
-Subfigure (e) shows the modelled relationships between \ac{din}, flow, and time.}\label{fig:dinc10}
+Subfigure (e) shows the modelled relationships between \ac{din}, flow, and time.
+Flow values in (e) are truncated by 5\textsuperscript{th} and 95\textsuperscript{th} percentiles of observed.}\label{fig:dinc10}
 \end{figure}

### 2017-01-13 sftrends_manu.Rnw
dinc10 figure description donein text

@@
 In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
-The difference between the two is shown in \cref{fig:dinc10}c such that years with model predictions greater or less than the flow-normalized values correspond wiht long-term trends in flow shown in \cref{fig:dinc10}d.
+The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
 For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.
 Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
-Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range
-
-Emphasize the information the model provides relative to the observed time series.
-A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
-The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
-Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
-The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
-Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
-Emphasize the summer/fall change in the 2000s, why is this?
-Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
-Relate to conceptual diagram.
-
-Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.
+Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.
+Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s.
+% The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}
+
+A comparison of trends with flow-normalized components relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
+These comparisons are made relative to changes in the magnitude, significance, and direction of the trend, all of which have important implications for decision-making.
+Overall, the differences in apparent trends underscore the critical importance of considering flow effects in decision-making, where flow is affected by complex relationships between inflow, water withdrawals, and exchange with the larger Bay.
 

### 2017-01-13 sftrends_manu.tex
dinc10 figure description donein text

@@
 In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
-The difference between the two is shown in \cref{fig:dinc10}c such that years with model predictions greater or less than the flow-normalized values correspond wiht long-term trends in flow shown in \cref{fig:dinc10}d.
+The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
 For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.
 Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
-Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range
+Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.
+Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s.
+% The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}
 
-Emphasize the information the model provides relative to the observed time series.
-A distinct annual trend with a maximum in the middle of the time series is observed, with lower values at the beginning and end of the period.
-The seasonal patterns generally showed that \ac{din} concentrations were highest in January with higher values at moderate to low flow rates depending on the year.
-Interestingly, summer and fall concentrations have showed a slight increase later in the time series (~2004-2009).
-The confounding effect of flow is also very apparent such that higher flows were associated with lower concentration.
-Dynaplot showed that there was always a negative assocation between the two (i.e., no modal response).
-Emphasize the summer/fall change in the 2000s, why is this?
-Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN?
-Relate to conceptual diagram.
-
-Effects of flow on trend interpration shown in \cref{fig:trndcomp1,fig:trndcomp2,fig:trndcomp3}.
+A comparison of trends with flow-normalized components relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
+These comparisons are made relative to changes in the magnitude, significance, and direction of the trend, all of which have important implications for decision-making.
+Overall, the differences in apparent trends underscore the critical importance of considering flow effects in decision-making, where flow is affected by complex relationships between inflow, water withdrawals, and exchange with the larger Bay.
 

### 2017-01-13 sftrends_manu.Rnw
working on flo obs text comparison

@@
 
-A comparison of trends with flow-normalized components relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
-These comparisons are made relative to changes in the magnitude, significance, and direction of the trend, all of which have important implications for decision-making.
-Overall, the differences in apparent trends underscore the critical importance of considering flow effects in decision-making, where flow is affected by complex relationships between inflow, water withdrawals, and exchange with the larger Bay.
+<<echo = F, cache = F>>=
+data(trnds_seasyr)
+
+cmp <- data.frame(trnds_obs, dat = 'obs') %>% 
+  rbind(., data.frame(trnds_nrm, dat = 'nrm')) %>% 
+  filter(cat %in% c('1976-1995', '1996-2013')) %>% 
+  select(dat, Site_Code, resvar, cat, ztest, perchg)
+
+zcmp <- select(cmp, -perchg) %>% 
+  mutate(ztest = ifelse(ztest < .05, 'sig', 'ns')) %>% 
+  spread(dat, ztest) %>% 
+  mutate(
+    sigtons = ifelse(obs == 'sig' & nrm == 'ns', 1, 0), 
+    nstosig = ifelse(obs == 'ns' & nrm == 'sig', 1, 0)
+  )
+
+chgcmp <- select(cmp, -ztest)# %>% 
+  # mutate(ztest = ifelse(ztest < .05, 'sig', 'ns')) %>% 
+  # spread(dat, ztest) %>% 
+  # mutate(
+  #   sigtons = ifelse(obs == 'sig' & nrm == 'ns', 1, 0), 
+  #   nstosig = ifelse(obs == 'ns' & nrm == 'sig', 1, 0)
+  # )
+
+@
+
+A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
+These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
+For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results.
+The magnitude of all comparisons changed
+
+Overall, the differences in apparent trends underscore the critical importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
 

### 2017-01-13 sftrends_manu.tex
working on flo obs text comparison

@@
 
-A comparison of trends with flow-normalized components relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
-These comparisons are made relative to changes in the magnitude, significance, and direction of the trend, all of which have important implications for decision-making.
-Overall, the differences in apparent trends underscore the critical importance of considering flow effects in decision-making, where flow is affected by complex relationships between inflow, water withdrawals, and exchange with the larger Bay.
+A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
+These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
+For all sixty trend comparisons in \cref{fig:trndcomp1} regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results.
+The magnitude of all comparisons changed
+
+Overall, the differences in apparent trends underscore the critical importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
 

### 2017-01-13 R/funcs.R

@@ -83,6 +83,16 @@ scinot <- function(x, pow = 2, digits = 2, showDollar = TRUE)
   
 }
 
+# wrapper for as.character to english function
+eng <- function(chr){
+  
+  out <- english(chr)
+  out <- as.character(out)
+  
+  return(out)
+  
+}
+
 ######
 # trend maps
 # 

### 2017-01-13 sftrends_manu.Rnw
version to title page

@@
 
+% get the version from git log
+<<echo = FALSE, cache = FALSE>>=
+raw <- system('git log -1', intern = TRUE)
+date <- raw[grep('^Date', raw)]
+date <- gsub('^Date:\\s*', '', date)
+commit <- raw[grep('^commit', raw)]
+commit <- gsub('^commit ', '', commit)
+commit <- paste0('\\href{https://github.com/fawda123/sftrends_manu/commit/', commit, '}{', commit, '}')
+vers <- paste0('Version: ', date, ', ', commit)
+@
+
 \author{Marcus W. Beck}
@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}}}
 

### 2017-01-13 sftrends_manu.tex
version to title page

@@
 
+% get the version from git log
+
 \author{Marcus W. Beck}
@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 13:54:46 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/63c599b38e357cdead0d5b008b8862f7b0d3ec57}{63c599b38e357cdead0d5b008b8862f7b0d3ec57}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.
-The most censored observations were observed for ammonium time series at sites C10 (25.4\%), D28 (17.8\%), D19 (12\%), D7 (7.9\%), and D4 (6.4\%).
+The most censored observations were observed for ammonium time series at sites C10 (25.4\%), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).
 
@@
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.16$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
 Mean observed concentrations were also higher later in the time series for all analytes.
-For example, average \ac{din} across all stations was $0.62$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.72$$\pm$$0.01$ for 1996-2013.
+For example, average \ac{din} across all stations was $0.61$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.7$$\pm$$0.01$ for 1996-2013.
 Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.

### 2017-01-13 sftrends_manu.Rnw
flow to obsdat fig

@@
 
-<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta.
-Data were collected monthly and evaluated with \\ac{wrtds} models from 1976 to 2013.
+<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
+Data were collected monthly and evaluated with \\ac{wrtds} models using daily flow estimates from 1976 to 2013.
 Note different y-axis scales.
-See \\cref{fig:delt_map} for station locations.', fig.height = 7, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
+See \\cref{fig:delt_map} for station locations.', fig.height = 7.5, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
 data(delt_dat)
+data(flow_dat)
+
+## 
+# obs data
 
@@
     panel.grid.major = element_line(size = 0.25),
-    panel.grid.minor = element_blank()
+    panel.grid.minor = element_blank(),
+    plot.margin = grid::unit(c(3, 6, 3, 6), 'pt')
   )
@@
 
-ggplot(toplo, aes(x = Date)) +
+p1 <- ggplot(toplo, aes(x = Date)) +
   geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
@@
   scale_fill_manual(values = cols, labels = leglabs) +
+  mythm %+replace% 
+  theme(axis.text.x = element_blank())
+
+## 
+# flo data
+toplo2 <- filter(flow_dat, 
+    !station %in% 'east' &
+    Date >= as.Date('1976-01-13') &
+    Date <= as.Date('2013-12-13')
+  ) %>% 
+  mutate(
+    q = log(q),
+    station = factor(station, levels = c('sac', 'sjr'), labels = c('Sac.', 'S. Jo.'))
+    )
+ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+p2 <- ggplot(toplo2, aes(x = Date, y = q)) +
+  geom_line(size = 0.25) +
+  scale_x_date(expand = c(0, 0)) + 
+  scale_y_continuous(ylab) + 
+  facet_grid(station~., scales = 'free_y') + 
   mythm
 
+# fix widths
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+
+maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
+
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+
+grid.arrange(pA, pB, ncol = 1, heights = c(1, 0.21))
 @

### 2017-01-13 sftrends_manu.tex
flow to obsdat fig

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 13:54:46 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/63c599b38e357cdead0d5b008b8862f7b0d3ec57}{63c599b38e357cdead0d5b008b8862f7b0d3ec57}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 15:40:02 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/f49c0a27672550b75425bf31fd1b561ddf59669e}{f49c0a27672550b75425bf31fd1b561ddf59669e}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
-\caption{Observed \ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \ac{sfe} Delta.
-Data were collected monthly and evaluated with \ac{wrtds} models from 1976 to 2013.
+\caption{Observed \ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
+Data were collected monthly and evaluated with \ac{wrtds} models using daily flow estimates from 1976 to 2013.
 Note different y-axis scales.

### 2017-01-15 sftrends_manu.Rnw
trndmap gone, almost done with trends text

@@
 Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
-As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.
+As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8.
 The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.
@@
 These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-Change over for time in the observed reflect could reflect mobilization or dilution effects of flow on concentration rather than an empirical system response to changes in nutrient sources.
-As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data.
+Change over for time in observed data could reflect mobilization or dilution effects of flow on concentration rather than an interpretable system response to changes in nutrient sources.
+As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10.
 Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b.
@@
 
+# descriptsions for sig changes
 zcmp <- select(cmp, -perchg) %>% 
@@
 
-chgcmp <- select(cmp, -ztest)# %>% 
-  # mutate(ztest = ifelse(ztest < .05, 'sig', 'ns')) %>% 
-  # spread(dat, ztest) %>% 
-  # mutate(
-  #   sigtons = ifelse(obs == 'sig' & nrm == 'ns', 1, 0), 
-  #   nstosig = ifelse(obs == 'ns' & nrm == 'sig', 1, 0)
-  # )
+# descriptions for magnitude changes
+chgcmp <- select(cmp, -ztest) %>% 
+  spread(dat, perchg) %>%
+  mutate(
+    signobs = sign(obs), 
+    signnrm = sign(nrm),
+    chgs = ifelse(
+      signobs == -1 & nrm < obs, 'moreneg', 
+      ifelse(signobs == 1 & nrm > obs, 'morepos',
+      ifelse(signobs == 1 & signnrm == -1, 'postoneg', 
+      ifelse(signobs == -1 & signnrm == 1, 'negtopos', 
+      ifelse(signobs == -1 & signnrm == -1 & nrm > obs, 'lessneg', 
+      ifelse(signobs == 1 & signnrm == 1 & nrm < obs, 'lesspos', 
+        'somechg')
+    )))))
+  )
+# table of results for shorter sexpr
+chgtb <- table(chgcmp$chgs)
 
@@
 These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
-For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results.
-The magnitude of all comparisons changed
-
-Overall, the differences in apparent trends underscore the critical importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
+For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
+Changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an inrease in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positve or less negative) compared to observed trends.
+More importantly, \Sexpr{eng(chgtb['postoneg'])} comparisons showed a trend reversal from positive to negative estimated change and \Sexpr{eng(chgtb['somechg'])} sites went from no change to negative estimated trends for the flow-normalized results.
+Differences by season in the observed relative to flow-normalized results from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
+The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013.
+
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
+
+Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta. Huge effect on trend interpration
+
+What do nitrogen trends mean?
+Have to interpret relative to trends in other variables.
+A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
+A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
+However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
+Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
+See Senn slide 14 (from burial?).
+The WRTDS model lets us at least address trends in the context of season, time, and flow.
+This allows for more improved interpretion relative to observing raw data.
+Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
+What about other variables (light level as suspended particulate matter, temperature)?
 
@@
 
-Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta.
-
 Second case study showed typical inverse relationships between nutrients and flow, more flow means greater flushing and dilution of nutrient concentrations.
@@
 But what are watershed loads?
-
-What do nitrogen trends mean?
-Have to interpret relative to trends in other variables.
-A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
-A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
-However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
-Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
-See Senn slide 14 (from burial?).
-The WRTDS model lets us at least address trends in the context of season, time, and flow.
-This allows for more improved interpretion relative to observing raw data.
-Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
-What about other variables (light level as suspended particulate matter, temperature)?
 
@@
   ncol = 1
-  )
-
-@
-
-<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013.
-Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
-Station names are shown in the top left panel.
-Station locations have been jittered to reduce overlap.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 5.3, cache = T, out.width = '0.7\\textwidth'>>=
-
-# load required data
-data(delt_dat)
-data(delt_map)
-data(trnds_seasyr)
-
-# combine all
-trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
-  select(Site_Code, resvar, cat, perchg, ann, trndvar) %>% 
-  spread(trndvar, perchg) %>% 
-  mutate(Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3')))
-
-##
-# get a bounding box for the stations
-statmeta <- select(delt_dat, Site_Code, Latitude, Longitude) %>% 
-  unique
-
-# bounding box
-buffx <- 0.00035
-buffy <- 0.0022
-lims<- select(statmeta, -Site_Code) %>% 
-  apply(., 2, function(x) range(x, na.rm = TRUE))
-lims <- c(
-  (1 + buffx) * lims[1, 2], 
-  (1 - buffy) * lims[1, 1], 
-  (1 - buffx) * lims[2, 2], 
-  (1 + buffy) * lims[2, 1]
-  )    
-names(lims) <- c('left', 'bottom', 'right', 'top')
-
-# need to shift locations of some stations
-statmeta[statmeta$Site_Code %in% 'D26', 'Latitude'] <- 38.14
-statmeta[statmeta$Site_Code %in% 'D26', 'Longitude'] <--121.55
-statmeta[statmeta$Site_Code %in% 'D7', 'Latitude'] <- 38.13
-statmeta[statmeta$Site_Code %in% 'D7', 'Longitude'] <--122.015
-statmeta[statmeta$Site_Code %in% 'C10', 'Latitude'] <-37.68
-statmeta[statmeta$Site_Code %in% 'C10', 'Longitude'] <--121.28
-statmeta[statmeta$Site_Code %in% 'D28', 'Latitude'] <-37.94
-statmeta[statmeta$Site_Code %in% 'MD10', 'Latitude'] <-38.1
-statmeta[statmeta$Site_Code %in% 'P8', 'Latitude'] <-37.93
-statmeta[statmeta$Site_Code %in% 'D19', 'Latitude'] <-38.06
-
-# response labels 
-reslabs <- c('din an', 'nh an', 'no23 an', 'din mobef', 'nh mobef', 'no23 mobef', 'din moaft', 'nh moaft', 'no23 moaft')
-reslevs <- rep(c(
-  'DIN', 
-  expression(paste(NH [4] ^ '+')),
-  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
-  ), 3)
-
-# convert delta map to ggplot compatible object
-delt_map <- fortify(delt_map)
-
-# base delta map
-alph <- 1
-pbase <- ggplot(delt_map, aes(x = long, y = lat)) + 
-  geom_polygon(aes(group = group, fill = hole), colour = alpha("lightblue", alph)) +
-  theme_bw() + 
-  theme(
-    plot.background = element_rect(fill='transparent', 
-      colour = NA),
-    panel.background = element_rect(fill='transparent', 
-      colour = NA),
-    legend.background = element_rect(fill='transparent', 
-      colour = NA),
-    legend.key = element_rect(fill = 'transparent', 
-      colour = NA),
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"), 
-    panel.grid.minor = element_blank(), 
-    axis.title.x = element_blank(), 
-    axis.title.y = element_blank(), 
-    legend.position = 'none', 
-    plot.margin = grid::unit(c(3, 3, 3, 3), 'pt'), 
-    axis.text.x = element_text(size = 8), 
-    axis.text.y = element_text(size = 8), 
-    strip.background=element_rect(fill = 'white', colour = 'white'),
-    panel.border = element_rect(colour = "grey")
-  ) +
-  coord_fixed(ratio = 1, xlim = lims[c(1, 3)], ylim = lims[c(2, 4)]) +
-  scale_fill_manual(values=c(alpha("lightblue", alph), "white"), guide="none")
-
-# bounds on limits
-yrng <- max(abs(range(trnds_seasyr$norm)))
-yrng <- c(-1 * yrng, yrng)
-
-# color ramp, red for pos, green for neg
-colpos <- RColorBrewer::brewer.pal(9, 'Reds')[3:9]
-colneg <- RColorBrewer::brewer.pal(9, 'Greens')[3:9]
-palneg <-  colorRampPalette(colneg)(1000)
-palpos <-   colorRampPalette(colpos)(1000)
-
-# plot(1:2000, col = c(palneg, palpos), pch = 16, cex = 4)
-# points(1000, 1000)
-
-# unnest data frame, assign colors by rank and sign
-toplo <- select(trnds_seasyr, Site_Code, resvar, ann, cat, norm) %>% 
-  rename(perchg = norm) %>% 
-  left_join(., statmeta, by = 'Site_Code') %>% 
-  data.frame %>% 
-  mutate(
-    perchg = pmax(yrng[1], perchg),
-    perchg = pmin(yrng[2], perchg),
-    negs = sign(perchg),  
-    rnks = scales::rescale(abs(perchg), c(1, 1000)), 
-    rnks = round(rnks), 
-    cols = ifelse(negs == 1, palpos[rnks], palneg[rnks]),
-    ind = 1:nrow(.)
-    ) %>%
-  filter(!cat %in% c('1976-1995', '1996-2013')) %>% # remove year categories
-  unite('cattype', cat, ann, sep = '', remove = F) %>%
-  mutate(
-    cattype = factor(cattype),
-    cattype = fct_collapse(cattype, 
-      mobef = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'bef'), 
-      moaft = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'aft') 
-    )
-  )
-
-# unite response varibles and summary period combos
-unis <- with(toplo, paste(resvar, cattype)) %>% 
-  unique %>% 
-  sort
-
-sites <- unique(toplo$Site_Code)
-
-plo_ls <- vector('list', length = length(unis))
-names(plo_ls) <- unis
-
-for(uni in unis){
-
-  # get variable, agg period
-  splits <- strsplit(uni, ' ')[[1]]
-  resvart <- splits[1]
-  cattypet <- splits[2]
-  toplo_tmp <- filter(toplo, resvar %in% resvart & cattype %in% cattypet)
-
-  # create individual bar plots as ggplot list
-  bars <- setNames(lapply(sites, function(i){
-
-    subplo <- filter(toplo_tmp, Site_Code %in% i)
-    ggplot(subplo, aes(x=cat, y = perchg, fill = factor(cat))) +
-      geom_bar(stat = 'identity', colour = 'black', size = 0.2) + 
-      scale_fill_manual(values = as.character(subplo$cols)) + 
-      theme_tree() + 
-      xlab(NULL) + 
-      ylab(NULL) + 
-      theme_transparent() +
-      theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
-      scale_y_continuous(limits = yrng)
-
-  }), sites)
-
-  ##
-  # add the barplots to the base map
-
-  # add site labels if plot is first
-  pall <- pbase
-  set.seed(5)
-  if(which(uni == unis) == 2)
-    pall <- pbase + 
-      geom_label_repel(
-        data = statmeta, 
-        aes(x = Longitude, y = Latitude, label = Site_Code), 
-        label.r = unit(0, "lines"),
-        box.padding = unit(2, "lines"),
-        label.padding = unit(0.15, "lines"), 
-        point.padding = unit(0.1, "lines"),
-        force = 2, size = 2,
-        fill = 'grey95'
-      )
-
-  # keep bar widths the same for cattypes
-  subwid <- 0.1
-  if(grepl('an$', uni))
-    subwid <- 0.06
-
-  for(i in 1:length(bars)){
-
-      nms <- names(bars)[i]
-      dat <- statmeta[statmeta$Site_Code %in% nms,]
-      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, subwid, 0.21) # change bar heights here
-
-  }
-
-  tval <- reslevs[which(uni == reslabs)]
-  pall <- pall + 
-    geom_text(data = data.frame(x = lims[1], y = lims[4], tval = as.character(tval)), 
-      aes(x = x, y = y, label = tval), parse = TRUE, hjust = 0, vjust = 1)
-  plo_ls[[uni]] <- pall
-
-}
-
-# setup legend
-rngs <- seq(yrng[1], yrng[2], length = length(c(colneg, colpos)))
-leg <- ggplot(data.frame(x = seq_along(rngs), y = seq_along(rngs), z = rngs), aes(x, y, colour = z)) + 
-  geom_point() + 
-  scale_colour_gradientn(colours = c(rev(colneg), colpos)) + 
-  guides(colour = guide_colourbar(title = expression('% chg yr' ^-1), barheight = 0.5, barwidth = 8)) + 
-  theme(legend.position = 'top')
-leg <- g_legend(leg)
-
-grid.arrange(
-  leg,
-  arrangeGrob(
-    arrangeGrob(plo_ls[[2]], plo_ls[[4]], plo_ls[[6]], ncol = 1, top = c('(a) Sp, Su, Fa, Wi (1976-1995)')),
-    arrangeGrob(plo_ls[[1]], plo_ls[[3]], plo_ls[[5]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1996-2013)')),
-    ncol = 2
-    ),
-  ncol = 1, 
-  heights = c(0.1, 1)
   )
@@
 
-% overall trends, years
-<<trndsann>>=
-data(trnds_seasyr)
-
-# percent changes
-tab <- trnds_nrm %>% 
-  filter(cat %in% c('1976-1995', '1996-2013')) %>% 
-  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      paste0('(', perchg, ')' ), 
-      paste0('(\\textit{\\textbf{', perchg, '}})')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  mutate(
-    perchg = paste0('\\footnotesize{', perchg, '}')
-  ) %>% 
-  unite('val', med, perchg, sep = ' ') %>% 
-  spread(cat, val) %>% 
-  arrange(resvar, Site_Code) %>% 
-  select(Site_Code, resvar, `1976-1995`, `1996-2013`)
-
-# table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \\cref{fig:trndmap} for a summary of spatial trends.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
-stats <- tab$Site_Code
-
-# table
-latex( 
-  tab[, -c(1, 2)],
-  file = '',
-  rowlabel = 'Analyte/Station',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = vars,
-  n.rgroup = rep(nrow(tab)/3, 3),
-  cgroup = c('Annual'),
-  n.cgroup = c(2),
-  rowname = stats,
-  label = 'tab:trndsann',
-  insert.bottom = foot.val
-  )
-
-@
-
-% overall trends, seasons, first annual periodyears
-<<trndsbef>>=
-data(trnds_seasyr)
-
-# percent changes
-tab <- trnds_nrm %>% 
-  filter(
-    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
-    ann %in% 'bef'
-    ) %>% 
-  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      paste0('(', perchg, ')' ), 
-      paste0('(\\textit{\\textbf{', perchg, '}})')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  mutate(
-    perchg = paste0('\\footnotesize{', perchg, '}')
-  ) %>% 
-  unite('val', med, perchg, sep = ' ') %>% 
-  spread(cat, val) %>% 
-  arrange(resvar, Site_Code)
-
-# table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \\cref{fig:trndmap} for a summary of spatial trends.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
-stats <- tab$Site_Code
-
-# table
-latex( 
-  tab[, -c(1, 2)],
-  file = '',
-  rowlabel = 'Analyte/Station',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = vars,
-  n.rgroup = rep(nrow(tab)/3, 3),
-  cgroup = c('Seasonal, 1976-1995'),
-  n.cgroup = c(4),
-  rowname = stats,
-  label = 'tab:trndsbef',
-  insert.bottom = foot.val
-  )
-
-@
-
-% overall trends, seasons, first annual periodyears
-<<trndsaft>>=
-data(trnds_seasyr)
-
-# percent changes
-tab <- trnds_nrm %>% 
-  filter(
-    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
-    ann %in% 'aft'
-    ) %>% 
-  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      paste0('(', perchg, ')' ), 
-      paste0('(\\textit{\\textbf{', perchg, '}})')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  mutate(
-    perchg = paste0('\\footnotesize{', perchg, '}')
-  ) %>% 
-  unite('val', med, perchg, sep = ' ') %>% 
-  spread(cat, val) %>% 
-  arrange(resvar, Site_Code)
-
-# table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \\cref{fig:trndmap} for a summary of spatial trends.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
-stats <- tab$Site_Code
-
-# table
-latex( 
-  tab[, -c(1, 2)],
-  file = '',
-  rowlabel = 'Analyte/Station',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = vars,
-  n.rgroup = rep(nrow(tab)/3, 3),
-  cgroup = c('Seasonal, 1996-2013'),
-  n.cgroup = c(4),
-  rowname = stats,
-  label = 'tab:trndsaft',
-  insert.bottom = foot.val
-  )
-
-@
-
 % trends in no23, nh at p8
@@
 
+% overall trends, years
+<<trndsann>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(cat %in% c('1976-1995', '1996-2013')) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code) %>% 
+  select(Site_Code, resvar, `1976-1995`, `1996-2013`)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+See \\cref{fig:trndmap} for a summary of spatial trends.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = rep(nrow(tab)/3, 3),
+  cgroup = c('Annual'),
+  n.cgroup = c(2),
+  rowname = stats,
+  label = 'tab:trndsann',
+  insert.bottom = foot.val
+  )
+
+@
+
+% overall trends, seasons, first annual periodyears
+<<trndsbef>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(
+    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
+    ann %in% 'bef'
+    ) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+See \\cref{fig:trndmap} for a summary of spatial trends.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = rep(nrow(tab)/3, 3),
+  cgroup = c('Seasonal, 1976-1995'),
+  n.cgroup = c(4),
+  rowname = stats,
+  label = 'tab:trndsbef',
+  insert.bottom = foot.val
+  )
+
+@
+
+% overall trends, seasons, first annual periodyears
+<<trndsaft>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(
+    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
+    ann %in% 'aft'
+    ) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+See \\cref{fig:trndmap} for a summary of spatial trends.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
+foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = rep(nrow(tab)/3, 3),
+  cgroup = c('Seasonal, 1996-2013'),
+  n.cgroup = c(4),
+  rowname = stats,
+  label = 'tab:trndsaft',
+  insert.bottom = foot.val
+  )
+
+@
+
 \end{suppinfo}

### 2017-01-15 sftrends_manu.tex
trndmap gone, almost done with trends text

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 15:40:02 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/f49c0a27672550b75425bf31fd1b561ddf59669e}{f49c0a27672550b75425bf31fd1b561ddf59669e}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 16:19:14 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/66d301f76d51fddfc7d4eeea2ed7b92995d7385f}{66d301f76d51fddfc7d4eeea2ed7b92995d7385f}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
-For all sixty trend comparisons in \cref{fig:trndcomp1} regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results.
-The magnitude of all comparisons changed
+For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
+Changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an inrease in magnitude (more negative or more positive) and twenty five had a decrease (less positve or less negative) compared to observed trends.
+More importantly, eleven comparisons showed a trend reversal from positive to negative estimated change and ten sites went from no change to negative estimated trends for the flow-normalized results.
+Differences by season in the observed relative to flow-normalized results from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
+The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013.
 
-Overall, the differences in apparent trends underscore the critical importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
+
+Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta. Huge effect on trend interpration
+
+What do nitrogen trends mean?
+Have to interpret relative to trends in other variables.
+A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
+A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
+However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
+Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
+See Senn slide 14 (from burial?).
+The WRTDS model lets us at least address trends in the context of season, time, and flow.
+This allows for more improved interpretion relative to observing raw data.
+Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
+What about other variables (light level as suspended particulate matter, temperature)?
 
@@
 
-{\centering \includegraphics[width=0.7\textwidth]{figs/trndmap-1}
-
-}
-
-\caption[Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013]{Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013.
-Changes are based on seasonal Kendall tests of flow-normalized results within each time period.
-Station names are shown in the top left panel.
-Station locations have been jittered to reduce overlap.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}
-\end{figure}
-
-\begin{figure}[!ht]
-
 {\centering \includegraphics[width=\textwidth]{figs/p8trnds-1}
@@
 % tables
+
+% trends in no23, nh at p8
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg", insert.bottom = foot.val)%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
+Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:p8chg}}
+\begin{center}
+\begin{tabular}{lllcll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
+\cline{2-3} \cline{5-6}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
+\hline
+{\bfseries Annual}&&&&&\tabularnewline
+~~1976-2006&1.3&\textit{\textbf{2}}**&&0.2&\textit{\textbf{2.8}}**\tabularnewline
+~~2007-2013&1.4&-1.9**&&0.1&-16.6**\tabularnewline
+\hline
+{\bfseries Seasonal, pre}&&&&&\tabularnewline
+~~Spring&1.2&\textit{\textbf{1.6}}**&&0.2&\textit{\textbf{1.4}}**\tabularnewline
+~~Summer&1&\textit{\textbf{2.4}}**&&0.1&\textit{\textbf{3.3}}**\tabularnewline
+~~Fall&1.3&\textit{\textbf{2.2}}**&&0.2&\textit{\textbf{4.9}}**\tabularnewline
+~~Winter&1.5&\textit{\textbf{2.1}}**&&0.7&\textit{\textbf{4.8}}**\tabularnewline
+\hline
+{\bfseries Seasonal, post}&&&&&\tabularnewline
+~~Spring&1.3&-1.6**&&0.1&-16.2**\tabularnewline
+~~Summer&0.9&-4.3**&&0.1&-15.7**\tabularnewline
+~~Fall&1.5&-1.7**&&0.1&-19.3**\tabularnewline
+~~Winter&2.2&-0.8**&&0.2&-26.7**\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
+
+% trends in din, chl, sio2 at d7
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:d7chg", insert.bottom = foot.val)%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:d7chg}}
+\begin{center}
+\begin{tabular}{lllcllcll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
+\cline{2-3} \cline{5-6} \cline{8-9}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
+\hline
+{\bfseries All}&&&&&&&&\tabularnewline
+~~1976-2013&0.4&\textit{\textbf{0.6}}**&&3&-6.7**&&12.7&\textit{\textbf{0.7}}**\tabularnewline
+\hline
+{\bfseries Annual}&&&&&&&&\tabularnewline
+~~1976-1985&0.4&-2.1**&&8.8&-10.7**&&10.2&-0.2\tabularnewline
+~~1986-1994&0.4&\textit{\textbf{3.6}}**&&2.6&-13.5**&&11.9&\textit{\textbf{2.3}}**\tabularnewline
+~~1995-2003&0.5&-0.1&&1.8&\textit{\textbf{1.9}}**&&13.3&\textit{\textbf{0.7}}**\tabularnewline
+~~2004-2013&0.5&-1.3**&&2.1&\textit{\textbf{2.9}}**&&13.1&-0.3**\tabularnewline
+\hline
+{\bfseries Seasonal}&&&&&&&&\tabularnewline
+~~Spring&0.5&-0.1&&3.4&-1&&14.7&\textit{\textbf{0.1}}**\tabularnewline
+~~Summer&0.4&\textit{\textbf{1.5}}**&&3.4&-8.8**&&12.2&\textit{\textbf{1.2}}**\tabularnewline
+~~Fall&0.4&\textit{\textbf{0.6}}**&&1.7&-8.8**&&12.1&\textit{\textbf{1}}**\tabularnewline
+~~Winter&0.6&-0.2&&1.4&-3.1**&&14.5&\textit{\textbf{0.3}}**\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\footnotesize *$p<0.05$; **$p<0.005$\end{table}
+
+\clearpage
+
+% supplementary material
+\begin{suppinfo}
+\beginsupplement
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1976-1995.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp2}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1996-2013.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp3}   
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\textwidth]{figs/stock-1}
+
+}
+
+\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
+\end{figure}
 
@@
 
-% trends in no23, nh at p8
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:p8chg}}
-\begin{center}
-\begin{tabular}{lllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
-\cline{2-3} \cline{5-6}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
-\hline
-{\bfseries Annual}&&&&&\tabularnewline
-~~1976-2006&1.3&\textit{\textbf{2}}**&&0.2&\textit{\textbf{2.8}}**\tabularnewline
-~~2007-2013&1.4&-1.9**&&0.1&-16.6**\tabularnewline
-\hline
-{\bfseries Seasonal, pre}&&&&&\tabularnewline
-~~Spring&1.2&\textit{\textbf{1.6}}**&&0.2&\textit{\textbf{1.4}}**\tabularnewline
-~~Summer&1&\textit{\textbf{2.4}}**&&0.1&\textit{\textbf{3.3}}**\tabularnewline
-~~Fall&1.3&\textit{\textbf{2.2}}**&&0.2&\textit{\textbf{4.9}}**\tabularnewline
-~~Winter&1.5&\textit{\textbf{2.1}}**&&0.7&\textit{\textbf{4.8}}**\tabularnewline
-\hline
-{\bfseries Seasonal, post}&&&&&\tabularnewline
-~~Spring&1.3&-1.6**&&0.1&-16.2**\tabularnewline
-~~Summer&0.9&-4.3**&&0.1&-15.7**\tabularnewline
-~~Fall&1.5&-1.7**&&0.1&-19.3**\tabularnewline
-~~Winter&2.2&-0.8**&&0.2&-26.7**\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
-
-% trends in din, chl, sio2 at d7
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:d7chg", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:d7chg}}
-\begin{center}
-\begin{tabular}{lllcllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
-\cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
-\hline
-{\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&0.4&\textit{\textbf{0.6}}**&&3&-6.7**&&12.7&\textit{\textbf{0.7}}**\tabularnewline
-\hline
-{\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&0.4&-2.1**&&8.8&-10.7**&&10.2&-0.2\tabularnewline
-~~1986-1994&0.4&\textit{\textbf{3.6}}**&&2.6&-13.5**&&11.9&\textit{\textbf{2.3}}**\tabularnewline
-~~1995-2003&0.5&-0.1&&1.8&\textit{\textbf{1.9}}**&&13.3&\textit{\textbf{0.7}}**\tabularnewline
-~~2004-2013&0.5&-1.3**&&2.1&\textit{\textbf{2.9}}**&&13.1&-0.3**\tabularnewline
-\hline
-{\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~Spring&0.5&-0.1&&3.4&-1&&14.7&\textit{\textbf{0.1}}**\tabularnewline
-~~Summer&0.4&\textit{\textbf{1.5}}**&&3.4&-8.8**&&12.2&\textit{\textbf{1.2}}**\tabularnewline
-~~Fall&0.4&\textit{\textbf{0.6}}**&&1.7&-8.8**&&12.1&\textit{\textbf{1}}**\tabularnewline
-~~Winter&0.6&-0.2&&1.4&-3.1**&&14.5&\textit{\textbf{0.3}}**\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
-
-\clearpage
-
-% supplementary material
-\begin{suppinfo}
-\beginsupplement
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1976-1995.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp2}   
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1996-2013.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp3}   
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\textwidth]{figs/stock-1}
-
-}
-
-\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
-\end{figure}
-
 \end{suppinfo}

### 2017-01-15 sftrends_manu.Rnw
almost done with first case study text

@@
 
-Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
-
-Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta. Huge effect on trend interpration
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes.
+Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed.
+A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.
+For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
+For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramentao or San Joaquin daily flow estimates.
+Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each locaiton.
+Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.
+
+\subsection{Selected examples}
+
+Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
+The selected case studies focused on \begin{inparaenum}[1\upshape)]
+\item effects of wastewater treatment upgrades upstream of P8, and
+\item effects of biological invasion on nutrient dynamics in Suisun Bay using D7 as an example.
+\end{inparaenum}
+Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
+
+\subsubsection{Effects of wastewater treatment}
+
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., )\cite{Cornwell14,Novick15}.
+Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}.
+Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
+\item a shift in nutrient contributions from the \ac{wwtp} before/after upgrade, and
+\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
+\end{inparaenum}
+
+<<echo = F>>=
+# for inline text and tab:p8chg
+
+data(h2dat)
+dt <- as.Date('2006-01-01')
+
+noraw_bf <- filter(h2dat, resvar %in% 'no23') %>% 
+  .$data %>% 
+  .[[1]] %>% 
+  .[which(.$date < dt), ]
+noraw_af <- filter(h2dat, resvar %in% 'no23') %>% 
+  .$data %>% 
+  .[[1]] %>% 
+  .[which(.$date >= dt), ]
+nhraw_bf <- filter(h2dat, resvar %in% 'nh') %>% 
+  .$data %>% 
+  .[[1]] %>% 
+  .[which(.$date < dt), ]
+nhraw_af <- filter(h2dat, resvar %in% 'nh') %>% 
+  .$data %>% 
+  .[[1]] %>% 
+  .[which(.$date >= dt), ]
+
+mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
+yrbrks_bf <- c(1975, 2005)
+yrbrks_af <- c(2006, 2013)
+molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
+yrlabs_bf <- '1976-2006'
+yrlabs_af <- '2007-2013'
+faclevs <- c('1976-2006_before', '2007-2013_after', paste0(molabs, '_before'), paste0(molabs, '_after'))
+
+# summarize din, sio2, chl trends
+no_bf <- wrtdstrnd_sk(noraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf) %>% 
+  mutate(period = 'before', var = 'no')
+no_af <- wrtdstrnd_sk(noraw_af, mobrks, yrbrks_af, molabs, yrlabs_af) %>% 
+  mutate(period = 'after', var = 'no')
+nh_bf <- wrtdstrnd_sk(nhraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf) %>% 
+  mutate(period = 'before', var = 'nh')
+nh_af <- wrtdstrnd_sk(nhraw_af, mobrks, yrbrks_af, molabs, yrlabs_af) %>% 
+  mutate(period = 'after', var = 'nh')
+
+@
+
+Effluent concentations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual shift in ammonium relative to the total (\cref{fig:stock}).
+Ammonium and nitrate concentrations were generally balanced prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite.
+As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase form 1976 followed by a large reduction in the 2000s.
+Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
+Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium, \cref{tab:p8chg}).
+Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%).
+Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate (\Sexpr{scinot(no_af[3, 'perchg'], 1, 1)}\%) and largest for ammonium in the winter (\Sexpr{scinot(nh_af[5, 'perchg'], 1, 1)}\%).
+
+Relationships of nitrogen with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing at high flow (\cref{fig:p8trnds}b).
+Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations at low flow in the winter (January).
+Additionally, strength of the flow/nutrient relationship changed between years.
+Nitrite/nitrate typically had the strongest relationshipwith flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
+Using \ac{wrtds}, an empirical link is created between upstream changes and observed effects downstream that is characterized by differences in analyte beween years and season.
+A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
+These dynamics cannot be characterized from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
+For example, estimated ammonium concentrations in July were low for all flow levels which could suggest either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
+As such, trends from \ac{wrtds} can be used to develop additional hypotheses for further testing cause and effect mechanisms of nutrient impacts on adverse conditions.
+
+\subsubsection{Effects of biological invasions}
+
+The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \cite{Cohen98}.
+The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \cite{Carlton90,Nichols90,Werner93,Crauder16}.
+In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
+Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}.
+The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
+
+We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
+\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
+\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
+\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
+\end{inparaenum}
+The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
+
+Data from \cite{Crauder16,Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
+Figure 10 in Ref.~\citenum{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
+
+Note the decrease in Potamocorbula abundance in 2011, 2012.
+These are wet years where abundance/biomass of the clams is driven down by lower salinity.
+Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.
+The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.
+This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
+See suggestions in Ref.~\citenum{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
+
+Further, chlorophyll trends early in the time series generally show a decrease with high flow with a distinct maximum at moderate flow.
+This may suggest stratification events at moderate flow contributed to phytoplankton blooms early in the time series.
+Water withdrawals later in the time series could have also altered environmental conditions to reduce the frequency occurrence of stratification events.
+Look into this more...
+
+What about biomass/density relationships for Potamocorbula?
+Although clam density increases throughout the period,  What about initial decrease in chlorophyll prior to clam invasion?
+Is this related to water withdrawals (i.e., decrease in stratification events at moderate flow)?
+
+\cref{fig:clmchl}, \cref{tab:d7chg}
+
+Second case study showed typical inverse relationships between nutrients and flow, more flow means greater flushing and dilution of nutrient concentrations.
+Conversely, low flow means less flushing and higher nutrient concentrations, although this may not always be observed if the available nutrients are biologically available.
+Low-flow events during warmer months show the lowest ammonium concentrations, which corresponds to seasonal maxima in chlorophyll concentration.
+A similar but weaker relationship was observed with nitrite/nitrate where increased flow was related to decreased concentration and lower concentrations overall were observed in the summer.
+However, low-flow events still had higher concentrations than high-flow events in July, as compared to ammonium which was low regardless of flow.
+This suggests that ammonium concentrations are driving phytoplankton production at P8.
+Annual trends in chlorophyll concentration (not shown) showed an overall decrease from the 1970s to present, although a slight peak is observed in the 2000s.
+This peak is likely related to the maximum ammonium concentration shown in \cref{fig:p8trnds}.
+Moreover, flow/chlorophyll relationships have generally been constant throughout the period of record such that a change in flow has not been related to a change in phytoplankton production.
+This suggests that nutrient loads that contribute to production at P8 are primarily from point sources at \ac{wwtp} outflows as a change in flow does not affect the load output.
+But what are watershed loads?
 
@@
 What about other variables (light level as suspended particulate matter, temperature)?
-
-\subsection{Selected examples}
-
-Two stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
-The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends over several years.
-The selected case studies focused on \begin{inparaenum}[1\upshape)]
-\item effects of wastewater treatment upgrades upstream of P8, and
-\item effects of biological invasion on nutrient dynamics in Suisun Bay.
-\end{inparaenum}
-Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
-
-\subsubsection{Effects of wastewater treatment}
-
-Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.
-As noted in \cite{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., )\cite{Cornwell14,Novick15}.
-Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}.
-Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
-\item a shift in load contributions before/after upgrade,
-\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
-\item different nitrogen species will have different changes depending on change in load outputs. 
-\end{inparaenum}
-The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
-%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
-
-Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:stock}).
-Nitrate is the primary constituent of total nitrogen after 2007.
-Organic nitrogen is a larger percentage of the total after nitrification.
-What was reduction in ammonium starting in 2002?
-
-Nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}), with ammonium showing the largest reduction.
-Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
-Percent changes are shown in \cref{tab:p8chg}, where both nitrogen species shows large percent increases prior to \ac{wwtp} upgrades followed by decreases after upgrades with ammonium showing the largest pecentage.
-Seasonally, increases prior to upgrades were most apparent in the \ac{jas} months for both analytes.
-Seasonal reductions post-upgrades were also largest in \ac{jas} for nitrite/nitrate, whereas percent reductions were similar across all monthly groupings for ammonium.
-
-Relationships of nitrogen with flow showed the typical inverse flow/concentration dynamic with flushing at high flow, although patterns differed by nitrogen species.
-Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations in the winter.
-Additionally, strength of the flow/nutrient relationship changed throughout the time series the year where the strongest relationship differed by analyte.
-Nitrite/nitrate typically had the strongest relationship flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
-
-\subsubsection{Effects of biological invasions}
-
-The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \cite{Cohen98}.
-The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \cite{Carlton90,Nichols90,Werner93,Crauder16}.
-In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
-Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}.
-The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
-
-We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
-\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
-\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
-\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
-\end{inparaenum}
-The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
-
-Data from \cite{Crauder16,Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
-Figure 10 in Ref.~\citenum{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
-
-Note the decrease in Potamocorbula abundance in 2011, 2012.
-These are wet years where abundance/biomass of the clams is driven down by lower salinity.
-Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.
-The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.
-This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
-See suggestions in Ref.~\citenum{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
-
-Further, chlorophyll trends early in the time series generally show a decrease with high flow with a distinct maximum at moderate flow.
-This may suggest stratification events at moderate flow contributed to phytoplankton blooms early in the time series.
-Water withdrawals later in the time series could have also altered environmental conditions to reduce the frequency occurrence of stratification events.
-Look into this more...
-
-What about biomass/density relationships for Potamocorbula?
-Although clam density increases throughout the period,  What about initial decrease in chlorophyll prior to clam invasion?
-Is this related to water withdrawals (i.e., decrease in stratification events at moderate flow)?
-
-\cref{fig:clmchl}, \cref{tab:d7chg}
-
-\subsection{Implications}
-
-Second case study showed typical inverse relationships between nutrients and flow, more flow means greater flushing and dilution of nutrient concentrations.
-Conversely, low flow means less flushing and higher nutrient concentrations, although this may not always be observed if the available nutrients are biologically available.
-Low-flow events during warmer months show the lowest ammonium concentrations, which corresponds to seasonal maxima in chlorophyll concentration.
-A similar but weaker relationship was observed with nitrite/nitrate where increased flow was related to decreased concentration and lower concentrations overall were observed in the summer.
-However, low-flow events still had higher concentrations than high-flow events in July, as compared to ammonium which was low regardless of flow.
-This suggests that ammonium concentrations are driving phytoplankton production at P8.
-Annual trends in chlorophyll concentration (not shown) showed an overall decrease from the 1970s to present, although a slight peak is observed in the 2000s.
-This peak is likely related to the maximum ammonium concentration shown in \cref{fig:p8trnds}.
-Moreover, flow/chlorophyll relationships have generally been constant throughout the period of record such that a change in flow has not been related to a change in phytoplankton production.
-This suggests that nutrient loads that contribute to production at P8 are primarily from point sources at \ac{wwtp} outflows as a change in flow does not affect the load output.
-But what are watershed loads?
 
@@
 <<p8chg>>=
-
-data(h2dat)
-dt <- as.Date('2006-01-01')
-
-noraw_bf <- filter(h2dat, resvar %in% 'no23') %>% 
-  .$data %>% 
-  .[[1]] %>% 
-  .[which(.$date < dt), ]
-noraw_af <- filter(h2dat, resvar %in% 'no23') %>% 
-  .$data %>% 
-  .[[1]] %>% 
-  .[which(.$date >= dt), ]
-nhraw_bf <- filter(h2dat, resvar %in% 'nh') %>% 
-  .$data %>% 
-  .[[1]] %>% 
-  .[which(.$date < dt), ]
-nhraw_af <- filter(h2dat, resvar %in% 'nh') %>% 
-  .$data %>% 
-  .[[1]] %>% 
-  .[which(.$date >= dt), ]
-
-mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
-yrbrks_bf <- c(1975, 2005)
-yrbrks_af <- c(2006, 2013)
-molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
-yrlabs_bf <- '1976-2006'
-yrlabs_af <- '2007-2013'
-faclevs <- c('1976-2006_before', '2007-2013_after', paste0(molabs, '_before'), paste0(molabs, '_after'))
-
-# summarize din, sio2, chl trends
-no_bf <- wrtdstrnd_sk(noraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf) %>% 
-  mutate(period = 'before', var = 'no')
-no_af <- wrtdstrnd_sk(noraw_af, mobrks, yrbrks_af, molabs, yrlabs_af) %>% 
-  mutate(period = 'after', var = 'no')
-nh_bf <- wrtdstrnd_sk(nhraw_bf, mobrks, yrbrks_bf, molabs, yrlabs_bf) %>% 
-  mutate(period = 'before', var = 'nh')
-nh_af <- wrtdstrnd_sk(nhraw_af, mobrks, yrbrks_af, molabs, yrlabs_af) %>% 
-  mutate(period = 'after', var = 'nh')
+# uses data from chunk in results test
 

### 2017-01-15 sftrends_manu.tex
almost done with first case study text

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 16:19:14 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/66d301f76d51fddfc7d4eeea2ed7b92995d7385f}{66d301f76d51fddfc7d4eeea2ed7b92995d7385f}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Sun Jan 15 13:39:13 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/82c4c81c70cde1a2ecfdf011fe349d76d42c89ec}{82c4c81c70cde1a2ecfdf011fe349d76d42c89ec}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
-Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.
-
-Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta. Huge effect on trend interpration
-
-What do nitrogen trends mean?
-Have to interpret relative to trends in other variables.
-A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
-A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
-However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
-Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
-See Senn slide 14 (from burial?).
-The WRTDS model lets us at least address trends in the context of season, time, and flow.
-This allows for more improved interpretion relative to observing raw data.
-Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
-What about other variables (light level as suspended particulate matter, temperature)?
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes.
+Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed.
+A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.
+For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
+For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramentao or San Joaquin daily flow estimates.
+Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each locaiton.
+Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.
 
@@
 
-Two stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
-The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends over several years.
+Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
 The selected case studies focused on \begin{inparaenum}[1\upshape)]
 \item effects of wastewater treatment upgrades upstream of P8, and
-\item effects of biological invasion on nutrient dynamics in Suisun Bay.
+\item effects of biological invasion on nutrient dynamics in Suisun Bay using D7 as an example.
 \end{inparaenum}
@@
 
-Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.
-As noted in \cite{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.
 Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., )\cite{Cornwell14,Novick15}.
-Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
+Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
 The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}.
 Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
-\item a shift in load contributions before/after upgrade,
-\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
-\item different nitrogen species will have different changes depending on change in load outputs. 
+\item a shift in nutrient contributions from the \ac{wwtp} before/after upgrade, and
+\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
 \end{inparaenum}
-The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
-%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}
 
-Overall reduction in total nitrogen load was observed as a result of reduction in ammonium (\cref{fig:stock}).
-Nitrate is the primary constituent of total nitrogen after 2007.
-Organic nitrogen is a larger percentage of the total after nitrification.
-What was reduction in ammonium starting in 2002?
+Effluent concentations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual shift in ammonium relative to the total (\cref{fig:stock}).
+Ammonium and nitrate concentrations were generally balanced prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite.
+As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase form 1976 followed by a large reduction in the 2000s.
+Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
+Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).
+Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%).
+Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate ($-4.3$\%) and largest for ammonium in the winter ($-26.7$\%).
 
-Nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}), with ammonium showing the largest reduction.
-Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
-Percent changes are shown in \cref{tab:p8chg}, where both nitrogen species shows large percent increases prior to \ac{wwtp} upgrades followed by decreases after upgrades with ammonium showing the largest pecentage.
-Seasonally, increases prior to upgrades were most apparent in the \ac{jas} months for both analytes.
-Seasonal reductions post-upgrades were also largest in \ac{jas} for nitrite/nitrate, whereas percent reductions were similar across all monthly groupings for ammonium.
-
-Relationships of nitrogen with flow showed the typical inverse flow/concentration dynamic with flushing at high flow, although patterns differed by nitrogen species.
-Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations in the winter.
-Additionally, strength of the flow/nutrient relationship changed throughout the time series the year where the strongest relationship differed by analyte.
-Nitrite/nitrate typically had the strongest relationship flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
+Relationships of nitrogen with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing at high flow (\cref{fig:p8trnds}b).
+Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations at low flow in the winter (January).
+Additionally, strength of the flow/nutrient relationship changed between years.
+Nitrite/nitrate typically had the strongest relationshipwith flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
+Using \ac{wrtds}, an empirical link is created between upstream changes and observed effects downstream that is characterized by differences in analyte beween years and season.
+A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
+These dynamics cannot be characterized from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
+For example, estimated ammonium concentrations in July were low for all flow levels which could suggest either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
+As such, trends from \ac{wrtds} can be used to develop additional hypotheses for further testing cause and effect mechanisms of nutrient impacts on adverse conditions.
 
@@
 
-\subsection{Implications}
-
 Second case study showed typical inverse relationships between nutrients and flow, more flow means greater flushing and dilution of nutrient concentrations.
@@
 But what are watershed loads?
+
+What do nitrogen trends mean?
+Have to interpret relative to trends in other variables.
+A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
+A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
+However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
+Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
+See Senn slide 14 (from burial?).
+The WRTDS model lets us at least address trends in the context of season, time, and flow.
+This allows for more improved interpretion relative to observing raw data.
+Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
+What about other variables (light level as suspended particulate matter, temperature)?
 

### 2017-01-16 sftrends_manu.Rnw
P8 case study text done

@@
 \item effects of wastewater treatment upgrades upstream of P8, and
-\item effects of biological invasion on nutrient dynamics in Suisun Bay using D7 as an example.
+\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7 as an example.
 \end{inparaenum}
@@
 As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase form 1976 followed by a large reduction in the 2000s.
-Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
+Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.
 Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium, \cref{tab:p8chg}).
@@
 
-Relationships of nitrogen with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing at high flow (\cref{fig:p8trnds}b).
+Relationships of nitrogen with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).
 Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations at low flow in the winter (January).
 Additionally, strength of the flow/nutrient relationship changed between years.
-Nitrite/nitrate typically had the strongest relationshipwith flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
-Using \ac{wrtds}, an empirical link is created between upstream changes and observed effects downstream that is characterized by differences in analyte beween years and season.
+Nitrite/nitrate typically had the strongest relationship with flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
+Using \ac{wrtds}, an empirical link is created between upstream changes and observed effects downstream that is characterized by differences in analytes between years and season.
 A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
-These dynamics cannot be characterized from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
-For example, estimated ammonium concentrations in July were low for all flow levels which could suggest either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
-As such, trends from \ac{wrtds} can be used to develop additional hypotheses for further testing cause and effect mechanisms of nutrient impacts on adverse conditions.
+These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
+For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer,or nitrogen was available and uptake by primary consumers was high.
+Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were generally associated with higher concentrations.
+This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the observed concentrations in July at low flow.
+As such, these simple observations from \ac{wrtds} can facilitate further testing of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to upstream management actions.
 
@@
 \end{inparaenum}
-The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
+The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
 
@@
 \cref{fig:clmchl}, \cref{tab:d7chg}
-
-Second case study showed typical inverse relationships between nutrients and flow, more flow means greater flushing and dilution of nutrient concentrations.
-Conversely, low flow means less flushing and higher nutrient concentrations, although this may not always be observed if the available nutrients are biologically available.
-Low-flow events during warmer months show the lowest ammonium concentrations, which corresponds to seasonal maxima in chlorophyll concentration.
-A similar but weaker relationship was observed with nitrite/nitrate where increased flow was related to decreased concentration and lower concentrations overall were observed in the summer.
-However, low-flow events still had higher concentrations than high-flow events in July, as compared to ammonium which was low regardless of flow.
-This suggests that ammonium concentrations are driving phytoplankton production at P8.
-Annual trends in chlorophyll concentration (not shown) showed an overall decrease from the 1970s to present, although a slight peak is observed in the 2000s.
-This peak is likely related to the maximum ammonium concentration shown in \cref{fig:p8trnds}.
-Moreover, flow/chlorophyll relationships have generally been constant throughout the period of record such that a change in flow has not been related to a change in phytoplankton production.
-This suggests that nutrient loads that contribute to production at P8 are primarily from point sources at \ac{wwtp} outflows as a change in flow does not affect the load output.
-But what are watershed loads?
 

### 2017-01-16 sftrends_manu.tex
P8 case study text done

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Sun Jan 15 13:39:13 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/82c4c81c70cde1a2ecfdf011fe349d76d42c89ec}{82c4c81c70cde1a2ecfdf011fe349d76d42c89ec}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Sun Jan 15 15:17:59 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/97e5231a2ff04fccd22ce6cba2309a1ba981b7d3}{97e5231a2ff04fccd22ce6cba2309a1ba981b7d3}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase form 1976 followed by a large reduction in the 2000s.
-Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease.
+Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.
 Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).

### 2017-01-16 sftrends_manu.Rnw
draft done, d7 table gone

@@
 This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the observed concentrations in July at low flow.
-As such, these simple observations from \ac{wrtds} can facilitate further testing of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to upstream management actions.
+As such, these simple observations from \ac{wrtds} can facilitate further testing of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
 
@@
 
-The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \cite{Cohen98}.
-The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \cite{Carlton90,Nichols90,Werner93,Crauder16}.
-In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
+Invasion of the upper \ac{sfe} estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
 Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
@@
 The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
-
 We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
@@
 
-Data from \cite{Crauder16,Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
-Figure 10 in Ref.~\citenum{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
-
-Note the decrease in Potamocorbula abundance in 2011, 2012.
-These are wet years where abundance/biomass of the clams is driven down by lower salinity.
-Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.
-The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.
-This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
-See suggestions in Ref.~\citenum{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
-
-Further, chlorophyll trends early in the time series generally show a decrease with high flow with a distinct maximum at moderate flow.
-This may suggest stratification events at moderate flow contributed to phytoplankton blooms early in the time series.
-Water withdrawals later in the time series could have also altered environmental conditions to reduce the frequency occurrence of stratification events.
-Look into this more...
-
-What about biomass/density relationships for Potamocorbula?
-Although clam density increases throughout the period,  What about initial decrease in chlorophyll prior to clam invasion?
-Is this related to water withdrawals (i.e., decrease in stratification events at moderate flow)?
-
-\cref{fig:clmchl}, \cref{tab:d7chg}
-
-What do nitrogen trends mean?
-Have to interpret relative to trends in other variables.
-A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
-A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
-However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
-Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
-See Senn slide 14 (from burial?).
-The WRTDS model lets us at least address trends in the context of season, time, and flow.
-This allows for more improved interpretion relative to observing raw data.
-Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
-What about other variables (light level as suspended particulate matter, temperature)?
+Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were mediated by a changes in flow over time (\cref{fig:clmchl}).
+Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
+amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
+The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
+A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) such that lower \ac{chla} was generally associated with higher biomass.
+However, changes in the \ac{chla}/flow relationship over time was dynamic such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f).
+Later in the time series, overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed.
+
+A general conclusion is that clam grazing reduced \ac{chla} concentration throughout the period of record, whereas the effect of flow as a top-down or bottom-up control on both was more dynamic.
+As an example, the relationship between flow and \ac{chla} earlier in the time period suggested a dilution effect at high flow and a peak of predicted \ac{chla} at moderate flows.
+In the absence of benthic grazing, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are moved downstream towards the lower bay.
+Following clam invasion later in the time series, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
+Concurrently, the increase in clam abundance shown in \ref{fig:clmchl}a is apparent, although variation between years is also observed.
+For example, clam abundance is reduced during high flow years in the late 1990s, 2006, and 2011 (\cref{fig:clmchl}a, flow in \cref{fig:obsdat}).
+In the same years, \ac{wrtds} predictions for \ac{chla} are higher than the flow-normalized component (\cref{fig:clmchl}c), which suggests a mobilization effect of increased flow on phytoplankton production.
+As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
+These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides additional to describe forcing factors of water quality in the Delta.
+
+As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.
+Changes in ambient conditions provide a preliminary but potentially incomplete description of trends depending on the question of interest.
+At a minimum, \ac{wrtds} focuses on high-level forcing factors of change by explicitly accounting for annual, seasonal, and flow effects in the description.
+We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trend sprovide a more powerful means of quantifying change.
+Combined with supporing data, \ac{wrtds} results can support additional hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
+Still, additional sources of variability must be considered as explicit factors that influence observed trends, including the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
+Statistical interpretations of multiple factors can provide quantitative links of nutrient loads with adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
 
@@
   mutate(
-    q = log(q),
     station = factor(station, levels = c('sac', 'sjr'), labels = c('Sac.', 'S. Jo.'))
     )
-ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+ylab <- expression(paste('Flow (', m^3, ' ', s^-1, ')'))
 p2 <- ggplot(toplo2, aes(x = Date, y = q)) +
@@
 <<clmchl, fig.height = 9, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
-Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means),with biomass linked to salinity (b).
+Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
 A coincident decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d).
@@
 
-% trends in din, chl, sio2 at d7
-<<d7chg>>=
-
-data(h3dat)
-
-d7dat <- filter(h3dat, Site_Code %in% 'D7')
-
-dinraw <- filter(d7dat, resvar %in% 'din') %>% 
-  .$data %>% 
-  .[[1]]
-chlraw <- filter(d7dat, resvar %in% 'chl') %>% 
-  .$data %>% 
-  .[[1]]
-sioraw <- filter(d7dat, resvar %in% 'sio2') %>% 
-  .$data %>% 
-  .[[1]]
-
-mobrks <- list(c(3, 4, 5), c(6, 7, 8), c(9, 10, 11), c(12, 1, 2))
-yrbrks <- c(1975, 1985, 1994, 2003, 2013)
-molabs <- c('Spring', 'Summer', 'Fall', 'Winter')
-yrlabs <- c('1976-1985', '1986-1994', '1995-2003', '2004-2013')
-
-# summarize din, sio2, chl trends
-din <- wrtdstrnd_sk(dinraw, mobrks, yrbrks, molabs, yrlabs)
-din <- wrtdstrnd_sk(dinraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
-  .[1, ] %>% 
-  rbind(., din) %>% 
-  data.frame(res = 'din', .)
-chl <- wrtdstrnd_sk(chlraw, mobrks, yrbrks, molabs, yrlabs)
-chl <- wrtdstrnd_sk(chlraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
-  .[1, ] %>% 
-  rbind(., chl) %>% 
-  data.frame(res = 'chl', .)
-sio <- wrtdstrnd_sk(sioraw, mobrks, yrbrks, molabs, yrlabs)
-sio <- wrtdstrnd_sk(sioraw, mobrks, yrbrks = c(-Inf, Inf), molabs, yrlabs = c('1976-2013')) %>% 
-  .[1, ] %>% 
-  rbind(., sio) %>% 
-  data.frame(res = 'sio', .)
-
-# combine for trends for table
-totab <- rbind(din, chl, sio) %>% 
-  select(res, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      perchg, 
-      paste0('\\textit{\\textbf{', perchg, '}}')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  gather('var', 'val', med:perchg) %>% 
-  unite('col', res, var) %>% 
-  spread(col, val) %>% 
-  mutate(cat = factor(cat, levels = c('1976-2013', yrlabs, molabs), labels = c('1976-2013', yrlabs, molabs))) %>% 
-  arrange(cat) %>% 
-  data.frame(sep = c(rep('Annual', 5), rep('Seasonal', 4)), .) %>% 
-  select(sep, cat, din_med, din_perchg, chl_med, chl_perchg, sio_med, sio_perchg)
-
-# final table formatting
-segs <- c('All', 'Annual', 'Seasonal')
-cats <- totab$cat
-totab <- totab[,-c(1:2)]
-vars <- c('DIN', 'Chl-\\textit{a}', 'SiO$_2$')
-names(totab) <- rep(c('Median', '\\% change'), 3)
-
-cap.val<-'Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-
-latex( 
-  totab,
-  file = '',
-  rowlabel = 'Period',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = segs,
-  n.rgroup = c(1, rep(4, 2)),
-  cgroup = vars,
-  n.cgroup = c(2, 2, 2),
-  rowname = cats,
-  label = 'tab:d7chg', 
-  insert.bottom = foot.val
-  )
-
-@
-
 \clearpage

### 2017-01-16 sftrends_manu.tex
draft done, d7 table gone

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Sun Jan 15 15:17:59 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/97e5231a2ff04fccd22ce6cba2309a1ba981b7d3}{97e5231a2ff04fccd22ce6cba2309a1ba981b7d3}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Mon Jan 16 07:59:41 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/586e750aea1bb38bbe57a715c876cade70cb62a9}{586e750aea1bb38bbe57a715c876cade70cb62a9}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
-Data from \cite{Crauder16,Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.
-Figure 10 in Ref.~\citenum{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.
+Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were mediated by a changes in flow over time (\cref{fig:clmchl}).
+Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
+amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
+The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
+A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) such that lower \ac{chla} was generally associated with higher biomass.
+However, changes in the \ac{chla}/flow relationship over time was dynamic such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f).
+Later in the time series, overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed.
 
-Note the decrease in Potamocorbula abundance in 2011, 2012.
-These are wet years where abundance/biomass of the clams is driven down by lower salinity.
-Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.
-The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.
-This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot).
-See suggestions in Ref.~\citenum{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
+A general conclusion is that clam grazing reduced \ac{chla} concentration throughout the period of record, whereas the effect of flow as a top-down or bottom-up control on both was more dynamic.
+As an example, the relationship between flow and \ac{chla} earlier in the time period suggested a dilution effect at high flow and a peak of predicted \ac{chla} at moderate flows.
+In the absence of benthic grazing, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are moved downstream towards the lower bay.
+Following clam invasion later in the time series, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
+Concurrently, the increase in clam abundance shown in \ref{fig:clmchl}a is apparent, although variation between years is also observed.
+For example, clam abundance is reduced during high flow years in the late 1990s, 2006, and 2011 (\cref{fig:clmchl}a, flow in \cref{fig:obsdat}).
+In the same years, \ac{wrtds} predictions for \ac{chla} are higher than the flow-normalized component (\cref{fig:clmchl}c), which suggests a mobilization effect of increased flow on phytoplankton production.
+As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
+These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides additional to describe forcing factors of water quality in the Delta.
 
-Further, chlorophyll trends early in the time series generally show a decrease with high flow with a distinct maximum at moderate flow.
-This may suggest stratification events at moderate flow contributed to phytoplankton blooms early in the time series.
-Water withdrawals later in the time series could have also altered environmental conditions to reduce the frequency occurrence of stratification events.
-Look into this more...
-
-What about biomass/density relationships for Potamocorbula?
-Although clam density increases throughout the period,  What about initial decrease in chlorophyll prior to clam invasion?
-Is this related to water withdrawals (i.e., decrease in stratification events at moderate flow)?
-
-\cref{fig:clmchl}, \cref{tab:d7chg}
-
-What do nitrogen trends mean?
-Have to interpret relative to trends in other variables.
-A decrease in nitrogen or constant nitrogen does not mean nitrogen inputs have stayed the same, they might actually be increasing if nitrogen.
-A change in chlorophyll relative to change in nitrogen could be informative, and even moreso, a change in silica relative to change in chlorophyll suggests diatom biomass has changed.
-However, there are mismatches in these trends that suggest other processes are at play, e.g., residence times and flow inputs, etc.
-Trends in Suisun relative to trends in Delta provide an example, e.g., Suisun is decrease in chlorophyll, increase in silica, increase in nitrogen, delta is decrease in silica, increase/decrease in DIN (depending on time period/season), decrease in chlorophyll, what's going on?
-See Senn slide 14 (from burial?).
-The WRTDS model lets us at least address trends in the context of season, time, and flow.
-This allows for more improved interpretion relative to observing raw data.
-Also explain more information by looking at ammonium, nitrative/nitrite, relative to DIN.
-What about other variables (light level as suspended particulate matter, temperature)?
+As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.
+Changes in ambient conditions provide a preliminary but potentially incomplete description of trends depending on the question of interest.
+At a minimum, \ac{wrtds} focuses on high-level forcing factors of change by explicitly accounting for annual, seasonal, and flow effects in the description.
+We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trend sprovide a more powerful means of quantifying change.
+Combined with supporing data, \ac{wrtds} results can support additional hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
+Still, additional sources of variability must be considered as explicit factors that influence observed trends, including the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
+Statistical interpretations of multiple factors can provide quantitative links of nutrient loads with adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
 
@@
 \caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
-Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means),with biomass linked to salinity (b).
+Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
 A coincident decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d).
@@
 ~~Winter&2.2&-0.8**&&0.2&-26.7**\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
-
-% trends in din, chl, sio2 at d7
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(1, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2, 2), rowname = cats,     label = "tab:d7chg", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in dissolved inorganic nitrogen (mg L$^{-1}$), chlorophyll ($\mu$g L$^{-1}$), and silicon dioxide (mg L$^{-1}$) concentrations for different time periods at station D7.
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:d7chg}}
-\begin{center}
-\begin{tabular}{lllcllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Chl-\textit{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries SiO$_2$}\tabularnewline
-\cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
-\hline
-{\bfseries All}&&&&&&&&\tabularnewline
-~~1976-2013&0.4&\textit{\textbf{0.6}}**&&3&-6.7**&&12.7&\textit{\textbf{0.7}}**\tabularnewline
-\hline
-{\bfseries Annual}&&&&&&&&\tabularnewline
-~~1976-1985&0.4&-2.1**&&8.8&-10.7**&&10.2&-0.2\tabularnewline
-~~1986-1994&0.4&\textit{\textbf{3.6}}**&&2.6&-13.5**&&11.9&\textit{\textbf{2.3}}**\tabularnewline
-~~1995-2003&0.5&-0.1&&1.8&\textit{\textbf{1.9}}**&&13.3&\textit{\textbf{0.7}}**\tabularnewline
-~~2004-2013&0.5&-1.3**&&2.1&\textit{\textbf{2.9}}**&&13.1&-0.3**\tabularnewline
-\hline
-{\bfseries Seasonal}&&&&&&&&\tabularnewline
-~~Spring&0.5&-0.1&&3.4&-1&&14.7&\textit{\textbf{0.1}}**\tabularnewline
-~~Summer&0.4&\textit{\textbf{1.5}}**&&3.4&-8.8**&&12.2&\textit{\textbf{1.2}}**\tabularnewline
-~~Fall&0.4&\textit{\textbf{0.6}}**&&1.7&-8.8**&&12.1&\textit{\textbf{1}}**\tabularnewline
-~~Winter&0.6&-0.2&&1.4&-3.1**&&14.5&\textit{\textbf{0.3}}**\tabularnewline
 \hline

### 2017-01-17 sftrends_manu.Rnw
working on minor text edits

@@
 library(english)
+library(forcats)
 @
@@
 As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years.
-Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes.
+Indicators of eutrophication can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
 For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}.
@@
 Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
-Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated recent development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}.
+Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}.
 These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
@@
 This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
-More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-infleunced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
-These studies have demonstrated potential for the use \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into drivers of change in aquatic systems.
+More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
+These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into eutrophication dynamics in aquatic systems.
 
@@
 The specific objectives were to \begin{inparaenum}[1\upshape)]
-\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation,
+\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
 \item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
 \end{inparaenum}
-The second objective evaluated two specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of \ac{wwtp} upgrades on water quality, and effects of biological invasion by benthic filter feeders on primary production.
-Although quantitative descriptions of change can be ends in themselves, the results were expected to have greater impact as a means to more detailed understanding of ecosystem properties.
+The second objective evaluated two specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
+Although quantitative descriptions of change can be ends in themselves, the results provide a means to more detailed understanding of ecosystem properties.
 Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.
-Overall, this analysis is expected to further an ecological and management-based understanding of dynamics in San Francisco Bay, with implications for water quality restoration and protection of this prominent system.
 
@@
 The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given many nutrients sources and the volume of water that is exchanged through the system with natural and anthropogenic processes.
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged through the system with natural and anthropogenic processes.
 A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}.
@@
 Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.
-The most censored observations were observed for ammonium time series at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\%), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).
+The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).
 
@@
 Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
-As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8.
-The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.
-Salinity has been used as a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.
+As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
+Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation given the stronger tidal influence at these stations.
+Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.
 
@@
 The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.
-The \ac{wrtds} model is a moving window regression that fits a unique set of parameters at each observation point in the time series.
-A unique set of weights is used for each regression to control the relevance of observations used to fit the model to the observation at the center of the window.
+The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
+A unique set of weights is used for each regression to control the importance of observations used to fit the model relative to the observation at the center of the window.
 The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow.
-The final vector used to fit the model at each point weights observations more similar to the center of the window with more importance.
 The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.
@@
 The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.
-All analyses used the WRTDStidal package for R \cite{Beck16b,RDCT16}.
+All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT16}.
 
 A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
-Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders\cite{Beck15}.
-Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
+Although variation in nutrients can be caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
 A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.
@@
 Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
-These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midway points in the time series.
-
-Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
+These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midpoints in the time series.
+
+Trends in each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
 This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82}.
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
-The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
-Seasonal Kendall tests were also used to describe trends in the model predictions for the observed data.
+The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
+Seasonal Kendall tests were also used to describe trends in the observed data.
 These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
@@
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisuan Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
 Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
@@
 For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}.
-Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.
-However, observed seasonality patterns were inconsistent between sites.
+Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.
 For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).
@@
 These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-Change over for time in observed data could reflect mobilization or dilution effects of flow on concentration rather than an interpretable system response to changes in nutrient sources.
+Change over time in observed data could reflect mobilization or dilution effects of flow on concentration rather than an interpretable system response to changes in nutrient sources.
 As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10.
@@
 In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
-The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
-For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.
-Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
+The difference between predicted and flow-normalized results is shown in \cref{fig:dinc10}c, such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
+For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrients.
+Further, \cref{fig:dinc10}e shows \ac{wrtds} estimates of seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
 Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.
-Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s.
-% The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}
-
+Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s, which is potentially linked to long-term climatic patterns\cite{Cloern07}.
 <<echo = F, cache = F>>=
@@
 A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
-These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
+These comparisons are made to identify changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
 For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
-Changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an inrease in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positve or less negative) compared to observed trends.
-More importantly, \Sexpr{eng(chgtb['postoneg'])} comparisons showed a trend reversal from positive to negative estimated change and \Sexpr{eng(chgtb['somechg'])} sites went from no change to negative estimated trends for the flow-normalized results.
-Differences by season in the observed relative to flow-normalized results from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
-The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013.
-
-Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes.
+This suggests that time series that include flow effects have sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
+Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.
+\Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change and \Sexpr{eng(chgtb['somechg'])} sites went from no change to negative estimated trends for the flow-normalized results.
+Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
+The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
+
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes., particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
 Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed.
@@
 For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramentao or San Joaquin daily flow estimates.
-Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each locaiton.
+Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
 Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.
@@
 \item effects of wastewater treatment upgrades upstream of P8, and
-\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7 as an example.
+\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.
 \end{inparaenum}
@@
 
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., )\cite{Cornwell14,Novick15}.
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc.)\cite{Cornwell14,Novick15}.
 Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}.
-Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 (\cref{fig:delt_map})\cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
+A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
 \item a shift in nutrient contributions from the \ac{wwtp} before/after upgrade, and
@@
 
-Effluent concentations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual shift in ammonium relative to the total (\cref{fig:stock}).
+Effluent concentations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to the total (\cref{fig:stock}).
 Ammonium and nitrate concentrations were generally balanced prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite.
@@
 Additionally, strength of the flow/nutrient relationship changed between years.
-Nitrite/nitrate typically had the strongest relationship with flow later in the time series, whereas ammonium had the strongest relationship with flow in the early 2000s.
+Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January.
 Using \ac{wrtds}, an empirical link is created between upstream changes and observed effects downstream that is characterized by differences in analytes between years and season.
@@
 These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
-For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer,or nitrogen was available and uptake by primary consumers was high.
+For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
 Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were generally associated with higher concentrations.
@@
 
-Invasion of the upper \ac{sfe} estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
-Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}.
+Invasion of the upper \ac{sfe} estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
+Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred with the most notable invasion effect being increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}.
 The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
 We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
-\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
-\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
-\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
+\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
+\item variation in the flow/chlorophyll relationship through indirect or direct controls.
 \end{inparaenum}
-The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
-
-Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were mediated by a changes in flow over time (\cref{fig:clmchl}).
+The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) provides an approach to assess the competing effects of climate variability, hydrology, and ecology on ambient conditions.
+
+Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were further affected by flow changes over time (\cref{fig:clmchl}).
 Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
@@
 As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.
-Changes in ambient conditions provide a preliminary but potentially incomplete description of trends depending on the question of interest.
-At a minimum, \ac{wrtds} focuses on high-level forcing factors of change by explicitly accounting for annual, seasonal, and flow effects in the description.
+At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects in the description.
 We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trend sprovide a more powerful means of quantifying change.
-Combined with supporing data, \ac{wrtds} results can support additional hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
-Still, additional sources of variability must be considered as explicit factors that influence observed trends, including the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
-Statistical interpretations of multiple factors can provide quantitative links of nutrient loads with adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
-
-\section{References} % omit on submit
+Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
+Still, additional sources of variability must be considered as explicit factors that influence observed trends and exploration of alternative time series analysis methods to address a wider range of questions should be the focus of further analyses.
+Additional factors to` consider could include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
+Statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
+
+% \section{References} % omit on submit
 \begin{singlespace}
@@
 
+\clearpage
+
 % overall trends, years
@@
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \\cref{fig:trndmap} for a summary of spatial trends.'
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.'
 foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
@@
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \\cref{fig:trndmap} for a summary of spatial trends.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'

### 2017-01-17 sftrends_manu.tex
working on minor text edits

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Mon Jan 16 07:59:41 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/586e750aea1bb38bbe57a715c876cade70cb62a9}{586e750aea1bb38bbe57a715c876cade70cb62a9}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Mon Jan 16 10:31:08 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/71477978e1431857840dcc15c6b5fcce60e9ffbd}{71477978e1431857840dcc15c6b5fcce60e9ffbd}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.
-The most censored observations were observed for ammonium time series at sites C10 (25.4\%), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).
+The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).
 
@@
 Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
-These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midway points in the time series.
+These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midpoints in the time series.
 
-Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
+Trends in each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
 This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82}.
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
-The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
-Seasonal Kendall tests were also used to describe trends in the model predictions for the observed data.
+The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
+Seasonal Kendall tests were also used to describe trends in the observed data.
 These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
@@
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
@@
 For example, average \ac{din} across all stations was $0.61$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.7$$\pm$$0.01$ for 1996-2013.
-Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.
-However, observed seasonality patterns were inconsistent between sites.
+Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.
 For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).
@@
 In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
-The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
-For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.
-Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
+The difference between predicted and flow-normalized results is shown in \cref{fig:dinc10}c, such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
+For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrients.
+Further, \cref{fig:dinc10}e shows \ac{wrtds} estimates of seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
 Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.
-Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s.
-% The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}
+Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s, which is potentially linked to long-term climatic patterns\cite{Cloern07}.
 
 A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
-These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
+These comparisons are made to identify changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
 For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
-Changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an inrease in magnitude (more negative or more positive) and twenty five had a decrease (less positve or less negative) compared to observed trends.
-More importantly, eleven comparisons showed a trend reversal from positive to negative estimated change and ten sites went from no change to negative estimated trends for the flow-normalized results.
-Differences by season in the observed relative to flow-normalized results from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
-The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013.
+This suggests that time series that include flow effects have sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
+Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an increase in magnitude (more negative or more positive) and twenty five had a decrease (less positive or less negative) compared to observed trends.
+Eleven comparisons showed a trend reversal from positive to negative estimated change and ten sites went from no change to negative estimated trends for the flow-normalized results.
+Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
+The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
 
-Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes.
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes., particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
 Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed.
@@
 
-Invasion of the upper \ac{sfe} estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.
-Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}.
+Invasion of the upper \ac{sfe} estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
+Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred with the most notable invasion effect being increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}.
 The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
@@
 \end{inparaenum}
-The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.
+The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) provides an approach to assess the competing effects of climate variability, hydrology, and ecology on ambient conditions.
 
@@
 As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.
-Changes in ambient conditions provide a preliminary but potentially incomplete description of trends depending on the question of interest.
-At a minimum, \ac{wrtds} focuses on high-level forcing factors of change by explicitly accounting for annual, seasonal, and flow effects in the description.
+At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects in the description.
 We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trend sprovide a more powerful means of quantifying change.
-Combined with supporing data, \ac{wrtds} results can support additional hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
-Still, additional sources of variability must be considered as explicit factors that influence observed trends, including the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
-Statistical interpretations of multiple factors can provide quantitative links of nutrient loads with adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
+Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
+Still, additional sources of variability must be considered as explicit factors that influence observed trends and exploration of alternative time series analysis methods to address a wider range of questions should be the focus of further analyses.
+Additional factors to` consider could include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
+Statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
 
-\section{References} % omit on submit
+% \section{References} % omit on submit
 \begin{singlespace}
@@
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \cref{fig:trndmap} for a summary of spatial trends.\label{tab:trndsann}}
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.\label{tab:trndsann}}
 \begin{center}
@@
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \cref{fig:trndmap} for a summary of spatial trends.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsbef}}
@@
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-See \cref{fig:trndmap} for a summary of spatial trends.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsaft}}

### 2017-01-18 sftrends_manu.Rnw
finished minor edits

@@
 These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-Change over time in observed data could reflect mobilization or dilution effects of flow on concentration rather than an interpretable system response to changes in nutrient sources.
 As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10.
@@
 
-Invasion of the upper \ac{sfe} estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
+Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
 Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
@@
 \item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
-\item variation in the flow/chlorophyll relationship through indirect or direct controls.
+\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
 \end{inparaenum}
@@
 A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) such that lower \ac{chla} was generally associated with higher biomass.
-However, changes in the \ac{chla}/flow relationship over time was dynamic such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f).
-Later in the time series, overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed.
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) with lower \ac{chla} associated with higher biomass.
+The \ac{chla}/flow relationship changed over time such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
 
 A general conclusion is that clam grazing reduced \ac{chla} concentration throughout the period of record, whereas the effect of flow as a top-down or bottom-up control on both was more dynamic.
-As an example, the relationship between flow and \ac{chla} earlier in the time period suggested a dilution effect at high flow and a peak of predicted \ac{chla} at moderate flows.
-In the absence of benthic grazing, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are moved downstream towards the lower bay.
+The relationship between flow and \ac{chla} earlier in the time period suggested a dilution effect at high flow and peak \ac{chla} at moderate flows.
+In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
 Following clam invasion later in the time series, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
-Concurrently, the increase in clam abundance shown in \ref{fig:clmchl}a is apparent, although variation between years is also observed.
-For example, clam abundance is reduced during high flow years in the late 1990s, 2006, and 2011 (\cref{fig:clmchl}a, flow in \cref{fig:obsdat}).
-In the same years, \ac{wrtds} predictions for \ac{chla} are higher than the flow-normalized component (\cref{fig:clmchl}c), which suggests a mobilization effect of increased flow on phytoplankton production.
+The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed (\ref{fig:clmchl}a).
+For example, clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\cref{fig:obsdat}).
+In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.
 As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
-These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides additional to describe forcing factors of water quality in the Delta.
+These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
 
 As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.
-At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects in the description.
-We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trend sprovide a more powerful means of quantifying change.
+At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.
+We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.
 Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
 Still, additional sources of variability must be considered as explicit factors that influence observed trends and exploration of alternative time series analysis methods to address a wider range of questions should be the focus of further analyses.
-Additional factors to` consider could include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
+Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
 Statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
@@
 \caption{The \acl{sfe} Delta and monitoring stations used for analysis.
-The Delta drains the combined watershed from the Sacramento and San Joaquin rivers (bottom left).
+The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (bottom left).
 All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.}
@@
 \end{figure}
-
-<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
-Data were collected monthly and evaluated with \\ac{wrtds} models using daily flow estimates from 1976 to 2013.
-Note different y-axis scales.
-See \\cref{fig:delt_map} for station locations.', fig.height = 7.5, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
-data(delt_dat)
-data(flow_dat)
-
-## 
-# obs data
-
-missvals <- c('1996-01-01', '2004-04-01') %>% 
-  as.Date
-missvals <- seq(missvals[1], missvals[2], by = 'month')
-missvals <- data.frame(Site_Code = 'D19', Date = missvals, nh = 0, no23 = 0)
-
-toplo <- delt_dat %>% 
-  select(Site_Code, Date, nh, no23) %>% 
-  rbind(., missvals) %>% 
-  gather('var', 'val', nh:no23) %>% 
-  mutate(Site_Code = factor(Site_Code, 
-    levels = rev(c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'MD10', 'C10', 'C3')))
-    )
-
-mythm <- theme_minimal() +
-  theme(
-    legend.position = 'top', 
-    legend.title = element_blank(),
-    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.title.x = element_blank(),
-    axis.text.y = element_text(size = 6),
-    axis.ticks.length = unit(.1, "cm"), 
-    panel.grid.major = element_line(size = 0.25),
-    panel.grid.minor = element_blank(),
-    plot.margin = grid::unit(c(3, 6, 3, 6), 'pt')
-  )
-
-cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
-  rev
-ylab <- expression(paste('DIN (mg ', L^-1, ')'))
-
-leglabs <- list(
-  expression(NH [4] ^ '+'),
-  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
-  )
-
-p1 <- ggplot(toplo, aes(x = Date)) +
-  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
-  scale_x_date(expand = c(0, 0)) + 
-  scale_y_continuous(ylab, expand = c(0, 0)) + 
-  facet_grid(Site_Code~., scales = 'free_y') +
-  scale_fill_manual(values = cols, labels = leglabs) +
-  mythm %+replace% 
-  theme(axis.text.x = element_blank())
-
-## 
-# flo data
-toplo2 <- filter(flow_dat, 
-    !station %in% 'east' &
-    Date >= as.Date('1976-01-13') &
-    Date <= as.Date('2013-12-13')
-  ) %>% 
-  mutate(
-    station = factor(station, levels = c('sac', 'sjr'), labels = c('Sac.', 'S. Jo.'))
-    )
-ylab <- expression(paste('Flow (', m^3, ' ', s^-1, ')'))
-p2 <- ggplot(toplo2, aes(x = Date, y = q)) +
-  geom_line(size = 0.25) +
-  scale_x_date(expand = c(0, 0)) + 
-  scale_y_continuous(ylab) + 
-  facet_grid(station~., scales = 'free_y') + 
-  mythm
-
-# fix widths
-# align widths of plots in first column, first two
-pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p2))
-
-maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
-
-pA$widths[2:3] <- maxWidth
-pB$widths[2:3] <- maxWidth
-
-grid.arrange(pA, pB, ncol = 1, heights = c(1, 0.21))
-@
 
@@
 trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
-  select(Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
+  select(Location, Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
   filter(!trndvar %in% 'fits') %>% 
   mutate(
+    Location = ifelse(
+      resvar %in% 'din' & Site_Code %in% c('MD10', 'D28', 'D7') & cat %in% c('1976-1995', 'Spring') & trndvar == 'res', 
+      Location , ''),
     Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3')),
@@
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
+  geom_text(aes(label = Location, x = -3.5), hjust = 0.2) + 
   scale_y_discrete('Stations')
@@
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
+  geom_text(aes(label = Location, x = -4), hjust = 0.2) + 
   scale_y_discrete('Stations')
@@
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
+  geom_text(aes(label = Location, x = -6), hjust = 0.2) + 
   scale_y_discrete('Stations')
@@
 Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
-Subfigure (e) shows the modelled relationships between \\ac{din}, flow, and time.
-Flow values in (e) are truncated by 5\\textsuperscript{th} and 95\\textsuperscript{th} percentiles of observed.'>>=
+Subfigure (e) shows the modelled relationships between \\ac{din}, flow (5\\textsuperscript{th} and 95\\textsuperscript{th} percentiles), and time.'>>=
 
@@
 
-<<p8trnds, fig.height = 8, fig.width = 9, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \\ac{wrtds} (a, bottom), and relationships with flow over time (b).
+<<p8trnds, fig.height = 8, fig.width = 9, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \\ac{wrtds}, and (b) relationships with flow over time from \\ac{wrtds}.
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.'>>=
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\\cref{fig:stock}).'>>=
 data(h2dat)
@@
 Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
-A coincident decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d).
+A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d) based on \\ac{wrtds} results.
 A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (e).
-Flow relationships with \\ac{chla} concentration have also changed over time (f, observations from June).'>>=
+Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (f, observations from June).'>>=
 # clam and model data at D7
@@
 
+<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
+Data were collected monthly and evaluated with \\ac{wrtds} models using daily flow estimates from 1976 to 2013.
+Note different y-axis scales.
+See \\cref{fig:delt_map} for station locations.', fig.height = 7.5, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
+data(delt_dat)
+data(flow_dat)
+
+## 
+# obs data
+
+missvals <- c('1996-01-01', '2004-04-01') %>% 
+  as.Date
+missvals <- seq(missvals[1], missvals[2], by = 'month')
+missvals <- data.frame(Site_Code = 'D19', Date = missvals, nh = 0, no23 = 0)
+
+toplo <- delt_dat %>% 
+  select(Site_Code, Date, nh, no23) %>% 
+  rbind(., missvals) %>% 
+  gather('var', 'val', nh:no23) %>% 
+  mutate(Site_Code = factor(Site_Code, 
+    levels = rev(c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'MD10', 'C10', 'C3')))
+    )
+
+mythm <- theme_minimal() +
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank(),
+    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.title.x = element_blank(),
+    axis.text.y = element_text(size = 6),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.major = element_line(size = 0.25),
+    panel.grid.minor = element_blank(),
+    plot.margin = grid::unit(c(3, 6, 3, 6), 'pt')
+  )
+
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
+  rev
+ylab <- expression(paste('DIN (mg ', L^-1, ')'))
+
+leglabs <- list(
+  expression(NH [4] ^ '+'),
+  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
+  )
+
+p1 <- ggplot(toplo, aes(x = Date)) +
+  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
+  scale_x_date(expand = c(0, 0)) + 
+  scale_y_continuous(ylab, expand = c(0, 0)) + 
+  facet_grid(Site_Code~., scales = 'free_y') +
+  scale_fill_manual(values = cols, labels = leglabs) +
+  mythm %+replace% 
+  theme(axis.text.x = element_blank())
+
+## 
+# flo data
+toplo2 <- filter(flow_dat, 
+    !station %in% 'east' &
+    Date >= as.Date('1976-01-13') &
+    Date <= as.Date('2013-12-13')
+  ) %>% 
+  mutate(
+    station = factor(station, levels = c('sac', 'sjr'), labels = c('Sac.', 'S. Jo.'))
+    )
+ylab <- expression(paste('Flow (', m^3, ' ', s^-1, ')'))
+p2 <- ggplot(toplo2, aes(x = Date, y = q)) +
+  geom_line(size = 0.25) +
+  scale_x_date(expand = c(0, 0)) + 
+  scale_y_continuous(ylab) + 
+  facet_grid(station~., scales = 'free_y') + 
+  mythm
+
+# fix widths
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+
+maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
+
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+
+grid.arrange(pA, pB, ncol = 1, heights = c(1, 0.21))
+@
+
 \begin{figure}

### 2017-01-18 sftrends_manu.tex
finished minor edits

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Mon Jan 16 10:31:08 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/71477978e1431857840dcc15c6b5fcce60e9ffbd}{71477978e1431857840dcc15c6b5fcce60e9ffbd}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Jan 17 17:20:06 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/11311aa862b4ee4a7c5d118a1736c0a932b3a26a}{11311aa862b4ee4a7c5d118a1736c0a932b3a26a}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
-\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
-\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
-\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
+\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
+\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
 \end{inparaenum}
@@
 
-Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were mediated by a changes in flow over time (\cref{fig:clmchl}).
+Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were further affected by flow changes over time (\cref{fig:clmchl}).
 Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
@@
 \caption{The \acl{sfe} Delta and monitoring stations used for analysis.
-The Delta drains the combined watershed from the Sacramento and San Joaquin rivers (bottom left).
+The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (bottom left).
 All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.}
 \label{fig:delt_map}   
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=0.9\textwidth]{figs/obsdat-1}
-
-}
-
-\caption{Observed \ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
-Data were collected monthly and evaluated with \ac{wrtds} models using daily flow estimates from 1976 to 2013.
-Note different y-axis scales.
-See \cref{fig:delt_map} for station locations.}\label{fig:obsdat}
 \end{figure}
@@
 Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
-Subfigure (e) shows the modelled relationships between \ac{din}, flow, and time.
-Flow values in (e) are truncated by 5\textsuperscript{th} and 95\textsuperscript{th} percentiles of observed.}\label{fig:dinc10}
+Subfigure (e) shows the modelled relationships between \ac{din}, flow (5\textsuperscript{th} and 95\textsuperscript{th} percentiles), and time.}\label{fig:dinc10}
 \end{figure}
@@
 
-\caption{Nitrogen trends at P8 as observed (a, top), predicted and flow-normalized estimates from \ac{wrtds} (a, bottom), and relationships with flow over time (b).
+\caption{Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}.
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\cref{fig:stock}), coincident with a dramatic decrease in ammonium at P8.}\label{fig:p8trnds}
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\cref{fig:stock}).}\label{fig:p8trnds}
 \end{figure}
@@
 Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
-A coincident decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d).
+A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d) based on \ac{wrtds} results.
 A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (e).
-Flow relationships with \ac{chla} concentration have also changed over time (f, observations from June).}\label{fig:clmchl}
+Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (f, observations from June).}\label{fig:clmchl}
 \end{figure}
@@
 \beginsupplement
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=0.9\textwidth]{figs/obsdat-1}
+
+}
+
+\caption{Observed \ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
+Data were collected monthly and evaluated with \ac{wrtds} models using daily flow estimates from 1976 to 2013.
+Note different y-axis scales.
+See \cref{fig:delt_map} for station locations.}\label{fig:obsdat}
+\end{figure}
 

### 2017-01-18 sftrends_manu.Rnw
minor edits again

@@
 
-Effluent concentations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to the total (\cref{fig:stock}).
+Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to the total (\cref{fig:stock}).
 Ammonium and nitrate concentrations were generally balanced prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite.
@@
 Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were generally associated with higher concentrations.
-This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the observed concentrations in July at low flow.
-As such, these simple observations from \ac{wrtds} can facilitate further testing of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
+This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow.
+As such, these simple observations from \ac{wrtds} provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
 

### 2017-01-18 sftrends_manu.tex
minor edits again

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Jan 17 17:20:06 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/11311aa862b4ee4a7c5d118a1736c0a932b3a26a}{11311aa862b4ee4a7c5d118a1736c0a932b3a26a}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Jan 18 13:43:59 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/85bc23721ff60c8b574e2e771764191be69cf211}{85bc23721ff60c8b574e2e771764191be69cf211}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}

### 2017-02-02 sftrends_manu.Rnw
abstract shortened to within est limits

@@
 \begin{abstract}
-\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of long-term time series.
-Combined with these methods, multi-decadal datasets of water quality in coastal systems can provide valuable opportunities to gain insights into ecosystem properties and drivers of change.
-This study describes use of an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}.
-This region is a complex mosaic of inflows that are primary sources of nutrients into the larger Bay.
-To date, a comprehensive evaluation of flow-normalized trends using the long-term monitoring dataset at multiple stations in the Delta has not been conducted despite the importance of nutrient transport from the region for water quality in the entire bay.
-The \ac{wrtds} technique is data-driven where the parameterization of the functional model changes smoothly over time following dynamic patterns of season and flow.
-Water quality trends that have not been previously quantified can be described, including variation in flow-normalized concentrations, frequency occurrence of extreme events, and response to historical changes in the watershed, all of which are important needs for understanding changes in the \ac{sfe}.
-Model results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
-This variation was described in the context of varying contributions of input flows from the Sacramento and San Joaquin rivers, as well as tidal exchange with the central \ac{sfe}.
-Conceptual relationships between water quality and drivers of change were used to generate and test hypotheses of mechanistic relationships using selected examples from the trend descriptions.
-Overall, this analysis provides an ecological and management-based understanding of historical trends in the \ac{sfe} as a means to interpret potential impacts of recent changes and expected trends in this dynamic system.
-An argument is also made for more comprehensive evaluations of long-term monitoring datasets to understand relationships between response endpoints and causal mechanisms in coastal waters.
+\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.
+We apply an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}.
+Results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
+Trend analysis with \ac{wrtds} flow-normalized data demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
+We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) effects of biological invasion on chlorophyll dynamics in Suisun Bay.
+Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
 \end{abstract}

### 2017-02-02 sftrends_manu.tex
abstract shortened to within est limits

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Jan 18 13:43:59 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/85bc23721ff60c8b574e2e771764191be69cf211}{85bc23721ff60c8b574e2e771764191be69cf211}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Jan 18 13:48:31 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/49568114dc54151cfc21515b651815124b3db256}{49568114dc54151cfc21515b651815124b3db256}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}

### 2017-02-09 sftrends_manu.Rnw
Pellerin14 added to refs

@@
 These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
-The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Zhang16}.
+The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
 This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.

### 2017-02-09 sftrends_manu.tex
Pellerin14 added to refs

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Jan 18 13:48:31 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/49568114dc54151cfc21515b651815124b3db256}{49568114dc54151cfc21515b651815124b3db256}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 2 16:00:40 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/4afb16203c36788d9f1a8c7b6ee43ed973d94495}{4afb16203c36788d9f1a8c7b6ee43ed973d94495}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}

### 2017-02-27 sftrends_manu.Rnw
fig 1 with whole bay

@@
 Trend analysis with \ac{wrtds} flow-normalized data demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
-We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) effects of biological invasion on chlorophyll dynamics in Suisun Bay.
+We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
 Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
@@
 Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.
-Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.
-Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
-As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years.
-Indicators of eutrophication can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
+Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.
+Climate, local, and regional environmental effects may act individually or together over time to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
+As a practical approach for water quality evaluation, trend analysis of ecosystem repsonse indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
+Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
 For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}.
 Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}.
-Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
-
-Appropriate methods for the analysis of change depend largely on the question of interest and on characteristics of the environmental dataset.
-Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}.
+Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
+
+Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.
+Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}.
 Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
@@
 More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
-These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into eutrophication dynamics in aquatic systems.
+These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.
+Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.
 
 The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \cite{Cloern15}.
-Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.
-Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
-The unique ecological and social context of the Bay provides a valuable opportunity to gain insight into ecosystem properties of estuaries that define water quality dynamics at different scales.
-The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay \cite{Jassby00,Jassby02,Jassby08}.
-A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades \cite{Jabusch16}.
-Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting an approach that explicitly considers flow effects is critical for trend analysis.
-To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
-
-The goal of this study was to provide a comprehensive description of nutrient trends in the Delta to inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay.
+Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
+Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
+The unique ecological and social context of the region provides a valuable opportunity to gain insight into ecosystem properties that define water quality dynamics at different scales.
+In particular, the Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \ac{sfe} \cite{Jassby00,Jassby02,Jassby08}.
+A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.
+Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
+To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
+
+The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
 We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts.
@@
 \end{inparaenum}
-The second objective evaluated two specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
+The second objective evaluated two specific water quality stations as additional case studies o demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
 Although quantitative descriptions of change can be ends in themselves, the results provide a means to more detailed understanding of ecosystem properties.
@@
 
-The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest bay on the Pacific coast of North America.
-The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land in the Central Valley.
-Water enters the Bay through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
-The \ac{sfe} system is divided into several sub-bays, including Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.
+The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest estuary on the Pacific coast of North America.
+The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.
+Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
+The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.
 Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.

### 2017-02-27 sftrends_manu.tex
fig 1 with whole bay

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 2 16:00:40 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/4afb16203c36788d9f1a8c7b6ee43ed973d94495}{4afb16203c36788d9f1a8c7b6ee43ed973d94495}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 9 09:38:49 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/dc55a82ecaa6c7579c80e824377410a6b61dc53f}{dc55a82ecaa6c7579c80e824377410a6b61dc53f}}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.
-Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.
-Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
-As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years.
-Indicators of eutrophication can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
+Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.
+Climate, local, and regional environmental effects may act individually or together over time to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
+As a practical approach for water quality evaluation, trend analysis of ecosystem repsonse indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
+Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
 For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}.
 Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}.
-Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
+Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
 
-Appropriate methods for the analysis of change depend largely on the question of interest and on characteristics of the environmental dataset.
-Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}.
+Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.
+Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}.
 Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
@@
 More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
-These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into eutrophication dynamics in aquatic systems.
+These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.
+Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.
 
 The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \cite{Cloern15}.
-Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.
-Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
-The unique ecological and social context of the Bay provides a valuable opportunity to gain insight into ecosystem properties of estuaries that define water quality dynamics at different scales.
-The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay \cite{Jassby00,Jassby02,Jassby08}.
-A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades \cite{Jabusch16}.
-Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting an approach that explicitly considers flow effects is critical for trend analysis.
-To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
+Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
+Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
+The unique ecological and social context of the region provides a valuable opportunity to gain insight into ecosystem properties that define water quality dynamics at different scales.
+In particular, the Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \ac{sfe} \cite{Jassby00,Jassby02,Jassby08}.
+A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.
+Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
+To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
 
-The goal of this study was to provide a comprehensive description of nutrient trends in the Delta to inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay.
+The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
 We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts.

### 2017-02-28 sftrends_manu.Rnw
sort of done with jabusch edits

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+Environmental Protection Agency.}}
 
@@
 This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
-More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
+More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
 These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.
@@
 
-The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest estuary on the Pacific coast of North America.
+The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.
 The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.
 Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
-The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.
-Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
-Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \cite{Cloern12b}.
-The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns.
+The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).
+Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
+Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}.
 Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.
-Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \cite{Cloern07}.
-
-Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication (e.g., Chesapeake Bay).\cite{Kemp05}  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
-Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.
+
+Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.
+Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.
 Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \cite{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.
-These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \cite{Cloern07}, biological invasions \cite{Cohen98}, and departures from the historical flow record \cite{Enright09,Cloern12b}.
-The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
-
-The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).
-The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged through the system with natural and anthropogenic processes.
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \cite{Thompson08,Crauder16}.
+However, recent water quality trends have suggested that resistence to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.
+These recent changes have been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
+
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
 A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}.
-For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.
+For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.
 Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
@@
 
-Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.
+Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{SFE} region.
 The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.
@@
 Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.
-Data were minimally processed with the exception of averaging replicates that occurred on the same day.
+Data were minimally processed, with the exception of averaging replicates that occurred on the same day.
 The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
@@
 As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
-Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation given the stronger tidal influence at these stations.
+Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.
 Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.
@@
 Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.
-For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).
+For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).
 
@@
 Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).
-More sites had decreasing trends for the observed data from 1996-2013.
+Decreasing trends were more common for the observed data from 1996-2013.
 \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing).
-Trends by location (upper Delta, middle, and Suisun stations) were not apparent, suggesting individual sites had trends that differed independent of relative location.
-For example, P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites.
+P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section.
 Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
@@
 
-Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
-These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10.
-Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b.
-In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
-The difference between predicted and flow-normalized results is shown in \cref{fig:dinc10}c, such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
-For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrients.
-Further, \cref{fig:dinc10}e shows \ac{wrtds} estimates of seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
-Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.
-Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s, which is potentially linked to long-term climatic patterns\cite{Cloern07}.
 <<echo = F, cache = F>>=
@@
       ifelse(signobs == 1 & signnrm == 1 & nrm < obs, 'lesspos', 
-        'somechg')
-    )))))
-  )
+      ifelse(signobs == 0 & signnrm == 1, 'zerotopos', 
+      ifelse(signobs == 0 & signnrm == -1, 'zerotoneg',
+      ifelse(signobs == -1 & signnrm == 0, 'negtozero', 
+      ifelse(signobs == 1 & signnrm == 0, 'postozero', 
+        NA)
+    )))))))))
+  )
+
 # table of results for shorter sexpr
@@
 @
-
-A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
-These comparisons are made to identify changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
+Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
+These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
+As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
 For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
-This suggests that time series that include flow effects have sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
+This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
 Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.
-\Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change and \Sexpr{eng(chgtb['somechg'])} sites went from no change to negative estimated trends for the flow-normalized results.
+\Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results.
 Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
@@
 Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes., particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
-Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed.
+Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
 A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.
 For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
-For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramentao or San Joaquin daily flow estimates.
+For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.
 Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
@@
 Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to the total (\cref{fig:stock}).
-Ammonium and nitrate concentrations were generally balanced prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite.
-As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase form 1976 followed by a large reduction in the 2000s.
+Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite.
+As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.
 Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.
@@
 
-Relationships of nitrogen with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).
-Seasonal variation was more apparent for ammonium, although both typically had the highest concentrations at low flow in the winter (January).
+Relationships of nirite/nitrate with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).
+Seasonal variation was even more apparent for ammonium, although both nitrite/nitrate and ammonium typically had the highest concentrations at low flow in the winter (January).
 Additionally, strength of the flow/nutrient relationship changed between years.
@@
 Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
-Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
+Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
 Changes in the physical environment have also occurred with the most notable invasion effect being increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}.
@@
 A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) with lower \ac{chla} associated with higher biomass.
-The \ac{chla}/flow relationship changed over time such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}  The \ac{chla}/flow relationship changed over time such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
 
@@
 Following clam invasion later in the time series, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
-The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed (\ref{fig:clmchl}a).
-For example, clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\cref{fig:obsdat}).
+The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.
+For example, clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a).
 In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.
@@
 Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
-Still, additional sources of variability must be considered as explicit factors that influence observed trends and exploration of alternative time series analysis methods to address a wider range of questions should be the focus of further analyses.
 Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
-Statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
+For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow-normalized results.
+Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models).\cite{Beck17}  Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
 
@@
 \includegraphics[width=0.8\textwidth,page=1]{figs/delt_map.pdf}
-\caption{The \acl{sfe} Delta and monitoring stations used for analysis.
-The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (bottom left).
+\caption{The \acl{sfe} and Delta region with monitoring stations used for analysis.
+The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset).
 All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.}
@@
 \end{figure}
+
+<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
+Data were collected monthly and evaluated with \\ac{wrtds} models using daily flow estimates from 1976 to 2013.
+Note different y-axis scales.
+See \\cref{fig:delt_map} for station locations.', fig.height = 7.5, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
+data(delt_dat)
+data(flow_dat)
+
+## 
+# obs data
+
+missvals <- c('1996-01-01', '2004-04-01') %>% 
+  as.Date
+missvals <- seq(missvals[1], missvals[2], by = 'month')
+missvals <- data.frame(Site_Code = 'D19', Date = missvals, nh = 0, no23 = 0)
+
+toplo <- delt_dat %>% 
+  select(Site_Code, Date, nh, no23) %>% 
+  rbind(., missvals) %>% 
+  gather('var', 'val', nh:no23) %>% 
+  mutate(Site_Code = factor(Site_Code, 
+    levels = rev(c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'MD10', 'C10', 'C3')))
+    )
+
+mythm <- theme_minimal() +
+  theme(
+    legend.position = 'top', 
+    legend.title = element_blank(),
+    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.title.x = element_blank(),
+    axis.text.y = element_text(size = 6),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.major = element_line(size = 0.25),
+    panel.grid.minor = element_blank(),
+    plot.margin = grid::unit(c(3, 6, 3, 6), 'pt')
+  )
+
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
+  rev
+ylab <- expression(paste('DIN (mg ', L^-1, ')'))
+
+leglabs <- list(
+  expression(NH [4] ^ '+'),
+  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
+  )
+
+p1 <- ggplot(toplo, aes(x = Date)) +
+  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
+  scale_x_date(expand = c(0, 0)) + 
+  scale_y_continuous(ylab, expand = c(0, 0)) + 
+  facet_grid(Site_Code~., scales = 'free_y') +
+  scale_fill_manual(values = cols, labels = leglabs) +
+  mythm %+replace% 
+  theme(axis.text.x = element_blank())
+
+## 
+# flo data
+toplo2 <- filter(flow_dat, 
+    !station %in% 'east' &
+    Date >= as.Date('1976-01-13') &
+    Date <= as.Date('2013-12-13')
+  ) %>% 
+  mutate(
+    station = factor(station, levels = c('sac', 'sjr'), labels = c('Sac.', 'S. Jo.'))
+    )
+ylab <- expression(paste('Flow (', m^3, ' ', s^-1, ')'))
+p2 <- ggplot(toplo2, aes(x = Date, y = q)) +
+  geom_line(size = 0.25) +
+  scale_x_date(expand = c(0, 0)) + 
+  scale_y_continuous(ylab) + 
+  facet_grid(station~., scales = 'free_y') + 
+  mythm
+
+# fix widths
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+
+maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
+
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+
+grid.arrange(pA, pB, ncol = 1, heights = c(1, 0.21))
+@
 
@@
     panel.grid.minor = element_blank(), 
-    legend.position = 'none'
+    legend.position = 'none', 
+    legend.text = element_text(size = 12), 
+    axis.title = element_text(size = 12)
   )
@@
 \end{figure}
-
-<<dinc10, fig.height = 9, fig.width = 7, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.
-Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} for the conditional mean response.
-The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.
-Subfigure (c) shows the difference between the model predictions and flow-normalized predictions.
-Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
-Subfigure (e) shows the modelled relationships between \\ac{din}, flow (5\\textsuperscript{th} and 95\\textsuperscript{th} percentiles), and time.'>>=
-
-# data
-data(h1dat)
-
-# din at c10
-toplo <- filter(h1dat, resvar == 'din') %>% 
-  .$data %>% 
-  .[[1]]
-
-# globals
-ylab <- expression(paste('DIN (mg ', L^-1, ')'))
-ptheme <- theme_minimal() + 
-  theme(
-    axis.line.x = element_line(),
-    axis.line.y = element_line(), 
-    axis.title.x = element_blank(),
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"), 
-    legend.position = 'top',    
-    legend.title = element_blank(),
-    axis.text.x = element_blank(),
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-  )
-col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
-col_lim <- c(0, 3.8)
-xlims <- range(toplo$date)
-txt <- data.frame(
-  x = max(toplo$date) - 300, 
-  y =  c(5, 2.4, 0.6, 7, 5.85), 
-  txt = c('(a)', '(b)', '(c)', '(d)', '(e)')
-)
-size <- 5
-
-# observed din
-toplo1 <- select(toplo, date, res) %>% 
-  na.omit()
-p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
-  geom_line() + 
-  scale_y_continuous(ylab) +
-  geom_text(data = txt[1, ], aes(x = x, y = y, label = txt), size = size) +
-  scale_x_date(limits = xlims, expand = c(0, 0)) +
-  ptheme
-
-# din prdnrmplot
-p2 <- prdnrmplot(toplo, logspace = F, col_vec = col_vec) + 
-  scale_y_continuous(ylab) +
-  geom_text(data = txt[2, ], aes(x = x, y = y, label = txt, colour = NULL), show.legend = F, size = size) +
-  scale_x_date(limits = xlims, expand = c(0, 0)) +
-  ptheme
-
-# flo/observed diffs
-toplo2 <- prdnrmplot(toplo, plot = F)
-toplo2 <- left_join(toplo2$fits, toplo2$nrms, by = c('date')) %>% 
-  mutate(flodiff = fits_variable - nrms_variable)
-toplo3 <- toplo
-toplo3$flo <- scales::rescale(toplo3$flo, to = attr(toplo, 'floobs_rng'))
-ylab <- expression(paste('Pred - norm DIN (mg ', L^-1, ')'))
-p3 <- ggplot(toplo2, aes(x = date, y = flodiff)) + 
-  geom_bar(stat = 'identity') + 
-  scale_y_continuous(ylab) +
-  scale_x_date(limits = xlims, expand = c(0, 0)) +
-  geom_text(data = txt[3, ], aes(x = x, y = y, label = txt), size = size) +
-  ptheme
-
-# flo with loess
-ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
-p4 <- ggplot(toplo3, aes(x = date, y = flo)) + 
-  geom_line() +
-  stat_smooth(se = F) + 
-  scale_y_continuous(ylab) +
-  scale_x_date(limits = range(toplo3$date), expand = c(0, 0)) +
-  geom_text(data = txt[4, ], aes(x = x, y = y, label = txt), size = size) +
-  ptheme
-
-# dynaplot, all
-txte <- txt[5, , drop = FALSE]
-txte[, 'x'] <- lubridate::decimal_date(txte[, 'x'])
-txte$res <- 0
-attr(toplo, 'reslab') <- expression(paste("DIN (mg ", L^-1, ")"))
-ylab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
-p5 <- gridplot(toplo, month = 'all', logspace = F, floscl = F) +
-  scale_y_continuous(ylab, expand = c(0, 0)) +
-  geom_text(data = txte, aes(x = x, y = y, label = txt), size = size) +
-  theme(
-    plot.margin = grid::unit(c(6, 6, 0, 6), 'pt')
-    )
-
-# align widths of plots in first column, first two
-pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p2))
-pC <- ggplot_gtable(ggplot_build(p3))
-pD <- ggplot_gtable(ggplot_build(p4))
-pE <- ggplot_gtable(ggplot_build(p5))
-maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[2:3], pD$widths[2:3], pE$widths[2:3])
-
-pA$widths[2:3] <- maxWidth
-pB$widths[2:3] <- maxWidth
-pC$widths[2:3] <- maxWidth
-pD$widths[2:3] <- maxWidth
-pE$widths[2:3] <- maxWidth
-
-grid.arrange(
-  pA, pB, pC, pD, pE,
-  heights = c(1, 1.3, 1, 1, 1.5),
-  ncol = 1
-  )
-
-@
 
@@
 
-<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
-Data were collected monthly and evaluated with \\ac{wrtds} models using daily flow estimates from 1976 to 2013.
-Note different y-axis scales.
-See \\cref{fig:delt_map} for station locations.', fig.height = 7.5, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
-data(delt_dat)
-data(flow_dat)
-
-## 
-# obs data
-
-missvals <- c('1996-01-01', '2004-04-01') %>% 
-  as.Date
-missvals <- seq(missvals[1], missvals[2], by = 'month')
-missvals <- data.frame(Site_Code = 'D19', Date = missvals, nh = 0, no23 = 0)
-
-toplo <- delt_dat %>% 
-  select(Site_Code, Date, nh, no23) %>% 
-  rbind(., missvals) %>% 
-  gather('var', 'val', nh:no23) %>% 
-  mutate(Site_Code = factor(Site_Code, 
-    levels = rev(c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'MD10', 'C10', 'C3')))
-    )
-
-mythm <- theme_minimal() +
-  theme(
-    legend.position = 'top', 
-    legend.title = element_blank(),
-    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.title.x = element_blank(),
-    axis.text.y = element_text(size = 6),
-    axis.ticks.length = unit(.1, "cm"), 
-    panel.grid.major = element_line(size = 0.25),
-    panel.grid.minor = element_blank(),
-    plot.margin = grid::unit(c(3, 6, 3, 6), 'pt')
-  )
-
-cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
-  rev
-ylab <- expression(paste('DIN (mg ', L^-1, ')'))
-
-leglabs <- list(
-  expression(NH [4] ^ '+'),
-  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
-  )
-
-p1 <- ggplot(toplo, aes(x = Date)) +
-  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
-  scale_x_date(expand = c(0, 0)) + 
-  scale_y_continuous(ylab, expand = c(0, 0)) + 
-  facet_grid(Site_Code~., scales = 'free_y') +
-  scale_fill_manual(values = cols, labels = leglabs) +
-  mythm %+replace% 
-  theme(axis.text.x = element_blank())
-
-## 
-# flo data
-toplo2 <- filter(flow_dat, 
-    !station %in% 'east' &
-    Date >= as.Date('1976-01-13') &
-    Date <= as.Date('2013-12-13')
-  ) %>% 
-  mutate(
-    station = factor(station, levels = c('sac', 'sjr'), labels = c('Sac.', 'S. Jo.'))
-    )
-ylab <- expression(paste('Flow (', m^3, ' ', s^-1, ')'))
-p2 <- ggplot(toplo2, aes(x = Date, y = q)) +
-  geom_line(size = 0.25) +
-  scale_x_date(expand = c(0, 0)) + 
-  scale_y_continuous(ylab) + 
-  facet_grid(station~., scales = 'free_y') + 
-  mythm
-
-# fix widths
-# align widths of plots in first column, first two
-pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p2))
-
-maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
-
-pA$widths[2:3] <- maxWidth
-pB$widths[2:3] <- maxWidth
-
-grid.arrange(pA, pB, ncol = 1, heights = c(1, 0.21))
-@
-
 \begin{figure}

### 2017-02-28 sftrends_manu.tex
sort of done with jabusch edits

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 9 09:38:49 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/dc55a82ecaa6c7579c80e824377410a6b61dc53f}{dc55a82ecaa6c7579c80e824377410a6b61dc53f}}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Mon Feb 27 14:06:24 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/874a6ef7bdc555832bfe2094074db64176390d6d}{874a6ef7bdc555832bfe2094074db64176390d6d}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+Environmental Protection Agency.}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
-The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest estuary on the Pacific coast of North America.
+The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.
 The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.
 Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
-The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.
-Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
-Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \cite{Cloern12b}.
-The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns.
+The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).
+Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
+Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}.
 Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.
-Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \cite{Cloern07}.
 
-Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication (e.g., Chesapeake Bay).\cite{Kemp05}  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.
-Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.
+Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.
+Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.
 Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \cite{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.
-These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \cite{Cloern07}, biological invasions \cite{Cohen98}, and departures from the historical flow record \cite{Enright09,Cloern12b}.
-The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \cite{Thompson08,Crauder16}.
+However, recent water quality trends have suggested that resistence to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.
+These recent changes have been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
 
-The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).
-The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged through the system with natural and anthropogenic processes.
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
 A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}.
-For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.
+For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.
 Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
@@
 Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).
-More sites had decreasing trends for the observed data from 1996-2013.
+Decreasing trends were more common for the observed data from 1996-2013.
 Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing).
-Trends by location (upper Delta, middle, and Suisun stations) were not apparent, suggesting individual sites had trends that differed independent of relative location.
-For example, P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites.
+P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section.
 Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
@@
 These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10.
-Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b.
-In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.
-The difference between predicted and flow-normalized results is shown in \cref{fig:dinc10}c, such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.
-For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrients.
-Further, \cref{fig:dinc10}e shows \ac{wrtds} estimates of seasonal variation in the relationships of \ac{din} with flow throughout the period of record.
-Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.
-Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s, which is potentially linked to long-term climatic patterns\cite{Cloern07}.
-
-A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.
-These comparisons are made to identify changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making.
+As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data was made to identify changes in the magnitude, significance, and direction of trends.
 For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
@@
 Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an increase in magnitude (more negative or more positive) and twenty five had a decrease (less positive or less negative) compared to observed trends.
-Eleven comparisons showed a trend reversal from positive to negative estimated change and ten sites went from no change to negative estimated trends for the flow-normalized results.
+Eleven comparisons showed a trend reversal from positive to negative estimated change, nine sites went from no change to negative estimated change, and one site went from no change to a positive trend for the flow-normalized results.
 Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
@@
 \includegraphics[width=0.8\textwidth,page=1]{figs/delt_map.pdf}
-\caption{The \acl{sfe} Delta and monitoring stations used for analysis.
-The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (bottom left).
+\caption{The \acl{sfe} and Delta region with monitoring stations used for analysis.
+The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset).
 All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.}
 \label{fig:delt_map}   
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=0.9\textwidth]{figs/obsdat-1}
+
+}
+
+\caption{Observed \ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
+Data were collected monthly and evaluated with \ac{wrtds} models using daily flow estimates from 1976 to 2013.
+Note different y-axis scales.
+See \cref{fig:delt_map} for station locations.}\label{fig:obsdat}
 \end{figure}
@@
 \label{fig:trndcomp1}   
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=0.75\textwidth]{figs/dinc10-1}
-
-}
-
-\caption{Time series of \ac{din} and flow at station C10.
-Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} for the conditional mean response.
-The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.
-Subfigure (c) shows the difference between the model predictions and flow-normalized predictions.
-Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend.
-Subfigure (e) shows the modelled relationships between \ac{din}, flow (5\textsuperscript{th} and 95\textsuperscript{th} percentiles), and time.}\label{fig:dinc10}
 \end{figure}
@@
 \beginsupplement
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=0.9\textwidth]{figs/obsdat-1}
-
-}
-
-\caption{Observed \ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \ac{sfe} Delta and flow from the Sacramento and San Joaquin rivers.
-Data were collected monthly and evaluated with \ac{wrtds} models using daily flow estimates from 1976 to 2013.
-Note different y-axis scales.
-See \cref{fig:delt_map} for station locations.}\label{fig:obsdat}
-\end{figure}
 

### 2017-02-28 sftrends_manu.Rnw
changed terms for station groupings

@@
 Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
-Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+Stations were grouped by location in the study area for comparison: peripheral stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior stations D19, D26, D28; and Suisun stations D4, D6, and D7.
 These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.
@@
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
 Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
@@
   mutate(
+    Location = gsub('Middle', 'Interior', Location), 
+    Location = gsub('Delta', 'Peripheral', Location),
     Location = ifelse(
@@
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
-  geom_text(aes(label = Location, x = -3.5), hjust = 0.2) + 
+  geom_text(aes(label = Location, x = -3.4), hjust = 0.1) + 
   scale_y_discrete('Stations')
@@
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
-  geom_text(aes(label = Location, x = -4), hjust = 0.2) + 
+  geom_text(aes(label = Location, x = -4), hjust = 0.1) + 
   scale_y_discrete('Stations')
@@
   geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
-  geom_text(aes(label = Location, x = -6), hjust = 0.2) + 
+  geom_text(aes(label = Location, x = -6), hjust = 0.1) + 
   scale_y_discrete('Stations')

### 2017-02-28 sftrends_manu.tex
changed terms for station groupings

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Mon Feb 27 14:06:24 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/874a6ef7bdc555832bfe2094074db64176390d6d}{874a6ef7bdc555832bfe2094074db64176390d6d}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Feb 28 10:33:48 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/e98d78ebdf5300498e36d9e0bc2fb4746fdd7f3c}{e98d78ebdf5300498e36d9e0bc2fb4746fdd7f3c}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}
@@
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
@@
 These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data was made to identify changes in the magnitude, significance, and direction of trends.
+As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
 For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
-This suggests that time series that include flow effects have sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
+This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
 Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an increase in magnitude (more negative or more positive) and twenty five had a decrease (less positive or less negative) compared to observed trends.

### 2017-02-28 sftrends_manu.Rnw
Jabusch edits done

@@
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
 Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
@@
 The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
-A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}  The \ac{chla}/flow relationship changed over time such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
-
-A general conclusion is that clam grazing reduced \ac{chla} concentration throughout the period of record, whereas the effect of flow as a top-down or bottom-up control on both was more dynamic.
-The relationship between flow and \ac{chla} earlier in the time period suggested a dilution effect at high flow and peak \ac{chla} at moderate flows.
+A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
+An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}f).
+These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer (also shown in \cref{fig:clmchl}d), whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}
+
+The effect of flow as a top-down or bottom-up control changed throughout the time series.
+The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}h), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
 In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
@@
 
-<<clmchl, fig.height = 9, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
+<<clmchl, fig.height = 9.3, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
 Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
-A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d) based on \\ac{wrtds} results.
-A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (e).
-Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (f, observations from June).'>>=
+A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (d) based on \\ac{wrtds} results.
+Reductions in \\ac{chla} concentration were coincident with an increase in SiO$_2$ concentration (e), with the greatest increase in August (f).
+A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (g).
+Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (h, observations from June).'>>=
+
 # clam and model data at D7
@@
 data(flow_dat)
+
 chld7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'chl')$mod[[1]]
+sio2d7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'sio2')$mod[[1]]
 
@@
 
-ylab <- expression(paste(italic(Potamocorbula), " biomass (g ", m^-2, ")"))
+ylab <- expression(paste(italic(Potamocorbula), " (g ", m^-2, ")"))
 xlab <- expression(paste("Sacramento daily flow (", m^3, ' ', s^-1, ")"))
@@
 
-# prdnrmplot
+# prdnrmplot, chl
 ylab <- expression(paste("ln-chlorophyll (", italic(mu), "g ", L^-1, ")"))
@@
 ##
+# sio2 plots
+
+# prdnrmplot, sio2
+# ylab <- expression(paste("ln-chlorophyll (", italic(mu), "g ", L^-1, ")"))
+
+p5 <- prdnrmplot(sio2d7, predicted = F, col_vec = cols, min_mo = 10) + 
+  mythm + 
+  scale_x_date('Year') +
+  # scale_y_continuous(ylab) +
+  labs(title = '(e)')
+
+# seasyr plot, sio2
+p6 <- seasyrplot(sio2d7, predicted = F, col_vec = cols, min_mo = 10) + 
+  theme(
+    axis.line.x = element_line(), 
+    axis.line.y = element_line(), 
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    legend.title = element_blank(), 
+    panel.border = element_blank(), 
+    axis.title.x = element_text(),
+    plot.title = element_text(hjust = thjust, vjust = tvjust),
+    plot.margin = pmar
+    ) + 
+  scale_x_date('Day of year', date_labels = "%m/%d") +
+  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) +
+  # scale_y_continuous(ylab) + 
+  labs(title = '(f)')
+
+##
 # chlorophyll by month
@@
 # scatter plot of concentration x biomass with model 
-p5 <- ggplot(toplo, aes(x = biomass, y = bt_norm)) + 
+p7 <- ggplot(toplo, aes(x = biomass, y = bt_norm)) + 
   geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
@@
   scale_y_continuous(ylab, limits = c(0.8, 4)) +
-  labs(title = '(e)')
+  labs(title = '(g)')
 
@@
 # dynaplot for chl, june
-p6 <- dynaplot(chld7, mo = 6, col_vec = cols, floscl = F, size = 0.5) +
+p8 <- dynaplot(chld7, mo = 6, col_vec = cols, floscl = F, size = 0.5) +
   mythm %+replace%
@@
   guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) + 
-  labs(title = '(f)')
+  labs(title = '(h)')
 
@@
 pC <- ggplot_gtable(ggplot_build(p5))
-maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[c(2:3)])
+pD <- ggplot_gtable(ggplot_build(p7))
+maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[c(2:3)], pD$widths[c(2:3)])
 
@@
 pC$widths[2:3] <- maxWidth
+pD$widths[2:3] <- maxWidth
 
 # align widths of plots in second column
-pD <- ggplot_gtable(ggplot_build(p2))
-pE <- ggplot_gtable(ggplot_build(p4))
-pF <- ggplot_gtable(ggplot_build(p6))
-maxWidth = unit.pmax(pD$widths[2:3], pE$widths[2:3], pF$widths[2:3])
+pE <- ggplot_gtable(ggplot_build(p2))
+pF <- ggplot_gtable(ggplot_build(p4))
+pG <- ggplot_gtable(ggplot_build(p6))
+pH <- ggplot_gtable(ggplot_build(p8))
+maxWidth = unit.pmax(pE$widths[2:3], pF$widths[2:3], pG$widths[2:3], pH$widths[2:3])
 
 # Set the widths
-pD$widths[2:3] <- maxWidth
 pE$widths[2:3] <- maxWidth
 pF$widths[2:3] <- maxWidth
+pG$widths[2:3] <- maxWidth
+pH$widths[2:3] <- maxWidth
 
@@
 grid.arrange(
-  arrangeGrob(pA, pB, pC, ncol = 1, heights = c(1, 1, 1)), 
-  arrangeGrob(pD, pE, pF, ncol = 1, heights = c(1, 1, 1)), 
+  arrangeGrob(pA, pB, pC, pD, ncol = 1, heights = c(1, 1, 1, 1)), 
+  arrangeGrob(pE, pF, pG, pH, ncol = 1, heights = c(1, 1, 1, 1)), 
   ncol = 2, widths = c(1, 1)

### 2017-02-28 sftrends_manu.tex
Jabusch edits done

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Feb 28 10:33:48 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/e98d78ebdf5300498e36d9e0bc2fb4746fdd7f3c}{e98d78ebdf5300498e36d9e0bc2fb4746fdd7f3c}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Feb 28 11:10:13 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/9449165d97e2f8ca7f7f1d47d3d92efe451d669a}{9449165d97e2f8ca7f7f1d47d3d92efe451d669a}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}
@@
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
@@
 The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
-A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}e) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}  The \ac{chla}/flow relationship changed over time such that increasing flow (decreasing salinity) showed a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}f), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
+A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
+An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}f).
+These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer (also shown in \cref{fig:clmchl}d), whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}
 
-A general conclusion is that clam grazing reduced \ac{chla} concentration throughout the period of record, whereas the effect of flow as a top-down or bottom-up control on both was more dynamic.
-The relationship between flow and \ac{chla} earlier in the time period suggested a dilution effect at high flow and peak \ac{chla} at moderate flows.
+The effect of flow as a top-down or bottom-up control changed throughout the time series.
+The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}h), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
 In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
@@
 Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
-A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends(d) based on \ac{wrtds} results.
-A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (e).
-Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (f, observations from June).}\label{fig:clmchl}
+A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (d) based on \ac{wrtds} results.
+Reductions in \ac{chla} concentration were coincident with an increase in SiO$_2$ concentration (e), with the greatest increase in August (f).
+A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (g).
+Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (h, observations from June).}\label{fig:clmchl}
 \end{figure}

### 2017-03-01 sftrends_manu.Rnw
showlabels removed

@@
 \usepackage{setspace}
-\usepackage{showlabels}
+% \usepackage{showlabels}
 
@@
 
-Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{SFE} region.
+Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{sfe} region.
 The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.

### 2017-03-01 sftrends_manu.tex
showlabels removed

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Feb 28 11:10:13 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/9449165d97e2f8ca7f7f1d47d3d92efe451d669a}{9449165d97e2f8ca7f7f1d47d3d92efe451d669a}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Feb 28 16:33:33 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/6213ed69c4113cd2ccdf0561273537009854ec0b}{6213ed69c4113cd2ccdf0561273537009854ec0b}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}

### 2017-03-03 sftrends_manu.Rnw
DIN eval back to clam case study

@@
 Climate, local, and regional environmental effects may act individually or together over time to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
-As a practical approach for water quality evaluation, trend analysis of ecosystem repsonse indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
+As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
 Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
@@
 These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
-The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
+The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
 This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
@@
 
-The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \cite{Cloern15}.
-Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
-Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
-The unique ecological and social context of the region provides a valuable opportunity to gain insight into ecosystem properties that define water quality dynamics at different scales.
-In particular, the Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \ac{sfe} \cite{Jassby00,Jassby02,Jassby08}.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \cite{Jassby00,Jassby02,Jassby08}.
+Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.
+Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
+Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
 A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.
@@
 \item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
-\item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
+\item provide detailed descriptions of two case studies in the context of conceptual relationships modeled with \ac{wrtds}.
 \end{inparaenum}
-The second objective evaluated two specific water quality stations as additional case studies o demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
-Although quantitative descriptions of change can be ends in themselves, the results provide a means to more detailed understanding of ecosystem properties.
+The second objective evaluated two specific water quality stations as additional case studies to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
+Although quantitative descriptions of change can be ends in themselves, the results provide means to a more detailed understanding of ecosystem properties.
 Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.
@@
 The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.
-The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.
-Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
-The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).
+The watershed provides drinking water to over 25 million people and includes irrigation for 18 thousand km$^2$ of agricultural land.
+Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
+The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).
 Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
@@
 Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
-Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.
-Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
+Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows and water exports directly from the system \cite{Jassby00,Jassby08}.
+Our analysis explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
 
@@
 All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.
-Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).
-Stations were grouped by location in the study area for comparison: peripheral stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior stations D19, D26, D28; and Suisun stations D4, D6, and D7.
-These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.
+Nutrient time series from 1976 to 2013 were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) for ten discrete sampling stations (\cref{fig:delt_map}).
+Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.
 Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.
@@
 \noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
-The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.
+The seasonal trend is modeled as a sinusoidal component to capture periodicity between years.
 The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
@@
 As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.
-The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method\cite{Beck15}.
-The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
-Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme events.
-The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.
+The \ac{wrtds} models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
 All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT16}.
@@
 
-The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
+The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
 In general, long-term (inter-annual) trends were not easily observed from the raw data.
@@
 A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
-\item a shift in nutrient contributions from the \ac{wwtp} before/after upgrade, and
+\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
 \item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
@@
 
-Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to the total (\cref{fig:stock}).
-Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite.
+Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).
+Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite.
 As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.
@@
 The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
-A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
-An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}f).
-These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer (also shown in \cref{fig:clmchl}d), whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
-This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}
-
-The effect of flow as a top-down or bottom-up control changed throughout the time series.
-The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}h), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
+A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).
+An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).
+These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (\cref{fig:clmchl}e, h) such that an increase in nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
+Overall, these results suggest the increase in nutrient concentrations are from decreased uptake by phytoplankton in Suisun Bay.
+
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
+However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
+The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
 In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
-Following clam invasion later in the time series, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
+Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
 The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.
-For example, clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a).
+Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a).
 In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.
@@
 \end{singlespace}
+\clearpage
+
+%%%%%%
+% figures
 
@@
 Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
-A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (d) based on \\ac{wrtds} results.
-Reductions in \\ac{chla} concentration were coincident with an increase in SiO$_2$ concentration (e), with the greatest increase in August (f).
-A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (g).
-Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (h, observations from June).'>>=
+A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \\ac{wrtds} results.
+Reductions in \\ac{chla} concentration were coincident with an increase in SiO$_2$ and \\ac{din} concentrations (d, e), with the greatest increases in August (g, h).
+A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (i).
+Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (j, observations from June).'>>=
 
@@
 sio2d7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'sio2')$mod[[1]]
+dind7 <- filter(h3dat, Site_Code %in% 'D7' & resvar %in% 'din')$mod[[1]]
 
@@
 p3 <- prdnrmplot(chld7, predicted = F, col_vec = cols, min_mo = 10) + 
-  mythm + 
+  mythm %+replace% 
+  theme(
+    axis.title.y = element_blank()
+    ) + 
+  guides(colour = guide_legend(override.aes = list(alpha = 0), label.theme = element_text(colour = 'white', angle = 0, size = 8))) +
+  scale_x_date('') +
+  labs(title = '(c)')
+
+# seasyrplot
+p6 <- seasyrplot(chld7, predicted = F, col_vec = cols, min_mo = 10) + 
+  mythm %+replace% 
+  theme(
+    legend.position = 'top', 
+    axis.title.y = element_blank()
+    ) + 
+  scale_x_date('', date_labels = "%m/%d") +
+  guides(colour = guide_colourbar(barheight = 0.25, barwidth = 0, label.theme = element_text(colour = 'white', angle = 0, size = 8))) +
+  scale_y_continuous(ylab) + 
+  labs(title = '(f)')
+
+##
+# sio2 plots
+
+# prdnrmplot, sio2
+p4 <- prdnrmplot(sio2d7, predicted = F, col_vec = cols, min_mo = 10) + 
+  mythm %+replace% 
+  theme(
+    axis.title.y = element_blank()
+    ) + 
+  guides(colour = guide_legend(override.aes = list(shape = c(16, NA), linetype = c(0, 1)), label.theme = element_text(colour = 'black', angle = 0, size = 8))) +
   scale_x_date('Year') +
-  scale_y_continuous(ylab) +
-  labs(title = '(c)')
-
-# seasyrplot
-p4 <- seasyrplot(chld7, predicted = F, col_vec = cols, min_mo = 10) + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"), 
-    legend.title = element_blank(), 
-    panel.border = element_blank(), 
-    axis.title.x = element_text(),
-    plot.title = element_text(hjust = thjust, vjust = tvjust),
-    plot.margin = pmar
+  labs(title = '(d)')
+
+# seasyr plot, sio2
+p7 <- seasyrplot(sio2d7, predicted = F, col_vec = cols, min_mo = 10) + 
+  mythm %+replace% 
+  theme(
+    axis.title.y = element_blank()
     ) + 
   scale_x_date('Day of year', date_labels = "%m/%d") +
-  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) +
-  scale_y_continuous(ylab) + 
-  labs(title = '(d)')
+  guides(colour = guide_colourbar(barheight = 0.25, barwidth = 7, label.theme = element_text(colour = 'black', angle = 0, size = 8))) +
+  labs(title = '(g)')
 
 ##
-# sio2 plots
-
-# prdnrmplot, sio2
-# ylab <- expression(paste("ln-chlorophyll (", italic(mu), "g ", L^-1, ")"))
-
-p5 <- prdnrmplot(sio2d7, predicted = F, col_vec = cols, min_mo = 10) + 
-  mythm + 
-  scale_x_date('Year') +
-  # scale_y_continuous(ylab) +
+# din plots
+p5 <- prdnrmplot(dind7, predicted = F, col_vec = cols) + 
+  mythm %+replace% 
+  theme(
+    axis.title.y = element_blank()
+    ) +  
+  guides(colour = guide_legend(override.aes = list(alpha = 0), label.theme = element_text(colour = 'white', angle = 0, size = 8))) +
+  scale_x_date('') +
   labs(title = '(e)')
 
-# seasyr plot, sio2
-p6 <- seasyrplot(sio2d7, predicted = F, col_vec = cols, min_mo = 10) + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"), 
-    legend.title = element_blank(), 
-    panel.border = element_blank(), 
-    axis.title.x = element_text(),
-    plot.title = element_text(hjust = thjust, vjust = tvjust),
-    plot.margin = pmar
+# seasyr plot, din
+p8 <- seasyrplot(dind7, predicted = F, col_vec = cols) + 
+  mythm %+replace% 
+  theme(
+    axis.title.y = element_blank()
     ) + 
-  scale_x_date('Day of year', date_labels = "%m/%d") +
-  guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) +
-  # scale_y_continuous(ylab) + 
-  labs(title = '(f)')
+  scale_x_date('', date_labels = "%m/%d") +
+  guides(colour = guide_colourbar(barheight = 0, barwidth = 7, label.theme = element_text(colour = 'white', angle = 0, size = 8))) +
+  labs(title = '(h)')
 
@@
 # scatter plot of concentration x biomass with model 
-p7 <- ggplot(toplo, aes(x = biomass, y = bt_norm)) + 
+p9 <- ggplot(toplo, aes(x = biomass, y = bt_norm)) + 
   geom_point(size = 3, alpha = 0.8, colour = 'darkgrey') + 
@@
   scale_y_continuous(ylab, limits = c(0.8, 4)) +
-  labs(title = '(g)')
+  labs(title = '(i)')
 
@@
 # dynaplot for chl, june
-p8 <- dynaplot(chld7, mo = 6, col_vec = cols, floscl = F, size = 0.5) +
+p10 <- dynaplot(chld7, mo = 6, col_vec = cols, floscl = F, size = 0.5) +
   mythm %+replace%
@@
   guides(colour = guide_colourbar(barheight = 0.5, barwidth = 10)) + 
-  labs(title = '(h)')
+  labs(title = '(j)')
 
@@
 
-# align widths of plots in first column
+# (1, 1), (2, 1), (3, 1), (4, 1)
 pA <- ggplot_gtable(ggplot_build(p1))
-pB <- ggplot_gtable(ggplot_build(p3))
-pC <- ggplot_gtable(ggplot_build(p5))
-pD <- ggplot_gtable(ggplot_build(p7))
-maxWidth = unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[c(2:3)], pD$widths[c(2:3)])
-
-# Set the widths
+pC <- ggplot_gtable(ggplot_build(p3))
+pF <- ggplot_gtable(ggplot_build(p6))
+pI <- ggplot_gtable(ggplot_build(p9))
+maxWidth = unit.pmax(pA$widths[2:3], pI$widths[2:3])
 pA$widths[2:3] <- maxWidth
+pI$widths[2:3] <- maxWidth
+maxWidth = unit.pmax(pC$widths[2:3], pF$widths[2:3])
+pC$widths[2:3] <- maxWidth
+pF$widths[2:3] <- maxWidth
+
+# (1, 2), (4, 2) 
+pB <- ggplot_gtable(ggplot_build(p2))
+pJ <- ggplot_gtable(ggplot_build(p10))
+maxWidth = unit.pmax(pB$widths[2:3], pJ$widths[c(2:3)])
 pB$widths[2:3] <- maxWidth
-pC$widths[2:3] <- maxWidth
+pJ$widths[2:3] <- maxWidth
+
+# (2, 2), (3, 2)
+pD <- ggplot_gtable(ggplot_build(p4))
+pG   <- ggplot_gtable(ggplot_build(p7))
+maxWidth = unit.pmax(pD$widths[2:3], pG$widths[c(2:3)])
 pD$widths[2:3] <- maxWidth
-
-# align widths of plots in second column
-pE <- ggplot_gtable(ggplot_build(p2))
-pF <- ggplot_gtable(ggplot_build(p4))
-pG <- ggplot_gtable(ggplot_build(p6))
+pG$widths[2:3] <- maxWidth
+
+# (2, 3), (3, 3)
+pE <- ggplot_gtable(ggplot_build(p5))
 pH   <- ggplot_gtable(ggplot_build(p8))
-maxWidth = unit.pmax(pE$widths[2:3], pF$widths[2:3], pG$widths[2:3], pH$widths[2:3])
-
-# Set the widths
+maxWidth = unit.pmax(pE$widths[2:3], pH$widths[c(2:3)])
 pE$widths[2:3] <- maxWidth
-pF$widths[2:3] <- maxWidth
-pG$widths[2:3] <- maxWidth
 pH$widths[2:3] <- maxWidth
+
+ylab1 <- expression(paste("ln-chlorophyll (", italic(mu), "g ", L^-1, ")"))
+ylab2 <- expression(paste("ln-silicon dioxide (mg ", L^-1, ")"))
+ylab3 <- expression(paste("ln-DIN (mg ", L^-1, ")"))
 
@@
 grid.arrange(
-  arrangeGrob(pA, pB, pC, pD, ncol = 1, heights = c(1, 1, 1, 1)), 
-  arrangeGrob(pE, pF, pG, pH, ncol = 1, heights = c(1, 1, 1, 1)), 
-  ncol = 2, widths = c(1, 1)
+  arrangeGrob(pA, pB, ncol = 2), 
+  arrangeGrob(
+    arrangeGrob(
+      textGrob(ylab1, rot = 90, gp = gpar(fontsize = 11)), 
+      arrangeGrob(pC, pF, ncol = 1), 
+      ncol = 2, widths = c(0.1, 1)
+      ), 
+    arrangeGrob(
+      textGrob(ylab2, rot = 90, gp = gpar(fontsize = 11)), 
+      arrangeGrob(pD, pG, ncol = 1), 
+      ncol = 2, widths = c(0.1, 1)
+      ),
+    arrangeGrob(
+      textGrob(ylab3, rot = 90, gp = gpar(fontsize = 11)), 
+      arrangeGrob(pE, pH, ncol = 1), 
+      ncol = 2, widths = c(0.1, 1)
+      ), 
+    ncol = 3
+    ),
+  arrangeGrob(pI, pJ, ncol = 2), 
+  ncol = 1, heights = c(1, 2, 1)
   )

### 2017-03-03 sftrends_manu.tex
DIN eval back to clam case study

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Feb 28 16:33:33 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/6213ed69c4113cd2ccdf0561273537009854ec0b}{6213ed69c4113cd2ccdf0561273537009854ec0b}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Mar 1 08:47:53 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/a3a14b892b843d68681697ef631e02a146fe6401}{a3a14b892b843d68681697ef631e02a146fe6401}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}
@@
 The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
-A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}d).
-An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}f).
-These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer (also shown in \cref{fig:clmchl}d), whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
-This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}
+A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).
+An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).
+These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (\cref{fig:clmchl}e, h) such that an increase in nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
+Overall, these results suggest the increase in nutrient concentrations are from decreased uptake by phytoplankton in Suisun Bay.
 
-The effect of flow as a top-down or bottom-up control changed throughout the time series.
-The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}h), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
+However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
+The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
 In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
-Following clam invasion later in the time series, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
+Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
 The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.
-For example, clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a).
+Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a).
 In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.
@@
 Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
-A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (d) based on \ac{wrtds} results.
-Reductions in \ac{chla} concentration were coincident with an increase in SiO$_2$ concentration (e), with the greatest increase in August (f).
-A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (g).
-Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (h, observations from June).}\label{fig:clmchl}
+A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \ac{wrtds} results.
+Reductions in \ac{chla} concentration were coincident with an increase in SiO$_2$ and \ac{din} concentrations (d, e), with the greatest increases in August (g, h).
+A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (i).
+Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (j, observations from June).}\label{fig:clmchl}
 \end{figure}

### 2017-03-05 sftrends_manu.Rnw
minor edits to din decrip

@@
 These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
-This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (\cref{fig:clmchl}e, h) such that an increase in nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (compare Figures \ref{fig:clmchl}e and \ref{fig:clmchl}e with \ref{fig:clmchl}d and \ref{fig:clmchl}g) such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
 Overall, these results suggest the increase in nutrient concentrations are from decreased uptake by phytoplankton in Suisun Bay.

### 2017-03-05 sftrends_manu.tex
minor edits to din decrip

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Mar 1 08:47:53 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/a3a14b892b843d68681697ef631e02a146fe6401}{a3a14b892b843d68681697ef631e02a146fe6401}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Mar 3 17:07:59 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/e74df4c4cebcbc4da36743397c352b1be42875e9}{e74df4c4cebcbc4da36743397c352b1be42875e9}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}

### 2017-03-09 sftrends_manu.tex
to STICS

@@
 \fax{+1 (850)9342401}
-\author{David Senn}
-\author{Thomas Jabusch}
+\author{David B. Senn}
+\author{Thomas W. Jabusch}
 \author{Phil Trowbridge}
@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Mar 3 17:07:59 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/e74df4c4cebcbc4da36743397c352b1be42875e9}{e74df4c4cebcbc4da36743397c352b1be42875e9}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Sun Mar 5 13:06:41 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/bf69bdbb5eef9d3a08d415c53dd23ab29126beda}{bf69bdbb5eef9d3a08d415c53dd23ab29126beda}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}

### 2017-03-14 sftrends_manu.Rnw
to STICS
minor yongshan edits

@@
 The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.
-The watershed provides drinking water to over 25 million people and includes irrigation for 18 thousand km$^2$ of agricultural land.
+The watershed provides water to over 25 million people and irrigation for 18 thousand km$^2$ of agricultural land.
 Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
@@
 
-Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes., particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
 Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).

### 2017-03-14 sftrends_manu.tex
minor yongshan edits

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Sun Mar 5 13:06:41 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/bf69bdbb5eef9d3a08d415c53dd23ab29126beda}{bf69bdbb5eef9d3a08d415c53dd23ab29126beda}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Mar 9 13:51:29 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/3a2258283f79a8ff3710c96c42f090e653e0a955}{3a2258283f79a8ff3710c96c42f090e653e0a955}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}

### 2017-03-15 sftrends_manu.tex
recompile draft

@@
 % get online bib file
-\begin{kframe}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in curl::curl\_fetch\_memory(url, handle = handle): Timeout was reached}}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in rawToChar(refs\$content): object 'refs' not found}}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in writeLines(refs, con = file("{}refs.bib"{})): object 'refs' not found}}\end{kframe}
 
@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Mar 9 13:51:29 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/3a2258283f79a8ff3710c96c42f090e653e0a955}{3a2258283f79a8ff3710c96c42f090e653e0a955}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Mar 14 09:11:12 2017 -0500, \href{https://github.com/fawda123/sftrends_manu/commit/7b5a17c006b01d83e6f284649e548d329f7ada14}{7b5a17c006b01d83e6f284649e548d329f7ada14}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}

### 2017-05-02 sftrends_manu.Rnw
fixed phil's name

@@
 \author{Thomas W. Jabusch}
-\author{Phil Trowbridge}
+\author{Philip R. Trowbridge}
 \affiliation{San Francisco Estuary Institute, Richmond, CA}
@@
 
-Relationships of nirite/nitrate with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).
+Relationships of nitrite/nitrate with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).
 Seasonal variation was even more apparent for ammonium, although both nitrite/nitrate and ammonium typically had the highest concentrations at low flow in the winter (January).

### 2017-05-02 sftrends_manu.tex
fixed phil's name

@@
 \author{Thomas W. Jabusch}
-\author{Phil Trowbridge}
+\author{Philip R. Trowbridge}
 \affiliation{San Francisco Estuary Institute, Richmond, CA}
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue Mar 14 09:11:12 2017 -0500, \href{https://github.com/fawda123/sftrends_manu/commit/7b5a17c006b01d83e6f284649e548d329f7ada14}{7b5a17c006b01d83e6f284649e548d329f7ada14}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Mar 15 17:49:38 2017 -0500, \href{https://github.com/fawda123/sftrends_manu/commit/835c0b4e439bcdbb0228b9b56799bfb4764b7d97}{835c0b4e439bcdbb0228b9b56799bfb4764b7d97}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}

### 2017-05-02 sftrends_manu.tex
david to last

@@
 \fax{+1 (850)9342401}
-\author{David B. Senn}
 \author{Thomas W. Jabusch}
 \author{Philip R. Trowbridge}
+\author{David B. Senn}
 \affiliation{San Francisco Estuary Institute, Richmond, CA}
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Wed Mar 15 17:49:38 2017 -0500, \href{https://github.com/fawda123/sftrends_manu/commit/835c0b4e439bcdbb0228b9b56799bfb4764b7d97}{835c0b4e439bcdbb0228b9b56799bfb4764b7d97}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue May 2 08:37:44 2017 -0500, \href{https://github.com/fawda123/sftrends_manu/commit/2a1fcbeba412e8ea16c8fea91f775c46617d667d}{2a1fcbeba412e8ea16c8fea91f775c46617d667d}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.}}

### 2017-05-02 sftrends_manu.Rnw
david to last
cover letter added, manuscript and supp info separated

@@
 % turn this off for submission
-\SectionNumbersOn
+% \SectionNumbersOn
 
@@
 \fax{+1 (850)9342401}
-\author{David B. Senn}
 \author{Thomas W. Jabusch}
 \author{Philip R. Trowbridge}
+\author{David B. Senn}
 \affiliation{San Francisco Estuary Institute, Richmond, CA}
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
-Environmental Protection Agency.}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
 
@@
 P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section.
-Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
+Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).
 Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.
@@
 As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
+For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in Table S1) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
@@
 \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results.
-Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
+Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (Figures S1 and S2 and Tables S2 and S3).
 The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
@@
 
-Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).
+Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (Figure S3).
 Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite.
@@
 Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models).\cite{Beck17}  Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
+
+\section*{Acknowledgments}
+
+We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program.
+We thank Larry Harding for providing comments on an earlier draft.
+This study was funded, reviewed, and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory.
+The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.
 
@@
 Trends are shown separately for different annual groupings.
-See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
+See Figures S1 and S2 for seasonal groupings.}
 \label{fig:trndcomp1}   
@@
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\\cref{fig:stock}).'>>=
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (Figure S3).'>>=
 data(h2dat)
@@
 cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, \\cref{fig:stock}).
+Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, Figure S3).
 Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
@@
 
-% supplementary material
-\begin{suppinfo}
-\beginsupplement
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1976-1995.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp2}   
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1996-2013.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp3}   
-\end{figure}
-
-<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
-data(stock_conc)
-
-stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))
-
-ptheme <- theme_minimal() + 
-  theme(
-    axis.line.x = element_line(), 
-    axis.line.y = element_line(), 
-    axis.ticks.x = element_line(),
-    axis.ticks.y = element_line(),
-    axis.ticks.length = unit(.1, "cm"),
-    axis.title.x = element_blank(),
-    legend.position = 'top', 
-    legend.title = element_blank()
-  )
-cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(3, 5, 8)] %>% 
-  rev
-
-toplo1 <- stock_conc %>% 
-  select(-val2) %>% 
-  spread(var, val) %>% 
-  mutate(TN = NH3 + NO2 + NO3)
-toplo2 <- stock_conc
-
-ylab1 <- expression(paste("Total nitrogen (mg ", L^-1, ")"))
-xrng <- range(c(toplo1$date, toplo2$date))
-yrng <- c(0, 50)
-leglabs <- list(
-  expression(NH [4] ^ '+'),
-  expression(NO [2] ^ '-'),
-  expression(NO [3] ^ '2-')
-  )
-
-lndate <- as.Date('2006/01/01')
-
-p1 <- ggplot(toplo2) + 
-  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
-  scale_x_date(expand = c(0, 0), limits = xrng) + 
-  scale_fill_manual(values = cols, labels = leglabs) + 
-  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
-  geom_label_repel(
-    data = data.frame(x = lndate, y = 35), 
-    aes(x, y, label = 'WWTP upgrades'), 
-    box.padding = grid::unit(2, "lines"), 
-    nudge_x = 1
-    ) +
-  scale_y_continuous(ylab1, expand = c(0, 0)) + 
-  ptheme
-
-p1
-@
-
-\clearpage
-
-% overall trends, years
-<<trndsann>>=
-data(trnds_seasyr)
-
-# percent changes
-tab <- trnds_nrm %>% 
-  filter(cat %in% c('1976-1995', '1996-2013')) %>% 
-  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      paste0('(', perchg, ')' ), 
-      paste0('(\\textit{\\textbf{', perchg, '}})')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  mutate(
-    perchg = paste0('\\footnotesize{', perchg, '}')
-  ) %>% 
-  unite('val', med, perchg, sep = ' ') %>% 
-  spread(cat, val) %>% 
-  arrange(resvar, Site_Code) %>% 
-  select(Site_Code, resvar, `1976-1995`, `1996-2013`)
-
-# table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
-stats <- tab$Site_Code
-
-# table
-latex( 
-  tab[, -c(1, 2)],
-  file = '',
-  rowlabel = 'Analyte/Station',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = vars,
-  n.rgroup = rep(nrow(tab)/3, 3),
-  cgroup = c('Annual'),
-  n.cgroup = c(2),
-  rowname = stats,
-  label = 'tab:trndsann',
-  insert.bottom = foot.val
-  )
-
-@
-
-% overall trends, seasons, first annual periodyears
-<<trndsbef>>=
-data(trnds_seasyr)
-
-# percent changes
-tab <- trnds_nrm %>% 
-  filter(
-    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
-    ann %in% 'bef'
-    ) %>% 
-  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      paste0('(', perchg, ')' ), 
-      paste0('(\\textit{\\textbf{', perchg, '}})')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  mutate(
-    perchg = paste0('\\footnotesize{', perchg, '}')
-  ) %>% 
-  unite('val', med, perchg, sep = ' ') %>% 
-  spread(cat, val) %>% 
-  arrange(resvar, Site_Code)
-
-# table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
-stats <- tab$Site_Code
-
-# table
-latex( 
-  tab[, -c(1, 2)],
-  file = '',
-  rowlabel = 'Analyte/Station',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = vars,
-  n.rgroup = rep(nrow(tab)/3, 3),
-  cgroup = c('Seasonal, 1976-1995'),
-  n.cgroup = c(4),
-  rowname = stats,
-  label = 'tab:trndsbef',
-  insert.bottom = foot.val
-  )
-
-@
-
-% overall trends, seasons, first annual periodyears
-<<trndsaft>>=
-data(trnds_seasyr)
-
-# percent changes
-tab <- trnds_nrm %>% 
-  filter(
-    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
-    ann %in% 'aft'
-    ) %>% 
-  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      paste0('(', perchg, ')' ), 
-      paste0('(\\textit{\\textbf{', perchg, '}})')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  mutate(
-    perchg = paste0('\\footnotesize{', perchg, '}')
-  ) %>% 
-  unite('val', med, perchg, sep = ' ') %>% 
-  spread(cat, val) %>% 
-  arrange(resvar, Site_Code)
-
-# table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
-vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
-stats <- tab$Site_Code
-
-# table
-latex( 
-  tab[, -c(1, 2)],
-  file = '',
-  rowlabel = 'Analyte/Station',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = vars,
-  n.rgroup = rep(nrow(tab)/3, 3),
-  cgroup = c('Seasonal, 1996-2013'),
-  n.cgroup = c(4),
-  rowname = stats,
-  label = 'tab:trndsaft',
-  insert.bottom = foot.val
-  )
-
-@
-
-\end{suppinfo}
-
 \end{document}

### 2017-05-02 sftrends_manu.tex
cover letter added, manuscript and supp info separated

@@
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue May 2 08:37:44 2017 -0500, \href{https://github.com/fawda123/sftrends_manu/commit/2a1fcbeba412e8ea16c8fea91f775c46617d667d}{2a1fcbeba412e8ea16c8fea91f775c46617d667d}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S.
-Environmental Protection Agency.}}
+\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section.
-Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
+Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).
 Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.
@@
 As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
+For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in Table S1) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
@@
 Eleven comparisons showed a trend reversal from positive to negative estimated change, nine sites went from no change to negative estimated change, and one site went from no change to a positive trend for the flow-normalized results.
-Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
+Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (Figures S1 and S2 and Tables S2 and S3).
 The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
@@
 
+\section*{Acknowledgments}
+
+We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program.
+We thank Larry Harding for providing comments on an earlier draft.
+This study was funded, reviewed, and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory.
+The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.
+
 % \section{References} % omit on submit
@@
 
-\caption{Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}.
+\caption[Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}]{Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}.
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\cref{fig:stock}).}\label{fig:p8trnds}
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (Figure S3).}\label{fig:p8trnds}
 \end{figure}
@@
 \caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
+Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, Figure S3).
 Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
@@
 \end{tabular}\end{center}
-
 \footnotesize *$p<0.05$; **$p<0.005$\end{table}
@@
 
-% supplementary material
-\begin{suppinfo}
-\beginsupplement
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1976-1995.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp2}   
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1996-2013.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp3}   
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\textwidth]{figs/stock-1}
-
-}
-
-\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
-\end{figure}
-
-\clearpage
-
-% overall trends, years
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.\label{tab:trndsann}}
-\begin{center}
-\begin{tabular}{lll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{2}{c}{\bfseries Annual}\tabularnewline
-\cline{2-3}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1995}&\multicolumn{1}{c}{1996-2013}\tabularnewline
-\hline
-{\bfseries DIN}&&\tabularnewline
-~~C10&1.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.4 \footnotesize{(-3.1)**}\tabularnewline
-~~C3&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.5 \footnotesize{(-0.1)**}\tabularnewline
-~~D19&0.4 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-1.2)**}\tabularnewline
-~~D28&0.4 \footnotesize{(\textit{\textbf{0.1}})**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
-~~D6&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.5 \footnotesize{(-0.3)**}\tabularnewline
-~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}\tabularnewline
-~~MD10&0.4 \footnotesize{(-1.1)**}&0.3 \footnotesize{(-2.4)**}\tabularnewline
-~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}\tabularnewline
-\hline
-{\bfseries NH$_{4}^{+}$}&&\tabularnewline
-~~C10&0.1 \footnotesize{(-3.4)**}&0 \footnotesize{(-5.2)**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{3.7}})**}&0.3 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
-~~D19&0 \footnotesize{(\textit{\textbf{0.4}})**}&0 \footnotesize{(-1.7)**}\tabularnewline
-~~D26&0.1 \footnotesize{(\textit{\textbf{2.2}})**}&0.1 \footnotesize{(-1.5)**}\tabularnewline
-~~D28&0 \footnotesize{(-1.1)**}&0 \footnotesize{(-1.4)**}\tabularnewline
-~~D4&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
-~~D6&0.1 \footnotesize{(\textit{\textbf{2.4}})**}&0.1 \footnotesize{(-0.5)**}\tabularnewline
-~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}\tabularnewline
-~~MD10&0.1 \footnotesize{(-2.8)**}&0 \footnotesize{(-1.1)**}\tabularnewline
-~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}\tabularnewline
-\hline
-{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&\tabularnewline
-~~C10&1.2 \footnotesize{(\textit{\textbf{1.4}})**}&1.4 \footnotesize{(-3)**}\tabularnewline
-~~C3&0.1 \footnotesize{(-0.1)**}&0.2 \footnotesize{(\textit{\textbf{0.7}})**}\tabularnewline
-~~D19&0.4 \footnotesize{(-0.1)**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
-~~D26&0.3 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1.1)**}\tabularnewline
-~~D28&0.4 \footnotesize{(-0.2)**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(-0.4)**}\tabularnewline
-~~D6&0.3 \footnotesize{(\textit{\textbf{1.3}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
-~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}\tabularnewline
-~~MD10&0.4 \footnotesize{(-1)**}&0.3 \footnotesize{(-2.5)**}\tabularnewline
-~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
-
-% overall trends, seasons, first annual periodyears
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsbef}}
-\begin{center}
-\begin{tabular}{lllll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1976-1995}\tabularnewline
-\cline{2-5}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
-\hline
-{\bfseries DIN}&&&&\tabularnewline
-~~C10&1.2 \footnotesize{(\textit{\textbf{1.1}})**}&1.2 \footnotesize{(\textit{\textbf{0.3}})}&1.3 \footnotesize{(\textit{\textbf{0.5}})**}&1.7 \footnotesize{(\textit{\textbf{1.2}})**}\tabularnewline
-~~C3&0.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.3 \footnotesize{(\textit{\textbf{2.3}})**}&0.4 \footnotesize{(\textit{\textbf{2.4}})**}&0.4 \footnotesize{(\textit{\textbf{1.9}})**}\tabularnewline
-~~D19&0.5 \footnotesize{(\textit{\textbf{0.3}})}&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.7 \footnotesize{(-0.2)}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(\textit{\textbf{0.4}})*}&0.4 \footnotesize{(\textit{\textbf{1}})**}&0.6 \footnotesize{(\textit{\textbf{0.3}})}\tabularnewline
-~~D28&0.5 \footnotesize{(\textit{\textbf{0.8}})*}&0.2 \footnotesize{(\textit{\textbf{0.3}})}&0.3 \footnotesize{(\textit{\textbf{0.5}})*}&0.8 \footnotesize{(-0.3)}\tabularnewline
-~~D4&0.4 \footnotesize{(\textit{\textbf{0.2}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.5 \footnotesize{(-0.5)}\tabularnewline
-~~D6&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{4.6}})**}&0.4 \footnotesize{(\textit{\textbf{1.4}})**}&0.5 \footnotesize{(-0.7)*}\tabularnewline
-~~D7&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.6 \footnotesize{(-2.4)**}\tabularnewline
-~~MD10&0.6 \footnotesize{(-0.3)}&0.2 \footnotesize{(-3.6)**}&0.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.3 \footnotesize{(-0.3)*}\tabularnewline
-~~P8&1.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.9 \footnotesize{(\textit{\textbf{2.4}})**}&1.3 \footnotesize{(\textit{\textbf{3.1}})**}&1.9 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
-\hline
-{\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
-~~C10&0.1 \footnotesize{(-2.3)**}&0 \footnotesize{(-6.8)**}&0.1 \footnotesize{(-7.1)**}&0.3 \footnotesize{(-1.5)**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.2 \footnotesize{(\textit{\textbf{4}})**}&0.3 \footnotesize{(\textit{\textbf{3.8}})**}&0.2 \footnotesize{(\textit{\textbf{2.9}})**}\tabularnewline
-~~D19&0.1 \footnotesize{(\textit{\textbf{0.4}})*}&0 \footnotesize{(-1.7)**}&0 \footnotesize{(\textit{\textbf{1.2}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}\tabularnewline
-~~D26&0.1 \footnotesize{(\textit{\textbf{1.4}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.1 \footnotesize{(\textit{\textbf{2.3}})**}\tabularnewline
-~~D28&0.1 \footnotesize{(-0.5)}&0 \footnotesize{(-3.7)**}&0 \footnotesize{(-1.6)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
-~~D4&0.1 \footnotesize{(\textit{\textbf{1.7}})**}&0 \footnotesize{(\textit{\textbf{1}})**}&0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textit{\textbf{2}})**}\tabularnewline
-~~D6&0.1 \footnotesize{(\textit{\textbf{2.9}})**}&0.1 \footnotesize{(\textit{\textbf{2.8}})**}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
-~~D7&0.1 \footnotesize{(\textit{\textbf{3.3}})**}&0 \footnotesize{(\textit{\textbf{2}})**}&0.1 \footnotesize{(-2.8)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
-~~MD10&0.1 \footnotesize{(-1.8)**}&0 \footnotesize{(-6.5)**}&0 \footnotesize{(-3.3)**}&0.2 \footnotesize{(\textit{\textbf{0.4}})}\tabularnewline
-~~P8&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.1 \footnotesize{(\textit{\textbf{1.8}})**}&0.2 \footnotesize{(\textit{\textbf{7}})**}&0.6 \footnotesize{(\textit{\textbf{7}})**}\tabularnewline
-\hline
-{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
-~~C10&1.1 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(\textit{\textbf{0.6}})**}&1.2 \footnotesize{(\textit{\textbf{1.3}})**}&1.5 \footnotesize{(\textit{\textbf{1.8}})**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{0.7}})**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
-~~D19&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.2 \footnotesize{(-0.3)}&0.3 \footnotesize{(\textit{\textbf{0.3}})}&0.6 \footnotesize{(-0.9)*}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.6}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.3}})*}&0.5 \footnotesize{(-0.3)}\tabularnewline
-~~D28&0.5 \footnotesize{(\textit{\textbf{0.7}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.2}})}&0.7 \footnotesize{(-1)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.1}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.4 \footnotesize{(-0.8)*}\tabularnewline
-~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.1}})**}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.4 \footnotesize{(-1)**}\tabularnewline
-~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textit{\textbf{3.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.4 \footnotesize{(-3.6)**}\tabularnewline
-~~MD10&0.5 \footnotesize{(-0.2)}&0.2 \footnotesize{(-3.6)**}&0.2 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(-0.5)*}\tabularnewline
-~~P8&1.2 \footnotesize{(\textit{\textbf{2}})**}&0.9 \footnotesize{(\textit{\textbf{2.3}})**}&1.1 \footnotesize{(\textit{\textbf{2}})**}&1.4 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
-
-% overall trends, seasons, first annual periodyears
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsaft}}
-\begin{center}
-\begin{tabular}{lllll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1996-2013}\tabularnewline
-\cline{2-5}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
-\hline
-{\bfseries DIN}&&&&\tabularnewline
-~~C10&1.1 \footnotesize{(-4.1)**}&1.3 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-2)**}&1.7 \footnotesize{(-3.4)**}\tabularnewline
-~~C3&0.5 \footnotesize{(\textit{\textbf{0.5}})}&0.4 \footnotesize{(\textit{\textbf{0.1}})}&0.6 \footnotesize{(-0.2)}&0.5 \footnotesize{(-0.6)**}\tabularnewline
-~~D19&0.5 \footnotesize{(-2.8)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.6)**}&0.7 \footnotesize{(-2.3)**}\tabularnewline
-~~D26&0.5 \footnotesize{(-1.9)**}&0.3 \footnotesize{(-1.7)**}&0.4 \footnotesize{(-1)**}&0.6 \footnotesize{(-0.8)**}\tabularnewline
-~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-4.9)**}&0.2 \footnotesize{(-4.9)**}&0.7 \footnotesize{(-2.1)**}\tabularnewline
-~~D4&0.4 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.9)**}&0.5 \footnotesize{(\textit{\textbf{0.6}})**}\tabularnewline
-~~D6&0.5 \footnotesize{(-0.2)*}&0.5 \footnotesize{(-1)**}&0.5 \footnotesize{(-0.3)*}&0.5 \footnotesize{(-0.1)}\tabularnewline
-~~D7&0.5 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-1.3)**}&0.4 \footnotesize{(-0.4)**}&0.6 \footnotesize{(-0.2)}\tabularnewline
-~~MD10&0.4 \footnotesize{(-2.3)**}&0.2 \footnotesize{(-3.7)**}&0.2 \footnotesize{(-4.4)**}&1 \footnotesize{(-1.8)**}\tabularnewline
-~~P8&1.5 \footnotesize{(-1.9)**}&1.2 \footnotesize{(-3.5)**}&1.8 \footnotesize{(-2.4)**}&2.7 \footnotesize{(-2.2)**}\tabularnewline
-\hline
-{\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
-~~C10&0 \footnotesize{(-4.2)**}&0 \footnotesize{(-6.1)**}&0 \footnotesize{(-8.5)**}&0.1 \footnotesize{(-7.3)**}\tabularnewline
-~~C3&0.3 \footnotesize{(\textit{\textbf{1}})**}&0.3 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-0.1)}\tabularnewline
-~~D19&0 \footnotesize{(-1.9)**}&0 \footnotesize{(-0.4)}&0 \footnotesize{(-2.2)**}&0.1 \footnotesize{(-1.8)**}\tabularnewline
-~~D26&0.1 \footnotesize{(-1.2)**}&0.1 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-1.9)**}&0.1 \footnotesize{(-1.4)**}\tabularnewline
-~~D28&0 \footnotesize{(-1.7)**}&0 \footnotesize{(-0.2)}&0 \footnotesize{(-2.4)**}&0.1 \footnotesize{(-3.1)**}\tabularnewline
-~~D4&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-0.3)}&0.1 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
-~~D6&0.1 \footnotesize{(-0.7)**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0.1 \footnotesize{(-0.3)**}\tabularnewline
-~~D7&0.1 \footnotesize{(-2.2)**}&0 \footnotesize{(-2.1)**}&0.1 \footnotesize{(\textit{\textbf{1.3}})**}&0.1 \footnotesize{(-0.4)*}\tabularnewline
-~~MD10&0 \footnotesize{(-1.4)*}&0 \footnotesize{(-0.1)}&0 \footnotesize{(-0.8)**}&0.1 \footnotesize{(-4.3)**}\tabularnewline
-~~P8&0.2 \footnotesize{(-8.7)**}&0.1 \footnotesize{(-6.3)**}&0.2 \footnotesize{(-10.4)**}&0.5 \footnotesize{(-13.1)**}\tabularnewline
-\hline
-{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
-~~C10&1.1 \footnotesize{(-4.2)**}&1.2 \footnotesize{(-3.2)**}&1.6 \footnotesize{(-1.9)**}&1.6 \footnotesize{(-3.3)**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.2 \footnotesize{(\textit{\textbf{1.7}})**}&0.2 \footnotesize{(-0.4)}\tabularnewline
-~~D19&0.4 \footnotesize{(-2.9)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.5)**}&0.6 \footnotesize{(-2.2)**}\tabularnewline
-~~D26&0.4 \footnotesize{(-1.9)**}&0.2 \footnotesize{(-1.6)**}&0.3 \footnotesize{(-0.6)*}&0.5 \footnotesize{(-0.6)**}\tabularnewline
-~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-5.4)**}&0.2 \footnotesize{(-5.2)**}&0.7 \footnotesize{(-1.7)**}\tabularnewline
-~~D4&0.3 \footnotesize{(-0.1)}&0.3 \footnotesize{(-1)**}&0.3 \footnotesize{(-1)**}&0.4 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
-~~D6&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.4)*}&0.4 \footnotesize{(-0.1)}\tabularnewline
-~~D7&0.4 \footnotesize{(-0.6)**}&0.4 \footnotesize{(-1.2)**}&0.4 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-0.3)*}\tabularnewline
-~~MD10&0.4 \footnotesize{(-2.6)**}&0.1 \footnotesize{(-4.5)**}&0.2 \footnotesize{(-5.4)**}&1 \footnotesize{(-1.4)**}\tabularnewline
-~~P8&1.3 \footnotesize{(-1.1)**}&1.1 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
-
-\end{suppinfo}
-
 \end{document}

### 2017-05-02 supporting.Rnw
cover letter added, manuscript and supp info separated

% \documentclass[journal = esthag, manuscript = article]{achemso}

\documentclass[letterpaper,12pt,oneside]{article}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[colorlinks  =true, allcolors=Blue]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{lineno}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage{paralist}
\usepackage{setspace}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyfoot[c]{S\thepage}
% \usepackage{showlabels}

% cleveref options
\crefname{table}{Table}{Tables}
\crefname{figure}{Figure}{Figures}
\renewcommand{\figurename}{Figure}

\renewcommand{\headrulewidth}{0pt}

%acronyms
\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
\acrodef{din}[DIN]{dissolved inorganic nitrogen}
\acrodef{emp}[EMP]{Environmental Monitoring Program}
\acrodef{iep}[IEP]{Interagency Ecological Program}
\acrodef{jas}[JAS]{July-August-September}
\acrodef{sfe}[SFE]{San Francisco Estuary}
\acrodef{wrtds}[WRTDS]{Weighted Regressions on Time, Discharge, and Season}
\acrodef{wwtp}[WWTP]{wastewater treatment plant}
\acrodefplural{wwtp}[WWTPs]{wastewater treatment plants}

%for supplemental figures/tables
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

% for unit macros
\newcommand{\mgl}{mg L$^{-1}$}
\newcommand{\mugl}{$\mu$g L$^{-1}$}

%knitr options
<<setup,include=F,cache=F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T)
options(replace.assign = TRUE, width = 90)

source('R/funcs.r')
library(WRTDStidal)
library(tidyverse)
library(gridExtra)
library(grid)
library(Hmisc)
library(ggrepel)
library(ggtree)
library(NADA)
library(english)
library(forcats)
@

\begin{document}

% supplementary material
\section*{Supplementary Information}

\noindent{\bf Four decades of water quality change in the upper San Francisco Estuary}
\vspace{0.1in}

\noindent Marcus W. Beck (beck.marcus@epa.gov), Thomas W. Jabusch,Philip R. Trowbridge, David B. Senn
\vspace{0.1in}

\beginsupplement

\noindent The following files are available free of charge: \cref{fig:trndcomp2,fig:trndcomp3,fig:stock}, \cref{tab:trndsann,tab:trndsbef,tab:trndsaft}

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different seasonal groupings from 1976-1995. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See Figure 3 for annual comparisons.}
\label{fig:trndcomp2}   
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different seasonal groupings from 1996-2013. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See Figure 3 for annual comparisons.}
\label{fig:trndcomp3}   
\end{figure}

<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.  Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
data(stock_conc)

stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))

ptheme <- theme_minimal() + 
  theme(
    axis.line.x = element_line(), 
    axis.line.y = element_line(), 
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line(),
    axis.ticks.length = unit(.1, "cm"),
    axis.title.x = element_blank(),
    legend.position = 'top', 
    legend.title = element_blank()
  )
cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(3, 5, 8)] %>% 
  rev

toplo1 <- stock_conc %>% 
  select(-val2) %>% 
  spread(var, val) %>% 
  mutate(TN = NH3 + NO2 + NO3)
toplo2 <- stock_conc

ylab1 <- expression(paste("Total nitrogen (mg ", L^-1, ")"))
xrng <- range(c(toplo1$date, toplo2$date))
yrng <- c(0, 50)
leglabs <- list(
  expression(NH [4] ^ '+'),
  expression(NO [2] ^ '-'),
  expression(NO [3] ^ '2-')
  )

lndate <- as.Date('2006/01/01')

p1 <- ggplot(toplo2) + 
  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
  scale_x_date(expand = c(0, 0), limits = xrng) + 
  scale_fill_manual(values = cols, labels = leglabs) + 
  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
  geom_label_repel(
    data = data.frame(x = lndate, y = 35), 
    aes(x, y, label = 'WWTP upgrades'), 
    box.padding = grid::unit(2, "lines"), 
    nudge_x = 1
    ) +
  scale_y_continuous(ylab1, expand = c(0, 0)) + 
  ptheme

p1
@

\clearpage

% overall trends, years
<<trndsann>>=
data(trnds_seasyr)

# percent changes
tab <- trnds_nrm %>% 
  filter(cat %in% c('1976-1995', '1996-2013')) %>% 
  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
  mutate(
    med = round(med, 1), 
    perchg = round(perchg, 1),
    perchg = ifelse(
      perchg < 0, 
      paste0('(', perchg, ')' ), 
      paste0('(\\textit{\\textbf{', perchg, '}})')
      ), 
    ztest = p_ast(ztest)
  ) %>% 
  unite('perchg', perchg, ztest, sep = '') %>% 
  mutate(
    perchg = paste0('\\footnotesize{', perchg, '}')
  ) %>% 
  unite('val', med, perchg, sep = ' ') %>% 
  spread(cat, val) %>% 
  arrange(resvar, Site_Code) %>% 
  select(Site_Code, resvar, `1976-1995`, `1996-2013`)

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.  Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.'
foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

# table
latex( 
  tab[, -c(1, 2)],
  file = '',
  rowlabel = 'Analyte/Station',
  caption = cap.val,
  caption.loc = 'top',
  rgroup = vars,
  n.rgroup = rep(nrow(tab)/3, 3),
  cgroup = c('Annual'),
  n.cgroup = c(2),
  rowname = stats,
  label = 'tab:trndsann',
  insert.bottom = foot.val
  )

@

% overall trends, seasons, first annual periodyears
<<trndsbef>>=
data(trnds_seasyr)

# percent changes
tab <- trnds_nrm %>% 
  filter(
    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
    ann %in% 'bef'
    ) %>% 
  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
  mutate(
    med = round(med, 1), 
    perchg = round(perchg, 1),
    perchg = ifelse(
      perchg < 0, 
      paste0('(', perchg, ')' ), 
      paste0('(\\textit{\\textbf{', perchg, '}})')
      ), 
    ztest = p_ast(ztest)
  ) %>% 
  unite('perchg', perchg, ztest, sep = '') %>% 
  mutate(
    perchg = paste0('\\footnotesize{', perchg, '}')
  ) %>% 
  unite('val', med, perchg, sep = ' ') %>% 
  spread(cat, val) %>% 
  arrange(resvar, Site_Code)

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995. Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

# table
latex( 
  tab[, -c(1, 2)],
  file = '',
  rowlabel = 'Analyte/Station',
  caption = cap.val,
  caption.loc = 'top',
  rgroup = vars,
  n.rgroup = rep(nrow(tab)/3, 3),
  cgroup = c('Seasonal, 1976-1995'),
  n.cgroup = c(4),
  rowname = stats,
  label = 'tab:trndsbef',
  insert.bottom = foot.val
  )

@

% overall trends, seasons, first annual periodyears
<<trndsaft>>=
data(trnds_seasyr)

# percent changes
tab <- trnds_nrm %>% 
  filter(
    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
    ann %in% 'aft'
    ) %>% 
  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
  mutate(
    med = round(med, 1), 
    perchg = round(perchg, 1),
    perchg = ifelse(
      perchg < 0, 
      paste0('(', perchg, ')' ), 
      paste0('(\\textit{\\textbf{', perchg, '}})')
      ), 
    ztest = p_ast(ztest)
  ) %>% 
  unite('perchg', perchg, ztest, sep = '') %>% 
  mutate(
    perchg = paste0('\\footnotesize{', perchg, '}')
  ) %>% 
  unite('val', med, perchg, sep = ' ') %>% 
  spread(cat, val) %>% 
  arrange(resvar, Site_Code)

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013. Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

# table
latex( 
  tab[, -c(1, 2)],
  file = '',
  rowlabel = 'Analyte/Station',
  caption = cap.val,
  caption.loc = 'top',
  rgroup = vars,
  n.rgroup = rep(nrow(tab)/3, 3),
  cgroup = c('Seasonal, 1996-2013'),
  n.cgroup = c(4),
  rowname = stats,
  label = 'tab:trndsaft',
  insert.bottom = foot.val
  )

@

\end{document}

### 2017-05-02 sftrends_manu.Rnw
slight shortening

@@
 Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.
-Climate, local, and regional environmental effects may act individually or together over time to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.
 As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
@@
 Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}.
-These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.
+These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time.
 The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
-This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.
 More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
 These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.
-Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.
 
@@
 Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
-To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
+To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utitilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
 
@@
 The second objective evaluated two specific water quality stations as additional case studies to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
-Although quantitative descriptions of change can be ends in themselves, the results provide means to a more detailed understanding of ecosystem properties.
-Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.
 
@@
 Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.
-The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).
+The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay (\cref{fig:delt_map}).
 Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
@@
 The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \cite{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resistence to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.
-These recent changes have been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
-
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
-A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}.
-For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.
-Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
-Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows and water exports directly from the system \cite{Jassby00,Jassby08}.
-Our analysis explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
+However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\cite{Lehman05,Cloern07,Lehman10}, which have have been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
 
@@
 
-Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{sfe} region.
-The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.
-The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.
-Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.
-All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.
-Nutrient time series from 1976 to 2013 were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) for ten discrete sampling stations (\cref{fig:delt_map}).
+Nutrient time series from 1976 to 2013 were obtained for ten discrete sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.
 Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
@@
 
-Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.
-Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.
+Daily flow estimates for the Delta region were obtained from the Dayflow software program \cite{IEP16}.
 The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends.
@@
 A unique set of weights is used for each regression to control the importance of observations used to fit the model relative to the observation at the center of the window.
-The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow.
 The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.
-As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.
 The \ac{wrtds} models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
@@
 Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
-These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midpoints in the time series.
+These annual and seasonal groupings were chosen for continuity with similar comparisons in Ref.~\citenum{Jabusch16} and as approximate twenty year midpoints in the time series.
 
@@
 The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
-In general, long-term (inter-annual) trends were not easily observed from the raw data.
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
 Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
@@
 
-Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years.
-Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Application of seasonal Kendall tests to evaluate trends in observed data provided information on the direction, magnitude, and statistical significance of changes between years.
+Trends estimated for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).
@@
 \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing).
-P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section.
+P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites (see next section).
 Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).
@@
 Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
-Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
+Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
 A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.
@@
 
-Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
-The selected case studies focused on \begin{inparaenum}[1\upshape)]
-\item effects of wastewater treatment upgrades upstream of P8, and
-\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.
-\end{inparaenum}
-Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
-
 \subsubsection{Effects of wastewater treatment}
 
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc.)\cite{Cornwell14,Novick15}.
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources \cite{Cornwell14,Novick15}.
 Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 (\cref{fig:delt_map})\cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
 A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
@@
 
-Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (Figure S3).
+Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (Figure S3).
 Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite.
@@
 Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January.
-Using \ac{wrtds}, an empirical link is created between upstream changes and observed effects downstream that is characterized by differences in analytes between years and season.
 A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
@@
 For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
-Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were generally associated with higher concentrations.
+Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.
 This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow.
-As such, these simple observations from \ac{wrtds} provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
+As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
 
@@
 Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred with the most notable invasion effect being increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}.
-The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
-We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
+Changes in the physical environment have also occurred, such as increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}.
+The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
+We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
 \item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
@@
 \end{inparaenum}
-The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) provides an approach to assess the competing effects of climate variability, hydrology, and ecology on ambient conditions.
-
-Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were further affected by flow changes over time (\cref{fig:clmchl}).
+
 Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
@@
 
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
 However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
@@
 We thank Larry Harding for providing comments on an earlier draft.
-This study was funded, reviewed, and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory.
+This study was reviewed and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory.
+The authors declare no competing financial interest.
 The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.
@@
 
-cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, Figure S3).
+cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8'
+foot.val<-'\\footnotesize Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, Figure S3).
 Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
@@
 Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$'
 

### 2017-05-02 sftrends_manu.tex
slight shortening

@@
 The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
-In general, long-term (inter-annual) trends were not easily observed from the raw data.
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
@@
 
-Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years.
-Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Application of seasonal Kendall tests to evaluate trends in observed data provided information on the direction, magnitude, and statistical significance of changes between years.
+Trends estimated for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).
@@
 Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing).
-P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section.
+P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites (see next section).
 Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).
@@
 \end{inparaenum}
-The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) provides an approach to assess the competing effects of climate variability, hydrology, and ecology on ambient conditions.
 
-Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were further affected by flow changes over time (\cref{fig:clmchl}).
 Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
@@
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
-Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, Figure S3).
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:p8chg}}
+\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8\label{tab:p8chg}} 
 \begin{center}
@@
 \end{tabular}\end{center}
-\footnotesize *$p<0.05$; **$p<0.005$\end{table}
+\footnotesize Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, Figure S3).
+Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$\end{table}
 

### 2017-05-02 supporting.Rnw
slight shortening

@@
 % supplementary material
-\section*{Supplementary Information}
+\section*{Supporting Information}
 
@@
 
-\noindent The following files are available free of charge: \cref{fig:trndcomp2,fig:trndcomp3,fig:stock}, \cref{tab:trndsann,tab:trndsbef,tab:trndsaft}
+\noindent The following files are available free of charge: \cref{fig:trndcomp2,fig:trndcomp3,fig:stock}, \cref{tab:trndsann,tab:trndsbef,tab:trndsaft} (7 pages)
 
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations'
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+*$p<0.05$; **$p<0.005$'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995'
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
-Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013'
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
-foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')

### 2017-05-03 sftrends_manu.Rnw
final senn edits added

@@
 Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred, such as increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}.
+Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply,\cite{Jassby08,Schoellhamer11,Cloern12b} although decreases in phytoplankton by clam biofiltration may have also increased clarity \cite{MacNally10}.
 The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
@@
 \end{inparaenum}
-
-Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
+Although the relationship between phytoplankton and clams have been well described in \ac{sfe}\cite{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.
+
+Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P.
 amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
-The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
+The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.
 A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).
@@
 This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (compare Figures \ref{fig:clmchl}e and \ref{fig:clmchl}e with \ref{fig:clmchl}d and \ref{fig:clmchl}g) such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
-Overall, these results suggest the increase in nutrient concentrations are from decreased uptake by phytoplankton in Suisun Bay.
+Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
+amurensis}.
 
@@
 We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.
+Moreover, trends in nutrient loads from point soucres in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases\cite{Jassby08} and exports to Suisun Bay\cite{Novick14}.
+The results from \ac{wrtds} demonstrating these changes are not unexpected, and consequently, we are not detracting from the potential implications of such increases.
+The important conclusion is that the physical/hydrological and biogeochemical factors that influence nutrient cycling and ambient concentrations in the Bay-Delta, and changes to those factors, are substantial enough that they can be comparable in magnitude to anthropogenic load increases or comparable to the effects of management actions to decrease nutrient levels.
+Therefore, methods that adjust for the effects of these factors are critical when studying long-term records to assess the impacts or effectiveness of load increases or management actions, respectively.
+
 Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
@@
 <<clmchl, fig.height = 9.3, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
-Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
+Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and reduction of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
 A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \\ac{wrtds} results.

### 2017-05-03 sftrends_manu.tex
final senn edits added

@@
 \end{inparaenum}
+Although the relationship between phytoplankton and clams have been well described in \ac{sfe}\cite{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.
 
-Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P.
+Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P.
 amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
-The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).
+The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.
 A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).
@@
 \caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
-Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
+Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and reduction of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
 A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \ac{wrtds} results.

### 2017-05-03 sftrends_manu.Rnw
fixed biblio non-article refs

@@
 
-% get the version based on commit date
-<<echo = FALSE, cache = FALSE>>=
-raw <- system('git log -1', intern = TRUE)
-raw <- raw[grep('^Date', raw)]
-raw <- paste('Version', raw)
-@
-
-% get online bib file
-<<echo = FALSE, cache = FALSE>>=
-refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
-refs <- rawToChar(refs$content)
-writeLines(refs, con = file('refs.bib'))
-@
-
-% get the version from git log
-<<echo = FALSE, cache = FALSE>>=
-raw <- system('git log -1', intern = TRUE)
-date <- raw[grep('^Date', raw)]
-date <- gsub('^Date:\\s*', '', date)
-commit <- raw[grep('^commit', raw)]
-commit <- gsub('^commit ', '', commit)
-commit <- paste0('\\href{https://github.com/fawda123/sftrends_manu/commit/', commit, '}{', commit, '}')
-vers <- paste0('Version: ', date, ', ', commit)
-@
-
 \author{Marcus W. Beck}
@@
 \noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
-The seasonal trend is modeled as a sinusoidal component to capture periodicity between years.
 The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
-A unique set of weights is used for each regression to control the importance of observations used to fit the model relative to the observation at the center of the window.
-The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.
-The \ac{wrtds} models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
+Models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
 All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT16}.
@@
 A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
-Although variation in nutrients can be caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends.
-A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.

### 2017-05-03 sftrends_manu.tex
fixed biblio non-article refs

@@
 %knitr options
-
-% get the version based on commit date
-
-% get online bib file
-
-% get the version from git log
 

### 2017-05-03 R/toc_abs.R
toc abstract added

@@ -1,0 +1,70 @@
+data(h2dat)
+
+nomod <-  filter(h2dat, resvar == 'no23') %>% 
+  .$data %>% 
+  .[[1]]
+
+# globals
+ylab1 <- expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-', ' (mg ', L^-1, ')'))
+flolab <- expression(paste('ln-Flow (', m^3, ' ', s^-1, ')'))
+
+# default theme
+mytheme <- theme_minimal() + 
+  theme(
+    plot.background = element_rect(fill='transparent', 
+      colour = NA),
+    panel.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.position = 'none',
+    legend.background = element_rect(fill='transparent', 
+      colour = NA),
+    legend.key = element_rect(fill = 'transparent', 
+      colour = NA),
+    axis.text.x = element_text(size = 16), 
+    axis.title.x = element_blank(),
+    axis.text.y = element_text(size = 16),
+    axis.title.y = element_text(size = 18), 
+    axis.line.x = element_line(), 
+    axis.line.y = element_line(), 
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"), 
+    panel.grid.major = element_line(colour = "#7BCCC4"),
+    panel.grid.minor = element_blank(), 
+    plot.margin = unit(c(3, 3, 3, 3), 'pt')
+  )   
+col_vec <- RColorBrewer::brewer.pal(9, 'Blues')[c(9, 8, 7)]
+xlims <- range(nomod$date)
+
+# observed time series
+toplo1 <- select(nomod, date, res) %>% 
+  na.omit()
+p1 <- ggplot(toplo1, aes(x = date, y = exp(res))) + 
+  geom_line() + 
+  mytheme + 
+  scale_y_continuous(ylab1, limits = c(0, 4)) +
+  scale_x_date(limits = xlims) 
+
+# prdnrmplots
+p3 <- prdnrmplot(nomod, logspace = F, min_mo = 10) + 
+  mytheme + 
+  scale_y_continuous(ylab1) +
+  scale_x_date(limits = xlims) 
+
+# align widths of plots in first column, first two
+pA <- ggplot_gtable(ggplot_build(p1))
+pC <- ggplot_gtable(ggplot_build(p3))
+
[…]

### 2017-05-04 sftrends_manu.Rnw
toc abstract added
submitted to ES&T

@@
 Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
+
+\centerline{\includegraphics[width=0.9\textwidth]{figs/tocabs.pdf}}
 \end{abstract}
@@
 Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
-To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utitilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
+To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
 
@@
 The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \cite{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\cite{Lehman05,Cloern07,Lehman10}, which have have been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
+However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\cite{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
 
@@
 
-Nutrient time series from 1976 to 2013 were obtained for ten discrete sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.
+Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.
 Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
@@
 Models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
-All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT16}.
-
-A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
+All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT17}.
+
+A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows\cite{Hirsch10}.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
@@
 Trends in each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
-This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82}.
+This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82,Millard13}.
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
@@
 These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
-Functions in the EnvStats package in R were used for the seasonal Kendall tests \cite{Millard13}.
 
@@
 
-Application of seasonal Kendall tests to evaluate trends in observed data provided information on the direction, magnitude, and statistical significance of changes between years.
-Trends estimated for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Estimated trends from Seasonal Kendall tests on the observed data varied considerably between sites and analytes (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).
@@
 Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
-Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.
+Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
 
@@
 These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
-This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (compare Figures \ref{fig:clmchl}e and \ref{fig:clmchl}e with \ref{fig:clmchl}d and \ref{fig:clmchl}g) such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
 Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
@@
 
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
 However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
@@
 We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.
-Moreover, trends in nutrient loads from point soucres in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases\cite{Jassby08} and exports to Suisun Bay\cite{Novick14}.
+Moreover, trends in nutrient loads from point sources in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases\cite{Jassby08} and exports to Suisun Bay\cite{Novick14}.
 The results from \ac{wrtds} demonstrating these changes are not unexpected, and consequently, we are not detracting from the potential implications of such increases.

### 2017-05-04 sftrends_manu.tex
toc abstract added
submitted to ES&T

@@
 Models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
-All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT16}.
+All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT17}.
 
-A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.
+A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows\cite{Hirsch10}.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
@@
 
-Application of seasonal Kendall tests to evaluate trends in observed data provided information on the direction, magnitude, and statistical significance of changes between years.
-Trends estimated for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Estimated trends from Seasonal Kendall tests on the observed data varied considerably between sites and analytes (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).

### 2017-09-28 sftrends_manu.Rnw
revision for resubmit, removed EST .sty

@@
-\documentclass[journal = esthag, manuscript = article]{achemso}
-
-% \usepackage[version=3]{mhchem}
-\usepackage[T1]{fontenc}
-\usepackage[colorlinks  =true, allcolors=Blue]{hyperref}
+\documentclass[letterpaper,12pt,oneside]{article}
+\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
+\usepackage{setspace}
+\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
 \usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{indentfirst}
+\usepackage{titlesec}
+\usepackage{multirow}
+\usepackage{booktabs}
+\usepackage{graphicx}
+\usepackage{verbatim}
+\usepackage{rotating}
+\usepackage{tabularx}
+\usepackage{outlines}
 \usepackage{lineno}
+\usepackage{array}
+\usepackage{times}
 \usepackage{cleveref}
 \usepackage{acronym}
+\usepackage[position=t]{subfig}
 \usepackage{paralist}
-\usepackage{setspace}
+\usepackage[noae]{Sweave}
+\usepackage{natbib}
+\usepackage{array}
+\usepackage{pdflscape}
+\usepackage{bm}
 % \usepackage{showlabels}
+\bibpunct{(}{)}{,}{a}{}{,}
+
+% page margins and section title formatting
+\linespread{1.5}
+\setlength{\footskip}{0.5in}
+\titleformat*{\section}{\Large\bf\em}
+\titleformat*{\subsection}{\singlespace\large\bf}
+\titleformat*{\subsubsection}{\singlespace\normalsize\bf\em}
+\titlespacing{\section}{0in}{0in}{0in}
+\titlespacing{\subsection}{0in}{0in}{0in}
+\titlespacing{\subsubsection}{0in}{0in}{0in}
 
@@
 \crefname{table}{Table}{Tables}
-\crefname{figure}{Figure}{Figures}
-\renewcommand{\figurename}{Figure}
+\crefname{figure}{Fig.}{Figs.}
+\renewcommand{\figurename}{Fig.}
 
@@
 
-%for supplemental figures/tables
-\newcommand{\beginsupplement}{%
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{S\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{S\arabic{figure}}%
-     }
-
 % for unit macros
@@
 \newcommand{\mugl}{$\mu$g L$^{-1}$}
-
-% turn this off for submission
-% \SectionNumbersOn
 
@@
 # set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T)
-options(replace.assign = TRUE, width = 90)
+opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
+options(replace.assign = TRUE, width = 90, digits = 1)
 
@@
 
-\author{Marcus W. Beck}
-\affiliation{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, Gulf Breeze, FL}
-\email{beck.marcus@epa.gov}
-\phone{+1 (850)9342480}
-\fax{+1 (850)9342401}
-\author{Thomas W. Jabusch}
-\author{Philip R. Trowbridge}
-\author{David B. Senn}
-\affiliation{San Francisco Estuary Institute, Richmond, CA}
-
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
+% get the version based on commit date
+<<echo = FALSE, cache = FALSE>>=
+raw <- system('git log -1', intern = TRUE)
+raw <- raw[grep('^Date', raw)]
+raw <- paste('Version', raw)
+@
+
+% get online bib file
+<<echo = FALSE, cache = FALSE>>=
+refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
+refs <- rawToChar(refs$content)
+writeLines(refs, con = file('refs.bib'))
+@
 
 \begin{document}
+
+\raggedbottom
 \linenumbers
-
-\begin{abstract}
+\raggedright
+\urlstyle{same}
+\setlength{\parindent}{0.5in}
+\renewcommand\refname{References \vspace{12pt}}
+
+\begin{singlespace}
+\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
+\author{
+  {\bf {\normalsize Marcus W. Beck$^1$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^3$, David B. Senn$^4$}}
+  \\\\{\textit {\normalsize $^1$Southern California Coastal Water Research Project}}
+  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
+	\\{\textit {\normalsize Phone: 714-755-3217, Fax: 714-755-3299, Email: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
+  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
+	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
+	\\{\textit {\normalsize Phone: 510-746-7340, Fax: 510-746-7300, Email: \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
+	\\\\{\textit {\normalsize $^3$San Francisco Estuary Institute}}
+	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
+	\\{\textit {\normalsize Phone: 510-746-7345, Fax: 510-746-7300, Email: \href{mailto:philt@sfei.org}{philt@sfei.org}}}
+	\\\\{\textit {\normalsize $^4$San Francisco Estuary Institute}}
+	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
+	\\{\textit {\normalsize Phone: 510-746-7366, Fax: 510-746-7300, Email: \href{mailto:davids@sfei.org}{davids@sfei.org}}}
+  \vspace{1in}
+  \\ \Sexpr{raw}
+	}
+\date{}
+\maketitle
+\end{singlespace}
+
+\section*{Abstract}
 \noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.
@@
 Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
-
-\centerline{\includegraphics[width=0.9\textwidth]{figs/tocabs.pdf}}
-\end{abstract}
 \acresetall
@@
 \begin{singlespace}
+\bibliographystyle{apalike_mine}
 \bibliography{refs}

### 2017-09-28 sftrends_manu.tex
revision for resubmit, removed EST .sty

@@
-\documentclass[journal = esthag, manuscript = article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[letterpaper,12pt,oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \usepackage{alltt}
-
-% \usepackage[version=3]{mhchem}
-\usepackage[T1]{fontenc}
-\usepackage[colorlinks  =true, allcolors=Blue]{hyperref}
+\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
+\usepackage{setspace}
+\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
 \usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{indentfirst}
+\usepackage{titlesec}
+\usepackage{multirow}
+\usepackage{booktabs}
+\usepackage{graphicx}
+\usepackage{verbatim}
+\usepackage{rotating}
+\usepackage{tabularx}
+\usepackage{outlines}
 \usepackage{lineno}
+\usepackage{array}
+\usepackage{times}
 \usepackage{cleveref}
 \usepackage{acronym}
+\usepackage[position=t]{subfig}
 \usepackage{paralist}
-\usepackage{setspace}
+\usepackage[noae]{Sweave}
+\usepackage{natbib}
+\usepackage{array}
+\usepackage{pdflscape}
+\usepackage{bm}
 % \usepackage{showlabels}
+\bibpunct{(}{)}{,}{a}{}{,}
+
+% page margins and section title formatting
+\linespread{1.5}
+\setlength{\footskip}{0.5in}
+\titleformat*{\section}{\Large\bf\em}
+\titleformat*{\subsection}{\singlespace\large\bf}
+\titleformat*{\subsubsection}{\singlespace\normalsize\bf\em}
+\titlespacing{\section}{0in}{0in}{0in}
+\titlespacing{\subsection}{0in}{0in}{0in}
+\titlespacing{\subsubsection}{0in}{0in}{0in}
 
@@
 
-% turn this off for submission
-% \SectionNumbersOn
-
 %knitr options
 
-\author{Marcus W. Beck}
-\affiliation{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, Gulf Breeze, FL}
-\email{beck.marcus@epa.gov}
-\phone{+1 (850)9342480}
-\fax{+1 (850)9342401}
-\author{Thomas W. Jabusch}
-\author{Philip R. Trowbridge}
-\author{David B. Senn}
-\affiliation{San Francisco Estuary Institute, Richmond, CA}
+% get the version based on commit date
 
-\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
+% get online bib file
+
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
+
+\raggedbottom
 \linenumbers
+\raggedright
+\urlstyle{same}
+\setlength{\parindent}{0.5in}
+\renewcommand\refname{References \vspace{12pt}}
 
-\begin{abstract}
+\begin{singlespace}
+\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
+\author{
+  {\bf {\normalsize Marcus W. Beck$^1$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^3$, David B. Senn$^4$}}
+  \\\\{\textit {\normalsize $^1$Southern California Coastal Water Research Project}}
+  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
+	\\{\textit {\normalsize Phone: 714-755-3217, Fax: 714-755-3299, Email: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
+  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
+	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
+	\\{\textit {\normalsize Phone: 510-746-7340, Fax: 510-746-7300, Email: \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
+	\\\\{\textit {\normalsize $^3$San Francisco Estuary Institute}}
+	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
+	\\{\textit {\normalsize Phone: 510-746-7345, Fax: 510-746-7300, Email: \href{mailto:philt@sfei.org}{philt@sfei.org}}}
+	\\\\{\textit {\normalsize $^4$San Francisco Estuary Institute}}
+	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
+	\\{\textit {\normalsize Phone: 510-746-7366, Fax: 510-746-7300, Email: \href{mailto:davids@sfei.org}{davids@sfei.org}}}
+  \vspace{1in}
+  \\ Version Date:   Thu May 4 07:59:19 2017 -0500
+	}
+\date{}
+\maketitle
+\end{singlespace}
+
+\section*{Abstract}
 \noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.

### 2017-09-28 sftrends_manu.Rnw
additional formatting away from ES&T

@@
 \author{
-  {\bf {\normalsize Marcus W. Beck$^1$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^3$, David B. Senn$^4$}}
-  \\\\{\textit {\normalsize $^1$Southern California Coastal Water Research Project}}
+  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
+  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
+  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
+  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
+  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
   \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
-	\\{\textit {\normalsize Phone: 714-755-3217, Fax: 714-755-3299, Email: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
   \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
 	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7340, Fax: 510-746-7300, Email: \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
-	\\\\{\textit {\normalsize $^3$San Francisco Estuary Institute}}
-	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7345, Fax: 510-746-7300, Email: \href{mailto:philt@sfei.org}{philt@sfei.org}}}
-	\\\\{\textit {\normalsize $^4$San Francisco Estuary Institute}}
-	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7366, Fax: 510-746-7300, Email: \href{mailto:davids@sfei.org}{davids@sfei.org}}}
   \vspace{1in}
@@
 \end{singlespace}
+\clearpage
 
@@
 Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.
-Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.
+Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
 As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
 Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
-For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}.
-Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}.
+For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}.
+Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
 Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
@@
 Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.
-Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}.
+Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}.
 Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
-Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}.
+Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}.
 These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time.
-The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
-More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}.
+The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
+More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
 These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.
 
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \cite{Jassby00,Jassby02,Jassby08}.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \citep{Jassby00,Jassby02,Jassby08}.
 Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.
 Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
-Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.
-A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.
-Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
+Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
+A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \citep{Jabusch16}.
+Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \citep{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
 To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
@@
 The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay (\cref{fig:delt_map}).
-Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.
-Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}.
-Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.
+Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \citep{Jassby00}.
+Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \citep{Cloern12b}.
+Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \citep{Cloern15}.
 
 Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.
-Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.
-Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \cite{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\cite{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}
+Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\citep{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
+Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.
+However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures, \cite{Cloern07})  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}.
+
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
+A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}.
+For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.
+Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
+Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows and water exports directly from the system \citep{Jassby00,Jassby08}.
+Our analysis explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
 
@@
 
-Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.
+Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).
 Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
@@
 
-Daily flow estimates for the Delta region were obtained from the Dayflow software program \cite{IEP16}.
+Daily flow estimates for the Delta region were obtained from the Dayflow software program \citep{IEP16}.
 The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends.
@@
 Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.
-Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.
+Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.
 
@@
 A total of thirty \ac{wrtds} models were created, one for each nitrogen analyte at each station.
-The functional form of \ac{wrtds} is a simple regression\cite{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:
+The functional form of \ac{wrtds} is a simple regression \citep{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:
 
@@
 The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
-Models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
-All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT17}.
-
-A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows\cite{Hirsch10}.
+Models applied herein were based on a tidal adaptation of the original method \citep{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
+All analyses used the WRTDStidal package written by the authors for the R statistical programming language \citep{Beck16b,RDCT17}.
+
+A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10}.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
 Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
-These annual and seasonal groupings were chosen for continuity with similar comparisons in Ref.~\citenum{Jabusch16} and as approximate twenty year midpoints in the time series.
+These annual and seasonal groupings were chosen for continuity with similar comparisons in \citet{Jabusch16} and as approximate twenty year midpoints in the time series.
 
 Trends in each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
-This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82,Millard13}.
+This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82,Millard13}.
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
-The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
+The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation \citep{Jassby08}.
+Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
 Seasonal Kendall tests were also used to describe trends in the observed data.
@@
 
-\section{Results and Discussion}
+\section{Results}
 
@@
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower \citep{Novick15}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
@@
 P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites (see next section).
-Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).
+Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
 Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.
@@
 @
-Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
-These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
+Relationships between flow and observed water quality are complex and can change significantly through space and time \citep{Hirsch10,Zhang16}.
+These principles have been demonstrated for monitoring data in the Delta region \citep{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
 As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in Table S1) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
+For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
@@
 \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results.
-Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (Figures S1 and S2 and Tables S2 and S3).
+Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
 The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
 
-Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
-Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
-A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.
-For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
-For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.
-Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
-Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
-
 \subsection{Selected examples}
 
+Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
+The selected case studies focused on \begin{inparaenum}[1\upshape)]
+\item effects of wastewater treatment upgrades upstream of P8, and		
+\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.		
+\end{inparaenum}		
+Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
+
 \subsubsection{Effects of wastewater treatment}
 
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources \cite{Cornwell14,Novick15}.
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.
 Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
 A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
@@
 
-Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (Figure S3).
+Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).
 Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite.
@@
 Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January.
-A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
-These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
-For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
-Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.
-This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow.
-As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
 
@@
 Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
-Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply,\cite{Jassby08,Schoellhamer11,Cloern12b} although decreases in phytoplankton by clam biofiltration may have also increased clarity \cite{MacNally10}.
-The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
+Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \citep{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply \citep{Jassby08,Schoellhamer11,Cloern12b}, although decreases in phytoplankton by clam biofiltration may have also increased clarity \citep{MacNally10}.
+The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}.
 We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
@@
 \end{inparaenum}
-Although the relationship between phytoplankton and clams have been well described in \ac{sfe}\cite{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.
+Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.
 
@@
 These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
-This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
+Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
 Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
@@
 
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies \citep{Alpine92,Thompson08}.
 However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
@@
 In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.
+
+\section{Discussion}
+
+% obs v flow norm trends
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
+Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
+A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station \citep{Novick15}.
+For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
+For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.
+Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
+Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
+
+% P8 discussion
+A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
+These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
+For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
+Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.
+This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow.
+As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
+
+% Suisun conclusions
 As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
-These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
+These relationships have been suggested by others \citep{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
 
@@
 We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.
-Moreover, trends in nutrient loads from point sources in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases\cite{Jassby08} and exports to Suisun Bay\cite{Novick14}.
+Moreover, trends in nutrient loads from point sources in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases \citep{Jassby08} and exports to Suisun Bay \citep{Novick14}.
 The results from \ac{wrtds} demonstrating these changes are not unexpected, and consequently, we are not detracting from the potential implications of such increases.
@@
 For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow-normalized results.
-Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models).\cite{Beck17}  Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
+Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models, \cite{Beck17}).
+Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
 
@@
 The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset).
-All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.}
+All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).}
 \label{fig:delt_map}   
@@
 Trends are shown separately for different annual groupings.
-See Figures S1 and S2 for seasonal groupings.}
+See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
 \label{fig:trndcomp1}   
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1976-1995.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See Figure 3 for annual comparisons.}
+\label{fig:trndcomp2}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1996-2013.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See Figure 3 for annual comparisons.}
+\label{fig:trndcomp3}   
+\end{figure}
+
+<<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
+data(stock_conc)
+
+stock_conc <- filter(stock_conc, date >= as.Date('2003-01-01'))
+
+ptheme <- theme_minimal() + 
+  theme(
+    axis.line.x = element_line(), 
+    axis.line.y = element_line(), 
+    axis.ticks.x = element_line(),
+    axis.ticks.y = element_line(),
+    axis.ticks.length = unit(.1, "cm"),
+    axis.title.x = element_blank(),
+    legend.position = 'top', 
+    legend.title = element_blank()
+  )
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(3, 5, 8)] %>% 
+  rev
+
+toplo1 <- stock_conc %>% 
+  select(-val2) %>% 
+  spread(var, val) %>% 
+  mutate(TN = NH3 + NO2 + NO3)
+toplo2 <- stock_conc
+
+ylab1 <- expression(paste("Total nitrogen (mg ", L^-1, ")"))
+xrng <- range(c(toplo1$date, toplo2$date))
+yrng <- c(0, 50)
+leglabs <- list(
+  expression(NH [4] ^ '+'),
+  expression(NO [2] ^ '-'),
+  expression(NO [3] ^ '2-')
+  )
+
+lndate <- as.Date('2006/01/01')
+
+p1 <- ggplot(toplo2) + 
+  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
+  scale_x_date(expand = c(0, 0), limits = xrng) + 
+  scale_fill_manual(values = cols, labels = leglabs) + 
+  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
+  geom_label_repel(
+    data = data.frame(x = lndate, y = 35), 
+    aes(x, y, label = 'WWTP upgrades'), 
+    box.padding = grid::unit(2, "lines"), 
+    nudge_x = 1
+    ) +
+  scale_y_continuous(ylab1, expand = c(0, 0)) + 
+  ptheme
+
+p1
+@
+
 <<p8trnds, fig.height = 8, fig.width = 9, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \\ac{wrtds}, and (b) relationships with flow over time from \\ac{wrtds}.
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (Figure S3).'>>=
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\\cref{fig:stock}).'>>=
 data(h2dat)
@@
 
+% overall trends, years
+<<trndsann>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(cat %in% c('1976-1995', '1996-2013')) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code) %>% 
+  select(Site_Code, resvar, `1976-1995`, `1996-2013`)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations'
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+*$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = rep(nrow(tab)/3, 3),
+  cgroup = c('Annual'),
+  n.cgroup = c(2),
+  rowname = stats,
+  label = 'tab:trndsann',
+  insert.bottom = foot.val
+  )
+
+@
+
+% overall trends, seasons, first annual periodyears
+<<trndsbef>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(
+    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
+    ann %in% 'bef'
+    ) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995'
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = rep(nrow(tab)/3, 3),
+  cgroup = c('Seasonal, 1976-1995'),
+  n.cgroup = c(4),
+  rowname = stats,
+  label = 'tab:trndsbef',
+  insert.bottom = foot.val
+  )
+
+@
+
+% overall trends, seasons, first annual periodyears
+<<trndsaft>>=
+data(trnds_seasyr)
+
+# percent changes
+tab <- trnds_nrm %>% 
+  filter(
+    cat %in% c('Spring', 'Summer', 'Fall', 'Winter') &
+    ann %in% 'aft'
+    ) %>% 
+  select(Site_Code, resvar, cat, med, perchg, ztest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      paste0('(', perchg, ')' ), 
+      paste0('(\\textit{\\textbf{', perchg, '}})')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  mutate(
+    perchg = paste0('\\footnotesize{', perchg, '}')
+  ) %>% 
+  unite('val', med, perchg, sep = ' ') %>% 
+  spread(cat, val) %>% 
+  arrange(resvar, Site_Code)
+
+# table stuff
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013'
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$'
+vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+stats <- tab$Site_Code
+
+# table
+latex( 
+  tab[, -c(1, 2)],
+  file = '',
+  rowlabel = 'Analyte/Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = vars,
+  n.rgroup = rep(nrow(tab)/3, 3),
+  cgroup = c('Seasonal, 1996-2013'),
+  n.cgroup = c(4),
+  rowname = stats,
+  label = 'tab:trndsaft',
+  insert.bottom = foot.val
+  )
+
+@
+
 % trends in no23, nh at p8
@@
 cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8'
-foot.val<-'\\footnotesize Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, Figure S3).
+foot.val<-'\\footnotesize Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, \\cref{fig:stock}).
 Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).

### 2017-09-28 sftrends_manu.tex
additional formatting away from ES&T

@@
 \author{
-  {\bf {\normalsize Marcus W. Beck$^1$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^3$, David B. Senn$^4$}}
-  \\\\{\textit {\normalsize $^1$Southern California Coastal Water Research Project}}
+  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
+  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
+  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
+  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
+  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
   \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
-	\\{\textit {\normalsize Phone: 714-755-3217, Fax: 714-755-3299, Email: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
   \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
 	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7340, Fax: 510-746-7300, Email: \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
-	\\\\{\textit {\normalsize $^3$San Francisco Estuary Institute}}
-	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7345, Fax: 510-746-7300, Email: \href{mailto:philt@sfei.org}{philt@sfei.org}}}
-	\\\\{\textit {\normalsize $^4$San Francisco Estuary Institute}}
-	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
-	\\{\textit {\normalsize Phone: 510-746-7366, Fax: 510-746-7300, Email: \href{mailto:davids@sfei.org}{davids@sfei.org}}}
   \vspace{1in}
-  \\ Version Date:   Thu May 4 07:59:19 2017 -0500
+  \\ Version Date:   Wed Sep 27 18:07:07 2017 -0700
 	}
@@
 The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
-Models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.
-All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT17}.
+Models applied herein were based on a tidal adaptation of the original method \citep{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
+All analyses used the WRTDStidal package written by the authors for the R statistical programming language \citep{Beck16b,RDCT17}.
 
-A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows\cite{Hirsch10}.
+A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10}.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
 Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
-These annual and seasonal groupings were chosen for continuity with similar comparisons in Ref.~\citenum{Jabusch16} and as approximate twenty year midpoints in the time series.
+These annual and seasonal groupings were chosen for continuity with similar comparisons in \citet{Jabusch16} and as approximate twenty year midpoints in the time series.
 
 Trends in each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions.
-This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82,Millard13}.
+This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82,Millard13}.
 Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.
-The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
+The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation \citep{Jassby08}.
+Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.
 Seasonal Kendall tests were also used to describe trends in the observed data.
@@
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower \citep{Novick15}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
@@
 P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites (see next section).
-Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).
+Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
 Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.
 
-Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}.
-These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
+Relationships between flow and observed water quality are complex and can change significantly through space and time \citep{Hirsch10,Zhang16}.
+These principles have been demonstrated for monitoring data in the Delta region \citep{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
 As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in Table S1) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
+For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
@@
 Eleven comparisons showed a trend reversal from positive to negative estimated change, nine sites went from no change to negative estimated change, and one site went from no change to a positive trend for the flow-normalized results.
-Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (Figures S1 and S2 and Tables S2 and S3).
+Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
 The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
-
-Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
-Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
-A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.
-For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
-For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.
-Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
-Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
 
@@
 
+Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
+The selected case studies focused on \begin{inparaenum}[1\upshape)]
+\item effects of wastewater treatment upgrades upstream of P8, and		
+\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.		
+\end{inparaenum}		
+Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
+
 \subsubsection{Effects of wastewater treatment}
 
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources \cite{Cornwell14,Novick15}.
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.
 Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
 A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
@@
 Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January.
+
+\subsubsection{Effects of biological invasions}
+
+Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
+Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \citep{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply \citep{Jassby08,Schoellhamer11,Cloern12b}, although decreases in phytoplankton by clam biofiltration may have also increased clarity \citep{MacNally10}.
+The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}.
+We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
+\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
+\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
+\end{inparaenum}
+Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.
+
+Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P.
+amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
+The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.
+A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).
+An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).
+These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
+Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
+Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
+amurensis}.
+
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies \citep{Alpine92,Thompson08}.
+However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
+The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
+In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
+Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
+The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.
+Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a).
+In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.
+
+\section{Discussion}
+
+% obs v flow norm trends
+Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
+Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
+A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station \citep{Novick15}.
+For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
+For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.
+Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
+Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
+
+% P8 discussion
 A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
@@
 
-\subsubsection{Effects of biological invasions}
-
-Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
-Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply,\cite{Jassby08,Schoellhamer11,Cloern12b} although decreases in phytoplankton by clam biofiltration may have also increased clarity \cite{MacNally10}.
-The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.
-We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
-\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
-\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
-\end{inparaenum}
-Although the relationship between phytoplankton and clams have been well described in \ac{sfe}\cite{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.
-
-Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P.
-amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
-The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.
-A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).
-An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).
-These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
-This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
-Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
-amurensis}.
-
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.
-However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
-The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
-In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
-Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
-The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.
-Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a).
-In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.
+% Suisun conclusions
 As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
-These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
+These relationships have been suggested by others \citep{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
 
@@
 Trends are shown separately for different annual groupings.
-See Figures S1 and S2 for seasonal groupings.}
+See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
 \label{fig:trndcomp1}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1976-1995.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See Figure 3 for annual comparisons.}
+\label{fig:trndcomp2}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1996-2013.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See Figure 3 for annual comparisons.}
+\label{fig:trndcomp3}   
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\textwidth]{figs/stock-1}
+
+}
+
+\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
+Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
 \end{figure}
@@
 
-\caption[Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}]{Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}.
+\caption{Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}.
 Nitrite/nitrate trends are on the left and ammonium trends are on the right.
-Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (Figure S3).}\label{fig:p8trnds}
+Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\cref{fig:stock}).}\label{fig:p8trnds}
 \end{figure}
@@
 % tables
+
+% overall trends, years
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann", insert.bottom = foot.val)%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations\label{tab:trndsann}} 
+\begin{center}
+\begin{tabular}{lll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{2}{c}{\bfseries Annual}\tabularnewline
+\cline{2-3}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1995}&\multicolumn{1}{c}{1996-2013}\tabularnewline
+\hline
+{\bfseries DIN}&&\tabularnewline
+~~C10&1.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.4 \footnotesize{(-3.1)**}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.5 \footnotesize{(-0.1)**}\tabularnewline
+~~D19&0.4 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-1.2)**}\tabularnewline
+~~D28&0.4 \footnotesize{(\textit{\textbf{0.1}})**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
+~~D6&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.5 \footnotesize{(-0.3)**}\tabularnewline
+~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}\tabularnewline
+~~MD10&0.4 \footnotesize{(-1.1)**}&0.3 \footnotesize{(-2.4)**}\tabularnewline
+~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}\tabularnewline
+\hline
+{\bfseries NH$_{4}^{+}$}&&\tabularnewline
+~~C10&0.1 \footnotesize{(-3.4)**}&0 \footnotesize{(-5.2)**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{3.7}})**}&0.3 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
+~~D19&0 \footnotesize{(\textit{\textbf{0.4}})**}&0 \footnotesize{(-1.7)**}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{2.2}})**}&0.1 \footnotesize{(-1.5)**}\tabularnewline
+~~D28&0 \footnotesize{(-1.1)**}&0 \footnotesize{(-1.4)**}\tabularnewline
+~~D4&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{2.4}})**}&0.1 \footnotesize{(-0.5)**}\tabularnewline
+~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}\tabularnewline
+~~MD10&0.1 \footnotesize{(-2.8)**}&0 \footnotesize{(-1.1)**}\tabularnewline
+~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}\tabularnewline
+\hline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{1.4}})**}&1.4 \footnotesize{(-3)**}\tabularnewline
+~~C3&0.1 \footnotesize{(-0.1)**}&0.2 \footnotesize{(\textit{\textbf{0.7}})**}\tabularnewline
+~~D19&0.4 \footnotesize{(-0.1)**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
+~~D26&0.3 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1.1)**}\tabularnewline
+~~D28&0.4 \footnotesize{(-0.2)**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(-0.4)**}\tabularnewline
+~~D6&0.3 \footnotesize{(\textit{\textbf{1.3}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
+~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}\tabularnewline
+~~MD10&0.4 \footnotesize{(-1)**}&0.3 \footnotesize{(-2.5)**}\tabularnewline
+~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}\tabularnewline
+\hline
+\end{tabular}\end{center}
+\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+*$p<0.05$; **$p<0.005$\end{table}
+
+% overall trends, seasons, first annual periodyears
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef", insert.bottom = foot.val)%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995\label{tab:trndsbef}} 
+\begin{center}
+\begin{tabular}{lllll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1976-1995}\tabularnewline
+\cline{2-5}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
+\hline
+{\bfseries DIN}&&&&\tabularnewline
+~~C10&1.2 \footnotesize{(\textit{\textbf{1.1}})**}&1.2 \footnotesize{(\textit{\textbf{0.3}})}&1.3 \footnotesize{(\textit{\textbf{0.5}})**}&1.7 \footnotesize{(\textit{\textbf{1.2}})**}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.3 \footnotesize{(\textit{\textbf{2.3}})**}&0.4 \footnotesize{(\textit{\textbf{2.4}})**}&0.4 \footnotesize{(\textit{\textbf{1.9}})**}\tabularnewline
+~~D19&0.5 \footnotesize{(\textit{\textbf{0.3}})}&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.7 \footnotesize{(-0.2)}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(\textit{\textbf{0.4}})*}&0.4 \footnotesize{(\textit{\textbf{1}})**}&0.6 \footnotesize{(\textit{\textbf{0.3}})}\tabularnewline
+~~D28&0.5 \footnotesize{(\textit{\textbf{0.8}})*}&0.2 \footnotesize{(\textit{\textbf{0.3}})}&0.3 \footnotesize{(\textit{\textbf{0.5}})*}&0.8 \footnotesize{(-0.3)}\tabularnewline
+~~D4&0.4 \footnotesize{(\textit{\textbf{0.2}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.5 \footnotesize{(-0.5)}\tabularnewline
+~~D6&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{4.6}})**}&0.4 \footnotesize{(\textit{\textbf{1.4}})**}&0.5 \footnotesize{(-0.7)*}\tabularnewline
+~~D7&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.6 \footnotesize{(-2.4)**}\tabularnewline
+~~MD10&0.6 \footnotesize{(-0.3)}&0.2 \footnotesize{(-3.6)**}&0.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.3 \footnotesize{(-0.3)*}\tabularnewline
+~~P8&1.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.9 \footnotesize{(\textit{\textbf{2.4}})**}&1.3 \footnotesize{(\textit{\textbf{3.1}})**}&1.9 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
+\hline
+{\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
+~~C10&0.1 \footnotesize{(-2.3)**}&0 \footnotesize{(-6.8)**}&0.1 \footnotesize{(-7.1)**}&0.3 \footnotesize{(-1.5)**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.2 \footnotesize{(\textit{\textbf{4}})**}&0.3 \footnotesize{(\textit{\textbf{3.8}})**}&0.2 \footnotesize{(\textit{\textbf{2.9}})**}\tabularnewline
+~~D19&0.1 \footnotesize{(\textit{\textbf{0.4}})*}&0 \footnotesize{(-1.7)**}&0 \footnotesize{(\textit{\textbf{1.2}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}\tabularnewline
+~~D26&0.1 \footnotesize{(\textit{\textbf{1.4}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.1 \footnotesize{(\textit{\textbf{2.3}})**}\tabularnewline
+~~D28&0.1 \footnotesize{(-0.5)}&0 \footnotesize{(-3.7)**}&0 \footnotesize{(-1.6)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
+~~D4&0.1 \footnotesize{(\textit{\textbf{1.7}})**}&0 \footnotesize{(\textit{\textbf{1}})**}&0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textit{\textbf{2}})**}\tabularnewline
+~~D6&0.1 \footnotesize{(\textit{\textbf{2.9}})**}&0.1 \footnotesize{(\textit{\textbf{2.8}})**}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
+~~D7&0.1 \footnotesize{(\textit{\textbf{3.3}})**}&0 \footnotesize{(\textit{\textbf{2}})**}&0.1 \footnotesize{(-2.8)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
+~~MD10&0.1 \footnotesize{(-1.8)**}&0 \footnotesize{(-6.5)**}&0 \footnotesize{(-3.3)**}&0.2 \footnotesize{(\textit{\textbf{0.4}})}\tabularnewline
+~~P8&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.1 \footnotesize{(\textit{\textbf{1.8}})**}&0.2 \footnotesize{(\textit{\textbf{7}})**}&0.6 \footnotesize{(\textit{\textbf{7}})**}\tabularnewline
+\hline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
+~~C10&1.1 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(\textit{\textbf{0.6}})**}&1.2 \footnotesize{(\textit{\textbf{1.3}})**}&1.5 \footnotesize{(\textit{\textbf{1.8}})**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{0.7}})**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+~~D19&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.2 \footnotesize{(-0.3)}&0.3 \footnotesize{(\textit{\textbf{0.3}})}&0.6 \footnotesize{(-0.9)*}\tabularnewline
+~~D26&0.4 \footnotesize{(\textit{\textbf{0.6}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.3}})*}&0.5 \footnotesize{(-0.3)}\tabularnewline
+~~D28&0.5 \footnotesize{(\textit{\textbf{0.7}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.2}})}&0.7 \footnotesize{(-1)**}\tabularnewline
+~~D4&0.3 \footnotesize{(\textit{\textbf{0.1}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.4 \footnotesize{(-0.8)*}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.1}})**}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.4 \footnotesize{(-1)**}\tabularnewline
+~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textit{\textbf{3.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.4 \footnotesize{(-3.6)**}\tabularnewline
+~~MD10&0.5 \footnotesize{(-0.2)}&0.2 \footnotesize{(-3.6)**}&0.2 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(-0.5)*}\tabularnewline
+~~P8&1.2 \footnotesize{(\textit{\textbf{2}})**}&0.9 \footnotesize{(\textit{\textbf{2.3}})**}&1.1 \footnotesize{(\textit{\textbf{2}})**}&1.4 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+\hline
+\end{tabular}\end{center}
+\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$\end{table}
+
+% overall trends, seasons, first annual periodyears
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft", insert.bottom = foot.val)%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013\label{tab:trndsaft}} 
+\begin{center}
+\begin{tabular}{lllll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1996-2013}\tabularnewline
+\cline{2-5}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
+\hline
+{\bfseries DIN}&&&&\tabularnewline
+~~C10&1.1 \footnotesize{(-4.1)**}&1.3 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-2)**}&1.7 \footnotesize{(-3.4)**}\tabularnewline
+~~C3&0.5 \footnotesize{(\textit{\textbf{0.5}})}&0.4 \footnotesize{(\textit{\textbf{0.1}})}&0.6 \footnotesize{(-0.2)}&0.5 \footnotesize{(-0.6)**}\tabularnewline
+~~D19&0.5 \footnotesize{(-2.8)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.6)**}&0.7 \footnotesize{(-2.3)**}\tabularnewline
+~~D26&0.5 \footnotesize{(-1.9)**}&0.3 \footnotesize{(-1.7)**}&0.4 \footnotesize{(-1)**}&0.6 \footnotesize{(-0.8)**}\tabularnewline
+~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-4.9)**}&0.2 \footnotesize{(-4.9)**}&0.7 \footnotesize{(-2.1)**}\tabularnewline
+~~D4&0.4 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.9)**}&0.5 \footnotesize{(\textit{\textbf{0.6}})**}\tabularnewline
+~~D6&0.5 \footnotesize{(-0.2)*}&0.5 \footnotesize{(-1)**}&0.5 \footnotesize{(-0.3)*}&0.5 \footnotesize{(-0.1)}\tabularnewline
+~~D7&0.5 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-1.3)**}&0.4 \footnotesize{(-0.4)**}&0.6 \footnotesize{(-0.2)}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.3)**}&0.2 \footnotesize{(-3.7)**}&0.2 \footnotesize{(-4.4)**}&1 \footnotesize{(-1.8)**}\tabularnewline
+~~P8&1.5 \footnotesize{(-1.9)**}&1.2 \footnotesize{(-3.5)**}&1.8 \footnotesize{(-2.4)**}&2.7 \footnotesize{(-2.2)**}\tabularnewline
+\hline
+{\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
+~~C10&0 \footnotesize{(-4.2)**}&0 \footnotesize{(-6.1)**}&0 \footnotesize{(-8.5)**}&0.1 \footnotesize{(-7.3)**}\tabularnewline
+~~C3&0.3 \footnotesize{(\textit{\textbf{1}})**}&0.3 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-0.1)}\tabularnewline
+~~D19&0 \footnotesize{(-1.9)**}&0 \footnotesize{(-0.4)}&0 \footnotesize{(-2.2)**}&0.1 \footnotesize{(-1.8)**}\tabularnewline
+~~D26&0.1 \footnotesize{(-1.2)**}&0.1 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-1.9)**}&0.1 \footnotesize{(-1.4)**}\tabularnewline
+~~D28&0 \footnotesize{(-1.7)**}&0 \footnotesize{(-0.2)}&0 \footnotesize{(-2.4)**}&0.1 \footnotesize{(-3.1)**}\tabularnewline
+~~D4&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-0.3)}&0.1 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+~~D6&0.1 \footnotesize{(-0.7)**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0.1 \footnotesize{(-0.3)**}\tabularnewline
+~~D7&0.1 \footnotesize{(-2.2)**}&0 \footnotesize{(-2.1)**}&0.1 \footnotesize{(\textit{\textbf{1.3}})**}&0.1 \footnotesize{(-0.4)*}\tabularnewline
+~~MD10&0 \footnotesize{(-1.4)*}&0 \footnotesize{(-0.1)}&0 \footnotesize{(-0.8)**}&0.1 \footnotesize{(-4.3)**}\tabularnewline
+~~P8&0.2 \footnotesize{(-8.7)**}&0.1 \footnotesize{(-6.3)**}&0.2 \footnotesize{(-10.4)**}&0.5 \footnotesize{(-13.1)**}\tabularnewline
+\hline
+{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
+~~C10&1.1 \footnotesize{(-4.2)**}&1.2 \footnotesize{(-3.2)**}&1.6 \footnotesize{(-1.9)**}&1.6 \footnotesize{(-3.3)**}\tabularnewline
+~~C3&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.2 \footnotesize{(\textit{\textbf{1.7}})**}&0.2 \footnotesize{(-0.4)}\tabularnewline
+~~D19&0.4 \footnotesize{(-2.9)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.5)**}&0.6 \footnotesize{(-2.2)**}\tabularnewline
+~~D26&0.4 \footnotesize{(-1.9)**}&0.2 \footnotesize{(-1.6)**}&0.3 \footnotesize{(-0.6)*}&0.5 \footnotesize{(-0.6)**}\tabularnewline
+~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-5.4)**}&0.2 \footnotesize{(-5.2)**}&0.7 \footnotesize{(-1.7)**}\tabularnewline
+~~D4&0.3 \footnotesize{(-0.1)}&0.3 \footnotesize{(-1)**}&0.3 \footnotesize{(-1)**}&0.4 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.4)*}&0.4 \footnotesize{(-0.1)}\tabularnewline
+~~D7&0.4 \footnotesize{(-0.6)**}&0.4 \footnotesize{(-1.2)**}&0.4 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-0.3)*}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.6)**}&0.1 \footnotesize{(-4.5)**}&0.2 \footnotesize{(-5.4)**}&1 \footnotesize{(-1.4)**}\tabularnewline
+~~P8&1.3 \footnotesize{(-1.1)**}&1.1 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
+\hline
+\end{tabular}\end{center}
+\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$; **$p<0.005$\end{table}
 
@@
 \end{tabular}\end{center}
-\footnotesize Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, Figure S3).
+\footnotesize Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
 Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).

### 2017-09-28 sftrends_manu.Rnw
abstract rewrite with key results

@@
 \section*{Abstract}
-\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.
-We apply an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}.
-Results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
-Trend analysis with \ac{wrtds} flow-normalized data demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
-We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
-Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
+\noindent Long-term monitoring data from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds} to describe historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
+Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
+We further described mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
+\ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions.
+Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and confounded by grazing pressure.
+Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
 \acresetall
@@
 Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
+
+% Station selection adnd flow-records? - comment from reviewer 1
+
+% limitations in the method - why we did not look at load?  Also, use of other tracers of flow/diversions - why we did not look at salinity?
 

### 2017-09-28 sftrends_manu.tex
abstract rewrite with key results

@@
   \vspace{1in}
-  \\ Version Date:   Wed Sep 27 18:07:07 2017 -0700
+  \\ Version Date:   Thu Sep 28 10:50:58 2017 -0700
 	}

### 2017-09-29 sftrends_manu.Rnw
working on better intro, lit review

@@
 % page margins and section title formatting
-\linespread{1.5}
+\linespread{2}
 \setlength{\footskip}{0.5in}
@@
 
-Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.
-Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the Central Valley watershed \citep{Jassby00,Jassby02,Jassby08}.
+Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.
+Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
+Changes in response to stressors suggests that recent conditions in \ac{sfe} have not followed past trajectories.
+For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in South Bay \citep{Lehman05,Cloern07,Sutula17}.
+These changes stem from multiple potential sources, although variation in freshwater inputs/withdrawals, establishment of invasive species, and effects of climate change have been implicated as key factors \citep{Cloern12b}.
+Many of these changes are linked to inputs from the Delta region.
+\citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.
+Changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
+
+Formal methods for trend analysis methods could help describe recent changes in water quality in the Delta region.
+Direct evaluation of observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
 As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
@@
 Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
-Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
-
-Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.
-Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}.
-Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
-Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}.
-These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time.
+
 The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
-More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
-These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.
-
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \citep{Jassby00,Jassby02,Jassby08}.
-Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.
-Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
-Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
-A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \citep{Jabusch16}.
-Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \citep{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
-To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
-
+The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}.
 The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
-We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts.
+We applied an estuarine adaptation of \ac{wrtds} to describe nitrogen trends in  different spatial and temporal contexts.
 The specific objectives were to \begin{inparaenum}[1\upshape)]
@@
 The second objective evaluated two specific water quality stations as additional case studies to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
+Our general hypothesis was that the results were expected to support previous descriptions of trends in this well-studied system, but that new insight into spatial and temporal variation in response endpoints was expected, particuarly in flow-normalized model predictions.
 
@@
 The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures, \cite{Cloern07})  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}.
+However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures, \cite{Cloern07})
+
+The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}. % see glibert14 description
 

### 2017-09-29 sftrends_manu.tex
working on better intro, lit review

@@
   \vspace{1in}
-  \\ Version Date:   Thu Sep 28 10:50:58 2017 -0700
+  \\ Version Date:   Thu Sep 28 11:58:28 2017 -0700
 	}
@@
 Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
-Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.
 
-Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.
-Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}.
-Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data.
-Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}.
-These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time.
 The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
-More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}.
-These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.
-
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \citep{Jassby00,Jassby02,Jassby08}.
-Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.
-Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
-Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.
-A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \citep{Jabusch16}.
-Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \citep{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.
-To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.
-
+The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}.
 The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
-We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts.
+We applied an estuarine adaptation of \ac{wrtds} to describe nitrogen trends in  different spatial and temporal contexts.
 The specific objectives were to \begin{inparaenum}[1\upshape)]
@@
 The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures, \cite{Cloern07})  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}.
+However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures, \cite{Cloern07})
+
+The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}.
 

### 2017-10-03 sftrends_manu.Rnw
better intro maybe done

@@
 \noindent Long-term monitoring data from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds} to describe historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
-Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
-We further described mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
+Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
+We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
 \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions.
@@
 The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the Central Valley watershed \citep{Jassby00,Jassby02,Jassby08}.
-Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.
-Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
-Changes in response to stressors suggests that recent conditions in \ac{sfe} have not followed past trajectories.
-For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in South Bay \citep{Lehman05,Cloern07,Sutula17}.
-These changes stem from multiple potential sources, although variation in freshwater inputs/withdrawals, establishment of invasive species, and effects of climate change have been implicated as key factors \citep{Cloern12b}.
-Many of these changes are linked to inputs from the Delta region.
+Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the larger Bay.
+Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories.
+For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in the lower bay \citep{Lehman05,Cloern07,Sutula17}.
+Although these changes are potentially linked to drivers at different spatial and temporal scales, inputs from the Delta remain a critical interest for understanding downstream effects.
+
+Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}.
+In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic and benthic species can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
+Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
 \citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.
-Changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
-
-Formal methods for trend analysis methods could help describe recent changes in water quality in the Delta region.
-Direct evaluation of observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
+Changes in phytoplankton biomass and the effects on pelagic organisms have also been of concern.
+\citet{Jassby08} described decadal trends in phytoplankton biomass to understand mechanisms of decline for pelagic fish populations in the Delta.
+Although phytoplankon concentrations have been relatively consistent in Suisan Bay, biomass in the upper Delta has been increasing.
+Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.
+
+A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep{IEP13}.
+Although these data have been used extensively, a systematic assessment of trends covering the full spatial and temporal coverage of the monitoring dataset has not been made.
+This information could inform both past and current research by providing a cohesive description of system-wide trends in nutrients.
+However, formal methods for trend analysis are required given that long-term changes can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
 As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
@@
 The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
-We applied an estuarine adaptation of \ac{wrtds} to describe nitrogen trends in  different spatial and temporal contexts.
 The specific objectives were to \begin{inparaenum}[1\upshape)]
@@
 \end{inparaenum}
-The second objective evaluated two specific water quality stations as additional case studies to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
+The second objective evaluated two specific water quality stations to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
 Our general hypothesis was that the results were expected to support previous descriptions of trends in this well-studied system, but that new insight into spatial and temporal variation in response endpoints was expected, particuarly in flow-normalized model predictions.
@@
 
-The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}. % see glibert14 description
-
 Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
@@
 Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
-These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.
+These stations include all but two of the long-term monitoring stations from the Delta and cover all of the major inflow and outflows \citep{Jabusch09}.
 Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.

### 2017-10-03 sftrends_manu.tex
better intro maybe done

@@
   \vspace{1in}
-  \\ Version Date:   Thu Sep 28 11:58:28 2017 -0700
+  \\ Version Date:   Fri Sep 29 12:02:06 2017 -0700
 	}
@@
 The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the Central Valley watershed \citep{Jassby00,Jassby02,Jassby08}.
-Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.
-Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.
-Changes in response to stressors suggests that recent conditions in \ac{sfe} have not followed past trajectories.
-For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in South Bay \citep{Lehman05,Cloern07,Sutula17}.
-These changes stem from multiple potential sources, although variation in freshwater inputs/withdrawals, establishment of invasive species, and effects of climate change have been implicated as key factors \citep{Cloern12b}.
-Many of these changes are linked to inputs from the Delta region.
+Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the larger Bay.
+Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories.
+For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in the lower bay \citep{Lehman05,Cloern07,Sutula17}.
+Although these changes are potentially linked to drivers at different spatial and temporal scales, inputs from the Delta remain a critical interest for understanding downstream effects.
+
+Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}.
+In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic and benthic species can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
+Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
 \citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.
-Changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
+Changes in phytoplankton biomass and the effects on pelagic organisms have also been of concern.
+\citet{Jassby08} described decadal trends in phytoplankton biomass to understand mechanisms of decline for pelagic fish populations in the Delta.
+Although phytoplankon concentrations have been relatively consistent in Suisan Bay, biomass in the upper Delta has been increasing.
+Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.
 
-Formal methods for trend analysis methods could help describe recent changes in water quality in the Delta region.
-Direct evaluation of observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
+A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep{IEP13}.
+Although these data have been used extensively, a systematic assessment of trends covering the full spatial and temporal coverage of the monitoring dataset has not been made.
+This information could inform both past and current research by providing a cohesive description of system-wide trends in nutrients.
+However, formal methods for trend analysis are required given that long-term changes can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
 As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
@@
 
-The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}.
-
 Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.

### 2017-10-06 sftrends_manu.Rnw
intro and methods text edits, working on annual agg figs

@@
 \section*{Abstract}
-\noindent Long-term monitoring data from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds} to describe historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).
+\noindent Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.
 Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
@@
 
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the Central Valley watershed \citep{Jassby00,Jassby02,Jassby08}.
-Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the larger Bay.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
+Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the larger Bay \citep{Dugdale07,Cornwell14}.
 Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories.
-For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in the lower bay \citep{Lehman05,Cloern07,Sutula17}.
-Although these changes are potentially linked to drivers at different spatial and temporal scales, inputs from the Delta remain a critical interest for understanding downstream effects.
+Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Shellenbarger08,Cloern07}.
+As a result, \ac{chla} thresholds to assess and manage levels of concern for phtoplankton biomass in the lower bay have been proposed \citep{Sutula17}.
+Although these changes are linked to drivers at different spatial and temporal scales, inputs from the Delta remain a critical interest for understanding downstream effects.
 
@@
 \citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.
+Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}.
 Changes in phytoplankton biomass and the effects on pelagic organisms have also been of concern.
@@
 Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.
-Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\citep{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
+Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
 Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
 The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures, \cite{Cloern07})
+However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales \citep[e.g., variation in sea surface temperatures,][]{Cloern07}.
 
@@
 
-Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).
+Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).
 Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
-These stations include all but two of the long-term monitoring stations from the Delta and cover all of the major inflow and outflows \citep{Jabusch09}.
-Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.
+These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}.
+Although many other stations are available for the region, the stations were chosen becaase they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}.
+Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.
 Data were minimally processed, with the exception of averaging replicates that occurred on the same day.
@@
 \noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
-The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
-Models applied herein were based on a tidal adaptation of the original method \citep{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
-All analyses used the WRTDStidal package written by the authors for the R statistical programming language \citep{Beck16b,RDCT17}.
+Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
+Rather than fitting a global model to the entire time series, one regression is fit at every point in the time series.
+Observations within a window used to fit each regression are weighted relative to annual, seasonal, and flow distances from the center of the window.
+Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.
+This approach allows for a type of smoothing where the observed fit is specific to each point in the time series.
+Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.
+All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
+Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September).
+All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.
 
@@
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
-Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
+Flow-normalization predictions for each month of each year were based on the average of predictions for flow values that occur in the same month across all years, weighted within each specific month and year for every observation.
+Flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
 These annual and seasonal groupings were chosen for continuity with similar comparisons in \citet{Jabusch16} and as approximate twenty year midpoints in the time series.
@@
 
+<<prddat, fig.cap = 'Model predictions.', fig.height = 8, fig.width = 8>>=
+library(tidyverse)
+library(grid)
+source('R/funcs.R')
+
+data(mods)
+
+toplo <- mods %>% 
+  mutate(
+    anns = map(data, annual_agg)
+  ) %>% 
+  select(-data, -flovar) %>% 
+  unnest %>% 
+  group_by(resvar) %>% 
+  nest %>% 
+  mutate(
+    plts = map(data, function(x){
+
+      p <- ggplot(x, aes(x = date, y = bt_fits, group = Site_Code, colour = Site_Code)) +
+        geom_point(alpha = 0.7) + 
+        geom_line(aes(y = bt_norm)) +
+        facet_wrap(~ Location, ncol = 1, scales = 'free_y') + 
+        theme_minimal() + 
+        theme(
+          legend.position = 'none', 
+          axis.title.x = element_blank(), 
+          axis.title.y = element_blank(), 
+          panel.border = element_rect(colour = 'black', fill = NA, size = 0.5), 
+          strip.text.x = element_text(size = 10)
+          )
+
+      return(p)
+
+    })
+  )
+
+pleg <- ggplot(toplo$data[[1]], aes(x = date, y = bt_fits, group = Site_Code, colour = Site_Code)) +
+  geom_point(alpha = 0.7) + 
+  geom_line(aes(y = bt_norm)) +
+  theme_minimal() + 
+  guides(colour = guide_legend(ncol = 2))
+pleg <- g_legend(pleg)
+
+ylab1 <- expression(paste(NH [4] ^ '+', ' (mg ', L^-1, ')'))
+ylab2 <- expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-', ' (mg ', L^-1, ')'))
+ylab3 <- expression(paste('DIN (mg ', L^-1, ')'))
+
+# (2, 3), (3, 3)
+pA <- ggplot_gtable(ggplot_build(toplo$plts[[1]]))
+pB<- ggplot_gtable(ggplot_build(toplo$plts[[2]]))
+pC <- ggplot_gtable(ggplot_build(toplo$plts[[3]]))
+maxWidth = unit.pmax(pA$widths[2:3], pB$widths[c(2:3)], pC$widths[c(2:3)])
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+pC$widths[2:3] <- maxWidth
+
+grid.arrange(
+  arrangeGrob(
+    arrangeGrob(pA, left = textGrob(ylab1, rot = 90)),
+    arrangeGrob(pC, left = textGrob(ylab3, rot = 90)), 
+    ncol = 1
+  ), 
+  arrangeGrob(
+    arrangeGrob(pB, left = textGrob(ylab2, rot = 90)), 
+    pleg, ncol = 1
+  ), 
+  ncol = 2
+  )
+
+@
+
 <<trndcomp, echo = F, warning = F, message = F, cache = T, results = 'hide'>>=
@@
 # (2, 3), (3, 3)
-pE <- ggplot_gtable(ggplot_build(p5))
+pE <- ggplot_gtable <- (ggplot_build(p5))
 pH   <- ggplot_gtable(ggplot_build(p8))

### 2017-10-06 sftrends_manu.tex
intro and methods text edits, working on annual agg figs

@@
   \vspace{1in}
-  \\ Version Date:   Fri Sep 29 12:02:06 2017 -0700
+  \\ Version Date:   Tue Oct 3 12:05:52 2017 -0700
 	}
@@
 
+\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figs/prddat-1}
+
+}
+
+\caption[Model predictions]{Model predictions.}\label{fig:prddat}
+\end{figure}
+
 \begin{figure}
@@
 
-\begin{figure}[!ht]
+\begin{kframe}
 
-{\centering \includegraphics[width=0.9\textwidth]{figs/clmchl-1}
-
-}
-
-\caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
-Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and reduction of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
-A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \ac{wrtds} results.
-Reductions in \ac{chla} concentration were coincident with an increase in SiO$_2$ and \ac{din} concentrations (d, e), with the greatest increases in August (g, h).
-A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (i).
-Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (j, observations from June).}\label{fig:clmchl}
-\end{figure}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in gList(structure(list(data = list(structure(list(colour = c("{}\#66C2A5"{}, : only 'grobs' allowed in "{}gList"{}}}\end{kframe}
 

### 2017-10-06 sftrends_manu.Rnw
a pointless figure

@@
 library(WRTDStidal)
+library(RColorBrewer)
 library(tidyverse)
@@
 
+cols <- brewer.pal(10, 'Set1') %>% 
+  colorRampPalette(.)
+
 toplo <- mods %>% 
   mutate(
-    anns = map(data, annual_agg)
+    anns = map(data, annual_agg), 
+    Location = factor(Location, 
+                      levels = c('Delta', 'Middle', 'Suisun'), 
+                      labels = c('Peripheral', 'Interior', 'Suisun')
+                      )
   ) %>% 
@@
           strip.text.x = element_text(size = 10)
-          )
+          ) + 
+        scale_colour_manual(values = cols(10))
 
@@
 
-pleg <- ggplot(toplo$data[[1]], aes(x = date, y = bt_fits, group = Site_Code, colour = Site_Code)) +
-  geom_point(alpha = 0.7) + 
-  geom_line(aes(y = bt_norm)) +
+stat_meta <- delt_dat %>% 
+  dplyr::select(Site_Code, Location) %>% 
+  unique %>% 
+  mutate(
+    Location = factor(Location, 
+           levels = c('Delta', 'Middle', 'Suisun'), 
+           labels = c('Peripheral', 'Interior', 'Suisun')
+    ),
+    colv = cols(10), 
+    pts = c(1:10)
+  )
+peri <- stat_meta %>% 
+  filter(Location == 'Peripheral')
+inte <- stat_meta %>% 
+  filter(Location == 'Interior') 
+suis <- stat_meta %>% 
+  filter(Location == 'Suisun')
+
+perileg <- ggplot(peri, aes(x = pts, y = pts, colour = Site_Code)) +
+  geom_point() +
+  geom_line() +
   theme_minimal() + 
-  guides(colour = guide_legend(ncol = 2))
-pleg <- g_legend(pleg)
+  scale_colour_manual(values = peri$colv, name = 'Peripheral') +
+  theme(legend.position = 'top')
+perileg <- g_legend(perileg)
+
+inteleg <- ggplot(inte, aes(x = pts, y = pts, colour = Site_Code)) +
+  geom_point() +
+  geom_line() +
+  theme_minimal() + 
+  scale_colour_manual(values = inte$colv, name = 'Interior') +
+  theme(legend.position = 'top')
+inteleg <- g_legend(inteleg)
+
+suisleg <- ggplot(suis, aes(x = pts, y = pts, colour = Site_Code)) +
+  geom_point() +
+  geom_line() +
+  theme_minimal() + 
+  scale_colour_manual(values = suis$colv, name = 'Suisun') +
+  theme(legend.position = 'top')
+suisleg <- g_legend(suisleg)
 
@@
     arrangeGrob(pB, left = textGrob(ylab2, rot = 90)), 
-    pleg, ncol = 1
+    arrangeGrob(perileg, inteleg, suisleg, ncol = 1)
   ), 

### 2017-10-09 sftrends_manu.Rnw
prddat fig is good now

--- 2017-10-06 sftrends_manu.Rnw
@@
 
-<<prddat, fig.cap = 'Model predictions.', fig.height = 8, fig.width = 8>>=
-library(tidyverse)
-library(grid)
-source('R/funcs.R')
-
+<<prddat, fig.cap = 'Model predictions for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \\cref{fig:delt_map}).
+Results are annually-averaged for each water year from October to September.
+Points are model predictions and lines are flow-normalized predictions.', fig.height = 8, fig.width = 8>>=
 data(mods)
+data(delt_dat)
+
+cols <-  brewer.pal(9, 'Blues')[c(4:9)] %>% 
+  colorRampPalette(.)
+
+stat_meta <- delt_dat %>% 
+  dplyr::select(Site_Code, Location) %>% 
+  unique %>% 
+  mutate(
+    Location = factor(Location, 
+           levels = c('Delta', 'Middle', 'Suisun'), 
+           labels = c('Peripheral', 'Interior', 'Suisun')
+    )
+  ) %>% 
+  arrange(Location, Site_Code) %>% 
+  mutate(
+    colv = c(cols(4), cols(3), cols(3)), 
+    pts = c(1:10), 
+    shps = c(c(16, 17, 18, 15), c(16:18), c(16:18))
+  )
 
@@
   mutate(
-    anns = map(data, annual_agg)
-  ) %>% 
+    anns = map(data, annual_agg), 
+    Location = factor(Location, 
+                      levels = c('Delta', 'Middle', 'Suisun'), 
+                      labels = c('Peripheral', 'Interior', 'Suisun')
+                      )
+  ) %>% 
+  left_join(stat_meta, by = c('Site_Code', 'Location')) %>% 
   select(-data, -flovar) %>% 
@@
 
-      p <- ggplot(x, aes(x = date, y = bt_fits, group = Site_Code, colour = Site_Code)) +
-        geom_point(alpha = 0.7) + 
-        geom_line(aes(y = bt_norm)) +
+      p <- ggplot(x, aes(x = date, y = bt_fits, group = Site_Code)) +
+        geom_point(alpha = 0.9, col = x$colv, pch = x$shps) + 
+        geom_line(aes(y = bt_norm), col = x$colv) +
         facet_wrap(~ Location, ncol = 1, scales = 'free_y') + 
@@
 
-pleg <- ggplot(toplo$data[[1]], aes(x = date, y = bt_fits, group = Site_Code, colour = Site_Code)) +
-  geom_point(alpha = 0.7) + 
-  geom_line(aes(y = bt_norm)) +
+peri <- stat_meta %>% 
+  filter(Location == 'Peripheral')
+inte <- stat_meta %>% 
+  filter(Location == 'Interior') 
+suis <- stat_meta %>% 
+  filter(Location == 'Suisun')
+
+perileg <- ggplot(peri, aes(x = pts, y = pts, colour = Site_Code, shape = Site_Code)) +
+  geom_point() +
+  geom_line() +
   theme_minimal() + 
-  guides(colour = guide_legend(ncol = 2))
-pleg <- g_legend(pleg)
+  scale_colour_manual(values = peri$colv, name = 'Peripheral') +
+  scale_shape_manual(values = peri$shps, name = 'Peripheral') +
+  theme(legend.position = 'top')
+perileg <- g_legend(perileg)
+
+inteleg <- ggplot(inte, aes(x = pts, y = pts, colour = Site_Code, shape = Site_Code)) +
+  geom_point() +
+  geom_line() +
+  theme_minimal() + 
+  scale_colour_manual(values = inte$colv, name = 'Interior') +
+  scale_shape_manual(values = inte$shps, name = 'Interior') +
+  theme(legend.position = 'top')
+inteleg <- g_legend(inteleg)
+
+suisleg <- ggplot(suis, aes(x = pts, y = pts, colour = Site_Code, shape = Site_Code)) +
+  geom_point() +
+  geom_line() +
+  theme_minimal() + 
+  scale_colour_manual(values = suis$colv, name = 'Suisun') +
+  scale_shape_manual(values = suis$shps, name = 'Suisun') +
+  theme(legend.position = 'top')
+suisleg <- g_legend(suisleg)
 
@@
 
-# (2, 3), (3, 3)
 pA <- ggplot_gtable(ggplot_build(toplo$plts[[1]]))
@@
     arrangeGrob(pB, left = textGrob(ylab2, rot = 90)), 
-    pleg, ncol = 1
+    arrangeGrob(perileg, inteleg, suisleg, ncol = 1, top = textGrob('\nStations'))
   ), 
@@
 # (2, 3), (3, 3)
-pE <- ggplot_gtable <- (ggplot_build(p5))
+pE <- ggplot_gtable(ggplot_build(p5))
 pH   <- ggplot_gtable(ggplot_build(p8))

### 2017-10-09 sftrends_manu.tex
a pointless figure
prddat fig is good now

@@
   \vspace{1in}
-  \\ Version Date:   Tue Oct 3 12:05:52 2017 -0700
+  \\ Version Date:   Fri Oct 6 15:33:58 2017 -0700
 	}
@@
 
-\caption[Model predictions]{Model predictions.}\label{fig:prddat}
+\caption{Model predictions for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \cref{fig:delt_map}).
+Results are annually-averaged for each water year from October to September.
+Points are model predictions and lines are flow-normalized predictions.}\label{fig:prddat}
 \end{figure}
@@
 
-\begin{kframe}
+\begin{figure}[!ht]
 
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in gList(structure(list(data = list(structure(list(colour = c("{}\#66C2A5"{}, : only 'grobs' allowed in "{}gList"{}}}\end{kframe}
+{\centering \includegraphics[width=0.9\textwidth]{figs/clmchl-1}
+
+}
+
+\caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.
+Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and reduction of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).
+A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \ac{wrtds} results.
+Reductions in \ac{chla} concentration were coincident with an increase in SiO$_2$ and \ac{din} concentrations (d, e), with the greatest increases in August (g, h).
+A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (i).
+Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (j, observations from June).}\label{fig:clmchl}
+\end{figure}
 

### 2017-10-09 sftrends_manu.Rnw
added text about trends in prddat fig

@@
 
-\subsection{Observed Data}
+\subsection{Observed and modelled Data}
 
@@
 
-\subsection{Trends}
+Long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).
+Although each station varied in the overall concentrations, patterns within the three Deltra regions (peripheral, interior, and Suisun) were observed.
+Concentrations for all nitrogen analytes were highest in the peripheral stations.
+Ammonium at P8 and C3 showed a consistent increase over time, followed by a reduction beggining in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.
+By contrast, \ac{din} and nitrite/nitrate concentrations at P8 and C10 showed increases followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did show any noticeable trends.
+Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for C3.
+Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.
+These trends were similar for the Suisun stations, although the reduction in the late 1990s was less pronounced.
+By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.
+
+\subsection{Trend tests}
 

### 2017-10-09 sftrends_manu.tex
added text about trends in prddat fig

@@
   \vspace{1in}
-  \\ Version Date:   Fri Oct 6 15:33:58 2017 -0700
+  \\ Version Date:   Mon Oct 9 10:06:49 2017 -0700
 	}
@@
 For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).
+
+Long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).
+Although each station varied in the overall concentrations, patterns within the three Deltra regions (peripheral, interior, and Suisun) were observed.
+Concentrations for all nitrogen analytes were highest in the peripheral stations.
+Ammonium at P8 and C3 showed a consistent increase over time, followed by a reduction beggining in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.
+By contrast, \ac{din} and nitrite/nitrate concentrations at P8 and C10 showed increases followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did show any noticeable trends.
+Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for C3.
+Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.
+These trends were similar for the Suisun stations, although the reduction in the late 1990s was less pronounced.
+By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.
 

### 2017-10-09 sftrends_manu.Rnw
fixed table formats - one star for sig, no bold-italic

@@
       paste0('(', perchg, ')' ), 
-      paste0('(\\textit{\\textbf{', perchg, '}})')
+      paste0('(\\textbf{', perchg, '})')
       ), 
@@
 cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations'
-foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-*$p<0.05$; **$p<0.005$'
+*$p<0.05$'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
 cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995'
-foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$; **$p<0.005$'
+*$p<0.05$'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
 cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013'
-foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$; **$p<0.005$'
+*$p<0.05$'
 vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
       perchg, 
-      paste0('\\textit{\\textbf{', perchg, '}}')
+      paste0('\\textbf{', perchg, '}')
       ), 
@@
 foot.val<-'\\footnotesize Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, \\cref{fig:stock}).
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Summaries are  medians and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
@@
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$; **$p<0.005$'
+*$p<0.05$'
 

### 2017-10-09 sftrends_manu.tex
fixed table formats - one star for sig, no bold-italic

@@
   \vspace{1in}
-  \\ Version Date:   Mon Oct 9 10:06:49 2017 -0700
+  \\ Version Date:   Mon Oct 9 10:41:34 2017 -0700
 	}
@@
 
-\subsection{Trends}
+\subsection{Trend tests}
 
@@
 {\bfseries DIN}&&\tabularnewline
-~~C10&1.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.4 \footnotesize{(-3.1)**}\tabularnewline
-~~C3&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.5 \footnotesize{(-0.1)**}\tabularnewline
-~~D19&0.4 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-1.2)**}\tabularnewline
-~~D28&0.4 \footnotesize{(\textit{\textbf{0.1}})**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
-~~D6&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.5 \footnotesize{(-0.3)**}\tabularnewline
-~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}\tabularnewline
-~~MD10&0.4 \footnotesize{(-1.1)**}&0.3 \footnotesize{(-2.4)**}\tabularnewline
-~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}\tabularnewline
+~~C10&1.3 \footnotesize{(\textbf{0.8})*}&1.4 \footnotesize{(-3.1)*}\tabularnewline
+~~C3&0.3 \footnotesize{(\textbf{2.2})*}&0.5 \footnotesize{(-0.1)*}\tabularnewline
+~~D19&0.4 \footnotesize{(\textbf{0.2})*}&0.4 \footnotesize{(-1.9)*}\tabularnewline
+~~D26&0.4 \footnotesize{(\textbf{0.4})*}&0.5 \footnotesize{(-1.2)*}\tabularnewline
+~~D28&0.4 \footnotesize{(\textbf{0.1})*}&0.4 \footnotesize{(-3.1)*}\tabularnewline
+~~D4&0.3 \footnotesize{(\textbf{0.6})*}&0.4 \footnotesize{(-0.3)*}\tabularnewline
+~~D6&0.4 \footnotesize{(\textbf{1.8})*}&0.5 \footnotesize{(-0.3)*}\tabularnewline
+~~D7&0.4 \footnotesize{(\textbf{1.7})*}&0.5 \footnotesize{(-0.7)*}\tabularnewline
+~~MD10&0.4 \footnotesize{(-1.1)*}&0.3 \footnotesize{(-2.4)*}\tabularnewline
+~~P8&1.3 \footnotesize{(\textbf{2.5})*}&1.7 \footnotesize{(-2)*}\tabularnewline
 \hline
 {\bfseries NH$_{4}^{+}$}&&\tabularnewline
-~~C10&0.1 \footnotesize{(-3.4)**}&0 \footnotesize{(-5.2)**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{3.7}})**}&0.3 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
-~~D19&0 \footnotesize{(\textit{\textbf{0.4}})**}&0 \footnotesize{(-1.7)**}\tabularnewline
-~~D26&0.1 \footnotesize{(\textit{\textbf{2.2}})**}&0.1 \footnotesize{(-1.5)**}\tabularnewline
-~~D28&0 \footnotesize{(-1.1)**}&0 \footnotesize{(-1.4)**}\tabularnewline
-~~D4&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
-~~D6&0.1 \footnotesize{(\textit{\textbf{2.4}})**}&0.1 \footnotesize{(-0.5)**}\tabularnewline
-~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}\tabularnewline
-~~MD10&0.1 \footnotesize{(-2.8)**}&0 \footnotesize{(-1.1)**}\tabularnewline
-~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}\tabularnewline
+~~C10&0.1 \footnotesize{(-3.4)*}&0 \footnotesize{(-5.2)*}\tabularnewline
+~~C3&0.2 \footnotesize{(\textbf{3.7})*}&0.3 \footnotesize{(\textbf{0})}\tabularnewline
+~~D19&0 \footnotesize{(\textbf{0.4})*}&0 \footnotesize{(-1.7)*}\tabularnewline
+~~D26&0.1 \footnotesize{(\textbf{2.2})*}&0.1 \footnotesize{(-1.5)*}\tabularnewline
+~~D28&0 \footnotesize{(-1.1)*}&0 \footnotesize{(-1.4)*}\tabularnewline
+~~D4&0 \footnotesize{(\textbf{0.9})*}&0.1 \footnotesize{(\textbf{0})}\tabularnewline
+~~D6&0.1 \footnotesize{(\textbf{2.4})*}&0.1 \footnotesize{(-0.5)*}\tabularnewline
+~~D7&0.1 \footnotesize{(\textbf{1.5})*}&0.1 \footnotesize{(-1.2)*}\tabularnewline
+~~MD10&0.1 \footnotesize{(-2.8)*}&0 \footnotesize{(-1.1)*}\tabularnewline
+~~P8&0.2 \footnotesize{(\textbf{4.9})*}&0.1 \footnotesize{(-10.3)*}\tabularnewline
 \hline
 {\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&\tabularnewline
-~~C10&1.2 \footnotesize{(\textit{\textbf{1.4}})**}&1.4 \footnotesize{(-3)**}\tabularnewline
-~~C3&0.1 \footnotesize{(-0.1)**}&0.2 \footnotesize{(\textit{\textbf{0.7}})**}\tabularnewline
-~~D19&0.4 \footnotesize{(-0.1)**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
-~~D26&0.3 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1.1)**}\tabularnewline
-~~D28&0.4 \footnotesize{(-0.2)**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(-0.4)**}\tabularnewline
-~~D6&0.3 \footnotesize{(\textit{\textbf{1.3}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
-~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}\tabularnewline
-~~MD10&0.4 \footnotesize{(-1)**}&0.3 \footnotesize{(-2.5)**}\tabularnewline
-~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}\tabularnewline
+~~C10&1.2 \footnotesize{(\textbf{1.4})*}&1.4 \footnotesize{(-3)*}\tabularnewline
+~~C3&0.1 \footnotesize{(-0.1)*}&0.2 \footnotesize{(\textbf{0.7})*}\tabularnewline
+~~D19&0.4 \footnotesize{(-0.1)*}&0.4 \footnotesize{(-1.9)*}\tabularnewline
+~~D26&0.3 \footnotesize{(\textbf{0})}&0.4 \footnotesize{(-1.1)*}\tabularnewline
+~~D28&0.4 \footnotesize{(-0.2)*}&0.4 \footnotesize{(-3.1)*}\tabularnewline
+~~D4&0.3 \footnotesize{(\textbf{0.7})*}&0.3 \footnotesize{(-0.4)*}\tabularnewline
+~~D6&0.3 \footnotesize{(\textbf{1.3})*}&0.4 \footnotesize{(-0.3)*}\tabularnewline
+~~D7&0.4 \footnotesize{(\textbf{0.7})*}&0.4 \footnotesize{(-0.7)*}\tabularnewline
+~~MD10&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(-2.5)*}\tabularnewline
+~~P8&1.2 \footnotesize{(\textbf{1.7})*}&1.5 \footnotesize{(-0.6)*}\tabularnewline
 \hline
 \end{tabular}\end{center}
-\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-*$p<0.05$; **$p<0.005$\end{table}
+*$p<0.05$\end{table}
 
@@
 {\bfseries DIN}&&&&\tabularnewline
-~~C10&1.2 \footnotesize{(\textit{\textbf{1.1}})**}&1.2 \footnotesize{(\textit{\textbf{0.3}})}&1.3 \footnotesize{(\textit{\textbf{0.5}})**}&1.7 \footnotesize{(\textit{\textbf{1.2}})**}\tabularnewline
-~~C3&0.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.3 \footnotesize{(\textit{\textbf{2.3}})**}&0.4 \footnotesize{(\textit{\textbf{2.4}})**}&0.4 \footnotesize{(\textit{\textbf{1.9}})**}\tabularnewline
-~~D19&0.5 \footnotesize{(\textit{\textbf{0.3}})}&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.7 \footnotesize{(-0.2)}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(\textit{\textbf{0.4}})*}&0.4 \footnotesize{(\textit{\textbf{1}})**}&0.6 \footnotesize{(\textit{\textbf{0.3}})}\tabularnewline
-~~D28&0.5 \footnotesize{(\textit{\textbf{0.8}})*}&0.2 \footnotesize{(\textit{\textbf{0.3}})}&0.3 \footnotesize{(\textit{\textbf{0.5}})*}&0.8 \footnotesize{(-0.3)}\tabularnewline
-~~D4&0.4 \footnotesize{(\textit{\textbf{0.2}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.5 \footnotesize{(-0.5)}\tabularnewline
-~~D6&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{4.6}})**}&0.4 \footnotesize{(\textit{\textbf{1.4}})**}&0.5 \footnotesize{(-0.7)*}\tabularnewline
-~~D7&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.6 \footnotesize{(-2.4)**}\tabularnewline
-~~MD10&0.6 \footnotesize{(-0.3)}&0.2 \footnotesize{(-3.6)**}&0.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.3 \footnotesize{(-0.3)*}\tabularnewline
-~~P8&1.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.9 \footnotesize{(\textit{\textbf{2.4}})**}&1.3 \footnotesize{(\textit{\textbf{3.1}})**}&1.9 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
+~~C10&1.2 \footnotesize{(\textbf{1.1})*}&1.2 \footnotesize{(\textbf{0.3})}&1.3 \footnotesize{(\textbf{0.5})*}&1.7 \footnotesize{(\textbf{1.2})*}\tabularnewline
+~~C3&0.3 \footnotesize{(\textbf{2.4})*}&0.3 \footnotesize{(\textbf{2.3})*}&0.4 \footnotesize{(\textbf{2.4})*}&0.4 \footnotesize{(\textbf{1.9})*}\tabularnewline
+~~D19&0.5 \footnotesize{(\textbf{0.3})}&0.2 \footnotesize{(\textbf{0.4})}&0.3 \footnotesize{(\textbf{0.7})*}&0.7 \footnotesize{(-0.2)}\tabularnewline
+~~D26&0.4 \footnotesize{(\textbf{0.7})*}&0.3 \footnotesize{(\textbf{0.4})*}&0.4 \footnotesize{(\textbf{1})*}&0.6 \footnotesize{(\textbf{0.3})}\tabularnewline
+~~D28&0.5 \footnotesize{(\textbf{0.8})*}&0.2 \footnotesize{(\textbf{0.3})}&0.3 \footnotesize{(\textbf{0.5})*}&0.8 \footnotesize{(-0.3)}\tabularnewline
+~~D4&0.4 \footnotesize{(\textbf{0.2})}&0.3 \footnotesize{(\textbf{1.4})*}&0.3 \footnotesize{(\textbf{1.1})*}&0.5 \footnotesize{(-0.5)}\tabularnewline
+~~D6&0.4 \footnotesize{(\textbf{0.4})}&0.3 \footnotesize{(\textbf{4.6})*}&0.4 \footnotesize{(\textbf{1.4})*}&0.5 \footnotesize{(-0.7)*}\tabularnewline
+~~D7&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textbf{4.2})*}&0.4 \footnotesize{(\textbf{1.5})*}&0.6 \footnotesize{(-2.4)*}\tabularnewline
+~~MD10&0.6 \footnotesize{(-0.3)}&0.2 \footnotesize{(-3.6)*}&0.3 \footnotesize{(\textbf{0.8})*}&1.3 \footnotesize{(-0.3)*}\tabularnewline
+~~P8&1.3 \footnotesize{(\textbf{2.4})*}&0.9 \footnotesize{(\textbf{2.4})*}&1.3 \footnotesize{(\textbf{3.1})*}&1.9 \footnotesize{(\textbf{2.1})*}\tabularnewline
 \hline
 {\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
-~~C10&0.1 \footnotesize{(-2.3)**}&0 \footnotesize{(-6.8)**}&0.1 \footnotesize{(-7.1)**}&0.3 \footnotesize{(-1.5)**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.2 \footnotesize{(\textit{\textbf{4}})**}&0.3 \footnotesize{(\textit{\textbf{3.8}})**}&0.2 \footnotesize{(\textit{\textbf{2.9}})**}\tabularnewline
-~~D19&0.1 \footnotesize{(\textit{\textbf{0.4}})*}&0 \footnotesize{(-1.7)**}&0 \footnotesize{(\textit{\textbf{1.2}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}\tabularnewline
-~~D26&0.1 \footnotesize{(\textit{\textbf{1.4}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.1 \footnotesize{(\textit{\textbf{2.3}})**}\tabularnewline
-~~D28&0.1 \footnotesize{(-0.5)}&0 \footnotesize{(-3.7)**}&0 \footnotesize{(-1.6)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
-~~D4&0.1 \footnotesize{(\textit{\textbf{1.7}})**}&0 \footnotesize{(\textit{\textbf{1}})**}&0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textit{\textbf{2}})**}\tabularnewline
-~~D6&0.1 \footnotesize{(\textit{\textbf{2.9}})**}&0.1 \footnotesize{(\textit{\textbf{2.8}})**}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
-~~D7&0.1 \footnotesize{(\textit{\textbf{3.3}})**}&0 \footnotesize{(\textit{\textbf{2}})**}&0.1 \footnotesize{(-2.8)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
-~~MD10&0.1 \footnotesize{(-1.8)**}&0 \footnotesize{(-6.5)**}&0 \footnotesize{(-3.3)**}&0.2 \footnotesize{(\textit{\textbf{0.4}})}\tabularnewline
-~~P8&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.1 \footnotesize{(\textit{\textbf{1.8}})**}&0.2 \footnotesize{(\textit{\textbf{7}})**}&0.6 \footnotesize{(\textit{\textbf{7}})**}\tabularnewline
+~~C10&0.1 \footnotesize{(-2.3)*}&0 \footnotesize{(-6.8)*}&0.1 \footnotesize{(-7.1)*}&0.3 \footnotesize{(-1.5)*}\tabularnewline
+~~C3&0.2 \footnotesize{(\textbf{3.9})*}&0.2 \footnotesize{(\textbf{4})*}&0.3 \footnotesize{(\textbf{3.8})*}&0.2 \footnotesize{(\textbf{2.9})*}\tabularnewline
+~~D19&0.1 \footnotesize{(\textbf{0.4})*}&0 \footnotesize{(-1.7)*}&0 \footnotesize{(\textbf{1.2})*}&0.1 \footnotesize{(\textbf{2.5})*}\tabularnewline
+~~D26&0.1 \footnotesize{(\textbf{1.4})*}&0.1 \footnotesize{(\textbf{2.5})*}&0.1 \footnotesize{(\textbf{3.1})*}&0.1 \footnotesize{(\textbf{2.3})*}\tabularnewline
+~~D28&0.1 \footnotesize{(-0.5)}&0 \footnotesize{(-3.7)*}&0 \footnotesize{(-1.6)*}&0.1 \footnotesize{(\textbf{1.7})*}\tabularnewline
+~~D4&0.1 \footnotesize{(\textbf{1.7})*}&0 \footnotesize{(\textbf{1})*}&0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textbf{2})*}\tabularnewline
+~~D6&0.1 \footnotesize{(\textbf{2.9})*}&0.1 \footnotesize{(\textbf{2.8})*}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(\textbf{2.1})*}\tabularnewline
+~~D7&0.1 \footnotesize{(\textbf{3.3})*}&0 \footnotesize{(\textbf{2})*}&0.1 \footnotesize{(-2.8)*}&0.1 \footnotesize{(\textbf{1.7})*}\tabularnewline
+~~MD10&0.1 \footnotesize{(-1.8)*}&0 \footnotesize{(-6.5)*}&0 \footnotesize{(-3.3)*}&0.2 \footnotesize{(\textbf{0.4})}\tabularnewline
+~~P8&0.2 \footnotesize{(\textbf{3.9})*}&0.1 \footnotesize{(\textbf{1.8})*}&0.2 \footnotesize{(\textbf{7})*}&0.6 \footnotesize{(\textbf{7})*}\tabularnewline
 \hline
 {\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
-~~C10&1.1 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(\textit{\textbf{0.6}})**}&1.2 \footnotesize{(\textit{\textbf{1.3}})**}&1.5 \footnotesize{(\textit{\textbf{1.8}})**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{0.7}})**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
-~~D19&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.2 \footnotesize{(-0.3)}&0.3 \footnotesize{(\textit{\textbf{0.3}})}&0.6 \footnotesize{(-0.9)*}\tabularnewline
-~~D26&0.4 \footnotesize{(\textit{\textbf{0.6}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.3}})*}&0.5 \footnotesize{(-0.3)}\tabularnewline
-~~D28&0.5 \footnotesize{(\textit{\textbf{0.7}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.2}})}&0.7 \footnotesize{(-1)**}\tabularnewline
-~~D4&0.3 \footnotesize{(\textit{\textbf{0.1}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.4 \footnotesize{(-0.8)*}\tabularnewline
-~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.1}})**}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.4 \footnotesize{(-1)**}\tabularnewline
-~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textit{\textbf{3.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.4 \footnotesize{(-3.6)**}\tabularnewline
-~~MD10&0.5 \footnotesize{(-0.2)}&0.2 \footnotesize{(-3.6)**}&0.2 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(-0.5)*}\tabularnewline
-~~P8&1.2 \footnotesize{(\textit{\textbf{2}})**}&0.9 \footnotesize{(\textit{\textbf{2.3}})**}&1.1 \footnotesize{(\textit{\textbf{2}})**}&1.4 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
+~~C10&1.1 \footnotesize{(\textbf{1.5})*}&1.2 \footnotesize{(\textbf{0.6})*}&1.2 \footnotesize{(\textbf{1.3})*}&1.5 \footnotesize{(\textbf{1.8})*}\tabularnewline
+~~C3&0.2 \footnotesize{(\textbf{0.7})*}&0.1 \footnotesize{(-1)*}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textbf{1})*}\tabularnewline
+~~D19&0.4 \footnotesize{(\textbf{0.4})}&0.2 \footnotesize{(-0.3)}&0.3 \footnotesize{(\textbf{0.3})}&0.6 \footnotesize{(-0.9)*}\tabularnewline
+~~D26&0.4 \footnotesize{(\textbf{0.6})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textbf{0.3})*}&0.5 \footnotesize{(-0.3)}\tabularnewline
+~~D28&0.5 \footnotesize{(\textbf{0.7})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textbf{0.2})}&0.7 \footnotesize{(-1)*}\tabularnewline
+~~D4&0.3 \footnotesize{(\textbf{0.1})}&0.3 \footnotesize{(\textbf{1.4})*}&0.3 \footnotesize{(\textbf{1.1})*}&0.4 \footnotesize{(-0.8)*}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textbf{4.1})*}&0.3 \footnotesize{(\textbf{1.4})*}&0.4 \footnotesize{(-1)*}\tabularnewline
+~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textbf{3.4})*}&0.4 \footnotesize{(\textbf{0.4})}&0.4 \footnotesize{(-3.6)*}\tabularnewline
+~~MD10&0.5 \footnotesize{(-0.2)}&0.2 \footnotesize{(-3.6)*}&0.2 \footnotesize{(\textbf{1.5})*}&1.2 \footnotesize{(-0.5)*}\tabularnewline
+~~P8&1.2 \footnotesize{(\textbf{2})*}&0.9 \footnotesize{(\textbf{2.3})*}&1.1 \footnotesize{(\textbf{2})*}&1.4 \footnotesize{(\textbf{1})*}\tabularnewline
 \hline
 \end{tabular}\end{center}
-\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$; **$p<0.005$\end{table}
+*$p<0.05$\end{table}
 
@@
 {\bfseries DIN}&&&&\tabularnewline
-~~C10&1.1 \footnotesize{(-4.1)**}&1.3 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-2)**}&1.7 \footnotesize{(-3.4)**}\tabularnewline
-~~C3&0.5 \footnotesize{(\textit{\textbf{0.5}})}&0.4 \footnotesize{(\textit{\textbf{0.1}})}&0.6 \footnotesize{(-0.2)}&0.5 \footnotesize{(-0.6)**}\tabularnewline
-~~D19&0.5 \footnotesize{(-2.8)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.6)**}&0.7 \footnotesize{(-2.3)**}\tabularnewline
-~~D26&0.5 \footnotesize{(-1.9)**}&0.3 \footnotesize{(-1.7)**}&0.4 \footnotesize{(-1)**}&0.6 \footnotesize{(-0.8)**}\tabularnewline
-~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-4.9)**}&0.2 \footnotesize{(-4.9)**}&0.7 \footnotesize{(-2.1)**}\tabularnewline
-~~D4&0.4 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.9)**}&0.5 \footnotesize{(\textit{\textbf{0.6}})**}\tabularnewline
-~~D6&0.5 \footnotesize{(-0.2)*}&0.5 \footnotesize{(-1)**}&0.5 \footnotesize{(-0.3)*}&0.5 \footnotesize{(-0.1)}\tabularnewline
-~~D7&0.5 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-1.3)**}&0.4 \footnotesize{(-0.4)**}&0.6 \footnotesize{(-0.2)}\tabularnewline
-~~MD10&0.4 \footnotesize{(-2.3)**}&0.2 \footnotesize{(-3.7)**}&0.2 \footnotesize{(-4.4)**}&1 \footnotesize{(-1.8)**}\tabularnewline
-~~P8&1.5 \footnotesize{(-1.9)**}&1.2 \footnotesize{(-3.5)**}&1.8 \footnotesize{(-2.4)**}&2.7 \footnotesize{(-2.2)**}\tabularnewline
+~~C10&1.1 \footnotesize{(-4.1)*}&1.3 \footnotesize{(-3.1)*}&1.6 \footnotesize{(-2)*}&1.7 \footnotesize{(-3.4)*}\tabularnewline
+~~C3&0.5 \footnotesize{(\textbf{0.5})}&0.4 \footnotesize{(\textbf{0.1})}&0.6 \footnotesize{(-0.2)}&0.5 \footnotesize{(-0.6)*}\tabularnewline
+~~D19&0.5 \footnotesize{(-2.8)*}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.6)*}&0.7 \footnotesize{(-2.3)*}\tabularnewline
+~~D26&0.5 \footnotesize{(-1.9)*}&0.3 \footnotesize{(-1.7)*}&0.4 \footnotesize{(-1)*}&0.6 \footnotesize{(-0.8)*}\tabularnewline
+~~D28&0.5 \footnotesize{(-3)*}&0.2 \footnotesize{(-4.9)*}&0.2 \footnotesize{(-4.9)*}&0.7 \footnotesize{(-2.1)*}\tabularnewline
+~~D4&0.4 \footnotesize{(\textbf{0})}&0.4 \footnotesize{(-1)*}&0.4 \footnotesize{(-0.9)*}&0.5 \footnotesize{(\textbf{0.6})*}\tabularnewline
+~~D6&0.5 \footnotesize{(-0.2)*}&0.5 \footnotesize{(-1)*}&0.5 \footnotesize{(-0.3)*}&0.5 \footnotesize{(-0.1)}\tabularnewline
+~~D7&0.5 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-1.3)*}&0.4 \footnotesize{(-0.4)*}&0.6 \footnotesize{(-0.2)}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.3)*}&0.2 \footnotesize{(-3.7)*}&0.2 \footnotesize{(-4.4)*}&1 \footnotesize{(-1.8)*}\tabularnewline
+~~P8&1.5 \footnotesize{(-1.9)*}&1.2 \footnotesize{(-3.5)*}&1.8 \footnotesize{(-2.4)*}&2.7 \footnotesize{(-2.2)*}\tabularnewline
 \hline
 {\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
-~~C10&0 \footnotesize{(-4.2)**}&0 \footnotesize{(-6.1)**}&0 \footnotesize{(-8.5)**}&0.1 \footnotesize{(-7.3)**}\tabularnewline
-~~C3&0.3 \footnotesize{(\textit{\textbf{1}})**}&0.3 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-0.1)}\tabularnewline
-~~D19&0 \footnotesize{(-1.9)**}&0 \footnotesize{(-0.4)}&0 \footnotesize{(-2.2)**}&0.1 \footnotesize{(-1.8)**}\tabularnewline
-~~D26&0.1 \footnotesize{(-1.2)**}&0.1 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-1.9)**}&0.1 \footnotesize{(-1.4)**}\tabularnewline
-~~D28&0 \footnotesize{(-1.7)**}&0 \footnotesize{(-0.2)}&0 \footnotesize{(-2.4)**}&0.1 \footnotesize{(-3.1)**}\tabularnewline
-~~D4&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0 \footnotesize{(-1.3)**}&0.1 \footnotesize{(-0.3)}&0.1 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
-~~D6&0.1 \footnotesize{(-0.7)**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(\textit{\textbf{0.3}})}&0.1 \footnotesize{(-0.3)**}\tabularnewline
-~~D7&0.1 \footnotesize{(-2.2)**}&0 \footnotesize{(-2.1)**}&0.1 \footnotesize{(\textit{\textbf{1.3}})**}&0.1 \footnotesize{(-0.4)*}\tabularnewline
-~~MD10&0 \footnotesize{(-1.4)*}&0 \footnotesize{(-0.1)}&0 \footnotesize{(-0.8)**}&0.1 \footnotesize{(-4.3)**}\tabularnewline
-~~P8&0.2 \footnotesize{(-8.7)**}&0.1 \footnotesize{(-6.3)**}&0.2 \footnotesize{(-10.4)**}&0.5 \footnotesize{(-13.1)**}\tabularnewline
+~~C10&0 \footnotesize{(-4.2)*}&0 \footnotesize{(-6.1)*}&0 \footnotesize{(-8.5)*}&0.1 \footnotesize{(-7.3)*}\tabularnewline
+~~C3&0.3 \footnotesize{(\textbf{1})*}&0.3 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-0.1)}\tabularnewline
+~~D19&0 \footnotesize{(-1.9)*}&0 \footnotesize{(-0.4)}&0 \footnotesize{(-2.2)*}&0.1 \footnotesize{(-1.8)*}\tabularnewline
+~~D26&0.1 \footnotesize{(-1.2)*}&0.1 \footnotesize{(-1.3)*}&0.1 \footnotesize{(-1.9)*}&0.1 \footnotesize{(-1.4)*}\tabularnewline
+~~D28&0 \footnotesize{(-1.7)*}&0 \footnotesize{(-0.2)}&0 \footnotesize{(-2.4)*}&0.1 \footnotesize{(-3.1)*}\tabularnewline
+~~D4&0.1 \footnotesize{(\textbf{0.3})}&0 \footnotesize{(-1.3)*}&0.1 \footnotesize{(-0.3)}&0.1 \footnotesize{(\textbf{1})*}\tabularnewline
+~~D6&0.1 \footnotesize{(-0.7)*}&0.1 \footnotesize{(-1)*}&0.1 \footnotesize{(\textbf{0.3})}&0.1 \footnotesize{(-0.3)*}\tabularnewline
+~~D7&0.1 \footnotesize{(-2.2)*}&0 \footnotesize{(-2.1)*}&0.1 \footnotesize{(\textbf{1.3})*}&0.1 \footnotesize{(-0.4)*}\tabularnewline
+~~MD10&0 \footnotesize{(-1.4)*}&0 \footnotesize{(-0.1)}&0 \footnotesize{(-0.8)*}&0.1 \footnotesize{(-4.3)*}\tabularnewline
+~~P8&0.2 \footnotesize{(-8.7)*}&0.1 \footnotesize{(-6.3)*}&0.2 \footnotesize{(-10.4)*}&0.5 \footnotesize{(-13.1)*}\tabularnewline
 \hline
 {\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
-~~C10&1.1 \footnotesize{(-4.2)**}&1.2 \footnotesize{(-3.2)**}&1.6 \footnotesize{(-1.9)**}&1.6 \footnotesize{(-3.3)**}\tabularnewline
-~~C3&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.2 \footnotesize{(\textit{\textbf{1.7}})**}&0.2 \footnotesize{(-0.4)}\tabularnewline
-~~D19&0.4 \footnotesize{(-2.9)**}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.5)**}&0.6 \footnotesize{(-2.2)**}\tabularnewline
-~~D26&0.4 \footnotesize{(-1.9)**}&0.2 \footnotesize{(-1.6)**}&0.3 \footnotesize{(-0.6)*}&0.5 \footnotesize{(-0.6)**}\tabularnewline
-~~D28&0.5 \footnotesize{(-3)**}&0.2 \footnotesize{(-5.4)**}&0.2 \footnotesize{(-5.2)**}&0.7 \footnotesize{(-1.7)**}\tabularnewline
-~~D4&0.3 \footnotesize{(-0.1)}&0.3 \footnotesize{(-1)**}&0.3 \footnotesize{(-1)**}&0.4 \footnotesize{(\textit{\textbf{0.4}})**}\tabularnewline
-~~D6&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-1)**}&0.4 \footnotesize{(-0.4)*}&0.4 \footnotesize{(-0.1)}\tabularnewline
-~~D7&0.4 \footnotesize{(-0.6)**}&0.4 \footnotesize{(-1.2)**}&0.4 \footnotesize{(-0.8)**}&0.4 \footnotesize{(-0.3)*}\tabularnewline
-~~MD10&0.4 \footnotesize{(-2.6)**}&0.1 \footnotesize{(-4.5)**}&0.2 \footnotesize{(-5.4)**}&1 \footnotesize{(-1.4)**}\tabularnewline
-~~P8&1.3 \footnotesize{(-1.1)**}&1.1 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
+~~C10&1.1 \footnotesize{(-4.2)*}&1.2 \footnotesize{(-3.2)*}&1.6 \footnotesize{(-1.9)*}&1.6 \footnotesize{(-3.3)*}\tabularnewline
+~~C3&0.2 \footnotesize{(\textbf{0.4})}&0.1 \footnotesize{(\textbf{3.1})*}&0.2 \footnotesize{(\textbf{1.7})*}&0.2 \footnotesize{(-0.4)}\tabularnewline
+~~D19&0.4 \footnotesize{(-2.9)*}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.5)*}&0.6 \footnotesize{(-2.2)*}\tabularnewline
+~~D26&0.4 \footnotesize{(-1.9)*}&0.2 \footnotesize{(-1.6)*}&0.3 \footnotesize{(-0.6)*}&0.5 \footnotesize{(-0.6)*}\tabularnewline
+~~D28&0.5 \footnotesize{(-3)*}&0.2 \footnotesize{(-5.4)*}&0.2 \footnotesize{(-5.2)*}&0.7 \footnotesize{(-1.7)*}\tabularnewline
+~~D4&0.3 \footnotesize{(-0.1)}&0.3 \footnotesize{(-1)*}&0.3 \footnotesize{(-1)*}&0.4 \footnotesize{(\textbf{0.4})*}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-1)*}&0.4 \footnotesize{(-0.4)*}&0.4 \footnotesize{(-0.1)}\tabularnewline
+~~D7&0.4 \footnotesize{(-0.6)*}&0.4 \footnotesize{(-1.2)*}&0.4 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.3)*}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.6)*}&0.1 \footnotesize{(-4.5)*}&0.2 \footnotesize{(-5.4)*}&1 \footnotesize{(-1.4)*}\tabularnewline
+~~P8&1.3 \footnotesize{(-1.1)*}&1.1 \footnotesize{(-3.1)*}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textbf{0})}\tabularnewline
 \hline
 \end{tabular}\end{center}
-\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic).
+\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$; **$p<0.005$\end{table}
+*$p<0.05$\end{table}
 
@@
 {\bfseries Annual}&&&&&\tabularnewline
-~~1976-2006&1.3&\textit{\textbf{2}}**&&0.2&\textit{\textbf{2.8}}**\tabularnewline
-~~2007-2013&1.4&-1.9**&&0.1&-16.6**\tabularnewline
+~~1976-2006&1.3&\textbf{2}*&&0.2&\textbf{2.8}*\tabularnewline
+~~2007-2013&1.4&-1.9*&&0.1&-16.6*\tabularnewline
 \hline
 {\bfseries Seasonal, pre}&&&&&\tabularnewline
-~~Spring&1.2&\textit{\textbf{1.6}}**&&0.2&\textit{\textbf{1.4}}**\tabularnewline
-~~Summer&1&\textit{\textbf{2.4}}**&&0.1&\textit{\textbf{3.3}}**\tabularnewline
-~~Fall&1.3&\textit{\textbf{2.2}}**&&0.2&\textit{\textbf{4.9}}**\tabularnewline
-~~Winter&1.5&\textit{\textbf{2.1}}**&&0.7&\textit{\textbf{4.8}}**\tabularnewline
+~~Spring&1.2&\textbf{1.6}*&&0.2&\textbf{1.4}*\tabularnewline
+~~Summer&1&\textbf{2.4}*&&0.1&\textbf{3.3}*\tabularnewline
+~~Fall&1.3&\textbf{2.2}*&&0.2&\textbf{4.9}*\tabularnewline
+~~Winter&1.5&\textbf{2.1}*&&0.7&\textbf{4.8}*\tabularnewline
 \hline
 {\bfseries Seasonal, post}&&&&&\tabularnewline
-~~Spring&1.3&-1.6**&&0.1&-16.2**\tabularnewline
-~~Summer&0.9&-4.3**&&0.1&-15.7**\tabularnewline
-~~Fall&1.5&-1.7**&&0.1&-19.3**\tabularnewline
-~~Winter&2.2&-0.8**&&0.2&-26.7**\tabularnewline
+~~Spring&1.3&-1.6*&&0.1&-16.2*\tabularnewline
+~~Summer&0.9&-4.3*&&0.1&-15.7*\tabularnewline
+~~Fall&1.5&-1.7*&&0.1&-19.3*\tabularnewline
+~~Winter&2.2&-0.8*&&0.2&-26.7*\tabularnewline
 \hline
@@
 \footnotesize Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
-Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).
+Summaries are  medians and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
@@
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$; **$p<0.005$\end{table}
+*$p<0.05$\end{table}
 

### 2017-10-09 R/funcs.R

@@ -29,8 +29,8 @@ p_form <- function(x, tex_out = F){
 # another stupid function for formatting p-values as asterisks
 p_ast <- function(x){
   
-  sig_cats <- c('**', '*', '')
-  sig_vals <- c(-Inf, 0.005, 0.05, Inf)
+  sig_cats <- c('*', '')
+  sig_vals <- c(-Inf, 0.05, Inf)
   
   out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
   out <- as.character(out)

### 2017-10-10 sftrends_manu.Rnw
results minor revision, ready to bulk up discussion

@@
 Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
-Rather than fitting a global model to the entire time series, one regression is fit at every point in the time series.
-Observations within a window used to fit each regression are weighted relative to annual, seasonal, and flow distances from the center of the window.
+Rather than fitting a global model to the entire time series, one regression is fit to every observation.
+Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.
 Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.
-This approach allows for a type of smoothing where the observed fit is specific to each point in the time series.
+This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows.
 Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.
@@
 All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.
+The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.
 
@@
 Concentrations for all nitrogen analytes were highest in the peripheral stations.
-Ammonium at P8 and C3 showed a consistent increase over time, followed by a reduction beggining in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.
-By contrast, \ac{din} and nitrite/nitrate concentrations at P8 and C10 showed increases followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did show any noticeable trends.
-Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for C3.
+Ammonium concentrations at P8 and C3 were highest and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.
+By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.
+Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.
 Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.
-These trends were similar for the Suisun stations, although the reduction in the late 1990s was less pronounced.
+These trends were similar for the Suisun stations, although the reduction in the late 1990s did not occur.
 By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.
@@
 As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.
-Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.
-Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium, \cref{tab:p8chg}).
+Interestingly, nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).
+Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium).
 Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%).
@@
 \end{inparaenum}
-Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.
+Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion and the relationship of phytoplankton with clam abundance was mediated by flow and climatic variation in recent years.
 
@@
 A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).
-An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).
-These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
-This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
+An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}d) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).
 Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.
-Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
-amurensis}.
 
@@
 The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series.
-In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
 Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow.
@@
 \section{Discussion}
+
+% overall decreasing trends in nitrogen concentrations, but more subtle variation depending on sites - and by season
+%Peripheral high, but opposing trends at some stations - P8 highest for both ammonium and nitrite/nitrage but C3 also high in ammonium but not nitrite/nitrate, C10 low in ammonium but high in nitrite/nitrate
+
+% Suisun increases in ammonium
 
@@
 Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
+% what about seasonal changes in the flow-normalized results? Does this demonstrate effects of normal seasonal rpecip on trend intepretations?
 
@@
 % Suisun conclusions
+These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
+Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
+amurensis}.
+
+In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
+
 As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.

### 2017-10-10 sftrends_manu.tex
results minor revision, ready to bulk up discussion

@@
   \vspace{1in}
-  \\ Version Date:   Mon Oct 9 10:41:34 2017 -0700
+  \\ Version Date:   Mon Oct 9 10:48:31 2017 -0700
 	}
@@
 As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.
-Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.
-Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).
+Interestingly, nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).
+Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium).
 Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%).

### 2017-10-10 sftrends_manu.Rnw
working on discussion

@@
 \usepackage{bm}
-% \usepackage{showlabels}
+\usepackage[inline]{showlabels}
 \bibpunct{(}{)}{,}{a}{}{,}
@@
 
+Water quality conditions in the Delta are dynamic and not easily characterized using observed time series.
+Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that would have not been possible with raw observations.
+A general conclusion is that nitrogen concentrations have decreased overall throughout the approximate forty years of observation.
+These results are confirmed visually from \ac{wrtds} results (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1, tab:trndsann}).
+Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis.
+Nutrient concentrations were highest at the peripheral stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento, San Joaquin, Cosumnes, and Mokelumne rivers.
+The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 as a direct consquence of \ac{wwtp} inputs upstream.
+Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento Regional County Sanitation District.
+By contrast, nitrite/nitrate concentrations were highest at C10 as a measure of contributions from the San Joaquin River to the south that drains a predominantly agricultural watershed.
+Although the Sacramento River drains a much larger agricultural watershed, the dominant ammonium signal as compared to nitrite/nitrate at C3 underscores the importance of \ac{wwtp} control for water quality issues downstream.
+
+Differing magnitudes of nitrogen analytes between stations as a function of source type can have an effect on the relationship between flow and nutrients.
+Both \citet{Hirsch10} and \citet{Beck15b} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.
+In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources.
+Increased flow may induce a change in nutrient concentrations with a reduction if flushing occurs or an increase if mobilization occurs.
+By contrast, nutrient loads are relatively chemostatic or invariant with changes in flow if point-sources are the dominant contributor.
+These relationships are modelled particularly well with \ac{wrtds}, which can provide a means of hypothesizing unknown sources or verifying trends in response to management actions.
+As noted above, C10 at the inflow of the San Joaquin River is dominated by nitrite/nitrate consistent with diffuse, agricultural inputs from the watershed.
+A logical expectation is that trends from observed data may vary considerably from trends with modelled results that are flow-normalized.
+Accordingly, trend analysis of nitrate/nitrate by season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).
+This is consistant with an expected effect of flow on raw time series, particularly
+
 % overall decreasing trends in nitrogen concentrations, but more subtle variation depending on sites - and by season
-%Peripheral high, but opposing trends at some stations - P8 highest for both ammonium and nitrite/nitrage but C3 also high in ammonium but not nitrite/nitrate, C10 low in ammonium but high in nitrite/nitrate
+%Peripheral high, but opposing trends at some stations - P8 highest for both ammonium and nitrite/nitrate but C3 also high in ammonium but not nitrite/nitrate, C10 low in ammonium but high in nitrite/nitrate
 

### 2017-10-10 sftrends_manu.tex
working on discussion

@@
   \vspace{1in}
-  \\ Version Date:   Mon Oct 9 10:48:31 2017 -0700
+  \\ Version Date:   Tue Oct 10 10:14:20 2017 -0700
 	}

### 2017-10-12 sftrends_manu.Rnw
susiun paragraph

@@
 Water quality conditions in the Delta are dynamic and not easily characterized using observed time series.
-Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that would have not been possible with raw observations.
+Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that is not possible with raw observations.
 A general conclusion is that nitrogen concentrations have decreased overall throughout the approximate forty years of observation.
-These results are confirmed visually from \ac{wrtds} results (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1, tab:trndsann}).
-Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis.
+These results are confirmed visually from \ac{wrtds} (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1}, \cref{tab:trndsann}).
+Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are of concern independent of overall trends.
 Nutrient concentrations were highest at the peripheral stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento, San Joaquin, Cosumnes, and Mokelumne rivers.
@@
 In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources.
-Increased flow may induce a change in nutrient concentrations with a reduction if flushing occurs or an increase if mobilization occurs.
+Increased flow may induce a change in nutrient concentrations, such that reudction may occur with flushing or an increase may occur through mobilization.
 By contrast, nutrient loads are relatively chemostatic or invariant with changes in flow if point-sources are the dominant contributor.
@@
 A logical expectation is that trends from observed data may vary considerably from trends with modelled results that are flow-normalized.
-Accordingly, trend analysis of nitrate/nitrate by season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).
-This is consistant with an expected effect of flow on raw time series, particularly
-
-% overall decreasing trends in nitrogen concentrations, but more subtle variation depending on sites - and by season
-%Peripheral high, but opposing trends at some stations - P8 highest for both ammonium and nitrite/nitrate but C3 also high in ammonium but not nitrite/nitrate, C10 low in ammonium but high in nitrite/nitrate
-
-% Suisun increases in ammonium
+Accordingly, trend analysis of nitrate/nitrate by year and season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).
+This is consistant with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly developed watersheds.
 
@@
 Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
+
+Long-term trends in nutrient concentations in Suisun Bay have also been the focus of intense study over the last two decades, particularly as it relates to phytoplankton production.
+Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 4 $\mu$mol L$^{-1}$  (0.072 mg L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.
+Energetic costs of ammonium for phytoplankton growth are lower than nitrate such that the former tends to stimulate growth in most cases.
+Although phytoplankton communities in \ac{sfb} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
+In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate as a primary form of nitrogen that stimulates growth.
+Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values for annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000.
+Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium for the period of record.
+However, decrease in recent years?
+This is a spring blooom thing, what would the potential nitrification at stockton?
 % what about seasonal changes in the flow-normalized results? Does this demonstrate effects of normal seasonal rpecip on trend intepretations?

### 2017-10-12 sftrends_manu.tex
susiun paragraph

@@
   \vspace{1in}
-  \\ Version Date:   Tue Oct 10 10:14:20 2017 -0700
+  \\ Version Date:   Tue Oct 10 12:00:57 2017 -0700
 	}

### 2017-10-17 sftrends_manu.Rnw
revised discussion done, need to proof

@@
 Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.
+The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08}, limits on phytoplankton growth from high turbidity, and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.
 However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales \citep[e.g., variation in sea surface temperatures,][]{Cloern07}.
@@
 As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
+Daily flow estimates were matched with the corresponding sample dates for the nutrients.
 Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.
 Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.
+Salinity was not used as a covariate at the interior and peripheral stations becuase initial analyses showed improved performance using flow estimates for the Sacramento and San Joaquin rivers.
 
@@
 
+% n min/max
+<<>>=
+nobs <- map(mods$data, ~ .x$res %>% na.omit %>% length) %>% 
+  unlist
+minobs <- min(nobs)
+minsta <- mods$Site_Code[which.min(nobs)]
+maxobs <- max(nobs)
+maxsta <- mods$Site_Code[which.max(nobs)]
+@
+
 \noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
-Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
+Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).
 Rather than fitting a global model to the entire time series, one regression is fit to every observation.
@@
 
-Water quality conditions in the Delta are dynamic and not easily characterized using observed time series.
+Water quality conditions in the Delta are dynamic and not easily characterized from observed time series.
 Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that is not possible with raw observations.
@@
 These results are confirmed visually from \ac{wrtds} (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1}, \cref{tab:trndsann}).
-Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are of concern independent of overall trends.
+Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are important independent of overall trends.
 Nutrient concentrations were highest at the peripheral stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento, San Joaquin, Cosumnes, and Mokelumne rivers.
@@
 By contrast, nitrite/nitrate concentrations were highest at C10 as a measure of contributions from the San Joaquin River to the south that drains a predominantly agricultural watershed.
-Although the Sacramento River drains a much larger agricultural watershed, the dominant ammonium signal as compared to nitrite/nitrate at C3 underscores the importance of \ac{wwtp} control for water quality issues downstream.
+Although the Sacramento River drains a much larger area, the dominant ammonium signal as compared to nitrite/nitrate at C3 underscores the importance of \ac{wwtp} control for water quality issues downstream.
 
@@
 Both \citet{Hirsch10} and \citet{Beck15b} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.
-In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources.
-Increased flow may induce a change in nutrient concentrations, such that reudction may occur with flushing or an increase may occur through mobilization.
+In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.
+Increased flow may induce a change in nutrient concentrations, such that reduction may occur with flushing or an increase may occur through mobilization.
 By contrast, nutrient loads are relatively chemostatic or invariant with changes in flow if point-sources are the dominant contributor.
@@
 Accordingly, trend analysis of nitrate/nitrate by year and season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).
-This is consistant with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly developed watersheds.
+This is consistant with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly developed watersheds \citep{Wan17}.
 
@@
 
-Long-term trends in nutrient concentations in Suisun Bay have also been the focus of intense study over the last two decades, particularly as it relates to phytoplankton production.
-Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 4 $\mu$mol L$^{-1}$  (0.072 mg L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.
+Long-term trends in nutrient and phytoplankton concentations in Suisun Bay have also been the focus of intense study for many years \citep{Cloern83,Lehman92,Dugdale07,Jassby08,Glibert14}.
+Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 0.072 mg L$^{-1}$ (4 $\mu$mol L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.
 Energetic costs of ammonium for phytoplankton growth are lower than nitrate such that the former tends to stimulate growth in most cases.
-Although phytoplankton communities in \ac{sfb} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
-In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate as a primary form of nitrogen that stimulates growth.
-Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values for annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000.
-Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium for the period of record.
-However, decrease in recent years?
-This is a spring blooom thing, what would the potential nitrification at stockton?
-% what about seasonal changes in the flow-normalized results? Does this demonstrate effects of normal seasonal rpecip on trend intepretations?
-
-% Station selection adnd flow-records? - comment from reviewer 1
-
-% limitations in the method - why we did not look at load?  Also, use of other tracers of flow/diversions - why we did not look at salinity?
-
-% P8 discussion
-A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
+Although phytoplankton communities in \ac{sfe} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
+In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate that stimulates bloom development.
+Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values of annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000 (\cref{fig:prddat}).
+Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium.
+However, a reversal of trends in in recent years may also be occuring in Suisun Bay, as model estimates suggest either relatively constant concetratrations or even a decrease at some stations beginnign around 2000 (e.g., D6, D7, \cref{fig:trndcomp1}).
+Combined with the shift towards a dominant spring peak in chloropyll growth (\cref{fig:clmchl}f), changing nitrogen ratios continue to be a concern for the management of production in the upper \ac{sfe}.
+
+\subsection{Interpretation of case studies}
+
+A general conclusion from the analysis of nitrogen trends at P8 is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
 These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
@@
 As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
-
-% Suisun conclusions
-These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+Additional research could investigathese hypotheses to better describe mechanisms of change as a basis for more informed management.
+
+In addition to the above comments, our results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, our
+results suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
 This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
-Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
+Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
 amurensis}.
-
-In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
-
+Flow effects on phytoplankton production have also changed over time.
+In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay (\cref{fig:clmchl}j).
+Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).
 As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
-These relationships have been suggested by others \citep{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
+These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
+
+\subsection{Conclusions}
 

### 2017-10-17 sftrends_manu.tex
revised discussion done, need to proof

@@
   \vspace{1in}
-  \\ Version Date:   Tue Oct 10 12:00:57 2017 -0700
+  \\ Version Date:   Thu Oct 12 12:24:31 2017 -0700
 	}
@@
 
+% n min/max
+
 \noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
-Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.
+Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from 433 at D19 to 571 at C3).
 Rather than fitting a global model to the entire time series, one regression is fit to every observation.
@@
 
-Long-term trends in nutrient concentations in Suisun Bay have also been the focus of intense study over the last two decades, particularly as it relates to phytoplankton production.
-Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 4 $\mu$mol L$^{-1}$  (0.072 mg L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.
+Long-term trends in nutrient and phytoplankton concentations in Suisun Bay have also been the focus of intense study for many years \citep{Cloern83,Lehman92,Dugdale07,Jassby08,Glibert14}.
+Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 0.072 mg L$^{-1}$ (4 $\mu$mol L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.
 Energetic costs of ammonium for phytoplankton growth are lower than nitrate such that the former tends to stimulate growth in most cases.
-Although phytoplankton communities in \ac{sfb} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
-In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate as a primary form of nitrogen that stimulates growth.
-Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values for annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000.
-Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium for the period of record.
-However, decrease in recent years?
-This is a spring blooom thing, what would the potential nitrification at stockton?
-% what about seasonal changes in the flow-normalized results? Does this demonstrate effects of normal seasonal rpecip on trend intepretations?
+Although phytoplankton communities in \ac{sfe} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
+In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate that stimulates bloom development.
+Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values of annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000 (\cref{fig:prddat}).
+Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium.
+However, a reversal of trends in in recent years may also be occuring in Suisun Bay, as model estimates suggest either relatively constant concetratrations or even a decrease at some stations beginnign around 2000 (e.g., D6, D7, \cref{fig:trndcomp1}).
+Combined with the shift towards a dominant spring peak in chloropyll growth (\cref{fig:clmchl}f), changing nitrogen ratios continue to be a concern for the management of production in the upper \ac{sfe}.
 
-% Station selection adnd flow-records? - comment from reviewer 1
+\subsection{Interpretation of case studies}
 
-% limitations in the method - why we did not look at load?  Also, use of other tracers of flow/diversions - why we did not look at salinity?
-
-% P8 discussion
-A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
+A general conclusion from the analysis of nitrogen trends at P8 is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
 These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
@@
 As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
+Additional research could investigathese hypotheses to better describe mechanisms of change as a basis for more informed management.
 
-% Suisun conclusions
-These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+In addition to the above comments, our results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, our
+results suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
 This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
-Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
+Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
 amurensis}.
+Flow effects on phytoplankton production have also changed over time.
+In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay (\cref{fig:clmchl}j).
+Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).
+As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
+These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
 
-In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.
-
-As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
-These relationships have been suggested by others \citep{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
+\subsection{Conclusions}
 

### 2017-10-17 sftrends_manu.Rnw
edits from EST done, need to proof

@@
 \noindent Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.
-Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
+Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
 We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
@@
 These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}.
-Although many other stations are available for the region, the stations were chosen becaase they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}.
+Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}.
 Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.
@@
 As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
-Daily flow estimates were matched with the corresponding sample dates for the nutrients.
+Daily flow estimates were matched with the corresponding sample dates for the nutrient data.
 Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.
 Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.
-Salinity was not used as a covariate at the interior and peripheral stations becuase initial analyses showed improved performance using flow estimates for the Sacramento and San Joaquin rivers.
+Salinity was not used as a covariate at the interior and peripheral stations becuase initial analyses showed improved performance using flow estimates from the Sacramento and San Joaquin rivers.
 
@@
 
-\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
+\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).
 Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).
@@
 Flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping.
-These annual and seasonal groupings were chosen for continuity with similar comparisons in \citet{Jabusch16} and as approximate twenty year midpoints in the time series.
+Annual groupings were chosen as approximate twenty year midpoints in the time series and seasonal groupings were chosen to evaluate inter-annual changes while keeping season constant.
 
@@
 Seasonal Kendall tests were also used to describe trends in the observed data.
-These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.
+These trends were compared with those based on the flow-normalized trends to evaluate potential differences in conclusions caused by flow effects.
 
@@
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower \citep{Novick15}.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when assimilation rates are lower \citep{Novick15}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
@@
 
-Estimated trends from Seasonal Kendall tests on the observed data varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Estimated trends from Seasonal Kendall tests on the raw time series varied considerably between sites and analytes (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).
@@
 \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing).
-P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites (see next section).
+P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites.
 Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
@@
 @
-Relationships between flow and observed water quality are complex and can change significantly through space and time \citep{Hirsch10,Zhang16}.
-These principles have been demonstrated for monitoring data in the Delta region \citep{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
+A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
+For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
@@
 Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
-Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
+Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
 A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station \citep{Novick15}.
@@
 These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
-For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high.
+For example, estimated ammonium concentrations in July were low for all flow levels, which suggests either nitrogen inputs were low in the summer or nitrogen was available but uptake by primary consumers and bacterial processing were high.
 Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.
@@
 We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program.
-We thank Larry Harding for providing comments on an earlier draft.
+We thank Larry Harding, Raphael Mazor, Yongshan Wan, and Elizabeth George for providing comments on an earlier draft.
 This study was reviewed and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory.

### 2017-10-17 sftrends_manu.tex
edits from EST done, need to proof

@@
   \vspace{1in}
-  \\ Version Date:   Thu Oct 12 12:24:31 2017 -0700
+  \\ Version Date:   Mon Oct 16 17:57:03 2017 -0700
 	}
@@
 
-\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).
+\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).
 Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from 433 at D19 to 571 at C3).
@@
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
-However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower \citep{Novick15}.
+However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when assimilation rates are lower \citep{Novick15}.
 By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
@@
 
-Estimated trends from Seasonal Kendall tests on the observed data varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Estimated trends from Seasonal Kendall tests on the raw time series varied considerably between sites and analytes (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).
@@
 Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing).
-P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites (see next section).
+P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites.
 Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).
@@
 
-Relationships between flow and observed water quality are complex and can change significantly through space and time \citep{Hirsch10,Zhang16}.
-These principles have been demonstrated for monitoring data in the Delta region \citep{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.
-As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
+A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
+For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.

### 2017-10-18 sftrends_manu.Rnw
intro edits done from Rafi

@@
 \section*{Abstract}
-\noindent Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.
+\noindent Quantitative descriptions of water quality, physical, and biological characteristics in estuaries is critical for developing an ecological understanding of drivers of change.
+Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.
 Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
@@
 \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions.
-Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and confounded by grazing pressure.
+Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure.
 Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
@@
 
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
-Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the larger Bay \citep{Dugdale07,Cornwell14}.
+Understanding drivers of water quality changes in large estuaries depends on accurate descriptions of source inputs from the watershed.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
+Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the estuary \citep{Dugdale07,Cornwell14}.
 Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories.
-Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Shellenbarger08,Cloern07}.
-As a result, \ac{chla} thresholds to assess and manage levels of concern for phtoplankton biomass in the lower bay have been proposed \citep{Sutula17}.
-Although these changes are linked to drivers at different spatial and temporal scales, inputs from the Delta remain a critical interest for understanding downstream effects.
-
+Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms (e.g., \textit{Pseudo-nitzschia}, \textit{Alexandrium}, \textit{Dinophysis spp.}) have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Cloern07,Shellenbarger08}.
+As a result, \ac{chla} thresholds to assess and manage levels of concern for phytoplankton biomass in the lower estuary have been proposed \citep{Sutula17}.
+Although these changes are linked to drivers at different spatial and temporal scales, describing inputs from the Delta is critical to understand downstream effects.
+
+Rates of primary production in coastal habitats are often defined by nutrient concentrations, although a simple relationship between enrichment and water quality changes can be difficult to determine \citep{Cloern01}.
 Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}.
-In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic and benthic species can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
+In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic fishes and benthic invertebrates can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
 Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
-\citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.
+\citet{Glibert14} attributed recent phytoplankton blooms in Suisun Bay to a drought, during which residence times and nitrification rates increased.
 Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}.
@@
 
-A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep{IEP13}.
-Although these data have been used extensively, a systematic assessment of trends covering the full spatial and temporal coverage of the monitoring dataset has not been made.
-This information could inform both past and current research by providing a cohesive description of system-wide trends in nutrients.
-However, formal methods for trend analysis are required given that long-term changes can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.
+Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change in coastal regions.
+A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.
+Although these data have been used extensively, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated.
+Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
+An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported \citep{Novick15}.
+Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.
+
+Formal methods for trend analysis are required to describe water quality changes that vary by space and time.
 As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years.
@@
 The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}.
-The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
+Although the \ac{wrtds} method has been effectively applied to describe changes in freshwater systems, use in coastal ecosystems has not been as extensive.
+Application of \ac{wrtds} to describe trends in estuaries could reveal new insights given the disproportionate effects of physical drivers, such as flow inputs and tidal exchange, on water quality.
+The effects of biological drivers may also be more apparent because hydrological effects can be removed by \ac{wrtds}.
+As such, application of \ac{wrtds} models for trend analysis could facilitate a broader discusion on the need to focus beyond nutrients to develop integrated plans for water quality management.
+
+The goal of this study was to provide a comprehensive description of nutrient trends in the Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
+This study also supports a general need to more effectively use long-term monitoring data to inform management priorities for water quality in coastal regions.
 The specific objectives were to \begin{inparaenum}[1\upshape)]
@@
 However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales \citep[e.g., variation in sea surface temperatures,][]{Cloern07}.
-
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
-A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}.
-For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.
-Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.
-Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows and water exports directly from the system \citep{Jassby00,Jassby08}.
-Our analysis explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.
 

### 2017-10-18 sftrends_manu.tex
intro edits done from Rafi

@@
   \vspace{1in}
-  \\ Version Date:   Mon Oct 16 17:57:03 2017 -0700
+  \\ Version Date:   Tue Oct 17 15:23:13 2017 -0700
 	}
@@
 
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
-Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the larger Bay \citep{Dugdale07,Cornwell14}.
+Understanding drivers of water quality changes in large estuaries depends on accurate descriptions of source inputs from the watershed.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
+Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the estuary \citep{Dugdale07,Cornwell14}.
 Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories.
-Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Shellenbarger08,Cloern07}.
-As a result, \ac{chla} thresholds to assess and manage levels of concern for phtoplankton biomass in the lower bay have been proposed \citep{Sutula17}.
-Although these changes are linked to drivers at different spatial and temporal scales, inputs from the Delta remain a critical interest for understanding downstream effects.
+Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms (e.g., \textit{Pseudo-nitzschia}, \textit{Alexandrium}, \textit{Dinophysis spp.}) have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Cloern07,Shellenbarger08}.
+As a result, \ac{chla} thresholds to assess and manage levels of concern for phytoplankton biomass in the lower estuary have been proposed \citep{Sutula17}.
+Although these changes are linked to drivers at different spatial and temporal scales, describing inputs from the Delta is critical to understand downstream effects.
 
+Rates of primary production in coastal habitats are often defined by nutrient concentrations, although a simple relationship between enrichment and water quality changes can be difficult to determine \citep{Cloern01}.
 Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}.
-In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic and benthic species can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
+In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic fishes and benthic invertebrates can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
 Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
-\citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.
+\citet{Glibert14} attributed recent phytoplankton blooms in Suisun Bay to a drought, during which residence times and nitrification rates increased.
 Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}.

### 2017-10-19 sftrends_manu.Rnw
added site description table

@@
 
-Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).
+Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.
 Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
@@
 The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset).
-All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).}
+All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).
+See \cref{tab:stdescrp} for station descriptions.}
 \label{fig:delt_map}   
@@
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See Figure 3 for annual comparisons.}
+See \cref{fig:trndcomp1} for annual comparisons.}
 \label{fig:trndcomp2}   
@@
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See Figure 3 for annual comparisons.}
+See \cref{fig:trndcomp1} for annual comparisons.}
 \label{fig:trndcomp3}   
@@
 
+% site descrip
+<<stdescrp>>=
+data(delt_dat)
+
+locs <- delt_dat %>% 
+  select(Location, Site_Code, Latitude, Longitude, nh, no23, din) %>% 
+  gather('var', 'val', nh:din) %>% 
+  group_by(Location, Site_Code, Latitude, Longitude, var) %>% 
+  summarise(
+    len = length(na.omit(val)),
+    val = mean(val, na.rm = TRUE), 
+    val = round(val, 2)
+  ) %>% 
+  ungroup %>% 
+  gather('sum', 'val', val:len) %>% 
+  unite('varsum', var, sum) %>% 
+  spread(varsum, val) %>% 
+  mutate(
+    Location = factor(Location, 
+                      levels = c('Delta', 'Middle', 'Suisun'), 
+                      labels = c('Peripheral', 'Interior', 'Suisun')
+                      ), 
+    Site_Code = factor(Site_Code, 
+                       levels = rev(c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'MD10', 'C10', 'C3'))
+                       ), 
+    Latitude = round(Latitude, 2),
+    Longitude = round(Longitude, 2)
+  ) %>% 
+  arrange(Location, Site_Code)
+
+# table stuff
+cap.val <- 'Monitoring stations in the upper San Francico Estuary used to evaluate nitrogen trends.
+Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station.
+Average units of nitrogen are reported in mg L$^{-1}$.'
+cgrps <- c('', '', 'DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
+rgrps <- locs$Location %>% unique
+stats <- locs$Site_Code
+totab <- locs[, -c(1, 2)]
+names(totab) <- c('Lat', 'Lon', '$n$', 'Ave.', '$n$', 'Ave.','$n$', 'Ave.')
+
+# table
+latex(
+  totab,
+  file = '',
+  rowlabel = 'Station',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = rgrps,
+  n.rgroup = c(4, 3, 3),
+  cgroup = cgrps,
+  n.cgroup = c(1, 1, 2, 2, 2),
+  rowname = stats,
+  label = 'tab:stdescrp'
+  ) 
+@
+
 % overall trends, years

### 2017-10-19 sftrends_manu.tex
new delt_map
added site description table

@@
   \vspace{1in}
-  \\ Version Date:   Tue Oct 17 15:23:13 2017 -0700
+  \\ Version Date:   Wed Oct 18 14:13:31 2017 -0700
 	}
@@
 % tables
+
+% site descrip
+%latex.default(totab, file = "", rowlabel = "Station", caption = cap.val,     caption.loc = "top", rgroup = rgrps, n.rgroup = c(4, 3, 3),     cgroup = cgrps, n.cgroup = c(1, 1, 2, 2, 2), rowname = stats,     label = "tab:stdescrp")%
+\begin{table}[!tbp]
+\caption{Monitoring stations in the upper San Francico Estuary used to evaluate nitrogen trends.
+Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station.
+Average units of nitrogen are reported in mg L$^{-1}$.\label{tab:stdescrp}}
+\begin{center}
+\begin{tabular}{lrcrcrrcrrcrr}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Station}&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries DIN}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}\tabularnewline
+\cline{6-7} \cline{9-10} \cline{12-13}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Lat}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Lon}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Ave.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Ave.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Ave.}\tabularnewline
+\hline
+{\bfseries Peripheral}&&&&&&&&&&&&\tabularnewline
+~~C3&$38.35$&&$-121.55$&&$569$&$0.39$&&$569$&$0.24$&&$571$&$0.15$\tabularnewline
+~~C10&$37.68$&&$-121.27$&&$539$&$1.42$&&$539$&$0.09$&&$558$&$1.34$\tabularnewline
+~~MD10&$38.04$&&$-121.42$&&$548$&$0.48$&&$548$&$0.06$&&$570$&$0.41$\tabularnewline
+~~P8&$37.98$&&$-121.38$&&$556$&$1.58$&&$556$&$0.28$&&$563$&$1.30$\tabularnewline
+\hline
+{\bfseries Interior}&&&&&&&&&&&&\tabularnewline
+~~D19&$38.04$&&$-121.61$&&$433$&$0.41$&&$435$&$0.05$&&$462$&$0.35$\tabularnewline
+~~D26&$38.08$&&$-121.57$&&$556$&$0.42$&&$556$&$0.10$&&$565$&$0.32$\tabularnewline
+~~D28&$37.97$&&$-121.57$&&$529$&$0.44$&&$535$&$0.05$&&$555$&$0.38$\tabularnewline
+\hline
+{\bfseries Suisun}&&&&&&&&&&&&\tabularnewline
+~~D4&$38.06$&&$-121.82$&&$546$&$0.40$&&$546$&$0.07$&&$565$&$0.33$\tabularnewline
+~~D6&$38.04$&&$-122.12$&&$534$&$0.44$&&$534$&$0.09$&&$562$&$0.35$\tabularnewline
+~~D7&$38.12$&&$-122.04$&&$535$&$0.44$&&$535$&$0.07$&&$561$&$0.36$\tabularnewline
+\hline
+\end{tabular}\end{center}
+\end{table}
 

### 2017-10-19 sftrends_manu.Rnw
med chl in table 1

@@
     len = length(na.omit(val)),
-    val = mean(val, na.rm = TRUE), 
+    val = median(val, na.rm = TRUE), 
     val = round(val, 2)
@@
 Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station.
-Average units of nitrogen are reported in mg L$^{-1}$.'
+Median values are reported in mg L$^{-1}$.'
 cgrps <- c('', '', 'DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
@@
 totab <- locs[, -c(1, 2)]
-names(totab) <- c('Lat', 'Lon', '$n$', 'Ave.', '$n$', 'Ave.','$n$', 'Ave.')
+names(totab) <- c('Lat', 'Lon', rep(c('$n$', 'Med.'), 3))
 

### 2017-10-20 sftrends_manu.Rnw
rafi edits done

@@
 Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
-An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported \citep{Novick15}.
+An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported to the estuary \citep{Novick15}.
 Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.
@@
 The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
-Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.
+Less than 3\% of all observations were below the deteciton limit (left-censored), although variation was observed between analytes and location.
 The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).
 
-Daily flow estimates for the Delta region were obtained from the Dayflow software program \citep{IEP16}.
-The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends.
+\ac{wrtds} models require flow data paired with nutrient data.
+At the peripheral and interior stations, daily flow estimates were matched with the corresponding sample dates for the nutrient data.
+Daily flow estimates were obtained from the Dayflow software program \citep{IEP16}.
+The Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
 Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.
 Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
-As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
-Daily flow estimates were matched with the corresponding sample dates for the nutrient data.
-Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.
+Salinity was used as a proxy for flow at sites D4, D6, and D7 where tidal influences were much stronger.
 Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.
-Salinity was not used as a covariate at the interior and peripheral stations becuase initial analyses showed improved performance using flow estimates from the Sacramento and San Joaquin rivers.
+Models were evaluated using salinity at the interior and peripheral stations but performance was reduced relative to models that used daily flow estimates.
 
@@
 Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.
-All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.
+All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data below the detection limit \citep{Tobin58}.
 Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September).
@@
 These trends were compared with those based on the flow-normalized trends to evaluate potential differences in conclusions caused by flow effects.
+
+\subsection{Selected examples}
+
+Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
+The selected case studies focused on \begin{inparaenum}[1\upshape)]
+\item effects of wastewater treatment upgrades upstream of P8, and		
+\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.		
+\end{inparaenum}		
+Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
+
+\subsubsection{Effects of wastewater treatment}
+
+Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.
+Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
+The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
+An increase and decrease of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
+\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
+\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
+\end{inparaenum}
+
+\subsubsection{Effects of biological invasions}
+
+Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
+Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \citep{Feyrer03,MacNally10}.
+Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply \citep{Jassby08,Schoellhamer11,Cloern12b}, although decreases in phytoplankton by clam biofiltration may have also increased clarity \citep{MacNally10}.
+The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}.
+We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
+\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
+\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
+\end{inparaenum}
+Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion and the relationship of phytoplankton with clam abundance was mediated by flow and climatic variation in recent years.
 
@@
 A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
+For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant ($p > 0.05$) with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
@@
 
-Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
-The selected case studies focused on \begin{inparaenum}[1\upshape)]
-\item effects of wastewater treatment upgrades upstream of P8, and		
-\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.		
-\end{inparaenum}		
-Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
-
 \subsubsection{Effects of wastewater treatment}
-
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.
-Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
-A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
-\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
-\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
-\end{inparaenum}
 
@@
 
-Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).
+Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had similar \ac{din} concentrations before and after upgrades, whereas ammonium concentrations were greatly reduced (\cref{fig:stock}).
 Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite.
 As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.
-Interestingly, nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).
+Nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).
 Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium).
@@
 
-Relationships of nitrite/nitrate with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).
+Nitrogen concentrations varied with flow, although relationships depended on season and year.
+Relationships of nitrite/nitrate with flow described by \ac{wrtds} showed flushing or dilution at higher flow (\cref{fig:p8trnds}b).
 Seasonal variation was even more apparent for ammonium, although both nitrite/nitrate and ammonium typically had the highest concentrations at low flow in the winter (January).
@@
 \subsubsection{Effects of biological invasions}
-
-Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
-Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \citep{Feyrer03,MacNally10}.
-Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply \citep{Jassby08,Schoellhamer11,Cloern12b}, although decreases in phytoplankton by clam biofiltration may have also increased clarity \citep{MacNally10}.
-The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}.
-We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
-\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
-\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
-\end{inparaenum}
-Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion and the relationship of phytoplankton with clam abundance was mediated by flow and climatic variation in recent years.
 
@@
 
-The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies \citep{Alpine92,Thompson08}.
+The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass.
 However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.
@@
 % obs v flow norm trends
+Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
 Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
-Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
-A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station \citep{Novick15}.
+An alternative evaluation of flow in the Delta demonstrated that flow contributions from different sources vary considerably over time at each station \citep{Novick15}.
 For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
@@
 
-A general conclusion from the analysis of nitrogen trends at P8 is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
-These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8.
-For example, estimated ammonium concentrations in July were low for all flow levels, which suggests either nitrogen inputs were low in the summer or nitrogen was available but uptake by primary consumers and bacterial processing were high.
+Seasonal timings of water quality improvements and the link to flow changes are difficult to characterize from the observed time series.
+\ac{wrtds} models were used to characterize these changes at P8 and to develop additional hypotheses of factors that influence nutrient concentrations.
+A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.
+Estimated ammonium concentrations in July were low for all flow levels, which suggests either nitrogen inputs were low in the summer or nitrogen was available but uptake by primary consumers and bacterial processing were high.
 Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.
@@
 As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
-Additional research could investigathese hypotheses to better describe mechanisms of change as a basis for more informed management.
-
-In addition to the above comments, our results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, our
-results suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+Additional research could investigate these hypotheses to better describe mechanisms of change as a basis for more informed management.
+
+The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.
+In general, clam biomass was associated with a decrease in \ac{chl} concentration, as shown by others \citep{Alpin92,Thompson08}.
+Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
 This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
@@
 Flow effects on phytoplankton production have also changed over time.
-In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay (\cref{fig:clmchl}j).
+In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the estuary (\cref{fig:clmchl}j).
 Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).
@@
 As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.
-At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.
+\ac{wrtds} models can facilitate descriptions of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.
 We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.
-Moreover, trends in nutrient loads from point sources in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases \citep{Jassby08} and exports to Suisun Bay \citep{Novick14}.
-The results from \ac{wrtds} demonstrating these changes are not unexpected, and consequently, we are not detracting from the potential implications of such increases.
+The results from \ac{wrtds} are also consistent with described trends of nutrient loads from point sources (e.g., Sacramento \ac{wwtp} increases \citep{Jassby08} and exports to Suisun Bay \citep{Novick14}) demonstrating that these changes are not unexpected.
+Consequently, we are not detracting from the potential implications of such increases.
 The important conclusion is that the physical/hydrological and biogeochemical factors that influence nutrient cycling and ambient concentrations in the Bay-Delta, and changes to those factors, are substantial enough that they can be comparable in magnitude to anthropogenic load increases or comparable to the effects of management actions to decrease nutrient levels.
@@
 
-Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
+Combined with additional data, our results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics.
 Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.
-For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow-normalized results.
-Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models, \cite{Beck17}).
-Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
+For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow effects.
+Alternative methods could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models, \cite{Beck17}).
+Overall, quantitative interpretations of multiple factors can provide a basis for identifying links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
 
@@
   rev
-ylab <- expression(paste('DIN (mg ', L^-1, ')'))
+ylab <- expression(paste('log - DIN (mg ', L^-1, ')'))
 
@@
 p1 <- ggplot(toplo, aes(x = Date)) +
-  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
+  geom_area(aes(y = log10(1 + val), group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
   scale_x_date(expand = c(0, 0)) + 
   scale_y_continuous(ylab, expand = c(0, 0)) + 
-  facet_grid(Site_Code~., scales = 'free_y') +
+  facet_grid(Site_Code~.)+ #, scales = 'free_y') +
   scale_fill_manual(values = cols, labels = leglabs) +

### 2017-10-20 sftrends_manu.tex
med chl in table 1
rafi edits done

@@
   \vspace{1in}
-  \\ Version Date:   Wed Oct 18 14:13:31 2017 -0700
+  \\ Version Date:   Thu Oct 19 15:58:58 2017 -0700
 	}
@@
 The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
-Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.
+Less than 3\% of all observations were below the deteciton limit (left-censored), although variation was observed between analytes and location.
 The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).
 
-Daily flow estimates for the Delta region were obtained from the Dayflow software program \citep{IEP16}.
-The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends.
+\ac{wrtds} models require flow data paired with nutrient data.
+At the peripheral and interior stations, daily flow estimates were matched with the corresponding sample dates for the nutrient data.
+Daily flow estimates were obtained from the Dayflow software program \citep{IEP16}.
+The Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
 Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.
 Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
-As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.
-Daily flow estimates were matched with the corresponding sample dates for the nutrient data.
-Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.
+Salinity was used as a proxy for flow at sites D4, D6, and D7 where tidal influences were much stronger.
 Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.
-Salinity was not used as a covariate at the interior and peripheral stations becuase initial analyses showed improved performance using flow estimates from the Sacramento and San Joaquin rivers.
+Models were evaluated using salinity at the interior and peripheral stations but performance was reduced relative to models that used daily flow estimates.
 
@@
 A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends.
-For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
+For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), thirteen comparisons had trends that were insignificant ($p > 0.05$) with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant.
 This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter.
@@
 
-Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.
-The selected case studies focused on \begin{inparaenum}[1\upshape)]
-\item effects of wastewater treatment upgrades upstream of P8, and		
-\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.		
-\end{inparaenum}		
-Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
-
 \subsubsection{Effects of wastewater treatment}
 
-Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.
-Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
-The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
-A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
-\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
-\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
-\end{inparaenum}
-
-Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).
+Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had similar \ac{din} concentrations before and after upgrades, whereas ammonium concentrations were greatly reduced (\cref{fig:stock}).
 Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite.
 As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.
-Interestingly, nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).
+Nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).
 Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium).
@@
 As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.
-Additional research could investigathese hypotheses to better describe mechanisms of change as a basis for more informed management.
+Additional research could investigate these hypotheses to better describe mechanisms of change as a basis for more informed management.
 
-In addition to the above comments, our results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, our
-results suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
+The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.
+In general, clam biomass was associated with a decrease in \ac{chl} concentration, as shown by others \citep{Alpin92,Thompson08}.
+Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
 This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
@@
 Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station.
-Average units of nitrogen are reported in mg L$^{-1}$.\label{tab:stdescrp}}
+Median values are reported in mg L$^{-1}$.\label{tab:stdescrp}}
 \begin{center}
@@
 \cline{6-7} \cline{9-10} \cline{12-13}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Lat}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Lon}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Ave.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Ave.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Ave.}\tabularnewline
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Lat}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Lon}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{$n$}&\multicolumn{1}{c}{Med.}\tabularnewline
 \hline
 {\bfseries Peripheral}&&&&&&&&&&&&\tabularnewline
-~~C3&$38.35$&&$-121.55$&&$569$&$0.39$&&$569$&$0.24$&&$571$&$0.15$\tabularnewline
-~~C10&$37.68$&&$-121.27$&&$539$&$1.42$&&$539$&$0.09$&&$558$&$1.34$\tabularnewline
-~~MD10&$38.04$&&$-121.42$&&$548$&$0.48$&&$548$&$0.06$&&$570$&$0.41$\tabularnewline
-~~P8&$37.98$&&$-121.38$&&$556$&$1.58$&&$556$&$0.28$&&$563$&$1.30$\tabularnewline
+~~C3&$38.35$&&$-121.55$&&$569$&$0.36$&&$569$&$0.22$&&$571$&$0.13$\tabularnewline
+~~C10&$37.68$&&$-121.27$&&$539$&$1.44$&&$539$&$0.04$&&$558$&$1.40$\tabularnewline
+~~MD10&$38.04$&&$-121.42$&&$548$&$0.31$&&$548$&$0.03$&&$570$&$0.28$\tabularnewline
+~~P8&$37.98$&&$-121.38$&&$556$&$1.46$&&$556$&$0.12$&&$563$&$1.20$\tabularnewline
 \hline
 {\bfseries Interior}&&&&&&&&&&&&\tabularnewline
-~~D19&$38.04$&&$-121.61$&&$433$&$0.41$&&$435$&$0.05$&&$462$&$0.35$\tabularnewline
-~~D26&$38.08$&&$-121.57$&&$556$&$0.42$&&$556$&$0.10$&&$565$&$0.32$\tabularnewline
-~~D28&$37.97$&&$-121.57$&&$529$&$0.44$&&$535$&$0.05$&&$555$&$0.38$\tabularnewline
+~~D19&$38.04$&&$-121.61$&&$433$&$0.35$&&$435$&$0.04$&&$462$&$0.31$\tabularnewline
+~~D26&$38.08$&&$-121.57$&&$556$&$0.38$&&$556$&$0.09$&&$565$&$0.29$\tabularnewline
+~~D28&$37.97$&&$-121.57$&&$529$&$0.38$&&$535$&$0.03$&&$555$&$0.33$\tabularnewline
 \hline
 {\bfseries Suisun}&&&&&&&&&&&&\tabularnewline
-~~D4&$38.06$&&$-121.82$&&$546$&$0.40$&&$546$&$0.07$&&$565$&$0.33$\tabularnewline
-~~D6&$38.04$&&$-122.12$&&$534$&$0.44$&&$534$&$0.09$&&$562$&$0.35$\tabularnewline
-~~D7&$38.12$&&$-122.04$&&$535$&$0.44$&&$535$&$0.07$&&$561$&$0.36$\tabularnewline
+~~D4&$38.06$&&$-121.82$&&$546$&$0.38$&&$546$&$0.05$&&$565$&$0.32$\tabularnewline
+~~D6&$38.04$&&$-122.12$&&$534$&$0.45$&&$534$&$0.08$&&$562$&$0.35$\tabularnewline
+~~D7&$38.12$&&$-122.04$&&$535$&$0.44$&&$535$&$0.06$&&$561$&$0.36$\tabularnewline
 \hline

### 2017-10-20 sftrends_manu.Rnw
some tables/figs back to supplement
no footnotes in tables, changed color sheme of clam plot

@@
 
+%for supplemental figures/tables
+\newcommand{\beginsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
+
 %knitr options
@@
 The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.
-In general, clam biomass was associated with a decrease in \ac{chl} concentration, as shown by others \citep{Alpin92,Thompson08}.
+In general, clam biomass was associated with a decrease in \ac{chla} concentration, as shown by others \citep{Alpine92,Thompson08}.
 Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1976-1995.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp2}   
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1996-2013.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp3}   
-\end{figure}
-
 <<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
@@
 p1 <- ggplot(clmdat, aes(x = yr, y = clams_smp, fill = species)) + 
-  geom_bar(stat = 'identity') + 
+  geom_bar(stat = 'identity', colour = 'black') + 
   mythm +
-  scale_fill_brewer(
-    palette = 'Set2',
+  scale_fill_manual(
+    values = c('white', 'lightgrey'),
     labels = parse(text = c('italic(Corbicula)', 'italic(Potamocorbula)'))
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations'
-foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
@@
   rowname = stats,
-  label = 'tab:trndsann',
-  insert.bottom = foot.val
-  )
-
-@
+  label = 'tab:trndsann'
+  )
+
+@
+
+% trends in no23, nh at p8
+<<p8chg>>=
+# uses data from chunk in results test
+
+# combine for trends for table
+totab <- rbind(no_bf, no_af, nh_bf, nh_af) %>% 
+  select(-tau, -slope, -chitest) %>% 
+  mutate(
+    med = round(med, 1), 
+    perchg = round(perchg, 1),
+    perchg = ifelse(
+      perchg < 0, 
+      perchg, 
+      paste0('\\textbf{', perchg, '}')
+      ), 
+    ztest = p_ast(ztest)
+  ) %>% 
+  unite('perchg', perchg, ztest, sep = '') %>% 
+  gather('est', 'val', med:perchg) %>% 
+  unite('col', var, est) %>% 
+  spread(col, val) %>% 
+  unite('period', cat, period) %>% 
+  mutate(period = factor(period, levels = faclevs, labels = faclevs)) %>% 
+  arrange(period) %>% 
+  mutate(cat = c(yrlabs_bf, yrlabs_af, molabs, molabs)) %>% 
+  data.frame(sep = c(rep('Annual', 2), rep('Seasonal, pre', 4), rep('Seasonal, post', 4)), .) %>% 
+  select(sep, cat, no_med, no_perchg, nh_med, nh_perchg)
+
+# final table formatting
+segs <- unique(totab$sep)
+cats <- totab$cat
+totab <- totab[,-c(1:2)]
+vars <- c('NO$_{2}^{-}$/NO$_{3}^{2-}$', 'NH$_{4}^{+}$')
+names(totab) <- rep(c('Median', '\\% change'), 2)
+
+cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8.
+Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, \\cref{fig:stock}).
+Summaries are  medians and percent change per year in parentheses (increasing in bold).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$'
+
+latex( 
+  totab,
+  file = '',
+  rowlabel = 'Period',
+  caption = cap.val,
+  caption.loc = 'top',
+  rgroup = segs,
+  n.rgroup = c(2, rep(4, 2)),
+  cgroup = vars,
+  n.cgroup = c(2, 2),
+  rowname = cats,
+  label = 'tab:p8chg'
+  )
+
+@
+
+\clearpage
+
+\beginsupplement
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1976-1995.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp2}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1996-2013.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp3}   
+\end{figure}
+
+\clearpage
 
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995'
-foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
@@
   rowname = stats,
-  label = 'tab:trndsbef',
-  insert.bottom = foot.val
+  label = 'tab:trndsbef'
   )
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013'
-foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
+cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
@@
   rowname = stats,
-  label = 'tab:trndsaft',
-  insert.bottom = foot.val
-  )
-
-@
-
-% trends in no23, nh at p8
-<<p8chg>>=
-# uses data from chunk in results test
-
-# combine for trends for table
-totab <- rbind(no_bf, no_af, nh_bf, nh_af) %>% 
-  select(-tau, -slope, -chitest) %>% 
-  mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      perchg, 
-      paste0('\\textbf{', perchg, '}')
-      ), 
-    ztest = p_ast(ztest)
-  ) %>% 
-  unite('perchg', perchg, ztest, sep = '') %>% 
-  gather('est', 'val', med:perchg) %>% 
-  unite('col', var, est) %>% 
-  spread(col, val) %>% 
-  unite('period', cat, period) %>% 
-  mutate(period = factor(period, levels = faclevs, labels = faclevs)) %>% 
-  arrange(period) %>% 
-  mutate(cat = c(yrlabs_bf, yrlabs_af, molabs, molabs)) %>% 
-  data.frame(sep = c(rep('Annual', 2), rep('Seasonal, pre', 4), rep('Seasonal, post', 4)), .) %>% 
-  select(sep, cat, no_med, no_perchg, nh_med, nh_perchg)
-
-# final table formatting
-segs <- unique(totab$sep)
-cats <- totab$cat
-totab <- totab[,-c(1:2)]
-vars <- c('NO$_{2}^{-}$/NO$_{3}^{2-}$', 'NH$_{4}^{+}$')
-names(totab) <- rep(c('Median', '\\% change'), 2)
-
-cap.val<-'Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \\ac{wwtp} upgrades upstream of station P8'
-foot.val<-'\\footnotesize Upgrades were completed in 2006 at the City of Stockton \\ac{wwtp} (San Joaquin County, \\cref{fig:stock}).
-Summaries are  medians and percent change per year in parentheses (increasing in bold).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$'
-
-latex( 
-  totab,
-  file = '',
-  rowlabel = 'Period',
-  caption = cap.val,
-  caption.loc = 'top',
-  rgroup = segs,
-  n.rgroup = c(2, rep(4, 2)),
-  cgroup = vars,
-  n.cgroup = c(2, 2),
-  rowname = cats,
-  label = 'tab:p8chg',
-  insert.bottom = foot.val
-  )
-
-@
-
-\clearpage
+  label = 'tab:trndsaft'
+  )
+
+@
 

### 2017-10-20 sftrends_manu.tex
some tables/figs back to supplement
no footnotes in tables, changed color sheme of clam plot

@@
   \vspace{1in}
-  \\ Version Date:   Thu Oct 19 15:58:58 2017 -0700
+  \\ Version Date:   Fri Oct 20 11:14:05 2017 -0700
 	}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1976-1995.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp2}   
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
-Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different seasonal groupings from 1996-2013.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-See \cref{fig:trndcomp1} for annual comparisons.}
-\label{fig:trndcomp3}   
-\end{figure}
-
 \begin{figure}[!ht]
@@
 % overall trends, years
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann", insert.bottom = foot.val)%
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann")%
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations\label{tab:trndsann}} 
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+*$p<0.05$\label{tab:trndsann}}
 \begin{center}
@@
 \end{tabular}\end{center}
-\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
+\end{table}
+
+% trends in no23, nh at p8
+%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg")%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8.
+Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
+Summaries are  medians and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-*$p<0.05$\end{table}
+Increasing values are in bold-italics.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$\label{tab:p8chg}}
+\begin{center}
+\begin{tabular}{lllcll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
+\cline{2-3} \cline{5-6}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
+\hline
+{\bfseries Annual}&&&&&\tabularnewline
+~~1976-2006&1.3&\textbf{2}*&&0.2&\textbf{2.8}*\tabularnewline
+~~2007-2013&1.4&-1.9*&&0.1&-16.6*\tabularnewline
+\hline
+{\bfseries Seasonal, pre}&&&&&\tabularnewline
+~~Spring&1.2&\textbf{1.6}*&&0.2&\textbf{1.4}*\tabularnewline
+~~Summer&1&\textbf{2.4}*&&0.1&\textbf{3.3}*\tabularnewline
+~~Fall&1.3&\textbf{2.2}*&&0.2&\textbf{4.9}*\tabularnewline
+~~Winter&1.5&\textbf{2.1}*&&0.7&\textbf{4.8}*\tabularnewline
+\hline
+{\bfseries Seasonal, post}&&&&&\tabularnewline
+~~Spring&1.3&-1.6*&&0.1&-16.2*\tabularnewline
+~~Summer&0.9&-4.3*&&0.1&-15.7*\tabularnewline
+~~Fall&1.5&-1.7*&&0.1&-19.3*\tabularnewline
+~~Winter&2.2&-0.8*&&0.2&-26.7*\tabularnewline
+\hline
+\end{tabular}\end{center}
+\end{table}
+
+\clearpage
+
+\beginsupplement
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1976-1995.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp2}   
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
+Trends are shown separately for different seasonal groupings from 1996-2013.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+See \cref{fig:trndcomp1} for annual comparisons.}
+\label{fig:trndcomp3}   
+\end{figure}
+
+\clearpage
 
 % overall trends, seasons, first annual periodyears
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef", insert.bottom = foot.val)%
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef")%
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995\label{tab:trndsbef}} 
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
+Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
+Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
+*$p<0.05$\label{tab:trndsbef}}
 \begin{center}
@@
 \end{tabular}\end{center}
-\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
+\end{table}
+
+% overall trends, seasons, first annual periodyears
+%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft")%
+\begin{table}[!tbp]
+\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
+Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
 Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
 Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$\end{table}
-
-% overall trends, seasons, first annual periodyears
-%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013\label{tab:trndsaft}} 
+*$p<0.05$\label{tab:trndsaft}}
 \begin{center}
@@
 \end{tabular}\end{center}
-\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$\end{table}
-
-% trends in no23, nh at p8
-%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg", insert.bottom = foot.val)%
-\begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8\label{tab:p8chg}} 
-\begin{center}
-\begin{tabular}{lllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
-\cline{2-3} \cline{5-6}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
-\hline
-{\bfseries Annual}&&&&&\tabularnewline
-~~1976-2006&1.3&\textbf{2}*&&0.2&\textbf{2.8}*\tabularnewline
-~~2007-2013&1.4&-1.9*&&0.1&-16.6*\tabularnewline
-\hline
-{\bfseries Seasonal, pre}&&&&&\tabularnewline
-~~Spring&1.2&\textbf{1.6}*&&0.2&\textbf{1.4}*\tabularnewline
-~~Summer&1&\textbf{2.4}*&&0.1&\textbf{3.3}*\tabularnewline
-~~Fall&1.3&\textbf{2.2}*&&0.2&\textbf{4.9}*\tabularnewline
-~~Winter&1.5&\textbf{2.1}*&&0.7&\textbf{4.8}*\tabularnewline
-\hline
-{\bfseries Seasonal, post}&&&&&\tabularnewline
-~~Spring&1.3&-1.6*&&0.1&-16.2*\tabularnewline
-~~Summer&0.9&-4.3*&&0.1&-15.7*\tabularnewline
-~~Fall&1.5&-1.7*&&0.1&-19.3*\tabularnewline
-~~Winter&2.2&-0.8*&&0.2&-26.7*\tabularnewline
-\hline
-\end{tabular}\end{center}
-\footnotesize Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).
-Summaries are  medians and percent change per year in parentheses (increasing in bold).
-Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.
-Increasing values are in bold-italics.
-Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.
-*$p<0.05$\end{table}
-
-\clearpage
+\end{table}
 

### 2017-10-27 sftrends_manu.Rnw
minor edits to maybe final draft

@@
 \noindent Quantitative descriptions of water quality, physical, and biological characteristics in estuaries is critical for developing an ecological understanding of drivers of change.
-Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.
+Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \acl{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.
 Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
+Modelled trends demonstrated that nutrient concentrations were on average higher in the last twenty years relative to the earlier periods of observation, although a consistent decline began in the mid-1990s and early 2000s.
 We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
-\ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions.
-Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure.
+\ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \acl{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions.
+Results for Suisun Bay showed that \acl{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure.
 Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
@@
 
-Understanding drivers of water quality changes in large estuaries depends on accurate descriptions of source inputs from the watershed.
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
-Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the estuary \citep{Dugdale07,Cornwell14}.
-Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories.
+Understanding drivers of water quality change in estuaries depends on accurate descriptions of source inputs.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes nutrient inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
+Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, recent responses to stressors suggests that ecosystem condition may be changing from past norms.
 Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms (e.g., \textit{Pseudo-nitzschia}, \textit{Alexandrium}, \textit{Dinophysis spp.}) have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Cloern07,Shellenbarger08}.
@@
 Although these changes are linked to drivers at different spatial and temporal scales, describing inputs from the Delta is critical to understand downstream effects.
-
 Rates of primary production in coastal habitats are often defined by nutrient concentrations, although a simple relationship between enrichment and water quality changes can be difficult to determine \citep{Cloern01}.
-Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}.
-In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic fishes and benthic invertebrates can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
+Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}, whereas light availability is the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}.
+Grazing pressure from pelagic fishes and benthic invertebrates can also reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.
 Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.
@@
 Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}.
-Changes in phytoplankton biomass and the effects on pelagic organisms have also been of concern.
-\citet{Jassby08} described decadal trends in phytoplankton biomass to understand mechanisms of decline for pelagic fish populations in the Delta.
-Although phytoplankon concentrations have been relatively consistent in Suisan Bay, biomass in the upper Delta has been increasing.
+Although phytoplankton concentrations have been relatively consistent in recent years in Suisan Bay, biomass in the upper Delta has been increasing \citep{Jassby08}.
 Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.
+Descriptions of nitrogen trends over several decades could be used to understand these recent changes, particularly in the context of primary production and physical drivers of change.
 
@@
 A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.
-Although these data have been used extensively, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated.
+Although these data have been used extensively \citep[e.g.,][]{Lehman92,Jassby08,Glibert10}, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated.
 Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
@@
 Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.
+A complete assessment of the Delta dataset that considers these variable effects is necessary to characterize the different trends in nitrogen analytes.
 
@@
 
-The goal of this study was to provide a comprehensive description of nutrient trends in the Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change.
-This study also supports a general need to more effectively use long-term monitoring data to inform management priorities for water quality in coastal regions.
+The goal of this study was to provide a comprehensive description of nutrient trends in the Delta over the last forty years.
+This information can inform the  understanding of ecosystem response dynamics and potential causes of water quality change.
 The specific objectives were to \begin{inparaenum}[1\upshape)]
@@
 Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
-The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08}, limits on phytoplankton growth from high turbidity, and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.
-However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales \citep[e.g., variation in sea surface temperatures,][]{Cloern07}.
 
@@
 Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.
-Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.
+Stations were grouped by location in the study area for comparison: \textit{peripheral} Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; \textit{interior} Delta stations D19, D26, D28; and \textit{Suisun} stations D4, D6, and D7.
 These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}.
@@
 The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
-Less than 3\% of all observations were below the deteciton limit (left-censored), although variation was observed between analytes and location.
+Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.
 The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).
@@
 
-\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).
-Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).
+\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal ($\beta_3$, $\beta_4$) trend, and $Q$ is the flow variable (either flow or salinity depending on station).
+
+The \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).
 Rather than fitting a global model to the entire time series, one regression is fit to every observation.
@@
 
-A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10}.
+A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10} or tidal variation \citep{Beck15}.
 Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.
@@
 
-\subsection{Observed and modelled Data}
+\subsection{Observed data and modelled trends}
 
@@
 
-Long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).
-Although each station varied in the overall concentrations, patterns within the three Deltra regions (peripheral, interior, and Suisun) were observed.
+Relative to the observed data, long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).
+Although each station varied in the overall concentrations, patterns within the three Delta regions (peripheral, interior, and Suisun) were observed.
 Concentrations for all nitrogen analytes were highest in the peripheral stations.
@@
 Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).
-The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
+The most notable change in the flow-normalized results was an overall decrease (less positive trend or greater negative trend) in concentrations for most sites in the summer and fall seasons for 1996-2013.
+More statistically significant trends were also observed with the flow-normalized results.
 
@@
 Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that is not possible with raw observations.
-A general conclusion is that nitrogen concentrations have decreased overall throughout the approximate forty years of observation.
+A general conclusion is that nitrogen concentrations have showed a consistent decrease beginning in the mid-1990s and early 2000s, although average concentrations remain above those observed earlier in the period of record.
 These results are confirmed visually from \ac{wrtds} (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1}, \cref{tab:trndsann}).
@@
 Differing magnitudes of nitrogen analytes between stations as a function of source type can have an effect on the relationship between flow and nutrients.
-Both \citet{Hirsch10} and \citet{Beck15b} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.
+Both \citet{Hirsch10} and \citet{Beck15} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.
 In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.
@@
 Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
+Our analysis is the first attempt to model nutrient dynamics related to flow in the entire Delta, such that additional work should focus on improving the characterization of flow signals at each station.
 
@@
 Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 0.072 mg L$^{-1}$ (4 $\mu$mol L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.
-Energetic costs of ammonium for phytoplankton growth are lower than nitrate such that the former tends to stimulate growth in most cases.
-Although phytoplankton communities in \ac{sfe} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
+Energetic costs of ammonium are lower than nitrate for phytoplankton growth.
+Although algal communities in \ac{sfe} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
 In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate that stimulates bloom development.
@@
 Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium.
-However, a reversal of trends in in recent years may also be occuring in Suisun Bay, as model estimates suggest either relatively constant concetratrations or even a decrease at some stations beginnign around 2000 (e.g., D6, D7, \cref{fig:trndcomp1}).
+However, a reversal of trends in recent years may also be occuring in Suisun Bay, as model estimates suggest either relatively constant concentrations or even a decrease at some stations beginning around 2000 (e.g., D6, D7, \cref{fig:trndcomp1}).
 Combined with the shift towards a dominant spring peak in chloropyll growth (\cref{fig:clmchl}f), changing nitrogen ratios continue to be a concern for the management of production in the upper \ac{sfe}.
@@
 As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure.
-These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.
+These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanisms demonstrated by \ac{wrtds} provide additonal quantitative evidence of factors that drive water quality in the Delta.
 
@@
 We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.
-The results from \ac{wrtds} are also consistent with described trends of nutrient loads from point sources (e.g., Sacramento \ac{wwtp} increases \citep{Jassby08} and exports to Suisun Bay \citep{Novick14}) demonstrating that these changes are not unexpected.
+The results from \ac{wrtds} are also consistent with described trends of nutrient loads from point sources \citep[e.g., Sacramento \ac{wwtp} increases and exports to Suisun Bay,][]{Jassby08,Novick14}, demonstrating that these changes are not unexpected.
 Consequently, we are not detracting from the potential implications of such increases.
@@
 For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow effects.
-Alternative methods could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models, \cite{Beck17}).
-Overall, quantitative interpretations of multiple factors can provide a basis for identifying links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.
+Alternative methods could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables \citep[e.g., generalized additive models,][]{Beck17}.
+Overall, quantitative interpretations of multiple factors can provide a more comprehensive understanding of relationships between nutrients and primary production, including adverse effects on ecosystem condition.
 
@@
 We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program.
-We thank Larry Harding, Raphael Mazor, Yongshan Wan, and Elizabeth George for providing comments on an earlier draft.
+We thank Larry Harding, Raphael Mazor, Yongshan Wan, and Susan Elizabeth George for providing comments on an earlier draft.
 This study was reviewed and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory.
@@
 
-<<prddat, fig.cap = 'Model predictions for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \\cref{fig:delt_map}).
-Results are annually-averaged for each water year from October to September.
-Points are model predictions and lines are flow-normalized predictions.', fig.height = 8, fig.width = 8>>=
+<<prddat, fig.cap = 'Model results (points) and flow-normalized predictions (lines) for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \\cref{fig:delt_map}).
+Results are annually-averaged for each water year from October to September.', fig.height = 8, fig.width = 8>>=
 data(mods)
@@
 Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station.
-Median values are reported in mg L$^{-1}$.'
+Median values (mg L$^{-1}$) are reported for the entire period of record.'
 cgrps <- c('', '', 'DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')

### 2017-10-27 sftrends_manu.tex
minor edits to maybe final draft

@@
   \vspace{1in}
-  \\ Version Date:   Fri Oct 20 11:14:05 2017 -0700
+  \\ Version Date:   Fri Oct 20 11:24:35 2017 -0700
 	}
@@
 The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
-Less than 3\% of all observations were below the deteciton limit (left-censored), although variation was observed between analytes and location.
+Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.
 The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).
@@
 
-\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).
-Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from 433 at D19 to 571 at C3).
+\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal ($\beta_3$, $\beta_4$) trend, and $Q$ is the flow variable (either flow or salinity depending on station).
+
+The \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from 433 at D19 to 571 at C3).
 Rather than fitting a global model to the entire time series, one regression is fit to every observation.
@@
 
-\caption{Model predictions for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \cref{fig:delt_map}).
-Results are annually-averaged for each water year from October to September.
-Points are model predictions and lines are flow-normalized predictions.}\label{fig:prddat}
+\caption{Model results (points) and flow-normalized predictions (lines) for ten stations grouped by nitrogen analyte and geographic location in the Delta region (locations in \cref{fig:delt_map}).
+Results are annually-averaged for each water year from October to September.}\label{fig:prddat}
 \end{figure}
@@
 Records from 1976 to 2013 were evaluated using the total observations ($n$) at each station.
-Median values are reported in mg L$^{-1}$.\label{tab:stdescrp}}
+Median values (mg L$^{-1}$) are reported for the entire period of record.\label{tab:stdescrp}}
 \begin{center}

### 2017-10-27 sftrends_manu.Rnw
version bump
abstract edit for idiocy

@@
 \usepackage{bm}
-\usepackage[inline]{showlabels}
+% \usepackage[inline]{showlabels}
 \bibpunct{(}{)}{,}{a}{}{,}
@@
 \section*{Abstract}
-\noindent Quantitative descriptions of water quality, physical, and biological characteristics in estuaries is critical for developing an ecological understanding of drivers of change.
+\noindent Quantitative descriptions of water quality, physical, and biological characteristics of estuaries are critical for developing an ecological understanding of drivers of change.
 Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \acl{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.

### 2017-10-27 sftrends_manu.tex
version bump
abstract edit for idiocy

@@
   \vspace{1in}
-  \\ Version Date:   Fri Oct 20 11:24:35 2017 -0700
+  \\ Version Date:   Fri Oct 27 15:55:09 2017 -0700
 	}

### 2018-01-24 sftrends_manu.Rnw
minor edits from Thomas and Phil comments

@@
 \section*{Abstract}
-\noindent Quantitative descriptions of water quality, physical, and biological characteristics of estuaries are critical for developing an ecological understanding of drivers of change.
+\noindent Quantitative descriptions of chemical, physical, and biological characteristics of estuaries are critical for developing an ecological understanding of drivers of change.
 Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \acl{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.
-Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.
-Modelled trends demonstrated that nutrient concentrations were on average higher in the last twenty years relative to the earlier periods of observation, although a consistent decline began in the mid-1990s and early 2000s.
-We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.
+Analysis of flow-normalized data revealed trends that were different from those in the observed time-series.
+Flow-normalized data exhibited changes in magnitude and even reversal of trends relative to the observed data.
+Modelled trends demonstrated that nutrient concentrations were on average higher in the last twenty years relative to the earlier periods of observation, although concentrations have been slowly declining since the mid-1990s and early 2000s.
+We further describe mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, macro-nutrients (nitrogen and silica), and flow in Suisun Bay.
 \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \acl{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions.
@@
 Understanding drivers of water quality change in estuaries depends on accurate descriptions of source inputs.
-The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes nutrient inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.
-Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, recent responses to stressors suggests that ecosystem condition may be changing from past norms.
-Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms (e.g., \textit{Pseudo-nitzschia}, \textit{Alexandrium}, \textit{Dinophysis spp.}) have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Cloern07,Shellenbarger08}.
-As a result, \ac{chla} thresholds to assess and manage levels of concern for phytoplankton biomass in the lower estuary have been proposed \citep{Sutula17}.
+The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes nutrient inputs from primarily urban sources, in addition to agricultural sources \citep{Jassby00,Jassby02,Jassby08}.
+Although water quality conditions in the \ac{sfe} symptomatic of eutrophication have historically been infrequent, recent responses to stressors suggests that ecosystem condition may be changing from past norms.
+Changes in phytoplankton biomass and composition, water clarity changes from sediment alteration, increases in harmful cyanobacterial blooms (\textit{Microcystis aeruginosa}), increases in non-native macrophytes, and periodic events of low dissolved oxygen have been a recent concern for management of the Delta \citep{Lehman05,Santos09,Hestir13,Lehman15,Dahm16,ASC17}.
 Although these changes are linked to drivers at different spatial and temporal scales, describing inputs from the Delta is critical to understand downstream effects.
+
 Rates of primary production in coastal habitats are often defined by nutrient concentrations, although a simple relationship between enrichment and water quality changes can be difficult to determine \citep{Cloern01}.
@@
 \citet{Glibert14} attributed recent phytoplankton blooms in Suisun Bay to a drought, during which residence times and nitrification rates increased.
-Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}.
-Although phytoplankton concentrations have been relatively consistent in recent years in Suisan Bay, biomass in the upper Delta has been increasing \citep{Jassby08}.
-Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.
-Descriptions of nitrogen trends over several decades could be used to understand these recent changes, particularly in the context of primary production and physical drivers of change.
-
-Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change in coastal regions.
-A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.
-Although these data have been used extensively \citep[e.g.,][]{Lehman92,Jassby08,Glibert10}, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated.
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
-An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported to the estuary \citep{Novick15}.
+Speciation changes in the dominant forms of nitrogen confound a direct interpretation of links with phytoplankton blooms and the relationships between nutrients and primary production in the upper estuary are not completely understood.
+Although phytoplankton concentrations have been relatively consistent in recent years in Suisun Bay, biomass trends in the Delta are mixed \citep{Jassby08,ASC17}.
+Descriptions of nitrogen trends over several decades could be used to better understand long-term and more recent changes, particularly in the context of primary production and physical drivers of change \citep{Dahm16}.
+
+Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change.
+A comprehensive water quality monitoring program has been in place in the upper \ac{sfe} for several decades \citep[\cref{fig:delt_map},][]{IEP13}.
+Although these data have been used extensively \citep[e.g.,][]{Lehman92,Jassby08,Glibert10}, water quality trends covering the full spatial and temporal coverage of the Delta have not been systematically evaluated.
+Quantitative descriptions of nutrient dynamics are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
+An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported \citep{Novick15}.
 Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.
-A complete assessment of the Delta dataset that considers these variable effects is necessary to characterize the different trends in nitrogen analytes.
+It is important to consider these variable effects to characterize different trends in nitrogen forms.
 
@@
 Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
-For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}.
-Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
-
-The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
+Similarly, nutrient trends that vary with hydrologic loading can vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
+Concentration of \ac{chla} as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}.
+Describing the relationship of a water quality variable as changing (chemodynamic) or invariant (chemostatic) with flow can isolate components, such as nutrient inputs, for a more direct assessment of causal factors \citep{Wan17}.
+
+The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in river systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
 The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}.
-Although the \ac{wrtds} method has been effectively applied to describe changes in freshwater systems, use in coastal ecosystems has not been as extensive.
+Although the \ac{wrtds} method has been effectively applied to describe changes in freshwater systems, use in tidally influenced systems has not been as extensive.
 Application of \ac{wrtds} to describe trends in estuaries could reveal new insights given the disproportionate effects of physical drivers, such as flow inputs and tidal exchange, on water quality.
@@
 
-The goal of this study was to provide a comprehensive description of nutrient trends in the Delta over the last forty years.
+The goal of this study was to provide a comprehensive description of nutrient trends in the Delta and Suisun Bay over the last forty years.
 This information can inform the  understanding of ecosystem response dynamics and potential causes of water quality change.
 The specific objectives were to \begin{inparaenum}[1\upshape)]
-\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
+\item quantify and interpret trends over four decades at ten stations in the Delta and Suisun Bay, including annual, seasonal, and spatial changes in nitrogen forms and response to flow variation, and
 \item provide detailed descriptions of two case studies in the context of conceptual relationships modeled with \ac{wrtds}.
 \end{inparaenum}
-The second objective evaluated two specific water quality stations to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.
+The second objective evaluated two specific water quality stations to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on nutrient cycling and primary production.
 Our general hypothesis was that the results were expected to support previous descriptions of trends in this well-studied system, but that new insight into spatial and temporal variation in response endpoints was expected, particuarly in flow-normalized model predictions.
@@
 Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.
-Stations were grouped by location in the study area for comparison: \textit{peripheral} Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; \textit{interior} Delta stations D19, D26, D28; and \textit{Suisun} stations D4, D6, and D7.
+Stations were grouped by location in the study area for comparison: \textit{peripheral} Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, and P8; \textit{interior} Delta stations D19, D26, and D28; and \textit{Suisun} stations D4, D6, and D7.
 These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}.
-Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}.
+Although many other stations are available for the region, the stations were chosen because they represent a consistent long-term dataset generated by a single program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}.
 Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.
 Data were minimally processed, with the exception of averaging replicates that occurred on the same day.
-The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
+The three nitrogen forms that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
 Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.
@@
 \ac{wrtds} models require flow data paired with nutrient data.
-At the peripheral and interior stations, daily flow estimates were matched with the corresponding sample dates for the nutrient data.
+At the Delta stations, daily flow estimates were matched with the corresponding sample dates for the nutrient data.
 Daily flow estimates were obtained from the Dayflow software program \citep{IEP16}.
@@
 Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.
-Salinity was used as a proxy for flow at sites D4, D6, and D7 where tidal influences were much stronger.
+Salinity was used as a proxy for flow at the Suisun Bay sites D4, D6, and D7 where tidal influences were much stronger.
 Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.
-Models were evaluated using salinity at the interior and peripheral stations but performance was reduced relative to models that used daily flow estimates.
+Models were evaluated using salinity at the Delta stations, but performance was reduced relative to models that used daily flow estimates.
 
@@
 
-A total of thirty \ac{wrtds} models were created, one for each nitrogen analyte at each station.
+A total of thirty \ac{wrtds} models were created, one for each nitrogen form at each station.
 The functional form of \ac{wrtds} is a simple regression \citep{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:
@@
 The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
-An increase and decrease of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
+A decrease of ammonium concentrations at P8 after 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
 \item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
@@
 \end{inparaenum}
-Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion and the relationship of phytoplankton with clam abundance was mediated by flow and climatic variation in recent years.
+Although the relationship between phytoplankton and clams in the upper \ac{sfe} has been well described \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion and the relationship of phytoplankton with clam abundance was mediated by flow and climatic variation in recent years.
 
@@
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when assimilation rates are lower \citep{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral Delta stations (C3, C10, MD10, P8; mean $\pm$ s.e.: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior Delta (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).
 Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
@@
 Relative to the observed data, long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).
-Although each station varied in the overall concentrations, patterns within the three Delta regions (peripheral, interior, and Suisun) were observed.
+Although each station varied in the overall concentrations, patterns within the three regions (peripheral Delta, interior Delta, and Suisun Bay) were observed.
 Concentrations for all nitrogen analytes were highest in the peripheral stations.
-Ammonium concentrations at P8 and C3 were highest and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.
-By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.
-Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.
-Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.
-These trends were similar for the Suisun stations, although the reduction in the late 1990s did not occur.
-By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.
+Ammonium concentrations were highest at P8 and C3 and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.
+By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral Delta stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.
+Trends at the interior Delta stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.
+Trends in \ac{din} and nitrite/nitrate for the interior Delta stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.
+These trends were similar for the Suisun Bay stations, although the reduction in the late 1990s did not occur.
+By contrast, ammonium concentrations were low in Suisun Bay but a gradual increase over the period of record was observed.
 
@@
 
-Water quality conditions in the Delta are dynamic and not easily characterized from observed time series.
-Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that is not possible with raw observations.
+Water quality conditions in the Delta-Suisun Bay region are dynamic and not easily characterized from observed time series.
+Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions that is not possible with raw observations.
 A general conclusion is that nitrogen concentrations have showed a consistent decrease beginning in the mid-1990s and early 2000s, although average concentrations remain above those observed earlier in the period of record.
@@
 Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are important independent of overall trends.
-Nutrient concentrations were highest at the peripheral stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento, San Joaquin, Cosumnes, and Mokelumne rivers.
-The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 as a direct consquence of \ac{wwtp} inputs upstream.
-Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento Regional County Sanitation District.
+Nutrient concentrations were highest at peripheral Delta stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento and San Joaquin rivers.
+The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 in the early 2000s as a direct consequence of \ac{wwtp} inputs upstream prior to infrastructure upgrades.
+Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento \ac{wwtp}.
 By contrast, nitrite/nitrate concentrations were highest at C10 as a measure of contributions from the San Joaquin River to the south that drains a predominantly agricultural watershed.
-Although the Sacramento River drains a much larger area, the dominant ammonium signal as compared to nitrite/nitrate at C3 underscores the importance of \ac{wwtp} control for water quality issues downstream.
 
@@
 Both \citet{Hirsch10} and \citet{Beck15} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.
-In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.
+In particular, a chemodynamic (i.e., changing) response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.
 Increased flow may induce a change in nutrient concentrations, such that reduction may occur with flushing or an increase may occur through mobilization.
@@
 Accordingly, trend analysis of nitrate/nitrate by year and season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).
-This is consistant with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly developed watersheds \citep{Wan17}.
+This is consistent with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly altered watersheds \citep{Wan17}.
 
 % obs v flow norm trends
-Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
+Our results underscore the importance of considering flow effects in the evaluation of water quality trends.
+There were important differences in trends between observed and flow-normalized data.
+These differences ranged from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}).
 Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.
 An alternative evaluation of flow in the Delta demonstrated that flow contributions from different sources vary considerably over time at each station \citep{Novick15}.
-For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.
-For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.
+For example, flow at MD10 represents seasonally changing inputs from the Sacramento and San Joaquin rivers and other sources.
+For simplicity, our analysis considered only the dominant water source and used the total daily inflow estimates from either the Sacramento River or the San Joaquin River at any of the Delta sites.
 Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location.
@@
 Long-term trends in nutrient and phytoplankton concentations in Suisun Bay have also been the focus of intense study for many years \citep{Cloern83,Lehman92,Dugdale07,Jassby08,Glibert14}.
-Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 0.072 mg L$^{-1}$ (4 $\mu$mol L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.
-Energetic costs of ammonium are lower than nitrate for phytoplankton growth.
-Although algal communities in \ac{sfe} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.
-In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate that stimulates bloom development.
 Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values of annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000 (\cref{fig:prddat}).
@@
 This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.
-Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
+A nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P.
 amurensis}.
@@
 We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program.
+All data are from the California Department of Water Resources Environmental Monitoring Program.
 We thank Larry Harding, Raphael Mazor, Yongshan Wan, and Susan Elizabeth George for providing comments on an earlier draft.
@@
 The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset).
+The grey outline is the legal boundary of the Delta.
 All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).
@@
 Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
-Trends are shown separately for different annual groupings.
+Trends are shown separately for different annual groupings from 1976 to 1995 (top) and 1996 to 2013 (bottom).
 See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
@@
 <<stock, fig.height = 3, fig.width = 8, cache = T, out.width = '\\textwidth', fig.cap = 'Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.'>>=
+Wastewater discharge requirements were implemented in 2006 to convert ammonium to nitrate.'>>=
 data(stock_conc)

### 2018-01-24 sftrends_manu.tex
minor edits from Thomas and Phil comments

@@
   \vspace{1in}
-  \\ Version Date:   Fri Oct 27 15:55:09 2017 -0700
+  \\ Version Date:   Fri Oct 27 15:58:09 2017 -0700
 	}
@@
 \citet{Glibert14} attributed recent phytoplankton blooms in Suisun Bay to a drought, during which residence times and nitrification rates increased.
-Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}.
-Although phytoplankton concentrations have been relatively consistent in recent years in Suisan Bay, biomass in the upper Delta has been increasing \citep{Jassby08}.
-Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.
-Descriptions of nitrogen trends over several decades could be used to understand these recent changes, particularly in the context of primary production and physical drivers of change.
+Speciation changes in the dominant forms of nitrogen confound a direct interpretation of links with phytoplankton blooms and the relationships between nutrients and primary production in the upper estuary are not completely understood.
+Although phytoplankton concentrations have been relatively consistent in recent years in Suisun Bay, biomass trends in the Delta are mixed \citep{Jassby08,ASC17}.
+Descriptions of nitrogen trends over several decades could be used to better understand long-term and more recent changes, particularly in the context of primary production and physical drivers of change \citep{Dahm16}.
 
-Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change in coastal regions.
-A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.
-Although these data have been used extensively \citep[e.g.,][]{Lehman92,Jassby08,Glibert10}, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated.
-Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
-An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported to the estuary \citep{Novick15}.
+Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change.
+A comprehensive water quality monitoring program has been in place in the upper \ac{sfe} for several decades \citep[\cref{fig:delt_map},][]{IEP13}.
+Although these data have been used extensively \citep[e.g.,][]{Lehman92,Jassby08,Glibert10}, water quality trends covering the full spatial and temporal coverage of the Delta have not been systematically evaluated.
+Quantitative descriptions of nutrient dynamics are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.
+An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported \citep{Novick15}.
 Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.
-A complete assessment of the Delta dataset that considers these variable effects is necessary to characterize the different trends in nitrogen analytes.
+It is important to consider these variable effects to characterize different trends in nitrogen forms.
 
@@
 Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes.
-For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}.
-Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
+Similarly, nutrient trends that vary with hydrologic loading can vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.
+Concentration of \ac{chla} as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}.
+Describing the relationship of a water quality variable as changing (chemodynamic) or invariant (chemostatic) with flow can isolate components, such as nutrient inputs, for a more direct assessment of causal factors \citep{Wan17}.
 
-The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
+The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in river systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
 The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}.
-Although the \ac{wrtds} method has been effectively applied to describe changes in freshwater systems, use in coastal ecosystems has not been as extensive.
+Although the \ac{wrtds} method has been effectively applied to describe changes in freshwater systems, use in tidally influenced systems has not been as extensive.
 Application of \ac{wrtds} to describe trends in estuaries could reveal new insights given the disproportionate effects of physical drivers, such as flow inputs and tidal exchange, on water quality.
@@
 However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when assimilation rates are lower \citep{Novick15}.
-By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
+By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral Delta stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior Delta (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
@@
 \caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.
-Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
+Wastewater discharge requirements were implemented in 2006 to convert ammonium to nitrate.}\label{fig:stock}
 \end{figure}

### 2018-02-02 sftrends_manu.Rnw
new cover letter for ecss, removed old est files
recompile with tinytex
zenodo doi badge

@@
 \usepackage{booktabs}
-\usepackage{graphicx}
-\usepackage{verbatim}
-\usepackage{rotating}
-\usepackage{tabularx}
 \usepackage{outlines}
 \usepackage{lineno}
-\usepackage{array}
-\usepackage{times}
 \usepackage{cleveref}
@@
 \usepackage{natbib}
-\usepackage{array}
-\usepackage{pdflscape}
-\usepackage{bm}
 % \usepackage[inline]{showlabels}
@@
   \vspace{1in}
+  \\{Declarations of interest: none}
+  \vspace{0.1in}
   \\ \Sexpr{raw}
@@
 Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
+
+\noindent \textit{Key words}: estuary, nitrogen, Sacramento - San Joaquin Delta, trend analysis, weighted regression
 \acresetall
@@
 The three nitrogen forms that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
-Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.
+Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between nitrogen forms and location.
 The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).
@@
 
-\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal ($\beta_3$, $\beta_4$) trend, and $Q$ is the flow variable (either flow or salinity depending on station).
+\noindent where $N$ is one of three nitrogen forms, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal ($\beta_3$, $\beta_4$) trend, and $Q$ is the flow variable (either flow or salinity depending on station).
 
@@
 
-The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).
+The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen forms and differences in apparent seasonal trends (\cref{fig:obsdat}).
 \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.
@@
 Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate.
-Mean observed concentrations were also higher later in the time series for all analytes.
+Mean observed concentrations were also higher later in the time series for all forms.
 For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}.
@@
 
-Relative to the observed data, long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).
+Relative to the observed data, long-term trends between stations for the different nitrogen forms were apparent from the modelled results (\cref{fig:prddat}).
 Although each station varied in the overall concentrations, patterns within the three regions (peripheral Delta, interior Delta, and Suisun Bay) were observed.
-Concentrations for all nitrogen analytes were highest in the peripheral stations.
+Concentrations for all nitrogen forms were highest in the peripheral stations.
 Ammonium concentrations were highest at P8 and C3 and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.
@@
 
-Estimated trends from Seasonal Kendall tests on the raw time series varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Estimated trends from Seasonal Kendall tests on the raw time series varied considerably between sites and nitrogen forms (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).
@@
 Water quality conditions in the Delta-Suisun Bay region are dynamic and not easily characterized from observed time series.
-Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions that is not possible with raw observations.
+Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen forms in three distinct regions that is not possible with raw observations.
 A general conclusion is that nitrogen concentrations have showed a consistent decrease beginning in the mid-1990s and early 2000s, although average concentrations remain above those observed earlier in the period of record.
@@
 Nutrient concentrations were highest at peripheral Delta stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento and San Joaquin rivers.
-The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 in the early 2000s as a direct consequence of \ac{wwtp} inputs upstream prior to infrastructure upgrades.
+The highest concentrations among all nitrogen forms, ammonium in particular, were observed at P8 in the early 2000s as a direct consequence of \ac{wwtp} inputs upstream prior to infrastructure upgrades.
 Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento \ac{wwtp}.
@@
 
-Differing magnitudes of nitrogen analytes between stations as a function of source type can have an effect on the relationship between flow and nutrients.
+Differing magnitudes of nitrogen forms between stations as a function of source type can have an effect on the relationship between flow and nutrients.
 Both \citet{Hirsch10} and \citet{Beck15} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.
@@
 \includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen forms.
 Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
+cap.val <- 'Summaries of flow-normalized trends in nitrogen forms for all stations and annual aggregations.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
@@
 \includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen forms.
 Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
@@
 \includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
-\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes.
+\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen forms.
 Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$).
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
+cap.val <- 'Summaries of flow-normalized trends in nitrogen forms for all stations and seasonal aggregations from 1976-1995.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
@@
 # table stuff
-cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
+cap.val <- 'Summaries of flow-normalized trends in nitrogen forms for all stations and seasonal aggregations from 1996-2013.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).

### 2018-02-02 sftrends_manu.tex
new cover letter for ecss, removed old est files
recompile with tinytex
zenodo doi badge

@@
   \vspace{1in}
-  \\ Version Date:   Fri Oct 27 15:58:09 2017 -0700
+  \\{Declarations of interest: none}
+  \vspace{0.1in}
+  \\ Version Date:   Sun Jan 28 10:22:13 2018 -0800
 	}
@@
 The three nitrogen forms that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).
-Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.
+Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between nitrogen forms and location.
 The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).
@@
 Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate.
-Mean observed concentrations were also higher later in the time series for all analytes.
+Mean observed concentrations were also higher later in the time series for all forms.
 For example, average \ac{din} across all stations was $0.61$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.7$$\pm$$0.01$ for 1996-2013.
@@
 
-Estimated trends from Seasonal Kendall tests on the raw time series varied considerably between sites and analytes (\cref{fig:trndcomp1}).
+Estimated trends from Seasonal Kendall tests on the raw time series varied considerably between sites and nitrogen forms (\cref{fig:trndcomp1}).
 Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).
@@
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.
+\caption{Summaries of flow-normalized trends in nitrogen forms for all stations and annual aggregations.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
@@
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995.
+\caption{Summaries of flow-normalized trends in nitrogen forms for all stations and seasonal aggregations from 1976-1995.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).
@@
 \begin{table}[!tbp]
-\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013.
+\caption{Summaries of flow-normalized trends in nitrogen forms for all stations and seasonal aggregations from 1996-2013.
 Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold).

### 2018-02-09 sftrends_manu.Rnw
SCCWRP as primary address

@@
   \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
-  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
-  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
-  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
+  \\\\{\textit {\normalsize $^1$Southern California Coastal Water Research Project}}
   \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
@@
 We thank Larry Harding, Raphael Mazor, Yongshan Wan, and Susan Elizabeth George for providing comments on an earlier draft.
-This study was reviewed and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory.
 The authors declare no competing financial interest.
-The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.
 

### 2018-02-09 sftrends_manu.tex
submitted to ECSS
SCCWRP as primary address

@@
   \vspace{0.1in}
-  \\ Version Date:   Sun Jan 28 10:22:13 2018 -0800
+  \\ Version Date:   Fri Feb 2 17:52:16 2018 -0800
 	}

### 2018-04-20 revision/review_response.tex
started ecss revision, figure 1 simplified

\documentclass[letterpaper,12pt]{article}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage{lineno}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage{paralist}
\usepackage{bm}
\usepackage{fixltx2e}

%acronyms
% \acrodef{}{}

\setlength{\parskip}{5mm}
\setlength{\parindent}{0in}

\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}

\begin{document}
\raggedright

% \title{}
% \author{}
% \maketitle

{\it Response to reviewer comments ``Four decades of water quality change in the upper San Francisco Estuary'' Beck MW, Jabusch, TW, Trowbridge PR, Senn DB.}

{\it We thank the editor and reviewer for providing helpful comments on our manuscript.  Responses to these comments are shown in italics.}

\Bigtxt{Editor's comments:}

Thank you for submitting your manuscript to Estuarine, Coastal and Shelf Science. We have completed the review of your manuscript. One reviewers recommended rejection, while the other was more lenient. If you can address the comments of Reviewer 1 adequately I will consider the MS for another round of reviews by the same reviewers. A summary is appended below. Particularly, the paper is not suitable for publication because ``.....in its present form unless there is a substantial revision to better explain/identify how this research advances our knowledge on nutrient / phytoplankton dynamics in estuarine ecosystems''.  While revising the paper please consider the reviewers' comments carefully. We look forward to receiving your detailed response and your revised manuscript.

{\it We appreciate the concern of reviewer 1 regarding significance of this work beyond the study system.  However, we argue that this work will have good readership in ECSS for a number of reasons.  San Francisco estuary, although a single system, is one of the largest and most studied estuaries in the world.  Although we hate to play the `large-prominent system' card as our burden of proof, the reality is that research conducted here has widespread implications for other systems and methods that are proven to work here are often adopted elsewhere.  See for example, ...  We also disagree that the application of WRTDS to tidal systems has been extensive in the literature, as compared to the original method developed for streams (Hirsch et al. 2010).  In fact, the only other applications in tidal waters were conducted by myself and colleagues.  Extension of WRTDS to SFE should be considered as a novel application given that the approach has only been applied to two other locations.  We emphasis in the text some of the unique challenges and finding in applying WRTDS to SFE.  Finally, we also want to mention that previous work specific to SFE has already been published in ECSS.  Sutula et al. (2017) described an approach for estimating biostimulatory nutrient targets for SFE.  The relevance of this work was of immediate regional importance and the broader importance was implicitly assumed.

That being said, we have revised the manuscript to help emphasize the relevance of this work outside of the region.  Please see my specific responses to reviewer 1. 
}

\Bigtxt{Reviewer 1:}

The authors conducted a detailed data/trend analysis of nutrient variability in the upper San Francisco Estuary (SFE) based on observations collected during the last four decades. Overall, this is a very nicely written, organized and presented manuscript and I don't have major comments related to the content or analysis presented in the manuscript. However, I am very concerned that the paper seems to be a study of only local significance as it is difficult to identify new methodologies or findings of widespread impact. The method implemented for the trend analyses, namely WRTDS, has been previously used in other investigations to describe decadal trends in rivers and tidal systems (see references in MS-Lines 92-96). So, unfortunately, it doesn't seems that the implementation of WRTDS for the SFE can be considered as a novel approach. The conclusions and analyses presented from the use of WRTDS seem to be supported by the data (e.g. DIN changes at P8 following the implementation of the WWTP), but again, it is difficult to identify new causal relationships or new information that can be used to better understand nutrient dynamics in other systems around the world.

Because of the limitations identified above, and having in mind that the journal discourages the submission of research of mainly local significance, I believe the paper should be declined in its present form unless there is a substantial revision to better explain/identify how this research advances our knowledge on nutrient / phytoplankton dynamics in estuarine ecosystems.

Minor comments

Check figure 1. Seems disorganized. Map of California in wrong place and not at scale.

{\it Figure 1 was simplified and the California map was given its own inset.}

Line  270 - 271. The figures don't seem to support this statement. If the colors are correct, then I see NH4 as the predominant form of Nitrogen in most stations.

\Bigtxt{Reviewer 2:}

This is a very well written article on an important subject.  It makes a significant contribution to the literature with regard to analysis of nutrient loading to an estuary.  This is a very well studied estuary already, and the results of the study are not at all unexpected, but it is an excellent application of new analysis methodology for estuaries subject to temporal, seasonal, and hydrologically influenced changes.  There is not much that I could find in the way of suggested improvements.  I would recommend it for publication with minor revisions as follows.

p. 7, line 130   Seasonally, inflow from the watershed ...

This a long sentence with three clauses that don't quite follow from one to the next, in my opinion.  I think the manuscript would read better by splitting the sentence into two.

p. 10, line 186 - I think readers would appreciate another sentence or two on how the weighting within the time window is accomplished.  I assume it is with some sort of exponentially decaying time function, but there isn't any description provided.  A short explanation is all that's needed.

p.10, line 195 - This is a follow-up to the previous comment.  Adding a mention of the time constant in the description of the method will help setup the presentation of the times used for averaging.  I think you should add some information here on how these time windows were chosen.

Figure 2.  More contrast in colors (perhaps blue and gray) would make it easier to distinguish ammonia and nitrate in the time history.  As is, it is somewhat difficult to see

p.18, line  In the text, I believe in the introduction, you mention one invading asian clam (potamocorbula), but the second clam genera (corbicula) isn't mentioned until the results and in figure and figure caption.  Is corbicula a native clam that was displaced?  It is not clear from the text what we are to make of the decrease in abundance of corbicula.  A few explanatory sentences in the introduction and/or the results sections are needed.

\end{document}

### 2018-04-20 revision/review_response.tex
more revisions

@@
 {\it We appreciate the concern of reviewer 1 regarding significance of this work beyond the study system.
-However, we argue that this work will have good readership in ECSS for a number of reasons.
-San Francisco estuary, although a single system, is one of the largest and most studied estuaries in the world.
-Although we hate to play the `large-prominent system' card as our burden of proof, the reality is that research conducted here has widespread implications for other systems and methods that are proven to work here are often adopted elsewhere.
-See for example, ...
-We also disagree that the application of WRTDS to tidal systems has been extensive in the literature, as compared to the original method developed for streams (Hirsch et al.
-2010).
-In fact, the only other applications in tidal waters were conducted by myself and colleagues.
-Extension of WRTDS to SFE should be considered as a novel application given that the approach has only been applied to two other locations.
-We emphasis in the text some of the unique challenges and finding in applying WRTDS to SFE.
-Finally, we also want to mention that previous work specific to SFE has already been published in ECSS.
-Sutula et al.
-(2017) described an approach for estimating biostimulatory nutrient targets for SFE.
-The relevance of this work was of immediate regional importance and the broader importance was implicitly assumed.
-
-That being said, we have revised the manuscript to help emphasize the relevance of this work outside of the region.  Please see my specific responses to reviewer 1. 
+Our revision addresses this primary concern by making the novel contributions of this work beyond San Francisco estuary more clear.
+We have also addressed the minor comments from both reviewers, including requested revisions to the figures.
 }
@@
 
+{\it We greatly appreciate these comments and we have revised our paper to address these concerns.
+In particular, the application of WRTDS to tidal systems is indeed a novel application that has only been reported in two publications (Beck and Hagy 2016, Beck and Murphy 2017).
+Our application of WRTDS to the San Francisco Delta presented new challenges and novel insights.
+We have revised the text to make these more clear:
+
+}
+
 Because of the limitations identified above, and having in mind that the journal discourages the submission of research of mainly local significance, I believe the paper should be declined in its present form unless there is a substantial revision to better explain/identify how this research advances our knowledge on nutrient / phytoplankton dynamics in estuarine ecosystems.
+
+{\it We have also included text in the discussion that articulates importance of this work for advancing the science of nutrient/phytoplankton dynamics at other locations:
+
+}
 
@@
 Line  270 - 271. The figures don't seem to support this statement. If the colors are correct, then I see NH4 as the predominant form of Nitrogen in most stations.
+
+{\it We have changed the color scheme to make this distinction more clear.
+This was also noted by reviewer 2.
+The dominance of nitrite/nitrate (grey) relative to ammonium (dark blue) is now apparent.}
 
@@
 
+{\it Changed to ``Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt.
+Long-term trends have shown that consumption, withdrawals, and export have steadily increased from 1960 to present, although climate effects have contributed to inter-annual variability (Cloern and Jassby 2012).''}
+
 p.
@@
 
+{\it Details were added: ``Weights within each window are estimated using a tri-cube weighting function, which is similar to a bell curve except the tail ends do not asymptote at zero (Hirsch et al.
+2010).
+This ensures that observations outside of the window are given zero weight, as compared to an infinitely small value.
+The final vector of weights used for each regression is the product of the three separate weight vectors for time, season, and flow.''}
+
 p.10, line 195 - This is a follow-up to the previous comment.
@@
 
+{\it An additional sentence was added to the previous paragraph to describe how time and season values are evaluated: ``Annual values are described using decimal time as a continuous numeric value (e.g., July 1st, 2017 is 2017.5) and season is described as a numeric value for day of year.''.
+
+The text on line 195 was also modified for clarity: `` Model predictions were evaluated as monthly values that were the same as the day of sampling in the observed data and as annual values that averaged monthly predictions within each water year (October to September).''
+}
+
 Figure 2.  More contrast in colors (perhaps blue and gray) would make it easier to distinguish ammonia and nitrate in the time history.  As is, it is somewhat difficult to see
+
+{\it Figure was modified, also in responese to reviewer 1.}
 
@@
 
+{\it Both are non-native invaders with different competitive advantages.
+This was clarified in the text: ``Invasion in the 1980s showed a clear increase of \textit{P.
+amurenusis} at D7 that coincided with a reduction in abundance of the non-native and previously established Asian Clam (\textit{Corbicula fluminea}) (Fig.
+7a).''}
+
 \end{document}

### 2018-04-20 sftrends_manu.Rnw
started ecss revision, figure 1 simplified
more revisions

@@
 Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \citep{Jassby00}.
-Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \citep{Cloern12b}.
+Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt.
+Long-term trends have shown that consumption, withdrawals, and export have steadily increased from 1960 to present, although climate effects have contributed to inter-annual variability \citep{Cloern12b}.
 Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \citep{Cloern15}.
@@
 Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.
+Annual values are described using decimal time as a continuous numeric value (e.g., July 1st, 2017 is 2017.5) and season is described as a numeric value for day of year.
+Weights within each window are estimated using a tri-cube weighting function, which is similar to a bell curve except the tail ends do not asymptote at zero \citep{Hirsch10}.
+This ensures that observations outside of the window are given zero weight, as compared to an infinitely small value.
+The final vector of weights used for each regression is the product of the three separate weight vectors for time, season, and flow.
 This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows.
+
 Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.
 All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data below the detection limit \citep{Tobin58}.
-Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September).
+Model predictions were evaluated as monthly values that were the same as the day of sampling in the observed data and as annual values that averaged monthly predictions within each water year (October to September).
 All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.
@@
 
-Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
+Invasion of the upper \ac{sfe} by the Amur River clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.
 Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \citep{Feyrer03,MacNally10}.
@@
 
-Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P.
-amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
+Invasion in the 1980s showed a clear increase of \textit{P.
+amurenusis} at D7 that coincided with a reduction in abundance of the non-native and previously established Asian Clam (\textit{Corbicula fluminea}) (\cref{fig:clmchl}a).
+Biomass of the \textit{P.
+amurensis} was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).
 The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.
@@
 \caption{The \acl{sfe} and Delta region with monitoring stations used for analysis.
-The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset).
+The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (dark grey region in top right inset).
 The grey outline is the legal boundary of the Delta.
@@
 
-cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
-  rev
+cols <- RColorBrewer::brewer.pal(9, 'Blues')[7] %>% 
+  c('grey')
+
 ylab <- expression(paste('log - DIN (mg ', L^-1, ')'))

### 2018-04-20 sftrends_manu.tex
started ecss revision, figure 1 simplified
more revisions

@@
   \vspace{0.1in}
-  \\ Version Date:   Fri Feb 2 17:52:16 2018 -0800
+  \\ Version Date:   Thu Apr 19 17:53:28 2018 -0700
 	}

### 2018-05-01 revision/review_response.tex
ECSS initial revision done

@@
 Our revision addresses this primary concern by making the novel contributions of this work beyond San Francisco estuary more clear.
+This is accomplished by emphasizing the novelty of application of WRTDS to tidal systems, both in practice and interpretation of the results.
 We have also addressed the minor comments from both reviewers, including requested revisions to the figures.
@@
 
+Because of the limitations identified above, and having in mind that the journal discourages the submission of research of mainly local significance, I believe the paper should be declined in its present form unless there is a substantial revision to better explain/identify how this research advances our knowledge on nutrient / phytoplankton dynamics in estuarine ecosystems.
+
 {\it We greatly appreciate these comments and we have revised our paper to address these concerns.
 In particular, the application of WRTDS to tidal systems is indeed a novel application that has only been reported in two publications (Beck and Hagy 2016, Beck and Murphy 2017).
+Our earlier draft did not sufficiently emphasize that WRTDS has not been fully explored in tidal systems.
 Our application of WRTDS to the San Francisco Delta presented new challenges and novel insights.
-We have revised the text to make these more clear:
+We have added a section to the discussion to make these more clear:
 
-}
+``Use of WRTDS to evaluate long-term trends in water quality has not been extensive in estuarine systems.
+To date, WRTDS has only been applied to evaluate eutrophication trends in Tampa Bay, Florida (Beck and Hagy 2015) and the Patuxent River Estuary (Chesapeake Bay, Beck and Murphy 2017).
+Unique lessons were learned from the application to these two systems, including \begin{inparaenum}[1\upshape)]
+\item use of salinity as an adequate tracer of both tidal forcing and freshwater inputs on indicators of water quality, 
+\item inclusion of quantile models as an extension to evaluate differences in the lower and upper conditional distributions of water quality changes, and 
+\item the ability to hypothesize pollutant sources with changes in the response of water quality to flow over time. 
+\end{inparaenum}
+The application of WRTDS to monitoring data from the Delta is the first to apply the method to a large temperate estuary on the Pacific coast.
 
-Because of the limitations identified above, and having in mind that the journal discourages the submission of research of mainly local significance, I believe the paper should be declined in its present form unless there is a substantial revision to better explain/identify how this research advances our knowledge on nutrient / phytoplankton dynamics in estuarine ecosystems.
+New challenges were encountered related to the geographic location and the unique characteristics of SFE that furthered our understanding of how the method can be used for trend analysis in tidal waters.
+The SFE is a macrotidal system that has daily variation in water depth typical of coastal systems on the Pacific Coast (i.e., $>$ 1 m).
+Previous applications of WRTDS in tidal waters were limited to systems with substantially smaller tidal amplitudes (i.e., $<$ 1 m), demonstrating that WRTDS can be applied in systems with a wide range of tidal influences.
+This point is particularly relevant given that hydrologic factors are often key determinants of water quality changes in coastal systems, suggesting the method is robust in the presence of large tidal swings.
+The SFE is also unique in that excess turbidity, rather than nutrients, is the dominant limiting factor of primary production.
+This differs substantially from Tampa Bay and Chesapeake Bay, neither of which are light-limited from sedimentation.
+Our case study in Suisun Bay provided a unique demonstration of WRTDS to describe chlorophyll dynamics as a function of multiple factors unrelated to nutrients by evaluating the effects of biological invasion on water quality and the complex interaction of flow on both.
+This result provided the first application of WRTDS in tidal waters to evaluate multiple factors that are included explicitly in the model (e.g, flow) and those that are not (e.g., biological forcing).
+As such, WRTDS has the potential to inform the understanding of nutrient and phytoplankton dynamics in coastal systems that vary greatly in physical, chemical, and biological characteristics.''
 
-{\it We have also included text in the discussion that articulates importance of this work for advancing the science of nutrient/phytoplankton dynamics at other locations:
-
+We also want to emphasize that the previous draft did indicate an important novel contribution of this work beyond the San Francisco Estuary.
+In particular, our result that showed trend analyses conducted on observed data could lead to different conclusions relative to flow-normalized data is particularly important and applies independent of location.
+This was described previously on lines 10-11 in the abstract, lines 319-337 in the results, and lines 431-439, 509-511 in the discussion.
 }

### 2018-05-01 sftrends_manu.Rnw
ECSS initial revision done

@@
 The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in river systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.
-The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}.
+The \ac{wrtds} method has more recently been adapted for trend analysis in tidal waters, with a limited focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}.
 Although the \ac{wrtds} method has been effectively applied to describe changes in freshwater systems, use in tidally influenced systems has not been as extensive.
@@
 Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.
-Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.
+Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional County Treatment Facility (hereafter Sacramento \ac{wwtp}) \citep{Jassby08}.
 Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.
@@
 Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.
-Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate.
+Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to add nitrification to convert biologically available ammonium to nitrate.
 The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream.
@@
 However, a reversal of trends in recent years may also be occuring in Suisun Bay, as model estimates suggest either relatively constant concentrations or even a decrease at some stations beginning around 2000 (e.g., D6, D7, \cref{fig:trndcomp1}).
-Combined with the shift towards a dominant spring peak in chloropyll growth (\cref{fig:clmchl}f), changing nitrogen ratios continue to be a concern for the management of production in the upper \ac{sfe}.
+At the same time, the modeled results show a shift towards a dominant spring peak in chloropyll growth (\cref{fig:clmchl}f).
 
@@
 These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanisms demonstrated by \ac{wrtds} provide additonal quantitative evidence of factors that drive water quality in the Delta.
+
+\subsection{Application of \ac{wrtds} to estuarine systems}
+
+Use of \ac{wrtds} to evaluate long-term trends in water quality has not been extensive in estuarine systems.
+To date, \ac{wrtds} has only been applied to evaluate eutrophication trends in Tampa Bay, Florida \citep{Beck15} and the Patuxent River Estuary (Chesapeake Bay, \citealt{Beck17}).
+Unique lessons were learned from the application to these two systems, including \begin{inparaenum}[1\upshape)]
+\item use of salinity as an adequate tracer of both tidal forcing and freshwater inputs on indicators of water quality, 
+\item inclusion of quantile models as an extension to evaluate differences in the lower and upper conditional distributions of water quality changes, and 
+\item the ability to hypothesize pollutant sources with changes in the response of water quality to flow over time. 
+\end{inparaenum}
+The application of \ac{wrtds} to monitoring data from the Delta is the first to apply the method to a large temperate estuary on the Pacific coast.
+
+New challenges were encountered related to the geographic location and the unique characteristics of \ac{sfe} that furthered our understanding of how the method can be used for trend analysis in tidal waters.
+The \ac{sfe} is a macrotidal system that has daily variation in water depth typical of coastal systems on the Pacific Coast (i.e., $>$ 1 m).
+Previous applications of \ac{wrtds} in tidal waters were limited to systems with substantially smaller tidal amplitudes (i.e., $<$ 1 m), demonstrating that \ac{wrtds} can be applied in systems with a wide range of tidal influences.
+This point is particularly relevant given that hydrologic factors are often key determinants of water quality changes in coastal systems, suggesting the method is robust in the presence of large tidal swings.
+The \ac{sfe} is also unique in that excess turbidity, rather than nutrients, is the dominant limiting factor of primary production.
+This differs substantially from Tampa Bay and Chesapeake Bay, neither of which are light-limited from sedimentation.
+Our case study in Suisun Bay provided a unique demonstration of \ac{wrtds} to describe chlorophyll dynamics as a function of multiple factors unrelated to nutrients by evaluating the effects of biological invasion on water quality and the complex interaction of flow on both.
+This result provided the first application of \ac{wrtds} in tidal waters to evaluate multiple factors that are included explicitly in the model (e.g, flow) and those that are not (e.g., biological forcing).
+As such, \ac{wrtds} has the potential to inform the understanding of nutrient and phytoplankton dynamics in coastal systems that vary greatly in physical, chemical, and biological characteristics.
 
@@
   mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      paste0('(', perchg, ')' ), 
+    med = format(round(med, 1), nsmall = 1), 
+    perchg = format(round(perchg, 1), nsmall = 1),
+    perchg = gsub('^\\s+', '', perchg),
+    perchg = ifelse(grepl('\\-', perchg),
+      paste0('(', perchg, ')'),
       paste0('(\\textbf{', perchg, '})')
@@
   mutate(
-    med = round(med, 1), 
-    perchg = round(perchg, 1),
-    perchg = ifelse(
-      perchg < 0, 
-      perchg, 
-      paste0('\\textbf{', perchg, '}')
+    med = format(round(med, 1), nsmall = 1), 
+    perchg = format(round(perchg, 1), nsmall = 1),
+    perchg = gsub('^\\s+', '', perchg),
+    perchg = ifelse(grepl('\\-', perchg),
+      paste0('(', perchg, ')'),
+      paste0('(\\textbf{', perchg, '})')
       ),

### 2018-05-01 sftrends_manu.tex
ECSS initial revision done

@@
   \vspace{0.1in}
-  \\ Version Date:   Thu Apr 19 17:53:28 2018 -0700
+  \\ Version Date:   Fri Apr 20 16:59:41 2018 -0700
 	}
@@
 ~~MD10&0.4 \footnotesize{(-1.1)*}&0.3 \footnotesize{(-2.4)*}\tabularnewline
-~~P8&1.3 \footnotesize{(\textbf{2.5})*}&1.7 \footnotesize{(-2)*}\tabularnewline
+~~P8&1.3 \footnotesize{(\textbf{2.5})*}&1.7 \footnotesize{(-2.0)*}\tabularnewline
 \hline
 {\bfseries NH$_{4}^{+}$}&&\tabularnewline
-~~C10&0.1 \footnotesize{(-3.4)*}&0 \footnotesize{(-5.2)*}\tabularnewline
-~~C3&0.2 \footnotesize{(\textbf{3.7})*}&0.3 \footnotesize{(\textbf{0})}\tabularnewline
-~~D19&0 \footnotesize{(\textbf{0.4})*}&0 \footnotesize{(-1.7)*}\tabularnewline
+~~C10&0.1 \footnotesize{(-3.4)*}&0.0 \footnotesize{(-5.2)*}\tabularnewline
+~~C3&0.2 \footnotesize{(\textbf{3.7})*}&0.3 \footnotesize{(\textbf{0.0})}\tabularnewline
+~~D19&0.0 \footnotesize{(\textbf{0.4})*}&0.0 \footnotesize{(-1.7)*}\tabularnewline
 ~~D26&0.1 \footnotesize{(\textbf{2.2})*}&0.1 \footnotesize{(-1.5)*}\tabularnewline
-~~D28&0 \footnotesize{(-1.1)*}&0 \footnotesize{(-1.4)*}\tabularnewline
-~~D4&0 \footnotesize{(\textbf{0.9})*}&0.1 \footnotesize{(\textbf{0})}\tabularnewline
+~~D28&0.0 \footnotesize{(-1.1)*}&0.0 \footnotesize{(-1.4)*}\tabularnewline
+~~D4&0.0 \footnotesize{(\textbf{0.9})*}&0.1 \footnotesize{(\textbf{0.0})}\tabularnewline
 ~~D6&0.1 \footnotesize{(\textbf{2.4})*}&0.1 \footnotesize{(-0.5)*}\tabularnewline
 ~~D7&0.1 \footnotesize{(\textbf{1.5})*}&0.1 \footnotesize{(-1.2)*}\tabularnewline
-~~MD10&0.1 \footnotesize{(-2.8)*}&0 \footnotesize{(-1.1)*}\tabularnewline
+~~MD10&0.1 \footnotesize{(-2.8)*}&0.0 \footnotesize{(-1.1)*}\tabularnewline
 ~~P8&0.2 \footnotesize{(\textbf{4.9})*}&0.1 \footnotesize{(-10.3)*}\tabularnewline
@@
 {\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&\tabularnewline
-~~C10&1.2 \footnotesize{(\textbf{1.4})*}&1.4 \footnotesize{(-3)*}\tabularnewline
+~~C10&1.2 \footnotesize{(\textbf{1.4})*}&1.4 \footnotesize{(-3.0)*}\tabularnewline
 ~~C3&0.1 \footnotesize{(-0.1)*}&0.2 \footnotesize{(\textbf{0.7})*}\tabularnewline
 ~~D19&0.4 \footnotesize{(-0.1)*}&0.4 \footnotesize{(-1.9)*}\tabularnewline
-~~D26&0.3 \footnotesize{(\textbf{0})}&0.4 \footnotesize{(-1.1)*}\tabularnewline
+~~D26&0.3 \footnotesize{(\textbf{0.0})}&0.4 \footnotesize{(-1.1)*}\tabularnewline
 ~~D28&0.4 \footnotesize{(-0.2)*}&0.4 \footnotesize{(-3.1)*}\tabularnewline
@@
 ~~D7&0.4 \footnotesize{(\textbf{0.7})*}&0.4 \footnotesize{(-0.7)*}\tabularnewline
-~~MD10&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(-2.5)*}\tabularnewline
+~~MD10&0.4 \footnotesize{(-1.0)*}&0.3 \footnotesize{(-2.5)*}\tabularnewline
 ~~P8&1.2 \footnotesize{(\textbf{1.7})*}&1.5 \footnotesize{(-0.6)*}\tabularnewline
@@
 {\bfseries Annual}&&&&&\tabularnewline
-~~1976-2006&1.3&\textbf{2}*&&0.2&\textbf{2.8}*\tabularnewline
-~~2007-2013&1.4&-1.9*&&0.1&-16.6*\tabularnewline
+~~1976-2006&1.3&(\textbf{2.0})*&&0.2&(\textbf{2.8})*\tabularnewline
+~~2007-2013&1.4&(-1.9)*&&0.1&(-16.6)*\tabularnewline
 \hline
 {\bfseries Seasonal, pre}&&&&&\tabularnewline
-~~Spring&1.2&\textbf{1.6}*&&0.2&\textbf{1.4}*\tabularnewline
-~~Summer&1&\textbf{2.4}*&&0.1&\textbf{3.3}*\tabularnewline
-~~Fall&1.3&\textbf{2.2}*&&0.2&\textbf{4.9}*\tabularnewline
-~~Winter&1.5&\textbf{2.1}*&&0.7&\textbf{4.8}*\tabularnewline
+~~Spring&1.2&(\textbf{1.6})*&&0.2&(\textbf{1.4})*\tabularnewline
+~~Summer&1.0&(\textbf{2.4})*&&0.1&(\textbf{3.3})*\tabularnewline
+~~Fall&1.3&(\textbf{2.2})*&&0.2&(\textbf{4.9})*\tabularnewline
+~~Winter&1.5&(\textbf{2.1})*&&0.7&(\textbf{4.8})*\tabularnewline
 \hline
 {\bfseries Seasonal, post}&&&&&\tabularnewline
-~~Spring&1.3&-1.6*&&0.1&-16.2*\tabularnewline
-~~Summer&0.9&-4.3*&&0.1&-15.7*\tabularnewline
-~~Fall&1.5&-1.7*&&0.1&-19.3*\tabularnewline
-~~Winter&2.2&-0.8*&&0.2&-26.7*\tabularnewline
+~~Spring&1.3&(-1.6)*&&0.1&(-16.2)*\tabularnewline
+~~Summer&0.9&(-4.3)*&&0.1&(-15.7)*\tabularnewline
+~~Fall&1.5&(-1.7)*&&0.1&(-19.3)*\tabularnewline
+~~Winter&2.2&(-0.8)*&&0.2&(-26.7)*\tabularnewline
 \hline
@@
 ~~D19&0.5 \footnotesize{(\textbf{0.3})}&0.2 \footnotesize{(\textbf{0.4})}&0.3 \footnotesize{(\textbf{0.7})*}&0.7 \footnotesize{(-0.2)}\tabularnewline
-~~D26&0.4 \footnotesize{(\textbf{0.7})*}&0.3 \footnotesize{(\textbf{0.4})*}&0.4 \footnotesize{(\textbf{1})*}&0.6 \footnotesize{(\textbf{0.3})}\tabularnewline
+~~D26&0.4 \footnotesize{(\textbf{0.7})*}&0.3 \footnotesize{(\textbf{0.4})*}&0.4 \footnotesize{(\textbf{1.0})*}&0.6 \footnotesize{(\textbf{0.3})}\tabularnewline
 ~~D28&0.5 \footnotesize{(\textbf{0.8})*}&0.2 \footnotesize{(\textbf{0.3})}&0.3 \footnotesize{(\textbf{0.5})*}&0.8 \footnotesize{(-0.3)}\tabularnewline
@@
 {\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
-~~C10&0.1 \footnotesize{(-2.3)*}&0 \footnotesize{(-6.8)*}&0.1 \footnotesize{(-7.1)*}&0.3 \footnotesize{(-1.5)*}\tabularnewline
-~~C3&0.2 \footnotesize{(\textbf{3.9})*}&0.2 \footnotesize{(\textbf{4})*}&0.3 \footnotesize{(\textbf{3.8})*}&0.2 \footnotesize{(\textbf{2.9})*}\tabularnewline
-~~D19&0.1 \footnotesize{(\textbf{0.4})*}&0 \footnotesize{(-1.7)*}&0 \footnotesize{(\textbf{1.2})*}&0.1 \footnotesize{(\textbf{2.5})*}\tabularnewline
+~~C10&0.1 \footnotesize{(-2.3)*}&0.0 \footnotesize{(-6.8)*}&0.1 \footnotesize{(-7.1)*}&0.3 \footnotesize{(-1.5)*}\tabularnewline
+~~C3&0.2 \footnotesize{(\textbf{3.9})*}&0.2 \footnotesize{(\textbf{4.0})*}&0.3 \footnotesize{(\textbf{3.8})*}&0.2 \footnotesize{(\textbf{2.9})*}\tabularnewline
+~~D19&0.1 \footnotesize{(\textbf{0.4})*}&0.0 \footnotesize{(-1.7)*}&0.0 \footnotesize{(\textbf{1.2})*}&0.1 \footnotesize{(\textbf{2.5})*}\tabularnewline
 ~~D26&0.1 \footnotesize{(\textbf{1.4})*}&0.1 \footnotesize{(\textbf{2.5})*}&0.1 \footnotesize{(\textbf{3.1})*}&0.1 \footnotesize{(\textbf{2.3})*}\tabularnewline
-~~D28&0.1 \footnotesize{(-0.5)}&0 \footnotesize{(-3.7)*}&0 \footnotesize{(-1.6)*}&0.1 \footnotesize{(\textbf{1.7})*}\tabularnewline
-~~D4&0.1 \footnotesize{(\textbf{1.7})*}&0 \footnotesize{(\textbf{1})*}&0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textbf{2})*}\tabularnewline
+~~D28&0.1 \footnotesize{(-0.5)}&0.0 \footnotesize{(-3.7)*}&0.0 \footnotesize{(-1.6)*}&0.1 \footnotesize{(\textbf{1.7})*}\tabularnewline
+~~D4&0.1 \footnotesize{(\textbf{1.7})*}&0.0 \footnotesize{(\textbf{1.0})*}&0.0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textbf{2.0})*}\tabularnewline
 ~~D6&0.1 \footnotesize{(\textbf{2.9})*}&0.1 \footnotesize{(\textbf{2.8})*}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(\textbf{2.1})*}\tabularnewline
-~~D7&0.1 \footnotesize{(\textbf{3.3})*}&0 \footnotesize{(\textbf{2})*}&0.1 \footnotesize{(-2.8)*}&0.1 \footnotesize{(\textbf{1.7})*}\tabularnewline
-~~MD10&0.1 \footnotesize{(-1.8)*}&0 \footnotesize{(-6.5)*}&0 \footnotesize{(-3.3)*}&0.2 \footnotesize{(\textbf{0.4})}\tabularnewline
-~~P8&0.2 \footnotesize{(\textbf{3.9})*}&0.1 \footnotesize{(\textbf{1.8})*}&0.2 \footnotesize{(\textbf{7})*}&0.6 \footnotesize{(\textbf{7})*}\tabularnewline
+~~D7&0.1 \footnotesize{(\textbf{3.3})*}&0.0 \footnotesize{(\textbf{2.0})*}&0.1 \footnotesize{(-2.8)*}&0.1 \footnotesize{(\textbf{1.7})*}\tabularnewline
+~~MD10&0.1 \footnotesize{(-1.8)*}&0.0 \footnotesize{(-6.5)*}&0.0 \footnotesize{(-3.3)*}&0.2 \footnotesize{(\textbf{0.4})}\tabularnewline
+~~P8&0.2 \footnotesize{(\textbf{3.9})*}&0.1 \footnotesize{(\textbf{1.8})*}&0.2 \footnotesize{(\textbf{7.0})*}&0.6 \footnotesize{(\textbf{7.0})*}\tabularnewline
 \hline
@@
 ~~C10&1.1 \footnotesize{(\textbf{1.5})*}&1.2 \footnotesize{(\textbf{0.6})*}&1.2 \footnotesize{(\textbf{1.3})*}&1.5 \footnotesize{(\textbf{1.8})*}\tabularnewline
-~~C3&0.2 \footnotesize{(\textbf{0.7})*}&0.1 \footnotesize{(-1)*}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textbf{1})*}\tabularnewline
+~~C3&0.2 \footnotesize{(\textbf{0.7})*}&0.1 \footnotesize{(-1.0)*}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textbf{1.0})*}\tabularnewline
 ~~D19&0.4 \footnotesize{(\textbf{0.4})}&0.2 \footnotesize{(-0.3)}&0.3 \footnotesize{(\textbf{0.3})}&0.6 \footnotesize{(-0.9)*}\tabularnewline
 ~~D26&0.4 \footnotesize{(\textbf{0.6})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textbf{0.3})*}&0.5 \footnotesize{(-0.3)}\tabularnewline
-~~D28&0.5 \footnotesize{(\textbf{0.7})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textbf{0.2})}&0.7 \footnotesize{(-1)*}\tabularnewline
+~~D28&0.5 \footnotesize{(\textbf{0.7})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textbf{0.2})}&0.7 \footnotesize{(-1.0)*}\tabularnewline
 ~~D4&0.3 \footnotesize{(\textbf{0.1})}&0.3 \footnotesize{(\textbf{1.4})*}&0.3 \footnotesize{(\textbf{1.1})*}&0.4 \footnotesize{(-0.8)*}\tabularnewline
-~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textbf{4.1})*}&0.3 \footnotesize{(\textbf{1.4})*}&0.4 \footnotesize{(-1)*}\tabularnewline
-~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textbf{3.4})*}&0.4 \footnotesize{(\textbf{0.4})}&0.4 \footnotesize{(-3.6)*}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textbf{4.1})*}&0.3 \footnotesize{(\textbf{1.4})*}&0.4 \footnotesize{(-1.0)*}\tabularnewline
+~~D7&0.4 \footnotesize{(-1.0)*}&0.3 \footnotesize{(\textbf{3.4})*}&0.4 \footnotesize{(\textbf{0.4})}&0.4 \footnotesize{(-3.6)*}\tabularnewline
 ~~MD10&0.5 \footnotesize{(-0.2)}&0.2 \footnotesize{(-3.6)*}&0.2 \footnotesize{(\textbf{1.5})*}&1.2 \footnotesize{(-0.5)*}\tabularnewline
-~~P8&1.2 \footnotesize{(\textbf{2})*}&0.9 \footnotesize{(\textbf{2.3})*}&1.1 \footnotesize{(\textbf{2})*}&1.4 \footnotesize{(\textbf{1})*}\tabularnewline
+~~P8&1.2 \footnotesize{(\textbf{2.0})*}&0.9 \footnotesize{(\textbf{2.3})*}&1.1 \footnotesize{(\textbf{2.0})*}&1.4 \footnotesize{(\textbf{1.0})*}\tabularnewline
 \hline
@@
 {\bfseries DIN}&&&&\tabularnewline
-~~C10&1.1 \footnotesize{(-4.1)*}&1.3 \footnotesize{(-3.1)*}&1.6 \footnotesize{(-2)*}&1.7 \footnotesize{(-3.4)*}\tabularnewline
+~~C10&1.1 \footnotesize{(-4.1)*}&1.3 \footnotesize{(-3.1)*}&1.6 \footnotesize{(-2.0)*}&1.7 \footnotesize{(-3.4)*}\tabularnewline
 ~~C3&0.5 \footnotesize{(\textbf{0.5})}&0.4 \footnotesize{(\textbf{0.1})}&0.6 \footnotesize{(-0.2)}&0.5 \footnotesize{(-0.6)*}\tabularnewline
-~~D19&0.5 \footnotesize{(-2.8)*}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.6)*}&0.7 \footnotesize{(-2.3)*}\tabularnewline
-~~D26&0.5 \footnotesize{(-1.9)*}&0.3 \footnotesize{(-1.7)*}&0.4 \footnotesize{(-1)*}&0.6 \footnotesize{(-0.8)*}\tabularnewline
-~~D28&0.5 \footnotesize{(-3)*}&0.2 \footnotesize{(-4.9)*}&0.2 \footnotesize{(-4.9)*}&0.7 \footnotesize{(-2.1)*}\tabularnewline
-~~D4&0.4 \footnotesize{(\textbf{0})}&0.4 \footnotesize{(-1)*}&0.4 \footnotesize{(-0.9)*}&0.5 \footnotesize{(\textbf{0.6})*}\tabularnewline
-~~D6&0.5 \footnotesize{(-0.2)*}&0.5 \footnotesize{(-1)*}&0.5 \footnotesize{(-0.3)*}&0.5 \footnotesize{(-0.1)}\tabularnewline
+~~D19&0.5 \footnotesize{(-2.8)*}&0.2 \footnotesize{(-1.0)*}&0.3 \footnotesize{(-1.6)*}&0.7 \footnotesize{(-2.3)*}\tabularnewline
+~~D26&0.5 \footnotesize{(-1.9)*}&0.3 \footnotesize{(-1.7)*}&0.4 \footnotesize{(-1.0)*}&0.6 \footnotesize{(-0.8)*}\tabularnewline
+~~D28&0.5 \footnotesize{(-3.0)*}&0.2 \footnotesize{(-4.9)*}&0.2 \footnotesize{(-4.9)*}&0.7 \footnotesize{(-2.1)*}\tabularnewline
+~~D4&0.4 \footnotesize{(\textbf{0.0})}&0.4 \footnotesize{(-1.0)*}&0.4 \footnotesize{(-0.9)*}&0.5 \footnotesize{(\textbf{0.6})*}\tabularnewline
+~~D6&0.5 \footnotesize{(-0.2)*}&0.5 \footnotesize{(-1.0)*}&0.5 \footnotesize{(-0.3)*}&0.5 \footnotesize{(-0.1)}\tabularnewline
 ~~D7&0.5 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-1.3)*}&0.4 \footnotesize{(-0.4)*}&0.6 \footnotesize{(-0.2)}\tabularnewline
-~~MD10&0.4 \footnotesize{(-2.3)*}&0.2 \footnotesize{(-3.7)*}&0.2 \footnotesize{(-4.4)*}&1 \footnotesize{(-1.8)*}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.3)*}&0.2 \footnotesize{(-3.7)*}&0.2 \footnotesize{(-4.4)*}&1.0 \footnotesize{(-1.8)*}\tabularnewline
 ~~P8&1.5 \footnotesize{(-1.9)*}&1.2 \footnotesize{(-3.5)*}&1.8 \footnotesize{(-2.4)*}&2.7 \footnotesize{(-2.2)*}\tabularnewline
@@
 {\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
-~~C10&0 \footnotesize{(-4.2)*}&0 \footnotesize{(-6.1)*}&0 \footnotesize{(-8.5)*}&0.1 \footnotesize{(-7.3)*}\tabularnewline
-~~C3&0.3 \footnotesize{(\textbf{1})*}&0.3 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-0.1)}\tabularnewline
-~~D19&0 \footnotesize{(-1.9)*}&0 \footnotesize{(-0.4)}&0 \footnotesize{(-2.2)*}&0.1 \footnotesize{(-1.8)*}\tabularnewline
+~~C10&0.0 \footnotesize{(-4.2)*}&0.0 \footnotesize{(-6.1)*}&0.0 \footnotesize{(-8.5)*}&0.1 \footnotesize{(-7.3)*}\tabularnewline
+~~C3&0.3 \footnotesize{(\textbf{1.0})*}&0.3 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.5)*}&0.2 \footnotesize{(-0.1)}\tabularnewline
+~~D19&0.0 \footnotesize{(-1.9)*}&0.0 \footnotesize{(-0.4)}&0.0 \footnotesize{(-2.2)*}&0.1 \footnotesize{(-1.8)*}\tabularnewline
 ~~D26&0.1 \footnotesize{(-1.2)*}&0.1 \footnotesize{(-1.3)*}&0.1 \footnotesize{(-1.9)*}&0.1 \footnotesize{(-1.4)*}\tabularnewline
-~~D28&0 \footnotesize{(-1.7)*}&0 \footnotesize{(-0.2)}&0 \footnotesize{(-2.4)*}&0.1 \footnotesize{(-3.1)*}\tabularnewline
-~~D4&0.1 \footnotesize{(\textbf{0.3})}&0 \footnotesize{(-1.3)*}&0.1 \footnotesize{(-0.3)}&0.1 \footnotesize{(\textbf{1})*}\tabularnewline
-~~D6&0.1 \footnotesize{(-0.7)*}&0.1 \footnotesize{(-1)*}&0.1 \footnotesize{(\textbf{0.3})}&0.1 \footnotesize{(-0.3)*}\tabularnewline
-~~D7&0.1 \footnotesize{(-2.2)*}&0 \footnotesize{(-2.1)*}&0.1 \footnotesize{(\textbf{1.3})*}&0.1 \footnotesize{(-0.4)*}\tabularnewline
-~~MD10&0 \footnotesize{(-1.4)*}&0 \footnotesize{(-0.1)}&0 \footnotesize{(-0.8)*}&0.1 \footnotesize{(-4.3)*}\tabularnewline
+~~D28&0.0 \footnotesize{(-1.7)*}&0.0 \footnotesize{(-0.2)}&0.0 \footnotesize{(-2.4)*}&0.1 \footnotesize{(-3.1)*}\tabularnewline
+~~D4&0.1 \footnotesize{(\textbf{0.3})}&0.0 \footnotesize{(-1.3)*}&0.1 \footnotesize{(-0.3)}&0.1 \footnotesize{(\textbf{1.0})*}\tabularnewline
+~~D6&0.1 \footnotesize{(-0.7)*}&0.1 \footnotesize{(-1.0)*}&0.1 \footnotesize{(\textbf{0.3})}&0.1 \footnotesize{(-0.3)*}\tabularnewline
+~~D7&0.1 \footnotesize{(-2.2)*}&0.0 \footnotesize{(-2.1)*}&0.1 \footnotesize{(\textbf{1.3})*}&0.1 \footnotesize{(-0.4)*}\tabularnewline
+~~MD10&0.0 \footnotesize{(-1.4)*}&0.0 \footnotesize{(-0.1)}&0.0 \footnotesize{(-0.8)*}&0.1 \footnotesize{(-4.3)*}\tabularnewline
 ~~P8&0.2 \footnotesize{(-8.7)*}&0.1 \footnotesize{(-6.3)*}&0.2 \footnotesize{(-10.4)*}&0.5 \footnotesize{(-13.1)*}\tabularnewline
@@
 ~~C3&0.2 \footnotesize{(\textbf{0.4})}&0.1 \footnotesize{(\textbf{3.1})*}&0.2 \footnotesize{(\textbf{1.7})*}&0.2 \footnotesize{(-0.4)}\tabularnewline
-~~D19&0.4 \footnotesize{(-2.9)*}&0.2 \footnotesize{(-1)*}&0.3 \footnotesize{(-1.5)*}&0.6 \footnotesize{(-2.2)*}\tabularnewline
+~~D19&0.4 \footnotesize{(-2.9)*}&0.2 \footnotesize{(-1.0)*}&0.3 \footnotesize{(-1.5)*}&0.6 \footnotesize{(-2.2)*}\tabularnewline
 ~~D26&0.4 \footnotesize{(-1.9)*}&0.2 \footnotesize{(-1.6)*}&0.3 \footnotesize{(-0.6)*}&0.5 \footnotesize{(-0.6)*}\tabularnewline
-~~D28&0.5 \footnotesize{(-3)*}&0.2 \footnotesize{(-5.4)*}&0.2 \footnotesize{(-5.2)*}&0.7 \footnotesize{(-1.7)*}\tabularnewline
-~~D4&0.3 \footnotesize{(-0.1)}&0.3 \footnotesize{(-1)*}&0.3 \footnotesize{(-1)*}&0.4 \footnotesize{(\textbf{0.4})*}\tabularnewline
-~~D6&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-1)*}&0.4 \footnotesize{(-0.4)*}&0.4 \footnotesize{(-0.1)}\tabularnewline
+~~D28&0.5 \footnotesize{(-3.0)*}&0.2 \footnotesize{(-5.4)*}&0.2 \footnotesize{(-5.2)*}&0.7 \footnotesize{(-1.7)*}\tabularnewline
+~~D4&0.3 \footnotesize{(-0.1)}&0.3 \footnotesize{(-1.0)*}&0.3 \footnotesize{(-1.0)*}&0.4 \footnotesize{(\textbf{0.4})*}\tabularnewline
+~~D6&0.4 \footnotesize{(-0.1)}&0.4 \footnotesize{(-1.0)*}&0.4 \footnotesize{(-0.4)*}&0.4 \footnotesize{(-0.1)}\tabularnewline
 ~~D7&0.4 \footnotesize{(-0.6)*}&0.4 \footnotesize{(-1.2)*}&0.4 \footnotesize{(-0.8)*}&0.4 \footnotesize{(-0.3)*}\tabularnewline
-~~MD10&0.4 \footnotesize{(-2.6)*}&0.1 \footnotesize{(-4.5)*}&0.2 \footnotesize{(-5.4)*}&1 \footnotesize{(-1.4)*}\tabularnewline
-~~P8&1.3 \footnotesize{(-1.1)*}&1.1 \footnotesize{(-3.1)*}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textbf{0})}\tabularnewline
+~~MD10&0.4 \footnotesize{(-2.6)*}&0.1 \footnotesize{(-4.5)*}&0.2 \footnotesize{(-5.4)*}&1.0 \footnotesize{(-1.4)*}\tabularnewline
+~~P8&1.3 \footnotesize{(-1.1)*}&1.1 \footnotesize{(-3.1)*}&1.6 \footnotesize{(-0.3)*}&2.2 \footnotesize{(\textbf{0.0})}\tabularnewline
 \hline
